### 2021-03-23 elsarticle-template-harv.tex
Initial Overleaf Import

%% 
%% Copyright 2007-2020 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references

\documentclass[preprint,12pt,authoryear]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

\journal{??????}

\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \affiliation for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \affiliation{organization={},
%%            addressline={}, 
%%            city={},
%%            postcode={}, 
%%            state={},
%%            country={}}
%% \fntext[label3]{}

\title{TIMES Ireland Model: Residential Sector}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \affiliation[label1]{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}
%%
%% \affiliation[label2]{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}

\author[inst1,inst2]{Jason Mc Guire}

\affiliation[inst1]{organization={Energy Policy and Modelling Group, MaREI Centre,},%Department and Organization
            addressline={Environmental Research Institute}, 
            city={Cork},
            country={Ireland}}

\affiliation[inst2]{organization={School of Engineering,},%Department and Organization
            addressline={University College Cork}, 
            city={Cork},
            country={Ireland}}

\author[inst1,inst2]{Fionn Rogan}
\author[inst1,inst2]{Olexandr Balyk}
\author[inst1,inst2]{Tomás Mac Uidhir}
\author[inst1,inst2]{Hannah Daly}
\author[inst1,inst2]{Brian Ó Gallachóir}

\begin{abstract}
%% Text of abstract
Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
\end{abstract}

%%Graphical abstract
%%\begin{graphicalabstract}
%%\includegraphics{grabs}
%%\end{graphicalabstract}

%%Research highlights
%%\begin{highlights}
%%\item Research highlight 1
%%\item Research highlight 2
%%\end{highlights}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
keyword one \sep keyword two
%% PACS codes here, in the form: \PACS code \sep code
\PACS 0000 \sep 1111
%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
\MSC 0000 \sep 1111
\end{keyword}

\end{frontmatter}

%% \linenumbers

%% main text
\section{Sample Section Title}
\label{sec:sample1}

%% For citations use: 
%%       \citet{<label>} ==> Jones et al. (2015)
%%       \citep{<label>} ==> (Jones et al., 2015)

Lorem ipsum dolor sit amet, consectetur adipiscing \citep{Fabioetal2013} elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud \citet{Blondeletal2008} exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit \citep{Blondeletal2008,FabricioLiang2013} anim id est laborum.

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum see appendix~\ref{sec:sample:appendix}.

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
\appendix

\section{Sample Appendix Section}
\label{sec:sample:appendix}
Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor section \ref{sec:sample1} incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

%% If you have bibdatabase file and want bibtex to generate the
%% bibitems, please use
%%
\bibliographystyle{elsarticle-harv} 
\bibliography{cas-refs}

%% else use the following coding to input the bibitems directly in the
%% TeX file.

% \begin{thebibliography}{00}

% %% \bibitem[Author(year)]{label}
% %% Text of bibliographic item

% \bibitem[ ()]{}

% \end{thebibliography}
\end{document}

\endinput
%%
%% End of file `elsarticle-template-harv.tex'.

### 2021-03-23 manuscript.tex
Starting

--- 2021-03-23 elsarticle-template-harv.tex
@@
 %% for the whole article with \linenumbers.
-%% \usepackage{lineno}
 
-\journal{??????}
+\usepackage{lineno}
+\usepackage{xcolor}
+\usepackage{textcomp}
+\usepackage{gensymb}
+\usepackage{multirow}
+\usepackage{makecell}
+\usepackage{float}
+\usepackage{longtable}
+\usepackage{threeparttable}
+\usepackage{biblatex}
+\addbibresource{references.bib} %Imports bibliography file
+
+\usepackage[subpreambles=true]{standalone}
+\usepackage{import}
+
+\usepackage[utf8]{inputenc}
+\usepackage[hyphens]{url}
+
+\journal{Journal X}
 
@@
 
-\author[inst1,inst2]{Jason Mc Guire}
+\author[inst1,inst2]{Jason Mc Guire\footnote{Contact: j.mcguire@ucc.ie}}
 
@@
 \author[inst1,inst2]{Olexandr Balyk}
-\author[inst1,inst2]{Tomás Mac Uidhir}
 \author[inst1,inst2]{Hannah Daly}
-\author[inst1,inst2]{Brian Ó Gallachóir}
+\author[inst1,inst2]{ \& Brian Ó Gallachóir}
 
@@
 %% Text of abstract
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Under the climate action bill, the Irish government have set legally binding targets to reduce greenhouse gas (GHG) emissions by 51\% in 2030, compared to 2018 and to achieve “net-zero” GHG emissions by 2050.
+These legally binding targets are set, so Ireland must comply with the objective of the Paris Agreement.
+Ireland and impartiular Ireland's residential sector is preforming poorly when compared to other EU countries. Ireland has the lowest renewable heat of 6.3\% \cite{2021EurostatExplorer}
+% In 2018 only 7.8% of the population of Ireland lived in flats, this is the lowest percentage in the EU-27, the second lowest was the Netherlands at 20.2% (Eurostat, 2018).
+The average EU-27 was 46% and the closest in comparison was the UK at 14.8% (Eurostat, 2018).
+Ireland has a low urbanisation percentage of 63.2%, where the European average is 74.5% and the UK is 83.4% (United Nations, Department of Economic and Social Affairs, Population Division .
+World Urbanization Prospects, 2018).
+Ireland has the 3rd largest residential living space per capita in Europe at 41.75 m2/capita, behind Denmark and Cyprus.
+The European average is 33.8 m2/capita (Average Floor Area per Capita, 2008).
+Ireland’s low urbanisation percentage has contributed to a low number of flats, and therefore a high percentage of detached, semi-detached and terrace dwellings.
+Ireland’s dwellings are heated primarily by individual systems sourced from fossil fuels and the large size of Ireland’s dwellings has compounded the challenge of Ireland’s settlement patterns on climate change obligations.
+The Energy system optimization models (ESOMs) have been used extensively in providing insights to decision makers on issues related to climate and energy policy.
+
+Reducing GHG emission in the residential sector is key..
+ five-year  carbon  budgets. 
+ Energy  systems  optimisa-tion modelling (ESOM) is a widely-used tool to inform pathways to addresslong-term energy challenges.
+This article describes a new ESOM developedto inform Ireland’s energy system decarbonisation challenge.
+The TIMES-Ireland Model (TIM) is an optimisation model of the Irish energy system,which  calculates  the  cost-optimal  fuel  and  technology  mix  to  meet  futureenergy service demands in the transport, buildings, industry and agriculturesectors,  while  respecting  constraints  in  greenhouse-gas  emissions,  primaryenergy  resources and  feasible  deployment  rates.
+TIM is  developed  to  takeinto account Ireland’s unique energy system context, including a very highpotential for offshore wind energy and the challenge of integrating this on arelatively isolated grid, a very ambitious decarbonisation target in the periodto 2030, the policy need to inform five-year carbon budgets to meet policytargets, and the challenge of decarbonising heat in the context of low buildingstock thermal efficiency and high reliance on fossil fuels.
+To that end, modelfeatures of note include “future proofing” with flexible timeslice and spatialdefinitions, with optional hourly time resolution in electricity generation and
 \end{abstract}
@@
 %% keywords here, in the form: keyword \sep keyword
-keyword one \sep keyword two
+Energy systems optimisation model (ESOM) \sep The integrated MARKAL-EFOM system (TIMES) \sep Residential Building Stock \sep Building Efficiency sep\ Renewable Heat \sep Model description 
 %% PACS codes here, in the form: \PACS code \sep code
-\PACS 0000 \sep 1111
+%%\PACS 0000 \sep 1111
 %% MSC codes here, in the form: \MSC code \sep code
 %% or \MSC[2008] code \sep code (2000 is the default)
-\MSC 0000 \sep 1111
+%% \MSC 0000 \sep 1111
 \end{keyword}
@@
 
-Lorem ipsum dolor sit amet, consectetur adipiscing \citep{Fabioetal2013} elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+%Lorem ipsum dolor sit amet, consectetur adipiscing \citep{Fabioetal2013} elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
 Ut enim ad minim veniam, quis nostrud \citet{Blondeletal2008} exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
@@
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+%Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
 Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
@@
 %%
-\bibliographystyle{elsarticle-harv} 
-\bibliography{cas-refs}
+
+\printbibliography
 
@@
 
-% \begin{thebibliography}{00}
+%%\begin{thebibliography}{00}
 
@@
 \end{document}
-
 \endinput

### 2021-03-23 manuscript.tex
Done for the day, in progress

@@
 \usepackage{biblatex}
-\addbibresource{references.bib} %Imports bibliography file
-
-\usepackage[subpreambles=true]{standalone}
-\usepackage{import}
-
-\usepackage[utf8]{inputenc}
-\usepackage[hyphens]{url}
+\bibliography{references.bib} %Imports bibliography file
 
@@
 
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{}
-%% \affiliation[label1]{organization={},
-%%             addressline={},
-%%             city={},
-%%             postcode={},
-%%             state={},
-%%             country={}}
-%%
-%% \affiliation[label2]{organization={},
-%%             addressline={},
-%%             city={},
-%%             postcode={},
-%%             state={},
-%%             country={}}
-
 \author[inst1,inst2]{Jason Mc Guire\footnote{Contact: j.mcguire@ucc.ie}}
 
-\affiliation[inst1]{organization={Energy Policy and Modelling Group, MaREI Centre,},%Department and Organization
+\affiliation[inst1]{organization={Energy Policy and Modelling Group, MaREI Centre},%Department and Organization
             addressline={Environmental Research Institute}, 
@@
 
-\affiliation[inst2]{organization={School of Engineering,},%Department and Organization
+\affiliation[inst2]{organization={School of Engineering},%Department and Organization
             addressline={University College Cork}, 
@@
 %% Text of abstract
-Under the climate action bill, the Irish government have set legally binding targets to reduce greenhouse gas (GHG) emissions by 51\% in 2030, compared to 2018 and to achieve “net-zero” GHG emissions by 2050.
-These legally binding targets are set, so Ireland must comply with the objective of the Paris Agreement.
-Ireland and impartiular Ireland's residential sector is preforming poorly when compared to other EU countries. Ireland has the lowest renewable heat of 6.3\% \cite{2021EurostatExplorer}
+Under the climate action bill, the Irish government have set legally binding targets to reduce greenhouse gas (GHG) emissions by 51\% in 2030, compared to 2018 and to achieve “net-zero” GHG emissions by 2050 \cite{2021Climate2021}.
+These legally binding targets are set to ensure Ireland complies the Paris Agreement.
+Ireland and particularly Ireland's residential sector is preforming poorly.
+Ireland has the lowest renewable heat of 6.3\% \cite{2021EurostatExplorer}, large and poor fabric houses is hindering progress
+Heat Pumps are offering a chance to change that Heat Pump renewable energy and decarbonising electricity secotr and high COPs with large rural detached settle patterns. 
 % In 2018 only 7.8% of the population of Ireland lived in flats, this is the lowest percentage in the EU-27, the second lowest was the Netherlands at 20.2% (Eurostat, 2018).
@@
 
-%% For citations use: 
-%%       \citet{<label>} ==> Jones et al. (2015)
-%%       \citep{<label>} ==> (Jones et al., 2015)
-
 %Lorem ipsum dolor sit amet, consectetur adipiscing \citep{Fabioetal2013} elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.

### 2021-05-07 manuscript.tex
Introduction in  Progress (70%) first draft

--- 2021-03-23 elsarticle-template-harv.tex
@@
 \documentclass[preprint,12pt,authoryear]{elsarticle}
-
+%\documentclass[10pt,final,journal,compsoc,a4paper]{IEEEtran}
 %% Use the option review to obtain double line spacing
@@
 \usepackage{amssymb}
+\usepackage{pgf-pie}
+\usepackage{calc}
+\usepackage{graphicx}
+\usepackage{float}
+\usepackage{anyfontsize}
+
 %% The amsthm package provides extended theorem environments
@@
 %% for the whole article with \linenumbers.
-%% \usepackage{lineno}
 
-\journal{??????}
+\addbibresource{references.bib} %Imports bibliography file
+
+\journal{Journal X}
 
@@
 
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{}
-%% \affiliation[label1]{organization={},
-%%             addressline={},
-%%             city={},
-%%             postcode={},
-%%             state={},
-%%             country={}}
-%%
-%% \affiliation[label2]{organization={},
-%%             addressline={},
-%%             city={},
-%%             postcode={},
-%%             state={},
-%%             country={}}
+\author[inst1,inst2]{Jason Mc Guire\footnote{Contact: j.mcguire@ucc.ie}}
 
-\author[inst1,inst2]{Jason Mc Guire}
-
-\affiliation[inst1]{organization={Energy Policy and Modelling Group, MaREI Centre,},%Department and Organization
+\affiliation[inst1]{organization={Energy Policy and Modelling Group, MaREI Centre},%Department and Organization
             addressline={Environmental Research Institute}, 
@@
 %% Text of abstract
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+Ireland and particularly the residential sector has underachieved in previous climate targets.
+In the EU, Ireland has the lowest renewable heat of 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
+The settlement patterns, floor area, occupancy behaviour, high reliance on fossil fuels and low thermal efficiency of Irelands housing stock has hindered climate action progress.
+Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges which provides insights to decision makers on issues related to climate and energy policy.
+The TIMES-Ireland Model (TIM) is an optimisation model of the Irish energy system,which  calculates  the  cost-optimal  fuel  and  technology  mix  to  meet  future energy service demands
+
 \end{abstract}
@@
 %% keywords here, in the form: keyword \sep keyword
-keyword one \sep keyword two
+Energy systems optimisation model (ESOM) \sep The integrated MARKAL-EFOM system (TIMES) \sep Residential Building Stock \sep Building Efficiency \sep Model description \sep Residential Energy Transition
 %% PACS codes here, in the form: \PACS code \sep code
-\PACS 0000 \sep 1111
+%%\PACS 0000 \sep 1111
 %% MSC codes here, in the form: \MSC code \sep code
 %% or \MSC[2008] code \sep code (2000 is the default)
-\MSC 0000 \sep 1111
+%% \MSC 0000 \sep 1111
 \end{keyword}
@@
 %% main text
-\section{Sample Section Title}
-\label{sec:sample1}
+\section{Introduction}
+\label{sec:Introduction}
+\subsection{Climate Targets}
+\label{Intro:Climate Targerts}
+\par % Paragraph 1 - Intro
+Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government have proposed legally binding targets to reduce greenhouse gas (GHG) emissions by 51\% in 2030 compared to 2018, and a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}.
+Ireland's national target accounts for all GHGs - both Emission Trading System (ETS) and non-ETS emissions and sets out to develop sectoral five year carbon budgets to aid progress of long-term targets.
+The Irish national targets are at the least consistent with the European Union's (EU's) updated first Nationally Determined Contribution (NDC), which increases EU's GHG reduction target from 40\% to at least 55\% by 2030 compared to 1990 \cite{SUBMISSIONSTATES}.
+Ireland's non-ETS decarbonisation pathway which is set out in the Climate Action Plan 2019 (CAP2019) compiles with EU's 40\% reduction.
+In CAP2019, Ireland's target was to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+To proportionally reflect the increased EU ambition would mean Ireland would get a new Non-ETS target of approx.
+45\%.
+While ETS emissions reduction would require a shift from a 43\% to approx.
+60\% ETS reduction in 2030 compared to 2005.
+This would still Ireland's national 2030 target is about 15\% more ambitious than the EU's updated first NDC.While this covers the GHG targets, there is also EU 2030 energy targets for transport (RES-T), heat (RES-H) and electricity (RES-E) under renewable directive (EU) 2018/2001, and energy efficiency targets under (EU) 2018/2002.
+The overview of Ireland's climate targets are shown in Fig.\ref{fig:IrelandClimatetargets}.
 
-%% For citations use: 
-%%       \citet{<label>} ==> Jones et al. (2015)
-%%       \citep{<label>} ==> (Jones et al., 2015)
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[scale=0.4]{Figures/Non-ETS-Share.png} 
+ \caption{Overview of Ireland's 2030 Climate Targets}
+ \label{fig:IrelandClimatetargets}
+\end{figure}
 
-Lorem ipsum dolor sit amet, consectetur adipiscing \citep{Fabioetal2013} elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ireland is a unique EU member state, located on the peripherals of Europe, with a the second highest GDP per captia at more than twice the average EU27 \cite{EurostatReal}, the highest proportion of agriculture Non-ETS GHG emmsions from agruclure at 34\% in 2018  \cite{EPA2020Irelands1990-2018}, and high dependency on fossil fuel imports with little electricity or gas grid interconnections.
+Under (EU) 2020/2126 Ireland has been afforded higher flexibility allowances in non-ETS targets, to compensate for some of the barriers.
+The Climate Change Advisory Council (CCAC) aims to set fair, robust and evidence-based sectoral five year carbon budgets, which comply with all of Ireland's climate targets.
+% Insert Figure about most ambitious 2030 targets -
+
+\subsection{Energy System Optimisation Models}
+\label{Intro:ESOM}
+\par % Paragraph 2 - ESOM
+A Energy System Optimisation Model (ESOM) can be a key tool in developing least cost decarbonisation pathways.
+ESOMs can take account of a large number of dynamic variables, and complex interconnections in each time period and economic sector.
+The Integrated MARKAL-EFOM System (TIMES) modelling framework is the ESOM developed in this paper.
+TIMES has been developed and maintained by ETSAP (Energy Technology Systems Analysis Program) under the aegis of the International Energy Agency (IEA), with the key focus to \textit{``establish, maintain and expand a consistent multi-country energy/economy/environment/ engineering analytical capability mainly based on TIMES"} \cite{InternationalEnergyAgency2015InternationalIA}.
+The contracting parties of ETSAP are the governments of eighteen countries and the European Commission.
+
+\par While other ESOM frameworks are available to explore Ireland's decarbonisation pathway such as  EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}.
+TIMES is the most understood and used, and because of the expertise among energy policy researchers in Ireland and the global network of ETSAP modellers, TIMES is the preferred ESOM.
+TIMES is a ``hybrid" ESOM with combines high technological detail and macroeconomic variables covering all energy sectors.
+The previous Irish TIMES model was extracted from the PET36 model (Pan European TIMES model) and was then updated with national data and assumptions, specific to Ireland.
+This previous Irish TIMES model was used extensively in the past to provide input to energy and climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Resources2015Irelands2015-2030,DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019,Deane2013TechnicalIreland,IrishGovernment2017National2017},however there was a need to design an ESOM to specifically explore Ireland's long term climate targets and a new model called TIMES Ireland Model (TIM) was built in 2021 to address this need.
+TIMES has also been used in UK \cite{Li2018IncorporatingModel,Kannan2009ModellingApproaches}, Denmark \cite{Petrovic2016ResidentialSystem,Tattini2018ImprovingModel}, Sweden \cite{Krook-Riekkola2014ModellingStudy,Aberg2011OptimisationBuildings} and China \cite{Shi2016ModellingModel} to find least cost pathways with a focus on the residential sector.While the many climate targets further add to complexity, the need for an ESOM is strengthened.
+An ESOM can provide insights to least cost energy system pathways, the energy system  transition policymakers must also consider the maximum benefits to the economy to truly optimise the transition.
+TIM can provide vital input to the discussions or further modelling required, and thus the transparency of an ESOM is critical to ensure absorptive capacity of all stakeholders.
+
+\subsection{Ireland: Residential Sector}
+\label{Intro:IreRSD}
+\par
+Under the United Nations Framework Convention on Climate Change (UNFCCC) GHG reporting guidelines, set out in ( decision 24/CP.19 ), states the 2006  Intergovernmental Panel on Climate Change (IPCC) guidelines are to be adhere to for national GHG reporting.
+Ireland’s residential sector accounted for 14.2\%of all GHGs in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+Fig.
+\ref{fig:Ireland2018GHG} shows an overview of Irelands GHG emissions, with a focus on the energy related GHG emissions.
+The residential sector accounts for 47\% of heat emissions 30\% of electricity emissions, therefore the residential sector plays a pivotal role in the energy-related GHG emissions in Ireland.
+
+\begin{figure}[h]
+    \begin{tikzpicture}[every node/.style={align=center}, pin distance=17mm]
+    \pie[cloud,text=inside, radius = 2.7]{34/Agriculture,7/Other,59/Energy-related}
+    \draw [->, line width=4pt] (5.8,-0.3) -- (6.8,-.3);
+    \hspace{9cm}
+    \pie[text=inside,radius=2.2,color={black!30,red!70,yellow!40}]{40/Transport,33/Heat,27/ Electricity}
+    \end{tikzpicture}
+    \caption{Ireland 2018 GHG Share \cite{Howley2005ENERGY-RELATEDIreland}}
+    \label{fig:Ireland2018GHG}
+\end{figure}
+
+ While the national GHG should reduce 51\% by 2030, the residential sector is expected to be reduce GHGs even further, due to the higher cost-effectiveness when compared with other sectors.
+The high ambition in residential is due to its dependency on receiving low carbon intensity electricity.
+The decarbonisation of the power generation sector is well underway, with carbon intensity falling annually to 375 gCO2/kWh in 2018, 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}.
+However the on-shore wind energy capacity is almost saturated, and policy looks to guide investment into off-shore wind to continue a reduction in carbon intensity.
+Under CAP2019 this continued growth and decarbonisation of the electricity will be the fundamental to lowering energy-related GHG in not only electricity, but heat and transport also.
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[scale=1.0]{Figures/CAP2019MACC.png} 
+ \caption{Climate Action Plan 2019 - Marginal Abatement Cost Curve}
+ \label{fig:CAP2019:MACC} \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
+\end{figure}
+
+\par
+The CAP2019 had a Marginal Abatement Cost Curve (MACC) is based on  based on Environmental Protection Agency (EPA) GHG emission projection report \cite{EPA2018EPAReport} and provides a solid analytical foundation on the most cost-effective pathway to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels.
+The MACC explores technological, behavioural and fuel switching.
+The x-axis (width of each bar chart) shows the potential reduction of annual MtCO2eq.
+emissions in 2030 from the technology or fuel switch.
+The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of CO2eq.
+over the 2021 to 2030 period.
+The columns are organised from the most economical (left side) to the most expensive technology (right side) in EUR/tCO2eq \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+In the least-cost pathway Ireland would increase Onshore wind by 8.2 GW and Offshore would grow to 3.5 GW,this would be expected to save 7 to 8 MtCO2eq.
+This would help to indirectly reduce the GHG in the residential sector.
+However specific residential solutions in the MACC include:
+\begin{itemize}
+    \item{Retrofit oil boiler existing dwellings to B2 equivalent}
+    \item{Switch from oil boilers to heat pumps in existing dwellings}
+    \item{Retrofit gas boiler and solid fuel stove existing dwellings to B2 equivalent}
+    \item{Introduce CO2-free heating in new buildings}
+\end{itemize}
+
+McKinsey and Company's modelling tools have helped develop the MACC shown in Fig.\ref{fig:CAP2019:MACC}, with more ambitious Non-ETS 2030 target due this year and a national GHG target for 2030, there is a requirement for a new MACC, from which a new residential five-year carbon budgets will arise.
+Sustainable Energy Authority of Ireland (SEAI) and others have been part of the renovation wave in Ireland, the uptake of heat pumps in existing homes is gradually building due to grants, awareness and increasing the ease of upgrades for homeowners, while new homes have a strict regulation to comply with, so heat pumps despite their high capital costs are preferred.
+The building regulations in 2021 will require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}.
+The CAP2019 plan to deliever a new retrofitting model to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030, highlighting the energy efficiency first approach in how Ireland aims to progress.
+
+\subsection{Ireland: Residential Sector Background}
+\label{Intro:IreRSDBack}
+
+\par
+Ireland’s decarbonisation pathway, is further separated from other EU member states when focusing on the residential sector, this is largely down to Ireland’s high reliance on fossil fuels and settlement patterns of low thermal efficiency of the housing stock.
+Ireland's lower than average urbanisation which is connected to only 5.8\% of the population of Ireland living in apartments in 2018, the lowest and well below the average EU-27 of 25.3\% \cite{EurostatDistributionSurveyilc_lvho01}, the lower density relates translates to more expensive mitigation.
+33\% of the population of the EU were deemed to live in dwellings too large for their needs in 2018, in Ireland the rate was 71.4\% \cite{EurostatShareSurveyilc_lvho50a}.
+Ireland set out national policy objectives under 10 strategic investment priorities \cite{Ireland2018ProjectFramework}.
+The first of strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, is a top priority".}, which will indirectly effect both electricity and transport GHG emission.
+In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved in the last 30 years, and in 2018 peat accounted for 12.3\% of residential GHG emissions.
+Residential coal GHG emissions has also declined to 6.9\% in 2018.
+However, the decline in peat and coal has been offset by more significant growth in oil boilers in rural dwellings in the same period, where oil accounts for 34.8\% of residential GHG emissions \cite{Howley2005ENERGY-RELATEDIreland}.
+Urban dwellings prefer gas as a heating fuel, which is cheaper, more secure ( 61\% of gas in 2018 was indigenous ) and less carbon intensive.
+Gas accounted for 15.5\% of residential GHG emission in 2018, significantly lower than oil due to the lower carbon intensity.
+Gas Network Ireland (GNI) anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply ) will be less than 30\% of 2018 levels \cite{Dineen2020ENERGYReport}
+In summary, the residential sector is highly reliant on fossil fuels as shown in Fig.\ref{fig:ResFuel}. 
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[scale=0.5]{Figures/ResidentialFuelUsage.png}
+ \caption{Historical Residential Fuel Consumption}
+ \label{fig:ResFuel}
+\end{figure}
+% NOTE THE Y-AXIS IS NOT CORRECT - DATA FROM SEAI ENERGY DATA PORTAL
+
+Historical fuel usage in one part of understanding residential energy, the other main part is understanding the low thermal efficiency of Irelands housing stock.
+As mentioned in section \ref{Intro:IreRSD}, Irelands new buildings produce 70\% less GHG emissions than building in 2005, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock.
+According to Central Statistics Office (CSO) approx.
+85\% of residential buildings are 2005 or older, the existing housing stock is too large to ignore in a climate transition which requires GHG emissions to more than half in the next decade.
+The BER database represents about half of the housing stock, Fig.
+\ref{fig:BERAge} highlights the amount of houses built pre-2005 and houses built in 2005 or after.
+Crucially it highlights the indication of energy consumption in $kWh/m^2/year$, which is then categorised into BER labels, with A1 representing the lowest energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest energy consumption ($ >450 kWh/m^2/year$) .
+Fig.\ref{fig:BERAge} highlights the impact the year of construction has on the expected energy consumption.
+The poor BERs consequently translates to higher energy bills to maintain comfortable temperature, but in reality it relates to low internal temperatures, which likely results in higher excessive winter mortality \cite{Clinch2000HousingMortality}.The BERs which come from the Energy Performance of Buildings Directive 2002/91/ EC, helps to better understand the housing stock.
+The BERs are required for new houses, rented houses, sold houses and renovated houses.
+However, BERs do not take account of actual usage, i.e occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
+The BER labels do therefore leave a gap between actual consumption and expected consumption, I will explain how I approached this problem in the Literature review.
+
+Fuel Poverty is when a household spends \textit{``10\% or more of income to achieve adequate warmth"}, an element energy reported 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}.
+While income supports and the \textit{Fuel Allowance and the Household Benefits} provides support to low income households to allievate some the fuel poverty.
+These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}.
+As Kerr points out the link between energy and social welfare policy is made explicitly \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty, which include increased risk of catastrophic cardiovascular or cerebrovascular events \cite{OMeara2016AIreland}.
+Ironically, this places low income households deeply dependent on fossil fuels for health.
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[scale=0.4]{Figures/BERAge.png}
+ \caption{Building Energy Rating by Age}
+ \label{fig:BERAge}
+\end{figure}
+
+\subsection{Residential Energy Services}
+\label{Intro:EnergyServices}
+
+Space heating accounted for the majority of final energy, Fig.
+\ref{fig:Ireland2016ServiceDemand} shows a snapshot of the residential energy services demands in 2018.
+As households become more thermal efficient and use higher efficient heating, space heating energy would be expected to decline.
+The new A-rated dwellings will use more energy for Pumps \& Fans, as mechanical ventilation is used over natural ventilation.
+%The energy used on appliances due to increasing information technology in households and the digitization of home appliances
+
+\begin{figure}[h]
+\centering
+    \begin{tikzpicture}
+    \tikzstyle{every node}=[font={\bf\large}]
+    \pie[   color = { yellow!90!black, green!60!black, blue!60,
+        red!70,
+        gray!80,
+        teal!20
+    }, explode = 0.4, text=legend, radius=4]{61.8/Space Heating, 2.2/Cooking, 18.5/Water Heating, 3.8/Lighting, 12.8/Appliances,0.9/Pumps \& Fans}
+    \end{tikzpicture}
+    \caption{Ireland 2018 Energy Service Demands}
+    \label{fig:Ireland2016ServiceDemands}
+\end{figure}
+
+The other service energy demands which are of importance in understanding the current residential usage, included lighting, pumps & fans, cooling, cooking and electrical appliances.
+SEAI 2018 residential energy report provides quantification on how each end use energy service demand, while also disaggregating demand based upon disposable income, location, owner/rented, BER label and age .
+While Ireland currently has a shortage of houses, it is expected that new builds will increase in the coming years, the new A-rated houses behave distinctively different which much less demand on space heating and with pumps and fans energy consumption usage increasing, while appliance end use is increasing across the BER ratings.
+The number of appliances has increased rapidly and while energy efficient improves are  reducing the energy per appliance, it is overshadowed by the growth in electrical appliances.
+There retrofitting grant which  requires building to have a HLI indication of 2 or less before a heat pump grant can be installs.
+Ireland’s aim of retrofitting 0.5 million homes from the existing 1.7 million in the before 2030.
+The effect of these targets also stretches out across the construction sector.
+The savings from retrofits will decrease fossil fuel energy usage, while most houses 00\% would need to get heat pumps too, to fulfil 0.4 million existing houses getting heat pumps.
+This energy efficiency first approach would allow for future technology choices, when current boiler are not at end of life.
+
+\subsection{Behavioural and Financial barriers}
+\label{Intro:barriers}
+\par
+In an attempt to influence fossil fuel demand, the Irish government introduced Carbon tax in 2010 XX.
+The current Carbon tax is €33.50 and this will rise ti €100/tCO2 in 2030.
+The tax applies to all fossil fuel consumption across different economic sectors and while many European countries have taken this approach, as Sweden lead by example and succefully implanted a Carbon tax in 1993??.
+Current carbon tax?
+Swedens GNP??
+Or PPP.
+The difficulty in carbon tax is that it can widen social inequality, whereby the lower income fossil fuelled homes who cannot afford upgrade and may already be in fuel poverty have to pay more on income on fuel and middle or high income homes can avoid the rising costs, by investing in alternative low carbon heating.
+The slow increase does afford time for lower income houses but with only expensive option currently available, some may have no option but to keep paying and receiving fuel subsidies.
+While the electricity network stretches to all corners of Ireland, The gas network is only reaches to 680,000 houses in towns and cities, with biomethance and renewable gas targets for 2030, X%.
+The district heating networks are new and currently only two in Dublin which reach about XX houses, 2030 target X%.
+large amount of data centres.
+Other future heating fuel options such as hydrogen, HVO, solar may also play a part.
+	ESOM can be used to start the conversation on..
+Look at other ESOM, brief explain…
+Other residential sectors..
+
+\subsection{TIMES Ireland Model (TIM)}
+\label{Intro:TIM}
+\par
+TIM was built to help Ireland provide a pathway to achieve 2030 and 2050 targets.
+The residential sector plays a key part of this and it was built from the SEAI 2018 energy balance, the BER database and CSO data.
+The data filtration process used in the residential sector, derived a lot of conversation of the BER database, a group know as Ireland Energy Wiki by the National Retrofitting Modelling Group grew from building the residential sector, questions about the filters, default values, assumptions around thermal equations (EC..) and uncertainties were used to improve the data before it was used in TIM .
+TIM has many internal project deadines, which required extensive meetings about the data, constraints and results as the modelling was progressing.
+The data and results were presented to 30 external reviewers to provide feedback, this process proved invaluable having different expertise with different presepectives enabled TIM to fill in gaps.
+TIM will also be open platform to help improve the model and provide transparency.
+TIM’s residential sector was built in parallel with LEAP’s residential sector, so that residential sector can easily simulate policy.
+The demand for results from TIM has increased significantly with increased government ambition, TIM will provide input into CAP 2021 and Ireland’s carbon budgets, working with DECC and the CCAC respectively will also provide feedback and improve the robustness of TIM.
+for these reasons TIM residential sector is a state of the art modelling platform.
+
+\section{Literature Review}
+\label{LitR}
+``How should Ireland set GHG targets in the residential sector?” 
+ % LIT REW INPUT DATA - EU \cite{Simoes2013TheProject}
+
+%Lorem ipsum dolor sit amet, consectetur adipiscing \citep{Fabioetal2013} elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
 Ut enim ad minim veniam, quis nostrud \citet{Blondeletal2008} exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
@@
 %% appendix sections are then done as normal sections
+\section{Acknowledgements}
+\label{sec:Acknowledgements}
+%CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC
+% https://www.marei.ie/project/capacity/
+CHIMERA is supported by a research grant from Science Foundation Ireland (SFI) and the National Natural Science Foundation of China (NSFC) under the SFI-NSFC Partnership Programme, grant no.
+17/NSFC/5181 and
+supported by MaREI, the SFI Research Centre for Energy, Climate, and Marine [Grant No: 12/RC/2302\_P2]
+
+https://www.marei.ie/project/chimera/
 \appendix
@@
 \label{sec:sample:appendix}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor section \ref{sec:sample1} incididunt ut labore et dolore magna aliqua.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor section \ref{sec:Introduction} incididunt ut labore et dolore magna aliqua.
 Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.

### 2021-05-10 manuscript.tex
Introduction - Financial Barriers Start

@@
 \usepackage{anyfontsize}
+\usepackage{pgfplots}
 
@@
 Gas accounted for 15.5\% of residential GHG emission in 2018, significantly lower than oil due to the lower carbon intensity.
-Gas Network Ireland (GNI) anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply ) will be less than 30\% of 2018 levels \cite{Dineen2020ENERGYReport}
+Gas Network Ireland (GNI) anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}
 In summary, the residential sector is highly reliant on fossil fuels as shown in Fig.\ref{fig:ResFuel}.
+
 \begin{figure}[!htbp]
@@
 According to Central Statistics Office (CSO) approx.
-85\% of residential buildings are 2005 or older, the existing housing stock is too large to ignore in a climate transition which requires GHG emissions to more than half in the next decade.
-The BER database represents about half of the housing stock, Fig.
-\ref{fig:BERAge} highlights the amount of houses built pre-2005 and houses built in 2005 or after.
-Crucially it highlights the indication of energy consumption in $kWh/m^2/year$, which is then categorised into BER labels, with A1 representing the lowest energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest energy consumption ($ >450 kWh/m^2/year$) .
-Fig.\ref{fig:BERAge} highlights the impact the year of construction has on the expected energy consumption.
+85\% of residential buildings are 2005 or older.
+The BER database, which represents half of the dwelling stock highlights the impact the year of construction has on the expected energy consumption, as shown in Fig.\ref{fig:BERAge}.The BERs are categorised into BER labels, with A1 representing the lowest energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest energy consumption ($ >450 kWh/m^2/year$) .
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[scale=0.4]{Figures/BERAge.png}
+ \caption{Building Energy Rating by Age}
+ \label{fig:BERAge}
+\end{figure}
+
 The poor BERs consequently translates to higher energy bills to maintain comfortable temperature, but in reality it relates to low internal temperatures, which likely results in higher excessive winter mortality \cite{Clinch2000HousingMortality}.The BERs which come from the Energy Performance of Buildings Directive 2002/91/ EC, helps to better understand the housing stock.
@@
 
-\begin{figure}[!htbp]
- \centering
- \includegraphics[scale=0.4]{Figures/BERAge.png}
- \caption{Building Energy Rating by Age}
- \label{fig:BERAge}
-\end{figure}
-
 \subsection{Residential Energy Services}
@@
 Space heating accounted for the majority of final energy, Fig.
-\ref{fig:Ireland2016ServiceDemand} shows a snapshot of the residential energy services demands in 2018.
+\ref{fig:Ireland2018ServiceDemands} shows a snapshot of the residential energy services demands in 2018.
 As households become more thermal efficient and use higher efficient heating, space heating energy would be expected to decline.
-The new A-rated dwellings will use more energy for Pumps \& Fans, as mechanical ventilation is used over natural ventilation.
-%The energy used on appliances due to increasing information technology in households and the digitization of home appliances
+The new A-rated dwellings will use less energy for space heating, but Pumps \& Fans energy will increase as mechanical ventilation is used over natural ventilation.
+
+\begin{figure}[!htbp]
+\centering
+    \begin{tikzpicture}
+    \tikzstyle{every node}=[font={\bf}]
+    \pie[   color = { yellow!90!black, green!60!black, blue!60,
+        red!70,
+        gray!80,
+        teal!20
+    }, explode = 0.4, text=legend, radius=3]{61.8/Space Heating, 2.2/Cooking, 18.5/Water Heating, 3.8/Lighting, 12.8/Appliances,0.9/Pumps \& Fans}
+    \end{tikzpicture}
+    \caption{Ireland 2018 Energy Service Demands}
+    \label{fig:Ireland2018ServiceDemands}
+\end{figure}
+
+Space heating is therefore central to residential energy policy, other important service energy demands are shown in Fig.\ref{fig:Ireland2018ServiceDemands}.
+Notably space cooling is absent for the pie chart, but space cooling is accounted for in appliances.
+The Policy Oriented Tool for Energy and Climate Change Impact Assessment (POTEnCIA) project disaggregated Ireland's residential space cooling from appliances and have output space cooling energy consumption data, shown in Fig.\ref{fig:Cooling} a significant amount of space cooling energy is expected in the Ireland residential sector.
 
@@
 \begin{tikzpicture}
-    \tikzstyle{every node}=[font={\bf\large}]
-    \pie[   color = { yellow!90!black, green!60!black, blue!60,
-        red!70,
-        gray!80,
-        teal!20
-    }, explode = 0.4, text=legend, radius=4]{61.8/Space Heating, 2.2/Cooking, 18.5/Water Heating, 3.8/Lighting, 12.8/Appliances,0.9/Pumps \& Fans}
+\begin{axis}[
+    x tick label style={
+		/pgf/number format/1000 sep=},xlabel={Year},
+	y tick label style={
+	    /pgf/number},
+		ylabel={Energy Consumption($ktoe$)},
+		 scaled y ticks = false,
+    xmin=2000, xmax=2050,
+    ymin=0, ymax=40,
+    xtick={2000,2010,2020,2030,2040,2050},
+    ytick={0,10,20,30,40},
+    ymajorgrids=true,
+]
+\addplot[
+    color=blue,
+    mark=square,
+    ]
+    coordinates {
+    (2000,0.6)(2010,3.1)(2020,8.4)(2030,16.3)(2040,26.1)(2050,37)
+    };
+\end{axis}
 \end{tikzpicture}
-    \caption{Ireland 2018 Energy Service Demands}
-    \label{fig:Ireland2016ServiceDemands}
+    \caption{Space Cooling Residential Energy Consumption (Source: POTEnCIA) }
+    \label{fig:Cooling}
 \end{figure}
 
-The other service energy demands which are of importance in understanding the current residential usage, included lighting, pumps & fans, cooling, cooking and electrical appliances.
-SEAI 2018 residential energy report provides quantification on how each end use energy service demand, while also disaggregating demand based upon disposable income, location, owner/rented, BER label and age .
-While Ireland currently has a shortage of houses, it is expected that new builds will increase in the coming years, the new A-rated houses behave distinctively different which much less demand on space heating and with pumps and fans energy consumption usage increasing, while appliance end use is increasing across the BER ratings.
-The number of appliances has increased rapidly and while energy efficient improves are  reducing the energy per appliance, it is overshadowed by the growth in electrical appliances.
-There retrofitting grant which  requires building to have a HLI indication of 2 or less before a heat pump grant can be installs.
-Ireland’s aim of retrofitting 0.5 million homes from the existing 1.7 million in the before 2030.
-The effect of these targets also stretches out across the construction sector.
-The savings from retrofits will decrease fossil fuel energy usage, while most houses 00\% would need to get heat pumps too, to fulfil 0.4 million existing houses getting heat pumps.
-This energy efficiency first approach would allow for future technology choices, when current boiler are not at end of life.
+The SEAI residential energy report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} provides high resolution detail on energy demands in the residential sector.
+Variables such as disposable income, appliance ownership, dwelling location, type of occupier status (owner/rented ), BER label and age of dwelling.
 
@@
 \par
-In an attempt to influence fossil fuel demand, the Irish government introduced Carbon tax in 2010 XX.
+In Ireland, the carbon tax was introduced in 2009 on petrol, diesel, oils and gas, and was extended to solid fuels (coal and peat) in 2013 and 2014.
+The rate per tonne of CO2 emissions was €15 from 2010 to 2011 and was
+increased to €20 per tonne of CO2 emissions in 2012.
+It applies exclusively on emissions in non-ETS sectors.
+As shown in Figure 1, Carbon Tax receipts from 2010 to
+2018 grew every year, reflecting strong economic growth,
+and amounted to €431 million in 2018. Auto Diesel
+accounts for 42% of carbon tax receipts, which is also
+reflective of the current composition of the vehicle fleet
+in Ireland with 60% of vehicles operating on diesel. It is
+interesting to note that unlike carbon tax receipts from
+The Irish government introduced Carbon tax in 2010, to influence fossil fuel demand, XX.
 The current Carbon tax is €33.50 and this will rise ti €100/tCO2 in 2030.
@@
  % LIT REW INPUT DATA - EU \cite{Simoes2013TheProject}
+
+%the new A-rated houses behave distinctively different which much less demand on space heating and with pumps and fans energy consumption usage increasing, while appliance end use is increasing across the BER ratings.
+The number of appliances has increased rapidly and while energy efficient improves are  reducing the energy per appliance, it is overshadowed by the growth in electrical appliances.
+There retrofitting grant which  requires building to have a HLI indication of 2 or less before a heat pump grant can be installs.
+Ireland’s aim of retrofitting 0.5 million homes from the existing 1.7 million in the before 2030.
+The effect of these targets also stretches out across the construction sector.
+The savings from retrofits will decrease fossil fuel energy usage, while most houses 00\% would need to get heat pumps too, to fulfil 0.4 million existing houses getting heat pumps.
+This energy efficiency first approach would allow for future technology choices, when current boiler are not at end of life.
 

### 2021-05-13 manuscript.tex
13 May 2021 - Update finishing Intro 1st draft

@@
 
+%Preliminary emissions data published by the EPA in January 2021 indicate that emissions from the residential sector increased by 9% in 2020, primarily due to the increase in home working and Covid-related restrictions on movement forcing people to spend more time at home.
 \begin{figure}[!htbp]
@@
 
-\subsection{Behavioural and Financial barriers}
-\label{Intro:barriers}
+\subsection{Current Energy Policy}
+\label{Intro:Policy}
+The 2019 Climate Action Plan identified 29 actions that would reduce emissions from the sector by 40-45%.  These actions included:
+Reducing fossil fuel use in the sector and moving to electricity to provide heating and hot water in buildings
+500,000 building retrofits to a BER B2 / cost optimal equivalent or carbon equivalent
+Installation of 600,000 heat pumps, with 400,000 of these in existing buildings
+Development of district heating, including two initiatives of municipal scale with the potential to provide heat equivalent to the needs of about 50,000 homes
+Complete the rollout of the Support Scheme for Renewable Heat
+Increasing the number of Sustainable Energy Communities to 1,500
+
+\subsubsection{Electricity}
 \par
-In Ireland, the carbon tax was introduced in 2009 on petrol, diesel, oils and gas, and was extended to solid fuels (coal and peat) in 2013 and 2014.
-The rate per tonne of CO2 emissions was €15 from 2010 to 2011 and was
-increased to €20 per tonne of CO2 emissions in 2012.
-It applies exclusively on emissions in non-ETS sectors.
-As shown in Figure 1, Carbon Tax receipts from 2010 to
-2018 grew every year, reflecting strong economic growth,
-and amounted to €431 million in 2018. Auto Diesel
-accounts for 42% of carbon tax receipts, which is also
-reflective of the current composition of the vehicle fleet
-in Ireland with 60% of vehicles operating on diesel. It is
-interesting to note that unlike carbon tax receipts from
-The Irish government introduced Carbon tax in 2010, to influence fossil fuel demand, XX.
-The current Carbon tax is €33.50 and this will rise ti €100/tCO2 in 2030.
-The tax applies to all fossil fuel consumption across different economic sectors and while many European countries have taken this approach, as Sweden lead by example and succefully implanted a Carbon tax in 1993??.
-Current carbon tax?
-Swedens GNP??
-Or PPP.
-The difficulty in carbon tax is that it can widen social inequality, whereby the lower income fossil fuelled homes who cannot afford upgrade and may already be in fuel poverty have to pay more on income on fuel and middle or high income homes can avoid the rising costs, by investing in alternative low carbon heating.
-The slow increase does afford time for lower income houses but with only expensive option currently available, some may have no option but to keep paying and receiving fuel subsidies.
-While the electricity network stretches to all corners of Ireland, The gas network is only reaches to 680,000 houses in towns and cities, with biomethance and renewable gas targets for 2030, X%.
+Decarbonising Electricity is key for Ireland, both the \textit{``stick"} (tax) approach which aims to deter investment in fossil fuel power generation, and the \textit{``carrot"} (subsidies) approach which aims to encourage renewable electricity is applied in Irish policy.
+In Ireland, a €15/tCO2 carbon tax was introduced in 2009 to influence fossil fuel demand.
+The carbon tax receipts grew every year from 2010 to 2018 \cite{AnOffice}.
+Ireland's current carbon tax of €33.50/tCO2 will rise to €100/tCO2 in 2030.
+Many countries have taken this approach, especially in Europe, where Finland was the first country to introduce a carbon tax in 1990, while Sweden were second to introduce a carbon tax after Finland, the carbon tax rate is the highest in the world at €108.81/tCO2.
+A carbon price or tax now accounts for approx.
+15\% of GHG emissions \cite{WorldBankGroup2020State2020}.
+The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentives investment in renewable electricity to achieve legally binding renewable targets under 2009/28/EC.
+REFIT operates  by  guaranteeing  new renewable generation and biomass a minimum price for electricity over a 15 year period.
+The last REFIT schemes closed in December 2015.
+In 2020 REFIT was replaced by the Renewable Electricity Support Scheme (RESS), which supports to renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target.
+RESS delivers for communities to participate in renewable energy projects and increasing technology diversity by broadening the renewable electricity technology mix, where almost 800 MW of solar has been approved in RESS-1 \cite{Eirgrid2020RenewableResults}, this builds alot on the current solar capacity of 16 MW, while making significant progress towards the CAP2019 target of 1500 MW by 2030.
+
+\subsubsection{Heat}
+Domestic supports for renewable heat or energy efficiency is through SEAI grants,\begin{itemize}
+    \item {Insulation}
+    \subitem {Attic}
+    \subitem {Cavity Wall}
+    \subitem {Internal - Dry Lining}
+    \subitem {External}
+    \item {Heat Pumps}
+    \subitem {Air to Water}
+    \subitem {Ground Source to Water}
+    \subitem {Exhaust Air to Water}
+    \subitem {Water to Water}
+    \subitem {Air to Air}
+    \item {Heating Controls}
+    \item {Solar Water Heating}
+\end{itemize}
+
+This report is a study of potential design options for a Renewable Heat Incentive (RHI) f
+
+-  critical for the development of renewable gas projects in Ireland.European Commission’s decision on State aid for the Support Scheme for Renewable Heat (SA.50807).
+GNI Vision 2050 - 20\% renewable gas by 2030
+While the electricity network reaches all residential dwellings, The gas network reaches 40\% or 680,000 dwellings in towns and cities. The 2030 gas decarbonisation targets, set by GNI  
+and  with biomethance and renewable gas targets for 2030, X\%.
 The district heating networks are new and currently only two in Dublin which reach about XX houses, 2030 target X%.
@@
 Other future heating fuel options such as hydrogen, HVO, solar may also play a part.
-	ESOM can be used to start the conversation on..
-Look at other ESOM, brief explain…
-Other residential sectors..
+Gas Networks Ireland (GNI) has a strategic plan to achieve 20% Renewable Gas on the network by 2030 
+Further investigate decarbonisation options such as hydrogen vehicles, biomethane and AD
+substitutes for natural gas - See \ref{fig:CAP2019:MACC} shows biomethane as the most expensive options..
 
@@
 for these reasons TIM residential sector is a state of the art modelling platform.
-
+%GRAPH OF IRELAND DECARBONISATION TARGETS  VS OTHER COUNTRIES - DIFFERENT INTRASTATE and RESOURCE AT BIDENS ENERGY SUBMIT - DENMAKR TOP 2018, 
+	ESOM can be used to start the conversation on..
+Look at other ESOM, brief explain…
+Other residential sectors..
+CAP2019 - Choosing the Pathways which Create the Least Burden
+and Offer the Most Opportunity for Ireland
 \section{Literature Review}

### 2021-05-15 manuscript.tex
Update from Friday

@@
 
-\journal{Journal X}
+\journal{Energy and Buildings}
 
@@
 
-Ireland and particularly the residential sector has underachieved in previous climate targets.
-In the EU, Ireland has the lowest renewable heat of 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
-The settlement patterns, floor area, occupancy behaviour, high reliance on fossil fuels and low thermal efficiency of Irelands housing stock has hindered climate action progress.
-Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges which provides insights to decision makers on issues related to climate and energy policy.
-The TIMES-Ireland Model (TIM) is an optimisation model of the Irish energy system,which  calculates  the  cost-optimal  fuel  and  technology  mix  to  meet  future energy service demands
+Write after paper is completed %Ireland's residential sector has particularly underachieved in previous climate targets.
+Located on the peripherals of Europe, with a high share of low thermal efficient detached dwellings with carbon intensive heating fuels, provides some distinctive barriers to decarbonising Ireland's residential sector.
+The disadvantageous starting position has impeded residential decarbonisation progress thus far.
+\par
+%Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges which provides insights to decision makers on issues related to climate and energy policy.
+\par
+%The TIMES-Ireland Model (TIM) is a newly developed optimisation model of the Irish energy system, which calculates the  cost-optimal decarbonisation pathway to meet future energy service demands while achieving Ireland's legally binding 2030 and 2050 climate targets.
 
@@
 %% keywords here, in the form: keyword \sep keyword
-Energy systems optimisation model (ESOM) \sep The integrated MARKAL-EFOM system (TIMES) \sep Residential Building Stock \sep Building Efficiency \sep Model description \sep Residential Energy Transition
+Energy systems optimisation model (ESOM) \sep The integrated MARKAL-EFOM system (TIMES) \sep Building decarbonisation \sep Model description \sep Residential energy transition
 %% PACS codes here, in the form: \PACS code \sep code
@@
 \par % Paragraph 1 - Intro
-Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government have proposed legally binding targets to reduce greenhouse gas (GHG) emissions by 51\% in 2030 compared to 2018, and a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}.
-Ireland's national target accounts for all GHGs - both Emission Trading System (ETS) and non-ETS emissions and sets out to develop sectoral five year carbon budgets to aid progress of long-term targets.
-The Irish national targets are at the least consistent with the European Union's (EU's) updated first Nationally Determined Contribution (NDC), which increases EU's GHG reduction target from 40\% to at least 55\% by 2030 compared to 1990 \cite{SUBMISSIONSTATES}.
-Ireland's non-ETS decarbonisation pathway which is set out in the Climate Action Plan 2019 (CAP2019) compiles with EU's 40\% reduction.
-In CAP2019, Ireland's target was to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-To proportionally reflect the increased EU ambition would mean Ireland would get a new Non-ETS target of approx.
-45\%.
-While ETS emissions reduction would require a shift from a 43\% to approx.
-60\% ETS reduction in 2030 compared to 2005.
-This would still Ireland's national 2030 target is about 15\% more ambitious than the EU's updated first NDC.While this covers the GHG targets, there is also EU 2030 energy targets for transport (RES-T), heat (RES-H) and electricity (RES-E) under renewable directive (EU) 2018/2001, and energy efficiency targets under (EU) 2018/2002.
-The overview of Ireland's climate targets are shown in Fig.\ref{fig:IrelandClimatetargets}.
+Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to reduce Greenhouse Gas (GHG) emissions by 51\% in 2030 compared to 2018.
+The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}.
+These targets accounts for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to develop sectoral five year carbon budgets to aid progress of the forementioned long-term targets.
+The EU's updated first Nationally Determined Contribution (NDC), increases EU's GHG reduction target from 40\% (first NDC) to at least 55\% (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}.
+Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+Climate Action Plan 2021 will have to be more ambitious than its predecessor, as it must at least reflect EU's updated NDC target.
+Assuming the share of contribution of ETS and Ireland's Non-ETS remains the same, to met the EU's updated NDC, Ireland's non-ETS GHG emissions target would change from 30\% to 45\% reduction in 2030 compared to 2005.
+Overall Ireland's national 2030 target would be about 15\% more ambitious than the EU's updated NDC.
+There are also renewable energy targets to consider for transport (RES-T), heat (RES-H) and electricity (RES-E) under renewable directive (EU) 2018/2001, and energy efficiency targets under (EU) 2018/2002.
+An overview of Ireland's climate targets are shown in Fig.\ref{fig:IrelandClimatetargets}.
 
-%Preliminary emissions data published by the EPA in January 2021 indicate that emissions from the residential sector increased by 9% in 2020, primarily due to the increase in home working and Covid-related restrictions on movement forcing people to spend more time at home.
 \begin{figure}[!htbp]
  \centering
- \includegraphics[scale=0.4]{Figures/Non-ETS-Share.png} 
+ \includegraphics[scale=0.35]{Figures/Non-ETS-Share.png} 
  \caption{Overview of Ireland's 2030 Climate Targets}
@@
 
-Ireland is a unique EU member state, located on the peripherals of Europe, with a the second highest GDP per captia at more than twice the average EU27 \cite{EurostatReal}, the highest proportion of agriculture Non-ETS GHG emmsions from agruclure at 34\% in 2018  \cite{EPA2020Irelands1990-2018}, and high dependency on fossil fuel imports with little electricity or gas grid interconnections.
-Under (EU) 2020/2126 Ireland has been afforded higher flexibility allowances in non-ETS targets, to compensate for some of the barriers.
-The Climate Change Advisory Council (CCAC) aims to set fair, robust and evidence-based sectoral five year carbon budgets, which comply with all of Ireland's climate targets.
-% Insert Figure about most ambitious 2030 targets -
+Ireland is a unique EU member state, located on the peripherals of Europe, with a the second highest GDP per captia at more than twice the average EU27 \cite{EurostatReal}, the highest proportion of agriculture GHG emissions at 34\% in 2018  \cite{EPA2020Irelands1990-2018}, and high dependency on fossil fuel imports with little electricity or gas grid interconnections.
+Under (EU) 2020/2126 - CHECK Ireland has been afforded higher flexibility allowances in non-ETS targets, to compensate for some of these barriers.
+The Climate Change Advisory Council (CCAC) will advise the government on setting sectoral five year carbon budgets, which comply with all of Ireland's climate targets.
+The most ambitious targets....
+overview in Fig \ref{fig:GHG2030A}, shows some of the most ambitious GHG targets globally and the share of GHG emissions from agriculture.
+Despite a large share of agriculture emissions, Ireland has set one of the most ambitious targets towards 2030.
+The CCAC will rely of some modelling tools to provide insights specific to Ireland, one of those tools is discussed in the next section.
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[scale=0.6]{Figures/GHGtargetAGR1.png} 
+ \caption{Most Ambitious 2030 GHG Targets}
+ \label{fig:GHG2030AGR}
+\end{figure}
 
@@
 \par
+
+%Preliminary emissions data published by the EPA in January 2021 indicate that emissions from the residential sector increased by 9% in 2020, primarily due to the increase in home working and Covid-related restrictions on movement forcing people to spend more time at homeThese estimates show Residential emissions increased by 9per cent (0.6 Mt CO2eq) in 2020, largely due to the impact of people working from home.
+2020 also saw low prices for home heating fuel,
 Under the United Nations Framework Convention on Climate Change (UNFCCC) GHG reporting guidelines, set out in ( decision 24/CP.19 ), states the 2006  Intergovernmental Panel on Climate Change (IPCC) guidelines are to be adhere to for national GHG reporting.
@@
 \label{Intro:IreRSDBack}
-
+Among European Union (EU) member states, Ireland has the lowest renewable heat of 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
 \par

### 2021-05-19 manuscript.tex
Updates

@@
 These targets accounts for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to develop sectoral five year carbon budgets to aid progress of the forementioned long-term targets.
-The EU's updated first Nationally Determined Contribution (NDC), increases EU's GHG reduction target from 40\% (first NDC) to at least 55\% (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}.
+The EU's updated Nationally Determined Contribution (NDC), increases EU's GHG reduction target from 40\% (first NDC) to at least 55\% (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}.
 Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
 Climate Action Plan 2021 will have to be more ambitious than its predecessor, as it must at least reflect EU's updated NDC target.
-Assuming the share of contribution of ETS and Ireland's Non-ETS remains the same, to met the EU's updated NDC, Ireland's non-ETS GHG emissions target would change from 30\% to 45\% reduction in 2030 compared to 2005.
+Assuming the share of contribution of ETS and Ireland's Non-ETS remains the same, Ireland's non-ETS GHG emissions target would change from 30\% to 45\% reduction in 2030 compared to 2005.
 Overall Ireland's national 2030 target would be about 15\% more ambitious than the EU's updated NDC.
@@
  \centering
- \includegraphics[scale=0.35]{Figures/Non-ETS-Share.png} 
+ \includegraphics[scale=0.4]{Figures/Non-ETS-Share.png} 
  \caption{Overview of Ireland's 2030 Climate Targets}
@@
 
-Ireland is a unique EU member state, located on the peripherals of Europe, with a the second highest GDP per captia at more than twice the average EU27 \cite{EurostatReal}, the highest proportion of agriculture GHG emissions at 34\% in 2018  \cite{EPA2020Irelands1990-2018}, and high dependency on fossil fuel imports with little electricity or gas grid interconnections.
-Under (EU) 2020/2126 - CHECK Ireland has been afforded higher flexibility allowances in non-ETS targets, to compensate for some of these barriers.
-The Climate Change Advisory Council (CCAC) will advise the government on setting sectoral five year carbon budgets, which comply with all of Ireland's climate targets.
-The most ambitious targets....
-overview in Fig \ref{fig:GHG2030A}, shows some of the most ambitious GHG targets globally and the share of GHG emissions from agriculture.
-Despite a large share of agriculture emissions, Ireland has set one of the most ambitious targets towards 2030.
-The CCAC will rely of some modelling tools to provide insights specific to Ireland, one of those tools is discussed in the next section.
+Ireland is a unique EU member state, located on the peripherals of Europe, with a the second highest GDP per capita \cite{EurostatReal} (Higher GDP per capita translates to higher non-ETS reduction target), the largest proportion of agriculture GHG emissions at 34\% in 2018  \cite{EPA2020Irelands1990-2018}, and a high dependency on fossil fuel imports.
+Ireland has been afforded high non-ETS flexibility allowances under (EU) 2020/2126, to compensate for some of these barriers.
+To put Irelands targets in perspective the ``Leaders' Summit on Climate" in April 2021, brought together the largest GHG emitters to update and raise ambition of their 2030 climate targets.
+Comparing these new targets with the same base year of 2018, Fig \ref{fig:GHG2030AGR} shows Ireland's national climate target is ambitious, especially when the share of agriculture GHG emissions is considered.
+Ireland has some distinct barriers and ambitious targets, the Climate Change Advisory Council (CCAC) will rely of some modelling tools to provide insights and set sectoral five year carbon budgets, one of those modelling tools is discussed in the next section.
 
@@
  \centering
- \includegraphics[scale=0.6]{Figures/GHGtargetAGR1.png} 
+ \includegraphics[scale=0.5]{Figures/GHGtargetAGR1.png} 
  \caption{Most Ambitious 2030 GHG Targets}
@@
 \par % Paragraph 2 - ESOM
-A Energy System Optimisation Model (ESOM) can be a key tool in developing least cost decarbonisation pathways.
-ESOMs can take account of a large number of dynamic variables, and complex interconnections in each time period and economic sector.
-The Integrated MARKAL-EFOM System (TIMES) modelling framework is the ESOM developed in this paper.
-TIMES has been developed and maintained by ETSAP (Energy Technology Systems Analysis Program) under the aegis of the International Energy Agency (IEA), with the key focus to \textit{``establish, maintain and expand a consistent multi-country energy/economy/environment/ engineering analytical capability mainly based on TIMES"} \cite{InternationalEnergyAgency2015InternationalIA}.
-The contracting parties of ETSAP are the governments of eighteen countries and the European Commission.
-
+A Energy System Optimisation Model (ESOM) can take account of a large number of dynamic variables, and complex interconnections in each time period and each economic sector and determine a least cost solution.
+The ESOM described in this paper is The Integrated MARKAL-EFOM System (TIMES) modelling framework, which has been developed and maintained by Energy Technology Systems Analysis Program (ETSAP) of the International Energy Agency (IEA).
+TIMES is a ``hybrid" ESOM, which combines high resolution ``bottom-up" technological detail and ``top-down" macroeconomic variables covering all energy sectors.
 \par While other ESOM frameworks are available to explore Ireland's decarbonisation pathway such as EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}.
-TIMES is the most understood and used, and because of the expertise among energy policy researchers in Ireland and the global network of ETSAP modellers, TIMES is the preferred ESOM.
-TIMES is a ``hybrid" ESOM with combines high technological detail and macroeconomic variables covering all energy sectors.
-The previous Irish TIMES model was extracted from the PET36 model (Pan European TIMES model) and was then updated with national data and assumptions, specific to Ireland.
-This previous Irish TIMES model was used extensively in the past to provide input to energy and climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Resources2015Irelands2015-2030,DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019,Deane2013TechnicalIreland,IrishGovernment2017National2017},however there was a need to design an ESOM to specifically explore Ireland's long term climate targets and a new model called TIMES Ireland Model (TIM) was built in 2021 to address this need.
-TIMES has also been used in UK \cite{Li2018IncorporatingModel,Kannan2009ModellingApproaches}, Denmark \cite{Petrovic2016ResidentialSystem,Tattini2018ImprovingModel}, Sweden \cite{Krook-Riekkola2014ModellingStudy,Aberg2011OptimisationBuildings} and China \cite{Shi2016ModellingModel} to find least cost pathways with a focus on the residential sector.While the many climate targets further add to complexity, the need for an ESOM is strengthened.
-An ESOM can provide insights to least cost energy system pathways, the energy system  transition policymakers must also consider the maximum benefits to the economy to truly optimise the transition.
-TIM can provide vital input to the discussions or further modelling required, and thus the transparency of an ESOM is critical to ensure absorptive capacity of all stakeholders.
+The expertise among energy policy researchers in Ireland and the global network of ETSAP modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways.
+\par Previous versions of the Irish TIMES model were extracted from the PET36 model (Pan European TIMES model), and then updated with specific national data and assumptions, these models were used to provide informative reports to Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.
+A new TIMES model called ``TIMES Ireland Model" (TIM) was built in 2021, TIM is structured to explore Ireland's new climate targets.
+TIMES has also been widely used to internationally, with some studies focusing on the residential sector.
+For example, Li studied the preference for heating technology choice in UK \cite{Li2018IncorporatingModel}, Petrovic examined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed how higher thermal efficiency in buildings might hamper co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}.
+TIMES has been used in Ireland and internationally to provide evidence-based and robust decarbonisation insights for climate policy discussions.
 
@@
 
-%Preliminary emissions data published by the EPA in January 2021 indicate that emissions from the residential sector increased by 9% in 2020, primarily due to the increase in home working and Covid-related restrictions on movement forcing people to spend more time at homeThese estimates show Residential emissions increased by 9per cent (0.6 Mt CO2eq) in 2020, largely due to the impact of people working from home.
-2020 also saw low prices for home heating fuel,
-Under the United Nations Framework Convention on Climate Change (UNFCCC) GHG reporting guidelines, set out in ( decision 24/CP.19 ), states the 2006  Intergovernmental Panel on Climate Change (IPCC) guidelines are to be adhere to for national GHG reporting.
-Ireland’s residential sector accounted for 14.2\%of all GHGs in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+The United Nations Framework Convention on Climate Change (UNFCCC) GHG reporting rules, as per Decision 24/CP.19, states that 2006  Intergovernmental Panel on Climate Change (IPCC) guidelines are to be adhere to for national GHG reporting.
+Under these guidelines, Ireland’s residential sector accounted for 14.2\%of GHGs emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+Focusing on energy-related GHG emissions, the residential sector directly accounts for 24\% of GHG emissions.
 Fig.
-\ref{fig:Ireland2018GHG} shows an overview of Irelands GHG emissions, with a focus on the energy related GHG emissions.
-The residential sector accounts for 47\% of heat emissions 30\% of electricity emissions, therefore the residential sector plays a pivotal role in the energy-related GHG emissions in Ireland.
+\ref{fig:Ireland2018GHG}.
+shows the energy related GHG disaggregated between transport, heat and electricity, where the residential sector is the direct source 47\% of heat and 30\% of electricity GHG emissions, the residential sector is therefore a pivotal sector in Ireland energy decarbonisation.
 
-\begin{figure}[h]
+\begin{figure}[!htbp]
     \begin{tikzpicture}[every node/.style={align=center}, pin distance=17mm]
-    \pie[cloud,text=inside, radius = 2.7]{34/Agriculture,7/Other,59/Energy-related}
-    \draw [->, line width=4pt] (5.8,-0.3) -- (6.8,-.3);
-    \hspace{9cm}
-    \pie[text=inside,radius=2.2,color={black!30,red!70,yellow!40}]{40/Transport,33/Heat,27/ Electricity}
+    \pie[cloud,text=inside,radius = 2.4]{34/Agriculture,7/Other,59/Energy-related}
+    \draw [->, line width=4pt] (5.4,-0.3) -- (6.5,-.3);
+    \hspace{8.5cm}
+    \pie[rotate =30,text=inside,radius=1.95,color={black!30,red!70,yellow!40}]{40/Transport,33/Heat,27/Electricity}
     \end{tikzpicture}
@@
 
- While the national GHG should reduce 51\% by 2030, the residential sector is expected to be reduce GHGs even further, due to the higher cost-effectiveness when compared with other sectors.
-The high ambition in residential is due to its dependency on receiving low carbon intensity electricity.
-The decarbonisation of the power generation sector is well underway, with carbon intensity falling annually to 375 gCO2/kWh in 2018, 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}.
-However the on-shore wind energy capacity is almost saturated, and policy looks to guide investment into off-shore wind to continue a reduction in carbon intensity.
-Under CAP2019 this continued growth and decarbonisation of the electricity will be the fundamental to lowering energy-related GHG in not only electricity, but heat and transport also.
+ \par The residential sector target is expected to be reduce GHG emissions beyond the 51\% national target by 2030 compared to 2018, due to the higher cost-effectiveness of abating GHG emissions when compared with other sectors.
+Despite these high target ambitions, preliminary results from the Environmental Protection Agency (EPA) has indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.
+The future decarbonisation of the residential sector is dependant on receiving low carbon intensity electricity.
+The decarbonisation of the power generation sector is well underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}.
+Under CAP2019 the continued growth and decarbonisation of the electricity will be the fundamental to lowering energy-related GHG in not only electricity and heat, but in transport also \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
 
@@
 \par
-The CAP2019 had a Marginal Abatement Cost Curve (MACC) is based on  based on Environmental Protection Agency (EPA) GHG emission projection report \cite{EPA2018EPAReport} and provides a solid analytical foundation on the most cost-effective pathway to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels.
-The MACC explores technological, behavioural and fuel switching.
-The x-axis (width of each bar chart) shows the potential reduction of annual MtCO2eq.
-emissions in 2030 from the technology or fuel switch.
-The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of CO2eq.
+
+The CAP2019 Marginal Abatement Cost Curve (MACC) developed by McKinsey and Company's modelling tools is based on EPA's GHG emission projection report \cite{EPA2018EPAReport} and provides an analytical foundation on the most cost-effective pathway to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels.
+The CAP2019 MACC in Fig.\ref{fig:CAP2019:MACC} explores technological, behavioural and fuel switching options.
+The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_e_q$.
+The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_e_q$.
 over the 2021 to 2030 period.
-The columns are organised from the most economical (left side) to the most expensive technology (right side) in EUR/tCO2eq \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-In the least-cost pathway Ireland would increase Onshore wind by 8.2 GW and Offshore would grow to 3.5 GW,this would be expected to save 7 to 8 MtCO2eq.
+The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_e_q$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+For example, increasing onshore-wind by $8.2 GW$ and offshore-wind would grow to $3.5 GW$,this would be expected to save $7 to 8 MtCO2_e_q$.
 This would help to indirectly reduce the GHG in the residential sector.
-However specific residential solutions in the MACC include:
+Solutions specific to the residential sector are:
 \begin{itemize}
@@
     \item{Retrofit gas boiler and solid fuel stove existing dwellings to B2 equivalent}
-    \item{Introduce CO2-free heating in new buildings}
+    \item{Introduce CO2-free heating in new buildings}\cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
 \end{itemize}
 
-McKinsey and Company's modelling tools have helped develop the MACC shown in Fig.\ref{fig:CAP2019:MACC}, with more ambitious Non-ETS 2030 target due this year and a national GHG target for 2030, there is a requirement for a new MACC, from which a new residential five-year carbon budgets will arise.
-Sustainable Energy Authority of Ireland (SEAI) and others have been part of the renovation wave in Ireland, the uptake of heat pumps in existing homes is gradually building due to grants, awareness and increasing the ease of upgrades for homeowners, while new homes have a strict regulation to comply with, so heat pumps despite their high capital costs are preferred.
+Sustainable Energy Authority of Ireland (SEAI) and others have been part of the renovation wave in Ireland, the uptake of heat pumps in existing homes is gradually building due to grants, awareness and increasing the ease of upgrades for homeowners.
+In parallel, new homes have a strict regulation to comply with, which means despite the high capital costs heat pumps are the preferred heating technology.
 The building regulations in 2021 will require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}.
-The CAP2019 plan to deliever a new retrofitting model to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030, highlighting the energy efficiency first approach in how Ireland aims to progress.
+The CAP2019 plan to deliver a new retrofitting model to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030, highlighting the energy efficiency first approach in how Ireland aims to progress.
 
@@
 \label{Intro:IreRSDBack}
-Among European Union (EU) member states, Ireland has the lowest renewable heat of 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
+
 \par
-Ireland’s decarbonisation pathway, is further separated from other EU member states when focusing on the residential sector, this is largely down to Ireland’s high reliance on fossil fuels and settlement patterns of low thermal efficiency of the housing stock.
-Ireland's lower than average urbanisation which is connected to only 5.8\% of the population of Ireland living in apartments in 2018, the lowest and well below the average EU-27 of 25.3\% \cite{EurostatDistributionSurveyilc_lvho01}, the lower density relates translates to more expensive mitigation.
-33\% of the population of the EU were deemed to live in dwellings too large for their needs in 2018, in Ireland the rate was 71.4\% \cite{EurostatShareSurveyilc_lvho50a}.
-Ireland set out national policy objectives under 10 strategic investment priorities \cite{Ireland2018ProjectFramework}.
-The first of strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, is a top priority".}, which will indirectly effect both electricity and transport GHG emission.
-In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved in the last 30 years, and in 2018 peat accounted for 12.3\% of residential GHG emissions.
+Ireland’s distinctive decarbonisation pathway, is further separated from other EU member states when focusing on the residential sector.
+This is largely down to Ireland’s high reliance on fossil fuels, larger dwellings and rural settlement patterns of low thermal efficiency of the housing stock.
+This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, and 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\%.
+\cite{EurostatShareSurveyilc_lvho50a}.
+The lower population density and large dwellings of the existing Irish housing stock, translates to more expensive GHG mitigation options compared to other residential sectors in the EU.
+
+In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential GHG emissions.
 Residential coal GHG emissions has also declined to 6.9\% in 2018.
-However, the decline in peat and coal has been offset by more significant growth in oil boilers in rural dwellings in the same period, where oil accounts for 34.8\% of residential GHG emissions \cite{Howley2005ENERGY-RELATEDIreland}.
-Urban dwellings prefer gas as a heating fuel, which is cheaper, more secure ( 61\% of gas in 2018 was indigenous ) and less carbon intensive.
-Gas accounted for 15.5\% of residential GHG emission in 2018, significantly lower than oil due to the lower carbon intensity.
-Gas Network Ireland (GNI) anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}
-In summary, the residential sector is highly reliant on fossil fuels as shown in Fig.\ref{fig:ResFuel}.
+However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+Gas which is only available in some urban areas, accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+Gas Network Ireland (GNI) anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}.
+In summary, the residential sector is highly reliant on fossil fuels as shown in Fig.\ref{fig:ResFuel}, consequently Ireland currently has the lowest renewable heat of  among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
 
@@
 
+%PICK UP HERE
 Historical fuel usage in one part of understanding residential energy, the other main part is understanding the low thermal efficiency of Irelands housing stock.
@@
 \label{Intro:Policy}
+Ireland has set out national policy objectives under 10 strategic investment priorities \cite{Ireland2018ProjectFramework}.
+The first strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}.
 The 2019 Climate Action Plan identified 29 actions that would reduce emissions from the sector by 40-45%.  These actions included:
@@
  % LIT REW INPUT DATA - EU \cite{Simoes2013TheProject} 
-
+- LI,er.alThe nationwide survey identified existing technologies, age,
+income, region, dwelling characteristics, and knowledge of eco- technology as the six most influential factors for determining homeowners' preferences for heating systems. Among
 %the new A-rated houses behave distinctively different which much less demand on space heating and with pumps and fans energy consumption usage increasing, while appliance end use is increasing across the BER ratings.
@@
 \label{sec:sample:appendix}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor section \ref{sec:Introduction} incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+XXXXXXXX
 

### 2021-05-20 manuscript.tex
20May2021

@@
 
-\subsection{Ireland: Residential Sector}
-\label{Intro:IreRSD}
-\par
-
-The United Nations Framework Convention on Climate Change (UNFCCC) GHG reporting rules, as per Decision 24/CP.19, states that 2006  Intergovernmental Panel on Climate Change (IPCC) guidelines are to be adhere to for national GHG reporting.
-Under these guidelines, Ireland’s residential sector accounted for 14.2\%of GHGs emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
-Focusing on energy-related GHG emissions, the residential sector directly accounts for 24\% of GHG emissions.
-Fig.
-\ref{fig:Ireland2018GHG}.
-shows the energy related GHG disaggregated between transport, heat and electricity, where the residential sector is the direct source 47\% of heat and 30\% of electricity GHG emissions, the residential sector is therefore a pivotal sector in Ireland energy decarbonisation.
-
-\begin{figure}[!htbp]
-    \begin{tikzpicture}[every node/.style={align=center}, pin distance=17mm]
-    \pie[cloud,text=inside,radius = 2.4]{34/Agriculture,7/Other,59/Energy-related}
-    \draw [->, line width=4pt] (5.4,-0.3) -- (6.5,-.3);
-    \hspace{8.5cm}
-    \pie[rotate =30,text=inside,radius=1.95,color={black!30,red!70,yellow!40}]{40/Transport,33/Heat,27/Electricity}
-    \end{tikzpicture}
-    \caption{Ireland 2018 GHG Share \cite{Howley2005ENERGY-RELATEDIreland}}
-    \label{fig:Ireland2018GHG}
-\end{figure}
-
- \par The residential sector target is expected to be reduce GHG emissions beyond the 51\% national target by 2030 compared to 2018, due to the higher cost-effectiveness of abating GHG emissions when compared with other sectors.
-Despite these high target ambitions, preliminary results from the Environmental Protection Agency (EPA) has indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.
-The future decarbonisation of the residential sector is dependant on receiving low carbon intensity electricity.
-The decarbonisation of the power generation sector is well underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}.
-Under CAP2019 the continued growth and decarbonisation of the electricity will be the fundamental to lowering energy-related GHG in not only electricity and heat, but in transport also \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-
-\begin{figure}[!htbp]
- \centering
- \includegraphics[scale=1.0]{Figures/CAP2019MACC.png} 
- \caption{Climate Action Plan 2019 - Marginal Abatement Cost Curve}
- \label{fig:CAP2019:MACC} \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
-\end{figure}
-
-\par
-
-The CAP2019 Marginal Abatement Cost Curve (MACC) developed by McKinsey and Company's modelling tools is based on EPA's GHG emission projection report \cite{EPA2018EPAReport} and provides an analytical foundation on the most cost-effective pathway to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels.
-The CAP2019 MACC in Fig.\ref{fig:CAP2019:MACC} explores technological, behavioural and fuel switching options.
-The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_e_q$.
-The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_e_q$.
-over the 2021 to 2030 period.
-The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_e_q$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-For example, increasing onshore-wind by $8.2 GW$ and offshore-wind would grow to $3.5 GW$,this would be expected to save $7 to 8 MtCO2_e_q$.
-This would help to indirectly reduce the GHG in the residential sector.
-Solutions specific to the residential sector are:
-\begin{itemize}
-    \item{Retrofit oil boiler existing dwellings to B2 equivalent}
-    \item{Switch from oil boilers to heat pumps in existing dwellings}
-    \item{Retrofit gas boiler and solid fuel stove existing dwellings to B2 equivalent}
-    \item{Introduce CO2-free heating in new buildings}\cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
-\end{itemize}
-
-Sustainable Energy Authority of Ireland (SEAI) and others have been part of the renovation wave in Ireland, the uptake of heat pumps in existing homes is gradually building due to grants, awareness and increasing the ease of upgrades for homeowners.
-In parallel, new homes have a strict regulation to comply with, which means despite the high capital costs heat pumps are the preferred heating technology.
-The building regulations in 2021 will require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}.
-The CAP2019 plan to deliver a new retrofitting model to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030, highlighting the energy efficiency first approach in how Ireland aims to progress.
-
 \subsection{Ireland: Residential Sector Background}
@@
 Ireland’s distinctive decarbonisation pathway, is further separated from other EU member states when focusing on the residential sector.
-This is largely down to Ireland’s high reliance on fossil fuels, larger dwellings and rural settlement patterns of low thermal efficiency of the housing stock.
-This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, and 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\%.
+This is largely down to Ireland’s high reliance on fossil fuels, low building thermal efficiency, larger dwellings and rural settlement patterns.
+This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\%.
 \cite{EurostatShareSurveyilc_lvho50a}.
-The lower population density and large dwellings of the existing Irish housing stock, translates to more expensive GHG mitigation options compared to other residential sectors in the EU.
+The low building thermal efficiency is highlighted in Fig.\ref{fig:BERAge} and home heating oil accounts for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport} .
 
@@
 
-%PICK UP HERE
-Historical fuel usage in one part of understanding residential energy, the other main part is understanding the low thermal efficiency of Irelands housing stock.
-As mentioned in section \ref{Intro:IreRSD}, Irelands new buildings produce 70\% less GHG emissions than building in 2005, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock.
+The building regulations in 2021 will require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock.
 According to Central Statistics Office (CSO) approx.
 85\% of residential buildings are 2005 or older.
-The BER database, which represents half of the dwelling stock highlights the impact the year of construction has on the expected energy consumption, as shown in Fig.\ref{fig:BERAge}.The BERs are categorised into BER labels, with A1 representing the lowest energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest energy consumption ($ >450 kWh/m^2/year$) .
+The BER database, which represents half of the dwelling stock, highlights the affect of construction year on the expected energy consumption, as shown in Fig.\ref{fig:BERAge}.
+The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
+The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, helps to better understand the housing stock.
+The BERs are required for new houses, rented houses, sold houses and renovated houses.
 
@@
  \centering
- \includegraphics[scale=0.4]{Figures/BERAge.png}
+ \includegraphics[scale=0.5]{Figures/BERAge.png}
  \caption{Building Energy Rating by Age}
@@
 
-The poor BERs consequently translates to higher energy bills to maintain comfortable temperature, but in reality it relates to low internal temperatures, which likely results in higher excessive winter mortality \cite{Clinch2000HousingMortality}.The BERs which come from the Energy Performance of Buildings Directive 2002/91/ EC, helps to better understand the housing stock.
-The BERs are required for new houses, rented houses, sold houses and renovated houses.
-However, BERs do not take account of actual usage, i.e occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
-The BER labels do therefore leave a gap between actual consumption and expected consumption, I will explain how I approached this problem in the Literature review.
+Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which results in higher excessive winter mortality \cite{Clinch2000HousingMortality}.
+BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
+The BER labels do therefore leave a gap between actual consumption and expected consumption, I will explain how I approached this problem in the literature review.
 
-Fuel Poverty is when a household spends \textit{``10\% or more of income to achieve adequate warmth"}, an element energy reported 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}.
-While income supports and the \textit{Fuel Allowance and the Household Benefits} provides support to low income households to allievate some the fuel poverty.
+Fuel Poverty which is a major contributor to the high excessive winter deaths, is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"}.
+An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}.
+Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty.
 These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}.
-As Kerr points out the link between energy and social welfare policy is made explicitly \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty, which include increased risk of catastrophic cardiovascular or cerebrovascular events \cite{OMeara2016AIreland}.
-Ironically, this places low income households deeply dependent on fossil fuels for health.
+As Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty, which include increased risk of catastrophic cardiovascular or cerebrovascular events \cite{OMeara2016AIreland}.
+Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
+
+The United Nations Framework Convention on Climate Change (UNFCCC) GHG reporting rules, as per Decision 24/CP.19, states that 2006  Intergovernmental Panel on Climate Change (IPCC) guidelines are to be adhere to for national GHG reporting.
+Under these guidelines, Ireland’s residential sector accounted for 14.2\%of GHGs emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+Focusing on energy-related GHG emissions, the residential sector directly accounts for 24\% of GHG emissions.
+Fig.
+\ref{fig:Ireland2018GHG} shows the energy related GHG disaggregated between transport, heat and electricity.
+Where the residential sector is the direct source 47\% of heat and 30\% of electricity GHG emissions, the residential sector is therefore a pivotal sector in Ireland energy decarbonisation.
+
+\begin{figure}[!htbp]
+    \begin{tikzpicture}[every node/.style={align=center}, pin distance=17mm]
+    \pie[cloud,text=inside,radius = 2.4]{34/Agriculture,7/Other,59/Energy-related}
+    \draw [->, line width=4pt] (5.4,-0.3) -- (6.5,-.3);
+    \hspace{8.5cm}
+    \pie[rotate =30,text=inside,radius=1.95,color={black!30,red!70,yellow!40}]{40/Transport,33/Heat,27/Electricity}
+    \end{tikzpicture}
+    \caption{Ireland 2018 GHG Share \cite{Howley2005ENERGY-RELATEDIreland}}
+    \label{fig:Ireland2018GHG}
+\end{figure}
 
@@
 
-Space heating accounted for the majority of final energy, Fig.
-\ref{fig:Ireland2018ServiceDemands} shows a snapshot of the residential energy services demands in 2018.
-As households become more thermal efficient and use higher efficient heating, space heating energy would be expected to decline.
-The new A-rated dwellings will use less energy for space heating, but Pumps \& Fans energy will increase as mechanical ventilation is used over natural ventilation.
+Space heating is the largest residential energy service demand, accounted for 61.8\% of final energy usage.
+Other residential energy services and their share of final energy in 2018 is shown in \ref{fig:Ireland2018ServiceDemands}.
+As building thermal efficiency and heating technology efficiency improves, both space and water heating energy would be expected to decline.
+The new A-rated dwellings can be used to predict the share of future energy service demands.
+From the BER database, it can be seen that A-rated dwellings use less than half space heating energy of the average dwelling and A-rated water heating energy decreases by about 40\% compared to the average A-rated dwelling.
+However Pumps \& Fans energy increases as mechanical ventilation is used over natural ventilation and Lighting usage in A-rated is more than double the average household.
+Cooking and Appliances energy usage would be expected to remain stable.
 
@@
 
-Space heating is therefore central to residential energy policy, other important service energy demands are shown in Fig.\ref{fig:Ireland2018ServiceDemands}.
-Notably space cooling is absent for the pie chart, but space cooling is accounted for in appliances.
-The Policy Oriented Tool for Energy and Climate Change Impact Assessment (POTEnCIA) project disaggregated Ireland's residential space cooling from appliances and have output space cooling energy consumption data, shown in Fig.\ref{fig:Cooling} a significant amount of space cooling energy is expected in the Ireland residential sector.
+Notably space cooling is absent from Fig \ref{fig:Ireland2018ServiceDemands}, but space cooling is accounted for in appliances.
+The Policy Oriented Tool for Energy and Climate Change Impact Assessment (POTEnCIA) project disaggregated Ireland's residential space cooling from appliances and have output space cooling energy consumption data, shown in Fig.\ref{fig:Cooling} a significant increase in space cooling energy is expected in the Ireland residential sector.
 
@@
 \label{Intro:Policy}
+
+The energy-related GHG emissions is expected to be reduce GHG emissions beyond the 51\% national target by 2030 compared to 2018, due to the higher cost-effectiveness of abating GHG emissions when compared with agriculture.
+Despite these high target ambitions in the energy-related GHG emissions, preliminary results from the Environmental Protection Agency (EPA) has indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.
+The CAP2019 Marginal Abatement Cost Curve (MACC), which was developed by McKinsey and Company's modelling tools and is based on EPA's GHG emission projection report \cite{EPA2018EPAReport} found electricity & transport are the most cost effective for decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
+The future decarbonisation of the residential sector is dependant on receiving low carbon intensity electricity.
+
 Ireland has set out national policy objectives under 10 strategic investment priorities \cite{Ireland2018ProjectFramework}.
 The first strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}.
-The 2019 Climate Action Plan identified 29 actions that would reduce emissions from the sector by 40-45%.  These actions included:
-Reducing fossil fuel use in the sector and moving to electricity to provide heating and hot water in buildings
-500,000 building retrofits to a BER B2 / cost optimal equivalent or carbon equivalent
-Installation of 600,000 heat pumps, with 400,000 of these in existing buildings
-Development of district heating, including two initiatives of municipal scale with the potential to provide heat equivalent to the needs of about 50,000 homes
-Complete the rollout of the Support Scheme for Renewable Heat
-Increasing the number of Sustainable Energy Communities to 1,500
+Some of these actions included:
+\begin{itemize}
+    \item{Reducing fossil fuel use in the sector and moving to electricity to provide heating and hot water in buildings}
+    \item{500,000 building retrofits to a BER B2 / cost optimal equivalent or carbon equivalent}
+    \item{Installation of 600,000 heat pumps, with 400,000 of these in existing buildings}
+    \item{Development of district heating, including two initiatives of municipal scale with the potential to provide heat equivalent to the needs of about 50,000 homes}
+    \item{Complete the roll-out of the Support Scheme for Renewable Heat}
+    \item{Increasing the number of Sustainable Energy Communities to 1,500}
+\end{itemize}
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[scale=1.0]{Figures/CAP2019MACC.png} 
+ \caption{Climate Action Plan 2019 - Marginal Abatement Cost Curve}
+ \label{fig:CAP2019:MACC} \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
+\end{figure}
+
+\par
+The CAP2019 MACC, shown in Fig.\ref{fig:CAP2019:MACC} provides $>300$ of the most cost-effective technologies options to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels.
+The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_e_q$.
+The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_e_q$.
+over the 2021 to 2030 period.
+The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_e_q$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+The options specific to the residential sector are:
+\begin{itemize}
+    \item{Retrofit oil boiler existing dwellings to B2 equivalent}
+    \item{Switch from oil boilers to heat pumps in existing dwellings}
+    \item{Retrofit gas boiler and solid fuel stove existing dwellings to B2 equivalent}
+    \item{Introduce CO2-free heating in new buildings}\cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
+\end{itemize}
+
+Where, the retrofitting of oil boilers is expected to reduce the Total Cost of Ownership (TCO) (ie.
+save money), but only abate a small amount of carbon, whereas switching oil boilers to heat pumps will increase TCO (ie.
+cost money) but the expected abatement of carbon is far greater.
 
@@
 \par
-Decarbonising Electricity is key for Ireland, both the \textit{``stick"} (tax) approach which aims to deter investment in fossil fuel power generation, and the \textit{``carrot"} (subsidies) approach which aims to encourage renewable electricity is applied in Irish policy.
-In Ireland, a €15/tCO2 carbon tax was introduced in 2009 to influence fossil fuel demand.
-The carbon tax receipts grew every year from 2010 to 2018 \cite{AnOffice}.
-Ireland's current carbon tax of €33.50/tCO2 will rise to €100/tCO2 in 2030.
-Many countries have taken this approach, especially in Europe, where Finland was the first country to introduce a carbon tax in 1990, while Sweden were second to introduce a carbon tax after Finland, the carbon tax rate is the highest in the world at €108.81/tCO2.
-A carbon price or tax now accounts for approx.
-15\% of GHG emissions \cite{WorldBankGroup2020State2020}.
-The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentives investment in renewable electricity to achieve legally binding renewable targets under 2009/28/EC.
+The decarbonisation of the power generation sector is well underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}.
+Under CAP2019 the continued growth and decarbonisation of the electricity will be the fundamental to lowering energy-related GHG in not only electricity, but in heat and transport also \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+Decarbonising electricity is key for Ireland, both the \textit{``stick"} (tax) approach which aims to deter investment in fossil fuel power generation, and the \textit{``carrot"} (subsidies) approach which aims to encourage renewable electricity is applied in Irish policy.
+In Ireland, a €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce fossil fuel demand and investment.
+The carbon tax receipts grew every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}.
+Ireland's current carbon tax of €33.50/ $tCO_2$ will rise to €100/ $tCO_2$ in 2030.
+Many countries have taken this approach, especially in Europe, where Sweden were second to introduce a carbon tax after Finland, the Swedish carbon tax rate is the highest in the world at €108.81/ $tCO_2$.
+About 15\% of global GHG emissions have either an associated cost or tax, this figure is increasing incrementally \cite{WorldBankGroup2020State2020}.
+The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise Ireland to invest in renewable electricity to achieve legally binding renewable targets under 2009/28/EC.
 REFIT operates by guaranteeing  new renewable generation and biomass a minimum price for electricity over a 15 year period.
@@
 \subsubsection{Heat}
-Domestic supports for renewable heat or energy efficiency is through SEAI grants,\begin{itemize}
+Sustainable Energy Authority of Ireland (SEAI) and others have been part of the renovation wave in Ireland, the uptake of heat pumps in existing homes is gradually growing due to grants, awareness and ease of upgrades for homeowners.
+In parallel, new homes have a strict regulation to comply with, which means despite the high capital costs heat pumps are the preferred heating technology.
+As mentioned in Section\ref{Intro:IreRSDBack} new buildings in Ireland produce 70\% less GHG emissions than building in 2005 and Section\ref{Intro:Policy} mentioned plans to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030.
+The energy efficiency first approach in how Ireland aims to
+decarbonize the residential sector. Domestic supports for renewable heat or energy efficiency include:
+\begin{itemize}
     \item {Insulation}
@@
 \end{itemize}
-
-This report is a study of potential design options for a Renewable Heat Incentive (RHI) f
+%PICKUPHERE
+This report is a study of potential design options for a Renewable Heat Incentive (RHI)
 

### 2021-05-21 manuscript.tex
Send 1st draft to Fionn, Hannah and Olex

@@
 These targets accounts for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to develop sectoral five year carbon budgets to aid progress of the forementioned long-term targets.
-The EU's updated Nationally Determined Contribution (NDC), increases EU's GHG reduction target from 40\% (first NDC) to at least 55\% (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}.
-Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+\par The EU's first Nationally Determined Contribution (NDC) which had a 40\% GHG reduction target (first NDC), has increased to at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}.
+Ireland's Climate Action Plan 2019 (CAP2019) which complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
 Climate Action Plan 2021 will have to be more ambitious than its predecessor, as it must at least reflect EU's updated NDC target.
-Assuming the share of contribution of ETS and Ireland's Non-ETS remains the same, Ireland's non-ETS GHG emissions target would change from 30\% to 45\% reduction in 2030 compared to 2005.
+Assuming the share of contribution of ETS and Ireland's Non-ETS remains the same, Ireland's non-ETS GHG emission 2030 target would change from 30\% to 45\%.
 Overall Ireland's national 2030 target would be about 15\% more ambitious than the EU's updated NDC.
@@
 Ireland is a unique EU member state, located on the peripherals of Europe, with a the second highest GDP per capita \cite{EurostatReal} (Higher GDP per capita translates to higher non-ETS reduction target), the largest proportion of agriculture GHG emissions at 34\% in 2018  \cite{EPA2020Irelands1990-2018}, and a high dependency on fossil fuel imports.
-Ireland has been afforded high non-ETS flexibility allowances under (EU) 2020/2126, to compensate for some of these barriers.
-To put Irelands targets in perspective the ``Leaders' Summit on Climate" in April 2021, brought together the largest GHG emitters to update and raise ambition of their 2030 climate targets.
-Comparing these new targets with the same base year of 2018, Fig \ref{fig:GHG2030AGR} shows Ireland's national climate target is ambitious, especially when the share of agriculture GHG emissions is considered.
-Ireland has some distinct barriers and ambitious targets, the Climate Change Advisory Council (CCAC) will rely of some modelling tools to provide insights and set sectoral five year carbon budgets, one of those modelling tools is discussed in the next section.
+Ireland has been afforded high non-ETS flexibility allowances under (EU) 2020/2126, to compensate for these barriers, but Ireland still has ambitious national climate targets.
+To put Ireland's climate targets in perspective Fig.\ref{fig:GHG2030AGR} compares different countries 2030 climate target with a 2018 base year and share of agriculture GHG emissions.
+Ireland has ambitious targets, despite distinct barriers.
+The Climate Change Advisory Council (CCAC) will rely of some modelling tools to provide insights and set sectoral five year carbon budgets.
 
@@
 \par % Paragraph 2 - ESOM
-A Energy System Optimisation Model (ESOM) can take account of a large number of dynamic variables, and complex interconnections in each time period and each economic sector and determine a least cost solution.
+A Energy System Optimisation Model (ESOM) can optimise the selection of a large number of dynamic variables to satisfy demand, while accounting for complex economic sectoral interconnections within each time period.
 The ESOM described in this paper is The Integrated MARKAL-EFOM System (TIMES) modelling framework, which has been developed and maintained by Energy Technology Systems Analysis Program (ETSAP) of the International Energy Agency (IEA).
 TIMES is a ``hybrid" ESOM, which combines high resolution ``bottom-up" technological detail and ``top-down" macroeconomic variables covering all energy sectors.
-\par While other ESOM frameworks are available to explore Ireland's decarbonisation pathway such as EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}.
-The expertise among energy policy researchers in Ireland and the global network of ETSAP modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways.
-\par Previous versions of the Irish TIMES model were extracted from the PET36 model (Pan European TIMES model), and then updated with specific national data and assumptions, these models were used to provide informative reports to Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.
-A new TIMES model called ``TIMES Ireland Model" (TIM) was built in 2021, TIM is structured to explore Ireland's new climate targets.
-TIMES has also been widely used to internationally, with some studies focusing on the residential sector.
-For example, Li studied the preference for heating technology choice in UK \cite{Li2018IncorporatingModel}, Petrovic examined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed how higher thermal efficiency in buildings might hamper co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}.
-TIMES has been used in Ireland and internationally to provide evidence-based and robust decarbonisation insights for climate policy discussions.
+\par In addition to TIMES there are other ESOM frameworks available such as EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}.
+However the expertise among energy policy researchers in Ireland combined with the global network of ETSAP modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways.
+\par Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model, and then updated with specific national data and assumptions.
+The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.
+While TIMES models can cover all the energy-related sectors, it can also be used to explore the residential sector in detail.
+Some studies focusing on the residential sector in TIMES include, Li who studied the preference for heating technology choice in UK \cite{Li2018IncorporatingModel}, Petrovic examined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed the affect higher thermal efficiency in buildings has on co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}.
+There has also been some studies residential sector in Ireland, using previous versions of Irish TIMES and now out-dated climate targets.
+Therefore, there is a need for a new TIMES model to be built specifically to explore Ireland's new 2030 and 2050 climate targets and sectoral five-year carbon budgets, the ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
 
@@
 This is largely down to Ireland’s high reliance on fossil fuels, low building thermal efficiency, larger dwellings and rural settlement patterns.
-This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\%.
-\cite{EurostatShareSurveyilc_lvho50a}.
-The low building thermal efficiency is highlighted in Fig.\ref{fig:BERAge} and home heating oil accounts for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport} .
+This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}.
+The low thermal efficiency of Ireland's dwelling is discussed later in this section and highlighted in Fig.\ref{fig:BERAge}.
+Furthermore Irelands fossil fuel dependency is underlined by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.
 
 In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential GHG emissions.
-Residential coal GHG emissions has also declined to 6.9\% in 2018.
+Residential coal GHG emissions has also declined to 6.9\% in 2018 \cite{EnergySEAI}.
 However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
 Gas which is only available in some urban areas, accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
-Gas Network Ireland (GNI) anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}.
 In summary, the residential sector is highly reliant on fossil fuels as shown in Fig.\ref{fig:ResFuel}, consequently Ireland currently has the lowest renewable heat of  among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
@@
 
-The building regulations in 2021 will require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock.
+The 2021 building regulations in 2021 require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock.
 According to Central Statistics Office (CSO) approx.
@@
 
-The energy-related GHG emissions is expected to be reduce GHG emissions beyond the 51\% national target by 2030 compared to 2018, due to the higher cost-effectiveness of abating GHG emissions when compared with agriculture.
+The energy-related GHG emissions are expected to be reduced beyond the 51\% national target by 2030 compared to 2018, due to the higher cost-effectiveness of abating GHG emissions when compared with agriculture.
 Despite these high target ambitions in the energy-related GHG emissions, preliminary results from the Environmental Protection Agency (EPA) has indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.
@@
 \subsubsection{Heat}
-Sustainable Energy Authority of Ireland (SEAI) and others have been part of the renovation wave in Ireland, the uptake of heat pumps in existing homes is gradually growing due to grants, awareness and ease of upgrades for homeowners.
-In parallel, new homes have a strict regulation to comply with, which means despite the high capital costs heat pumps are the preferred heating technology.
-As mentioned in Section\ref{Intro:IreRSDBack} new buildings in Ireland produce 70\% less GHG emissions than building in 2005 and Section\ref{Intro:Policy} mentioned plans to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030.
-The energy efficiency first approach in how Ireland aims to
-decarbonize the residential sector. Domestic supports for renewable heat or energy efficiency include:
+%Gas Network Ireland (GNI) anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}.
+Sustainable Energy Authority of Ireland (SEAI) have been played a leading role in the uptake of building fabric upgrades and heat pumps installations in existing homes.
+SEAI provide numerous grants to homeowners, increase public awareness of the benefits and are working on simplifying the process.
+In parallel, new dwellings must achieve a minimum of 20\% renewable energy, the ambient heat used in heat pumps is included as a renewable energy, often heat pumps are the preferred heating technology despite the high capital costs.
+As mentioned in Section\ref{Intro:IreRSDBack} new buildings in Ireland produce 70\% less GHG emissions than building in 2005 and Section\ref{Intro:Policy} mentioned CAP2019 plans to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030.
+The energy efficiency first approach in how Ireland aims to decarbonize the residential sector.
+Domestic supports for renewable heat or energy efficiency include:
 \begin{itemize}
@@
 \end{itemize}
-%PICKUPHERE
-This report is a study of potential design options for a Renewable Heat Incentive (RHI)
 
--  critical for the development of renewable gas projects in Ireland.European Commission’s decision on State aid for the Support Scheme for Renewable Heat (SA.50807).
-GNI Vision 2050 - 20\% renewable gas by 2030
-While the electricity network reaches all residential dwellings, The gas network reaches 40\% or 680,000 dwellings in towns and cities. The 2030 gas decarbonisation targets, set by GNI  
-and  with biomethance and renewable gas targets for 2030, X\%.
-The district heating networks are new and currently only two in Dublin which reach about XX houses, 2030 target X%.
-large amount of data centres.
-Other future heating fuel options such as hydrogen, HVO, solar may also play a part.
-Gas Networks Ireland (GNI) has a strategic plan to achieve 20% Renewable Gas on the network by 2030 
-Further investigate decarbonisation options such as hydrogen vehicles, biomethane and AD
-substitutes for natural gas - See \ref{fig:CAP2019:MACC} shows biomethane as the most expensive options..
+\subsubsection{Gas \& District Heating}
 
-\subsection{TIMES Ireland Model (TIM)}
-\label{Intro:TIM}
-\par
-TIM was built to help Ireland provide a pathway to achieve 2030 and 2050 targets.
-The residential sector plays a key part of this and it was built from the SEAI 2018 energy balance, the BER database and CSO data.
-The data filtration process used in the residential sector, derived a lot of conversation of the BER database, a group know as Ireland Energy Wiki by the National Retrofitting Modelling Group grew from building the residential sector, questions about the filters, default values, assumptions around thermal equations (EC..) and uncertainties were used to improve the data before it was used in TIM .
-TIM has many internal project deadines, which required extensive meetings about the data, constraints and results as the modelling was progressing.
-The data and results were presented to 30 external reviewers to provide feedback, this process proved invaluable having different expertise with different presepectives enabled TIM to fill in gaps.
-TIM will also be open platform to help improve the model and provide transparency.
-TIM’s residential sector was built in parallel with LEAP’s residential sector, so that residential sector can easily simulate policy.
-The demand for results from TIM has increased significantly with increased government ambition, TIM will provide input into CAP 2021 and Ireland’s carbon budgets, working with DECC and the CCAC respectively will also provide feedback and improve the robustness of TIM.
-for these reasons TIM residential sector is a state of the art modelling platform.
-%GRAPH OF IRELAND DECARBONISATION TARGETS  VS OTHER COUNTRIES - DIFFERENT INTRASTATE and RESOURCE AT BIDENS ENERGY SUBMIT - DENMAKR TOP 2018, 
-	ESOM can be used to start the conversation on..
-Look at other ESOM, brief explain…
-Other residential sectors..
-CAP2019 - Choosing the Pathways which Create the Least Burden
-and Offer the Most Opportunity for Ireland
+In a bid not to be left behind in a ``net-zero" future, Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019AIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+GNI plans to met ``net-zero" in 2050 by using 50\% Carbon Capture and Storage (CCS), 37\% biomethane and about 13\% Hydrogen\cite{Ervia2019AIreland}.
+The gas network currently reaches 40\% or 680,000 dwellings in urban areas, but with the ``Compact Growth" as mentioned in section \ref{Intro:Policy} a higher share of building may have access to the gas network in the future.
+Increasing biomethane from an agricultural source in the gas network is the most expensive option in CAP2019's MACC shown in Fig.\ref{fig:CAP2019:MACC}.
+
+Ireland has one of the lowest shares of District Heating (DH) in Europe at less than 1\%  \cite{Gartland2016AIreland} \cite{Harney2020DeterminingIreland}
+In Ireland,  DH networks are a new, despite being used for decade across central Europe. CAP2019 plans to have approx. 0.4\% of heating provided by DH in 2030, where the  Research in Ireland h
+IrDEA Up to 57\% feasible - same as Denmark DH
+37% of the Irish heat demand to district heatingshows that the district heating option provides a solution that ismore efficient than an individual heating solution for 80% 2050 reduction 
+potential for 37% district heating nationally (Connollyet al., 2013a).
+HVO, solar may also play a part.
+
 \section{Literature Review}
 \label{LitR}
+
 ``How should Ireland set GHG targets in the residential sector?” 
+Choosing the Pathways which Create the Least Burden
+and Offer the Most Opportunity for Ireland
+\subsection{Modelling Approach}
+\label{LitR:Approach}
+\subsection{Supply}
+\label{LitR:Supply}
+\subsection{Demand}
+\label{LitR:Demand}
+\subsection{Assumptions}
+\label{LitR:Assumptions}
  % LIT REW INPUT DATA - EU \cite{Simoes2013TheProject} 
@@
 
-%Lorem ipsum dolor sit amet, consectetur adipiscing \citep{Fabioetal2013} elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud \citet{Blondeletal2008} exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit \citep{Blondeletal2008,FabricioLiang2013} anim id est laborum.
-
-%Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum see appendix~\ref{sec:sample:appendix}.
-
+TIM was built to help Ireland provide a pathway to achieve 2030 and 2050 targets.
+The residential sector plays a key part of this and it was built from the SEAI 2018 energy balance, the BER database and CSO data.
+The data filtration process used in the residential sector, derived a lot of conversation of the BER database, a group know as Ireland Energy Wiki by the National Retrofitting Modelling Group grew from building the residential sector, questions about the filters, default values, assumptions around thermal equations (EC..) and uncertainties were used to improve the data before it was used in TIM .
+TIM has many internal project deadines, which required extensive meetings about the data, constraints and results as the modelling was progressing.
+The data and results were presented to 30 external reviewers to provide feedback, this process proved invaluable having different expertise with different presepectives enabled TIM to fill in gaps.
+TIM will also be open platform to help improve the model and provide transparency.
+TIM’s residential sector was built in parallel with LEAP’s residential sector, so that residential sector can easily simulate policy.
+The demand for results from TIM has increased significantly with increased government ambition, TIM will provide input into CAP 2021 and Ireland’s carbon budgets, working with DECC and the CCAC respectively will also provide feedback and improve the robustness of TIM.
+for these reasons TIM residential sector is a state of the art modelling platform.
+\section{Results}
+\label{Results}
+\subsection{Trajectories}
+\label{LitR:Traj}
 %% The Appendices part is started with the command \appendix;
 %% appendix sections are then done as normal sections
+\section{Conclusion}
+\label{Conclusion}
+\section{Discussion}
+\label{Discussion}
 \section{Acknowledgements}

### 2021-05-21 manuscript.tex
Update 21 may 2021

@@
 \subsection{Climate Targets}
-\label{Intro:Climate Targerts}
+
 \par % Paragraph 1 - Intro
@@
 
-\subsection{Ireland: Residential Sector Background}
-\label{Intro:IreRSDBack}
-
-\par
-Ireland’s distinctive decarbonisation pathway, is further separated from other EU member states when focusing on the residential sector.
-This is largely down to Ireland’s high reliance on fossil fuels, low building thermal efficiency, larger dwellings and rural settlement patterns.
-This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}.
-The low thermal efficiency of Ireland's dwelling is discussed later in this section and highlighted in Fig.\ref{fig:BERAge}.
-Furthermore Irelands fossil fuel dependency is underlined by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.
-
-In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential GHG emissions.
-Residential coal GHG emissions has also declined to 6.9\% in 2018 \cite{EnergySEAI}.
-However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
-Gas which is only available in some urban areas, accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
-In summary, the residential sector is highly reliant on fossil fuels as shown in Fig.\ref{fig:ResFuel}, consequently Ireland currently has the lowest renewable heat of  among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
-
-\begin{figure}[!htbp]
- \centering
- \includegraphics[scale=0.5]{Figures/ResidentialFuelUsage.png}
- \caption{Historical Residential Fuel Consumption}
- \label{fig:ResFuel}
-\end{figure}
-% NOTE THE Y-AXIS IS NOT CORRECT - DATA FROM SEAI ENERGY DATA PORTAL
-
-The 2021 building regulations in 2021 require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock.
-According to Central Statistics Office (CSO) approx.
-85\% of residential buildings are 2005 or older.
-The BER database, which represents half of the dwelling stock, highlights the affect of construction year on the expected energy consumption, as shown in Fig.\ref{fig:BERAge}.
-The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
-The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, helps to better understand the housing stock.
-The BERs are required for new houses, rented houses, sold houses and renovated houses.
-
-\begin{figure}[!htbp]
- \centering
- \includegraphics[scale=0.5]{Figures/BERAge.png}
- \caption{Building Energy Rating by Age}
- \label{fig:BERAge}
-\end{figure}
-
-Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which results in higher excessive winter mortality \cite{Clinch2000HousingMortality}.
-BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
-The BER labels do therefore leave a gap between actual consumption and expected consumption, I will explain how I approached this problem in the literature review.
-
-Fuel Poverty which is a major contributor to the high excessive winter deaths, is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"}.
-An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}.
-Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty.
-These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}.
-As Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty, which include increased risk of catastrophic cardiovascular or cerebrovascular events \cite{OMeara2016AIreland}.
-Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
+\subsection{Ireland: Residential Sector}
+\label{Intro:IreRSD}
 
@@
 \end{figure}
+
+\subsubsection{Background}
+\label{Intro:IreRSDBack}
+Ireland’s distinctive decarbonisation pathway, is further separated from other EU member states when focusing on the residential sector.
+This is largely down to Ireland’s high reliance on fossil fuels, low building thermal efficiency, larger dwellings and rural settlement patterns.
+This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}.
+The low thermal efficiency of Ireland's dwelling is discussed later in this section and highlighted in Fig.\ref{fig:BERAge}.
+Furthermore Irelands fossil fuel dependency is underlined by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.
+
+In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential GHG emissions.
+Residential coal GHG emissions has also declined to 6.9\% in 2018 \cite{EnergySEAI}.
+However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+Gas which is only available in some urban areas, accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+In summary, the residential sector is highly reliant on fossil fuels, consequently Ireland currently has the lowest renewable heat of  among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
+
+The 2021 building regulations require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock.
+According to Central Statistics Office (CSO) approx.
+85\% of residential buildings are 2005 or older.
+The BER database, which represents half of the dwelling stock, highlights the affect of construction year on the expected energy consumption, as shown in Fig.\ref{fig:BERAge}.
+The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
+The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, which helps to better understand the housing stock.
+BERs are required for new houses, rented houses, sold houses and renovated houses.
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[scale=0.5]{Figures/BERAge.png}
+ \caption{Building Energy Rating by Age}
+ \label{fig:BERAge}
+\end{figure}
+
+Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which results in higher excessive winter mortality \cite{Clinch2000HousingMortality}.
+BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
+The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
+
+Fuel Poverty which is a major contributor to the high excessive winter deaths, is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"}.
+An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}.
+Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty.
+These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}.
+As Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty, which include increased risk of catastrophic cardiovascular or cerebrovascular events \cite{OMeara2016AIreland}.
+Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
 
@@
 IrDEA Up to 57\% feasible - same as Denmark DH
-37% of the Irish heat demand to district heatingshows that the district heating option provides a solution that ismore efficient than an individual heating solution for 80% 2050 reduction 
-potential for 37% district heating nationally (Connollyet al., 2013a).
+37\% Connollyet al., 2013a).
+of the Irish heat demand to district heatingshows that the district heating option provides a solution that ismore efficient than an individual heating solution for 80\% 2050 reduction
+
 HVO, solar may also play a part.
@@
 and Offer the Most Opportunity for Ireland
-\subsection{Modelling Approach}
-\label{LitR:Approach}
+\subsection{Modelling Methodology or Approach}
+\label{LitR:Model}
 \subsection{Supply}

### 2021-05-24 manuscript.tex
24May2021

@@
 To put Ireland's climate targets in perspective Fig.\ref{fig:GHG2030AGR} compares different countries 2030 climate target with a 2018 base year and share of agriculture GHG emissions.
-Ireland has ambitious targets, despite distinct barriers.
-The Climate Change Advisory Council (CCAC) will rely of some modelling tools to provide insights and set sectoral five year carbon budgets.
+It can be observed that Ireland has ambitious targets, despite distinct barriers.
 
@@
 \par % Paragraph 2 - ESOM
-A Energy System Optimisation Model (ESOM) can optimise the selection of a large number of dynamic variables to satisfy demand, while accounting for complex economic sectoral interconnections within each time period.
+The Climate Change Advisory Council (CCAC) will rely a range of modelling tools to provide insights to the Irish policymakers.
+One such model, is an Energy System Optimisation Model (ESOM), which can optimise the selection of a large number of dynamic variables to satisfy demand, while accounting for complex interconnections within each time period.
 The ESOM described in this paper is The Integrated MARKAL-EFOM System (TIMES) modelling framework, which has been developed and maintained by Energy Technology Systems Analysis Program (ETSAP) of the International Energy Agency (IEA).
@@
 The United Nations Framework Convention on Climate Change (UNFCCC) GHG reporting rules, as per Decision 24/CP.19, states that 2006  Intergovernmental Panel on Climate Change (IPCC) guidelines are to be adhere to for national GHG reporting.
-Under these guidelines, Ireland’s residential sector accounted for 14.2\%of GHGs emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
-Focusing on energy-related GHG emissions, the residential sector directly accounts for 24\% of GHG emissions.
+Under these guidelines, Ireland’s residential sector accounted for 14.2\% of GHGs emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+Focusing on energy-related GHG emissions only, the residential sector directly accounts for 24\% of GHG emissions.
 Fig.
 \ref{fig:Ireland2018GHG} shows the energy related GHG disaggregated between transport, heat and electricity.
-Where the residential sector is the direct source 47\% of heat and 30\% of electricity GHG emissions, the residential sector is therefore a pivotal sector in Ireland energy decarbonisation.
 
@@
 \end{figure}
+
+Where the residential sector is the direct source 47\% of heat and 30\% of electricity GHG emissions, the residential sector is therefore a pivotal sector in decarbonising energy in Ireland.
 
@@
 Gas which is only available in some urban areas, accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+Residential GHG emissions are still growing as preliminary results from the Environmental Protection Agency (EPA) has indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.
 In summary, the residential sector is highly reliant on fossil fuels, consequently Ireland currently has the lowest renewable heat of  among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
@@
 85\% of residential buildings are 2005 or older.
-The BER database, which represents half of the dwelling stock, highlights the affect of construction year on the expected energy consumption, as shown in Fig.\ref{fig:BERAge}.
+The BER database, which represents half of the dwelling stock, highlights the affect year of construction has on expected energy consumption, as shown in Fig.\ref{fig:BERAge}.
 The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
@@
 
-Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which results in higher excessive winter mortality \cite{Clinch2000HousingMortality}.
-BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
-The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
-
-Fuel Poverty which is a major contributor to the high excessive winter deaths, is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"}.
+Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which is a reason for Ireland's high excessive winter mortality \cite{Clinch2000HousingMortality}.
+Fuel Poverty which which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths.
 An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}.
@@
 These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}.
-As Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty, which include increased risk of catastrophic cardiovascular or cerebrovascular events \cite{OMeara2016AIreland}.
+Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}.
 Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
-
+\par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
+The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
 \subsection{Residential Energy Services}
@@
 
-Space heating is the largest residential energy service demand, accounted for 61.8\% of final energy usage.
-Other residential energy services and their share of final energy in 2018 is shown in \ref{fig:Ireland2018ServiceDemands}.
+Space heating is the largest residential energy service demand, accounted for 61.8\% of final energy usage in 2018.
+Other residential energy services and their share of final energy is shown in Fig.\ref{fig:Ireland2018ServiceDemands}.
 As building thermal efficiency and heating technology efficiency improves, both space and water heating energy would be expected to decline.
-The new A-rated dwellings can be used to predict the share of future energy service demands.
+The new A-rated dwellings can be used as a guide to show the share of future energy service demands.
 From the BER database, it can be seen that A-rated dwellings use less than half space heating energy of the average dwelling and A-rated water heating energy decreases by about 40\% compared to the average A-rated dwelling.
@@
         teal!20
-    }, explode = 0.4, text=legend, radius=3]{61.8/Space Heating, 2.2/Cooking, 18.5/Water Heating, 3.8/Lighting, 12.8/Appliances,0.9/Pumps \& Fans}
+    }, explode = 0.6, text=legend, radius=2.4]{61.8/Space Heating, 2.2/Cooking, 18.5/Water Heating, 3.8/Lighting, 12.8/Appliances,0.9/Pumps \& Fans}
     \end{tikzpicture}
@@
 
-Notably space cooling is absent from Fig \ref{fig:Ireland2018ServiceDemands}, but space cooling is accounted for in appliances.
-The Policy Oriented Tool for Energy and Climate Change Impact Assessment (POTEnCIA) project disaggregated Ireland's residential space cooling from appliances and have output space cooling energy consumption data, shown in Fig.\ref{fig:Cooling} a significant increase in space cooling energy is expected in the Ireland residential sector.
+Notably absent from Fig.\ref{fig:Ireland2018ServiceDemands}, is space cooling, which is accounted for in appliances.
+The Policy Oriented Tool for Energy and Climate Change Impact Assessment (POTEnCIA) project disaggregated Ireland's residential space cooling from appliances and have output expected space cooling energy consumption projections are shown in Fig.\ref{fig:Cooling}.
 
-\begin{figure}[h]
+\begin{figure}[!htbp]
 \centering
@@
 The SEAI residential energy report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} provides high resolution detail on energy demands in the residential sector.
-Variables such as disposable income, appliance ownership, dwelling location, type of occupier status (owner/rented ), BER label and age of dwelling.
+Variables such as disposable income, appliance ownership, dwelling location, type of occupier status (owner/rented ), BER labels and age of dwelling.
 
@@
 The energy-related GHG emissions are expected to be reduced beyond the 51\% national target by 2030 compared to 2018, due to the higher cost-effectiveness of abating GHG emissions when compared with agriculture.
-Despite these high target ambitions in the energy-related GHG emissions, preliminary results from the Environmental Protection Agency (EPA) has indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.
-The CAP2019 Marginal Abatement Cost Curve (MACC), which was developed by McKinsey and Company's modelling tools and is based on EPA's GHG emission projection report \cite{EPA2018EPAReport} found electricity & transport are the most cost effective for decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
-The future decarbonisation of the residential sector is dependant on receiving low carbon intensity electricity.
+The CAP2019 Marginal Abatement Cost Curve (MACC), which was developed by McKinsey and Company's modelling tools and is based on EPA's GHG emission projection report \cite{EPA2018EPAReport} found electricity \& transport are the most cost effective for decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
+Therefore a cost effective solution to decarbonising the residential sector, is to receive low carbon electricity and increasing the share of electricity usage.
 
@@
 The first strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}.
-Some of these actions included:
-\begin{itemize}
-    \item{Reducing fossil fuel use in the sector and moving to electricity to provide heating and hot water in buildings}
-    \item{500,000 building retrofits to a BER B2 / cost optimal equivalent or carbon equivalent}
-    \item{Installation of 600,000 heat pumps, with 400,000 of these in existing buildings}
-    \item{Development of district heating, including two initiatives of municipal scale with the potential to provide heat equivalent to the needs of about 50,000 homes}
-    \item{Complete the roll-out of the Support Scheme for Renewable Heat}
-    \item{Increasing the number of Sustainable Energy Communities to 1,500}
-\end{itemize}
+The densification of Ireland's main cities, will provide opportunities to improve the efficiency of heating, electricity and transport within and between cities.
 
-\begin{figure}[!htbp]
- \centering
- \includegraphics[scale=1.0]{Figures/CAP2019MACC.png} 
- \caption{Climate Action Plan 2019 - Marginal Abatement Cost Curve}
- \label{fig:CAP2019:MACC} \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
-\end{figure}
-
-\par
 The CAP2019 MACC, shown in Fig.\ref{fig:CAP2019:MACC} provides $>300$ of the most cost-effective technologies options to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels.
 The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_e_q$.
-The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_e_q$.
-over the 2021 to 2030 period.
+The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_e_q$ over the 2021 to 2030 period.
 The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_e_q$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
@@
 
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[scale=1]{Figures/CAP2019MACC.png} 
+ \caption{Climate Action Plan 2019 - Marginal Abatement Cost Curve}
+ \label{fig:CAP2019:MACC} \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
+\end{figure}
 Where, the retrofitting of oil boilers is expected to reduce the Total Cost of Ownership (TCO) (ie.
@@
 \par
-The decarbonisation of the power generation sector is well underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}.
+In Ireland the decarbonisation of the electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}.
 Under CAP2019 the continued growth and decarbonisation of the electricity will be the fundamental to lowering energy-related GHG in not only electricity, but in heat and transport also \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
@@
 About 15\% of global GHG emissions have either an associated cost or tax, this figure is increasing incrementally \cite{WorldBankGroup2020State2020}.
-The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise Ireland to invest in renewable electricity to achieve legally binding renewable targets under 2009/28/EC.
+\par The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise Ireland to invest in renewable electricity to achieve legally binding renewable targets under 2009/28/EC.
 REFIT operates by guaranteeing  new renewable generation and biomass a minimum price for electricity over a 15 year period.
@@
 In 2020 REFIT was replaced by the Renewable Electricity Support Scheme (RESS), which supports to renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target.
-RESS delivers for communities to participate in renewable energy projects and increasing technology diversity by broadening the renewable electricity technology mix, where almost 800 MW of solar has been approved in RESS-1 \cite{Eirgrid2020RenewableResults}, this builds alot on the current solar capacity of 16 MW, while making significant progress towards the CAP2019 target of 1500 MW by 2030.
+RESS prioritizes communities to participate in renewable energy projects and increasing technology diversity by broadening the renewable electricity technology mix, where almost 800 MW of solar has been approved in RESS-1 \cite{Eirgrid2020RenewableResults}, this builds alot on the current solar capacity of 16 MW, while making significant progress towards the CAP2019 target of 1500 MW solar by 2030.
 
 \subsubsection{Heat}
-%Gas Network Ireland (GNI) anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}.
-Sustainable Energy Authority of Ireland (SEAI) have been played a leading role in the uptake of building fabric upgrades and heat pumps installations in existing homes.
+Sustainable Energy Authority of Ireland (SEAI) have played a leading role in the uptake of building fabric upgrades and heat pumps installations in existing homes.
 SEAI provide numerous grants to homeowners, increase public awareness of the benefits and are working on simplifying the process.
-In parallel, new dwellings must achieve a minimum of 20\% renewable energy, the ambient heat used in heat pumps is included as a renewable energy, often heat pumps are the preferred heating technology despite the high capital costs.
-As mentioned in Section\ref{Intro:IreRSDBack} new buildings in Ireland produce 70\% less GHG emissions than building in 2005 and Section\ref{Intro:Policy} mentioned CAP2019 plans to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030.
-The energy efficiency first approach in how Ireland aims to decarbonize the residential sector.
-Domestic supports for renewable heat or energy efficiency include:
+In parallel, new dwellings must achieve a minimum of 20\% renewable energy.
+The ambient heat used in heat pumps is included as a renewable energy, heat pumps are therefore the preferred heating technology despite the high capital costs.
+As mentioned in Section\ref{Intro:IreRSDBack} new buildings in Ireland produce 70\% less GHG emissions than building in 2005 and with CAP2019 plans to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030, shows Irelands energy efficiency first approach in the residential sector.
+in how Ireland aims to decarbonize the residential sector.
+Domestic supports for renewable heating technologies and energy efficiency include:
 \begin{itemize}
@@
     \item {Solar Water Heating}
+    \item{ Solar Photovoltaics }
 \end{itemize}
 
-\subsubsection{Gas \& District Heating}
+\subsubsection{Gas}
 
-In a bid not to be left behind in a ``net-zero" future, Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019AIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-GNI plans to met ``net-zero" in 2050 by using 50\% Carbon Capture and Storage (CCS), 37\% biomethane and about 13\% Hydrogen\cite{Ervia2019AIreland}.
+In a bid not to be left behind in a ``net-zero" future, Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019VisionIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+GNI anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}, with lower indigenous gas expected GNI plans to met ``net-zero" in 2050 by using 50\% Carbon Capture and Storage (CCS), 37\% biomethane and 13\% Hydrogen\cite{Ervia2019VisionIreland}.
 The gas network currently reaches 40\% or 680,000 dwellings in urban areas, but with the ``Compact Growth" as mentioned in section \ref{Intro:Policy} a higher share of building may have access to the gas network in the future.
-Increasing biomethane from an agricultural source in the gas network is the most expensive option in CAP2019's MACC shown in Fig.\ref{fig:CAP2019:MACC}.
+ALthough increasing biomethane from an agricultural source in the gas network is the most expensive option in CAP2019's MACC shown in Fig.\ref{fig:CAP2019:MACC}.
 
-Ireland has one of the lowest shares of District Heating (DH) in Europe at less than 1\%  \cite{Gartland2016AIreland} \cite{Harney2020DeterminingIreland}
-In Ireland,  DH networks are a new, despite being used for decade across central Europe. CAP2019 plans to have approx. 0.4\% of heating provided by DH in 2030, where the  Research in Ireland h
-IrDEA Up to 57\% feasible - same as Denmark DH
-37\% Connollyet al., 2013a).
-of the Irish heat demand to district heatingshows that the district heating option provides a solution that ismore efficient than an individual heating solution for 80\% 2050 reduction
+\subsubsection{District Heating}
+Ireland has one of the lowest shares of District Heating (DH) in Europe at less than 1\%  \cite{Gartland2016AIreland}.
+DH networks are a mature technology, being used for almost 100 years across Europe.
+``Compact Growth" and advancement in DH control will strengthen the case for DH networks to play a part in decarbonisaing the residential sector.
+CAP2019 plans to have approx.
+0.4\% of heating provided by DH in 2030, research using TIMES and EnergyPLAN has shown that 37\% of heating provided by DH is most efficient if Ireland is to reduce GHG emissions by 80\% in 2050, compared to 1990 \cite{Thellufsen2019ImplementingIreland}.
+While Irish District Energy Association (IrDEA) estimate
+to 57\% of heat can be sourced from DH, if supporting policy and regulation was in place.
+District Heating will likely play a role, but Gartland \cite{Gartland2016AIreland} points out a range of organisational, technical, regulatory and economic barriers to DH growth in Ireland.
+The most prolific barriers being the lack of guidelines, regulations, policies, frameworks or standards for DH which creates high risk and uncertainty when planning DH systems \cite{Gartland2016AIreland}.
+Gartland further explains that \textit{choices made regarding energy supply have been, and continue to be, economically evaluated on a very short-term basis}, the use of an long-term ESOM may provide new insights to climate policy.
 
-HVO, solar may also play a part.
+\subsubsection{Hydrotreated Vegetable Oil}
+While it maybe optimal to use Hydrotreated Vegetable Oil (HVO) in transport, it is also considered in the residential sector, as it can provide a cheap alternative for the 700,000 Irish dwellings using home heating oil ( mostly kerosene).
+HVO which meets the EU sustainability criteria, reduce GHG emissions by 60\% as compared to fossil fuels \cite{Soam2019FactorsSweden}.
+The capacity of HVO is growing in Europe, but HVO is currently about twice the price of kerosene.
+HVO can be used to replace kerosene, albeit with cost of up to €400 per oil boiler, however there is no policy for HVO in the residential sector.
 
@@
 \label{LitR}
-
+%PICKUPHERE
 ``How should Ireland set GHG targets in the residential sector?” 
@@
 \label{LitR:Model}
+The TIM approach.....
 \subsection{Supply}

### 2021-05-26 manuscript.tex
26 May2021

@@
 %% Text of abstract
-
-Write after paper is completed %Ireland's residential sector has particularly underachieved in previous climate targets.
+%Ireland's residential sector has particularly underachieved in previous climate targets.
 Located on the peripherals of Europe, with a high share of low thermal efficient detached dwellings with carbon intensive heating fuels, provides some distinctive barriers to decarbonising Ireland's residential sector.
@@
 %The TIMES-Ireland Model (TIM) is a newly developed optimisation model of the Irish energy system, which calculates the  cost-optimal decarbonisation pathway to meet future energy service demands while achieving Ireland's legally binding 2030 and 2050 climate targets.
+Write after paper is completed
 
@@
 \label{sec:Introduction}
+
+Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways, and interpreting the results.
+This section looks at Ireland's climate targets, energy policy and distinctive residential sector, before examining how researchers have used Energy system optimization models (ESOMs) to analyse the residential sector.
+Section \ref{LitR} describes the methodology used to develop the residential sector of TIMES Ireland Model (TIM). The sector outlines the data used and calculations used ......
+What does new model do that previous didn’t?
+Does new policy context require new modelling approach?
+How does new modelling structure provide insights (give examples) in a way that previous model didn’t?
+% ``How should Ireland set GHG targets in the residential sector?” 
+This is done by examining the current climate targets, EU dirictives and Irish and EU policy, then analysiing future policy and looking at ESOMs
+
 \subsection{Climate Targets}
@@
 The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
-\subsection{Residential Energy Services}
-\label{Intro:EnergyServices}
-
-Space heating is the largest residential energy service demand, accounted for 61.8\% of final energy usage in 2018.
-Other residential energy services and their share of final energy is shown in Fig.\ref{fig:Ireland2018ServiceDemands}.
-As building thermal efficiency and heating technology efficiency improves, both space and water heating energy would be expected to decline.
-The new A-rated dwellings can be used as a guide to show the share of future energy service demands.
-From the BER database, it can be seen that A-rated dwellings use less than half space heating energy of the average dwelling and A-rated water heating energy decreases by about 40\% compared to the average A-rated dwelling.
-However Pumps \& Fans energy increases as mechanical ventilation is used over natural ventilation and Lighting usage in A-rated is more than double the average household.
-Cooking and Appliances energy usage would be expected to remain stable.
-
-\begin{figure}[!htbp]
-\centering
-    \begin{tikzpicture}
-    \tikzstyle{every node}=[font={\bf}]
-    \pie[   color = { yellow!90!black, green!60!black, blue!60,
-        red!70,
-        gray!80,
-        teal!20
-    }, explode = 0.6, text=legend, radius=2.4]{61.8/Space Heating, 2.2/Cooking, 18.5/Water Heating, 3.8/Lighting, 12.8/Appliances,0.9/Pumps \& Fans}
-    \end{tikzpicture}
-    \caption{Ireland 2018 Energy Service Demands}
-    \label{fig:Ireland2018ServiceDemands}
-\end{figure}
-
-Notably absent from Fig.\ref{fig:Ireland2018ServiceDemands}, is space cooling, which is accounted for in appliances.
-The Policy Oriented Tool for Energy and Climate Change Impact Assessment (POTEnCIA) project disaggregated Ireland's residential space cooling from appliances and have output expected space cooling energy consumption projections are shown in Fig.\ref{fig:Cooling}.
-
-\begin{figure}[!htbp]
-\centering
-\begin{tikzpicture}
-\begin{axis}[
-    x tick label style={
-		/pgf/number format/1000 sep=},xlabel={Year},
-	y tick label style={
-	    /pgf/number},
-		ylabel={Energy Consumption($ktoe$)},
-		 scaled y ticks = false,
-    xmin=2000, xmax=2050,
-    ymin=0, ymax=40,
-    xtick={2000,2010,2020,2030,2040,2050},
-    ytick={0,10,20,30,40},
-    ymajorgrids=true,
-]
-\addplot[
-    color=blue,
-    mark=square,
-    ]
-    coordinates {
-    (2000,0.6)(2010,3.1)(2020,8.4)(2030,16.3)(2040,26.1)(2050,37)
-    };
-\end{axis}
-\end{tikzpicture}
-    \caption{Space Cooling Residential Energy Consumption (Source: POTEnCIA) }
-    \label{fig:Cooling}
-\end{figure}
-
-The SEAI residential energy report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} provides high resolution detail on energy demands in the residential sector.
-Variables such as disposable income, appliance ownership, dwelling location, type of occupier status (owner/rented ), BER labels and age of dwelling.
 
@@
 
-The energy-related GHG emissions are expected to be reduced beyond the 51\% national target by 2030 compared to 2018, due to the higher cost-effectiveness of abating GHG emissions when compared with agriculture.
-The CAP2019 Marginal Abatement Cost Curve (MACC), which was developed by McKinsey and Company's modelling tools and is based on EPA's GHG emission projection report \cite{EPA2018EPAReport} found electricity \& transport are the most cost effective for decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
+The CAP2019 Marginal Abatement Cost Curve (MACC),  \cite{EPA2018EPAReport} found electricity \& transport are the most cost effective for decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
 Therefore a cost effective solution to decarbonising the residential sector, is to receive low carbon electricity and increasing the share of electricity usage.
@@
 HVO can be used to replace kerosene, albeit with cost of up to €400 per oil boiler, however there is no policy for HVO in the residential sector.
+biofuel obligation scheme - Internationally there is a limited supply of HVO available and the
+demand for the fuel is high.HVO has been produced in the past at Ireland’s only refinery at Whitegate and
+potentially could be again with the appropriate investment. Not HVO100, but higher concentration of HVO than current blending capacity
 
@@
 \label{LitR}
-%PICKUPHERE
+
 ``How should Ireland set GHG targets in the residential sector?” 
@@
 The TIM approach.....
+%a newly formed group with members from different universities, industries and semi-state agencies within Ireland.
+Following this, other datasets were then used to develop the full housing stock, based on the correlation of year of construction and building energy ratings.
+\subsection{Residential Energy Services}
+\label{Intro:EnergyServices}
+
+Space heating is the largest residential energy service demand, accounted for 61.8\% of final energy usage in 2018.
+Other residential energy services and their share of final energy is shown in Fig.\ref{fig:Ireland2018ServiceDemands}.
+As building thermal efficiency and heating technology efficiency improves, both space and water heating energy would be expected to decline.
+The new A-rated dwellings can be used as a guide to show the share of future energy service demands.
+From the BER database, it can be seen that A-rated dwellings use less than half space heating energy of the average dwelling and A-rated water heating energy decreases by about 40\% compared to the average A-rated dwelling.
+However Pumps \& Fans energy increases as mechanical ventilation is used over natural ventilation and Lighting usage in A-rated is more than double the average household.
+Cooking and Appliances energy usage would be expected to remain stable.
+
+\begin{figure}[!htbp]
+\centering
+    \begin{tikzpicture}
+    \tikzstyle{every node}=[font={\bf}]
+    \pie[   color = { yellow!90!black, green!60!black, blue!60,
+        red!70,
+        gray!80,
+        teal!20
+    }, explode = 0.6, text=legend, radius=2.4]{61.8/Space Heating, 2.2/Cooking, 18.5/Water Heating, 3.8/Lighting, 12.8/Appliances,0.9/Pumps \& Fans}
+    \end{tikzpicture}
+    \caption{Ireland 2018 Energy Service Demands}
+    \label{fig:Ireland2018ServiceDemands}
+\end{figure}
+
+Notably absent from Fig.\ref{fig:Ireland2018ServiceDemands}, is space cooling, which is accounted for in appliances.
+The Policy Oriented Tool for Energy and Climate Change Impact Assessment (POTEnCIA) project disaggregated Ireland's residential space cooling from appliances and have output expected space cooling energy consumption projections are shown in Fig.\ref{fig:Cooling}.
+
+\begin{figure}[!htbp]
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    x tick label style={
+		/pgf/number format/1000 sep=},xlabel={Year},
+	y tick label style={
+	    /pgf/number},
+		ylabel={Energy Consumption($ktoe$)},
+		 scaled y ticks = false,
+    xmin=2000, xmax=2050,
+    ymin=0, ymax=40,
+    xtick={2000,2010,2020,2030,2040,2050},
+    ytick={0,10,20,30,40},
+    ymajorgrids=true,
+]
+\addplot[
+    color=blue,
+    mark=square,
+    ]
+    coordinates {
+    (2000,0.6)(2010,3.1)(2020,8.4)(2030,16.3)(2040,26.1)(2050,37)
+    };
+\end{axis}
+\end{tikzpicture}
+    \caption{Space Cooling Residential Energy Consumption (Source: POTEnCIA) }
+    \label{fig:Cooling}
+\end{figure}
+
+The SEAI residential energy report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} provides high resolution detail on energy demands in the residential sector.
+Variables such as disposable income, appliance ownership, dwelling location, type of occupier status (owner/rented ), BER labels and age of dwelling.
+
 \subsection{Supply}
@@
 \label{LitR:Demand}
+%Space Heating demand - BER space heating is only within certain months and assumes 21C heating..
 \subsection{Assumptions}
 \label{LitR:Assumptions}
+Macroeconomic Assumptions 
+Capacity
+Costs
+Efficiency
+Savings
  % LIT REW INPUT DATA - EU \cite{Simoes2013TheProject} 

### 2021-05-27 manuscript.tex
27 May 2021

@@
 \usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{amsfonts}
 \usepackage{pgf-pie}
@@
 \usepackage{pgfplots}
+\usepackage{amsmath}
 
@@
 Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways, and interpreting the results.
-This section looks at Ireland's climate targets, energy policy and distinctive residential sector, before examining how researchers have used Energy system optimization models (ESOMs) to analyse the residential sector.
-Section \ref{LitR} describes the methodology used to develop the residential sector of TIMES Ireland Model (TIM). The sector outlines the data used and calculations used ......
-What does new model do that previous didn’t?
-Does new policy context require new modelling approach?
-How does new modelling structure provide insights (give examples) in a way that previous model didn’t?
-% ``How should Ireland set GHG targets in the residential sector?” 
-This is done by examining the current climate targets, EU dirictives and Irish and EU policy, then analysiing future policy and looking at ESOMs
-
-\subsection{Climate Targets}
-
-\par % Paragraph 1 - Intro
-Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to reduce Greenhouse Gas (GHG) emissions by 51\% in 2030 compared to 2018.
-The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}.
-These targets accounts for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to develop sectoral five year carbon budgets to aid progress of the forementioned long-term targets.
-\par The EU's first Nationally Determined Contribution (NDC) which had a 40\% GHG reduction target (first NDC), has increased to at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}.
-Ireland's Climate Action Plan 2019 (CAP2019) which complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-Climate Action Plan 2021 will have to be more ambitious than its predecessor, as it must at least reflect EU's updated NDC target.
-Assuming the share of contribution of ETS and Ireland's Non-ETS remains the same, Ireland's non-ETS GHG emission 2030 target would change from 30\% to 45\%.
-Overall Ireland's national 2030 target would be about 15\% more ambitious than the EU's updated NDC.
-There are also renewable energy targets to consider for transport (RES-T), heat (RES-H) and electricity (RES-E) under renewable directive (EU) 2018/2001, and energy efficiency targets under (EU) 2018/2002.
-An overview of Ireland's climate targets are shown in Fig.\ref{fig:IrelandClimatetargets}.
-
-\begin{figure}[!htbp]
- \centering
- \includegraphics[scale=0.4]{Figures/Non-ETS-Share.png} 
- \caption{Overview of Ireland's 2030 Climate Targets}
- \label{fig:IrelandClimatetargets}
-\end{figure}
-
-Ireland is a unique EU member state, located on the peripherals of Europe, with a the second highest GDP per capita \cite{EurostatReal} (Higher GDP per capita translates to higher non-ETS reduction target), the largest proportion of agriculture GHG emissions at 34\% in 2018  \cite{EPA2020Irelands1990-2018}, and a high dependency on fossil fuel imports.
-Ireland has been afforded high non-ETS flexibility allowances under (EU) 2020/2126, to compensate for these barriers, but Ireland still has ambitious national climate targets.
-To put Ireland's climate targets in perspective Fig.\ref{fig:GHG2030AGR} compares different countries 2030 climate target with a 2018 base year and share of agriculture GHG emissions.
-It can be observed that Ireland has ambitious targets, despite distinct barriers.
-
-\begin{figure}[!htbp]
- \centering
- \includegraphics[scale=0.5]{Figures/GHGtargetAGR1.png} 
- \caption{Most Ambitious 2030 GHG Targets}
- \label{fig:GHG2030AGR}
-\end{figure}
-
-\subsection{Energy System Optimisation Models}
-\label{Intro:ESOM}
-\par % Paragraph 2 - ESOM
-The Climate Change Advisory Council (CCAC) will rely a range of modelling tools to provide insights to the Irish policymakers.
-One such model, is an Energy System Optimisation Model (ESOM), which can optimise the selection of a large number of dynamic variables to satisfy demand, while accounting for complex interconnections within each time period.
-The ESOM described in this paper is The Integrated MARKAL-EFOM System (TIMES) modelling framework, which has been developed and maintained by Energy Technology Systems Analysis Program (ETSAP) of the International Energy Agency (IEA).
-TIMES is a ``hybrid" ESOM, which combines high resolution ``bottom-up" technological detail and ``top-down" macroeconomic variables covering all energy sectors.
-\par In addition to TIMES there are other ESOM frameworks available such as EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}.
-However the expertise among energy policy researchers in Ireland combined with the global network of ETSAP modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways.
-\par Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model, and then updated with specific national data and assumptions.
-The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.
-While TIMES models can cover all the energy-related sectors, it can also be used to explore the residential sector in detail.
-Some studies focusing on the residential sector in TIMES include, Li who studied the preference for heating technology choice in UK \cite{Li2018IncorporatingModel}, Petrovic examined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed the affect higher thermal efficiency in buildings has on co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}.
-There has also been some studies residential sector in Ireland, using previous versions of Irish TIMES and now out-dated climate targets.
-Therefore, there is a need for a new TIMES model to be built specifically to explore Ireland's new 2030 and 2050 climate targets and sectoral five-year carbon budgets, the ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
+This section looks at Ireland's climate targets, energy policy and distinctive residential sector, before examining how researchers have previously used Energy system optimization models (ESOMs) to analyse the residential sector.
+\par Section \ref{LitR} describes the methodology used to develop the residential sector of TIMES Ireland Model (TIM).
+The sector outlines the data used, along with the applied constraints, technologies, costs and demands.
+The section outlines the strengths and weakness of the newly developed model and how it differs from other models.
+\par Section \ref{Results} discusses the results- scenarios
+\par Section \ref{Conclusion} Conclusion
 
@@
 
-The United Nations Framework Convention on Climate Change (UNFCCC) GHG reporting rules, as per Decision 24/CP.19, states that 2006  Intergovernmental Panel on Climate Change (IPCC) guidelines are to be adhere to for national GHG reporting.
-Under these guidelines, Ireland’s residential sector accounted for 14.2\% of GHGs emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+Ireland’s distinctive energy sectors are further separated from other EU member states when focusing on the residential sector.
+This is largely down to Ireland’s high reliance on fossil fuels, low building thermal efficiency, larger dwellings and rural settlement patterns.
+This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}.
+The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts, firstly the 2021 building regulations require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}, and secondly according to Central Statistics Office (CSO) data approx.
+85\% of residential buildings are 2005 or older.
+Furthermore Irelands fossil fuel dependency is underlined by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.
+
+In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential GHG emissions.
+Residential coal GHG emissions has also declined to 6.9\% in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+Gas availability is limited to some urban areas, and accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+The residential sector relied on fossil fuels for over 90\% of thermal energy in 2018, consequently Ireland currently has the lowest renewable heat of among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
+Preliminary results from the Environmental Protection Agency (EPA) have indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.
+
+\subsubsection{Current Greenhouse Gas Emissions}
+\label{Intro:GHG}
+
+Under 2006  Intergovernmental Panel on Climate Change (IPCC) guidelines, Ireland’s residential sector accounted for 14.2\% of GHGs emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
 Focusing on energy-related GHG emissions only, the residential sector directly accounts for 24\% of GHG emissions.
@@
 
-\subsubsection{Background}
-\label{Intro:IreRSDBack}
-Ireland’s distinctive decarbonisation pathway, is further separated from other EU member states when focusing on the residential sector.
-This is largely down to Ireland’s high reliance on fossil fuels, low building thermal efficiency, larger dwellings and rural settlement patterns.
-This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}.
-The low thermal efficiency of Ireland's dwelling is discussed later in this section and highlighted in Fig.\ref{fig:BERAge}.
-Furthermore Irelands fossil fuel dependency is underlined by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.
+\subsection{Climate Targets}
 
-In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential GHG emissions.
-Residential coal GHG emissions has also declined to 6.9\% in 2018 \cite{EnergySEAI}.
-However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
-Gas which is only available in some urban areas, accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
-Residential GHG emissions are still growing as preliminary results from the Environmental Protection Agency (EPA) has indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.
-In summary, the residential sector is highly reliant on fossil fuels, consequently Ireland currently has the lowest renewable heat of  among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
+Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to reduce Greenhouse Gas (GHG) emissions by 51\% in 2030 compared to 2018.
+The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}.
+These targets accounts for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out make sectoral five year carbon budgets  a legal requirement to aid progress of the forementioned long-term targets.
+\par The EU's first Nationally Determined Contribution (NDC) which had a 40\% GHG reduction target (first NDC), has increased to at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}.
+Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+Climate Action Plan 2021 will have to be more ambitious than its predecessor, CAP2019 as it must reflect EU's updated NDC target and take account of the national set target.
+There are also renewable energy targets to consider for transport (RES-T), heat (RES-H) and electricity (RES-E) under renewable directive (EU) 2018/2001, and energy efficiency targets under (EU) 2018/2002.
 
-The 2021 building regulations require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock.
-According to Central Statistics Office (CSO) approx.
-85\% of residential buildings are 2005 or older.
-The BER database, which represents half of the dwelling stock, highlights the affect year of construction has on expected energy consumption, as shown in Fig.\ref{fig:BERAge}.
-The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
-The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, which helps to better understand the housing stock.
-BERs are required for new houses, rented houses, sold houses and renovated houses.
+\subsection{Energy System Optimisation Models}
+\label{Intro:ESOM}
+An Energy System Optimisation Model (ESOM), can be used to optimise the selection of a large number of dynamic variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
+\par Some of the main ESOM include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016EnergyProgramme}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}.
+The Integrated MARKAL-EFOM System (TIMES) modelling framework was selected to develop an Irish model, a description of the framework is outlined in the following subsection.
 
-\begin{figure}[!htbp]
- \centering
- \includegraphics[scale=0.5]{Figures/BERAge.png}
- \caption{Building Energy Rating by Age}
- \label{fig:BERAge}
-\end{figure}
+\subsubsection{TIMES}
+\label{Intro:TIMES}
 
-Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which is a reason for Ireland's high excessive winter mortality \cite{Clinch2000HousingMortality}.
-Fuel Poverty which which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths.
-An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}.
-Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty.
-These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}.
-Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}.
-Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
-\par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
-The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
+\begin{equation} 
+\begin{split}
+NPV = {\sum_{r=1}^{R}  \sum_{y\in YEARS}(1 + d_{r,y})^{REFYR-y}}
+\cdot {ANNCOST(r,y)} 
+\end{split}
+%\[ \sum_{n=1}^{\infty} 2^{-n} = 1 \]
+\label{Eq:ObjectiveFunction}
+\end{equation}
+where: $NPV$ is the net present value of the total cost for all regions (the TIMES objective function);
+$ANNCOST(r,y)$ is the total annual cost in region r and year y; $d_{r,y}$ is the general discount rate;
+$REFYR$ is the reference year for discounting;
+YEARS is the set of years for which there are costs, including all years in the horizon, plus past years (before the initial period) if costs have been defined for past investments, plus a number of years after end of the horizon EOH where some investment and dismantling costs are still being incurred, as well as the Salvage Value; and
+R is the set of regions in the area of study.
+$ANNCOST$ is quite complex and is postponed until section 6.2 of PART II
+Due to the data driven nature of TIMES (see section 2.5), all TIMES constraints are activated and defined by specifying some attributes.
+Attributes are attached to processes, to commodities, to flows, or to special variables that have been created to define new TIMES features.
+Indeed, TIMES
+
+TIMES has been developed and maintained by Energy Technology Systems Analysis Program (ETSAP) of the International Energy Agency (IEA).
+is a bottom-up optimisation model generator for energy-environment sys-tems analysis at various levels of spatial, temporal and sectoral resolution ...which combines high resolution ``bottom-up" technological detail and ``top-down" macroeconomic variables covering all energy sectors.
+ % It calculates the lowest-cost configuration of energy fuels and technologies whichmeet future energy demands, while respecting technical, environmental, economic, social and policy constraints.
+Key inputsand  constraints  include  primary  energy  resource  availability  and  costs,  the  technical  and  cost  evolution  of  new  mitigationoptions and maximum feasible uptake rates of new technologies.
+Alternatively, TIM can be used to assess the implications ofcertain policies, namely regulatory or technology target-setting (for example, biofuels blending obligation or sales/stock share80target for electric vehicles).
+\par  However the expertise among energy policy researchers in Ireland combined with the global network of ETSAP modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways.
+\par Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model, and then updated with specific national data and assumptions.
+The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.
+Therefore, there is a need for a new TIMES model to be built specifically to explore Ireland's new 2030 and 2050 climate targets and sectoral five-year carbon budgets, the ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
+GROWTH CONSTRAINT -A specific type of constraint may be defined to limit the share of process (p) in the total production of commodity (c).
+The constraint indicates that the flow of commodity (c) from/to process (p) - CommodiTypically, a growth constraint is of the following generic form (ignoring several indices for clarity: \cite{Loulou2016EnergyProgramme}
+\begin{equation} 
+\begin{split}
+VAR\_CAP(t + 1) \leq (1 + GROWTH^{M(t+1)-M(t)}) \cdot VAR\_CAP(t)+K
+\end{split}
+\label{Eq:GrowthConstraint}
+\end{equation}
+The GROWTH coefficient is defined as a new attribute of the technology, and represents the maximum annual growth allowed for the capacity.
+The quantity M(t+1)-M(t) is the number of years between the milestones of periods t and t+1.
+The constant K is useful whenever the technology has no capacity initially, in order to allow capacity to build over time (if K were absent and initial capacity is zero, the technology would never acquire any capacity)
+Note that the sign of the constraint may also be of the "larger than or equal to" type to express a maximum rate of abandonment, in which case the "+" sign is replaced by a "–" sign in the right-hand-side of the constraint.
+Equality is also allowed, but must be used only exceptionally in order to avoid railroading of the model.
+5.4.11
+\subsection{Residential Sector research using ESOMs}
+\label{Intro:ESOMRes}
+While TIMES models can cover all the energy-related sectors, it can also be used to explore the residential sector in detail.
+Some studies focusing on the residential sector in TIMES include, Li who studied the preference for heating technology choice in UK \cite{Li2018IncorporatingModel}, Petrovic examined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed the affect higher thermal efficiency in buildings has on co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}.
+There has also been some studies residential sector in Ireland, using previous versions of Irish TIMES and now out-dated climate targets.
 
@@
 The CAP2019 Marginal Abatement Cost Curve (MACC),  \cite{EPA2018EPAReport} found electricity \& transport are the most cost effective for decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
-Therefore a cost effective solution to decarbonising the residential sector, is to receive low carbon electricity and increasing the share of electricity usage.
+Therefore the most cost effective solution to decarbonising the residential sector, is to receive higher amounts of low carbon electricity.
 
@@
 The CAP2019 MACC, shown in Fig.\ref{fig:CAP2019:MACC} provides $>300$ of the most cost-effective technologies options to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels. 
-The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_e_q$.
-The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_e_q$ over the 2021 to 2030 period.
-The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_e_q$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_{eq}$.
+The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_{eq}$ over the 2021 to 2030 period.
+The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_{eq}$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
 The options specific to the residential sector are:
+
 \begin{itemize}
@@
 
-\section{Literature Review}
+\section{Methodology}
 \label{LitR}
+\subsection{Overview}
+\label{Overview}
+%The TIMES-Ireland Model (TIM) is an optimisation model of the Irish energy system, which calculates the cost-5optimal fuel and technology mix to meet future energy service demands in the transport, buildings, industry and agriculturesectors, while respecting constraints in greenhouse-gas emissions, primary energy resources and feasible deployment rates.
 
-``How should Ireland set GHG targets in the residential sector?” 
+%What does new model do that previous didn’t?
+%old one - Ireland’s only fully integrated long-term energy systems optimisation model.
+It is unique in its ability to provide least-cost, cross-sectoral evidenced-based analysis for climate and energy policy.\cite{Gallachoir2020The3}
+%TIM residential improvements - DH, HVO, Solar ( in progress ), better graps the cost and savings of retrofits and heat pumps, which can infrom MACC. 
+%Build with 2018 baseyear, not 2010 as previous versions had - this suits the 2030 target
+%THe implications of net-zero
+%Does new policy context require new modelling approach?
+%Yes, do account for ongoing investments in climate mitigation opitions, the approach should provide higher granularity for likely pathways
+%How does new modelling structure provide insights (give examples) in a way that previous model didn’t?
+%Demands - 
+%Improved more detailed Energy Service Demands
+
+% ``How should Ireland set GHG targets in the residential sector?”
+% ``How should Ireland set GHG targets in the residential sector?”
+``How should Ireland set GHG targets in the residential sector?” feasible deployment rates
 Choosing the Pathways which Create the Least Burden
 and Offer the Most Opportunity for Ireland
-\subsection{Modelling Methodology or Approach}
-\label{LitR:Model}
-The TIM approach.....
+
+The TIM approach.....data and calibration
 %a newly formed group with members from different universities, industries and semi-state agencies within Ireland.
 Following this, other datasets were then used to develop the full housing stock, based on the correlation of year of construction and building energy ratings.
-\subsection{Residential Energy Services}
-\label{Intro:EnergyServices}
+
+\subsection{Residential}
+\label{LitR:RES}
+\subsubsection{Demands}
+\label{LitR:Demands }
+Macroeconomic 
+ The BER database, which represents half of the dwelling stock, highlights the affect year of construction has on expected energy consumption, as shown in Fig.\ref{fig:BERAge}.
+The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
+The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, which helps to better understand the housing stock.
+BERs are required for new houses, rented houses, sold houses and renovated houses.
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[scale=0.5]{Figures/BERAge.png}
+ \caption{Building Energy Rating by Age}
+ \label{fig:BERAge}
+\end{figure}
+
+Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which is a reason for Ireland's high excessive winter mortality \cite{Clinch2000HousingMortality}.
+Fuel Poverty which which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths.
+An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}.
+Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty.
+These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}.
+Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}.
+Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
+\par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
+The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
+\subsubsection{Technology}
+Capacity, Efficiency, Constraints, Uptake, adapation
+
+\subsubsection{Costs}
+\subsubsection{Savings}
 
@@
     x tick label style={
-		/pgf/number format/1000 sep=},xlabel={Year},
+		/pgf/number format/1000 sep=}, xlabel={Year},
 	y tick label style={
@@
     };
+
 \end{axis}
@@
 
-\subsection{Supply}
-\label{LitR:Supply}
-\subsection{Demand}
-\label{LitR:Demand}
 %Space Heating demand - BER space heating is only within certain months and assumes 21C heating..
-\subsection{Assumptions}
-\label{LitR:Assumptions}
-Macroeconomic Assumptions 
-Capacity
-Costs
-Efficiency
-Savings
+\subsubsection{Constraints}
+\label{LitR:Constraints}
+
  % LIT REW INPUT DATA - EU \cite{Simoes2013TheProject} 
@@
 \label{Results}
+\subsection{Scenarios}
+\label{LitR:Sce}
 \subsection{Trajectories}
@@
 %% appendix sections are then done as normal sections
+\section{Discussion}
+\label{Discussion}
+Soft-linking to LEAP and PLEXOS
+\subsection{Strengths}
+\label{Dis:Strengths}
+\subsection{Limitations}
+\label{Dis:Limit}
+
 \section{Conclusion}
 \label{Conclusion}
-\section{Discussion}
-\label{Discussion}
 \section{Acknowledgements}

### 2021-05-27 manuscript.tex
Done for 27May2021

@@
 \label{Intro:TIMES}
-
+The objective function may be thought of as the discounted sum of net annual costs (i.e. costs minus revenues),
 \begin{equation} 
@@
 \label{LitR:RES}
-\subsubsection{Demands}
-\label{LitR:Demands }
-Macroeconomic 
+
  The BER database, which represents half of the dwelling stock, highlights the affect year of construction has on expected energy consumption, as shown in Fig.\ref{fig:BERAge}.
@@
 \subsubsection{Costs}
+\label{LitR:Costs}
 \subsubsection{Savings}
+\label{LitR:Savings}
+\subsubsection{Demands}
+\label{LitR:Demands}
+Macroeconomic 
+Check main TIM paper 
+Historical population estimates and future projections are obtained from the Central Statistics Office (CSO), Ireland \citep{CentralStatisticsOffice2020}.
+We use the M2F1 scenario since it represents medium growth in population and is in line with population projections used in other national sources \citep{Yakut2020}.
+
+\begin{table}[htbp]
+\footnotesize
+ \centering
+ \caption{Energy service demands in TIM}
+
+ \begin{tabular}{p{18em}lccc}
+ \hline
+    \multirow{Energy Service Demand} & \multirow{Driver} & \multicolumn{2}{c}{Value} & \multirow{Unit} \\
+          &       & \multicolumn{1}{c}{2018} & \multicolumn{1}{c}{2050} &  \\ \hline
+    Residential Apartment Demand & Population & 206.80 & 628.93 & 000' \\
+    Residential Attached Demand & Population & 766.35 & 1056.50 & 000' \\
+    Residential Detached Demand & Population & 724.43 & 889.76 & 000' \\
+    \hline
+    \end{tabular}%
+
+  \label{tab:ESD_list}%
+\end{table}%
+
+% Table generated by Excel2LaTeX from sheet 'Population'
+\begin{table}[htbp]
+ \centering
+ \footnotesize
+ \caption{Population}
+ \begin{tabular}{cc}
+ \hline
+ Year & Population (millions) \\
+ \hline
+ 2018 & 4.85 \\
+ 2020 & 4.98 \\
+ 2030 & 5.40 \\
+ 2040 & 5.82 \\
+ 2050 & 6.19 \\
+ \hline
+ \end{tabular}%
+ \label{tab:pop}%
+\end{table}%
+
+The residential stock projections upto 2040 are taken from ESRI's housing demand estimates \citep{Bergin2020}.
+The stock is expected to increase by 40\% from 2018 level with a CAGR of 2\%.
+This results in an average of 27,600 new houses per annum between between 2021-2040.
+Beyond 2040, population is used as a driver to project housing stock.
+The total housing stock obtained in 2050 is 2.57 million which implies 8\% increase from 2040.
+% Table generated by Excel2LaTeX from sheet 'Sheet1'
+\begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Number of dwellings by type ('000)}
+    \begin{tabular}{rrrr}
+    \hline
+    \multicolumn{1}{l}{Year} & \multicolumn{1}{l}{Apartment} & \multicolumn{1}{l}{Attached } & \multicolumn{1}{l}{Detached} \\ \hline
+    2018  & 207   & 766   & 724 \\
+    2030  & 355   & 918   & 833 \\
+    2040  & 493   & 1003  & 878 \\
+    2050  & 629   & 1057  & 890 \\ \hline
+    \end{tabular}%
+  \label{tab:addlabel}%
+\end{table}%
 

### 2021-05-31 manuscript.tex
31 May 2021

@@
 \title{TIMES Ireland Model: Residential Sector}
-
+% The changing TIMES of Irish Homes
 \author[inst1,inst2]{Jason Mc Guire\footnote{Contact: j.mcguire@ucc.ie}}
@@
 Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways, and interpreting the results.
-This section looks at Ireland's climate targets, energy policy and distinctive residential sector, before examining how researchers have previously used Energy system optimization models (ESOMs) to analyse the residential sector.
-\par Section \ref{LitR} describes the methodology used to develop the residential sector of TIMES Ireland Model (TIM).
-The sector outlines the data used, along with the applied constraints, technologies, costs and demands.
-The section outlines the strengths and weakness of the newly developed model and how it differs from other models.
-\par Section \ref{Results} discusses the results- scenarios
-\par Section \ref{Conclusion} Conclusion
+Section \ref{sec:Introduction} looks at Ireland's climate targets, energy policy and distinctive residential sector, before examining how researchers have previously used Energy system optimization models (ESOMs) to analyse the residential sector.
+\par Section \ref{LitR} describes the methodology used to develop the residential sector while also describing the input data, along with the applied constraints, technologies, costs and demands.  
+Section \ref{Results} outlines the results obtained from selected scenarios.
+Section \ref{Discussion} discusses the strengths and weakness of the newly developed model and how it differs from other models.
+Lastly, section \ref{Conclusion} draws some conclusions on the model, results and suggestions on future work which would be beneficially for policy makers in Ireland.
 
@@
 85\% of residential buildings are 2005 or older.
-Furthermore Irelands fossil fuel dependency is underlined by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.
+Furthermore Irelands fossil fuel dependency is underpinned by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.
 
@@
 These targets accounts for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out make sectoral five year carbon budgets  a legal requirement to aid progress of the forementioned long-term targets.
-\par The EU's first Nationally Determined Contribution (NDC) which had a 40\% GHG reduction target (first NDC), has increased to at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}.
+\par The EU's first Nationally Determined Contribution (NDC) which had a 40\% GHG reduction target (first NDC), has been succeeded by a new target of at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}.
 Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-Climate Action Plan 2021 will have to be more ambitious than its predecessor, CAP2019 as it must reflect EU's updated NDC target and take account of the national set target.
+Climate Action Plan 2021 will have to be more ambitious than its predecessor, CAP2019 as it must reflect EU's updated NDC target and take account of the national target also.
 There are also renewable energy targets to consider for transport (RES-T), heat (RES-H) and electricity (RES-E) under renewable directive (EU) 2018/2001, and energy efficiency targets under (EU) 2018/2002.
 
-\subsection{Energy System Optimisation Models}
+\subsection{Current Energy Policy}
+\label{Intro:Policy}
+Ireland has set out national policy objectives under 10 strategic investment priorities \cite{Ireland2018ProjectFramework}.
+The first strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}.
+The densification of Ireland's main cities, will provide opportunities to improve the efficiency of heating, electricity and transport within and between cities.
+\par Specifically focusing on the energy sector, the Marginal Abatement Cost Curve (MACC) produced in CAP2019  \cite{EPA2018EPAReport} found electricity \& transport are most cost effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
+CAP2019 findings show the most cost effective solution to decarbonising the residential sector, is to receive higher amounts of low carbon electricity.
+
+%The CAP2019 MACC, shown in Fig.\ref{fig:CAP2019:MACC} provides $>300$ of the most cost-effective technologies options to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels. 
+%The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_{eq}$.
+The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_{eq}$ over the 2021 to 2030 period.
+The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_{eq}$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+%The options specific to the residential sector are:
+
+%\begin{itemize}
+%    \item{Retrofit oil boiler existing dwellings to B2 equivalent}
+ %   \item{Switch from oil boilers to heat pumps in existing dwellings}
+ %   \item{Retrofit gas boiler and solid fuel stove existing dwellings to B2 equivalent}
+ %   \item{Introduce CO2-free heating in new buildings}\cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
+%\end{itemize}
+
+%\begin{figure}[!htbp]
+ %\centering
+ %\includegraphics[scale=1]{Figures/CAP2019MACC.png} 
+ %\caption{Climate Action Plan 2019 - Marginal Abatement Cost Curve}
+% \label{fig:CAP2019:MACC} \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
+%\end{figure}
+%For example, Figure \ref{fig:CAP2019:MACC} shows the retrofitting of oil boilers is expected to reduce the Total Cost of Ownership (TCO) (ie.
+save money), but only abate a small amount of carbon, whereas switching oil boilers to heat pumps will increase TCO (ie.
+cost money) but the expected abatement of carbon is far greater.
+
+In Ireland the decarbonisation of the electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}.
+A €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}.
+Ireland's current carbon tax of €33.50/ $tCO_2$ will rise to €100/ $tCO_2$ in 2030.
+The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise renewable electricity investment to achieve legally binding renewable targets under 2009/28/EC.
+REFIT operates by guaranteeing a minimum price for electricity over a 15 year period.
+The Renewable Electricity Support Scheme (RESS) replaced REFIT in 2020.
+RESS supports to renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target.
+% Almost 800 MW of solar has been approved in RESS-1 \cite{Eirgrid2020RenewableResults}, this greatly exceeds the current solar capacity of 16 MW, while making significant progress towards the CAP2019 target of 1500 MW solar by 2030.
+\par Sustainable Energy Authority of Ireland (SEAI) have a key role in the rollout of retrofitting and heat pumps in existing homes.
+In parallel, new dwellings must achieve a minimum of 20\% renewable energy.
+The ambient heat used in heat pumps is included as a renewable energy, heat pumps are therefore the preferred heating technology despite the high capital costs.
+As mentioned in Section \ref{Intro:IreRSD}, new buildings in Ireland produce 70\% less GHG emissions than building in 2005, and with CAP2019 plans to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030, highlights the energy efficiency first approach in the residential sector.
+\par Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019VisionIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+GNI anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}.
+The gas network currently reaches 40\% or 680,000 dwellings in urban areas, but with the ``Compact Growth" policy strategy a higher share of building may have access to the gas network in the future.
+A major barrier to increasing biomethane from an agricultural source in the gas network is the cost, it was the most expensive option in the Marginal Abatement Cost Curve (MACC) produced in CAP2019.
+\par Ireland has one of the lowest shares of District Heating (DH) in Europe at less than 1\%  \cite{Gartland2016AIreland}.
+The ``Compact Growth" policy strategy and advancement in DH control will strengthen the case for DH networks to play a part in decarbonisaing the residential sector.
+CAP2019 plans to have approx.
+0.4\% of heating provided by DH in 2030, research using TIMES and EnergyPLAN has shown that 37\% of heating provided by DH is optimal if Ireland is to reduce GHG emissions by 80\% in 2050, compared to 1990 \cite{Thellufsen2019ImplementingIreland}.
+While Irish District Energy Association (IrDEA) suggest 57\% of heat can be sourced from DH, if supporting policy and regulation was in place.
+District Heating will likely play a role, however Gartland points out a range of organisational, technical, regulatory and economic barriers to DH growth in Ireland \cite{Gartland2016AIreland}.
+\par While it maybe optimal to use Hydrotreated Vegetable Oil (HVO) in transport, it is also considered in the residential sector, as it can provide a cheap alternative for the 700,000 Irish dwellings using home heating oil ( mostly kerosene).
+HVO which meets the EU sustainability criteria, reduces GHG emissions by 60\% as compared to fossil fuels \cite{Soam2019FactorsSweden}.
+The capacity of HVO is growing in Europe, but HVO is currently about twice the price of kerosene.
+HVO can be used to replace kerosene, albeit with cost of up to €400 per oil boiler, however there is no policy for HVO in the residential sector.
+
+\subsection{Residential Modelling}
 \label{Intro:ESOM}
-An Energy System Optimisation Model (ESOM), can be used to optimise the selection of a large number of dynamic variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
-\par Some of the main ESOM include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016EnergyProgramme}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}.
-The Integrated MARKAL-EFOM System (TIMES) modelling framework was selected to develop an Irish model, a description of the framework is outlined in the following subsection.
+%Convert the ESOM section in the intro to a “residential modelling section” which I would use to “describe” how residential sector has been model with TIMES and other models (I put describe in “” because I would just mention the approaches and cite specific studies), as well as mention their advantages/disadvantages;- Olex Email
+Energy System Optimisation Models (ESOMs) can be used to optimise the selection of a large number of dynamic variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
+\par Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}.
 
-\subsubsection{TIMES}
-\label{Intro:TIMES}
-The objective function may be thought of as the discounted sum of net annual costs (i.e. costs minus revenues),
+While ESOM can cover all the energy-related sectors, it can also be used to explore the residential sector in detail.
+
+Some studies focusing on the residential sector in TIMES include, Li who studied the preference for heating technology choice in UK \cite{Li2018IncorporatingModel}, Petrovic examined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed the affect higher thermal efficiency in buildings has on co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}.
+There has also been some studies residential sector in Ireland, using previous versions of Irish TIMES and now out-dated climate targets.
+
+\section{Methodology}
+\label{LitR}
+Goal of TIM
+
+\subsection{TIM Overview}
+\label{Overview}
+NOT DONE 
+The Integrated MARKAL-EFOM System (TIMES) modelling framework was utilised to develop an Irish model, a description and selection process of the framework are outlined in the following subsection. 
+\par TIMES has been developed and maintained by Energy Technology Systems Analysis Program (ETSAP) of the International Energy Agency (IEA).
+TIMES is a bottom-up optimisation model generator for energy-environment systems analysis at various levels of spatial, temporal and sectoral resolution.
+\par TIMES uses The GAMS source code to optimise a least cost solution with perfect foresight.
+The objective function may be thought of as the discounted sum of net annual costs (i.e.
+costs minus revenues),
 \begin{equation} 
@@
 
-TIMES has been developed and maintained by Energy Technology Systems Analysis Program (ETSAP) of the International Energy Agency (IEA).
-is a bottom-up optimisation model generator for energy-environment sys-tems analysis at various levels of spatial, temporal and sectoral resolution ...which combines high resolution ``bottom-up" technological detail and ``top-down" macroeconomic variables covering all energy sectors.
  % It calculates the lowest-cost configuration of energy fuels and technologies whichmeet future energy demands, while respecting technical, environmental, economic, social and policy constraints.
@@
 5.4.11
-\subsection{Residential Sector research using ESOMs}
-\label{Intro:ESOMRes}
-While TIMES models can cover all the energy-related sectors, it can also be used to explore the residential sector in detail.
-Some studies focusing on the residential sector in TIMES include, Li who studied the preference for heating technology choice in UK \cite{Li2018IncorporatingModel}, Petrovic examined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed the affect higher thermal efficiency in buildings has on co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}.
-There has also been some studies residential sector in Ireland, using previous versions of Irish TIMES and now out-dated climate targets.
 
-\subsection{Current Energy Policy}
-\label{Intro:Policy}
-
-The CAP2019 Marginal Abatement Cost Curve (MACC),  \cite{EPA2018EPAReport} found electricity \& transport are the most cost effective for decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
-Therefore the most cost effective solution to decarbonising the residential sector, is to receive higher amounts of low carbon electricity.
-
-Ireland has set out national policy objectives under 10 strategic investment priorities \cite{Ireland2018ProjectFramework}.
-The first strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}.
-The densification of Ireland's main cities, will provide opportunities to improve the efficiency of heating, electricity and transport within and between cities.
-
-The CAP2019 MACC, shown in Fig.\ref{fig:CAP2019:MACC} provides $>300$ of the most cost-effective technologies options to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels. 
-The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_{eq}$.
-The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_{eq}$ over the 2021 to 2030 period.
-The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_{eq}$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-The options specific to the residential sector are:
-
-\begin{itemize}
-    \item{Retrofit oil boiler existing dwellings to B2 equivalent}
-    \item{Switch from oil boilers to heat pumps in existing dwellings}
-    \item{Retrofit gas boiler and solid fuel stove existing dwellings to B2 equivalent}
-    \item{Introduce CO2-free heating in new buildings}\cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
-\end{itemize}
-
-\begin{figure}[!htbp]
- \centering
- \includegraphics[scale=1]{Figures/CAP2019MACC.png} 
- \caption{Climate Action Plan 2019 - Marginal Abatement Cost Curve}
- \label{fig:CAP2019:MACC} \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
-\end{figure}
-Where, the retrofitting of oil boilers is expected to reduce the Total Cost of Ownership (TCO) (ie.
-save money), but only abate a small amount of carbon, whereas switching oil boilers to heat pumps will increase TCO (ie.
-cost money) but the expected abatement of carbon is far greater.
-
-\subsubsection{Electricity}
-\par
-In Ireland the decarbonisation of the electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}.
-Under CAP2019 the continued growth and decarbonisation of the electricity will be the fundamental to lowering energy-related GHG in not only electricity, but in heat and transport also \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-Decarbonising electricity is key for Ireland, both the \textit{``stick"} (tax) approach which aims to deter investment in fossil fuel power generation, and the \textit{``carrot"} (subsidies) approach which aims to encourage renewable electricity is applied in Irish policy.
-In Ireland, a €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce fossil fuel demand and investment.
-The carbon tax receipts grew every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}.
-Ireland's current carbon tax of €33.50/ $tCO_2$ will rise to €100/ $tCO_2$ in 2030.
-Many countries have taken this approach, especially in Europe, where Sweden were second to introduce a carbon tax after Finland, the Swedish carbon tax rate is the highest in the world at €108.81/ $tCO_2$.
-About 15\% of global GHG emissions have either an associated cost or tax, this figure is increasing incrementally \cite{WorldBankGroup2020State2020}.
-\par The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise Ireland to invest in renewable electricity to achieve legally binding renewable targets under 2009/28/EC.
-REFIT operates by guaranteeing  new renewable generation and biomass a minimum price for electricity over a 15 year period.
-The last REFIT schemes closed in December 2015.
-In 2020 REFIT was replaced by the Renewable Electricity Support Scheme (RESS), which supports to renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target.
-RESS prioritizes communities to participate in renewable energy projects and increasing technology diversity by broadening the renewable electricity technology mix, where almost 800 MW of solar has been approved in RESS-1 \cite{Eirgrid2020RenewableResults}, this builds alot on the current solar capacity of 16 MW, while making significant progress towards the CAP2019 target of 1500 MW solar by 2030.
-
-\subsubsection{Heat}
-Sustainable Energy Authority of Ireland (SEAI) have played a leading role in the uptake of building fabric upgrades and heat pumps installations in existing homes.
-SEAI provide numerous grants to homeowners, increase public awareness of the benefits and are working on simplifying the process.
-In parallel, new dwellings must achieve a minimum of 20\% renewable energy.
-The ambient heat used in heat pumps is included as a renewable energy, heat pumps are therefore the preferred heating technology despite the high capital costs.
-As mentioned in Section\ref{Intro:IreRSDBack} new buildings in Ireland produce 70\% less GHG emissions than building in 2005 and with CAP2019 plans to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030, shows Irelands energy efficiency first approach in the residential sector.
-in how Ireland aims to decarbonize the residential sector.
-Domestic supports for renewable heating technologies and energy efficiency include:
-\begin{itemize}
-    \item {Insulation}
-    \subitem {Attic}
-    \subitem {Cavity Wall}
-    \subitem {Internal - Dry Lining}
-    \subitem {External}
-    \item {Heat Pumps}
-    \subitem {Air to Water}
-    \subitem {Ground Source to Water}
-    \subitem {Exhaust Air to Water}
-    \subitem {Water to Water}
-    \subitem {Air to Air}
-    \item {Heating Controls}
-    \item {Solar Water Heating}
-    \item{ Solar Photovoltaics }
-\end{itemize}
-
-\subsubsection{Gas}
-
-In a bid not to be left behind in a ``net-zero" future, Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019VisionIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-GNI anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}, with lower indigenous gas expected GNI plans to met ``net-zero" in 2050 by using 50\% Carbon Capture and Storage (CCS), 37\% biomethane and 13\% Hydrogen\cite{Ervia2019VisionIreland}.
-The gas network currently reaches 40\% or 680,000 dwellings in urban areas, but with the ``Compact Growth" as mentioned in section \ref{Intro:Policy} a higher share of building may have access to the gas network in the future.
-ALthough increasing biomethane from an agricultural source in the gas network is the most expensive option in CAP2019's MACC shown in Fig.\ref{fig:CAP2019:MACC}.
-
-\subsubsection{District Heating}
-Ireland has one of the lowest shares of District Heating (DH) in Europe at less than 1\%  \cite{Gartland2016AIreland}.
-DH networks are a mature technology, being used for almost 100 years across Europe.
-``Compact Growth" and advancement in DH control will strengthen the case for DH networks to play a part in decarbonisaing the residential sector.
-CAP2019 plans to have approx.
-0.4\% of heating provided by DH in 2030, research using TIMES and EnergyPLAN has shown that 37\% of heating provided by DH is most efficient if Ireland is to reduce GHG emissions by 80\% in 2050, compared to 1990 \cite{Thellufsen2019ImplementingIreland}.
-While Irish District Energy Association (IrDEA) estimate
-to 57\% of heat can be sourced from DH, if supporting policy and regulation was in place.
-District Heating will likely play a role, but Gartland \cite{Gartland2016AIreland} points out a range of organisational, technical, regulatory and economic barriers to DH growth in Ireland.
-The most prolific barriers being the lack of guidelines, regulations, policies, frameworks or standards for DH which creates high risk and uncertainty when planning DH systems \cite{Gartland2016AIreland}.
-Gartland further explains that \textit{choices made regarding energy supply have been, and continue to be, economically evaluated on a very short-term basis}, the use of an long-term ESOM may provide new insights to climate policy.
-
-\subsubsection{Hydrotreated Vegetable Oil}
-While it maybe optimal to use Hydrotreated Vegetable Oil (HVO) in transport, it is also considered in the residential sector, as it can provide a cheap alternative for the 700,000 Irish dwellings using home heating oil ( mostly kerosene).
-HVO which meets the EU sustainability criteria, reduce GHG emissions by 60\% as compared to fossil fuels \cite{Soam2019FactorsSweden}.
-The capacity of HVO is growing in Europe, but HVO is currently about twice the price of kerosene.
-HVO can be used to replace kerosene, albeit with cost of up to €400 per oil boiler, however there is no policy for HVO in the residential sector.
-biofuel obligation scheme - Internationally there is a limited supply of HVO available and the
-demand for the fuel is high.HVO has been produced in the past at Ireland’s only refinery at Whitegate and
-potentially could be again with the appropriate investment. Not HVO100, but higher concentration of HVO than current blending capacity
-
-\section{Methodology}
-\label{LitR}
-\subsection{Overview}
-\label{Overview}
+%Add description of TIMES and how residential sector is represented to the Methodology section; - Olex Email
 %The TIMES-Ireland Model (TIM) is an optimisation model of the Irish energy system, which calculates the cost-5optimal fuel and technology mix to meet future energy service demands in the transport, buildings, industry and agriculturesectors, while respecting constraints in greenhouse-gas emissions, primary energy resources and feasible deployment rates.
@@
 Following this, other datasets were then used to develop the full housing stock, based on the correlation of year of construction and building energy ratings.
-
+\label{Results}
+\subsubsection{Scenarios}
+Behavioral Constraints here 
 \subsection{Residential}
 \label{LitR:RES}
-
+ENERGY CONSUMPTION CHARACTERISTICS of Houses - 
  The BER database, which represents half of the dwelling stock, highlights the affect year of construction has on expected energy consumption, as shown in Fig.\ref{fig:BERAge}.
@@
 Macroeconomic 
+% textit{To accommodate an expanding population, the number of dwellings has also increased to currently stand at 1.7 million households, although this fgure remains signifcantly short of projected housing need, withannual average demand estimated to be up to 36,000 units for the next 30 years (IBEC 2018).
+Yet despite the number of new homes built to ever increasing energy performance standards, the Irish housing stock is among the poorest in Europe in terms of energy effciency (Goggins et al.
+2016).
+Current trends also show that households are getting bigger, with an average 15\% increase in foor area across all homes between 2000 and 2016 (SEAI 2018) \cite{FahyEnergyPractice}}
 Check main TIM paper 
@@
 for these reasons TIM residential sector is a state of the art modelling platform.
+
 \section{Results}
+
 \label{Results}
-\subsection{Scenarios}
-\label{LitR:Sce}
+Electricity Supply, Technology and Fuel Mix, Retrofitting, Cost
 \subsection{Trajectories}
@@
 \label{Dis:Limit}
-
-\section{Conclusion}
+Consumers’ final investment decision can be
+represented by “willingness-to-pay” curves – for
+instance, around 10\% of owner-occupiers are willing
+to invest in energy efficiency measures for a payback
+time of 4 years, falling to 0\% for a payback time of
+6 years for some consumer cohorts
+\section{Conclusions}
 \label{Conclusion}
+This paper reports results on ambitious mitigation target in the period to 2050 for the Irish energy system. The analysis have been performed using the Irish TIMES model
+\textit{We stress the importance of differentiating between technical con-
+straints and behavioural constraints in optimisation models.
+Including both conflates what is considered desirable under the economic welfare paradigm (i.e.
+cost-minimising decisions) with “realistic decisions”, complicating the interpretation of the results.
+Optimisation models are by design well suited to study cost-optimal behaviour, and running scenarios just with technical constraints facilitates the identification of areas where market barriers are more entrenched, and interventions to accelerate the uptake of low-carbon practices may be needed.} \cite{Astudillo2017CanModelb}
 \section{Acknowledgements}

### 2021-06-01 manuscript.tex
Residential ENergy Modelling in Progress

@@
 % The changing TIMES of Irish Homes
+
 \author[inst1,inst2]{Jason Mc Guire\footnote{Contact: j.mcguire@ucc.ie}}
@@
 In Ireland the decarbonisation of the electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}.
+Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private invesvestment towards low carbon or renewable energy.
 A €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}.
@@
 \label{Intro:ESOM}
-%Convert the ESOM section in the intro to a “residential modelling section” which I would use to “describe” how residential sector has been model with TIMES and other models (I put describe in “” because I would just mention the approaches and cite specific studies), as well as mention their advantages/disadvantages;- Olex Email
-Energy System Optimisation Models (ESOMs) can be used to optimise the selection of a large number of dynamic variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
-\par Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}.
 
-While ESOM can cover all the energy-related sectors, it can also be used to explore the residential sector in detail.
+Energy System Optimisation Models (ESOMs) can be used to explore least cost pathways, as they consider a large amount of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
+Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.
 
-Some studies focusing on the residential sector in TIMES include, Li who studied the preference for heating technology choice in UK \cite{Li2018IncorporatingModel}, Petrovic examined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed the affect higher thermal efficiency in buildings has on co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}.
+While ESOM cover all the energy-related sectors, they can also be used to specifically explore the residential sector in detail.
+For example, Li et al.
+\cite{Li2018IncorporatingModel} investigated the heating technology preference in UK, using UK TIMES Model (UKTM).
+As TIMES assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
+Li set out to incorporate a Discrete Choice Model (DCM) into TIMES to better reflect how homeowners choose heating technologies.
+The DCM was developed from a nationwide survey with 442 homeowners completing the survey.
+The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems.
+The DCM then estimated the probability of a heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices.
+While Li's researcher has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights, through more surveys and analysis.
+
+Leibowicz et al.
+developed an expanded version of the Open Source Energy Modeling System (OSeMOSYS) that integrates electricity generation and storage, residential building energy service demands, and end-use technologies for providing those services \cite{Leibowicz2018OptimalServices}.
+Endogenous constraints on the market penetration rate of each technology and hourly service demand profiles are established and applied to the residential sector of Austin, Texas.
+Austin was chosen because of the high granularity data availability \cite{EnergyInc.}.
+Leibowicz et al.
+find, net present costs in the \textit{``Policy scenarios"} ( emissions must decline linearly toward 10\% of their current level by 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations will be imposed over the model timeframe), but the percentage increase in net present cost due to the policy diminishes with a more thermally efficient building stock.
+Leibowicz et al.
+also finds electrification to be the dominant strategy for cost-effectively decarbonizing urban residential dwellings.
+These insights show the strength of an ESOM, but the not all costs are taken into account in Leibowicz study, the capital cost associated with more thermally efficient buildings are not included.
+
+Hedegard and Münster used....
+
+Patankar et al. TEMOA south sudan
+
+Petrovic xamined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed the affect higher thermal efficiency in buildings has on co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}.
 There has also been some studies residential sector in Ireland, using previous versions of Irish TIMES and now out-dated climate targets.
@@
 
-\subsection{TIM Overview}
+\subsection{TIMES Energy System Optimisation Model}
 \label{Overview}
@@
 costs minus revenues),
+% The market impacts of grid tariffs are analysed using the energy
+system model Balmorel (Wiese et al.
+(2018)) [25].
+Balmorel is an open source, bottom-up, linear partial equilibrium model.
+It opti- mizes social economic surplus of production, transmission, con- sumption, and investments in electricity and DH under the assumption ofperfect competition, with perfect foresight, and with respect to exogenous technical, physical, and regulatory constraints per geographically defined area.
+Generation technologies have specified fuel types, efficiencies, fixed costs, variable costs, and environmental characteristics.
+CHP
 \begin{equation} 
@@
 \label{LitR:RES}
+RES FIGURE
 ENERGY CONSUMPTION CHARACTERISTICS of Houses - 
@@
 The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
-\subsubsection{Technology}
-Capacity, Efficiency, Constraints, Uptake, adapation
+\subsubsection{House Stock Data}
+BER Database, CSO and Calculations and Filtration
 
@@
 \label{LitR:Savings}
-\subsubsection{Demands}
+\subsubsection{Energy Service Demands Projections}
 \label{LitR:Demands}
 Macroeconomic 
+Time slices - 40 etc. 
 % textit{To accommodate an expanding population, the number of dwellings has also increased to currently stand at 1.7 million households, although this fgure remains signifcantly short of projected housing need, withannual average demand estimated to be up to 36,000 units for the next 30 years (IBEC 2018).
@@
 %Space Heating demand - BER space heating is only within certain months and assumes 21C heating..
-\subsubsection{Constraints}
+\subsubsection{Space and Water Heating}
+Capacity, Efficiency, Constraints, Uptake, adapation
+\subsubsection{Lighting, Appliances and Other}
+Capacity, Efficiency, Constraints, Uptake, adapation
+\subsubsection{User Constraints}
 \label{LitR:Constraints}
@@
 Electricity Supply, Technology and Fuel Mix, Retrofitting, Cost
-\subsection{Trajectories}
-\label{LitR:Traj}
+
 %% The Appendices part is started with the command \appendix;
@@
 Soft-linking to LEAP and PLEXOS
+%o	Discuss and interpret the results
+%o	Focus on what added value the model brings
+%o	Also include future development priorities here
+%o	What can’t the model do? Where is it more suitable to use another model - Hannah Email
+
 \subsection{Strengths}

### 2021-06-01 manuscript.tex
Break

@@
 
-Hedegard and Münster used....
+Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised Balmorel model to analyse how heat pumps with  complementing heat storages affect wind power investments in Denmark.
+A westeren and eastern region is used and the model covers norway,sweden, finland and germay in order to cover import/export
+Thermal building model add-on: Theoptimisation is done subject to a number of constraints includingsatisfaction of demands for electricity and heat in each time period,renewable energy potentials, and technical restrictions on units
+almorel is used for handling theintegrated power and district heating system.
+
+nd reduce the need for peak and reservecapacity
+Results show thatthe heat pumps, even without flexible operation, can contribute significantly to facilitating larger windpower investments and reducing system costs, fuel consumption, and CO2emissions
+NOTE- The model is deterministic assuming perfect foresightand the linear version applied does not include start-up costs, min-imum load requirements, or part load efficiencies
+Modelled indivaully heated - RESULT When invest-ments in individual heat pumps are allowed (iHP and iHP-Flex sce-nario), air–water heat pumps are installed in all individuallyheated areas.
+This represents a significant electricity demand
+inidentifying individual heat pumps as highly socio-economicallycompetitive in non-district heating areas.
+Our results indicate thatin terms of making wind power investments more attractive andreducing system costs, fuel consumption, and CO2emissions, theinstallation of the heat pumps is the most important step, whilethe system benefits of adding flexibility to the heat pumps aremoderate.
+When investing in individual heat storages, the heat pumpscontribute in peak load shaving, and their operation is prioritisedfor hours with low electricity prices.
+However, the flexible opera-tion provides only moderate system benefits.
+Hereof, the mainbenefit is a reduced need for peak/reserve capacity investmentsof about 300–600 MW for the case analysed, corresponding tothe size of a large power plant.
 
@@
 Capacity, Efficiency, Constraints, Uptake, adapation
+heat output (Qt) and electricity input in a given time per-iod (Pt) is expressed asQt= COPPt, where COP is the average an-nual coefficient of performance for the heat pump
 \subsubsection{Lighting, Appliances and Other}

### 2021-06-01 manuscript.tex
Done for the day

--- 2021-06-01 manuscript.tex
@@
 \cite{Li2018IncorporatingModel} investigated the heating technology preference in UK, using UK TIMES Model (UKTM).
-As TIMES assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
+As TIMES \cite{Loulou2016DocumentationI} assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
 Li set out to incorporate a Discrete Choice Model (DCM) into TIMES to better reflect how homeowners choose heating technologies.
@@
 Leibowicz et al.
-developed an expanded version of the Open Source Energy Modeling System (OSeMOSYS) that integrates electricity generation and storage, residential building energy service demands, and end-use technologies for providing those services \cite{Leibowicz2018OptimalServices}.
+developed an expanded version of the Open Source Energy Modeling System (OSeMOSYS) \cite{Howells2011OSeMOSYS:Development.} that integrates electricity generation and storage, residential building energy service demands, and end-use technologies for providing those services \cite{Leibowicz2018OptimalServices}.
 Endogenous constraints on the market penetration rate of each technology and hourly service demand profiles are established and applied to the residential sector of Austin, Texas.
@@
 
-Hedegard and Münster used....
+Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised the ability of Balmorel \cite{Wiese2018BalmorelModel} model to handle the integrated power and district heating system to analyse how heat pumps with  complementing heat storages affect wind power investments in Denmark.
+The model was disaggreated between a western and eastern Denmark region and the model covers Norway, Sweden, Finland and Germany, in order to cover import/export.
+Exogenous user constraints applied to the model, based on current Danish policy includes: Wind power must in Denmark by 2020 correspond to at least 50\% of the national electricity consumption, coal is phased out in heat and power plants by 2030, Individual oil boilers are phased out by 2030 and 1.2 Mtonnes straw and 0.2 Mtonnes wood chips must be incinerated at Danish heat and power plants annually.
+The results by Hedegard and Münster find when investments in individual heat pumps are allowed, air–water heat pumps are installed in non-district heating areas.
+This represents a significant electricity demand in identifying individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments and reducing system costs, fuel consumption, and $CO_2$ emissions.
+Hedegard and Münster found the system benefits of adding individual heat storage to the heat pumps are moderate, the main benefit is a reduced need for peak/reserve capacity investments of about 300–600 MW, corresponding to the size of a large power plant.
+While Balmoral model has provided insights, it does not cover all aspects of the power and heat system, such as start-up costs, minimum load requirements, or part load efficiencies.
+Nevertheless, the insights are valuable and the results can be soft-linked to other models for deeper analysis.
 
-Patankar et al. TEMOA south sudan
+Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model \cite{stergaard2015ReviewingSimulations} to explore the feasibility of photovoltaic (PV) systems installed at household level in combination with batteries in Samsø, which is known for being Denmark’s renewable energy island.
+The two approaches are observed for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
+The PV/battery combination is but one out of many potential energy technology combinations, but it insights to the optimisation of the energy system in Samsø.
+Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy systems perspective, one reason for this is the wind power could also be stored in communal batteries.
+From the households’ perspective, the first leads to higher, one results showed with identical battery capacity 64\% of electricity demand is satisfied by a residential battery and 36\% by a one communal battery, but the advantage of a energy system model is the integration of other economic sectors in the energy system.
+While the battery location was explored, the PV system location was not, further investigation into the optimal PV system location could provide further insights.
+The household occupancy behaviour with and without household batteries on the PV system could alter the optimal pathway.
 
-Petrovic xamined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed the affect higher thermal efficiency in buildings has on co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}.
-There has also been some studies residential sector in Ireland, using previous versions of Irish TIMES and now out-dated climate targets.
+% NOT INCLUDE HAVE ENOUGH Petrovic and Karlsson \cite{Petrovic2016ResidentialSystem}  examined a new approached to represent residential heat pumps in Denmark
+
+PREVIOUS RESIDENTIAL ESOM FOR IRELAND 
+\cite{Uidhir2020ResidentialIreland} \cite{Uidhir2020ImprovingDecisions}
+\cite{Dineen2011ModellingHeating}
+\cite{Dineen2015ImprovedSource}
+\cite{Rogan2014LEAPsSystem}
+\cite{Chiodi2013ModellingSystem} - more general
+\cite{Connolly2021AalborgEnergy-System}- EnergyPLAN more general
+\cite{Thellufsen2019ImplementingIreland}-EnergyPLAN DH
+\cite{Goodbody2013RegionalCommunities} - HOMER (Hybrid Optimization of Multiple Energy Resources)
+---- There has also been some studies residential sector in Ireland, using previous versions of Irish TIMES and now out-dated climate targets.
 
@@
 \label{LitR}
-Goal of TIM
+understand, collect and filter robust data along with feasible deployment rates into TIMES framework to provide results, which are analysed internally and externally to give insights on optimally achieving 2030 and 2050 targets and five-year carbon budgets.
+FIGURE?
+
+``How should Ireland set GHG targets in the residential sector?” 
+Choosing the Pathways which Create the Least Burden
+and Offer the Most Opportunity for Ireland
 
@@
 %Improved more detailed Energy Service Demands
-
-% ``How should Ireland set GHG targets in the residential sector?”
-% ``How should Ireland set GHG targets in the residential sector?”
-``How should Ireland set GHG targets in the residential sector?” feasible deployment rates
-Choosing the Pathways which Create the Least Burden
-and Offer the Most Opportunity for Ireland
 
@@
 \subsubsection{Space and Water Heating}
+ISO EN 13790 (Energy Performance of Buildings Directive).
 Capacity, Efficiency, Constraints, Uptake, adapation
+heat output (Qt) and electricity input in a given time per-iod (Pt) is expressed asQt= COPPt, where COP is the average an-nual coefficient of performance for the heat pump
 \subsubsection{Lighting, Appliances and Other}
@@
 \label{sec:Acknowledgements}
-%CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC
+Tomas %CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC
 % https://www.marei.ie/project/capacity/

### 2021-06-02 manuscript.tex
Finished for the day

@@
 
-\title{TIMES Ireland Model: Residential Sector}
+\title{Developing sustainable Irish homes in changing TIMES}
+%TIMES Ireland Model: Residential Sector
 % The changing TIMES of Irish Homes
+% Developing sustainable homes in changing TIMES
 
@@
 Energy System Optimisation Models (ESOMs) can be used to explore least cost pathways, as they consider a large amount of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
-Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.
-
-While ESOM cover all the energy-related sectors, they can also be used to specifically explore the residential sector in detail.
-For example, Li et al.
+Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM cover all the energy-related sectors, they can also be used to specifically explore the residential sector in higher detail, some examples are outlined in this section.
+\par For example, Li et al.
 \cite{Li2018IncorporatingModel} investigated the heating technology preference in UK, using UK TIMES Model (UKTM).
@@
 
-PREVIOUS RESIDENTIAL ESOM FOR IRELAND 
-\cite{Uidhir2020ResidentialIreland} \cite{Uidhir2020ImprovingDecisions}
-\cite{Dineen2011ModellingHeating}
-\cite{Dineen2015ImprovedSource}
-\cite{Rogan2014LEAPsSystem}
-\cite{Chiodi2013ModellingSystem} - more general
-\cite{Connolly2021AalborgEnergy-System}- EnergyPLAN more general
-\cite{Thellufsen2019ImplementingIreland}-EnergyPLAN DH
-\cite{Goodbody2013RegionalCommunities} - HOMER (Hybrid Optimization of Multiple Energy Resources)
----- There has also been some studies residential sector in Ireland, using previous versions of Irish TIMES and now out-dated climate targets.
+Many bottom-up models have been used previously to specifically explore Ireland's residential sector \cite{Dineen2011ModellingHeating}\cite{Ahern2013StateStock}\cite{Ahern2015EnergySystems}\cite{Dineen2015ImprovedSource}\cite{Uidhir2020ResidentialIreland}\cite{Uidhir2020ImprovingDecisions} and some models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, but very few models have used energy system optimisation model to explore the Irish residential sector.
+In 2014, Rogan et al used OSeMOYOS to explore energy service demands \cite{Rogan2014LEAPsSystem}, and Thellufsen examining DH with EnergyPLAN and TIMES in 2019 \cite{Thellufsen2019ImplementingIreland}.
 
@@
 \label{LitR}
-understand, collect and filter robust data along with feasible deployment rates into TIMES framework to provide results, which are analysed internally and externally to give insights on optimally achieving 2030 and 2050 targets and five-year carbon budgets.
-FIGURE?
+% The technical bottom-up development approach of the residential sector is combined with other economic sectors to meet macroeconomic defined demands, to provide GHG targets insights both the energy system and residential sector.
 
-``How should Ireland set GHG targets in the residential sector?” 
-Choosing the Pathways which Create the Least Burden
-and Offer the Most Opportunity for Ireland
+%``How should Ireland set GHG targets in the residential sector?” 
+%Choosing the Pathways which Create the Least Burden and Offer the Most Opportunity for Ireland
 
@@
 \label{Overview}
-NOT DONE 
-The Integrated MARKAL-EFOM System (TIMES) modelling framework was utilised to develop an Irish model, a description and selection process of the framework are outlined in the following subsection. 
-\par TIMES has been developed and maintained by Energy Technology Systems Analysis Program (ETSAP) of the International Energy Agency (IEA).
-TIMES is a bottom-up optimisation model generator for energy-environment systems analysis at various levels of spatial, temporal and sectoral resolution.
-\par TIMES uses The GAMS source code to optimise a least cost solution with perfect foresight.
-The objective function may be thought of as the discounted sum of net annual costs (i.e.
-costs minus revenues),
-% The market impacts of grid tariffs are analysed using the energy
-system model Balmorel (Wiese et al.
-(2018)) [25].
-Balmorel is an open source, bottom-up, linear partial equilibrium model.
-It opti- mizes social economic surplus of production, transmission, con- sumption, and investments in electricity and DH under the assumption ofperfect competition, with perfect foresight, and with respect to exogenous technical, physical, and regulatory constraints per geographically defined area.
-Generation technologies have specified fuel types, efficiencies, fixed costs, variable costs, and environmental characteristics.
-CHP
+\par  The expertise among energy policy researchers in Ireland combined with the global network of Energy Technology Systems Analysis Program (ETSAP) modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways.
+Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model, and then updated with specific national data and assumptions.
+The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.
+There is a need for a new ESOM to be built specifically to explore Ireland's new 2030 and 2050 climate targets and sectoral five-year carbon budgets, the new``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
+\par The Integrated MARKAL-EFOM System (TIMES) modelling framework was utilised to develop the Irish model outlined in this paper.
+TIMES has been developed and maintained by ETSAP of the International Energy Agency (IEA).
+TIMES is a bottom-up optimisation energy-environment model with perfect foresight which is used to analyse various levels of spatial, temporal and sectoral resolution.
+Numerous exogenous technical, physical, and regulatory constraints can be are applied to each region and generation technologies have specified fuel types, efficiencies, fixed costs, variable costs, lifetime, capacities, availability factors and environmental characteristics.
+TIMES uses a linear programming optimizer matrix in General Algebraic Modeling System (GAMS), to compute all calculations.
+
+\par The function of partial equilibrium operates whereby prices and quantities in each time period are such that the suppliers produce exactly the quantities demanded by the consumers and therefore the total economic surplus  (sum of the suppliers’ and consumers’ surpluses) is maximized \cite{Loulou2016DocumentationI}.
+The partial equilibrium creates a stable market situation, from which the maximized surplus is first transformed into an equivalent cost minimization objective by taking the negative of the surplus, and calling this value the total system cost.
+The TIMES objective is therefore to minimize the total `cost' of the system, the TIMES objective function is:
 \begin{equation} 
@@
 \end{equation}
-where: $NPV$ is the net present value of the total cost for all regions (the TIMES objective function);
-$ANNCOST(r,y)$ is the total annual cost in region r and year y; $d_{r,y}$ is the general discount rate;
-$REFYR$ is the reference year for discounting;
-YEARS is the set of years for which there are costs, including all years in the horizon, plus past years (before the initial period) if costs have been defined for past investments, plus a number of years after end of the horizon EOH where some investment and dismantling costs are still being incurred, as well as the Salvage Value; and
-R is the set of regions in the area of study.
-$ANNCOST$ is quite complex and is postponed until section 6.2 of PART II
-Due to the data driven nature of TIMES (see section 2.5), all TIMES constraints are activated and defined by specifying some attributes.
-Attributes are attached to processes, to commodities, to flows, or to special variables that have been created to define new TIMES features.
-Indeed, TIMES
+Where: $NPV$ is the net present value of the total cost for all regions ; $ANNCOST(r,y)$ is the total annual cost in region r and year y; $d_{r,y}$ is the general discount rate; $REFYR$ is the reference year for discounting; YEARS is the set of years for which there are costs; and $R$ is the set of regions in the area of study.
 
- % It calculates the lowest-cost configuration of energy fuels and technologies whichmeet future energy demands, while respecting technical, environmental, economic, social and policy constraints.
-Key inputsand  constraints  include  primary  energy  resource  availability  and  costs,  the  technical  and  cost  evolution  of  new  mitigationoptions and maximum feasible uptake rates of new technologies.
-Alternatively, TIM can be used to assess the implications ofcertain policies, namely regulatory or technology target-setting (for example, biofuels blending obligation or sales/stock share80target for electric vehicles).
-\par  However the expertise among energy policy researchers in Ireland combined with the global network of ETSAP modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways.
-\par Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model, and then updated with specific national data and assumptions.
-The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.
-Therefore, there is a need for a new TIMES model to be built specifically to explore Ireland's new 2030 and 2050 climate targets and sectoral five-year carbon budgets, the ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
-GROWTH CONSTRAINT -A specific type of constraint may be defined to limit the share of process (p) in the total production of commodity (c).
-The constraint indicates that the flow of commodity (c) from/to process (p) - CommodiTypically, a growth constraint is of the following generic form (ignoring several indices for clarity: \cite{Loulou2016EnergyProgramme}
-\begin{equation} 
-\begin{split}
-VAR\_CAP(t + 1) \leq (1 + GROWTH^{M(t+1)-M(t)}) \cdot VAR\_CAP(t)+K
-\end{split}
-\label{Eq:GrowthConstraint}
-\end{equation}
-The GROWTH coefficient is defined as a new attribute of the technology, and represents the maximum annual growth allowed for the capacity.
-The quantity M(t+1)-M(t) is the number of years between the milestones of periods t and t+1.
-The constant K is useful whenever the technology has no capacity initially, in order to allow capacity to build over time (if K were absent and initial capacity is zero, the technology would never acquire any capacity)
-Note that the sign of the constraint may also be of the "larger than or equal to" type to express a maximum rate of abandonment, in which case the "+" sign is replaced by a "–" sign in the right-hand-side of the constraint.
-Equality is also allowed, but must be used only exceptionally in order to avoid railroading of the model.
-5.4.11
+With update technological and macroeconomic data, structured specifically to analyse for sectoral five-year carbon budgets towards 2030 and 2050, it is hoped TIM will provide further and more robust insights than previous Irish TIMES model.
+While also accounting for more fuels and technological options, the residential sector in TIM is structured with around retrofitting, not vice-versa as previous versions, so that it is more aligned to ongoing climate mitigation investments.
+Also, the residential sector in TIM has also been structured concurrently and similarly to a new Irish LEAP model, to allow more accessible soft-linking.
 
 %Add description of TIMES and how residential sector is represented to the Methodology section; - Olex Email
-%The TIMES-Ireland Model (TIM) is an optimisation model of the Irish energy system, which calculates the cost-5optimal fuel and technology mix to meet future energy service demands in the transport, buildings, industry and agriculturesectors, while respecting constraints in greenhouse-gas emissions, primary energy resources and feasible deployment rates.
+%The TIMES-Ireland Model (TIM) is an optimisation model of the Irish energy system, which calculates the cost-5optimal fuel and technology mix to meet future energy service demands in the transport, buildings, industry and agriculture sectors, while respecting constraints in greenhouse-gas emissions, primary energy resources and feasible deployment rates.
 
 %What does new model do that previous didn’t?
-%old one - Ireland’s only fully integrated long-term energy systems optimisation model.
-It is unique in its ability to provide least-cost, cross-sectoral evidenced-based analysis for climate and energy policy.\cite{Gallachoir2020The3}
-%TIM residential improvements - DH, HVO, Solar ( in progress ), better graps the cost and savings of retrofits and heat pumps, which can infrom MACC. 
-%Build with 2018 baseyear, not 2010 as previous versions had - this suits the 2030 target
+%TIM residential improvements - DH, HVO, Solar ( in progress ), better graps the cost and savings of retrofits and heat pumps, which can infrom MACC.Build with 2018 baseyear, not 2010 as previous versions had - this suits the 2030 target
 %THe implications of net-zero
@@
 
-The TIM approach.....data and calibration
-%a newly formed group with members from different universities, industries and semi-state agencies within Ireland.
-Following this, other datasets were then used to develop the full housing stock, based on the correlation of year of construction and building energy ratings.
-\label{Results}
-\subsubsection{Scenarios}
-Behavioral Constraints here 
 \subsection{Residential}
 \label{LitR:RES}
+
+%PICKUPHERE
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[scale=1]{Figures/TIM_Residential_RES (1).jpg}
+ \caption{Residential Sector - Reference Energy System}
+ \label{fig:RES}
+\end{figure}
+Behavioral Constraints here 
 RES FIGURE 
 ENERGY CONSUMPTION CHARACTERISTICS of Houses - 
- The BER database, which represents half of the dwelling stock, highlights the affect year of construction has on expected energy consumption, as shown in Fig.\ref{fig:BERAge}.
+ %The BER database, which represents half of the dwelling stock, highlights the affect year of construction has on expected energy consumption, as shown in Fig.\ref{fig:BERAge}.
 The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
@@
 BER Database, CSO and Calculations and Filtration 
+FIGURE - Expected and Actual Consumption per m2
+Than Actual
 
@@
 \label{LitR:Costs}
+Elephant in the room - Fuel Poverty
+%JF -The tax / subsidy inclusive prices are important in driving behaviour, and taxes and subsidies will be a vital policy instrument.
+Thus, uptake of technologies such as retrofitting will be driven, inter alia, by the net price for consumers.
 \subsubsection{Savings}
@@
 %Space Heating demand - BER space heating is only within certain months and assumes 21C heating..
+Thermostats are not widely used for temperature regulation in private dwellings in Ireland; instead the domestic heating schedule was assumed to be 07.00–09.00 h and 17.00–23.00 h for a heating season of October to May inclusive, based on guidelines issued by the Sustainable Energy Authority of Ireland (SEAI) [32].
+The thermal load consists of both space heating and a baseline demand for sustaining water heating.
+For non-domestic premises the thermal load is defined by office hours, in line with the electrical demand.
+The domestic thermal loads considered in this study are listed in Table 2; the thermal load for the non-domestic
 \subsubsection{Space and Water Heating}
@@
 \label{LitR:Constraints}
+LIST OF SCENARIOS?
+GROWTH CONSTRAINT -A specific type of constraint may be defined to limit the share of process (p) in the total production of commodity (c).
+The constraint indicates that the flow of commodity (c) from/to process (p) - CommodiTypically, a growth constraint is of the following generic form (ignoring several indices for clarity: \cite{Loulou2016EnergyProgramme}
+\begin{equation} 
+\begin{split}
+VAR\_CAP(t + 1) \leq (1 + GROWTH^{M(t+1)-M(t)}) \cdot VAR\_CAP(t)+K
+\end{split}
+\label{Eq:GrowthConstraint}
+\end{equation}
+The GROWTH coefficient is defined as a new attribute of the technology, and represents the maximum annual growth allowed for the capacity.
+The quantity M(t+1)-M(t) is the number of years between the milestones of periods t and t+1.
+The constant K is useful whenever the technology has no capacity initially, in order to allow capacity to build over time (if K were absent and initial capacity is zero, the technology would never acquire any capacity)
+Note that the sign of the constraint may also be of the "larger than or equal to" type to express a maximum rate of abandonment, in which case the "+" sign is replaced by a "–" sign in the right-hand-side of the constraint.
+Equality is also allowed, but must be used only exceptionally in order to avoid railroading of the model.
+5.4.11
 

### 2021-06-22 manuscript.tex
Done for the day

@@
 \usepackage{amsmath}
+\usepackage{comment}
+% \usepackage{cite}
+
+\usepackage[numeric-comp]{biblatex}
+
+\addbibresource{references.bib}%Imports bibliography file
 
@@
 %% for the whole article with \linenumbers.
-
-\addbibresource{references.bib} %Imports bibliography file
 
@@
 
-\title{Developing sustainable Irish homes in changing TIMES}
+\title{Sustainable Irish homes in changing TIMES}
 %TIMES Ireland Model: Residential Sector
@@
 
-Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways, and interpreting the results.
-Section \ref{sec:Introduction} looks at Ireland's climate targets, energy policy and distinctive residential sector, before examining how researchers have previously used Energy system optimization models (ESOMs) to analyse the residential sector.
-\par Section \ref{LitR} describes the methodology used to develop the residential sector while also describing the input data, along with the applied constraints, technologies, costs and demands.  
+Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways, and interpreting the results, so Section \ref{sec:Introduction} of this paper looks at Ireland's climate targets, energy policy and distinctive residential sector.
+While also examining how researchers have previously used Energy System Optimization Models (ESOMs) to analyse the residential sector.
+Section \ref{LitR} describes the methodology used to develop the residential sector while also outlining the input data.
+The process of applying constraints, technologies, costs and demands in the model.
 Section \ref{Results} outlines the results obtained from selected scenarios.
 Section \ref{Discussion} discusses the strengths and weakness of the newly developed model and how it differs from other models.
-Lastly, section \ref{Conclusion} draws some conclusions on the model, results and suggestions on future work which would be beneficially for policy makers in Ireland.
+Lastly, section \ref{Conclusion} draws some conclusions on the model, results and discussion on future work.
 
@@
 Ireland’s distinctive energy sectors are further separated from other EU member states when focusing on the residential sector.
-This is largely down to Ireland’s high reliance on fossil fuels, low building thermal efficiency, larger dwellings and rural settlement patterns.
+This is largely down to Ireland’s higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings and rural settlement patterns.
 This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}.
-The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts, firstly the 2021 building regulations require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}, and secondly according to Central Statistics Office (CSO) data approx.
+The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of $CO_2$ emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousingGov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx.
 85\% of residential buildings are 2005 or older.
-Furthermore Irelands fossil fuel dependency is underpinned by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.
+Furthermore, Ireland's fossil fuel dependency is underpinned by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.
 
-In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential GHG emissions.
-Residential coal GHG emissions has also declined to 6.9\% in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential Greenhouse Gas (GHG) emissions.
+GHG emissions from coal in the residential sector also declined to 6.9\% in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
 However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
-Gas availability is limited to some urban areas, and accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+While, gas availability is limited to some urban areas, and accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
 The residential sector relied on fossil fuels for over 90\% of thermal energy in 2018, consequently Ireland currently has the lowest renewable heat of among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
@@
 Fig.
-\ref{fig:Ireland2018GHG} shows the energy related GHG disaggregated between transport, heat and electricity.
+\ref{fig:Ireland2018GHG} shows the energy-related GHG disaggregated between transport, heat and electricity.
 
@@
 Leibowicz et al.
-developed an expanded version of the Open Source Energy Modeling System (OSeMOSYS) \cite{Howells2011OSeMOSYS:Development.} that integrates electricity generation and storage, residential building energy service demands, and end-use technologies for providing those services \cite{Leibowicz2018OptimalServices}.
-Endogenous constraints on the market penetration rate of each technology and hourly service demand profiles are established and applied to the residential sector of Austin, Texas.
+developed an expanded a version of the Open Source Energy Modeling System (OSeMOSYS) \cite{Howells2011OSeMOSYS:Development.} that integrates electricity generation and storage, residential building energy service demands, and end-use technologies \cite{Leibowicz2018OptimalServices}.Endogenous constraints on the market penetration rate of each technology and hourly service demand profiles are established and applied to the residential sector of Austin, Texas.
 Austin was chosen because of the high granularity data availability \cite{EnergyInc.}.
 Leibowicz et al.
-find, net present costs in the \textit{``Policy scenarios"} ( emissions must decline linearly toward 10\% of their current level by 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations will be imposed over the model timeframe), but the percentage increase in net present cost due to the policy diminishes with a more thermally efficient building stock.
+results find net present costs in the \textit{``Policy scenarios"} ( emissions declining linearly toward 10\% of their current level by 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations imposed ).
+However, the percentage increase in net present cost due to the policy diminishes with a more thermally efficient building stock.
 Leibowicz et al.
@@
 
-Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised the ability of Balmorel \cite{Wiese2018BalmorelModel} model to handle the integrated power and district heating system to analyse how heat pumps with  complementing heat storages affect wind power investments in Denmark.
+Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised  Balmorel model \cite{Wiese2018BalmorelModel} to handle the integrated power and district heating system, to analyse how heat pumps with complementing heat storages affect wind power investments in Denmark.
 The model was disaggreated between a western and eastern Denmark region and the model covers Norway, Sweden, Finland and Germany, in order to cover import/export.
-Exogenous user constraints applied to the model, based on current Danish policy includes: Wind power must in Denmark by 2020 correspond to at least 50\% of the national electricity consumption, coal is phased out in heat and power plants by 2030, Individual oil boilers are phased out by 2030 and 1.2 Mtonnes straw and 0.2 Mtonnes wood chips must be incinerated at Danish heat and power plants annually.
-The results by Hedegard and Münster find when investments in individual heat pumps are allowed, air–water heat pumps are installed in non-district heating areas.
+Exogenous user constraints applied to the model, based on current Danish policy includes: Wind power must provide at least 50\% of the national electricity consumption by 2020, coal is phased out in heat and power plants by 2030 and Individual oil boilers are phased out by 2030.
+The results by find when investments in individual heat pumps are allowed, air to water heat pumps are installed in non-district heating areas.
 This represents a significant electricity demand in identifying individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments and reducing system costs, fuel consumption, and $CO_2$ emissions.
-Hedegard and Münster found the system benefits of adding individual heat storage to the heat pumps are moderate, the main benefit is a reduced need for peak/reserve capacity investments of about 300–600 MW, corresponding to the size of a large power plant.
-While Balmoral model has provided insights, it does not cover all aspects of the power and heat system, such as start-up costs, minimum load requirements, or part load efficiencies.
+Hedegard and Münster found the system benefits of adding individual heat storage to the heat pumps are moderate, the main benefit is a reduced need for peak/reserve capacity investments.
+The Balmorel model has provided insights, but it does not cover all aspects of the power and heat system, such as start-up costs, minimum load requirements, or part load efficiencies.
 Nevertheless, the insights are valuable and the results can be soft-linked to other models for deeper analysis.
@@
 The two approaches are observed for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
-The PV/battery combination is but one out of many potential energy technology combinations, but it insights to the optimisation of the energy system in Samsø.
 Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy systems perspective, one reason for this is the wind power could also be stored in communal batteries.
-From the households’ perspective, the first leads to higher, one results showed with identical battery capacity 64\% of electricity demand is satisfied by a residential battery and 36\% by a one communal battery, but the advantage of a energy system model is the integration of other economic sectors in the energy system.
+When identical communal and individual battery capacities where compared, it was fond  64\% of residential electricity demand is satisfied by a individual battery and only 36\% by a communal battery.
 While the battery location was explored, the PV system location was not, further investigation into the optimal PV system location could provide further insights.
@@
 
-Many bottom-up models have been used previously to specifically explore Ireland's residential sector \cite{Dineen2011ModellingHeating}\cite{Ahern2013StateStock}\cite{Ahern2015EnergySystems}\cite{Dineen2015ImprovedSource}\cite{Uidhir2020ResidentialIreland}\cite{Uidhir2020ImprovingDecisions} and some models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, but very few models have used energy system optimisation model to explore the Irish residential sector.
+% Many bottom-up models have been used previously to specifically explore Ireland's residential sector \cite{Dineen2011ModellingHeating}\cite{Ahern2013StateStock}\cite{Ahern2015EnergySystems}\cite{Dineen2015ImprovedSource}\cite{Uidhir2020ResidentialIreland}\cite{Uidhir2020ImprovingDecisions} and some models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, but very few models have used energy system optimisation model to explore the Irish residential sector.
 In 2014, Rogan et al used OSeMOYOS to explore energy service demands \cite{Rogan2014LEAPsSystem}, and Thellufsen examining DH with EnergyPLAN and TIMES in 2019 \cite{Thellufsen2019ImplementingIreland}.
@@
 \label{LitR}
+
 % The technical bottom-up development approach of the residential sector is combined with other economic sectors to meet macroeconomic defined demands, to provide GHG targets insights both the energy system and residential sector.
@@
 
+%FROM TIM Paper
+%TIM has been co-developed with the LEAP-Ireland model (Mac Uidhir et al., 2020), which is a bottom-up simulation modelof the Irish energy system which simulates the impact of different policy measures and targets on overall GHG emissions inIreland, with a particularly granular representation of transport and residential heat demand.
+Underlying data for the relevantsectors are shared between TIM and LEAP using a Data Repository and shared coding convention dictionary to improve the consistency between the models, give more robust analytical insights for policy and to share and exchange expertise between themodelling teams.
+This also facilitates multi-model approaches to energy systems modelling, which can make use of harmonisedhybrid frameworks coupling simulation and optimisation simultaneously (Rogan et al., 2014)
 \subsection{TIMES Energy System Optimisation Model}
@@
 The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.
-There is a need for a new ESOM to be built specifically to explore Ireland's new 2030 and 2050 climate targets and sectoral five-year carbon budgets, the new``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
-\par The Integrated MARKAL-EFOM System (TIMES) modelling framework was utilised to develop the Irish model outlined in this paper.
-TIMES has been developed and maintained by ETSAP of the International Energy Agency (IEA).
-TIMES is a bottom-up optimisation energy-environment model with perfect foresight which is used to analyse various levels of spatial, temporal and sectoral resolution.
-Numerous exogenous technical, physical, and regulatory constraints can be are applied to each region and generation technologies have specified fuel types, efficiencies, fixed costs, variable costs, lifetime, capacities, availability factors and environmental characteristics.
+There is a need for a new ESOM to be built specifically to explore Ireland's new 2030 and 2050 climate targets and sectoral five-year carbon budgets, the new ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
+\par TIMES is a bottom-up optimisation energy-environment model with perfect foresight which is used to analyse various levels of spatial, temporal and sectoral resolution.
+Exogenous technical, physical, and regulatory constraints can be applied and technologies have specified fuel types, efficiencies, lifetimes, availability, environmental characteristics and fixed, variable and operation costs.
 TIMES uses a linear programming optimizer matrix in General Algebraic Modeling System (GAMS), to compute all calculations.
@@
 \par The function of partial equilibrium operates whereby prices and quantities in each time period are such that the suppliers produce exactly the quantities demanded by the consumers and therefore the total economic surplus  (sum of the suppliers’ and consumers’ surpluses) is maximized \cite{Loulou2016DocumentationI}.
-The partial equilibrium creates a stable market situation, from which the maximized surplus is first transformed into an equivalent cost minimization objective by taking the negative of the surplus, and calling this value the total system cost.
-The TIMES objective is therefore to minimize the total `cost' of the system, the TIMES objective function is:
+The TIMES objective is therefore to minimize the total cost of the system, by solving for the following:
 \begin{equation} 
@@
 \end{equation}
-Where: $NPV$ is the net present value of the total cost for all regions ; $ANNCOST(r,y)$ is the total annual cost in region r and year y; $d_{r,y}$ is the general discount rate; $REFYR$ is the reference year for discounting; YEARS is the set of years for which there are costs; and $R$ is the set of regions in the area of study.
+Where, $NPV$ is the net present value of the total cost for all regions ; $ANNCOST(r,y)$ is the total annual cost in region r and year y; $d_{r,y}$ is the general discount rate; $REFYR$ is the reference year for discounting; YEARS is the set of years for which there are costs; and $R$ is the set of regions in the area of study.
 
-With update technological and macroeconomic data, structured specifically to analyse for sectoral five-year carbon budgets towards 2030 and 2050, it is hoped TIM will provide further and more robust insights than previous Irish TIMES model.
-While also accounting for more fuels and technological options, the residential sector in TIM is structured with around retrofitting, not vice-versa as previous versions, so that it is more aligned to ongoing climate mitigation investments.
-Also, the residential sector in TIM has also been structured concurrently and similarly to a new Irish LEAP model, to allow more accessible soft-linking.
+%With update technological and macroeconomic data, structured specifically to analyse for sectoral five-year carbon budgets towards 2030 and 2050, it is hoped TIM will provide further and more robust insights than previous Irish TIMES model.
 
@@
 \label{LitR:RES}
+The residential sector was developed in parallel with other sectors in TIM, and also been structured concurrently and similarly to the residential sector in the new Irish LEAP model, to allow more accessible soft-linking.
+Fig.\ref{fig:RES} shows the schematic overview of the residential sector in TIM, illustrating the flow of energy from fuel to demand.
 
-%PICKUPHERE
 \begin{figure}[!htbp]
  \centering
- \includegraphics[scale=1]{Figures/TIM_Residential_RES (1).jpg}
+ \includegraphics[scale=0.45]{Figures/TIM_RES3.png}
  \caption{Residential Sector - Reference Energy System}
@@
 \end{figure}
+
+The 2018 SEAI Energy Balance supplies the base year fuel data, both the quantity and type of fuel.
+Where 15 residential fuel types are found in the SEAI Energy Balance, TIM uses 10 residential fuel types.
+For example, bituminous coal, anthracite + manufactured ovoids and lignite \ brown coal briquettes are combined into one fuel called ``Coal" in TIM.
+There are two types of energy services in TIM - Non-Archetype and Archetype.
+The six non-archetype energy services, are not dependant on the type of dwelling and the base year quantity is the total national consumption of that energy service, this data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA (Policy Oriented Tool for Energy and Climate Change Impact Assessment) database.
+The demand for non-archetype energy services is based on the growth of residential dwellings, which is detailed in section \ref{LitR:Demands}.
+The SEAI BER Public database, which is Ireland's national Energy Performance Certificate (EPC) register, complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD).
+The four energy service type are derived from the database - space heating, water heating, pumps \& fans and lighting.
+The space heating calculation (IS EN 13790) used to estimate the expected energy consumption per dwelling, makes some assumptions.
+The internal temperature assumption has a large error, therefore TIM has made adjustments so that the total consumption of energy services is calibrated to the base year data.
+IS EN 13790 assumes all living areas ( living room, sitting room and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C.
+We know this is not the case, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions in TIM.
+
+\begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Internal Temperature Assumptions}
+    \begin{tabular}{rrr}
+    \hline
+    \multicolumn{1}{l}{BER Rating} & \multicolumn{1}{l}{Living Area} & \multicolumn{1}{l}{Non-Living Area }  \\ \hline
+    A  & 23$^{\circ}$C   & 20$^{\circ}$C  \\
+    B  & 21$^{\circ}$C   & 18$^{\circ}$C  \\
+    C  & 18$^{\circ}$C  & 15$^{\circ}$C  \\
+    D  & 18$^{\circ}$C   & 15$^{\circ}$C  \\
+    E  & 18$^{\circ}$C   & 15$^{\circ}$C  \\ 
+    F  & 18$^{\circ}$C   & 15$^{\circ}$C  \\ 
+    G  & 18$^{\circ}$C   & 13$^{\circ}$C  \\ \hline
+    \end{tabular}
+  \label{Table:InternalTemps}
+\end{table}
+
+The Archetype Dwelling Energy Model SQL (ArDEM-SQL) is a simulation model built on IS EN 13790, which converts the BER public database into Microsoft SQL language and then calculates annual energy consumption for space heating, water heating, pumps \& fans and lighting.
+ArDEM-SQL, is only used to change internal temperatures and that data is then used in TIM.
+NOTE large variation in energy rating due to occupancy behaviour -
+%Healy & Clinch reported that fuel-poor households were more likely to be experiencing colder temperatures than other households.
+In total almost 30% of fuelpoor households had living room temperatures below 18°C and more than two thirds had living room temperatures below 20°C.
+Overall nearly half of all non-fuel-poor homes had living room temperatures below 20°C.
+In households with persons aged over 65 years, 16% had living room temperatures below 18°C and over 50% had temperatures below 20°C.
+On the other end of the scale over 5% of households occupied by persons over 65 years experienced living room temperatures of ≥24°C.https://arrow.tudublin.ie/cgi/viewcontent.cgi?article=1089&context=scienmas
+%The BER for the sample dwellings ranged from C1 to F. Over two thirds of the
+sample dwellings had a BER of E or F. The remaining dwellings had a BER of C or
+D. In total 4 of the sample dwellings had undergone energy efficiency upgrades in
+recent years, including cavity wall insulation, attic insulation where applicable and
+heating control upgrades. In the case of two sample dwellings, these upgrade
+works were completed between monitoring period 1 and 2, and therefore results
+could be compared before and after retrofit.
+
+Non-Archetype - lack of data
+The fuel is calibrated to the
+
+DEMAND Space cooling was disaggregated from electricity in POTEnCIA, this space cooling this was used in TIM.
+POTEnCIA provides high resolution energy service demands in 2018, along with energy service projections to 2050.
 Behavioral Constraints here
-RES FIGURE 
+
+While also accounting for more fuels and technological options, the residential sector in TIM is structured with around retrofitting, not vice-versa as previous versions, so that it is more aligned to ongoing climate mitigation investments.
+.
+The Non-Archetype Demands are so, because of lack of data
 ENERGY CONSUMPTION CHARACTERISTICS of Houses - 
@@
 \subsubsection{House Stock Data}
+\label{LitR:StockData}
+CIARAs Paper
+Every BER displays over 200 energy related variables per dwelling, and over half of all residential dwellings are accounted. This large and detailed BER database does
 BER Database, CSO and Calculations and Filtration 
@@
 Than Actual 
-
+DEAP - The space heat use is defined in IS EN 13790 as the heat delivered to the heated space by an ideal heating
+system to maintain the set-point temperature during a given period of time. The DEAP calculation of space
+heat use is done on a monthly basis based on the procedure described in IS EN 13790. A single-zone
+calculation is used, using a single average value of mean internal temperature as described in Section 7.In DEAP, the heating season is defined as running from October to May inclusive.
+The annual space heat use
+is the sum of monthly values for these eight months.
+The heating hours and required internal temperatures in DEAP are based on the requirements of a typical
+household. The schedule is as follows.
+• Weekdays: 07.00 to 09.00 and 17.00 to 23.00
+• Weekends: 07.00 to 09.00 and 17.00 to 23.00
+This standardised schedule represents a total heating period of 56 hours per week.
 \subsubsection{Costs}
@@
 Current trends also show that households are getting bigger, with an average 15\% increase in foor area across all homes between 2000 and 2016 (SEAI 2018) \cite{FahyEnergyPractice}}
-Check main TIM paper 
+%
+% The growth rate of the housing stock in the period to 2030 is obtained from the ESRI's macroeconomic estimates \citep{Yakut2020}. 
+% A multiple linear regression is calculated to project future housing stock based on population size and occupancy rate beyond 2030.
+The total housing stock is expected to increase by 79\% in 2050 relative to 2018 to accommodate the growing population and declining occupancy rates.
+The share of apartments changes from 12\% in 2018 to 24\% in 2050, while the shares of attached and detached dwellings decline by 4\% and 8\% respectively.
 Historical population estimates and future projections are obtained from the Central Statistics Office (CSO), Ireland \citep{CentralStatisticsOffice2020}.
@@
 
+% Energy service demands in the residential sector are driven by the size of the housing stock by type (apartment, attached or detached), which in turn is driven by population growth and the occupancy rate, the average number of people per dwelling.
+The occupancy rate is expected to fall to 2.12 in 2051 \citep{PropertyIndustryIreland2019}, from 2.75 in 2016 (CSO).
+Intermediate values are estimated using Equation \ref{eq:OR}.
+% \begin{equation}
+% \label{eq:OR}
+%  OR= a+b*log(\Delta t)
+% \end{equation}
+% here, $OR$ is the occupancy rate, $\Delta t$ is difference between years, parameters $a$ \& $b$ are estimated using historical values. 
 \begin{table}[htbp]

### 2021-06-24 manuscript.tex
24-June-2021

@@
 
-\documentclass[preprint,12pt,authoryear]{elsarticle}
+\documentclass[authoryear,preprint,12pt]{elsarticle}
 %\documentclass[10pt,final,journal,compsoc,a4paper]{IEEEtran}
@@
 \usepackage{comment}
-% \usepackage{cite}
+%\usepackage{cite}
 
-\usepackage[numeric-comp]{biblatex}
-
-\addbibresource{references.bib}%Imports bibliography file
+\usepackage{biblatex}
+\addbibresource{references.bib} %Imports bibliography file
 
@@
 
-\title{Sustainable Irish homes in changing TIMES}
+\title{Sustainable Pathways for Irish homes in changing TIMES}
 %TIMES Ireland Model: Residential Sector
@@
 
+\subsection{Residential Modelling}
+\label{Intro:ESOM}
+
+Energy System Optimisation Models (ESOMs) can be used to explore least cost pathways, as they consider a large amount of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
+Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM cover all the energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section.
+\par For example, Li et al.
+\cite{Li2018IncorporatingModel} investigated the heating technology preference in UK, using UK TIMES Model (UKTM).
+As TIMES \cite{Loulou2016DocumentationI} assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
+Li set out to incorporate a Discrete Choice Model (DCM) into TIMES to better reflect how homeowners choose heating technologies.
+The DCM was developed from a nationwide survey with 442 homeowners completing the survey.
+The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems.
+The DCM then estimated the probability of a heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices.
+While Li's researcher has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights, through more surveys and analysis.
+
+Leibowicz et al.
+developed an expanded a version of the Open Source Energy Modeling System (OSeMOSYS) \cite{Howells2011OSeMOSYS:Development.} that integrates electricity generation and storage, residential building energy service demands, and end-use technologies \cite{Leibowicz2018OptimalServices}.Endogenous constraints on the market penetration rate of each technology and hourly service demand profiles are established and applied to the residential sector of Austin, Texas.
+Austin was chosen because of the high granularity data availability \cite{EnergyInc.}.
+Leibowicz et al.
+results find net present costs in the \textit{``Policy scenarios"} ( emissions declining linearly toward 10\% of their current level by 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations imposed ).
+However, the percentage increase in net present cost due to the policy diminishes with a more thermally efficient building stock.
+Leibowicz et al.
+also finds electrification to be the dominant strategy for cost-effectively decarbonizing urban residential dwellings.
+These insights show the strength of an ESOM, but the not all costs are taken into account in Leibowicz study, the capital cost associated with more thermally efficient buildings are not included.
+
+Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised  Balmorel model \cite{Wiese2018BalmorelModel} to handle the integrated power and district heating system, to analyse how heat pumps with complementing heat storages affect wind power investments in Denmark.
+The model was disaggreated between a western and eastern Denmark region and the model covers Norway, Sweden, Finland and Germany, in order to cover import/export.
+Exogenous user constraints applied to the model, based on current Danish policy includes: Wind power must provide at least 50\% of the national electricity consumption by 2020, coal is phased out in heat and power plants by 2030 and Individual oil boilers are phased out by 2030.
+The results by find when investments in individual heat pumps are allowed, air to water heat pumps are installed in non-district heating areas.
+This represents a significant electricity demand in identifying individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments and reducing system costs, fuel consumption, and $CO_2$ emissions.
+Hedegard and Münster found the system benefits of adding individual heat storage to the heat pumps are moderate, the main benefit is a reduced need for peak/reserve capacity investments.
+The Balmorel model has provided insights, but it does not cover all aspects of the power and heat system, such as start-up costs, minimum load requirements, or part load efficiencies.
+Nevertheless, the insights are valuable and the results can be soft-linked to other models for deeper analysis.
+
+Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model \cite{stergaard2015ReviewingSimulations} to explore the feasibility of photovoltaic (PV) systems installed at household level in combination with batteries in Samsø, which is known for being Denmark’s renewable energy island.
+The two approaches are observed for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
+Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy systems perspective, one reason for this is the wind power could also be stored in communal batteries.
+When identical communal and individual battery capacities where compared, it was fond  64\% of residential electricity demand is satisfied by a individual battery and only 36\% by a communal battery.
+While the battery location was explored, the PV system location was not, further investigation into the optimal PV system location could provide further insights.
+The household occupancy behaviour with and without household batteries on the PV system could alter the optimal pathway.
+
+%The Cambridge Housing Model (BEIS) ses English Housing Survey data, coupled to a SAP ( UK's BER equivalent ) energy calculator, to estimate energy use and CO2 emissions for all homes in England, broken down by final use.
+The CHM reads in the EHS dwelling for each case and performs building physics calculations to determine energy consumption and associated CO2 emissions, by use and by fuel type.
+Multiplying the energy use and CO2 emissions by the associated weighting and summing across all cases gives total values for England.
+Housing Data is cleaned, Climate data for region is 8 energy services - space heating ( main \& secondary ), water heating, space cooling, lighting, electrical appliances, cooking and pumps \& fan.
+6 fuels - gas, oil, solid, biomass, electriicty and renewable
+
+% NOT INCLUDE HAVE ENOUGH Petrovic and Karlsson \cite{Petrovic2016ResidentialSystem}  examined a new approached to represent residential heat pumps in Denmark
+
+\subsubsection{Ireland: Residential Modelling}
+\label{Intro:ESOMIreland}
+
+In Ireland, previous modelling on the residential sector has often involved bottom-up residential models, rather than energy system models specifically examining the residential sector.
+Nevertheless, the residential models have provided insights on input data and methodology approach used in Ireland.
+In 2011 Dineen and Ó Gallachóir \cite{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP).
+The model focused on energy efficiency of space and water heating by examining technology and building fabric efficiency.
+The building fabric efficiency of newly occupied dwellings developed takes into account the theoretical increase in energy efficiency due to improved 2008 and 2010 building regulations.
+Space heating demand accounted for 70\% of household energy consumption in newly built dwellings in 1997, down to just under 60\% in 2006 and the model is predicts a reduction to 27\% by 2015.The paper acknowledges retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined.
+Nor is the affect of internal temperatures on energy efficiency improvements.
+\par In 2013 Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures to the existing detached, oil centrally heated, rural housing stock.
+A base geometry according to age bands and thermal characteristics was created in the study.
+The heat load for a statistical occupancy and year was established.
+The default values used in some of the BER database variables where highlighted.
+For example,  there are assumed U-values when the roof insulation thickness is unknown, this is based on age, but there are no U-values for uninsulated roofs and a 2001 survey found 18\% of detached housing in Ireland had no roof insulation.
+The internal temperature assumed for all dwelling was 20$^{\circ}$C.
+Ahern et al.
+found that newer dwellings, consume the greatest amount of energy in kWh/annum than all the older dwellings, due to the fact that their large floor area.
+Concluding the majority of energy efficiency progress for thermal use in Irish dwellings is being offset.
+
+The reduction in heating costs and CO2 emissions for houses constructed prior to 1979 (coinciding with the introduction of building regulations) and around 26% for newer homes, thus offering a significant contribution (44%) to Ireland’s residential carbon abatement projections and hence in meeting the EU’s directives on energy and carbon.
+
+through energy efficiency measures has been offset by the so called size effect. The
+The model predicts a reduction of 65% in running costs and
+CO2 emissions for house types A–E and a reduction of 26% in running costs and CO2 emissions for house types F–J from retrofit measures, and forecasts a total saving of 276 million euro and a CO2 abatement of 1.21 MtCO2.
+By
+the
+potential exists to reduce running costs and CO2 emissions by an average of 63% for housing constructed prior to the 1979 building regulations (house types A–E) and by 26% for newer housing.
+The ratio of amounts of CO2 which can be potentially saved to
+the amount of money saved was calculated as 4.35 kgCO2/h.
+Nationally an annual saving of h183 million euro is theoretically achievable with a resultant saving of 0.80 MtCO2 arising from insulation retrofits, the replacement of single glazed windows and increasing air tightness.
+The
+\cite{Ahern2015EnergySystems}\cite{Dineen2015ImprovedSource}\cite{Uidhir2020ResidentialIreland}\cite{Uidhir2020ImprovingDecisions} and some models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, but very few models have used energy system optimisation model to explore the Irish residential sector.
+In 2014, Rogan et al used OSeMOYOS to explore energy service demands \cite{Rogan2014LEAPsSystem}, and Thellufsen examining DH with EnergyPLAN and TIMES in 2019 \cite{Thellufsen2019ImplementingIreland}.
+
 \subsection{Ireland: Residential Sector}
@@
 This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}.
-The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of $CO_2$ emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousingGov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx.
+The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of CO\textsubscript{2} emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousingGov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx.
 85\% of residential buildings are 2005 or older.
@@
 
-\subsection{Climate Targets}
+\subsubsection{Climate Targets}
 
@@
 
-\subsection{Current Energy Policy}
+\subsubsection{Current Energy Policy}
 \label{Intro:Policy}
@@
 
-\subsection{Residential Modelling}
-\label{Intro:ESOM}
-
-Energy System Optimisation Models (ESOMs) can be used to explore least cost pathways, as they consider a large amount of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
-Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM cover all the energy-related sectors, they can also be used to specifically explore the residential sector in higher detail, some examples are outlined in this section.
-\par For example, Li et al.
-\cite{Li2018IncorporatingModel} investigated the heating technology preference in UK, using UK TIMES Model (UKTM).
-As TIMES \cite{Loulou2016DocumentationI} assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
-Li set out to incorporate a Discrete Choice Model (DCM) into TIMES to better reflect how homeowners choose heating technologies.
-The DCM was developed from a nationwide survey with 442 homeowners completing the survey.
-The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems.
-The DCM then estimated the probability of a heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices.
-While Li's researcher has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights, through more surveys and analysis.
-
-Leibowicz et al.
-developed an expanded a version of the Open Source Energy Modeling System (OSeMOSYS) \cite{Howells2011OSeMOSYS:Development.} that integrates electricity generation and storage, residential building energy service demands, and end-use technologies \cite{Leibowicz2018OptimalServices}.Endogenous constraints on the market penetration rate of each technology and hourly service demand profiles are established and applied to the residential sector of Austin, Texas.
-Austin was chosen because of the high granularity data availability \cite{EnergyInc.}.
-Leibowicz et al.
-results find net present costs in the \textit{``Policy scenarios"} ( emissions declining linearly toward 10\% of their current level by 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations imposed ).
-However, the percentage increase in net present cost due to the policy diminishes with a more thermally efficient building stock.
-Leibowicz et al.
-also finds electrification to be the dominant strategy for cost-effectively decarbonizing urban residential dwellings.
-These insights show the strength of an ESOM, but the not all costs are taken into account in Leibowicz study, the capital cost associated with more thermally efficient buildings are not included.
-
-Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised  Balmorel model \cite{Wiese2018BalmorelModel} to handle the integrated power and district heating system, to analyse how heat pumps with complementing heat storages affect wind power investments in Denmark.
-The model was disaggreated between a western and eastern Denmark region and the model covers Norway, Sweden, Finland and Germany, in order to cover import/export.
-Exogenous user constraints applied to the model, based on current Danish policy includes: Wind power must provide at least 50\% of the national electricity consumption by 2020, coal is phased out in heat and power plants by 2030 and Individual oil boilers are phased out by 2030.
-The results by find when investments in individual heat pumps are allowed, air to water heat pumps are installed in non-district heating areas.
-This represents a significant electricity demand in identifying individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments and reducing system costs, fuel consumption, and $CO_2$ emissions.
-Hedegard and Münster found the system benefits of adding individual heat storage to the heat pumps are moderate, the main benefit is a reduced need for peak/reserve capacity investments.
-The Balmorel model has provided insights, but it does not cover all aspects of the power and heat system, such as start-up costs, minimum load requirements, or part load efficiencies.
-Nevertheless, the insights are valuable and the results can be soft-linked to other models for deeper analysis.
-
-Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model \cite{stergaard2015ReviewingSimulations} to explore the feasibility of photovoltaic (PV) systems installed at household level in combination with batteries in Samsø, which is known for being Denmark’s renewable energy island.
-The two approaches are observed for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
-Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy systems perspective, one reason for this is the wind power could also be stored in communal batteries.
-When identical communal and individual battery capacities where compared, it was fond  64\% of residential electricity demand is satisfied by a individual battery and only 36\% by a communal battery.
-While the battery location was explored, the PV system location was not, further investigation into the optimal PV system location could provide further insights.
-The household occupancy behaviour with and without household batteries on the PV system could alter the optimal pathway.
-
-% NOT INCLUDE HAVE ENOUGH Petrovic and Karlsson \cite{Petrovic2016ResidentialSystem}  examined a new approached to represent residential heat pumps in Denmark
-
-% Many bottom-up models have been used previously to specifically explore Ireland's residential sector \cite{Dineen2011ModellingHeating}\cite{Ahern2013StateStock}\cite{Ahern2015EnergySystems}\cite{Dineen2015ImprovedSource}\cite{Uidhir2020ResidentialIreland}\cite{Uidhir2020ImprovingDecisions} and some models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, but very few models have used energy system optimisation model to explore the Irish residential sector.
-In 2014, Rogan et al used OSeMOYOS to explore energy service demands \cite{Rogan2014LEAPsSystem}, and Thellufsen examining DH with EnergyPLAN and TIMES in 2019 \cite{Thellufsen2019ImplementingIreland}.
-
 \section{Methodology}
@@
 Fig.\ref{fig:RES} shows the schematic overview of the residential sector in TIM, illustrating the flow of energy from fuel to demand.
-
+First iteration - of model will need further development DH, BEhavour, higher granularity and spatial will be easily done with dataset. 
 \begin{figure}[!htbp]
@@
 
-The Archetype Dwelling Energy Model SQL (ArDEM-SQL) is a simulation model built on IS EN 13790, which converts the BER public database into Microsoft SQL language and then calculates annual energy consumption for space heating, water heating, pumps \& fans and lighting.
-ArDEM-SQL, is only used to change internal temperatures and that data is then used in TIM.
-NOTE large variation in energy rating due to occupancy behaviour -
+The Archetype Dwelling Energy Model SQL (ArDEM-SQL) is a simulation model built on IS EN 13790, it converts the BER public database into Microsoft SQL language and then calculates annual energy consumption for space heating, water heating, pumps \& fans and lighting.
+ArDEM-SQL is used to clean the input data in TIM, by changing internal temperatures.
+The internal temperature is highly dependent on occupancy behaviour, a thermal comfort  thermal discomfort are higher among the over-65s than for other household \cite{Healy2002FuelIreland} - living room temperatues between 12 and 26...s  and therefore there is are large variations per BER rating.
+The International Standard ISO 7730 [6] uses these PMV and PPD indices to predict the thermal sensation of people exposed to moderate thermal environments, as well as to specify acceptable thermal environmental conditions for comfort...
+Despite this, it is
+worth noting that the most inefficient dwellings i.e. band F, managed to achieve an
+average temperature during period 2 that was only 0.5°C lower than the most
+efficient dwellings i.e. band C.
+C = 19.2, D = 18, E = 18.2 and F= 18.7 older occupants dublin Ho
 %Healy & Clinch reported that fuel-poor households were more likely to be experiencing colder temperatures than other households.
@@
 \label{LitR:StockData}
+EPC register EU - Ireland
+%The quality of Swedish EPC data was initially studied by Göransson [41], subsequently by Claesson [42], Mangold et al.
+[21] and Hårsman et al.
+[43], and most recently by Pasichnyi et al.
+[6].
+These studies have found a couple of important deficiencies in Swedish EPC data.
+First, the energy expert, and the firm he or she belongs to, has been shown to generate variations in Swedish EPC data [43], [44].
+More specifically, Hårsman et al.
+[43] found that the energy expert along with the characteristics of the firm caused a variation of ±20% in the assessment of energy consumption.
+%Second, through interviews with energy experts, Claesson [42] found that most EPCs contain uncertain estimates of energy use as well as arbitrarily distributed values—thus challenging the notion that Swedish EPC data quality, while allegedly measured, should be superior to the quality of calculated values in other member states.
 CIARAs Paper
@@
 Macroeconomic 
+no elasticity - fuel prices and weather compensation?
 Time slices - 40 etc. 
@@
 \label{Dis:Strengths}
+
+Detailed bottom up model - using 2018 base year ( in line with policy )
+Updated technology
+Combines energy system effects
+Robust filtered input data about housing stock
+
 \subsection{Limitations}
@@
 6 years for some consumer cohorts
+Type of Tenure is ignored
+BER aggeragateed
+housing stock u-values
+behaviour and occupancy
+Area
 \section{Conclusions}

### 2021-06-25 manuscript.tex
25 June 2021

@@
 Nevertheless, the residential models have provided insights on input data and methodology approach used in Ireland.
+Early work by Healy and CLinch ---
+2009 Hull
+
 In 2011 Dineen and Ó Gallachóir \cite{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP).
@@
 The default values used in some of the BER database variables where highlighted.
-For example,  there are assumed U-values when the roof insulation thickness is unknown, this is based on age, but there are no U-values for uninsulated roofs and a 2001 survey found 18\% of detached housing in Ireland had no roof insulation.
-The internal temperature assumed for all dwelling was 20$^{\circ}$C.
+For example,  there are assumed U-values when the roof insulation thickness is unknown, this is based on age, but there are no U-values for uninsulated roofs despite a 2001 survey found 18\% of detached housing in Ireland had no roof insulation.
 Ahern et al.
-found that newer dwellings, consume the greatest amount of energy in kWh/annum than all the older dwellings, due to the fact that their large floor area.
-Concluding the majority of energy efficiency progress for thermal use in Irish dwellings is being offset.
+found dwellings built post-2000 consume the greatest amount of energy (kWh/annum), due to their large floor area.
+Concluding that the majority of energy efficiency progress for thermal use in Irish dwellings is being offset.
+The model thermal improvement measures resulted a reduction of running costs and CO2 emissions by an average of 65\% for houses constructed prior to 1979, this resulted in an average 12-year payback.
+Houses built in 1979 or after resulted in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures.
 
-The reduction in heating costs and CO2 emissions for houses constructed prior to 1979 (coinciding with the introduction of building regulations) and around 26% for newer homes, thus offering a significant contribution (44%) to Ireland’s residential carbon abatement projections and hence in meeting the EU’s directives on energy and carbon.
+\par In 2015 Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examing new dwelling, to examining energy savings potential of energy efficiency improvement measures of the existing dwelling stock.
+The model relied on the BER database, which accounted for 18\% of all Irish dwellings, currently the database has more than doubled and accounts for over 50\%.
+The model used 175 archetype dwellings ( 5 building types, 7 energy bands and 5 wall construction types ) and 6 different retrofit measures were applied.
+BER database filters were used in this model, the BER filters from this study, were updated and applied in TIM.
+A sensitivity analysis is performed on the effects of internal temperature and rebound.
+It was found that there was approximately a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift.
+Concluding, BER standard assumptions overestimate dwelling energy consumption and that the average internal temperature across the Irish dwelling stock may be closer to 18$^{\circ}$C than 21$^{\circ}$C.
 
-through energy efficiency measures has been offset by the so called size effect. The
-The model predicts a reduction of 65% in running costs and
-CO2 emissions for house types A–E and a reduction of 26% in running costs and CO2 emissions for house types F–J from retrofit measures, and forecasts a total saving of 276 million euro and a CO2 abatement of 1.21 MtCO2.
-By
-the
-potential exists to reduce running costs and CO2 emissions by an average of 63% for housing constructed prior to the 1979 building regulations (house types A–E) and by 26% for newer housing.
-The ratio of amounts of CO2 which can be potentially saved to
-the amount of money saved was calculated as 4.35 kgCO2/h.
-Nationally an annual saving of h183 million euro is theoretically achievable with a resultant saving of 0.80 MtCO2 arising from insulation retrofits, the replacement of single glazed windows and increasing air tightness.
-The
-\cite{Ahern2015EnergySystems}\cite{Dineen2015ImprovedSource}\cite{Uidhir2020ResidentialIreland}\cite{Uidhir2020ImprovingDecisions} and some models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, but very few models have used energy system optimisation model to explore the Irish residential sector.
+Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}  created Archetype Dwelling Energy Model - Structured Query Language (ArDEM-SQL), to simulate the potential for improved energy efficiency gains within the existing retrofit programs, specifically the Better Energy Homes (BEH) scheme.
+BER database had grown to in ex- cess of 735,0 0 0 records
+We introduce a new methodology that provides insights into sub-optimal retrofit choices.
+The five most common retrofit combinations are simulated for each of the nine archetypes. The 3 energy rating types and 3 archetypes
+The results show that the alternative retrofit combination differs by archetype and that additional energy effi- ciency gains of up to 86% can be achieved due to alternative retrofit choices
+while maintaining the total number of retrofits completed, we estimate an energy savings potential of 1389 GWh, realising an additional 643 GWh or an 86% additional energy savings potential relative to those in the baseline data, see Fig.
+7
+The simulation of alternative archetype retrofit measures
+yielded additional energy savings of 86% when compared with the observed energy savings as they appear in the baseline sce- nario.
+The additional energy savings are not evenly distributed across all building archetypes with detached CD/EFG and terraced CD/EFG account for 98% of the potential additional energy sav- ings when using the alternative retrofits identified, suggesting that greater energy savings are possible when retrofit measures are ap- plied to less energy efficient dwellings.
+It
+\citep{Ahern2020ADatabases}
+and some models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, but very few models have used energy system optimisation model to explore the Irish residential sector.
 In 2014, Rogan et al used OSeMOYOS to explore energy service demands \cite{Rogan2014LEAPsSystem}, and Thellufsen examining DH with EnergyPLAN and TIMES in 2019 \cite{Thellufsen2019ImplementingIreland}.
@@
 
-The Archetype Dwelling Energy Model SQL (ArDEM-SQL) is a simulation model built on IS EN 13790, it converts the BER public database into Microsoft SQL language and then calculates annual energy consumption for space heating, water heating, pumps \& fans and lighting.
+The ArDEM-SQL is a simulation model built on IS EN 13790, it converts the BER public database into Microsoft SQL language and then calculates annual energy consumption for space heating, water heating, pumps \& fans and lighting.
 ArDEM-SQL is used to clean the input data in TIM, by changing internal temperatures.
@@
 This standardised schedule represents a total heating period of 56 hours per week.
+
+Scaling BER to CSO
+\begin{equation} 
+\begin{split}
+S_y =  \frac{C_y} {B_y}
+\end{split}
+%\[ \sum_{n=1}^{\infty} 2^{-n} = 1 \]
+\label{Eq:BERtoCSO}
+\end{equation}
+The number of dwellings by year of construction within the
+BER database (B y ) is then scaled to align with the total number of dwellings as they appear in the CSO dataset (C y ) –for each archetype.
+Eq.
+(8) describes the scaling factor (S y ), used for each year of construction bracket
+
+\begin{equation} 
+\begin{split}
+A_y =  {S_y} \cdot {B_{yarc}}
+\end{split}
+\label{Eq:BERtoCSO2}
+\end{equation}
+Number of archetypes dwellings by year of construction , where A k = Frequency of each archetype, S y = Scaling Factor, B yarc = Number Dwellings by BER category, A k = Frequency of each archetype.
+
+Model calibration: archetype demand The
 \subsubsection{Costs}
@@
 \subsubsection{Energy Service Demands Projections}
+\begin{equation} 
+\begin{split}
+E_y = {\sum_{A,M}} N_{A,y} \cdot S_{A,M}
+\end{split}
+%\[ \sum_{n=1}^{\infty} 2^{-n} = 1 \]
+\label{Eq:RetrofitSaving}
+\end{equation}
+$E_y$ = Energy Savings in Year y
+$N_{A,y}$ = Number of dwellings of archetype A retrofitted in year y
+$S_{A,M}$ = ¼ Energy Savings per annum for retrofit measureM carried
+out on archetype dwelling A
+A = For each archetype
+M = For each retrofit measure carried out
+
+Next - 
+\begin{equation} 
+\begin{split}
+D_M = {\sum_{A,M}^n} (A_kI_{ki} - A_kI_{kT} )
+\end{split}
+%\[ \sum_{n=1}^{\infty} 2^{-n} = 1 \]
+\label{Eq:EnergyandRetrofitSaving}
+\end{equation}
+Total Energy Demand , where D T = Total Energy Demand, n = total number of archetypes, A k = Frequency of each archetype, I k = Energy Intensity of each archetype - D M = Energy Savings by retrofit measure (M), n = total number of archetypes, A k = Frequency of each archetype dwellings eligible for retrofit measure, I ki = Pre-retrofit archetype energy intensity, I kr - post-retrofit archetype intensity - Tomas Paper
+
+Heterogeneous Occupancy Behaviour > House Stock 
 \label{LitR:Demands}
+Residential Energy was >50\%higher before internal temperature adjustments
 Macroeconomic 
@@
 Area
+expected energy consumption weakness of BER database as 
+Actual energy consumption for dwellings of
+similar or near identical construction may vary by up to 600%
+depending on usage \cite{Sunikka-Blank2012BuildingConsumption}
 \section{Conclusions}

### 2021-06-25 manuscript.tex
Done for 25-June-2021

@@
 
-In Ireland, previous modelling on the residential sector has often involved bottom-up residential models, rather than energy system models specifically examining the residential sector.
-Nevertheless, the residential models have provided insights on input data and methodology approach used in Ireland.
-Early work by Healy and CLinch ---
-2009 Hull
+%and some models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, but very
+
+In 2014, Rogan et al. \cite{Rogan2014LEAPsSystem} used OSeMOYOS to explore residential energy service demands in Ireland
+ LEAP writes an input file for OSeMOSYS, runs it and imports its results back into LEAP.
+Currently, the LEAP input file only contains the aggregated electricity demand as the main driver for the optimization within OSeMOSYS.
+ the model examines the
+ 3 main sceanrios - exploring transtport, residential, services and industy. The reference sceanrio - 
+REF - TFC increase by 6.7\% in 2020, base year is 2008.
+space and water heating TFC have decreased in share of total residential TFC from 82 to 77 \%, whereas lighting, cooking and appliances have increased in TFC from 18 to 23 \%.
+assume that 1.4 \% of existing dwellings per annum will undergo a shallow retrofit in the reference
+case, corresponding to just below 300,000 dwellings
+
+In the energy efficiency scenario, we assume that 800,000 residential dwellings will be retrofitted between 2009 and 2020, in line with government targets for the BEH scheme,  Other policies in the residential sector are building regulations 2011 and building regulations 2013 and CFL lighting,  The combined impact of these policies is a reduction in TFC of 9.9 \% compared to the reference scenario.
+
+The energy efficiency+scenario explores the impact of an ambitious retrofitting scenario beyond its current rate of implementation: retrofitbest is realised by retrofitting the same overall number of dwellings but undertaking a deeper, more comprehensive package of
+energy efficiency measures on each, resulting in 7.2 \% savings compared to the energy efficiency scenario and 16.5 \% savings compared to the reference scenario.
+  Insights - A challenging issue for energy modelling of policy is the
+barriers and obstacles to investment, particularly in
+residential energy efficiency measures such as split
+incentives and information gaps, which lead to suboptimal
+uptake of economically beneficial energy
+efficiency improvements.
+
+Thellufsen examining DH with EnergyPLAN and TIMES in 2019 \cite{Thellufsen2019ImplementingIreland}.
+
+In Ireland few models have used energy system optimisation model to explore the residential sector.
+The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector only models, rather than energy system models specifically examining the residential sector.
+Nevertheless, the residential models have provided insights on input data and methodology approach used in Ireland, some of which are outlined below.
+
+Early work by Healy and CLinch ---\cite{Clinch2000HousingMortality,Healy2002FuelIreland,Clinch2003ValuingModel,Clinch2000DomesticFailure,Clinch2001Cost-benefitEfficiency}
 
@@
 
-Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}  created Archetype Dwelling Energy Model - Structured Query Language (ArDEM-SQL), to simulate the potential for improved energy efficiency gains within the existing retrofit programs, specifically the Better Energy Homes (BEH) scheme.
-BER database had grown to in ex- cess of 735,0 0 0 records
-We introduce a new methodology that provides insights into sub-optimal retrofit choices.
-The five most common retrofit combinations are simulated for each of the nine archetypes. The 3 energy rating types and 3 archetypes
-The results show that the alternative retrofit combination differs by archetype and that additional energy effi- ciency gains of up to 86% can be achieved due to alternative retrofit choices
-while maintaining the total number of retrofits completed, we estimate an energy savings potential of 1389 GWh, realising an additional 643 GWh or an 86% additional energy savings potential relative to those in the baseline data, see Fig.
-7
-The simulation of alternative archetype retrofit measures
-yielded additional energy savings of 86% when compared with the observed energy savings as they appear in the baseline sce- nario.
-The additional energy savings are not evenly distributed across all building archetypes with detached CD/EFG and terraced CD/EFG account for 98% of the potential additional energy sav- ings when using the alternative retrofits identified, suggesting that greater energy savings are possible when retrofit measures are ap- plied to less energy efficient dwellings.
-It
-\citep{Ahern2020ADatabases}
-and some models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, but very few models have used energy system optimisation model to explore the Irish residential sector.
-In 2014, Rogan et al used OSeMOYOS to explore energy service demands \cite{Rogan2014LEAPsSystem}, and Thellufsen examining DH with EnergyPLAN and TIMES in 2019 \cite{Thellufsen2019ImplementingIreland}.
+The Archetype Dwelling Energy Model (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}.
+ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and then calculates annual energy consumption to simulate the potential for improved energy efficiency gains within the existing retrofit programs.
+Mac Uidhir et al.
+specifically examined the Better Energy Homes (BEH) scheme.
+The five most common retrofit combinations are simulated for each of 9 archetypes ( 3 energy rating types and 3 building types ).
+The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\% can be achieved when compared the baseline scenario.
+This shows the the type of retrofit and type of house can have a huge affect on the energy saved.
+Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is considered negligible.
+
+SEAI’s National Energy Modelling Framework (NEMF) is a suite of modelling tools used for modelling the Irish
+energy system, both supply and demand side, to inform Irish Government energy policy and project Ireland’s
+energy use based on these policie - https://www.seai.ie/publications/National-Energy-Projections-to-2030.pdf
+The macro-economic model currently in use is called COSMO (Core Structural Model of the Irish Economy),
+which creates a baseline forecast of energy consumption by sector (ESRI, 2017c). A range of sectoral models
+are used to analyse both supply and demand side energy. Electricity supply is modelled using a commercial
+power system software called PLEXOS47. A BioHeat model is used for techno-economic modelling of
+bioenergy supply chains in Ireland for the electricity generation, transport and heat sectors48. There is also a
+separate transport energy simulation model, which considers biofuels blending and the growth of low
+emissions vehicles in the vehicle stock, including electric vehicles. In addition, separate modelling by an
+energy-efficiency policy analysis tool (Unlocking model) informed the energy efficiency estimates included in
+the NEMF49.
 
@@
 
-The ArDEM-SQL is a simulation model built on IS EN 13790, it converts the BER public database into Microsoft SQL language and then calculates annual energy consumption for space heating, water heating, pumps \& fans and lighting.
+The ArDEM-SQL, which was described in Section \ref{Intro:ESOMIreland} is a simulation model built on IS EN 13790.
 ArDEM-SQL is used to clean the input data in TIM, by changing internal temperatures.

### 2021-06-28 manuscript.tex
Done -

@@
 
-In 2014, Rogan et al. \cite{Rogan2014LEAPsSystem} used OSeMOYOS to explore residential energy service demands in Ireland
- LEAP writes an input file for OSeMOSYS, runs it and imports its results back into LEAP.
-Currently, the LEAP input file only contains the aggregated electricity demand as the main driver for the optimization within OSeMOSYS.
- the model examines the
- 3 main sceanrios - exploring transtport, residential, services and industy. The reference sceanrio - 
-REF - TFC increase by 6.7\% in 2020, base year is 2008.
-space and water heating TFC have decreased in share of total residential TFC from 82 to 77 \%, whereas lighting, cooking and appliances have increased in TFC from 18 to 23 \%.
-assume that 1.4 \% of existing dwellings per annum will undergo a shallow retrofit in the reference
-case, corresponding to just below 300,000 dwellings
+In 2014, Rogan et al.
+\cite{Rogan2014LEAPsSystem} linked OSeMOYOS, an ESOM to the Long Range Energy Alternatives Planning System (LEAP) simulation model, to explore specific energy efficiency policies not only in the residential sector, but also in transport, industry and services sector in Ireland.
+The bottom up development of the residential sector in LEAP was combined with the cost optimisation capabilities of OSeMOYOS.
+Three scenarios are explored, namely, the reference scenario, energy efficiency scenario and energy efficiency+ scenario.
+The reference assume that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020.
+In the reference scenario the residential total energy consumption rose 6.7 \% in 2020, compared to 2008.
+The share of space and water heating decreased from 82 \% to 77 \% in that period.
+In the energy efficiency scenario 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulted in a reduction in of energy consumption of 9.9 \% in 2020, compared to the reference scenario.
+The energy efficiency+ scenario also allows for 800,000 residential dwellings to be retrofitting, but accounts for deeper retrofits.
+The results in a reduction in of energy consumption of 16.5 \% in 2020, compared to the reference scenario.Rogan et al.
+discussed the barriers, obstacles to investment, split incentives and information gaps which restrict energy efficiency improvements in the residential sector.
+The importance for policy to not only look at number of retrofit, but also the depth of retrofit works, is shown by the difference in scenarios.
+The model also found considerable scope for energy savings to be made through improved building regulations.
 
-In the energy efficiency scenario, we assume that 800,000 residential dwellings will be retrofitted between 2009 and 2020, in line with government targets for the BEH scheme,  Other policies in the residential sector are building regulations 2011 and building regulations 2013 and CFL lighting,  The combined impact of these policies is a reduction in TFC of 9.9 \% compared to the reference scenario.
+In 2018, Thellufsen et al.
+\cite{Thellufsen2019ImplementingIreland} investigated the potential for the implementation of district heating in both the commercial and residential sectors of Ireland.
+The paper analyses the $CO_2-80$ scenario ( 80 \% $CO_2$ reduction in 2050, compared to 1990 ) with and without district heating using the previous Irish TIMES model and EnergyPLAN.
+The previous version of Irish TIMES did not account for district heating, so EnergyPLAN is used to investigate the operation and  feasibility of district heating in Ireland.
+A sensitivity analysis was preformed on investment costs, heat grid losses and discount rates, however in all cases the inclusion of district heating is optimal.
+The final results find district heating reduces fuel consumption by 3.5 \% and the conversion of 37 \% of the Irish heat demand to district heating is found to be optimal.
+Both the TIMES model and climate target are superseded and the study did not include the future heat savings expected from retrofitting, nevertheless this study has provided insights on the potential of district heating.
 
-The energy efficiency+scenario explores the impact of an ambitious retrofitting scenario beyond its current rate of implementation: retrofitbest is realised by retrofitting the same overall number of dwellings but undertaking a deeper, more comprehensive package of
-energy efficiency measures on each, resulting in 7.2 \% savings compared to the energy efficiency scenario and 16.5 \% savings compared to the reference scenario.
-  Insights - A challenging issue for energy modelling of policy is the
-barriers and obstacles to investment, particularly in
-residential energy efficiency measures such as split
-incentives and information gaps, which lead to suboptimal
-uptake of economically beneficial energy
-efficiency improvements.
-
-Thellufsen examining DH with EnergyPLAN and TIMES in 2019 \cite{Thellufsen2019ImplementingIreland}.
-
-In Ireland few models have used energy system optimisation model to explore the residential sector.
+\par In Ireland few models have used energy system optimisation model to explore the residential sector.
 The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector only models, rather than energy system models specifically examining the residential sector.
@@
 
-Early work by Healy and CLinch ---\cite{Clinch2000HousingMortality,Healy2002FuelIreland,Clinch2003ValuingModel,Clinch2000DomesticFailure,Clinch2001Cost-benefitEfficiency}
+Clinch and Healy analysed different aspects of thermal comfort in the Irish residential sector.
+This was the first comprehensive analysis of the residential sector, their studies examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure} and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2000DomesticFailure,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}.
+In 2000 Clinch and Healy developed and published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Mircsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating.
+A cost-benefit model (CBM) was fitted to the EAM to facilitate the conversion of the physical estimates into monetary amounts.
+Under the predicted scenario, energy savings easily outweighing programme costs, representing a benefit-cost ratio of 1.7.The benefits of reduced sickness, hospitalisation and drugs and mortality, were quantified, this moved the  benefit-cost ratio up to 3.0.
+The same model was used by Clinch, Healy and King \cite{Clinch2001ModellingEfficiency} in addition three models of householder spending on energy are developed here to to evaluate the human behavioural component
+into the computer model.
+the modelling process involved developing a computer model of the energy performance of the existing dwelling stock.
+It was noted, energy consumption predicted by the model is considerably higher than given in the national energy balance, due to intertal temperature assumptions.
+The model showed the energy-efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compared to the reference scenarios.
 
@@
 
-SEAI’s National Energy Modelling Framework (NEMF) is a suite of modelling tools used for modelling the Irish
-energy system, both supply and demand side, to inform Irish Government energy policy and project Ireland’s
-energy use based on these policie - https://www.seai.ie/publications/National-Energy-Projections-to-2030.pdf
-The macro-economic model currently in use is called COSMO (Core Structural Model of the Irish Economy),
-which creates a baseline forecast of energy consumption by sector (ESRI, 2017c). A range of sectoral models
-are used to analyse both supply and demand side energy. Electricity supply is modelled using a commercial
-power system software called PLEXOS47. A BioHeat model is used for techno-economic modelling of
-bioenergy supply chains in Ireland for the electricity generation, transport and heat sectors48. There is also a
-separate transport energy simulation model, which considers biofuels blending and the growth of low
-emissions vehicles in the vehicle stock, including electric vehicles. In addition, separate modelling by an
-energy-efficiency policy analysis tool (Unlocking model) informed the energy efficiency estimates included in
-the NEMF49.
+Finnaly, it is worth mentioning SEAI’s National Energy Modelling Framework (NEMF), which is a suite of modelling tools used for modelling the Irish energy system.
+Some of the main models used in the NEMF include:
+\begin{itemize}
+\item {Core Structural Model of the Irish Economy (COSMO)  is a macro-economic model used by the Economic and Social Research Institute (ESRI) to create a baseline forecast of energy consumption by sector} \cite{Bergin2017COSMO:Ireland}
+\item{PLEXOS is a power systems modeling tool used for electricity modeling and planning}\cite{PLEXOSExemplar}
+\item{Energy-efficiency policy analysis tool informs the energy efficiency estimates} \cite{Scheer2015UnlockingTob}
+\begin{itemize}
+\item{Consumer Choice Model determines the attractiveness of energy efficiency finance options}\cite{SustainableEnergyAuthorityofIreland2017BehaviouralSector}
+\end{itemize}
+\item{BioHEAT is a techno-economic simulation model of bioenergy and heat, used by SEAI to support policy decision in Ireland.}\cite{Durusut2018BioHEAT:Sectors}
+\end{itemize}
+
+The listed NEMF models and more are used by SEAI to model the impact of incentives and opportunities created by specific policy instruments. The models are a combination of top down, ..........
 
@@
 \label{Dis:Limit}
+%Discrepancies were investigated, and assumptions reviewed intuitively in order to achieve a good match.=
 Consumers’ final investment decision can be

### 2021-07-01 manuscript.tex
1-July-2021 Intro done, working on housing stock data.

@@
 \usepackage{amssymb}
+\usepackage{pifont}
 \usepackage{amsmath}
@@
 \usepackage{comment}
+\usepackage{booktabs}
+%\usepackage{multirow}
+
+%\usepackage[table]{xcolor}
 %\usepackage{cite}
@@
 Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM cover all the energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section.
+
 \par For example, Li et al.
@@
 
-%and some models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, but very
+ESOM models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, however few Irish studies use ESOM models to explore the residential sector, two examples of Irish ESOM focusing on the residential sector are detailed here.
 
@@
 
-\par In Ireland few models have used energy system optimisation model to explore the residential sector.
-The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector only models, rather than energy system models specifically examining the residential sector.
+\par  The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector only models, rather than energy system models specifically examining the residential sector.
 Nevertheless, the residential models have provided insights on input data and methodology approach used in Ireland, some of which are outlined below.
@@
 
-\par In 2015 Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examing new dwelling, to examining energy savings potential of energy efficiency improvement measures of the existing dwelling stock.
-The model relied on the BER database, which accounted for 18\% of all Irish dwellings, currently the database has more than doubled and accounts for over 50\%.
+\par In 2015 Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examining new dwelling, to examining energy savings potential of energy efficiency improvement measures of the existing dwelling stock.
+The Dwelling Energy Assessment Procedure (DEAP) model used, relied on the BER database, which accounted for 18\% of all Irish dwellings, currently the database has more than doubled and accounts for over 50\%.
+DEAP is the national tool for BER calculation and checking building regulation compliance.
 The model used 175 archetype dwellings ( 5 building types, 7 energy bands and 5 wall construction types ) and 6 different retrofit measures were applied.
@@
 
-Finnaly, it is worth mentioning SEAI’s National Energy Modelling Framework (NEMF), which is a suite of modelling tools used for modelling the Irish energy system.
+Finally, it is worth mentioning SEAI’s National Energy Modelling Framework (NEMF), which is a suite of modelling tools used for modelling the Irish energy system.
 Some of the main models used in the NEMF include:
@@
 
-The listed NEMF models and more are used by SEAI to model the impact of incentives and opportunities created by specific policy instruments. The models are a combination of top down, ..........
+The listed models are some of models used within SEAI's NEMF to model the impact of incentives and opportunities created by specific policy instruments.
+
+The models described in this section included a range of national ESOMs and bottom-up residential models.
+The models have varied objectives such as exploring consumer choices, PV battery systems location, evaluating health effects of retrofitting and integration of bio-energy and heat.
+The thermal efficiency of dwellings is a common objective in Irish models, due to the poor thermal efficiency of the dwelling stock and the every growing BER database is providing high data granularity input to these models.
+The internal temperature is also analysed in some of the studies, a recurring theme is the wide range of internal temperatures.
+The bottom-up residential models are typically simulation models looking at specific policy.
+The complexity of the models have increased significantly in the last 20 years, in Ireland the models starting out in Microsoft Excel and switching to Microsoft SQL.
+There has been few studies on the Irish residential sector using ESOMs, as the residential bottom-up models are preferred when focusing on that sector.
+The studies on Irish ESOM typically do not focus on the residential sector, instead they focus on the overall energy system, electricity generation or transport.
+The use of ESOM to explore DH in Ireland, allows for insights which the bottom-up simulation models cannot provide.
+The connection between waste heat in electricity and industry can show the optimal solution for district heating in commercial and residential sectors.
 
@@
 In Ireland the decarbonisation of the electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}.
-Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private invesvestment towards low carbon or renewable energy.
+Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private investment towards low carbon or renewable energy.
 A €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}.
@@
 The ambient heat used in heat pumps is included as a renewable energy, heat pumps are therefore the preferred heating technology despite the high capital costs.
-As mentioned in Section \ref{Intro:IreRSD}, new buildings in Ireland produce 70\% less GHG emissions than building in 2005, and with CAP2019 plans to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030, highlights the energy efficiency first approach in the residential sector.
+As mentioned in Section \ref{Intro:IreRSD}, new buildings in Ireland produce 70\% less GHG emissions than buildings in 2005, along with CAP2019 plans to retrofit 500,000 existing homes to at least a  ‘B2’ energy rating by 2030, highlights the energy efficiency first approach in the residential sector.
 \par Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019VisionIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
@@
 
-% The technical bottom-up development approach of the residential sector is combined with other economic sectors to meet macroeconomic defined demands, to provide GHG targets insights both the energy system and residential sector.
+%``How should Ireland set GHG targets in the residential sector?”
 
-%``How should Ireland set GHG targets in the residential sector?” 
-%Choosing the Pathways which Create the Least Burden and Offer the Most Opportunity for Ireland
-
-%FROM TIM Paper
-%TIM has been co-developed with the LEAP-Ireland model (Mac Uidhir et al., 2020), which is a bottom-up simulation modelof the Irish energy system which simulates the impact of different policy measures and targets on overall GHG emissions inIreland, with a particularly granular representation of transport and residential heat demand.
-Underlying data for the relevantsectors are shared between TIM and LEAP using a Data Repository and shared coding convention dictionary to improve the consistency between the models, give more robust analytical insights for policy and to share and exchange expertise between themodelling teams.
-This also facilitates multi-model approaches to energy systems modelling, which can make use of harmonisedhybrid frameworks coupling simulation and optimisation simultaneously (Rogan et al., 2014)
 \subsection{TIMES Energy System Optimisation Model}
@@
 
-%With update technological and macroeconomic data, structured specifically to analyse for sectoral five-year carbon budgets towards 2030 and 2050, it is hoped TIM will provide further and more robust insights than previous Irish TIMES model.
-
 %Add description of TIMES and how residential sector is represented to the Methodology section; - Olex Email
@@
 
+\subsection{Residential}
+\label{LitR:RES}
+The new structure and updated technological and macroeconomic data will provide more robust insights on sectoral five-year carbon budgets towards 2030 and 2050 than previous models.
+The residential sector was developed in parallel with other sectors in TIM, and concurrently with the new Irish LEAP model, which allows for more accessible soft-linking to provide more robust analytical insights by combining simulation and optimisation, and to exchange expertise between modelling teams.
+Fig.\ref{fig:RES} illustrates the flow of energy from fuel to demand in the residential sector.
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[scale=0.45]{Figures/TIM_RES3.png}
+ \caption{Residential Sector - Reference Energy System}
+ \label{fig:RES}
+\end{figure}
+
+The BER database provides the base year bottom-up energy consumption data, this is extrapolated to represent the entire stock and cross-checked with the top-down energy consumption from the 2018 SEAI Energy Balance.
+To compare from the two databases, fuels were were combined and condensed down to 13 fuels in TIM.
+Table \ref{Table:Fuels} shows an overview of the fuels combined from the different data sources.
+
+\begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Residential Fuels}
+   \begin{tabular}{lll}
+\hline
+\textbf{SEAI Energy Balance}  & \textbf{BER Database}  & \textbf{TIM}                  \\ \hline \hline
+Bituminous Coal & House   Coal & Coal                  \\
+Anthracite + Manufactured & Anthracite &         \\
+Lignite \textbackslash Brown Coal  & Manufactured Smokeless    &                       \\ \hline
+Sod Peat & Sod Peat & Peat                                 \\
+Briquettes  & Peat   Briquettes   &                       \\ \hline
+Kerosene & Heating   Oil   & Kerosene              \\
+Gasoil / Diesel /DERV   &       &                       \\
+Petroleum Coke            &                &         \\ \hline
+Natural Gas    & Mains   Gas          & Natural Gas           \\ \hline
+Biomass     & Wood   Chips    & Biomass               \\
+  & Wood   Logs   &       \\
+  & Wood   Pellets (bulk)  &                       \\
+ & Wood   Pellets (in bags ) &                       \\ \hline
+Solar Thermal &     & Solar                 \\
+Solar Photovoltaic             &         &            \\ \hline
+Ambient Heat     &   & Ambient Heat (Space)  \\
+   &    & Ambient Heat* (Water) \\ \hline
+Electricity        & Electricity  & Electricity           \\ \hline
+Heat            & Unknown     & Heat                  \\ \hline
+LPG     & Bottled   LPG        & LPG                   \\
+  & Bulk LPG       &               \\ \hline
+Biodiesel      & Biodiesel  & Biodiesel             \\ \hline
+  Bioethanol     & Bioethanol   & Ethanol               \\ \hline
+ & Solid Multi Fuel    & Coal                  \\
+ &      & Peat                  \\
+ &     & Wood                  \\ \hline
+\end{tabular}
+\label{Table:Fuels}
+\end{table}
+
+In TIM fuels such as coal, represent a combination of coal fuels from the two different databases.
+Kerosene accounts for over 70 \% of oil usage, while Gasoil accounts for most of the remaining share, the fuels have similar properties and there was no evidence to support keeping separate fuels would provide insights to residential decarbonisation.
+The exception is ambient heat, which could potentially provide insights to residential decarbonisation by seperating it into ambient heat for space heating and ambient heat for water heating.
+The solid multi-fuel from the BER database represents an open fire or stove which is heated by either coal, peat or wood.
+Through analysis of the SEAI Energy Balance and BER database, the fuel share assigned for coal, peat and wood is 19.4 \%, 74.4 \% and 6.2 \% respectively.
+There are two types of energy services represented in Table \ref{fig:RES} - Non-Archetype and Archetype.
+Non-archetype energy services fill in the lack of data on archetype energy services.
+The six non-archetype energy services, are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service, this bottom-up data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA (Policy Oriented Tool for Energy and Climate Change Impact Assessment) database and the SEAI Energy Balance, more details on non-archeype services are found in Section \ref{LitR:Lighting,Appliances}.
+The four archetype energy services are dependent on  building archetype and data is sourced from the BER database, which is discussed in Section \ref{LitR:StockData}.
+The internal temperature adjustment for space heating is detailed in Section \ref{LitR:StockData}.
+The demand for archetype and non-archetype energy services is based on the growth of residential dwellings, which is detailed in section \ref{LitR:Demands}.
+
+First iteration - of model will need further development DH, BEhavour, higher granularity and spatial will be easily done with dataset. 
 %What does new model do that previous didn’t?
@@
 
-\subsection{Residential}
-\label{LitR:RES}
-The residential sector was developed in parallel with other sectors in TIM, and also been structured concurrently and similarly to the residential sector in the new Irish LEAP model, to allow more accessible soft-linking.
-Fig.\ref{fig:RES} shows the schematic overview of the residential sector in TIM, illustrating the flow of energy from fuel to demand.
-First iteration - of model will need further development DH, BEhavour, higher granularity and spatial will be easily done with dataset. 
+\subsubsection{House Stock Data}
+\label{LitR:StockData}
+
+The BER database, which represents half of the dwelling stock, highlights the affect year of construction has on expected energy consumption, as shown in Fig.\ref{fig:BERAge}.
+The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
+The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, which helps to better understand the housing stock.
+BERs are required for new houses, rented houses, sold houses and renovated houses.
+
 \begin{figure}[!htbp]
  \centering
- \includegraphics[scale=0.45]{Figures/TIM_RES3.png}
- \caption{Residential Sector - Reference Energy System}
- \label{fig:RES}
+ \includegraphics[scale=0.5]{Figures/BERAge.png}
+ \caption{Building Energy Rating by Age}
+ \label{fig:BERAge}
 \end{figure}
 
-The 2018 SEAI Energy Balance supplies the base year fuel data, both the quantity and type of fuel.
-Where 15 residential fuel types are found in the SEAI Energy Balance, TIM uses 10 residential fuel types.
-For example, bituminous coal, anthracite + manufactured ovoids and lignite \ brown coal briquettes are combined into one fuel called ``Coal" in TIM.
-There are two types of energy services in TIM - Non-Archetype and Archetype.
-The six non-archetype energy services, are not dependant on the type of dwelling and the base year quantity is the total national consumption of that energy service, this data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA (Policy Oriented Tool for Energy and Climate Change Impact Assessment) database.
-The demand for non-archetype energy services is based on the growth of residential dwellings, which is detailed in section \ref{LitR:Demands}.
-The SEAI BER Public database, which is Ireland's national Energy Performance Certificate (EPC) register, complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD).
+ENERGY CONSUMPTION CHARACTERISTICS of Houses - Table?
+
+Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which is a reason for Ireland's high excessive winter mortality \cite{Clinch2000HousingMortality}.
+Fuel Poverty which which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths.
+An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}.
+Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty.
+These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}.
+Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}.
+Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
+\par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
+The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
+
+The SEAI BER Public database, which is Ireland's national Energy Performance Certificate (EPC) register, which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD).
 The four energy service type are derived from the database - space heating, water heating, pumps \& fans and lighting.
@@
 The ArDEM-SQL, which was described in Section \ref{Intro:ESOMIreland} is a simulation model built on IS EN 13790.
-ArDEM-SQL is used to clean the input data in TIM, by changing internal temperatures.
-The internal temperature is highly dependent on occupancy behaviour, a thermal comfort  thermal discomfort are higher among the over-65s than for other household \cite{Healy2002FuelIreland} - living room temperatues between 12 and 26...s  and therefore there is are large variations per BER rating.
-The International Standard ISO 7730 [6] uses these PMV and PPD indices to predict the thermal sensation of people exposed to moderate thermal environments, as well as to specify acceptable thermal environmental conditions for comfort...
-Despite this, it is
-worth noting that the most inefficient dwellings i.e. band F, managed to achieve an
-average temperature during period 2 that was only 0.5°C lower than the most
-efficient dwellings i.e. band C.
-C = 19.2, D = 18, E = 18.2 and F= 18.7 older occupants dublin Ho
+ArDEM-SQL alters primary energy consumption from the BER database, by changing internal temperatures which are used in TIM.
 %Healy & Clinch reported that fuel-poor households were more likely to be experiencing colder temperatures than other households.
@@
 %The BER for the sample dwellings ranged from C1 to F. Over two thirds of the
-sample dwellings had a BER of E or F. The remaining dwellings had a BER of C or
-D. In total 4 of the sample dwellings had undergone energy efficiency upgrades in
-recent years, including cavity wall insulation, attic insulation where applicable and
-heating control upgrades. In the case of two sample dwellings, these upgrade
-works were completed between monitoring period 1 and 2, and therefore results
-could be compared before and after retrofit.
 
-Non-Archetype - lack of data
-The fuel is calibrated to the
-
-DEMAND Space cooling was disaggregated from electricity in POTEnCIA, this space cooling this was used in TIM.
-POTEnCIA provides high resolution energy service demands in 2018, along with energy service projections to 2050.
-Behavioral Constraints here
-
-While also accounting for more fuels and technological options, the residential sector in TIM is structured with around retrofitting, not vice-versa as previous versions, so that it is more aligned to ongoing climate mitigation investments.
-.
-The Non-Archetype Demands are so, because of lack of data
-ENERGY CONSUMPTION CHARACTERISTICS of Houses - 
- %The BER database, which represents half of the dwelling stock, highlights the affect year of construction has on expected energy consumption, as shown in Fig.\ref{fig:BERAge}.
-The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
-The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, which helps to better understand the housing stock.
-BERs are required for new houses, rented houses, sold houses and renovated houses.
-
-\begin{figure}[!htbp]
- \centering
- \includegraphics[scale=0.5]{Figures/BERAge.png}
- \caption{Building Energy Rating by Age}
- \label{fig:BERAge}
-\end{figure}
-
-Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which is a reason for Ireland's high excessive winter mortality \cite{Clinch2000HousingMortality}.
-Fuel Poverty which which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths.
-An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}.
-Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty.
-These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}.
-Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}.
-Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
-\par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
-The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
-\subsubsection{House Stock Data}
-\label{LitR:StockData}
 EPC register EU - Ireland
@@
 Model calibration: archetype demand The
+The fuel is calibrated to the before calibration the fuel was 50\% too high
+
+DEMAND Space cooling was disaggregated from electricity in POTEnCIA, this space cooling this was used in TIM.
+POTEnCIA provides high resolution energy service demands in 2018, along with energy service projections to 2050.
+Behavioral Constraints here
+
 \subsubsection{Costs}
@@
 \label{LitR:Savings}
+While also accounting for more fuels and technological options, the residential sector in TIM is structured with around retrofitting, not vice-versa as previous versions, so that it is more aligned to ongoing climate mitigation investments.
+
 \subsubsection{Energy Service Demands Projections}
+\label{LitR:Demands}
+% The technical bottom-up development approach of the residential sector is combined with other economic sectors to meet macroeconomic defined demands, to provide GHG targets insights both the energy system and residential sector.
 \begin{equation} 
@@
 \subsubsection{Lighting, Appliances and Other}
+\label{LitR:Lighting,Appliances}
 Capacity, Efficiency, Constraints, Uptake, adapation

### 2021-07-05 manuscript.tex
5 July 2021

@@
 \usepackage{booktabs}
+\usepackage{rotating}
+\usepackage{siunitx}
+\usepackage{eurosym}
 %\usepackage{multirow}
+
+\DeclareSIUnit{\EUR}{\text{\euro}}
 
@@
 
-%Add description of TIMES and how residential sector is represented to the Methodology section; - Olex Email
-%The TIMES-Ireland Model (TIM) is an optimisation model of the Irish energy system, which calculates the cost-5optimal fuel and technology mix to meet future energy service demands in the transport, buildings, industry and agriculture sectors, while respecting constraints in greenhouse-gas emissions, primary energy resources and feasible deployment rates.
+TIM is an optimisation model of the Irish energy system, which calculates the cost-optimal fuel and technology mix to meet future energy service demands.
+The residential, power generation and transport sectors are rebuild with high granularity from the bottom-up, while industry, agriculture and services sectors have been updated.
+The model operates within the limits of GHG constraints, resources constraints, and feasible deployment rates.
 
@@
 
-The BER database provides the base year bottom-up energy consumption data, this is extrapolated to represent the entire stock and cross-checked with the top-down energy consumption from the 2018 SEAI Energy Balance.
+The BER database provides the base year bottom-up energy consumption data, this is scaled up to represent the entire stock and cross-checked with the top-down energy consumption from the 2018 SEAI Energy Balance.
 To compare from the two databases, fuels were were combined and condensed down to 13 fuels in TIM.
@@
 
-In TIM fuels such as coal, represent a combination of coal fuels from the two different databases.
-Kerosene accounts for over 70 \% of oil usage, while Gasoil accounts for most of the remaining share, the fuels have similar properties and there was no evidence to support keeping separate fuels would provide insights to residential decarbonisation.
-The exception is ambient heat, which could potentially provide insights to residential decarbonisation by seperating it into ambient heat for space heating and ambient heat for water heating.
+In TIM, fuels such as coal, represent a combination of coal fuels from the SEAI Energy Balance and BER database.
+Kerosene accounts for over 70 \% of home heating oil usage in 2018, while Gasoil accounts for most of the remaining share, the fuels have similar properties and there was no evidence to support keeping separate fuels would provide further insights to residential decarbonisation.
+The exception is ambient heat, which could potentially provide insights to residential decarbonisation by separating it into ambient heat for space heating and ambient heat for water heating.
 The solid multi-fuel from the BER database represents an open fire or stove which is heated by either coal, peat or wood.
@@
 Non-archetype energy services fill in the lack of data on archetype energy services.
-The six non-archetype energy services, are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service, this bottom-up data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA (Policy Oriented Tool for Energy and Climate Change Impact Assessment) database and the SEAI Energy Balance, more details on non-archeype services are found in Section \ref{LitR:Lighting,Appliances}.
+The six non-archetype energy services, are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service, this bottom-up data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA (Policy Oriented Tool for Energy and Climate Change Impact Assessment) database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{LitR:Lighting,Appliances}.
 The four archetype energy services are dependent on  building archetype and data is sourced from the BER database, which is discussed in Section \ref{LitR:StockData}.
@@
 
-First iteration - of model will need further development DH, BEhavour, higher granularity and spatial will be easily done with dataset. 
-%What does new model do that previous didn’t?
-%TIM residential improvements - DH, HVO, Solar ( in progress ), better graps the cost and savings of retrofits and heat pumps, which can infrom MACC.Build with 2018 baseyear, not 2010 as previous versions had - this suits the 2030 target
-%THe implications of net-zero
-%Does new policy context require new modelling approach?
-%Yes, do account for ongoing investments in climate mitigation opitions, the approach should provide higher granularity for likely pathways
-%How does new modelling structure provide insights (give examples) in a way that previous model didn’t?
-%Demands - 
-%Improved more detailed Energy Service Demands
-
-\subsubsection{House Stock Data}
+\subsubsection{Dwelling Stock Data}
 \label{LitR:StockData}
 
-The BER database, which represents half of the dwelling stock, highlights the affect year of construction has on expected energy consumption, as shown in Fig.\ref{fig:BERAge}.
-The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
-The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, which helps to better understand the housing stock.
-BERs are required for new houses, rented houses, sold houses and renovated houses.
-
-\begin{figure}[!htbp]
- \centering
- \includegraphics[scale=0.5]{Figures/BERAge.png}
- \caption{Building Energy Rating by Age}
- \label{fig:BERAge}
-\end{figure}
-
-ENERGY CONSUMPTION CHARACTERISTICS of Houses - Table?
-
-Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which is a reason for Ireland's high excessive winter mortality \cite{Clinch2000HousingMortality}.
-Fuel Poverty which which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths.
-An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}.
-Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty.
-These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}.
-Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}.
-Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
-\par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
-The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
-
-The SEAI BER Public database, which is Ireland's national Energy Performance Certificate (EPC) register, which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD).
-The four energy service type are derived from the database - space heating, water heating, pumps \& fans and lighting.
-The space heating calculation (IS EN 13790) used to estimate the expected energy consumption per dwelling, makes some assumptions.
-The internal temperature assumption has a large error, therefore TIM has made adjustments so that the total consumption of energy services is calibrated to the base year data.
-IS EN 13790 assumes all living areas ( living room, sitting room and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C.
-We know this is not the case, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions in TIM.
+The dwelling stock data is obtained mainly from the SEAI BER Public database, which is firstly filtered, then scaled up using Central Statistics Office (CSO) data. 
+The BER database is Ireland's national Energy Performance Certificate (EPC) register, which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD).
+The BERs are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
+Over half of Irish dwellings are captured in the BER database, every BER displays over 200 energy related variables.
+The DEAP model, which was outlined by Dineen et al.\cite{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM} provides the results for the energy rating in the BER database.
+However, the space heating calculation used to estimate the expected energy consumption per dwelling, which makes some assumptions.
+DEAP assumes that each dwelling is single zoned, the heating season is frin October to May inclusive, and the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season.
+One assumption, the internal temperature assumption results in a large error, where space and water heating are overestimated in DEAP.
+Space heating consumption reduced by 26 \% after internal temperature adjustment were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance.
+DEAP assumes all living areas ( living room, sitting room, and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C.
+Through analysis of previous residential modelling in Ireland this is not the case, as the internal temperatures are often much lower, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions applied in TIM.
 
@@
 The ArDEM-SQL, which was described in Section \ref{Intro:ESOMIreland} is a simulation model built on IS EN 13790.
-ArDEM-SQL alters primary energy consumption from the BER database, by changing internal temperatures which are used in TIM.
-%Healy & Clinch reported that fuel-poor households were more likely to be experiencing colder temperatures than other households.
-In total almost 30% of fuelpoor households had living room temperatures below 18°C and more than two thirds had living room temperatures below 20°C.
-Overall nearly half of all non-fuel-poor homes had living room temperatures below 20°C.
-In households with persons aged over 65 years, 16% had living room temperatures below 18°C and over 50% had temperatures below 20°C.
-On the other end of the scale over 5% of households occupied by persons over 65 years experienced living room temperatures of ≥24°C.https://arrow.tudublin.ie/cgi/viewcontent.cgi?article=1089&context=scienmas
-%The BER for the sample dwellings ranged from C1 to F. Over two thirds of the
-
-EPC register EU - Ireland
-%The quality of Swedish EPC data was initially studied by Göransson [41], subsequently by Claesson [42], Mangold et al.
-[21] and Hårsman et al.
-[43], and most recently by Pasichnyi et al.
-[6].
-These studies have found a couple of important deficiencies in Swedish EPC data.
-First, the energy expert, and the firm he or she belongs to, has been shown to generate variations in Swedish EPC data [43], [44].
-More specifically, Hårsman et al.
-[43] found that the energy expert along with the characteristics of the firm caused a variation of ±20% in the assessment of energy consumption.
-%Second, through interviews with energy experts, Claesson [42] found that most EPCs contain uncertain estimates of energy use as well as arbitrarily distributed values—thus challenging the notion that Swedish EPC data quality, while allegedly measured, should be superior to the quality of calculated values in other member states.
-CIARAs Paper
-Every BER displays over 200 energy related variables per dwelling, and over half of all residential dwellings are accounted. This large and detailed BER database does
-BER Database, CSO and Calculations and Filtration 
-FIGURE - Expected and Actual Consumption per m2
-Than Actual 
-DEAP - The space heat use is defined in IS EN 13790 as the heat delivered to the heated space by an ideal heating
-system to maintain the set-point temperature during a given period of time. The DEAP calculation of space
-heat use is done on a monthly basis based on the procedure described in IS EN 13790. A single-zone
-calculation is used, using a single average value of mean internal temperature as described in Section 7.In DEAP, the heating season is defined as running from October to May inclusive.
-The annual space heat use
-is the sum of monthly values for these eight months.
-The heating hours and required internal temperatures in DEAP are based on the requirements of a typical
-household. The schedule is as follows.
-• Weekdays: 07.00 to 09.00 and 17.00 to 23.00
-• Weekends: 07.00 to 09.00 and 17.00 to 23.00
-This standardised schedule represents a total heating period of 56 hours per week.
-
-Scaling BER to CSO
+ArDEM-SQL alters primary energy consumption from the BER database by changing the internal temperatures.
+
+Other concerns for uncertainties in the database, which are not related to DEAP calculation are statistically anomalous spikes in the data, caused by applying default values \cite{Ahern2020ADatabases} and different companies generating variations of ±20\% in the assessment of energy consumption \cite{Harsman2016OnCertificates}.While the BER databases provides high granularity data, it must be filtered to improve the representation of the dwelling stock data.
+The filters applied to the raw BER database are outlined in \ref{sec:appendixA}.
+
+The Census 2016 data \cite{CensusOffice} from the CSO is also used in calculating the dwelling stock.
+Dataset E1005 provides a split of seven archetypes per county, which are aggregated to form three archetypes.
+The total number of building archetypes per county in 2016 is used to scale up the BER database.
+The gap of total dwelling between the Cenus year in 2016 and the model base year in 2018 was initially ignored due to lack of data.
+However, later it was found that data on new dwelling electrical connections per archetype and per county are available from CSO dataset NDQ06 this would have filled the gap, the first iteration of the dwelling stock has a 2 \% error.
+
+Once both database were finalised, the next step in this process, was to scale up the BER database.
+
+This was done by first finding the ratio of the building archetype per county between the BER database and CSO data, as shown in Equation \ref{Eq:BERtoCSO1}.
+
 \begin{equation} 
 \begin{split}
-S_y =  \frac{C_y} {B_y}
+R_{a,c} = {\frac{CSO_{a-nsa,c}} {BER_{a,c}} + \frac{CSO_{nsa,c}} {CSO_{a,c}}}
 \end{split}
-%\[ \sum_{n=1}^{\infty} 2^{-n} = 1 \]
-\label{Eq:BERtoCSO}
+\label{Eq:BERtoCSO1}
 \end{equation}
-The number of dwellings by year of construction within the
-BER database (B y ) is then scaled to align with the total number of dwellings as they appear in the CSO dataset (C y ) –for each archetype.
-Eq.
-(8) describes the scaling factor (S y ), used for each year of construction bracket
-
+
+$R_{a,c}$ is the ratio per archetype (a) and per county (c), there are 26 counties and 3 archetypes.
+$CSO_{a-nsa,c}$ represents the number of dwelling archetypes minus the non-stated archetypes (nsa) per county in the CSO database.
+$BER_{a,c}$ is the number of archetypes per county  in the BER database.
+The next step is to multiply the number of BERs per archetype and per county by the relevant ratio.
 \begin{equation} 
 \begin{split}
-A_y =  {S_y} \cdot {B_{yarc}}
+BER_{a,c} * R_{a,c} = TIM_{a,c}
 \end{split}
@@
 \end{equation}
-Number of archetypes dwellings by year of construction , where A k = Frequency of each archetype, S y = Scaling Factor, B yarc = Number Dwellings by BER category, A k = Frequency of each archetype.
-
-Model calibration: archetype demand The
-The fuel is calibrated to the before calibration the fuel was 50\% too high
-
-DEMAND Space cooling was disaggregated from electricity in POTEnCIA, this space cooling this was used in TIM.
-POTEnCIA provides high resolution energy service demands in 2018, along with energy service projections to 2050.
-Behavioral Constraints here
+
+$TIM_{a,c}$ represents the final dwelling stock data per archetype and per county, the first iteration of TIM does not disaggregate the dwellings per county, instead a national total per archetype is input to TIM ( $TIM_a$ ).
+
+Once the final dwelling stock is complete, the next step was to assign a BER rating to all dwellings.
+This was done by comparing 9 different dwelling age ranges for each of the 3 archetypes between the BER and CSO datasets.
+
+\begin{equation} 
+\begin{split}
+TIM_{r,y,a} = CSO_{y,a} * \frac{BER_{r,y,a}} {BER_{y,a}} * \frac{TIM_{a}} {BER_{a}}
+\end{split}
+\label{Eq:AgeBER}
+\end{equation}
+
+Where $TIM_{r,y,a}$ is the total dwellings in TIM by BER rating (r), year of construction (y) and archerype (a).
+$CSO_{y,a}$ is the number of dwellings by year of construction and archetype in the CSO dataset, the CSO dataset does not contain BER rating types.
+$BER_{r,y,a}$ is the number of BER ratings by year of construction and archetype from the BER database and dividing this value by $BER_{y,a}$ results in a probability of a certain age type falling within a BER rating, and the final part of the equation $\frac{TIM_{a}} {BER_{a}}$ compensates for the non-stated archetypes in the CSO data, $TIM_a$ was already calculated in equation \ref{Eq:BERtoCSO2}.
+The results of each of the correlations are in  \ref{sec:appendixA2}.
+Table \ref{Table:FinalDwellingStock} outlines the final dwelling stock used in TIM.
+
+\begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Final Dwelling Stock}
+\begin{tabular}{l|lll}
+BER Rating & Apartment & Attached & Detached \\ \hline
+A          & 9,419     & 15,472   & 20,379   \\
+B1         & 7,459     & 9,538    & 9,434    \\
+B2         & 14,042    & 17,545   & 20,091   \\
+B3         & 19,924    & 49,769   & 53,466   \\
+C          & 58,905    & 282,152  & 251,319  \\
+D          & 43,739    & 187,627  & 166,668  \\
+E          & 25,768    & 101,419  & 81,397   \\
+F          & 12,331    & 50,124   & 43,241   \\
+G          & 15,211    & 52,707   & 78,436   \\ \hline
+Total      & 206,799   & 766,352  & 724,430 
+\end{tabular}
+\label{Table:FinalDwellingStock}
+\end{table}
 
@@
 \label{LitR:Costs}
-Elephant in the room - Fuel Poverty
+
+This section only covers the costs applied in the residential sector, which includes space and water heating technologies, retrofitting, cooking, lighting, pumps, fans, space cooling, and other electrical appliances.
+The fuel costs are not applied in residential sector, there are applied in the supply sector, and the data is sourced from World Energy Outlook reports from 2019 and 2020. 
+The majority of the space and water heating cost data is obtained from the
+\textit{``Techno-economic projections until 2050 for smaller heating and cooling technologies in the residential and tertiary sectors in the EU”} report \cite{JointResearchCentre2017Techno-economicEU}.
+The cost data used in TIM was based on the north EU region (Denmark) and where data was missing, the central EU region (Germany) cost data was used.
+The high granularity of data, allows for cost of a technology to be disaggregated between apartment and non-apartment, costs and efficiencies are projected every five years from 2020 to 2050.
+The lifetime and emissions/pollutants are also included, while the cost data is further disaggregated between capital expenditure ( equipment, installation \& additional capital costs ) and Operational \& Maintenance (O\&M) costs.
+\par The model has two types of retrofitting - shallow ( BER is improved by $ \leq 35 \%$ and deep ( BER is improved by $>35 \%$ ).
+The cost data was sourced from Ali et.
+al \cite{Ali2020ABuildings} who derived costs from SEAI's Better Energy Homes (BEH) and Better Energy Warmer Homes (BEWH) datasets, which are Ireland's largest retrofitting dataset ( over 285,000 retrofits ).
+To provide more robust costs, costs data is also used from the Typology Approach for Building Stock Energy Assessment (TABULA) report \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} which has more detailed cost data on 2 stages of retrofit for a small sample of 31 dwellings.
+The formula used to find the average cost of retrofit per type of retrofit and archetype, are shown in Equations \ref{Eq:RetrofitCost1} and \ref{Eq:RetrofitCost2}.
+
+\begin{equation} 
+\begin{split}
+ \text{\euro}_{r,a,t} = Average ( (Ali_{r,t} * AF),TABULA_{r,a,t} )
+\end{split}
+\label{Eq:RetrofitCost1}
+\end{equation}
+
+Where $ {\text{\euro}}{}_{r,a,t}$ is the TIM cost for retrofit by rating (r), archetype (a) and type of retrofit (t).
+$Ali_{r,t}$ is the cost data from Ali et al.
+\cite{Ali2020ABuildings}, it contains all ratings and types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments and 1.1 for detached dwellings.
+The $TABULA_{r,a,t}$ is the average cost of retrofit from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} per rating, archetype and type of retrofit, however with only 62 samples in total not all retrofits are included.
+In this case TABULA is ignored and Ali et al.
+is the only source of retrofit costs.
+
+\begin{equation} 
+\begin{split}
+\text{\euro}_{a,t} = \text{\euro}_{r,a,t} * \frac{TIM_{r,a}} {TIM_{a}}
+\end{split}
+\label{Eq:RetrofitCost2}
+\end{equation}
+
+The model uses the weighted average cost of retrofit per archetype
+
+in the residential and tertiary sector in the EU 
+Topten -
+
+\subsubsection{Savings}
+\label{LitR:Savings}
+Retrofit Savings - SEAI dataset not published..papers Usman,etc.
+While also accounting for more fuels and technological options, the residential sector in TIM is structured with around retrofitting, not vice-versa as previous versions, so that it is more aligned to ongoing climate mitigation investments.
+
+\subsubsection{Energy Service Demands Projections}
+\label{LitR:Demands}
 %JF -The tax / subsidy inclusive prices are important in driving behaviour, and taxes and subsidies will be a vital policy instrument.
 Thus, uptake of technologies such as retrofitting will be driven, inter alia, by the net price for consumers.
-\subsubsection{Savings}
-\label{LitR:Savings}
-While also accounting for more fuels and technological options, the residential sector in TIM is structured with around retrofitting, not vice-versa as previous versions, so that it is more aligned to ongoing climate mitigation investments.
-
-\subsubsection{Energy Service Demands Projections}
-\label{LitR:Demands}
+ENERGY CONSUMPTION CHARACTERISTICS of Houses - Table?
 % The technical bottom-up development approach of the residential sector is combined with other economic sectors to meet macroeconomic defined demands, to provide GHG targets insights both the energy system and residential sector.
@@
 \subsubsection{Space and Water Heating}
+\label{LitR:Heating}
+
+Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which is a reason for Ireland's high excessive winter mortality \cite{Clinch2000HousingMortality}.
+Fuel Poverty which which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths.
+An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}.
+Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty.
+These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}.
+Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}.
+Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
+\par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
+The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
+
+ Build national fuel
+The CSO data, is mainly data from the 2016 Cenus. 
+Fuel Calibration
+ E1055 - Year, Archetype ( Detached, Semi-detached, terraced, Flat or apartment, converted to flat or apartment, bed-sit and not stated )
+E1008 Fuel by Archetype 
+ EXXX - Age, Archetype and Fuel 
+ dataset provides the central heating type or main fuel and
+
+Model calibration: archetype demand The
+The fuel is calibrated to the before calibration the fuel was 50\% too high
+
+DEMAND Space cooling was disaggregated from electricity in POTEnCIA, this space cooling this was used in TIM.
+POTEnCIA provides high resolution energy service demands in 2018, along with energy service projections to 2050.
+Behavioral Constraints here
+
+FIGURE - Expected and Actual Consumption per m2
 ISO EN 13790 (Energy Performance of Buildings Directive).
@@
 
+First iteration - of model will need further development DH, BEhavour, higher granularity and spatial will be easily done with dataset. 
+%What does new model do that previous didn’t?
+%TIM residential improvements - DH, HVO, Solar ( in progress ), better graps the cost and savings of retrofits and heat pumps, which can infrom MACC.Build with 2018 baseyear, not 2010 as previous versions had - this suits the 2030 target
+%THe implications of net-zero
+%Does new policy context require new modelling approach?
+%Yes, do account for ongoing investments in climate mitigation opitions, the approach should provide higher granularity for likely pathways
+%How does new modelling structure provide insights (give examples) in a way that previous model didn’t?
+%Demands - 
+%Improved more detailed Energy Service Demands
 Detailed bottom up model - using 2018 base year ( in line with policy )
@@
 
-\section{Sample Appendix Section}
-\label{sec:sample:appendix}
-XXXXXXXX
+\section{}
+\label{sec:appendix}
+
+\subsection{}
+\label{sec:appendixA}
+BER Database Filters
+\begin{table}[htbp]
+ \centering
+ \footnotesize
+ \caption{BER Filters}
+ \begin{tabular}{ccc}
+ \hline
+ Filter Name & Data Removed & Reason \\
+ \hline
+ Ground Floor Area &  $<30m^2$ \& $>1000m^2$ & Considered erroneous \\
+ Provisional Records & All & Higher Inaccuracies  \\
+ Thermal Bridging Factor & $\leq0.15 W/m^2K$ & Outlier \\
+ Declared Loss Factor & $>20kWh/day$ & Considered erroneous \\
+ Living Area Percent & $<5\%$ \& $>90\%$ & Outlier \\
+ Heat System Supply Heat Fraction & $>21\%$ & Outlier  \\
+ Heat System Efficiency Adjustment Factor & $<0.7$ & Considered erroneous \\
+ Main Space System Efficiency & $\leq19\%$ & Considered erroneous \\
+Water System Efficiency Adjustment Factor & $<0.7$ & Considered erroneous\\
+Main Water System Efficiency & $\leq19\%$ \& $>320\%$ & Considered erroneous \\
+ \hline
+ \end{tabular}%
+ \label{BERFilters}%
+\end{table}%
+
+\subsection{}
+\label{sec:appendixA2}
+Age to BER Rating Correlation Index
+\begin{sidewaystable}
+
+\centering
+ \footnotesize
+ \caption{Apartment Age to BER Rating Correlation}
+\begin{tabular}{l|lllllllll}
+Year   & A      & B1     & B2     & B3     & C      & D      & E      & F      & G      \\ \hline
+Before\_1919 & 0.0005 & 0.0021 & 0.0066 & 0.0139 & 0.0865 & 0.1618 & 0.1889 & 0.1447 & 0.3950 \\
+1919-1945    & 0.0007 & 0.0075 & 0.0457 & 0.0689 & 0.1646 & 0.1775 & 0.1614 & 0.1100 & 0.2636 \\
+1946-1960    & 0.0069 & 0.0146 & 0.0434 & 0.0553 & 0.1610 & 0.2085 & 0.1843 & 0.1139 & 0.2122 \\
+1961-1970    & 0.0097 & 0.0173 & 0.0692 & 0.0736 & 0.2060 & 0.2208 & 0.1484 & 0.0957 & 0.1593 \\
+1971-1980    & 0.0005 & 0.0014 & 0.0215 & 0.0305 & 0.2556 & 0.2809 & 0.2032 & 0.1009 & 0.1055 \\
+1981-1990    & 0.0002 & 0.0003 & 0.0035 & 0.0249 & 0.2278 & 0.3223 & 0.2589 & 0.1036 & 0.0585 \\
+1991-2000    & 0.0003 & 0.0016 & 0.0095 & 0.0293 & 0.3052 & 0.3728 & 0.2069 & 0.0546 & 0.0199 \\
+2001-2010    & 0.0049 & 0.0310 & 0.0938 & 0.1504 & 0.4272 & 0.2033 & 0.0711 & 0.0140 & 0.0042 \\
+Later\_2010  & 0.7996 & 0.0837 & 0.0713 & 0.0246 & 0.0140 & 0.0039 & 0.0019 & 0.0004 & 0.0005
+\end{tabular}
+
+\centering
+ \footnotesize
+ \caption{Attached Age to BER Rating Correlation}
+\begin{tabular}{l|lllllllll}
+Year         & A      & B1     & B2     & B3     & C      & D      & E      & F      & G      \\ \hline
+Before\_1919 & 0.0021 & 0.0027 & 0.0091 & 0.0245 & 0.1239 & 0.1974 & 0.2109 & 0.1458 & 0.2837 \\
+1919-1945    & 0.0019 & 0.0028 & 0.0102 & 0.0310 & 0.1835 & 0.2348 & 0.2046 & 0.1340 & 0.1973 \\
+1946-1960    & 0.0023 & 0.0050 & 0.0130 & 0.0305 & 0.1991 & 0.2585 & 0.2309 & 0.1312 & 0.1295 \\
+1961-1970    & 0.0032 & 0.0023 & 0.0100 & 0.0363 & 0.2757 & 0.3134 & 0.2021 & 0.0915 & 0.0654 \\
+1971-1980    & 0.0022 & 0.0024 & 0.0069 & 0.0382 & 0.3452 & 0.3329 & 0.1738 & 0.0619 & 0.0365 \\
+1981-1990    & 0.0010 & 0.0010 & 0.0069 & 0.0447 & 0.4263 & 0.3486 & 0.1198 & 0.0326 & 0.0190 \\
+1991-2000    & 0.0011 & 0.0043 & 0.0084 & 0.0469 & 0.4833 & 0.3389 & 0.0835 & 0.0250 & 0.0087 \\
+2001-2010    & 0.0045 & 0.0144 & 0.0438 & 0.1439 & 0.6330 & 0.1208 & 0.0303 & 0.0077 & 0.0016 \\
+Later\_2010  & 0.9359 & 0.0391 & 0.0147 & 0.0064 & 0.0032 & 0.0006 & 0.0000 & 0.0000 & 0.0000
+\end{tabular}
+
+\centering
+ \footnotesize
+ \caption{Detached Age to BER Rating Correlation}
+\begin{tabular}{l|lllllllll}
+Detached     & A      & B1     & B2     & B3     & C      & D      & E      & F      & G      \\ \hline
+Before\_1919 & 0.0035 & 0.0036 & 0.0078 & 0.0159 & 0.0912 & 0.1570 & 0.1791 & 0.1422 & 0.3996 \\
+1919-1945    & 0.0044 & 0.0035 & 0.0083 & 0.0200 & 0.1059 & 0.1728 & 0.1762 & 0.1354 & 0.3734 \\
+1946-1960    & 0.0055 & 0.0053 & 0.0127 & 0.0215 & 0.1298 & 0.2189 & 0.1928 & 0.1353 & 0.2782 \\
+1961-1970    & 0.0061 & 0.0046 & 0.0118 & 0.0229 & 0.1770 & 0.2910 & 0.2240 & 0.1132 & 0.1494 \\
+1971-1980    & 0.0034 & 0.0027 & 0.0090 & 0.0270 & 0.2663 & 0.3586 & 0.1942 & 0.0729 & 0.0660 \\
+1981-1990    & 0.0027 & 0.0028 & 0.0091 & 0.0422 & 0.3641 & 0.3920 & 0.1285 & 0.0349 & 0.0237 \\
+1991-2000    & 0.0019 & 0.0027 & 0.0152 & 0.0754 & 0.5372 & 0.2911 & 0.0519 & 0.0163 & 0.0082 \\
+2001-2010    & 0.0081 & 0.0216 & 0.0643 & 0.1785 & 0.5903 & 0.1112 & 0.0172 & 0.0059 & 0.0029 \\
+Later\_2010  & 0.8131 & 0.0868 & 0.0509 & 0.0295 & 0.0165 & 0.0029 & 0.0002 & 0.0002 & 0.0001
+\end{tabular}
+\end{sidewaystable}
 
@@
 %%
-
+\newpage
 \printbibliography

### 2021-07-05 manuscript.tex
Writeful Free - Done

@@
 %The TIMES-Ireland Model (TIM) is a newly developed optimisation model of the Irish energy system, which calculates the  cost-optimal decarbonisation pathway to meet future energy service demands while achieving Ireland's legally binding 2030 and 2050 climate targets.
-Write after paper is completed
+Write after the paper is completed
 
@@
 Section \ref{LitR} describes the methodology used to develop the residential sector while also outlining the input data.
-The process of applying constraints, technologies, costs and demands in the model.
-Section \ref{Results} outlines the results obtained from selected scenarios.
+The process of applying constraints, technologies, costs, and demands in the model.
+Section \ref{Results} outlines the results obtained from the selected scenarios.
 Section \ref{Discussion} discusses the strengths and weakness of the newly developed model and how it differs from other models.
@@
 
-Energy System Optimisation Models (ESOMs) can be used to explore least cost pathways, as they consider a large amount of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
-Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM cover all the energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section.
+Energy System Optimisation Models (ESOMs) can be used to explore the least cost pathways, as they consider a large number of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
+Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM covers all energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section.
 
@@
 As TIMES \cite{Loulou2016DocumentationI} assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
-Li set out to incorporate a Discrete Choice Model (DCM) into TIMES to better reflect how homeowners choose heating technologies.
+Li set out to incorporate a Discrete Choice Model (DCM) for TIMES to better reflect how homeowners choose heating technologies.
 The DCM was developed from a nationwide survey with 442 homeowners completing the survey.
 The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems.
-The DCM then estimated the probability of a heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices.
+The DCM then estimated the probability of heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices.
 While Li's researcher has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights, through more surveys and analysis.
@@
 Leibowicz et al.
-also finds electrification to be the dominant strategy for cost-effectively decarbonizing urban residential dwellings.
-These insights show the strength of an ESOM, but the not all costs are taken into account in Leibowicz study, the capital cost associated with more thermally efficient buildings are not included.
-
-Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised  Balmorel model \cite{Wiese2018BalmorelModel} to handle the integrated power and district heating system, to analyse how heat pumps with complementing heat storages affect wind power investments in Denmark.
-The model was disaggreated between a western and eastern Denmark region and the model covers Norway, Sweden, Finland and Germany, in order to cover import/export.
-Exogenous user constraints applied to the model, based on current Danish policy includes: Wind power must provide at least 50\% of the national electricity consumption by 2020, coal is phased out in heat and power plants by 2030 and Individual oil boilers are phased out by 2030.
-The results by find when investments in individual heat pumps are allowed, air to water heat pumps are installed in non-district heating areas.
+also found electrification to be the dominant strategy for cost-effectively decarbonizing urban residential dwellings.
+These insights show the strength of an ESOM, but not all costs are taken into account in Leibowicz study, the capital costs associated with more thermally efficient buildings are not included.
+
+Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised  Balmorel model \cite{Wiese2018BalmorelModel} to handle the integrated power and district heating system, to analyse how heat pumps with complementing heat storage affect wind power investments in Denmark.
+The model was disaggreated between a western and eastern Denmark region and the model covers Norway, Sweden, Finland, and Germany, in order to cover import/export.
+Exogenous user constraints applied to the model, based on the current Danish policy includes: Wind power must provide at least 50\% of the national electricity consumption by 2020, coal is phased out in heat and power plants by 2030, and individual oil boilers are phased out by 2030.
+The results find when investments in individual heat pumps are allowed, air-to-water heat pumps are installed in non-district heating areas.
 This represents a significant electricity demand in identifying individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments and reducing system costs, fuel consumption, and $CO_2$ emissions.
@@
 
-Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model \cite{stergaard2015ReviewingSimulations} to explore the feasibility of photovoltaic (PV) systems installed at household level in combination with batteries in Samsø, which is known for being Denmark’s renewable energy island.
+Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model \cite{stergaard2015ReviewingSimulations} to explore the feasibility of photovoltaic (PV) systems installed at the household level in combination with batteries in Samsø, which is known for being Denmark’s renewable energy island.
 The two approaches are observed for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
-Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy systems perspective, one reason for this is the wind power could also be stored in communal batteries.
-When identical communal and individual battery capacities where compared, it was fond  64\% of residential electricity demand is satisfied by a individual battery and only 36\% by a communal battery.
+Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy system perspective, one reason for this is the wind power could also be stored in communal batteries.
+When identical communal and individual battery capacities where compared, it was fond  64\% of residential electricity demand is satisfied by an individual battery and only 36\% by a communal battery.
 While the battery location was explored, the PV system location was not, further investigation into the optimal PV system location could provide further insights.
@@
 In 2014, Rogan et al.
-\cite{Rogan2014LEAPsSystem} linked OSeMOYOS, an ESOM to the Long Range Energy Alternatives Planning System (LEAP) simulation model, to explore specific energy efficiency policies not only in the residential sector, but also in transport, industry and services sector in Ireland.
-The bottom up development of the residential sector in LEAP was combined with the cost optimisation capabilities of OSeMOYOS.
+\cite{Rogan2014LEAPsSystem} linked OSeMOYOS, an ESOM to the Long Range Energy Alternatives Planning System (LEAP) simulation model, to explore specific energy efficiency policies not only in the residential sector, but also in transport, industry and services sectors in Ireland.
+The bottom-up development of the residential sector in LEAP was combined with the cost optimisation capabilities of OSeMOYOS.
 Three scenarios are explored, namely, the reference scenario, energy efficiency scenario and energy efficiency+ scenario.
@@
 The share of space and water heating decreased from 82 \% to 77 \% in that period.
-In the energy efficiency scenario 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulted in a reduction in of energy consumption of 9.9 \% in 2020, compared to the reference scenario.
+In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulting in a 9.9 \% reduction of energy consumption in 2020, compared to the reference scenario.
 The energy efficiency+ scenario also allows for 800,000 residential dwellings to be retrofitting, but accounts for deeper retrofits.
 The results in a reduction in of energy consumption of 16.5 \% in 2020, compared to the reference scenario.Rogan et al.
-discussed the barriers, obstacles to investment, split incentives and information gaps which restrict energy efficiency improvements in the residential sector.
-The importance for policy to not only look at number of retrofit, but also the depth of retrofit works, is shown by the difference in scenarios.
+discussed the barriers, obstacles to investment, split incentives, and information gaps which restrict energy efficiency improvements in the residential sector.
+The importance for the policy to not only look at the number of retrofits, but also the depth of retrofit work, is shown by the difference in scenarios.
 The model also found considerable scope for energy savings to be made through improved building regulations.
@@
 The previous version of Irish TIMES did not account for district heating, so EnergyPLAN is used to investigate the operation and  feasibility of district heating in Ireland.
-A sensitivity analysis was preformed on investment costs, heat grid losses and discount rates, however in all cases the inclusion of district heating is optimal.
-The final results find district heating reduces fuel consumption by 3.5 \% and the conversion of 37 \% of the Irish heat demand to district heating is found to be optimal.
+A sensitivity analysis was preformed on investment costs, heat grid losses, and discount rates, however in all cases the inclusion of district heating is optimal.
+The final results find that district heating reduces fuel consumption by 3.5 \% and the conversion of 37 \% of the Irish heat demand to district heating is found to be optimal.
 Both the TIMES model and climate target are superseded and the study did not include the future heat savings expected from retrofitting, nevertheless this study has provided insights on the potential of district heating.
 
-\par  The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector only models, rather than energy system models specifically examining the residential sector.
-Nevertheless, the residential models have provided insights on input data and methodology approach used in Ireland, some of which are outlined below.
+\par  The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector models, rather than energy system models specifically examining the residential sector.
+Nevertheless, the residential models have provided insights on the input data and methodology approach used in Ireland, some of which are outlined below.
 
@@
 This was the first comprehensive analysis of the residential sector, their studies examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure} and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2000DomesticFailure,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}.
-In 2000 Clinch and Healy developed and published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Mircsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating.
+In 2000, Clinch and Healy developed and published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Microsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating.
 A cost-benefit model (CBM) was fitted to the EAM to facilitate the conversion of the physical estimates into monetary amounts.
@@
 into the computer model.
-the modelling process involved developing a computer model of the energy performance of the existing dwelling stock.
-It was noted, energy consumption predicted by the model is considerably higher than given in the national energy balance, due to intertal temperature assumptions.
-The model showed the energy-efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compared to the reference scenarios.
-
-In 2011 Dineen and Ó Gallachóir \cite{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP).
-The model focused on energy efficiency of space and water heating by examining technology and building fabric efficiency.
+The modelling process involved developing a computer model of the energy performance of the existing dwelling stock.
+It was noted that energy consumption predicted by the model is considerably higher than given in the national energy balance, due to internal temperature assumptions.
+The model showed that the energy efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compared to the reference scenarios.
+
+In 2011, Dineen and Ó Gallachóir \cite{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP).
+The model focused on the energy efficiency of space and water heating by examining technology and building fabric efficiency.
 The building fabric efficiency of newly occupied dwellings developed takes into account the theoretical increase in energy efficiency due to improved 2008 and 2010 building regulations.
-Space heating demand accounted for 70\% of household energy consumption in newly built dwellings in 1997, down to just under 60\% in 2006 and the model is predicts a reduction to 27\% by 2015.The paper acknowledges retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined.
-Nor is the affect of internal temperatures on energy efficiency improvements.
-\par In 2013 Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures to the existing detached, oil centrally heated, rural housing stock.
+Space heating demand accounted for 70\% of household energy consumption in newly built dwellings in 1997, down to just under 60\% in 2006 and the model predicts a reduction to 27\% by 2015.The paper acknowledges retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperatures affect on energy efficiency improvements.
+\par In 2013, Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures to the existing detached, oil centrally heated, rural housing stock.
 A base geometry according to age bands and thermal characteristics was created in the study.
@@
 Concluding that the majority of energy efficiency progress for thermal use in Irish dwellings is being offset.
-The model thermal improvement measures resulted a reduction of running costs and CO2 emissions by an average of 65\% for houses constructed prior to 1979, this resulted in an average 12-year payback.
+The model thermal improvement measures resulted a reduction of running costs and CO2 emissions by an average of 65\% for houses constructed prior to 1979, and this resulted in an average 12-year payback.
 Houses built in 1979 or after resulted in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures.
 
-\par In 2015 Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examining new dwelling, to examining energy savings potential of energy efficiency improvement measures of the existing dwelling stock.
+\par In 2015, Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examining new dwellings, by examining energy savings potential of energy efficiency improvement measures of the existing dwelling stock.
 The Dwelling Energy Assessment Procedure (DEAP) model used, relied on the BER database, which accounted for 18\% of all Irish dwellings, currently the database has more than doubled and accounts for over 50\%.
@@
 The model used 175 archetype dwellings ( 5 building types, 7 energy bands and 5 wall construction types ) and 6 different retrofit measures were applied.
-BER database filters were used in this model, the BER filters from this study, were updated and applied in TIM.
-A sensitivity analysis is performed on the effects of internal temperature and rebound.
-It was found that there was approximately a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift.
+BER database filters were used in this model, the BER filters from this study were updated and applied in TIM.
+A sensitivity analysis is performed and it was found that there was approximately a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift.
 Concluding, BER standard assumptions overestimate dwelling energy consumption and that the average internal temperature across the Irish dwelling stock may be closer to 18$^{\circ}$C than 21$^{\circ}$C.
@@
 specifically examined the Better Energy Homes (BEH) scheme.
-The five most common retrofit combinations are simulated for each of 9 archetypes ( 3 energy rating types and 3 building types ).
-The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\% can be achieved when compared the baseline scenario.
+The five most common retrofit combinations are simulated for each of the 9 archetypes ( 3 energy rating types and 3 building types ).
+The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\% can be achieved when comparing the baseline scenario.
 This shows the the type of retrofit and type of house can have a huge affect on the energy saved.
-Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is considered negligible.
+Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F, and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is considered negligible.
 
@@
 
-The listed models are some of models used within SEAI's NEMF to model the impact of incentives and opportunities created by specific policy instruments.
+The listed models are some of the models used within SEAI's NEMF to model the impact of incentives and opportunities created by specific policy instruments.
 
 The models described in this section included a range of national ESOMs and bottom-up residential models.
-The models have varied objectives such as exploring consumer choices, PV battery systems location, evaluating health effects of retrofitting and integration of bio-energy and heat.
+The models have varied objectives such as exploring consumer choices, PV battery system location, evaluating health effects of retrofitting, and integration of bioenergy and heat.
 The thermal efficiency of dwellings is a common objective in Irish models, due to the poor thermal efficiency of the dwelling stock and the every growing BER database is providing high data granularity input to these models.
 The internal temperature is also analysed in some of the studies, a recurring theme is the wide range of internal temperatures.
-The bottom-up residential models are typically simulation models looking at specific policy.
-The complexity of the models have increased significantly in the last 20 years, in Ireland the models starting out in Microsoft Excel and switching to Microsoft SQL.
+The bottom-up residential models are typically simulation models looking at a specific policy.
+The complexity of the models has increased significantly in the last 20 years, in Ireland, the models starting out in Microsoft Excel and switching to Microsoft SQL.
 There has been few studies on the Irish residential sector using ESOMs, as the residential bottom-up models are preferred when focusing on that sector.
-The studies on Irish ESOM typically do not focus on the residential sector, instead they focus on the overall energy system, electricity generation or transport.
-The use of ESOM to explore DH in Ireland, allows for insights which the bottom-up simulation models cannot provide.
+The studies on Irish ESOM typically do not focus on the residential sector, instead they focus on the overall energy system, electricity generation, or transport.
+The use of ESOM to explore DH in Ireland, allows for insights which bottom-up simulation models cannot provide.
 The connection between waste heat in electricity and industry can show the optimal solution for district heating in commercial and residential sectors.
@@
 Ireland’s distinctive energy sectors are further separated from other EU member states when focusing on the residential sector.
-This is largely down to Ireland’s higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings and rural settlement patterns.
+This is largely down to Ireland’s higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings, and rural settlement patterns.
 This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}.
@@
 85\% of residential buildings are 2005 or older.
-Furthermore, Ireland's fossil fuel dependency is underpinned by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.
+Furthermore, Ireland's fossil fuel dependency is underpinned by home heating oil, accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.
 
@@
 While, gas availability is limited to some urban areas, and accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
-The residential sector relied on fossil fuels for over 90\% of thermal energy in 2018, consequently Ireland currently has the lowest renewable heat of among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
+The residential sector relied on fossil fuels for over 90\% of thermal energy in 2018, consequently Ireland currently has the lowest renewable heat among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
 Preliminary results from the Environmental Protection Agency (EPA) have indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.
@@
 The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}.
-These targets accounts for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out make sectoral five year carbon budgets  a legal requirement to aid progress of the forementioned long-term targets.
-\par The EU's first Nationally Determined Contribution (NDC) which had a 40\% GHG reduction target (first NDC), has been succeeded by a new target of at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}.
+These targets account for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to make sectoral five year carbon budgets a legal requirement to aid progress of the forementioned long-term targets.
+\par The EU's first Nationally Determined Contribution (NDC), which had a 40\% GHG reduction target (first NDC), has been succeeded by a new target of at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}.
 Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
@@
 The densification of Ireland's main cities, will provide opportunities to improve the efficiency of heating, electricity and transport within and between cities.
-\par Specifically focusing on the energy sector, the Marginal Abatement Cost Curve (MACC) produced in CAP2019  \cite{EPA2018EPAReport} found electricity \& transport are most cost effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
-CAP2019 findings show the most cost effective solution to decarbonising the residential sector, is to receive higher amounts of low carbon electricity.
+\par Specifically focusing on the energy sector, the Marginal Abatement Cost Curve (MACC) produced in CAP2019  \cite{EPA2018EPAReport} found electricity \& transport are the most cost effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
+CAP2019 findings show the most cost-effective solution to decarbonising the residential sector, is to receive higher amounts of low carbon electricity.
 
@@
 
-In Ireland the decarbonisation of the electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}.
+In Ireland, the decarbonisation of electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}.
 Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private investment towards low carbon or renewable energy.
@@
 The Renewable Electricity Support Scheme (RESS) replaced REFIT in 2020.
-RESS supports to renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target.
+RESS supports renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target.
 % Almost 800 MW of solar has been approved in RESS-1 \cite{Eirgrid2020RenewableResults}, this greatly exceeds the current solar capacity of 16 MW, while making significant progress towards the CAP2019 target of 1500 MW solar by 2030.
-\par Sustainable Energy Authority of Ireland (SEAI) have a key role in the rollout of retrofitting and heat pumps in existing homes.
-In parallel, new dwellings must achieve a minimum of 20\% renewable energy.
+\par Sustainable Energy Authority of Ireland (SEAI) has a key role in the rollout of retrofitting and heat pumps in existing homes.
+In parallel, new dwellings must achieve a minimum share of 20\% renewable energy.
 The ambient heat used in heat pumps is included as a renewable energy, heat pumps are therefore the preferred heating technology despite the high capital costs.
@@
 While Irish District Energy Association (IrDEA) suggest 57\% of heat can be sourced from DH, if supporting policy and regulation was in place.
-District Heating will likely play a role, however Gartland points out a range of organisational, technical, regulatory and economic barriers to DH growth in Ireland \cite{Gartland2016AIreland}.
+District Heating will likely play a role, however Gartland points out a range of organisational, technical, regulatory, and economic barriers to DH growth in Ireland \cite{Gartland2016AIreland}.
 \par While it maybe optimal to use Hydrotreated Vegetable Oil (HVO) in transport, it is also considered in the residential sector, as it can provide a cheap alternative for the 700,000 Irish dwellings using home heating oil ( mostly kerosene).
@@
 The capacity of HVO is growing in Europe, but HVO is currently about twice the price of kerosene.
-HVO can be used to replace kerosene, albeit with cost of up to €400 per oil boiler, however there is no policy for HVO in the residential sector.
+HVO can be used to replace kerosene, albeit at a cost of up to €400 per oil boiler, however there is no policy for HVO in the residential sector.
 
@@
 \par  The expertise among energy policy researchers in Ireland combined with the global network of Energy Technology Systems Analysis Program (ETSAP) modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways.
-Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model, and then updated with specific national data and assumptions.
+Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model and then updated with specific national data and assumptions.
 The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.
@@
 
-\par The function of partial equilibrium operates whereby prices and quantities in each time period are such that the suppliers produce exactly the quantities demanded by the consumers and therefore the total economic surplus  (sum of the suppliers’ and consumers’ surpluses) is maximized \cite{Loulou2016DocumentationI}.
+\par The function of partial equilibrium operates whereby the prices and quantities in each time period are such that the suppliers produce exactly the quantities demanded by the consumers and therefore the total economic surplus  (sum of the suppliers’ and consumers’ surpluses) is maximized \cite{Loulou2016DocumentationI}.
 The TIMES objective is therefore to minimize the total cost of the system, by solving for the following:
@@
 The exception is ambient heat, which could potentially provide insights to residential decarbonisation by separating it into ambient heat for space heating and ambient heat for water heating.
-The solid multi-fuel from the BER database represents an open fire or stove which is heated by either coal, peat or wood.
-Through analysis of the SEAI Energy Balance and BER database, the fuel share assigned for coal, peat and wood is 19.4 \%, 74.4 \% and 6.2 \% respectively.
+The solid multi-fuel from the BER database represents an open fire or stove which is heated by either coal, peat, or wood.
+Through examining the SEAI Energy Balance and BER database, the fuel share assigned for coal, peat, and wood is 19.4 \%, 74.4 \% and 6.2 \% respectively.
 There are two types of energy services represented in Table \ref{fig:RES} - Non-Archetype and Archetype.
 Non-archetype energy services fill in the lack of data on archetype energy services.
-The six non-archetype energy services, are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service, this bottom-up data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA (Policy Oriented Tool for Energy and Climate Change Impact Assessment) database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{LitR:Lighting,Appliances}.
-The four archetype energy services are dependent on  building archetype and data is sourced from the BER database, which is discussed in Section \ref{LitR:StockData}.
+The six non-archetype energy services are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service, this bottom-up data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA (Policy Oriented Tool for Energy and Climate Change Impact Assessment) database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{LitR:Lighting,Appliances}.
+The four archetype energy services are dependent on  building archetypes and data is sourced from the BER database, which is discussed in Section \ref{LitR:StockData}.
 The internal temperature adjustment for space heating is detailed in Section \ref{LitR:StockData}.
@@
 
-Where $TIM_{r,y,a}$ is the total dwellings in TIM by BER rating (r), year of construction (y) and archerype (a).
+Where $TIM_{r,y,a}$ is the total dwellings in TIM by BER rating (r), year of construction (y), and archetype (a).
 $CSO_{y,a}$ is the number of dwellings by year of construction and archetype in the CSO dataset, the CSO dataset does not contain BER rating types.
@@
 To provide more robust costs, costs data is also used from the Typology Approach for Building Stock Energy Assessment (TABULA) report \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} which has more detailed cost data on 2 stages of retrofit for a small sample of 31 dwellings.
-The formula used to find the average cost of retrofit per type of retrofit and archetype, are shown in Equations \ref{Eq:RetrofitCost1} and \ref{Eq:RetrofitCost2}.
+The formulas used to find the average cost of a retrofit per type of retrofit and archetype are shown in Equations \ref{Eq:RetrofitCost1} and \ref{Eq:RetrofitCost2}.
 
@@
 \cite{Ali2020ABuildings}, it contains all ratings and types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments and 1.1 for detached dwellings.
-The $TABULA_{r,a,t}$ is the average cost of retrofit from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} per rating, archetype and type of retrofit, however with only 62 samples in total not all retrofits are included.
-In this case TABULA is ignored and Ali et al.
+The $TABULA_{r,a,t}$ is the average cost of retrofit from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} per rating, archetype and type of retrofit, however with only 62 samples in total, not all retrofits are included.
+In this case, TABULA is ignored and Ali et al.
 is the only source of retrofit costs.
+\par To simplify the model, the cost of retrofit per BER rating is not used in TIM. Instead an average weighted cost per archetype and per retrofit is calculated.
 
@@
 
-The model uses the weighted average cost of retrofit per archetype
-
-in the residential and tertiary sector in the EU 
+${TIM_{r,a}}$ is the number of dwellings by rating and archetype and ${TIM_{a}}$ is the total number of dwellings of that archetype, however A-rated dwellings are excluded from the total, as it is assumed they cannot be retrofitted further.
+
+THe cost data from appliances 
 Topten -
@@
 Retrofit Savings - SEAI dataset not published..papers Usman,etc.
-While also accounting for more fuels and technological options, the residential sector in TIM is structured with around retrofitting, not vice-versa as previous versions, so that it is more aligned to ongoing climate mitigation investments.
-
-\subsubsection{Energy Service Demands Projections}
+While also accounting for more fuels and technological options, the residential sector in TIM is structured around retrofitting, not vice versa as previous versions, so that it is more aligned to ongoing climate mitigation investments.
+
+\subsubsection{Energy Service Demand Projections}
 \label{LitR:Demands}
@@
 Macroeconomic 
-no elasticity - fuel prices and weather compensation?
+no elasticity - fuel prices, and weather compensation?
 Time slices - 40 etc. 
@@
 This results in an average of 27,600 new houses per annum between between 2021-2040.
-Beyond 2040, population is used as a driver to project housing stock.
-The total housing stock obtained in 2050 is 2.57 million which implies 8\% increase from 2040.
+Beyond 2040, the population is used as a driver to project housing stock.
+The total housing stock obtained in 2050 is 2.57 million, which implies 8\% increase from 2040.
 % Table generated by Excel2LaTeX from sheet 'Sheet1'
@@
 Space heating is the largest residential energy service demand, accounted for 61.8\% of final energy usage in 2018.
-Other residential energy services and their share of final energy is shown in Fig.\ref{fig:Ireland2018ServiceDemands}.
+Other residential energy services and their share of final energy are shown in Fig.\ref{fig:Ireland2018ServiceDemands}.
 As building thermal efficiency and heating technology efficiency improves, both space and water heating energy would be expected to decline.
@@
 From the BER database, it can be seen that A-rated dwellings use less than half space heating energy of the average dwelling and A-rated water heating energy decreases by about 40\% compared to the average A-rated dwelling.
-However Pumps \& Fans energy increases as mechanical ventilation is used over natural ventilation and Lighting usage in A-rated is more than double the average household.
-Cooking and Appliances energy usage would be expected to remain stable.
+However Pumps \& Fans energy increases as mechanical ventilation is used over natural ventilation and lighting usage in A-rated is more than double the average household.
+Cooking and appliance energy usage would be expected to remain stable.
 
@@
 
-The SEAI residential energy report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} provides high resolution detail on energy demands in the residential sector.
+The SEAI Residential energy report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} provides high resolution detail on energy demands in the residential sector.
 Variables such as disposable income, appliance ownership, dwelling location, type of occupier status (owner/rented ), BER labels and age of dwelling.
@@
 %Space Heating demand - BER space heating is only within certain months and assumes 21C heating..
-Thermostats are not widely used for temperature regulation in private dwellings in Ireland; instead the domestic heating schedule was assumed to be 07.00–09.00 h and 17.00–23.00 h for a heating season of October to May inclusive, based on guidelines issued by the Sustainable Energy Authority of Ireland (SEAI) [32].
+Thermostats are not widely used for temperature regulation in private dwellings in Ireland; instead, the domestic heating schedule was assumed to be 07.00–09.00 h and 17.00–23.00 h for a heating season of October to May inclusive, based on guidelines issued by the Sustainable Energy Authority of Ireland (SEAI) [32].
 The thermal load consists of both space heating and a baseline demand for sustaining water heating.
-For non-domestic premises the thermal load is defined by office hours, in line with the electrical demand.
+For non-domestic premises, the thermal load is defined by office hours, in line with the electrical demand.
 The domestic thermal loads considered in this study are listed in Table 2; the thermal load for the non-domestic
@@
 An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}.
-Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty.
+Fuel Allowance and Household Benefits provide support to low-income households to alleviate fuel poverty.
 These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}.
 Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}.
-Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
+Ironically, it can be argued this places low-income households deeply dependent on fossil fuels for health.
 \par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
@@
  Build national fuel
-The CSO data, is mainly data from the 2016 Cenus. 
+The CSO data, which is mainly data from the 2016 Cenus. 
 Fuel Calibration
@@
 
-Model calibration: archetype demand The
-The fuel is calibrated to the before calibration the fuel was 50\% too high
+Model calibration: archetypes demand The
 
@@
 Capacity, Efficiency, Constraints, Uptake, adapation
-heat output (Qt) and electricity input in a given time per-iod (Pt) is expressed asQt= COPPt, where COP is the average an-nual coefficient of performance for the heat pump
+heat output (Qt) and electricity input in a given time period (Pt) is expressed asQt= COPPt, where COP is the average annual coefficient of performance for the heat pump
 \subsubsection{Lighting, Appliances and Other}
@@
 \end{equation}
-The GROWTH coefficient is defined as a new attribute of the technology, and represents the maximum annual growth allowed for the capacity.
+The GROWTH coefficient is defined as a new attribute of the technology and represents the maximum annual growth allowed for the capacity.
 The quantity M(t+1)-M(t) is the number of years between the milestones of periods t and t+1.
@@
 TIM was built to help Ireland provide a pathway to achieve 2030 and 2050 targets.
-The residential sector plays a key part of this and it was built from the SEAI 2018 energy balance, the BER database and CSO data.
-The data filtration process used in the residential sector, derived a lot of conversation of the BER database, a group know as Ireland Energy Wiki by the National Retrofitting Modelling Group grew from building the residential sector, questions about the filters, default values, assumptions around thermal equations (EC..) and uncertainties were used to improve the data before it was used in TIM .
+The residential sector plays a key part of this and it was built from the SEAI 2018 energy balance, the BER database, and CSO data.
+The data filtration process used in the residential sector derived a lot of conversation from the BER database, a group know as Ireland Energy Wiki by the National Retrofitting Modelling Group grew from building the residential sector, questions about the filters, default values, assumptions around thermal equations (EC..) and uncertainties were used to improve the data before it was used in TIM .
 TIM has many internal project deadines, which required extensive meetings about the data, constraints and results as the modelling was progressing.
-The data and results were presented to 30 external reviewers to provide feedback, this process proved invaluable having different expertise with different presepectives enabled TIM to fill in gaps.
-TIM will also be open platform to help improve the model and provide transparency.
-TIM’s residential sector was built in parallel with LEAP’s residential sector, so that residential sector can easily simulate policy.
-The demand for results from TIM has increased significantly with increased government ambition, TIM will provide input into CAP 2021 and Ireland’s carbon budgets, working with DECC and the CCAC respectively will also provide feedback and improve the robustness of TIM.
-for these reasons TIM residential sector is a state of the art modelling platform.
+The data and results were presented to 30 external reviewers to provide feedback, this process proved invaluable having different expertise with different perspectives enabling TIM to fill in gaps.
+TIM will also be an open platform to help improve the model and provide transparency.
+TIM’s residential sector was built in parallel with LEAP’s residential sector, so that the residential sector can easily simulate policy.
+The demand for results from TIM has increased significantly with increasing government ambitions, TIM will provide input into CAP 2021 and Ireland’s carbon budgets, working with DECC and the CCAC respectively, will also provide feedback and improve the robustness of TIM.
+For these reasons, the TIM residential sector is at the cutting edge of modelling platforms.
 
@@
 \label{Discussion}
-Soft-linking to LEAP and PLEXOS
+Soft-linking with LEAP and PLEXOS
 %o	Discuss and interpret the results
@@
 
-First iteration - of model will need further development DH, BEhavour, higher granularity and spatial will be easily done with dataset. 
+First iteration - of model will need further development, DH, BEhavour, higher granularity and spatial will be easily done with the dataset. 
 %What does new model do that previous didn’t?
@@
 %Improved more detailed Energy Service Demands
-Detailed bottom up model - using 2018 base year ( in line with policy )
+Detailed bottom-up model - using the 2018 base year ( in line with policy )
 Updated technology
@@
 \label{Conclusion}
-This paper reports results on ambitious mitigation target in the period to 2050 for the Irish energy system. The analysis have been performed using the Irish TIMES model
+This paper reports results on an ambitious mitigation target in the period to 2050 for the Irish energy system. The analysis has been performed using the Irish TIMES model
 \textit{We stress the importance of differentiating between technical con-

### 2021-07-06 manuscript.tex
6  July 2021

@@
 
-Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways, and interpreting the results, so Section \ref{sec:Introduction} of this paper looks at Ireland's climate targets, energy policy and distinctive residential sector.
+Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways and interpreting the results, so Section \ref{sec:Introduction} of this paper looks at Ireland's climate targets, energy policy, and distinctive residential sector.
 While also examining how researchers have previously used Energy System Optimization Models (ESOMs) to analyse the residential sector.
@@
 \par For example, Li et al.
-\cite{Li2018IncorporatingModel} investigated the heating technology preference in UK, using UK TIMES Model (UKTM).
+\cite{Li2018IncorporatingModel} investigated the heating technology preference in UK using UK TIMES Model (UKTM).
 As TIMES \cite{Loulou2016DocumentationI} assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
@@
 The DCM then estimated the probability of heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices.
-While Li's researcher has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights, through more surveys and analysis.
+While Li's researcher has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights through more surveys and analysis.
 
@@
 
-Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised  Balmorel model \cite{Wiese2018BalmorelModel} to handle the integrated power and district heating system, to analyse how heat pumps with complementing heat storage affect wind power investments in Denmark.
-The model was disaggreated between a western and eastern Denmark region and the model covers Norway, Sweden, Finland, and Germany, in order to cover import/export.
+Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised  Balmorel model \cite{Wiese2018BalmorelModel} to handle the integrated power and district heating system to analyse how heat pumps with complementing heat storage affect wind power investments in Denmark.
+The model was disaggreated between a western and eastern Denmark region and the model covers Norway, Sweden, Finland, and Germany to cover import/export.
 Exogenous user constraints applied to the model, based on the current Danish policy includes: Wind power must provide at least 50\% of the national electricity consumption by 2020, coal is phased out in heat and power plants by 2030, and individual oil boilers are phased out by 2030.
-The results find when investments in individual heat pumps are allowed, air-to-water heat pumps are installed in non-district heating areas.
+The results find that when investments in individual heat pumps are allowed, air-to-water heat pumps are installed in non-district heating areas.
 This represents a significant electricity demand in identifying individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments and reducing system costs, fuel consumption, and $CO_2$ emissions.
@@
 Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model \cite{stergaard2015ReviewingSimulations} to explore the feasibility of photovoltaic (PV) systems installed at the household level in combination with batteries in Samsø, which is known for being Denmark’s renewable energy island.
-The two approaches are observed for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
+Two approaches are observed for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
 Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy system perspective, one reason for this is the wind power could also be stored in communal batteries.
@@
 In 2014, Rogan et al.
-\cite{Rogan2014LEAPsSystem} linked OSeMOYOS, an ESOM to the Long Range Energy Alternatives Planning System (LEAP) simulation model, to explore specific energy efficiency policies not only in the residential sector, but also in transport, industry and services sectors in Ireland.
+\cite{Rogan2014LEAPsSystem} linked OSeMOYOS, an ESOM, to the Long Range Energy Alternatives Planning System (LEAP) simulation model, to explore specific energy efficiency policies not only in the residential sector, but also in the transport, industry and services sectors in Ireland.
 The bottom-up development of the residential sector in LEAP was combined with the cost optimisation capabilities of OSeMOYOS.
-Three scenarios are explored, namely, the reference scenario, energy efficiency scenario and energy efficiency+ scenario.
-The reference assume that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020.
-In the reference scenario the residential total energy consumption rose 6.7 \% in 2020, compared to 2008.
+Three scenarios are explored, namely, the reference scenario, energy efficiency scenario, and energy efficiency+ scenario.
+The reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020.
+In the reference scenario, the residential total energy consumption rose 6.7 \% in 2020, compared to 2008.
 The share of space and water heating decreased from 82 \% to 77 \% in that period.
@@
 The previous version of Irish TIMES did not account for district heating, so EnergyPLAN is used to investigate the operation and  feasibility of district heating in Ireland.
-A sensitivity analysis was preformed on investment costs, heat grid losses, and discount rates, however in all cases the inclusion of district heating is optimal.
+A sensitivity analysis was performed on investment costs, heat grid losses, and discount rates, however in all cases the inclusion of district heating is optimal.
 The final results find that district heating reduces fuel consumption by 3.5 \% and the conversion of 37 \% of the Irish heat demand to district heating is found to be optimal.
@@
 Clinch and Healy analysed different aspects of thermal comfort in the Irish residential sector.
-This was the first comprehensive analysis of the residential sector, their studies examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure} and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2000DomesticFailure,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}.
+This was the first comprehensive analysis of the residential sector, and their studies examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure} and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2000DomesticFailure,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}.
 In 2000, Clinch and Healy developed and published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Microsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating.
 A cost-benefit model (CBM) was fitted to the EAM to facilitate the conversion of the physical estimates into monetary amounts.
-Under the predicted scenario, energy savings easily outweighing programme costs, representing a benefit-cost ratio of 1.7.The benefits of reduced sickness, hospitalisation and drugs and mortality, were quantified, this moved the  benefit-cost ratio up to 3.0.
-The same model was used by Clinch, Healy and King \cite{Clinch2001ModellingEfficiency} in addition three models of householder spending on energy are developed here to to evaluate the human behavioural component
+Under the predicted scenario, energy savings easily outweighed programme costs, representing a benefit-cost ratio of 1.7.The benefits of reduced sickness, hospitalisation, drugs, and mortality were quantified and this moved the  benefit-cost ratio up to 3.0.
+The same model was used by Clinch, Healy, and King \cite{Clinch2001ModellingEfficiency} in addition three models of householder spending on energy are developed here to to evaluate the human behavioural component
 into the computer model.
 The modelling process involved developing a computer model of the energy performance of the existing dwelling stock.
-It was noted that energy consumption predicted by the model is considerably higher than given in the national energy balance, due to internal temperature assumptions.
+It was noted that the energy consumption predicted by the model is considerably higher than given in the national energy balance, due to internal temperature assumptions.
 The model showed that the energy efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compared to the reference scenarios.
@@
 Concluding that the majority of energy efficiency progress for thermal use in Irish dwellings is being offset.
-The model thermal improvement measures resulted a reduction of running costs and CO2 emissions by an average of 65\% for houses constructed prior to 1979, and this resulted in an average 12-year payback.
+The model thermal improvement measures resulted in a reduction of running costs and CO2 emissions by an average of 65\% for houses constructed prior to 1979, and this resulted in an average 12-year payback.
 Houses built in 1979 or after resulted in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures.
 
-\par In 2015, Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examining new dwellings, by examining energy savings potential of energy efficiency improvement measures of the existing dwelling stock.
-The Dwelling Energy Assessment Procedure (DEAP) model used, relied on the BER database, which accounted for 18\% of all Irish dwellings, currently the database has more than doubled and accounts for over 50\%.
-DEAP is the national tool for BER calculation and checking building regulation compliance.
-The model used 175 archetype dwellings ( 5 building types, 7 energy bands and 5 wall construction types ) and 6 different retrofit measures were applied.
+\par In 2015, Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examining new dwellings, by examining the energy savings potential of energy efficiency improvement measures of the existing dwelling stock.
+The Dwelling Energy Assessment Procedure (DEAP) model was used, this is the national tool for BER calculation and checking building regulation compliance.
+The DEAP model used 175 archetype dwellings ( 5 building types, 7 energy bands, and 5 wall construction types ) and 6 different retrofit measures were applied.
 BER database filters were used in this model, the BER filters from this study were updated and applied in TIM.
@@
 The Archetype Dwelling Energy Model (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}.
-ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and then calculates annual energy consumption to simulate the potential for improved energy efficiency gains within the existing retrofit programs.
+ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and then calculates the annual energy consumption to simulate the potential for improved energy efficiency gains within the existing retrofit programs.
 Mac Uidhir et al.
@@
 The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\% can be achieved when comparing the baseline scenario.
-This shows the the type of retrofit and type of house can have a huge affect on the energy saved.
+This shows the the type of retrofit and type of house can have a large affect on the energy saved.
 Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F, and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is considered negligible.
@@
 
-The models described in this section included a range of national ESOMs and bottom-up residential models.
-The models have varied objectives such as exploring consumer choices, PV battery system location, evaluating health effects of retrofitting, and integration of bioenergy and heat.
-The thermal efficiency of dwellings is a common objective in Irish models, due to the poor thermal efficiency of the dwelling stock and the every growing BER database is providing high data granularity input to these models.
+The models described in this section include a range of national ESOMs and bottom-up residential models.
+The models have varied objectives such as exploring consumer choices, PV battery system location, evaluating the health effects of retrofitting, and integration of bioenergy and heat.
+The thermal efficiency of dwellings is a common objective in Irish models, due to the poor thermal efficiency of the dwelling stock and the growing BER database is providing high data granularity input to these models.
 The internal temperature is also analysed in some of the studies, a recurring theme is the wide range of internal temperatures.
 The bottom-up residential models are typically simulation models looking at a specific policy.
-The complexity of the models has increased significantly in the last 20 years, in Ireland, the models starting out in Microsoft Excel and switching to Microsoft SQL.
-There has been few studies on the Irish residential sector using ESOMs, as the residential bottom-up models are preferred when focusing on that sector.
+The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switching to Microsoft SQL.
+There has been few studies on the Irish residential sector using ESOMs, as residential bottom-up models are preferred when focusing on that sector.
 The studies on Irish ESOM typically do not focus on the residential sector, instead they focus on the overall energy system, electricity generation, or transport.
-The use of ESOM to explore DH in Ireland, allows for insights which bottom-up simulation models cannot provide.
+The use of ESOM to explore DH in Ireland allows for insights which bottom-up simulation models cannot provide.
 The connection between waste heat in electricity and industry can show the optimal solution for district heating in commercial and residential sectors.
@@
 GHG emissions from coal in the residential sector also declined to 6.9\% in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
-However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
-While, gas availability is limited to some urban areas, and accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+However, the decline in peat and coal has been offset by a more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+While, gas availability is limited to some urban areas and accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
 The residential sector relied on fossil fuels for over 90\% of thermal energy in 2018, consequently Ireland currently has the lowest renewable heat among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
@@
 Under 2006 Intergovernmental Panel on Climate Change (IPCC) guidelines, Ireland’s residential sector accounted for 14.2\% of GHGs emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
-Focusing on energy-related GHG emissions only, the residential sector directly accounts for 24\% of GHG emissions.
+Focusing only on energy-related GHG emissions, the residential sector directly accounts for 24\% of GHG emissions.
 Fig.
@@
 The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}.
-These targets account for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to make sectoral five year carbon budgets a legal requirement to aid progress of the forementioned long-term targets.
+These targets account for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to make sectoral five-year carbon budgets a legal requirement to aid progress of the forementioned long-term targets.
 \par The EU's first Nationally Determined Contribution (NDC), which had a 40\% GHG reduction target (first NDC), has been succeeded by a new target of at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}.
@@
 Climate Action Plan 2021 will have to be more ambitious than its predecessor, CAP2019 as it must reflect EU's updated NDC target and take account of the national target also.
-There are also renewable energy targets to consider for transport (RES-T), heat (RES-H) and electricity (RES-E) under renewable directive (EU) 2018/2001, and energy efficiency targets under (EU) 2018/2002.
+There are also renewable energy targets to consider for transport (RES-T), heat (RES-H), and electricity (RES-E) under the renewable directive (EU) 2018/2001, and the energy efficiency targets under (EU) 2018/2002.
 
@@
 The densification of Ireland's main cities, will provide opportunities to improve the efficiency of heating, electricity and transport within and between cities.
-\par Specifically focusing on the energy sector, the Marginal Abatement Cost Curve (MACC) produced in CAP2019  \cite{EPA2018EPAReport} found electricity \& transport are the most cost effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
+\par Specifically focusing on the energy sector, the Marginal Abatement Cost Curve (MACC) produced in CAP2019  \cite{EPA2018EPAReport} found electricity \& transport are the most cost-effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
 CAP2019 findings show the most cost-effective solution to decarbonising the residential sector, is to receive higher amounts of low carbon electricity.
@@
 The ambient heat used in heat pumps is included as a renewable energy, heat pumps are therefore the preferred heating technology despite the high capital costs.
-As mentioned in Section \ref{Intro:IreRSD}, new buildings in Ireland produce 70\% less GHG emissions than buildings in 2005, along with CAP2019 plans to retrofit 500,000 existing homes to at least a  ‘B2’ energy rating by 2030, highlights the energy efficiency first approach in the residential sector.
+As mentioned in Section \ref{Intro:IreRSD}, new buildings in Ireland produce 70\% less GHG emissions than buildings in 2005, along with CAP2019 plans to retrofit 500,000 existing homes to at least a  ‘B2’ energy rating by 2030, which highlights the energy efficiency first approach in the residential sector.
 \par Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019VisionIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
@@
 \par TIMES is a bottom-up optimisation energy-environment model with perfect foresight which is used to analyse various levels of spatial, temporal and sectoral resolution.
-Exogenous technical, physical, and regulatory constraints can be applied and technologies have specified fuel types, efficiencies, lifetimes, availability, environmental characteristics and fixed, variable and operation costs.
+Exogenous technical, physical, and regulatory constraints can be applied and technologies have specified fuel types, efficiencies, lifetimes, availability, environmental characteristics, and fixed, variable and operation costs.
 TIMES uses a linear programming optimizer matrix in General Algebraic Modeling System (GAMS), to compute all calculations.
@@
 TIM is an optimisation model of the Irish energy system, which calculates the cost-optimal fuel and technology mix to meet future energy service demands.
-The residential, power generation and transport sectors are rebuild with high granularity from the bottom-up, while industry, agriculture and services sectors have been updated.
-The model operates within the limits of GHG constraints, resources constraints, and feasible deployment rates.
+The residential, power generation and transport sectors are rebuilt with high granularity from the bottom-up, while industry, agriculture, and services sectors have been updated.
+The model operates within the limits of GHG constraints, resource constraints, and feasible deployment rates.
 
@@
 In TIM, fuels such as coal, represent a combination of coal fuels from the SEAI Energy Balance and BER database.
-Kerosene accounts for over 70 \% of home heating oil usage in 2018, while Gasoil accounts for most of the remaining share, the fuels have similar properties and there was no evidence to support keeping separate fuels would provide further insights to residential decarbonisation.
+Kerosene accounts for over 70 \% of home heating oil usage in 2018, while Gasoil accounts for most of the remaining share, the fuels have similar properties, and there was no evidence to support keeping separate fuels would provide further insight into residential decarbonisation.
 The exception is ambient heat, which could potentially provide insights to residential decarbonisation by separating it into ambient heat for space heating and ambient heat for water heating.
@@
 The DEAP model, which was outlined by Dineen et al.\cite{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM} provides the results for the energy rating in the BER database.
-However, the space heating calculation used to estimate the expected energy consumption per dwelling, which makes some assumptions.
+However, the space heating calculation is used to estimate the expected energy consumption per dwelling, which makes some assumptions.
 DEAP assumes that each dwelling is single zoned, the heating season is frin October to May inclusive, and the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season.
 One assumption, the internal temperature assumption results in a large error, where space and water heating are overestimated in DEAP.
-Space heating consumption reduced by 26 \% after internal temperature adjustment were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance.
+Space heating consumption reduced by 26 \% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance.
 DEAP assumes all living areas ( living room, sitting room, and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C.
-Through analysis of previous residential modelling in Ireland this is not the case, as the internal temperatures are often much lower, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions applied in TIM.
+Through the analysis of previous residential modelling in Ireland this is not the case, as the internal temperatures are often much lower, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions applied in TIM.
 
@@
 
-The ArDEM-SQL, which was described in Section \ref{Intro:ESOMIreland} is a simulation model built on IS EN 13790.
-ArDEM-SQL alters primary energy consumption from the BER database by changing the internal temperatures.
-
-Other concerns for uncertainties in the database, which are not related to DEAP calculation are statistically anomalous spikes in the data, caused by applying default values \cite{Ahern2020ADatabases} and different companies generating variations of ±20\% in the assessment of energy consumption \cite{Harsman2016OnCertificates}.While the BER databases provides high granularity data, it must be filtered to improve the representation of the dwelling stock data.
+The ArDEM-SQL, which was described in Section \ref{Intro:ESOMIreland} is a simulation model built on EN 13790.
+ArDEM-SQL alters the primary energy consumption of the BER database by changing the internal temperatures.
+
+Other concerns for uncertainties in the database, which are not related to DEAP calculation are statistically anomalous spikes in the data, caused by applying default values \cite{Ahern2020ADatabases} and different companies generating variations of ±20\% in the assessment of energy consumption \cite{Harsman2016OnCertificates}.While the BER database provides high granularity data, it must be filtered to improve the representation of the dwelling stock data.
 The filters applied to the raw BER database are outlined in \ref{sec:appendixA}.
@@
 The total number of building archetypes per county in 2016 is used to scale up the BER database.
-The gap of total dwelling between the Cenus year in 2016 and the model base year in 2018 was initially ignored due to lack of data.
-However, later it was found that data on new dwelling electrical connections per archetype and per county are available from CSO dataset NDQ06 this would have filled the gap, the first iteration of the dwelling stock has a 2 \% error.
-
-Once both database were finalised, the next step in this process, was to scale up the BER database.
+The gap in total dwellings between the census year in 2016 and the model base year in 2018 was initially ignored due to lack of data.
+However, later it was found that data on new dwelling electrical connections per archetype and per county are available from CSO dataset NDQ06, this would have filled the gap, the first iteration of the dwelling stock has a 2 \% error.
+
+Once both databases were finalised, the next step in this process was to scale up the BER database.
 
@@
 
-Once the final dwelling stock is complete, the next step was to assign a BER rating to all dwellings.
+Once the final dwelling stock is complete, the next step is to assign a BER rating to all dwellings.
 This was done by comparing 9 different dwelling age ranges for each of the 3 archetypes between the BER and CSO datasets.
@@
 This section only covers the costs applied in the residential sector, which includes space and water heating technologies, retrofitting, cooking, lighting, pumps, fans, space cooling, and other electrical appliances.
-The fuel costs are not applied in residential sector, there are applied in the supply sector, and the data is sourced from World Energy Outlook reports from 2019 and 2020. 
+The fuel costs are not applied in the residential sector, there are applied in the supply sector, and the data is sourced from World Energy Outlook reports from 2019 and 2020. 
 The majority of the space and water heating cost data is obtained from the
@@
 The cost data used in TIM was based on the north EU region (Denmark) and where data was missing, the central EU region (Germany) cost data was used.
-The high granularity of data, allows for cost of a technology to be disaggregated between apartment and non-apartment, costs and efficiencies are projected every five years from 2020 to 2050.
+The high granularity of data allows for the cost of a technology to be disaggregated between apartments and non-apartments, costs and efficiencies are projected every five years from 2020 to 2050.
 The lifetime and emissions/pollutants are also included, while the cost data is further disaggregated between capital expenditure ( equipment, installation \& additional capital costs ) and Operational \& Maintenance (O\&M) costs.
@@
 al \cite{Ali2020ABuildings} who derived costs from SEAI's Better Energy Homes (BEH) and Better Energy Warmer Homes (BEWH) datasets, which are Ireland's largest retrofitting dataset ( over 285,000 retrofits ).
-To provide more robust costs, costs data is also used from the Typology Approach for Building Stock Energy Assessment (TABULA) report \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} which has more detailed cost data on 2 stages of retrofit for a small sample of 31 dwellings.
+To provide more robust retrofit costs, data is also used from the Typology Approach for Building Stock Energy Assessment (TABULA) report \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} which has more detailed cost data on 2 stages of retrofit for a small sample of 31 dwellings.
 The formulas used to find the average cost of a retrofit per type of retrofit and archetype are shown in Equations \ref{Eq:RetrofitCost1} and \ref{Eq:RetrofitCost2}.
@@
 \cite{Ali2020ABuildings}, it contains all ratings and types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments and 1.1 for detached dwellings.
-The $TABULA_{r,a,t}$ is the average cost of retrofit from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} per rating, archetype and type of retrofit, however with only 62 samples in total, not all retrofits are included.
+The $TABULA_{r,a,t}$ is the average cost of a retrofit from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} per rating, archetype and type of retrofit, however, with only 62 samples in total, not all retrofits are included.
 In this case, TABULA is ignored and Ali et al.
 is the only source of retrofit costs.
-\par To simplify the model, the cost of retrofit per BER rating is not used in TIM. Instead an average weighted cost per archetype and per retrofit is calculated.
+\par To simplify the model, the cost of retrofit per BER rating is not used in TIM. Instead, an average weighted cost per archetype and per retrofit is calculated.
 
@@
 
-THe cost data from appliances 
+THe cost data of appliances 
 Topten -
@@
 Retrofit Savings - SEAI dataset not published..papers Usman,etc.
-While also accounting for more fuels and technological options, the residential sector in TIM is structured around retrofitting, not vice versa as previous versions, so that it is more aligned to ongoing climate mitigation investments.
+While also accounting for more fuels and technological options, the residential sector in TIM is structured around retrofitting, not vice versa as previous versions, so that it is more aligned with ongoing climate mitigation investments.
 
@@
 Thus, uptake of technologies such as retrofitting will be driven, inter alia, by the net price for consumers.
-ENERGY CONSUMPTION CHARACTERISTICS of Houses - Table?
+ENERGY CONSUMPTION CHARACTERISTICS of House - Table?
 % The technical bottom-up development approach of the residential sector is combined with other economic sectors to meet macroeconomic defined demands, to provide GHG targets insights both the energy system and residential sector.
@@
 
-The residential stock projections upto 2040 are taken from ESRI's housing demand estimates \citep{Bergin2020}.
+The residential stock projections up to 2040 are taken from ESRI's housing demand estimates \citep{Bergin2020}.
 The stock is expected to increase by 40\% from 2018 level with a CAGR of 2\%.
@@
 The new A-rated dwellings can be used as a guide to show the share of future energy service demands.
-From the BER database, it can be seen that A-rated dwellings use less than half space heating energy of the average dwelling and A-rated water heating energy decreases by about 40\% compared to the average A-rated dwelling.
+From the BER database, it can be seen that A-rated dwellings use less than half the space heating energy of the average dwelling and A-rated water heating energy decreases by about 40\% compared to the average A-rated dwelling.
 However Pumps \& Fans energy increases as mechanical ventilation is used over natural ventilation and lighting usage in A-rated is more than double the average household.
@@
 
-The SEAI Residential energy report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} provides high resolution detail on energy demands in the residential sector.
-Variables such as disposable income, appliance ownership, dwelling location, type of occupier status (owner/rented ), BER labels and age of dwelling.
+The SEAI Residential Energy report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} provides high resolution details on energy demands in the residential sector.
+Variables such as disposable income, appliance ownership, dwelling location, type of occupier status (owner/rented ), BER labels, and age of dwelling.
 
 %Space Heating demand - BER space heating is only within certain months and assumes 21C heating..
-Thermostats are not widely used for temperature regulation in private dwellings in Ireland; instead, the domestic heating schedule was assumed to be 07.00–09.00 h and 17.00–23.00 h for a heating season of October to May inclusive, based on guidelines issued by the Sustainable Energy Authority of Ireland (SEAI) [32].
+Thermostats are not widely used for temperature regulation in private dwellings in Ireland; instead, the domestic heating schedule was assumed to be 07.00–09.00 h and 17.00–23.00 h for the heating season of October to May inclusive, based on the guidelines issued by the Sustainable Energy Authority of Ireland (SEAI) [32].
 The thermal load consists of both space heating and a baseline demand for sustaining water heating.
-For non-domestic premises, the thermal load is defined by office hours, in line with the electrical demand.
+For non-domestic premises, the thermal load is defined by office hours in line with the electrical demand.
 The domestic thermal loads considered in this study are listed in Table 2; the thermal load for the non-domestic
@@
 Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which is a reason for Ireland's high excessive winter mortality \cite{Clinch2000HousingMortality}.
-Fuel Poverty which which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths.
+Fuel Poverty which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths.
 An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}.
@@
 Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}.
-Ironically, it can be argued this places low-income households deeply dependent on fossil fuels for health.
+Ironically, it can be argued that this places low-income households deeply dependent on fossil fuels for health.
 \par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
@@
  Build national fuel
-The CSO data, which is mainly data from the 2016 Cenus. 
+The CSO data, which is mainly data from the 2016 Census. 
 Fuel Calibration
@@
 The quantity M(t+1)-M(t) is the number of years between the milestones of periods t and t+1.
-The constant K is useful whenever the technology has no capacity initially, in order to allow capacity to build over time (if K were absent and initial capacity is zero, the technology would never acquire any capacity)
+The constant K is useful whenever the technology has no capacity initially, to allow the capacity to build over time (if K were absent and the initial capacity is zero, the technology would never acquire any capacity)
 Note that the sign of the constraint may also be of the "larger than or equal to" type to express a maximum rate of abandonment, in which case the "+" sign is replaced by a "–" sign in the right-hand-side of the constraint.
@@
 The residential sector plays a key part of this and it was built from the SEAI 2018 energy balance, the BER database, and CSO data.
-The data filtration process used in the residential sector derived a lot of conversation from the BER database, a group know as Ireland Energy Wiki by the National Retrofitting Modelling Group grew from building the residential sector, questions about the filters, default values, assumptions around thermal equations (EC..) and uncertainties were used to improve the data before it was used in TIM .
+The data filtration process used in the residential sector derived a lot of conversation from the BER database, a group known as Ireland Energy Wiki by the National Retrofitting Modelling Group grew from building the residential sector, questions about the filters, default values, assumptions around thermal equations (EC..) and uncertainties were used to improve the data before it was used in TIM .
 TIM has many internal project deadines, which required extensive meetings about the data, constraints and results as the modelling was progressing.
@@
 TIM’s residential sector was built in parallel with LEAP’s residential sector, so that the residential sector can easily simulate policy.
-The demand for results from TIM has increased significantly with increasing government ambitions, TIM will provide input into CAP 2021 and Ireland’s carbon budgets, working with DECC and the CCAC respectively, will also provide feedback and improve the robustness of TIM.
+The demand for results from TIM has increased significantly with increasing government ambitions, TIM will provide input into CAP 2021 and Ireland’s carbon budgets, working with DECC and the CCAC, respectively, will also provide feedback and improve the robustness of TIM.
 For these reasons, the TIM residential sector is at the cutting edge of modelling platforms.
@@
 
-First iteration - of model will need further development, DH, BEhavour, higher granularity and spatial will be easily done with the dataset. 
+First iteration - of the model will need further development, DH, BEhavour, higher granularity, and spatial will be easily done with the dataset. 
 %What does new model do that previous didn’t?

### 2021-07-08 manuscript.tex
8 July 2021

@@
 \usepackage{eurosym}
-%\usepackage{multirow}
+\usepackage{multirow}
+\usepackage{xcolor} 
+\usepackage{colortbl}
 
@@
 Energy System Optimisation Models (ESOMs) can be used to explore the least cost pathways, as they consider a large number of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
-Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM covers all energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section.
+Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM cover all energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section.
 
@@
 The DCM then estimated the probability of heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices.
-While Li's researcher has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights through more surveys and analysis.
+While Li's research has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights through more surveys and analysis.
 
@@
 The results find that when investments in individual heat pumps are allowed, air-to-water heat pumps are installed in non-district heating areas.
-This represents a significant electricity demand in identifying individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments and reducing system costs, fuel consumption, and $CO_2$ emissions.
-Hedegard and Münster found the system benefits of adding individual heat storage to the heat pumps are moderate, the main benefit is a reduced need for peak/reserve capacity investments.
+This represents a significant electricity demand in identifying individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments and orangeucing system costs, fuel consumption, and $CO_2$ emissions.
+Hedegard and Münster found the system benefits of adding individual heat storage to the heat pumps are moderate, the main benefit is a orangeuced need for peak/reserve capacity investments.
 The Balmorel model has provided insights, but it does not cover all aspects of the power and heat system, such as start-up costs, minimum load requirements, or part load efficiencies.
@@
 Two approaches are observed for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
-Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy system perspective, one reason for this is the wind power could also be stored in communal batteries.
-When identical communal and individual battery capacities where compared, it was fond  64\% of residential electricity demand is satisfied by an individual battery and only 36\% by a communal battery.
-While the battery location was explored, the PV system location was not, further investigation into the optimal PV system location could provide further insights.
+Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy system perspective, one reason for this is the wind power could also be stoorange in communal batteries.
+When identical communal and individual battery capacities where compaorange, it was fond  64\% of residential electricity demand is satisfied by an individual battery and only 36\% by a communal battery.
+While the battery location was exploorange, the PV system location was not, further investigation into the optimal PV system location could provide further insights.
 The household occupancy behaviour with and without household batteries on the PV system could alter the optimal pathway.
@@
 
-ESOM models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, however few Irish studies use ESOM models to explore the residential sector, two examples of Irish ESOM focusing on the residential sector are detailed here.
+ESOM models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem,Connolly2021AalborgEnergy-System}, however few Irish studies use ESOM models to explore the residential sector, two examples of Irish ESOM focusing on the residential sector are detailed here.
 
@@
 The bottom-up development of the residential sector in LEAP was combined with the cost optimisation capabilities of OSeMOYOS.
-Three scenarios are explored, namely, the reference scenario, energy efficiency scenario, and energy efficiency+ scenario.
+Three scenarios are exploorange, namely, the reference scenario, energy efficiency scenario, and energy efficiency+ scenario.
 The reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020.
-In the reference scenario, the residential total energy consumption rose 6.7 \% in 2020, compared to 2008.
+In the reference scenario, the residential total energy consumption rose 6.7 \% in 2020, compaorange to 2008.
 The share of space and water heating decreased from 82 \% to 77 \% in that period.
-In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulting in a 9.9 \% reduction of energy consumption in 2020, compared to the reference scenario.
+In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulting in a 9.9 \% orangeuction of energy consumption in 2020, compaorange to the reference scenario.
 The energy efficiency+ scenario also allows for 800,000 residential dwellings to be retrofitting, but accounts for deeper retrofits.
-The results in a reduction in of energy consumption of 16.5 \% in 2020, compared to the reference scenario.Rogan et al.
+The results in a orangeuction in of energy consumption of 16.5 \% in 2020, compaorange to the reference scenario.Rogan et al.
 discussed the barriers, obstacles to investment, split incentives, and information gaps which restrict energy efficiency improvements in the residential sector.
@@
 \cite{Thellufsen2019ImplementingIreland} investigated the potential for the implementation of district heating in both the commercial and residential sectors of Ireland.
-The paper analyses the $CO_2-80$ scenario ( 80 \% $CO_2$ reduction in 2050, compared to 1990 ) with and without district heating using the previous Irish TIMES model and EnergyPLAN.
+The paper analyses the $CO_2-80$ scenario ( 80 \% $CO_2$ orangeuction in 2050, compaorange to 1990 ) with and without district heating using the previous Irish TIMES model and EnergyPLAN.
 The previous version of Irish TIMES did not account for district heating, so EnergyPLAN is used to investigate the operation and  feasibility of district heating in Ireland.
 A sensitivity analysis was performed on investment costs, heat grid losses, and discount rates, however in all cases the inclusion of district heating is optimal.
-The final results find that district heating reduces fuel consumption by 3.5 \% and the conversion of 37 \% of the Irish heat demand to district heating is found to be optimal.
+The final results find that district heating orangeuces fuel consumption by 3.5 \% and the conversion of 37 \% of the Irish heat demand to district heating is found to be optimal.
 Both the TIMES model and climate target are superseded and the study did not include the future heat savings expected from retrofitting, nevertheless this study has provided insights on the potential of district heating.
@@
 A cost-benefit model (CBM) was fitted to the EAM to facilitate the conversion of the physical estimates into monetary amounts.
-Under the predicted scenario, energy savings easily outweighed programme costs, representing a benefit-cost ratio of 1.7.The benefits of reduced sickness, hospitalisation, drugs, and mortality were quantified and this moved the  benefit-cost ratio up to 3.0.
+Under the porangeicted scenario, energy savings easily outweighed programme costs, representing a benefit-cost ratio of 1.7.The benefits of orangeuced sickness, hospitalisation, drugs, and mortality were quantified and this moved the  benefit-cost ratio up to 3.0.
 The same model was used by Clinch, Healy, and King \cite{Clinch2001ModellingEfficiency} in addition three models of householder spending on energy are developed here to to evaluate the human behavioural component
@@
 The modelling process involved developing a computer model of the energy performance of the existing dwelling stock.
-It was noted that the energy consumption predicted by the model is considerably higher than given in the national energy balance, due to internal temperature assumptions.
-The model showed that the energy efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compared to the reference scenarios.
+It was noted that the energy consumption porangeicted by the model is considerably higher than given in the national energy balance, due to internal temperature assumptions.
+The model showed that the energy efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compaorange to the reference scenarios.
 
@@
 The building fabric efficiency of newly occupied dwellings developed takes into account the theoretical increase in energy efficiency due to improved 2008 and 2010 building regulations.
-Space heating demand accounted for 70\% of household energy consumption in newly built dwellings in 1997, down to just under 60\% in 2006 and the model predicts a reduction to 27\% by 2015.The paper acknowledges retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperatures affect on energy efficiency improvements.
+Space heating demand accounted for 70\% of household energy consumption in newly built dwellings in 1997, down to just under 60\% in 2006 and the model porangeicts a orangeuction to 27\% by 2015.The paper acknowledges retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperatures affect on energy efficiency improvements.
 \par In 2013, Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures to the existing detached, oil centrally heated, rural housing stock.
@@
 Concluding that the majority of energy efficiency progress for thermal use in Irish dwellings is being offset.
-The model thermal improvement measures resulted in a reduction of running costs and CO2 emissions by an average of 65\% for houses constructed prior to 1979, and this resulted in an average 12-year payback.
-Houses built in 1979 or after resulted in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures.
+The model thermal improvement measures resulted in a orangeuction of running costs and CO2 emissions by an average of 65\% for houses constructed prior to 1979, and this resulted in an average 12-year payback.
+Houses built in 1979 or after resulted in an average 26\% orangeuction in running costs and $CO_2$ emissions from retrofit measures.
 
@@
 This shows the the type of retrofit and type of house can have a large affect on the energy saved.
-Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F, and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is considered negligible.
+Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F, and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is consideorange negligible.
 
@@
 The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switching to Microsoft SQL.
-There has been few studies on the Irish residential sector using ESOMs, as residential bottom-up models are preferred when focusing on that sector.
+There has been few studies on the Irish residential sector using ESOMs, as residential bottom-up models are preferorange when focusing on that sector.
 The studies on Irish ESOM typically do not focus on the residential sector, instead they focus on the overall energy system, electricity generation, or transport.
@@
 This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}.
-The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of CO\textsubscript{2} emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousingGov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx.
+The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% orangeuction of CO\textsubscript{2} emissions in new dwellings compaorange to dwellings built in 2005 \cite{DepartmentofHousingGov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx.
 85\% of residential buildings are 2005 or older.
@@
     \hspace{8.5cm}
-    \pie[rotate =30,text=inside,radius=1.95,color={black!30,red!70,yellow!40}]{40/Transport,33/Heat,27/Electricity}
+    \pie[rotate =30,text=inside,radius=1.95,color={black!30,orange!70,yellow!40}]{40/Transport,33/Heat,27/Electricity}
     \end{tikzpicture}
@@
 
-Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to reduce Greenhouse Gas (GHG) emissions by 51\% in 2030 compared to 2018.
+Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to orangeuce Greenhouse Gas (GHG) emissions by 51\% in 2030 compaorange to 2018.
 The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}.
 These targets account for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to make sectoral five-year carbon budgets a legal requirement to aid progress of the forementioned long-term targets.
-\par The EU's first Nationally Determined Contribution (NDC), which had a 40\% GHG reduction target (first NDC), has been succeeded by a new target of at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}.
-Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-Climate Action Plan 2021 will have to be more ambitious than its predecessor, CAP2019 as it must reflect EU's updated NDC target and take account of the national target also.
+\par The EU's first Nationally Determined Contribution (NDC), which had a 40\% GHG orangeuction target (first NDC), has been succeeded by a new target of at least a 55\% GHG orangeuction target (updated NDC) by 2030 compaorange to 1990 levels \cite{SUBMISSIONSTATES}.
+Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG orangeuction target in 2030 compaorange to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+Climate Action Plan 2021 will have to be more ambitious than its porangeecessor, CAP2019 as it must reflect EU's updated NDC target and take account of the national target also.
 There are also renewable energy targets to consider for transport (RES-T), heat (RES-H), and electricity (RES-E) under the renewable directive (EU) 2018/2001, and the energy efficiency targets under (EU) 2018/2002.
@@
 
-%The CAP2019 MACC, shown in Fig.\ref{fig:CAP2019:MACC} provides $>300$ of the most cost-effective technologies options to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels. 
-%The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_{eq}$.
+%The CAP2019 MACC, shown in Fig.\ref{fig:CAP2019:MACC} provides $>300$ of the most cost-effective technologies options to orangeuce Non-ETS GHG by 30\% in 2030 compaorange to 2005 levels. 
+%The x-axis (width of each bar chart) shows the potential orangeuction of annual $MtCO2_{eq}$.
 The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_{eq}$ over the 2021 to 2030 period.
@@
 %\end{figure}
-%For example, Figure \ref{fig:CAP2019:MACC} shows the retrofitting of oil boilers is expected to reduce the Total Cost of Ownership (TCO) (ie.
+%For example, Figure \ref{fig:CAP2019:MACC} shows the retrofitting of oil boilers is expected to orangeuce the Total Cost of Ownership (TCO) (ie.
 save money), but only abate a small amount of carbon, whereas switching oil boilers to heat pumps will increase TCO (ie.
@@
 Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private investment towards low carbon or renewable energy.
-A €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}.
+A €15/ $tCO_2$ carbon tax was introduced in 2009 to orangeuce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}.
 Ireland's current carbon tax of €33.50/ $tCO_2$ will rise to €100/ $tCO_2$ in 2030.
@@
 In parallel, new dwellings must achieve a minimum share of 20\% renewable energy.
-The ambient heat used in heat pumps is included as a renewable energy, heat pumps are therefore the preferred heating technology despite the high capital costs.
+The ambient heat used in heat pumps is included as a renewable energy, heat pumps are therefore the preferorange heating technology despite the high capital costs.
 As mentioned in Section \ref{Intro:IreRSD}, new buildings in Ireland produce 70\% less GHG emissions than buildings in 2005, along with CAP2019 plans to retrofit 500,000 existing homes to at least a  ‘B2’ energy rating by 2030, which highlights the energy efficiency first approach in the residential sector.
@@
 CAP2019 plans to have approx.
-0.4\% of heating provided by DH in 2030, research using TIMES and EnergyPLAN has shown that 37\% of heating provided by DH is optimal if Ireland is to reduce GHG emissions by 80\% in 2050, compared to 1990 \cite{Thellufsen2019ImplementingIreland}.
+0.4\% of heating provided by DH in 2030, research using TIMES and EnergyPLAN has shown that 37\% of heating provided by DH is optimal if Ireland is to orangeuce GHG emissions by 80\% in 2050, compaorange to 1990 \cite{Thellufsen2019ImplementingIreland}.
 While Irish District Energy Association (IrDEA) suggest 57\% of heat can be sourced from DH, if supporting policy and regulation was in place.
 District Heating will likely play a role, however Gartland points out a range of organisational, technical, regulatory, and economic barriers to DH growth in Ireland \cite{Gartland2016AIreland}.
-\par While it maybe optimal to use Hydrotreated Vegetable Oil (HVO) in transport, it is also considered in the residential sector, as it can provide a cheap alternative for the 700,000 Irish dwellings using home heating oil ( mostly kerosene).
-HVO which meets the EU sustainability criteria, reduces GHG emissions by 60\% as compared to fossil fuels \cite{Soam2019FactorsSweden}.
+\par While it maybe optimal to use Hydrotreated Vegetable Oil (HVO) in transport, it is also consideorange in the residential sector, as it can provide a cheap alternative for the 700,000 Irish dwellings using home heating oil ( mostly kerosene).
+HVO which meets the EU sustainability criteria, orangeuces GHG emissions by 60\% as compaorange to fossil fuels \cite{Soam2019FactorsSweden}.
 The capacity of HVO is growing in Europe, but HVO is currently about twice the price of kerosene.
@@
 \label{Overview}
-\par  The expertise among energy policy researchers in Ireland combined with the global network of Energy Technology Systems Analysis Program (ETSAP) modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways.
+\par  The expertise among energy policy researchers in Ireland combined with the global network of Energy Technology Systems Analysis Program (ETSAP) modellers, makes TIMES the preferorange ESOM to explore Ireland's decarbonisation pathways.
 Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model and then updated with specific national data and assumptions.
@@
 Bituminous Coal & House   Coal & Coal                  \\
-Anthracite + Manufactured & Anthracite &         \\
-Lignite \textbackslash Brown Coal  & Manufactured Smokeless    &                       \\ \hline
+Anthracite + Manufactuorange & Anthracite &         \\
+Lignite \textbackslash Brown Coal  & Manufactuorange Smokeless    &                       \\ \hline
 Sod Peat & Sod Peat & Peat                                 \\
@@
 
-The dwelling stock data is obtained mainly from the SEAI BER Public database, which is firstly filtered, then scaled up using Central Statistics Office (CSO) data. 
+The dwelling stock data is obtained mainly from the SEAI BER Public database, which is firstly filteorange, then scaled up using Central Statistics Office (CSO) data. 
 The BER database is Ireland's national Energy Performance Certificate (EPC) register, which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD).
 The BERs are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
-Over half of Irish dwellings are captured in the BER database, every BER displays over 200 energy related variables.
+Over half of Irish dwellings are captuorange in the BER database, every BER displays over 200 energy related variables.
 The DEAP model, which was outlined by Dineen et al.\cite{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM} provides the results for the energy rating in the BER database.
@@
 One assumption, the internal temperature assumption results in a large error, where space and water heating are overestimated in DEAP.
-Space heating consumption reduced by 26 \% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance.
+Space heating consumption orangeuced by 26 \% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance.
 DEAP assumes all living areas ( living room, sitting room, and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C.
@@
 
-Other concerns for uncertainties in the database, which are not related to DEAP calculation are statistically anomalous spikes in the data, caused by applying default values \cite{Ahern2020ADatabases} and different companies generating variations of ±20\% in the assessment of energy consumption \cite{Harsman2016OnCertificates}.While the BER database provides high granularity data, it must be filtered to improve the representation of the dwelling stock data.
+Other concerns for uncertainties in the database, which are not related to DEAP calculation are statistically anomalous spikes in the data, caused by applying default values \cite{Ahern2020ADatabases} and different companies generating variations of ±20\% in the assessment of energy consumption \cite{Harsman2016OnCertificates}.While the BER database provides high granularity data, it must be filteorange to improve the representation of the dwelling stock data.
 The filters applied to the raw BER database are outlined in \ref{sec:appendixA}.
@@
 The total number of building archetypes per county in 2016 is used to scale up the BER database.
-The gap in total dwellings between the census year in 2016 and the model base year in 2018 was initially ignored due to lack of data.
+The gap in total dwellings between the census year in 2016 and the model base year in 2018 was initially ignoorange due to lack of data.
 However, later it was found that data on new dwelling electrical connections per archetype and per county are available from CSO dataset NDQ06, this would have filled the gap, the first iteration of the dwelling stock has a 2 \% error.
@@
 The lifetime and emissions/pollutants are also included, while the cost data is further disaggregated between capital expenditure ( equipment, installation \& additional capital costs ) and Operational \& Maintenance (O\&M) costs.
-\par The model has two types of retrofitting - shallow ( BER is improved by $ \leq 35 \%$ and deep ( BER is improved by $>35 \%$ ).
-The cost data was sourced from Ali et.
-al \cite{Ali2020ABuildings} who derived costs from SEAI's Better Energy Homes (BEH) and Better Energy Warmer Homes (BEWH) datasets, which are Ireland's largest retrofitting dataset ( over 285,000 retrofits ).
-To provide more robust retrofit costs, data is also used from the Typology Approach for Building Stock Energy Assessment (TABULA) report \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} which has more detailed cost data on 2 stages of retrofit for a small sample of 31 dwellings.
+\par The model has two types of retrofitting options per archetype,  a shallow retrofit is when the primary energy ($kWh/m^2/yr$) is reduced by $ \leq 35 \%$, and deep retrofit is where the primary energy is reduced by $>35 \%$, a reduction of $ \> 75 \%$ is considered infeasible.
+The different between the pre-retrofit and post-retrofit primary energy determines the type of retrofit, as shown in \ref{sec:appendixA3}.
+
+The retrofit cost data was sourced from Ali et.
+al \cite{Ali2020ABuildings} who derived costs from SEAI's Better Energy Homes (BEH) and Better Energy Warmer Homes (BEWH) datasets, combined this is Ireland's largest retrofitting dataset ( over 285,000 retrofits ).
+To provide more robust retrofit costs, data is also used from the Typology Approach for Building Stock Energy Assessment (TABULA) report \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} which has more detailed cost data with two stages of retrofit (standard and advanced) for a sample of 31 dwellings.
 The formulas used to find the average cost of a retrofit per type of retrofit and archetype are shown in Equations \ref{Eq:RetrofitCost1} and \ref{Eq:RetrofitCost2}.
@@
 \begin{split}
- \text{\euro}_{r,a,t} = Average ( (Ali_{r,t} * AF),TABULA_{r,a,t} )
+ \text{\euro}_{rf-rt,a} = Average ( (Ali_{rf-rt} * AF),TABULA_{rf-rt,a} )
 \end{split}
@@
 
-Where $ {\text{\euro}}{}_{r,a,t}$ is the TIM cost for retrofit by rating (r), archetype (a) and type of retrofit (t).
+Where $ {\text{\euro}}{}_{rf-rt,a}$ is the retrofit cost from pre-retrofit BER rating (rf) to post-retrofit rating (rt) and archetype (a).
 $Ali_{r,t}$ is the cost data from Ali et al.
-\cite{Ali2020ABuildings}, it contains all ratings and types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments and 1.1 for detached dwellings.
-The $TABULA_{r,a,t}$ is the average cost of a retrofit from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} per rating, archetype and type of retrofit, however, with only 62 samples in total, not all retrofits are included.
+\cite{Ali2020ABuildings}, it contains all ratings and types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments, 1 for attached, and 1.1 for detached dwellings.
+$TABULA_{rf-rt,a}$ is the average retrofit cost from pre-retrofit BER rating to post-retrofit BER rating and archetype sourced from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland}.
+However with only 62 samples in total not all retrofits are included.
 In this case, TABULA is ignored and Ali et al.
 is the only source of retrofit costs.
+The retrofit costs are cross-checked with the residential cost-optimal report \cite{AECOM2020ReportBuildings}.
 \par To simplify the model, the cost of retrofit per BER rating is not used in TIM. Instead, an average weighted cost per archetype and per retrofit is calculated.
@@
 \begin{split}
-\text{\euro}_{a,t} = \text{\euro}_{r,a,t} * \frac{TIM_{r,a}} {TIM_{a}}
+\text{\euro}_{a,t} = Average(\text{\euro}_{rf,a}) * \frac{TIM_{r,a}} {TIM_{a}}
 \end{split}
@@
 
-THe cost data of appliances 
-Topten -
+\par The cost data for different electrical appliances was sourced from Topten EU database.
+Topten has a large database covering electrical appliances available in the European market, the data collected includes cost and efficiency.
+While some Topten reports have looked at sales and market shares \cite{Michel2016EnergyREPORT}.
 
@@
 \label{LitR:Savings}
-Retrofit Savings - SEAI dataset not published..papers Usman,etc.
-While also accounting for more fuels and technological options, the residential sector in TIM is structured around retrofitting, not vice versa as previous versions, so that it is more aligned with ongoing climate mitigation investments.
+
+The previous section outlined the cost of retrofitting, and this section looks at the expected savings of retrofitting existing dwellings and building new dwellings.  
+The annual heating costs per archetype and per BER rating are sourced from SEAI's indicative energy running costs and the internal temperature assumptions in Table \ref{Table:InternalTemps} are included to calculate the costs, which are outlined in Table \ref{Table:FuelCost}
+
+\begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Heating fuel cost per BER rating and archetype }
+    \begin{tabular}{l|lcc}
+    \hline
+    BER Rating & Attached & \multicolumn{1}{l}{Detached} & \multicolumn{1}{l}{Apartment} \\ \hline
+A1         & $\text{\euro}$323      & $\text{\euro}$539                          & $\text{\euro}$200               \\
+A2         & $\text{\euro}$646      & $\text{\euro}$1078                         &$\text{\euro}$400                \\
+A3         & $\text{\euro}$804      & $\text{\euro}$1213                         & $\text{\euro}$500                 \\
+B1         & $\text{\euro}$745      & $\text{\euro}$1200                         & $\text{\euro}$440              \\
+B2         & $\text{\euro}$950      & $\text{\euro}$1500                         & $\text{\euro}$570                \\
+B3         & $\text{\euro}$1150     & $\text{\euro}$1900                         & $\text{\euro}$700                \\
+C1         & $\text{\euro}$862      & $\text{\euro}$1432                         & $\text{\euro}$499              \\
+C2         & $\text{\euro}$1022     & $\text{\euro}$1693                         & $\text{\euro}$624           \\
+C3         & $\text{\euro}$1181     & $\text{\euro}$1888                         & $\text{\euro}$686                \\
+D1         & $\text{\euro}$1434     & $\text{\euro}$2364                         & $\text{\euro}$836                \\
+D2         & $\text{\euro}$1701     & $\text{\euro}$2769                         & $\text{\euro}$964               \\
+E1         & $\text{\euro}$1979     & $\text{\euro}$3233                         & $\text{\euro}$1187               \\
+E2         & $\text{\euro}$2252     & $\text{\euro}$3646                         & $\text{\euro}$1319                \\
+F          & $\text{\euro}$2740     & $\text{\euro}$4403                         & $\text{\euro}$1626              \\
+G          & $\text{\euro}$2702     & $\text{\euro}$4398                         & $\text{\euro}$1687        \\                  \hline
+    \end{tabular}
+  \label{Table:FuelCost}
+\end{table}
+
+As TIM only inputs the savings per type of retrofit and archetype, some calculations must be performed on Table \ref{Table:FuelCost}.
+The first calculation is to find the savings from pre-retrofit running cost to post-retrofit running cost, as shown in Equation \ref{Eq:RetrofitSaving1}
+
+\begin{equation} 
+\begin{split}
+ Saving(\%)_{a,rf-rt} = 1 - ( \frac {Average(\text{\euro}_{rt,a}/year) } {Average(\text{\euro}_{rf,a}/year) } )
+\end{split}
+\label{Eq:RetrofitSaving1}
+\end{equation}
+
+Where $Saving(\%)_{a,rf-rt}$ are the savings ( or reduced running cost) per archetype (a), where the pre-retrofit BER rating is defined as the rating from (rf), and the post-retrofit BER rating is defined as the rating to (rt).
+
+Allocating each retrofit change to either shallow or deep is the next step, this is shown in \ref{sec:appendixA3}.
+The next calculation is to find the weighted average of each retrofit depth per archetype, this is then a TIM input.
+
+\begin{equation} 
+\begin{split}
+Saving(\%)_{a,d} = Average(Saving(\%)_{a,rf}) * \frac{TIM_{r,a}} {TIM_{a}}
+\end{split}
+\label{Eq:RetrofitSaving2}
+\end{equation}
+
+Where, t is the type of retrofit.
+${TIM_{r,a}}$ is the number of dwellings by rating and archetype and ${TIM_{a}}$ is the total number of dwellings of that archetype.The expected savings from each type of retrofit and for each archetype is outlined in Table \ref{Table:RetrofitSavings}.
+
+\begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Retrofit Savings}
+\begin{tabular}{l|lcc}
+Retrofit & Attached & \multicolumn{1}{l}{Detached} & \multicolumn{1}{l}{Apartment} \\ \hline
+Shallow  & 20\%     & 20\%                         & 22\%                          \\
+Deep     & 44\%     & 42\%                         & 37\%
+
+The higher thermal building efficiency in new dwellings, is also expected to reduce energy running costs.
+
+\end{tabular}
+\label{Table:RetrofitSavings}
+\end{table}
+
+The
 
@@
 
- \begin{tabular}{p{18em}lccc}
- \hline
-    \multirow{Energy Service Demand} & \multirow{Driver} & \multicolumn{2}{c}{Value} & \multirow{Unit} \\
-          &       & \multicolumn{1}{c}{2018} & \multicolumn{1}{c}{2050} &  \\ \hline
-    Residential Apartment Demand & Population & 206.80 & 628.93 & 000' \\
-    Residential Attached Demand & Population & 766.35 & 1056.50 & 000' \\
-    Residential Detached Demand & Population & 724.43 & 889.76 & 000' \\
-    \hline
-    \end{tabular}%
+% \begin{tabular}{p{18em}lccc}
+% \hline
+%    \multirow{Energy Service Demand} & \multirow{Driver} & %\multicolumn{2}{c}{Value} & \multirow{Unit} \\
+%          &       & \multicolumn{1}{c}{2018} & \multicolumn{1}{c}{2050} &  \\ \hline
+%    Residential Apartment Demand & Population & 206.80 & 628.93 & 000' \\
+ %   Residential Attached Demand & Population & 766.35 & 1056.50 & 000' \\
+ %   Residential Detached Demand & Population & 724.43 & 889.76 & 000' \\
+%    \hline
+%    \end{tabular}%
 
@@
 The new A-rated dwellings can be used as a guide to show the share of future energy service demands.
-From the BER database, it can be seen that A-rated dwellings use less than half the space heating energy of the average dwelling and A-rated water heating energy decreases by about 40\% compared to the average A-rated dwelling.
+From the BER database, it can be seen that A-rated dwellings use less than half the space heating energy of the average dwelling and A-rated water heating energy decreases by about 40\% compaorange to the average A-rated dwelling.
 However Pumps \& Fans energy increases as mechanical ventilation is used over natural ventilation and lighting usage in A-rated is more than double the average household.
@@
     \pie[   color = { yellow!90!black, green!60!black, blue!60,
-        red!70,
+        orange!70,
         gray!80,
@@
 For non-domestic premises, the thermal load is defined by office hours in line with the electrical demand.
-The domestic thermal loads considered in this study are listed in Table 2; the thermal load for the non-domestic
+The domestic thermal loads consideorange in this study are listed in Table 2; the thermal load for the non-domestic
 \subsubsection{Space and Water Heating}
@@
 %the new A-rated houses behave distinctively different which much less demand on space heating and with pumps and fans energy consumption usage increasing, while appliance end use is increasing across the BER ratings.
-The number of appliances has increased rapidly and while energy efficient improves are  reducing the energy per appliance, it is overshadowed by the growth in electrical appliances.
+The number of appliances has increased rapidly and while energy efficient improves are  orangeucing the energy per appliance, it is overshadowed by the growth in electrical appliances.
 There retrofitting grant which  requires building to have a HLI indication of 2 or less before a heat pump grant can be installs.
@@
 The data filtration process used in the residential sector derived a lot of conversation from the BER database, a group known as Ireland Energy Wiki by the National Retrofitting Modelling Group grew from building the residential sector, questions about the filters, default values, assumptions around thermal equations (EC..) and uncertainties were used to improve the data before it was used in TIM .
-TIM has many internal project deadines, which required extensive meetings about the data, constraints and results as the modelling was progressing.
+TIM has many internal project deadines, which requiorange extensive meetings about the data, constraints and results as the modelling was progressing.
 The data and results were presented to 30 external reviewers to provide feedback, this process proved invaluable having different expertise with different perspectives enabling TIM to fill in gaps.
@@
 \label{Discussion}
+5\% increase of residential energy in 2030 when internal temperatures are adjusted. 
 Soft-linking with LEAP and PLEXOS
@@
 %o	What can’t the model do? Where is it more suitable to use another model - Hannah Email
-
+Retrofitting ``rebound" saving orangeuces energy other is more difficult to measure like improved physical and mental health. 
 \subsection{Strengths}
@@
 Combines energy system effects
-Robust filtered input data about housing stock
+Robust filteorange input data about housing stock
 
@@
 \label{Dis:Limit}
+BER Database - No PV Solar capacity, No Heat Pump ready, Large defaults, wall types check?)
 %Discrepancies were investigated, and assumptions reviewed intuitively in order to achieve a good match.=
@@
 6 years for some consumer cohorts
-Type of Tenure is ignored
+Type of Tenure is ignoorange
 BER aggeragateed
@@
 straints and behavioural constraints in optimisation models.
-Including both conflates what is considered desirable under the economic welfare paradigm (i.e.
+Including both conflates what is consideorange desirable under the economic welfare paradigm (i.e.
 cost-minimising decisions) with “realistic decisions”, complicating the interpretation of the results.
@@
  \hline
- Ground Floor Area &  $<30m^2$ \& $>1000m^2$ & Considered erroneous \\
+ Ground Floor Area &  $<30m^2$ \& $>1000m^2$ & Consideorange erroneous \\
  Provisional Records & All & Higher Inaccuracies  \\
  Thermal Bridging Factor & $\leq0.15 W/m^2K$ & Outlier \\
- Declared Loss Factor & $>20kWh/day$ & Considered erroneous \\
+ Declaorange Loss Factor & $>20kWh/day$ & Consideorange erroneous \\
  Living Area Percent & $<5\%$ \& $>90\%$ & Outlier \\
  Heat System Supply Heat Fraction & $>21\%$ & Outlier  \\
- Heat System Efficiency Adjustment Factor & $<0.7$ & Considered erroneous \\
- Main Space System Efficiency & $\leq19\%$ & Considered erroneous \\
-Water System Efficiency Adjustment Factor & $<0.7$ & Considered erroneous\\
-Main Water System Efficiency & $\leq19\%$ \& $>320\%$ & Considered erroneous \\
+ Heat System Efficiency Adjustment Factor & $<0.7$ & Consideorange erroneous \\
+ Main Space System Efficiency & $\leq19\%$ & Consideorange erroneous \\
+Water System Efficiency Adjustment Factor & $<0.7$ & Consideorange erroneous\\
+Main Water System Efficiency & $\leq19\%$ \& $>320\%$ & Consideorange erroneous \\
  \hline
@@
 \newpage
+
+\subsection{}
+\label{sec:appendixA3}
+Depth of Retrofit defined by expected energy savings. The red cells are unfeasible, orange cells show deep retrofits and yellow represent shallow retrofits. 
+%The model has two types of retrofitting options per archetype,  a shallow retrofit is when the primary energy ($kWh/m^2/yr$) is reduced by $ \leq 35 \%$, and deep retrofit is where the primary energy is reduced by $>35 \%$, a reduction of $ \> 75 \%$ is considered infeasible.
+The different between the pre-retrofit and post-retrofit primary energy determines the type of retrofit, as shown i
+
+\begin{table}[htbp]
+ \centering
+ \footnotesize
+ \caption{Energy Savings per BER rating}
+ \begin{tabular}{|cl|llllllll}
+\hline   
+&  \multicolumn{8}{c}{Pre-Retrofit Rating}     \\ 
+ & & G  & F & E & D  & C & B3  & B2  & B1 \\ \hline 
+\multirow{8}{*}{\rotatebox[origin=c]{90}{Post-Retrofit Rating}} & A       & \cellcolor{red}83\% & \cellcolor{orange}73\% & \cellcolor{orange}67\% & \cellcolor{orange}57\% & \cellcolor{orange}38\%  & \cellcolor{orange}46\% & \cellcolor{yellow}35\% & \cellcolor{yellow}17\% \\
+ & B1 & \cellcolor{red}79\% & \cellcolor{orange}68\%  & \cellcolor{orange}61\% & \cellcolor{orange}48\% & \cellcolor{yellow}25\%  & \cellcolor{yellow}35\% & \cellcolor{yellow}22\% &      \\
+  & B2      & \cellcolor{orange}73\% & \cellcolor{orange}59\% & \cellcolor{orange}49\% & \cellcolor{yellow}33\% & \cellcolor{yellow}4\%   & \cellcolor{yellow}17\% &      &      \\
+  & B3      & \cellcolor{orange}68\% & \cellcolor{orange}50\% & \cellcolor{orange}39\% & \cellcolor{yellow}19\% & \cellcolor{yellow}0\% &      &      &      \\
+  & C       & \cellcolor{orange}72\% & \cellcolor{orange}57\% & \cellcolor{orange}47\% & \cellcolor{yellow}30\% &       &      &      &      \\
+ & D       & \cellcolor{orange}60\% & \cellcolor{orange}\cellcolor{orange}38\% & \cellcolor{yellow}25\% &      &       &      &      &      \\
+ & E       & \cellcolor{orange}47\% & \cellcolor{yellow}18\% &      &      &       &      &      &      \\
+  & F       & \cellcolor{yellow}35\% &      &      &      &       &      &      &
+
+ \end{tabular}%
+ \label{DepthofRetrofit}%
+\end{table}
+
 \printbibliography

### 2021-07-08 manuscript.tex
Done on Energy Service Demands

@@
 \usepackage{pgf-pie}
+\usepackage{pgfplots}
 \usepackage{calc}
@@
 \usepackage{pgfplots}
-\usepackage{amsmath}
 \usepackage{comment}
@@
 \DeclareSIUnit{\EUR}{\text{\euro}}
-
+ \pgfplotsset{compat=1.17}
 %\usepackage[table]{xcolor}
 %\usepackage{cite}
-
-\usepackage{biblatex}
+%\usepackage{harvard}
+%\usepackage{biblatex}
 \addbibresource{references.bib} %Imports bibliography file
@@
 
-The Census 2016 data \cite{CensusOffice} from the CSO is also used in calculating the dwelling stock.
+The Census 2016 data \cite{2016CensusIreland} from the CSO is also used in calculating the dwelling stock.
 Dataset E1005 provides a split of seven archetypes per county, which are aggregated to form three archetypes.
@@
 
-The higher thermal building efficiency in new dwellings, is also expected to reduce energy running costs.
-
 \end{tabular}
@@
 \end{table}
-
-The
 
@@
 Thus, uptake of technologies such as retrofitting will be driven, inter alia, by the net price for consumers.
+The residential energy service demand projections are a critical component of TIM. The future population projections are obtained from CSO's \textit{``M2F1"} scenario  %\cite{}. 
+ .
+
+HDD and warmer weather to offset each other???
+
 ENERGY CONSUMPTION CHARACTERISTICS of House - Table?
+The higher thermal building efficiency in new dwellings is also expected to reduce energy running costs. The average existing space and water heating and water h
 % The technical bottom-up development approach of the residential sector is combined with other economic sectors to meet macroeconomic defined demands, to provide GHG targets insights both the energy system and residential sector.
-\begin{equation} 
-\begin{split}
-E_y = {\sum_{A,M}} N_{A,y} \cdot S_{A,M}
-\end{split}
-%\[ \sum_{n=1}^{\infty} 2^{-n} = 1 \]
-\label{Eq:RetrofitSaving}
-\end{equation}
-$E_y$ = Energy Savings in Year y
-$N_{A,y}$ = Number of dwellings of archetype A retrofitted in year y
-$S_{A,M}$ = ¼ Energy Savings per annum for retrofit measureM carried
-out on archetype dwelling A
-A = For each archetype
-M = For each retrofit measure carried out
-
-Next - 
-\begin{equation} 
-\begin{split}
-D_M = {\sum_{A,M}^n} (A_kI_{ki} - A_kI_{kT} )
-\end{split}
-%\[ \sum_{n=1}^{\infty} 2^{-n} = 1 \]
-\label{Eq:EnergyandRetrofitSaving}
-\end{equation}
-Total Energy Demand , where D T = Total Energy Demand, n = total number of archetypes, A k = Frequency of each archetype, I k = Energy Intensity of each archetype - D M = Energy Savings by retrofit measure (M), n = total number of archetypes, A k = Frequency of each archetype dwellings eligible for retrofit measure, I ki = Pre-retrofit archetype energy intensity, I kr - post-retrofit archetype intensity - Tomas Paper
 
 Heterogeneous Occupancy Behaviour > House Stock
-\label{LitR:Demands}
+
 Residential Energy was >50\%higher before internal temperature adjustments
@@
 Time slices - 40 etc. 
-% textit{To accommodate an expanding population, the number of dwellings has also increased to currently stand at 1.7 million households, although this fgure remains signifcantly short of projected housing need, withannual average demand estimated to be up to 36,000 units for the next 30 years (IBEC 2018).
+% textit{To accommodate an expanding population, the number of dwellings has also increased to currently stand at 1.7 million households, although this fgure remains signifcantly short of projected housing need, with annual average demand estimated to be up to 36,000 units for the next 30 years (IBEC 2018).
 Yet despite the number of new homes built to ever increasing energy performance standards, the Irish housing stock is among the poorest in Europe in terms of energy effciency (Goggins et al.
@@
 The share of apartments changes from 12\% in 2018 to 24\% in 2050, while the shares of attached and detached dwellings decline by 4\% and 8\% respectively.
-Historical population estimates and future projections are obtained from the Central Statistics Office (CSO), Ireland \citep{CentralStatisticsOffice2020}.
-We use the M2F1 scenario since it represents medium growth in population and is in line with population projections used in other national sources \citep{Yakut2020}.
 
@@
 
-The residential stock projections up to 2040 are taken from ESRI's housing demand estimates \citep{Bergin2020}.
+The residential stock projections up to 2040 are taken from ESRI's housing demand estimates \cite{Bergin2020}.
 The stock is expected to increase by 40\% from 2018 level with a CAGR of 2\%.
@@
 
-\begin{figure}[!htbp]
-\centering
+%\begin{figure}[!htbp]
+%\centering
+%\begin{tikzpicture}
+%\begin{axis}[
+%    x tick label style={
+%		/pgf/number format/1000 sep=}, xlabel={Year},
+%	y tick label style={
+%	    /pgf/number},
+%		ylabel={Energy Consumption($ktoe$)},
+%		 scaled y ticks = false,
+%    xmin=2000, xmax=2050,
+%    ymin=0, ymax=40,
+%    xtick={2000,2010,2020,2030,2040,2050},
+ %   ytick={0,10,20,30,40},
+ %   ymajorgrids=true,
+%]
+%\addplot[
+%    color=blue,
+%    mark=square,
+%    ]
+%    coordinates {
+%    (2000,0.6)(2010,3.1)(2020,8.4)(2030,16.3)(2040,26.1)(2050,37)
+ %   };
+%    
+%\end{axis}
+%\end{tikzpicture}
+%    \caption{Space Cooling Residential Energy Consumption (Source: POTEnCIA) }
+ %   \label{fig:Cooling}
+%\end{figure}
+
 \begin{tikzpicture}
-\begin{axis}[
-    x tick label style={
-		/pgf/number format/1000 sep=}, xlabel={Year},
-	y tick label style={
-	    /pgf/number},
-		ylabel={Energy Consumption($ktoe$)},
-		 scaled y ticks = false,
-    xmin=2000, xmax=2050,
-    ymin=0, ymax=40,
-    xtick={2000,2010,2020,2030,2040,2050},
-    ytick={0,10,20,30,40},
-    ymajorgrids=true,
-]
-\addplot[
-    color=blue,
-    mark=square,
-    ]
-    coordinates {
-    (2000,0.6)(2010,3.1)(2020,8.4)(2030,16.3)(2040,26.1)(2050,37)
-    };
-
+\begin{axis}
+\addplot[color=red]{exp(x)};
 \end{axis}
 \end{tikzpicture}
-    \caption{Space Cooling Residential Energy Consumption (Source: POTEnCIA) }
-    \label{fig:Cooling}
-\end{figure}
 
@@
 Capacity, Efficiency, Constraints, Uptake, adapation
-heat output (Qt) and electricity input in a given time period (Pt) is expressed asQt= COPPt, where COP is the average annual coefficient of performance for the heat pump
+%heat output (Qt) and electricity input in a given time period (Pt) is expressed asQt= COPPt, where COP is the average annual coefficient of performance for the heat pump
 \subsubsection{Lighting, Appliances and Other}
@@
 \label{LitR:Constraints}
-LIST OF SCENARIOS?
+In Power Sector too! 
 GROWTH CONSTRAINT -A specific type of constraint may be defined to limit the share of process (p) in the total production of commodity (c).
@@
 \section{Results}
-
 \label{Results}

### 2021-07-09 manuscript.tex
Energy Service Demands

@@
 \label{LitR:Demands}
-%JF -The tax / subsidy inclusive prices are important in driving behaviour, and taxes and subsidies will be a vital policy instrument.
-Thus, uptake of technologies such as retrofitting will be driven, inter alia, by the net price for consumers.
-The residential energy service demand projections are a critical component of TIM. The future population projections are obtained from CSO's \textit{``M2F1"} scenario  %\cite{}. 
- .
+
+Energy service demands in the residential sector are driven by the total dwelling stock for non-archetype energy services or apartment, attached or detached dwelling stock for archetype energy services, which in turn is driven by population growth and the occupancy rate (the average number of people per dwelling).
+The residential stock projections up to 2040 are taken from Economic and Social Research Institute's (ESRI's) housing demand estimates \cite{BerginREGIONALLEVEL}.
+Archetype energy service demands are dependant on both the type of dwelling and whether it is existing or new.
+The average archetype energy service in new and existing per archetpe is shown in Table X.
+\begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Archetype Energy Service Demands}
+ \begin{tabular}{c|cc|cc|cc}
+ \hline
+ \multicolumn{1}{c|}{Energy Service}  & \multicolumn{2}{c|}{Apartments} & \multicolumn{2}{c|}{Attached} & \multicolumn{2}{c}{Detached}  \\
+
+  & Existing & New & Existing & New & Existing & New \\
+ \hline
+ Space Heating & 2.79 & 81\% & 2.79 & 81\% & 2.79 & 81\% \% \\
+ Water Heating & 3.63 & 81\% & 3.63 & 81\% & 3.63 & 81\% \% \\
+ Lighting & 7.5 & 307\% & 6.83 & 338\% & 6.21 & 379\%  \\
+ Pumps \& Fans  & 3.25 & 82\% & 3.25 & 82\% & 3.25 & 82\%  \\
+ Space Cooling  & 4.23 & 82\% & 4.23 & 82\% & 4.23 & 82\%  \\
+ \hline
+ \end{tabular}
+
+ \label{Archetype Energy Service Demands}
+\end{table}
+%%PICK UP HERE
+TABLE Y - Apartment Growth - 
+%% Ask Ankita for more info about Demand Calculations!!
+%Compound Annual Growth Rate (CAGR)
+%The occupancy rate is expected to fall to 2.12 in 2051 \citep{PropertyIndustryIreland2019}, from 2.75 in 2016 (CSO).
+
+%  Intermediate values are estimated using Equation \ref{eq:OR}.
+% \begin{equation}
+% \label{eq:OR}
+%  OR= a+b*log(\Delta t)
+% \end{equation}
+% here, $OR$ is the occupancy rate, $\Delta t$ is difference between years, parameters $a$ \& $b$ are estimated using historical values.
+
+ The stock is expected to increase by 40\% from 2018 level with a CAGR of 2\%.
+This results in an average of 27,600 new houses per annum between between 2021-2040.
+Beyond 2040, the population is used as a driver to project housing stock.
+The total housing stock obtained in 2050 is 2.57 million, which implies 8\% increase from 2040.
+% Table generated by Excel2LaTeX from sheet 'Sheet1'
+\begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Number of dwellings by type ('000)}
+    \begin{tabular}{rrrr}
+    \hline
+    \multicolumn{1}{l}{Year} & \multicolumn{1}{l}{Apartment} & \multicolumn{1}{l}{Attached } & \multicolumn{1}{l}{Detached} \\ \hline
+    2018  & 207   & 766   & 724 \\
+    2030  & 355   & 918   & 833 \\
+    2040  & 493   & 1003  & 878 \\
+    2050  & 629   & 1057  & 890 \\ \hline
+    \end{tabular}%
+  \label{tab:addlabel}%
+\end{table}%
 
@@
  \label{tab:pop}%
-\end{table}%
-
-The residential stock projections up to 2040 are taken from ESRI's housing demand estimates \cite{Bergin2020}.
-The stock is expected to increase by 40\% from 2018 level with a CAGR of 2\%.
-This results in an average of 27,600 new houses per annum between between 2021-2040.
-Beyond 2040, the population is used as a driver to project housing stock.
-The total housing stock obtained in 2050 is 2.57 million, which implies 8\% increase from 2040.
-% Table generated by Excel2LaTeX from sheet 'Sheet1'
-\begin{table}[htbp]
-  \centering
-  \footnotesize
-  \caption{Number of dwellings by type ('000)}
-    \begin{tabular}{rrrr}
-    \hline
-    \multicolumn{1}{l}{Year} & \multicolumn{1}{l}{Apartment} & \multicolumn{1}{l}{Attached } & \multicolumn{1}{l}{Detached} \\ \hline
-    2018  & 207   & 766   & 724 \\
-    2030  & 355   & 918   & 833 \\
-    2040  & 493   & 1003  & 878 \\
-    2050  & 629   & 1057  & 890 \\ \hline
-    \end{tabular}%
-  \label{tab:addlabel}%
 \end{table}%

### 2021-07-12 manuscript.tex
Technology Appendix Table and working on subsection heating

@@
 Deep     & 44\%   & 42\%     & 37\%               
-
 \end{tabular}
@@
 
+Demand TIMESLICE
+
 Energy service demands in the residential sector are driven by the total dwelling stock for non-archetype energy services or apartment, attached or detached dwelling stock for archetype energy services, which in turn is driven by population growth and the occupancy rate (the average number of people per dwelling).
@@
 Archetype energy service demands are dependant on both the type of dwelling and whether it is existing or new.
-The average archetype energy service in new and existing per archetpe is shown in Table X.
+The average archetype energy service in new and existing is shown in Table \ref{Archetype Energy Service Demands}
 \begin{table}[htbp]
@@
  \multicolumn{1}{c|}{Energy Service}  & \multicolumn{2}{c|}{Apartments} & \multicolumn{2}{c|}{Attached} & \multicolumn{2}{c}{Detached}  \\
-
+  \multicolumn{1}{c|}{}  & \multicolumn{2}{c|}{($kWh/house/yr$)} & \multicolumn{2}{c|}{($kWh/house/yr$)} & \multicolumn{2}{c}{($kWh/house/yr$)}  \\
   & Existing & New & Existing & New & Existing & New \\
  \hline
- Space Heating & 2.79 & 81\% & 2.79 & 81\% & 2.79 & 81\% \% \\
- Water Heating & 3.63 & 81\% & 3.63 & 81\% & 3.63 & 81\% \% \\
- Lighting & 7.5 & 307\% & 6.83 & 338\% & 6.21 & 379\%  \\
- Pumps \& Fans  & 3.25 & 82\% & 3.25 & 82\% & 3.25 & 82\%  \\
- Space Cooling  & 4.23 & 82\% & 4.23 & 82\% & 4.23 & 82\%  \\
+ Space Heating & 3,853 & 1,228 & 7,852 & 2,845 & 14,405 & 6,745 \\
+ Water Heating & 3,409 & 2,197 & 3,011 & 2,351 & 3,806 & 2,493 \\
+ Lighting & 179 & 156 & 254 & 229 & 419 & 392  \\
+ Pumps \& Fans  & 55  & 60 & 106 & 104 & 131 & 110  \\
+ Space Cooling  & 0 & 60 & 0 & 104 & 0 & 110  \\
  \hline
  \end{tabular}
-
  \label{Archetype Energy Service Demands}
 \end{table}
+
+Space and water heating, and lighting energy service demands for existing dwellings are average values of the base year dwelling stock and as all new dwellings are required to be A-rated, the new is the average for the existing A-rated dwellings.
+The Pumps \& Fans are similar, except in the new dwellings the value is halved and the remaining half is added to space cooling which is not seperately defined in the base year dwelling stock.
+This is an assumption which was cross-checked with POTEnCIA space cooling projections per household.
+
+While the energy service demand changes, so to does the number of dwellings by archetype and age. The existing dwellings decrease, oblsense rate of 0.35\%
+1.7 million in 2018, to 2.1 million 2030 to almost 2.6 million in 2050
+Share of new to old 23\ of buildings in 2030 will be new and 41\% in 2050.
+Apartment share moves dramitically from 12\% to 24\%. Detached 43\% to 35\% and Attached 45\% to 41\% 
 %%PICK UP HERE
-TABLE Y - Apartment Growth - 
+\begin{figure}[!htbp]
+ \centering
+ \caption{Number of dwellings by type and age ('000)}
+ \includegraphics[scale=0.42]{Figures/TIM_DwellingStock.png}
+ \label{fig:DwellingStoc}
+\end{figure}
+
 %% Ask Ankita for more info about Demand Calculations!!
@@
 
+\subsubsection{Space and Water Heating}
+\label{LitR:Heating}
+
+The demand for space and water heating is outlined in the previous section, and this section explains how water and space heating are achieved through technology and retrofits.
+While the demands are exogenous, the choice of technologies to at least satisfy the demand are endogenous and is captured in Equation \ref{Eq:Technology}
+
+\begin{equation} 
+\begin{split}
+Min \text{\euro} \sum_k VAR\_ACT_{k,i}(y)  \geq DM_i(y) 
+\end{split}
+\label{Eq:Technology}
+\end{equation}
+
+Where VAR\_ACT are the activity levels of end-use technology (k) for demand category (i) in year (y).
+DM is the exogenous demand for that energy service for year, y.
+The choice of technology is minimised while also satisfying all constraints.
+
+All technologies and retrofits are archetype specific and some technologies deliver both water and space heating, while other technologies can only deliver one type of heating.
+In addition, some technologies use more than one fuel to meet demand.
+The list of space and water heating technologies in TIM, is summarised in 
+Scope of technologies
+\cite{JointResearchCentre2017Techno-economicEU}
+Danish Energy Agency - Technology Catalogue
+Granularity
+Calculation is minised and obeies contrstaints
+
 %Space Heating demand - BER space heating is only within certain months and assumes 21C heating..
-Thermostats are not widely used for temperature regulation in private dwellings in Ireland; instead, the domestic heating schedule was assumed to be 07.00–09.00 h and 17.00–23.00 h for the heating season of October to May inclusive, based on the guidelines issued by the Sustainable Energy Authority of Ireland (SEAI) [32].
+%Thermostats are not widely used for temperature regulation in private dwellings in Ireland; instead, the domestic heating schedule was assumed to be 07.00–09.00 h and 17.00–23.00 h for the heating season of October to May inclusive, based on the guidelines issued by the Sustainable Energy Authority of Ireland (SEAI) [32].
 The thermal load consists of both space heating and a baseline demand for sustaining water heating.
@@
 The domestic thermal loads consideorange in this study are listed in Table 2; the thermal load for the non-domestic
-\subsubsection{Space and Water Heating}
-\label{LitR:Heating}
-
-Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which is a reason for Ireland's high excessive winter mortality \cite{Clinch2000HousingMortality}.
-Fuel Poverty which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths.
-An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}.
-Fuel Allowance and Household Benefits provide support to low-income households to alleviate fuel poverty.
-These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}.
-Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}.
-Ironically, it can be argued that this places low-income households deeply dependent on fossil fuels for health.
-\par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling.
-The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
 
@@
 %heat output (Qt) and electricity input in a given time period (Pt) is expressed asQt= COPPt, where COP is the average annual coefficient of performance for the heat pump
+
 \subsubsection{Lighting, Appliances and Other}
 \label{LitR:Lighting,Appliances}
+
 Capacity, Efficiency, Constraints, Uptake, adapation
+
 \subsubsection{User Constraints}
 \label{LitR:Constraints}
+
 In Power Sector too! 
@@
 
+\begin{table}[htbp]
+  \centering
+  \caption{Space and Water Heating Technologies}
+  \resizebox{\textwidth}{!}{
+ \begin{tabular}{cc|cc|cc|cc}
+ \hline
+  \multicolumn{2}{c|}{Technology} & \multicolumn{2}{c|}{Apartments} & \multicolumn{2}{c|}{Attached} & \multicolumn{2}{c}{Detached}  \\
+ Name & Description & Space  & Water  & Space  & Water   & Space  & Water \\
+ \hline
+ R-SH\_?\_KER\_N1 & Kerosene Boiler & \checkmark & \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55} \\
+R-SW\_?\_KER\_N1 & Kerosene Boiler & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
+ \hline
+ \end{tabular}
+}
+ \label{Heating Technology}
+\end{table}
+
 \printbibliography

### 2021-07-14 manuscript.tex
Starting Results

@@
 The total number of building archetypes per county in 2016 is used to scale up the BER database.
-The gap in total dwellings between the census year in 2016 and the model base year in 2018 was initially ignoorange due to lack of data.
+The gap in total dwellings between the census year in 2016 and the model base year in 2018 was initially ignoored  due to lack of data.
 However, later it was found that data on new dwelling electrical connections per archetype and per county are available from CSO dataset NDQ06, this would have filled the gap, the first iteration of the dwelling stock has a 2 \% error.
@@
 \end{figure}
+
+Also non- archetype energy servies demands -
 
@@
 
-\begin{tikzpicture}
-\begin{axis}
-\addplot[color=red]{exp(x)};
-\end{axis}
-\end{tikzpicture}
-
-The SEAI Residential Energy report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} provides high resolution details on energy demands in the residential sector.
-Variables such as disposable income, appliance ownership, dwelling location, type of occupier status (owner/rented ), BER labels, and age of dwelling.
-
 \subsubsection{Space and Water Heating}
@@
 The demand for space and water heating is outlined in the previous section, and this section explains how water and space heating are achieved through technology and retrofits.
-While the demands are exogenous, the choice of technologies to at least satisfy the demand are endogenous and is captured in Equation \ref{Eq:Technology}
+While the demands are exogenous, the choice of technologies to at least satisfy the demand are endogenous and is captured in Equation \ref{Eq:Technology}.
 
@@
 Where VAR\_ACT are the activity levels of end-use technology (k) for demand category (i) in year (y).
-DM is the exogenous demand for that energy service for year, y.
-The choice of technology is minimised while also satisfying all constraints.
-
-All technologies and retrofits are archetype specific and some technologies deliver both water and space heating, while other technologies can only deliver one type of heating.
-In addition, some technologies use more than one fuel to meet demand.
-The list of space and water heating technologies in TIM, is summarised in 
-Scope of technologies
-\cite{JointResearchCentre2017Techno-economicEU}
-Danish Energy Agency - Technology Catalogue
-Granularity
-Calculation is minised and obeies contrstaints
-
-%Space Heating demand - BER space heating is only within certain months and assumes 21C heating..
-%Thermostats are not widely used for temperature regulation in private dwellings in Ireland; instead, the domestic heating schedule was assumed to be 07.00–09.00 h and 17.00–23.00 h for the heating season of October to May inclusive, based on the guidelines issued by the Sustainable Energy Authority of Ireland (SEAI) [32].
-The thermal load consists of both space heating and a baseline demand for sustaining water heating.
-For non-domestic premises, the thermal load is defined by office hours in line with the electrical demand.
-The domestic thermal loads consideorange in this study are listed in Table 2; the thermal load for the non-domestic
-
- Build national fuel
-The CSO data, which is mainly data from the 2016 Census. 
-Fuel Calibration
- E1055 - Year, Archetype ( Detached, Semi-detached, terraced, Flat or apartment, converted to flat or apartment, bed-sit and not stated )
-E1008 Fuel by Archetype 
- EXXX - Age, Archetype and Fuel 
- dataset provides the central heating type or main fuel and
-
-Model calibration: archetypes demand The
-
-DEMAND Space cooling was disaggregated from electricity in POTEnCIA, this space cooling this was used in TIM.
-POTEnCIA provides high resolution energy service demands in 2018, along with energy service projections to 2050.
-Behavioral Constraints here
-
-FIGURE - Expected and Actual Consumption per m2
-ISO EN 13790 (Energy Performance of Buildings Directive).
-Capacity, Efficiency, Constraints, Uptake, adapation
-%heat output (Qt) and electricity input in a given time period (Pt) is expressed asQt= COPPt, where COP is the average annual coefficient of performance for the heat pump
+DM is the exogenous demand per energy service per year.
+The cost of the technology used is minimised while also satisfying all constraints.
+
+All technologies are archetype specific and some technologies deliver both water and space heating, while other technologies can only deliver one type of heating, the list of available technologies is outlined in \ref{sec:appendixA4}.
+Retrofits are included as they satisfy the residential space heating demand, without consuming fuel.
+The Danish Energy Agency's Technology Data for Individual Heat Production dataset was input to TIM, this dataset is also used in the EU TIMES model, the dataset has high granularity and is continuously updated.
+
+The number of main heating systems equals the number of dwellings, this assumption ignores any impact of secondary system heating.
+One expection in the rule is , which are only secondary heating systems.
+
+The base year heating energy consumption uses census data \cite{2016CensusIreland} from the CSO, dataset E1008 is used to obtain the fuel used in heating systems per archetype.
+The fuel usage is calibrated using the SEAI Energy Balance 2018.
 
@@
 
-Capacity, Efficiency, Constraints, Uptake, adapation
+Lighting, appliances, and pumps \& fans are all fueled by electricity, whereas cooking can be fueled by electricity, gas or LPG.
+As shown in Figure \ref{fig:Ireland2018ServiceDemands}, these energy services accounted for 20\% of energy consumption in 2018, however this share of consumption is expected to increase.
+Refrigeration, cloth washing, cloth drying, dish washing are under the appliances category and data is sourced from Topten \cite{Michel2016EnergyREPORT} and SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
+Cooking is also sourced from Eurostat's ``Energy consumption in households by type and end-used" excel datasheet and cross-checked with SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}  Appliances not included in this list are simply under the ``Other Appliances" category in TIM and these final values are cross-checked with POTEnCIA.
+Other Appliances include television, multimedia, and computer technology.
+Lighting and pumps \& fans are archetype specific energy services as the data is sourced from the BER database.
+Equation \ref{Eq:Technology} is also used, but unlike the heating energy service demands there is little or no choice in different technologies (k) for some of the demand categories (i), therefore, the model has no choice but to invest in a technology, however  cost and efficiency of each technology does determine the overall investment.
 
@@
 \label{LitR:Constraints}
-
-In Power Sector too! 
-GROWTH CONSTRAINT -A specific type of constraint may be defined to limit the share of process (p) in the total production of commodity (c).
-The constraint indicates that the flow of commodity (c) from/to process (p) - CommodiTypically, a growth constraint is of the following generic form (ignoring several indices for clarity: \cite{Loulou2016EnergyProgramme}
+The TIM model operates on the basis of perfect foresight, but
+\textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel} for this reason ``User Constraints" are applied in the residential sector to better reflect reality.
+The User Constraints applied to the residential sector include:
+\begin{itemize}
+ \item One Space and Water Heating System per household ( stoves excluded ) 
+ \item Maximum 60\% of existing buildings to be retrofitted by 2030, and 95 \% by 2070. Maximum \% interpolated between 2030 and 2070. 
+ \item The natural gas grid is currently connected to 39\% of apartments, 52\% of attached dwellings and 12\% of detached dwellings.
+The total share of dwellings connected to the gas grid is an upper constraint to grow by 50\% for each archetype.
+\item Maximum fuel share in cooking for Gas and LPG is 40\% and 10\% respectively. Base year fuel share for Gas is 32\% and LPG is 1.5 \%.
+\item Maximum 10\% solar input to dual fuel heating technology
+\item Maximum 30\% gas input into gas hybrid heating pumps
+\end{itemize}
+
+While other constraints are applied in different sectors, which affect the residetinal sector and vice-versa, they are not listed in this paper.
+The growth constraints listed above are calculated using the following \cite{Loulou2016DocumentationI} :
 \begin{equation} 
 \begin{split}
-VAR\_CAP(t + 1) \leq (1 + GROWTH^{M(t+1)-M(t)}) \cdot VAR\_CAP(t)+K
+VAR\_CAP(t + 1) \leq (1 + G^{M(t+1)-M(t)}) \cdot VAR\_CAP(t)+K
 \end{split}
@@
 \end{equation}
-The GROWTH coefficient is defined as a new attribute of the technology and represents the maximum annual growth allowed for the capacity.
-The quantity M(t+1)-M(t) is the number of years between the milestones of periods t and t+1.
-The constant K is useful whenever the technology has no capacity initially, to allow the capacity to build over time (if K were absent and the initial capacity is zero, the technology would never acquire any capacity)
-Note that the sign of the constraint may also be of the "larger than or equal to" type to express a maximum rate of abandonment, in which case the "+" sign is replaced by a "–" sign in the right-hand-side of the constraint.
-Equality is also allowed, but must be used only exceptionally in order to avoid railroading of the model.
-5.4.11
-
- % LIT REW INPUT DATA - EU \cite{Simoes2013TheProject} 
-- LI,er.alThe nationwide survey identified existing technologies, age,
-income, region, dwelling characteristics, and knowledge of eco- technology as the six most influential factors for determining homeowners' preferences for heating systems. Among
-%the new A-rated houses behave distinctively different which much less demand on space heating and with pumps and fans energy consumption usage increasing, while appliance end use is increasing across the BER ratings.
-The number of appliances has increased rapidly and while energy efficient improves are  orangeucing the energy per appliance, it is overshadowed by the growth in electrical appliances.
-There retrofitting grant which  requires building to have a HLI indication of 2 or less before a heat pump grant can be installs.
-Ireland’s aim of retrofitting 0.5 million homes from the existing 1.7 million in the before 2030.
-The effect of these targets also stretches out across the construction sector.
-The savings from retrofits will decrease fossil fuel energy usage, while most houses 00\% would need to get heat pumps too, to fulfil 0.4 million existing houses getting heat pumps.
-This energy efficiency first approach would allow for future technology choices, when current boiler are not at end of life.
-
-TIM was built to help Ireland provide a pathway to achieve 2030 and 2050 targets.
-The residential sector plays a key part of this and it was built from the SEAI 2018 energy balance, the BER database, and CSO data.
-The data filtration process used in the residential sector derived a lot of conversation from the BER database, a group known as Ireland Energy Wiki by the National Retrofitting Modelling Group grew from building the residential sector, questions about the filters, default values, assumptions around thermal equations (EC..) and uncertainties were used to improve the data before it was used in TIM .
-TIM has many internal project deadines, which requiorange extensive meetings about the data, constraints and results as the modelling was progressing.
-The data and results were presented to 30 external reviewers to provide feedback, this process proved invaluable having different expertise with different perspectives enabling TIM to fill in gaps.
-TIM will also be an open platform to help improve the model and provide transparency.
-TIM’s residential sector was built in parallel with LEAP’s residential sector, so that the residential sector can easily simulate policy.
-The demand for results from TIM has increased significantly with increasing government ambitions, TIM will provide input into CAP 2021 and Ireland’s carbon budgets, working with DECC and the CCAC, respectively, will also provide feedback and improve the robustness of TIM.
-For these reasons, the TIM residential sector is at the cutting edge of modelling platforms.
+
+Where $VAR\_CAP$ is the technology capacity in time period (t), G is the growth coefficient, defined as a new attribute of the technology and represents the maximum annual growth allowed for the capacity.
+The quantity $M(t+1)-M(t)$ is the number of years between the milestones of time periods t and t+1.
+The constant K is useful whenever the technology has no capacity initially, to allow the capacity to build over time.
 
@@
 
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
+The final version of the newly developed TIM provides some valuable insights on each sector and the sectoral interconnections.
+The results outlined in this paper are based upon six scenarios.
+TIM can only analysis the energy system, the other large source of GHG emissions is from agriculture, therefore some scenarios are labelled based on agriculture GHG reduction (AXX) and Energy System GHG reduction (EXX), the six scenarios are:
+\begin{itemize}
+    \item \textbf{No Mitigation} Least cost solution, where no emissions constraints are imposed
+    \item \textbf{With Additional Measures (WAM)} includes the implementation of Ireland’s CAP 2019.
+    \item \textbf{A25E65} Agriculture GHG emissions reduce by 25\%, Energy system GHG emissions reduces by 65\% by 2030 compared to 2018.
+    \item \textbf{A33E61} Agriculture GHGs reduces by 33\% and Energy system GHGs by 61\%
+    \item \textbf{A40E57} Agriculture GHGs reduces by 33\% and Energy system GHGs by 61\%
+    \item \textbf{A51E51} Equal GHG reduction share in of 51 \% in Agriculture and Energy Systems. 
+\end{itemize}
+
+While the first two scenarios do not comply with the overall 51 \% GHG reduction in 2030 compared to 2018 or net-zero by 2050, however they do provide insights on the incremental steps required.
+The last four scenarios comply with the overall 51 \% target and net-zero by 2050 and offer insights about different trajectories to achieve those targets.
+Under CAP 2019 agriculture, GHG emission reduction is 10-15\% on the projected levels in 2030 relative to 2017. The higher agriculture ambition explored GHG reductions of 25-51\%.
+
+XXX % PICK UP HERE
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[width=\textwidth]{Figures/CO2Waterfall.png}
+ \caption{$CO_2$ Waterfall}
+ \label{fig:$CO_2$ Waterfall}
+\end{figure}
+
 \section{Discussion}
@@
 \label{sec:appendix}
-
+%% The Appendices part is started with the command \appendix;
+%% appendix sections are then done as normal sections
 \subsection{}
@@
 
+\newpage
+
+\subsection{}
+\label{sec:appendixA4}
+
+List of water and space heating technologies available per archetype. The ``???" denotes either APT for Apartments, ATT for Attached, or DET for detached dwellings. 
 \begin{table}[htbp]
@@
   \multicolumn{2}{c|}{Technology} & \multicolumn{2}{c|}{Apartments} & \multicolumn{2}{c|}{Attached} & \multicolumn{2}{c}{Detached}  \\
- Name & Description & Space  & Water  & Space  & Water   & Space  & Water \\
- \hline
- R-SH\_?\_KER\_N1 & Kerosene Boiler & \checkmark & \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55} \\
-R-SW\_?\_KER\_N1 & Kerosene Boiler & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
- \hline
- \end{tabular}
+ Name & Fuel/Description & Space  & Water  & Space  & Water   & Space  & Water \\
+ \hline
+  \multicolumn{8}{c}{Boiler or Stove}  \\
+ \hline
+ R-SH\_???\_KER\_N1 & Oil & \checkmark & \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55} \\
+R-SW\_???\_KER\_N1 & Oil  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
+R-SW\_???\_KER\_N2 & Oil/Solar & \ding{55} & \ding{55} & \checkmark & \checkmark & \checkmark & \checkmark \\
+R-SW\_???\_KER\_N3 & Oil/Biomass & \ding{55} & \ding{55} & \checkmark & \checkmark & \checkmark & \checkmark \\
+ R-SH\_???\_GAS\_N1 & Gas  & \checkmark & \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55} \\
+R-SW\_???\_GAS\_N1 & Gas  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
+R-SW\_???\_GAS\_N2 & Gas/Solar  & \ding{55} & \ding{55} & \checkmark & \checkmark & \checkmark & \checkmark \\
+R-SW\_???\_GAS\_N3 & Gas/Biomass  & \ding{55} & \ding{55} & \checkmark & \checkmark & \checkmark & \checkmark \\
+ R-SH\_???\_LPG\_N1 & LPG  & \checkmark & \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55} \\
+R-SW\_???\_LPG\_N1 & LPG  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
+ R-SH\_???\_WOO\_N1 & Biomass  & \checkmark & \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55} \\
+R-SW\_???\_WOO\_N1 & Biomass  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
+R-SH\_???\_FPL\_N1 & Stove & \ding{55} & \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55} \\
+R-SW\_???\_FPL\_N1 & Stove & \ding{55} & \ding{55} & \checkmark & \checkmark & \checkmark & \checkmark\\
+R-SH\_???\_HVO\_N1 & HVO  & \ding{55} & \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55} \\
+R-SW\_???\_HVO\_N1 & HVO  & \ding{55} & \ding{55} & \checkmark & \checkmark & \checkmark & \checkmark \\
+ \hline
+  \multicolumn{8}{c}{Electrical}  \\
+ \hline
+R-SH\_???\_ELC\_N1 & Resistance Heater & \checkmark & \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55} \\
+ \hline
+  \multicolumn{8}{c}{Electrical Heat Pumps}  \\
+ \hline
+R-SH\_???\_ELC\_HPN1 & A-to-A  & \ding{55} & \ding{55} & \ding{55} & \ding{55} & \ding{55} & \ding{55} \\
+R-HC\_???\_ELC\_HPN1 & A-to-A  & \ding{55} & \ding{55} & \ding{55} & \ding{55} & \ding{55} & \ding{55} \\
+R-SH\_???\_ELC\_HPN2 & A-to-W & \checkmark & \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55} \\
+R-SW\_???\_ELC\_HPN1 & A-to-W & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
+R-SW\_???\_ELC\_HPN2 & A-to-W  w/Solar & \ding{55} & \ding{55} & \checkmark & \checkmark & \checkmark & \checkmark \\
+R-SH\_???\_ELC\_HPN3 & G-to-W  & \ding{55} & \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55}\\
+R-HC\_???\_ELC\_HPN2 & G-to-W & \checkmark& \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55}\\
+ \hline
+  \multicolumn{8}{c}{Gas and Hybrid Heat Pumps}  \\
+ \hline
+ R-SW\_???\_GAS\_HPN1 & Gas Absorption & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\
+ R-SW\_???\_GAS\_HPN2 & Gas Engine & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\
+ R-SW\_???\_GAS\_HHPN1 & Gas Hybrid & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\
+  \hline
+  \multicolumn{8}{c}{District Heating}  \\
+ \hline
+R-SW\_???\_HET\_N1 & Centralized & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\
+R-SW\_???\_HET\_N2 & Decentralized & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
+  \hline
+  \multicolumn{8}{c}{Water Heat Only}  \\
+ \hline
+ R-WH\_???\_ELC\_N1 & Resistance Heater  & \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55} & \checkmark \\
+ R-WH\_???\_ELC\_N2 & Solar Thermal  & \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55} & \checkmark \\
+   \hline
+  \multicolumn{8}{c}{Retrofits}  \\
+ \hline
+ R-FTFT-???\_Shallow & Shallow Retrofit & \checkmark & \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55} \\
+ R-FTFT-???\_Deep & Deep Retrofit & \checkmark & \ding{55} & \checkmark & \ding{55} & \checkmark & \ding{55} \\
+ \end{tabular}%
  }
- \label{Heating Technology}
+ \label{HeatingTech}%
 \end{table}
+
+\newpage
 

### 2021-07-16 manuscript.tex
Finishing Results!

@@
 
-Demand TIMESLICE
-
-Energy service demands in the residential sector are driven by the total dwelling stock for non-archetype energy services or apartment, attached or detached dwelling stock for archetype energy services, which in turn is driven by population growth and the occupancy rate (the average number of people per dwelling).
+Energy service demands in the residential sector are driven by the total dwelling stock for non-archetype energy services and apartment, attached or detached dwelling stock for archetype energy services, which in turn is driven by population growth and the occupancy rate (the average number of people per dwelling).
 The residential stock projections up to 2040 are taken from Economic and Social Research Institute's (ESRI's) housing demand estimates \cite{BerginREGIONALLEVEL}.
+However, there is archetpye specfic growth is limited or unavailable, for this reason a simple linear regression is used to calculate archertype growth.
+Table \ref{Table:DwellingGrowth} shows the growth rate for historical years (2016-2020) and the future years (2020 onwards).
+The growth rate of apartments is the highest and increasing up to 2040, while attached and detached dwelling types is decreasing, indicating a preference for smaller dwellings and compact urban planning to rise in the future.
+
+\begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Dwelling Growth}
+\begin{tabular}{l|lcc}
+Years & Apartment & \multicolumn{1}{l}{Attached } & \multicolumn{1}{l}{Detached} \\ \hline
+2016-2020  & 2.89\%   & 1.69\%     & 1.58\%     \\
+\hline
+2020-2030     & 4.26\%   & 1.15\%     & 0.81\%   \\
+2030-2040     & 3.33\%   & 0.89\%     & 0.53\%   \\
+2040-2050     & 2.47\%   & 0.52\%     & 0.14\%   \\
+\end{tabular}
+\label{Table:DwellingGrowth}
+\end{table}
 Archetype energy service demands are dependant on both the type of dwelling and whether it is existing or new.
@@
 \end{figure}
-
-Also non- archetype energy servies demands -
-
-%% Ask Ankita for more info about Demand Calculations!!
-%Compound Annual Growth Rate (CAGR)
-%The occupancy rate is expected to fall to 2.12 in 2051 \citep{PropertyIndustryIreland2019}, from 2.75 in 2016 (CSO).
-
-%  Intermediate values are estimated using Equation \ref{eq:OR}.
-% \begin{equation}
-% \label{eq:OR}
-%  OR= a+b*log(\Delta t)
-% \end{equation}
-% here, $OR$ is the occupancy rate, $\Delta t$ is difference between years, parameters $a$ \& $b$ are estimated using historical values.
 
@@
 % here, $OR$ is the occupancy rate, $\Delta t$ is difference between years, parameters $a$ \& $b$ are estimated using historical values.
-\begin{table}[htbp]
-\footnotesize
- \centering
- \caption{Energy service demands in TIM}
-
-% \begin{tabular}{p{18em}lccc}
-% \hline
-%    \multirow{Energy Service Demand} & \multirow{Driver} & %\multicolumn{2}{c}{Value} & \multirow{Unit} \\
-%          &       & \multicolumn{1}{c}{2018} & \multicolumn{1}{c}{2050} &  \\ \hline
-%    Residential Apartment Demand & Population & 206.80 & 628.93 & 000' \\
- %   Residential Attached Demand & Population & 766.35 & 1056.50 & 000' \\
- %   Residential Detached Demand & Population & 724.43 & 889.76 & 000' \\
-%    \hline
-%    \end{tabular}%
-
-  \label{tab:ESD_list}%
-\end{table}%
-
-% Table generated by Excel2LaTeX from sheet 'Population'
-\begin{table}[htbp]
- \centering
- \footnotesize
- \caption{Population}
- \begin{tabular}{cc}
- \hline
- Year & Population (millions) \\
- \hline
- 2018 & 4.85 \\
- 2020 & 4.98 \\
- 2030 & 5.40 \\
- 2040 & 5.82 \\
- 2050 & 6.19 \\
- \hline
- \end{tabular}%
- \label{tab:pop}%
-\end{table}%
 
@@
 
-%\begin{figure}[!htbp]
-%\centering
-%\begin{tikzpicture}
-%\begin{axis}[
-%    x tick label style={
-%		/pgf/number format/1000 sep=}, xlabel={Year},
-%	y tick label style={
-%	    /pgf/number},
-%		ylabel={Energy Consumption($ktoe$)},
-%		 scaled y ticks = false,
-%    xmin=2000, xmax=2050,
-%    ymin=0, ymax=40,
-%    xtick={2000,2010,2020,2030,2040,2050},
- %   ytick={0,10,20,30,40},
- %   ymajorgrids=true,
-%]
-%\addplot[
-%    color=blue,
-%    mark=square,
-%    ]
-%    coordinates {
-%    (2000,0.6)(2010,3.1)(2020,8.4)(2030,16.3)(2040,26.1)(2050,37)
- %   };
-%    
-%\end{axis}
-%\end{tikzpicture}
-%    \caption{Space Cooling Residential Energy Consumption (Source: POTEnCIA) }
- %   \label{fig:Cooling}
-%\end{figure}
+Demand TIMESLICE 40
 
@@
  \item Maximum 60\% of existing buildings to be retrofitted by 2030, and 95 \% by 2070. Maximum \% interpolated between 2030 and 2070. 
+ \item A prerequisite electrical heat pump requirement is applied to each archetype, the percentage of dwellings which requires a shallow or deep retrofit before a heat pump can be installed (  B2 rating for heat pump ready dwelling ).
+Hybrid heat pumps are not included in this constraint.
+ \begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Prerequisite Heat Pump Requirement}
+    \begin{tabular}{r|rrr}
+    \hline
+    \multicolumn{1}{l}{Requirement} & \multicolumn{1}{l}{Apartment} & \multicolumn{1}{l}{Attached } & \multicolumn{1}{l}{Detached} \\ \hline
+    No Retrofit  & 24.5 \%   & 14.3 \%   & 12 \% \\
+    Shallow Retrofit  & 28.5 \%   & 34.7 \%  & 36.8 \% \\
+    Deep Retrofit  & 47 \%   & 51 \%  & 51.2 \% \\
+    \end{tabular}%
+  \label{tab:HeatPumpConstraint}%
+\end{table}%
  \item The natural gas grid is currently connected to 39\% of apartments, 52\% of attached dwellings and 12\% of detached dwellings.
@@
 \label{Results}
-Electricity Supply, Technology and Fuel Mix, Retrofitting, Cost
-
-The final version of the newly developed TIM provides some valuable insights on each sector and the sectoral interconnections.
+
+The newly developed TIM provides some valuable insights on each sector and the sectoral interconnections.
 The results outlined in this paper are based upon six scenarios.
-TIM can only analysis the energy system, the other large source of GHG emissions is from agriculture, therefore some scenarios are labelled based on agriculture GHG reduction (AXX) and Energy System GHG reduction (EXX), the six scenarios are:
+While TIM can only analysis the energy system, the other large source of GHG emissions is from agriculture, therefore some scenarios are labelled based on agriculture GHG reduction percentage and Energy System GHG reduction percentage, the six scenarios are:
 \begin{itemize}
@@
 
-While the first two scenarios do not comply with the overall 51 \% GHG reduction in 2030 compared to 2018 or net-zero by 2050, however they do provide insights on the incremental steps required.
-The last four scenarios comply with the overall 51 \% target and net-zero by 2050 and offer insights about different trajectories to achieve those targets.
+While the first two scenarios do not comply with the Climate Action and Low Carbon Development (Amendment) Act 2021 target of an overall 51 \% GHG reduction in 2030 compared to 2018 or net-zero by 2050, however they do provide insights on the incremental steps required.
+The last four scenarios comply with 2030 and 2050 targets and offer insights about different trajectories to achieve those targets.
 Under CAP 2019 agriculture, GHG emission reduction is 10-15\% on the projected levels in 2030 relative to 2017. The higher agriculture ambition explored GHG reductions of 25-51\%.
 
-XXX % PICK UP HERE
+Figure \ref{fig:$CO_2$ Waterfall} provides an overview of  scenario affects on each sector in 2030, as the energy system ambition increases incrementally from left to right.
 
@@
 
+The share of sectoral $CO_2$ emissions is not evenly spread in 2018 and the optimal share of sectoral $CO_2$ reductions in 2030 is also unevenly distributed.
+In the A51E51 scenario, the optimal electricity annual emissions reduction is $7.1 MtCO_2$ in 2030, while transport currently emits the similiar $CO_2$ levels only reduces by $4.5 MtCO_2$.
+The residential sector reduces $CO_2$ emission by 63\% in A51E51 in 2030, electricity is 60 \% and transport is 38\%.
+Not only does Figure \ref{fig:$CO_2$ Waterfall} show the different optimal sectoral reductions, but also the ``low hanging fruit" in the large electricity reduction to the    ``high coconuts" in the step from A33E61 to A25E65, transport mitigates the majority of $CO_2$ emissions required.
+
+Zooming in and focusing on the fuel-mix within the residential sector gives further insights on the optimal pathways to achieve 2030 targets, Figure \ref{fig:ResFuel} shows this, with the base year stacked chart on the top row, the four climate bill compliant scenarios on rows 2-5 and the last two rows are non-compliant.
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[width=\textwidth]{Figures/RSD_Fuel2.png}
+ \caption{Residential Fuel per scenario}
+ \label{fig:ResFuel}
+\end{figure}
+
+In 2018, the residential sector had 124 PJ of primary energy that figure needs to decrease by 8-13 \% in 2030, to comply with climate targets.
+While Figure \ref{fig:$CO_2$ Waterfall} shows the large $CO_2$ reduction in electricity, this translates to a Figure \ref{fig:ResFuel} representing a large uptake of electricity in the residential sector within climate bill compliant scenarios.
+The large increase in ambient energy indicates the type of technology deployed in the residential sector, more on this in Figure XXX.
+The overall share of ambient heat increases from 1.4 \% in 2019 to between 13-26 \% in 2030.
+as a fuel  In climate bill compliant sceanrios peat and coal are completely phased out, while oil use decreases 90-93 \% by 2030.
+Gas energy increase by 34 \% in the A51E51 scenario, but remains relatively stable amoung other climate bill compliant sceanrios in 2030.
+By comparing compliant and noncompliant sceanrios, we can observe the favoured fuels and technologies, while also observing the incremental steps of emission reduction and overall cost change with each sceanrio.
+
+While Figure \ref{fig:ResFuel} shows the primary energy redcution required, it must be noted that retrofits are used to reduce the primary energy required ( reduced energy through behavioural change is not modelled ).
+Figure \ref{fig:Retrofits} shows both the number of retrofits per type and the energy savings for each of the four climate bill compliant scenarios.
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[width=\textwidth]{Figures/Retrofits.png}
+ \caption{Retrofits by type per scenario}
+ \label{fig:Retrofits}
+\end{figure}
+
+The number of retrofits required varies from 393,000 in A51E51 to 615,000 in A25E65, meaning between 23 - 36 \% of existing dwellings will require some form of retrofit.  
+As mentioned in Table \ref{tab:HeatPumpConstraint}, the rollout of retrofits constraints the amount of heat pumps which can be installed, however, a certain share of existing dwellings and all new dwellings are heat pump ready, i.e B2 rating or above.
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[width=\textwidth]{Figures/HeatPumps.png}
+ \caption{Heat Pump by type per scenario}
+ \label{fig:HeatPump}
+\end{figure}
+
+Where SW is space and water heating, HC is space heating and space cooling and A-to-W represents air-to-water type heat pumps.
+The optimal number of heat pumps in each scenario ranges from 492,000 in A51E51 to 938,000 in A25E65, which equates to 23 - 45 \% of dwellings having a heat pump in 2030.
+Close to 45,000 air to water heat pumps are installed in apartments across all scenarios, as there are no retrofits in apartments, these heat pumps are installed in the existing B2 rating or above apartments or new apartments.
+The detached air-to-water heat pumps are the most prevalent across scenarios and maybe considered the low hanging fruit.
+
+The advantage of energy sytem model is to oberve the intersectoral impacts for further insights...
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[width=\textwidth]{Figures/SectorCost.png}
+ \caption{Heat Pump by type per scenario}
+ \label{fig:SectorCost}
+\end{figure}
+
+Final Fuels share in each secotr Figure ....
+
 \section{Discussion}
@@
 Soft-linking with LEAP and PLEXOS
+Gas is a faoured fuel from Fig \ref{fig:RESFuel} this provides insights on potential to decarbonise
 %o	Discuss and interpret the results
@@
 BER Database - No PV Solar capacity, No Heat Pump ready, Large defaults, wall types check?)
+Community Retrofits - Better Energy Communities is our national retrofit initiative with grant support of up to €28 million for 2021.
+We support new approaches to achieving energy efficiency in Irish communities.
 %Discrepancies were investigated, and assumptions reviewed intuitively in order to achieve a good match.=

### 2021-07-19 manuscript.tex
Figure for Results - almost finished

@@
 \usepackage{colortbl}
+\usepackage{subcaption}
 
@@
 
-The advantage of energy sytem model is to oberve the intersectoral impacts for further insights...
+The advantage of an energy system model is to observe the intersectoral connections.
+Figure \ref{fig:SectorFuel} presents a boxplot of the 2030 fuel consumption in four different sectors, under the four scenarios which comply with the 51 \% reduction.....
 \begin{figure}[!htbp]
  \centering
+ \includegraphics[width=\textwidth]{Figures/SectorFuel.png}
+ \caption{Sector Fuel per scenario}
+ \label{fig:SectorFuel}
+\end{figure}
+
+for further insights...
+
+\begin{figure}[!htbp]
+ \centering
  \includegraphics[width=\textwidth]{Figures/SectorCost.png}
- \caption{Heat Pump by type per scenario}
+ \caption{Sector Cost per scenario}
  \label{fig:SectorCost}

### 2021-07-20 manuscript.tex
Draft 0 almosts done - redo Demands, the done and re-read.

--- 2021-07-16 manuscript.tex
@@
 %The TIMES-Ireland Model (TIM) is a newly developed optimisation model of the Irish energy system, which calculates the  cost-optimal decarbonisation pathway to meet future energy service demands while achieving Ireland's legally binding 2030 and 2050 climate targets.
-Write after the paper is completed
+\par Through historical settlement patterns, thermally inefficient building fabrics, high fuel poverty, and the social aspects of solid fuel heating, the Irish residential sector has proved hard to decarbonise.
+Despite this, Ireland has set one of the most ambitious climate targets into law, a 51 \% greenhouse gas reduction in 2030, compared to 2018.
+There is a need for a new state of the art model to provide insights on the optimal pathway for the energy system.
+This paper reviews residential and energy system models and describes the methodology applied in the residential sector in the newly developed TIMES Ireland Model (TIM).
+The paper includes some results to provide a discussion on some of the insights which need to be further investigated in the residential sector.
 
@@
 %% keywords here, in the form: keyword \sep keyword
-Energy systems optimisation model (ESOM) \sep The integrated MARKAL-EFOM system (TIMES) \sep Building decarbonisation \sep Model description \sep Residential energy transition
+Energy Systems Optimisation Model \sep TIMES \sep Residential decarbonisation \sep Model description 
 %% PACS codes here, in the form: \PACS code \sep code
@@
 
-Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways and interpreting the results, so Section \ref{sec:Introduction} of this paper looks at Ireland's climate targets, energy policy, and distinctive residential sector.
-While also examining how researchers have previously used Energy System Optimization Models (ESOMs) to analyse the residential sector.
-Section \ref{LitR} describes the methodology used to develop the residential sector while also outlining the input data.
-The process of applying constraints, technologies, costs, and demands in the model.
-Section \ref{Results} outlines the results obtained from the selected scenarios.
-Section \ref{Discussion} discusses the strengths and weakness of the newly developed model and how it differs from other models.
+Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways and interpreting the results.
+Section \ref{sec:Introduction} of this paper looks at Ireland's climate targets, energy policy, and distinctive residential sector.
+While also examining how researchers have previously used bottom-up residential models and Energy System Optimization Models (ESOMs) to analyse the residential sector.
+Section \ref{LitR} describes the methodology used to develop the residential sector.
+The treatment of input data is described in the model,  while the calculations used to develop the base year residential dwelling stock, and cost, saving and demand estimates are outlined, as is the process of applying constraints in the model.
+Section \ref{Results} outlines the results obtained from the six selected scenarios.
+Section \ref{Discussion} discusses the results, strengths, and weakness of the newly developed model and how it differs from other models.
 Lastly, section \ref{Conclusion} draws some conclusions on the model, results and discussion on future work.
@@
 Energy System Optimisation Models (ESOMs) can be used to explore the least cost pathways, as they consider a large number of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
-Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM cover all energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section.
+Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOMs cover all energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section.
 
@@
 Leibowicz et al.
-developed an expanded a version of the Open Source Energy Modeling System (OSeMOSYS) \cite{Howells2011OSeMOSYS:Development.} that integrates electricity generation and storage, residential building energy service demands, and end-use technologies \cite{Leibowicz2018OptimalServices}.Endogenous constraints on the market penetration rate of each technology and hourly service demand profiles are established and applied to the residential sector of Austin, Texas.
-Austin was chosen because of the high granularity data availability \cite{EnergyInc.}.
+developed an expanded version of the Open Source Energy Modeling System (OSeMOSYS) \cite{Howells2011OSeMOSYS:Development.} that integrates electricity generation and storage, residential building energy service demands, and end-use technologies \cite{Leibowicz2018OptimalServices}.Endogenous constraints on the market penetration rate of each technology and hourly service demand profiles are established and applied to the residential sector of Austin, Texas.
+Austin was chosen because of the high granularity of available data \cite{EnergyInc.}.
 Leibowicz et al.
@@
 This represents a significant electricity demand in identifying individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments and orangeucing system costs, fuel consumption, and $CO_2$ emissions.
-Hedegard and Münster found the system benefits of adding individual heat storage to the heat pumps are moderate, the main benefit is a orangeuced need for peak/reserve capacity investments.
+Hedegard and Münster found the system benefits of adding individual heat storage to the heat pumps are moderate, the main benefit is an reduced need for peak/reserve capacity investments.
 The Balmorel model has provided insights, but it does not cover all aspects of the power and heat system, such as start-up costs, minimum load requirements, or part load efficiencies.
@@
 Two approaches are observed for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
-Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy system perspective, one reason for this is the wind power could also be stoorange in communal batteries.
-When identical communal and individual battery capacities where compaorange, it was fond  64\% of residential electricity demand is satisfied by an individual battery and only 36\% by a communal battery.
+Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy system perspective, one reason for this is the wind power could also be storage in communal batteries.
+When identical communal and individual battery capacities where compared, it was fond  64\% of residential electricity demand is satisfied by an individual battery and only 36\% by a communal battery.
 While the battery location was exploorange, the PV system location was not, further investigation into the optimal PV system location could provide further insights.
@@
 In 2014, Rogan et al.
-\cite{Rogan2014LEAPsSystem} linked OSeMOYOS, an ESOM, to the Long Range Energy Alternatives Planning System (LEAP) simulation model, to explore specific energy efficiency policies not only in the residential sector, but also in the transport, industry and services sectors in Ireland.
+\cite{Rogan2014LEAPsSystem} linked OSeMOYOS, an ESOM, to the Long range Energy Alternatives Planning system (LEAP) simulation model, to explore specific energy efficiency policies not only in the residential sector, but also in the transport, industry and services sectors in Ireland.
 The bottom-up development of the residential sector in LEAP was combined with the cost optimisation capabilities of OSeMOYOS.
@@
 The reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020.
-In the reference scenario, the residential total energy consumption rose 6.7 \% in 2020, compaorange to 2008.
+In the reference scenario, the residential total energy consumption rose 6.7 \% in 2020, compared to 2008.
 The share of space and water heating decreased from 82 \% to 77 \% in that period.
-In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulting in a 9.9 \% orangeuction of energy consumption in 2020, compaorange to the reference scenario.
+In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulting in a 9.9 \% reduction of energy consumption in 2020, compared to the reference scenario.
 The energy efficiency+ scenario also allows for 800,000 residential dwellings to be retrofitting, but accounts for deeper retrofits.
-The results in a orangeuction in of energy consumption of 16.5 \% in 2020, compaorange to the reference scenario.Rogan et al.
+This results in an energy consumption reduction of 16.5 \% in 2020, compared to the reference scenario.Rogan et al.
 discussed the barriers, obstacles to investment, split incentives, and information gaps which restrict energy efficiency improvements in the residential sector.
@@
 \cite{Thellufsen2019ImplementingIreland} investigated the potential for the implementation of district heating in both the commercial and residential sectors of Ireland.
-The paper analyses the $CO_2-80$ scenario ( 80 \% $CO_2$ orangeuction in 2050, compaorange to 1990 ) with and without district heating using the previous Irish TIMES model and EnergyPLAN.
+The paper analyses the $CO_2-80$ scenario ( 80 \% $CO_2$ reduction in 2050, compared to 1990 ) with and without district heating using the previous Irish TIMES model and EnergyPLAN.
 The previous version of Irish TIMES did not account for district heating, so EnergyPLAN is used to investigate the operation and  feasibility of district heating in Ireland.
 A sensitivity analysis was performed on investment costs, heat grid losses, and discount rates, however in all cases the inclusion of district heating is optimal.
-The final results find that district heating orangeuces fuel consumption by 3.5 \% and the conversion of 37 \% of the Irish heat demand to district heating is found to be optimal.
+The final results find that district heating compared fuel consumption by 3.5 \% and the conversion of 37 \% of the Irish heat demand to district heating is found to be optimal.
 Both the TIMES model and climate target are superseded and the study did not include the future heat savings expected from retrofitting, nevertheless this study has provided insights on the potential of district heating.
@@
 A cost-benefit model (CBM) was fitted to the EAM to facilitate the conversion of the physical estimates into monetary amounts.
-Under the porangeicted scenario, energy savings easily outweighed programme costs, representing a benefit-cost ratio of 1.7.The benefits of orangeuced sickness, hospitalisation, drugs, and mortality were quantified and this moved the  benefit-cost ratio up to 3.0.
+Under the predicted scenario, energy savings easily outweighed programme costs, representing a benefit-cost ratio of 1.7.The benefits of reduced sickness, hospitalisation, drugs, and mortality were quantified and this moved the  benefit-cost ratio up to 3.0.
 The same model was used by Clinch, Healy, and King \cite{Clinch2001ModellingEfficiency} in addition three models of householder spending on energy are developed here to to evaluate the human behavioural component
@@
 The modelling process involved developing a computer model of the energy performance of the existing dwelling stock.
-It was noted that the energy consumption porangeicted by the model is considerably higher than given in the national energy balance, due to internal temperature assumptions.
-The model showed that the energy efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compaorange to the reference scenarios.
+It was noted that the energy consumption predicted by the model is considerably higher than given in the national energy balance, due to internal temperature assumptions.
+The model showed that the energy efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compared to the reference scenarios.
 
@@
 The building fabric efficiency of newly occupied dwellings developed takes into account the theoretical increase in energy efficiency due to improved 2008 and 2010 building regulations.
-Space heating demand accounted for 70\% of household energy consumption in newly built dwellings in 1997, down to just under 60\% in 2006 and the model porangeicts a orangeuction to 27\% by 2015.The paper acknowledges retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperatures affect on energy efficiency improvements.
+Space heating demand accounted for 70\% of household energy consumption in newly built dwellings in 1997, down to just under 60\% in 2006 and the model predicts a reduction to 27\% by 2015.The paper acknowledges retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperatures affect on energy efficiency improvements.
 \par In 2013, Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures to the existing detached, oil centrally heated, rural housing stock.
@@
 Concluding that the majority of energy efficiency progress for thermal use in Irish dwellings is being offset.
-The model thermal improvement measures resulted in a orangeuction of running costs and CO2 emissions by an average of 65\% for houses constructed prior to 1979, and this resulted in an average 12-year payback.
-Houses built in 1979 or after resulted in an average 26\% orangeuction in running costs and $CO_2$ emissions from retrofit measures.
+The model thermal improvement measures resulted in a reduction of running costs and CO2 emissions by an average of 65\% for houses constructed prior to 1979, and this resulted in an average 12-year payback.
+Houses built in 1979 or after resulted in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures.
 
@@
 The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switching to Microsoft SQL.
-There has been few studies on the Irish residential sector using ESOMs, as residential bottom-up models are preferorange when focusing on that sector.
+There has been few studies on the Irish residential sector using ESOMs, as residential bottom-up models are preferred when focusing on that sector.
 The studies on Irish ESOM typically do not focus on the residential sector, instead they focus on the overall energy system, electricity generation, or transport.
@@
 This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}.
-The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% orangeuction of CO\textsubscript{2} emissions in new dwellings compaorange to dwellings built in 2005 \cite{DepartmentofHousingGov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx.
+The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of CO\textsubscript{2} emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousingGov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx.
 85\% of residential buildings are 2005 or older.
@@
 
-Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to orangeuce Greenhouse Gas (GHG) emissions by 51\% in 2030 compaorange to 2018.
+Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to reduce Greenhouse Gas (GHG) emissions by 51\% in 2030 compared to 2018.
 The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}.
 These targets account for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to make sectoral five-year carbon budgets a legal requirement to aid progress of the forementioned long-term targets.
-\par The EU's first Nationally Determined Contribution (NDC), which had a 40\% GHG orangeuction target (first NDC), has been succeeded by a new target of at least a 55\% GHG orangeuction target (updated NDC) by 2030 compaorange to 1990 levels \cite{SUBMISSIONSTATES}.
-Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG orangeuction target in 2030 compaorange to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+\par The EU's first Nationally Determined Contribution (NDC), which had a 40\% GHG reduction target (first NDC), has been succeeded by a new target of at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}.
+Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
 Climate Action Plan 2021 will have to be more ambitious than its porangeecessor, CAP2019 as it must reflect EU's updated NDC target and take account of the national target also.
@@
 
-%The CAP2019 MACC, shown in Fig.\ref{fig:CAP2019:MACC} provides $>300$ of the most cost-effective technologies options to orangeuce Non-ETS GHG by 30\% in 2030 compaorange to 2005 levels. 
-%The x-axis (width of each bar chart) shows the potential orangeuction of annual $MtCO2_{eq}$.
+%The CAP2019 MACC, shown in Fig.\ref{fig:CAP2019:MACC} provides $>300$ of the most cost-effective technologies options to orangeuce Non-ETS GHG by 30\% in 2030 compared to 2005 levels. 
+%The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_{eq}$.
 The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_{eq}$ over the 2021 to 2030 period.
@@
 In parallel, new dwellings must achieve a minimum share of 20\% renewable energy.
-The ambient heat used in heat pumps is included as a renewable energy, heat pumps are therefore the preferorange heating technology despite the high capital costs.
+The ambient heat used in heat pumps is included as a renewable energy, heat pumps are therefore the preferred heating technology despite the high capital costs.
 As mentioned in Section \ref{Intro:IreRSD}, new buildings in Ireland produce 70\% less GHG emissions than buildings in 2005, along with CAP2019 plans to retrofit 500,000 existing homes to at least a  ‘B2’ energy rating by 2030, which highlights the energy efficiency first approach in the residential sector.
@@
 CAP2019 plans to have approx.
-0.4\% of heating provided by DH in 2030, research using TIMES and EnergyPLAN has shown that 37\% of heating provided by DH is optimal if Ireland is to orangeuce GHG emissions by 80\% in 2050, compaorange to 1990 \cite{Thellufsen2019ImplementingIreland}.
+0.4\% of heating provided by DH in 2030, research using TIMES and EnergyPLAN has shown that 37\% of heating provided by DH is optimal if Ireland is to orangeuce GHG emissions by 80\% in 2050, compared to 1990 \cite{Thellufsen2019ImplementingIreland}.
 While Irish District Energy Association (IrDEA) suggest 57\% of heat can be sourced from DH, if supporting policy and regulation was in place.
@@
 \par While it maybe optimal to use Hydrotreated Vegetable Oil (HVO) in transport, it is also consideorange in the residential sector, as it can provide a cheap alternative for the 700,000 Irish dwellings using home heating oil ( mostly kerosene).
-HVO which meets the EU sustainability criteria, orangeuces GHG emissions by 60\% as compaorange to fossil fuels \cite{Soam2019FactorsSweden}.
+HVO, which meets the EU sustainability criteria, reduces GHG emissions by 60\% as compared to fossil fuels \cite{Soam2019FactorsSweden}.
 The capacity of HVO is growing in Europe, but HVO is currently about twice the price of kerosene.
@@
 \label{Overview}
-\par  The expertise among energy policy researchers in Ireland combined with the global network of Energy Technology Systems Analysis Program (ETSAP) modellers, makes TIMES the preferorange ESOM to explore Ireland's decarbonisation pathways.
+\par  The expertise among energy policy researchers in Ireland combined with the global network of Energy Technology Systems Analysis Program (ETSAP) modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways.
 Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model and then updated with specific national data and assumptions.
@@
 \par TIMES is a bottom-up optimisation energy-environment model with perfect foresight which is used to analyse various levels of spatial, temporal and sectoral resolution.
-Exogenous technical, physical, and regulatory constraints can be applied and technologies have specified fuel types, efficiencies, lifetimes, availability, environmental characteristics, and fixed, variable and operation costs.
+Exogenous technical, physical, and regulatory constraints can be applied and technologies have specified fuel types, efficiency, lifetime, availability, environmental characteristics, and fixed, variable and operation costs.
 TIMES uses a linear programming optimizer matrix in General Algebraic Modeling System (GAMS), to compute all calculations.
@@
 However, the space heating calculation is used to estimate the expected energy consumption per dwelling, which makes some assumptions.
-DEAP assumes that each dwelling is single zoned, the heating season is frin October to May inclusive, and the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season.
-One assumption, the internal temperature assumption results in a large error, where space and water heating are overestimated in DEAP.
-Space heating consumption orangeuced by 26 \% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance.
+DEAP assumes that each dwelling is single zoned, the heating season is from October to May inclusive, and the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season.
+One assumption, the internal temperature assumption, results in a large error, where space and water heating are overestimated in DEAP.
+Space heating consumption is reduced by 26 \% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance.
 DEAP assumes all living areas ( living room, sitting room, and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C.
@@
 
-Where, t is the type of retrofit.
-${TIM_{r,a}}$ is the number of dwellings by rating and archetype and ${TIM_{a}}$ is the total number of dwellings of that archetype.The expected savings from each type of retrofit and for each archetype is outlined in Table \ref{Table:RetrofitSavings}.
+Where t is the type of retrofit.
+${TIM_{r,a}}$ is the number of dwellings by rating and archetype and ${TIM_{a}}$ is the total number of dwellings of that archetype.The expected savings from each type of retrofit and for each archetype are outlined in Table \ref{Table:RetrofitSavings}.
 
@@
 The residential stock projections up to 2040 are taken from Economic and Social Research Institute's (ESRI's) housing demand estimates \cite{BerginREGIONALLEVEL}.
-However, there is archetpye specfic growth is limited or unavailable, for this reason a simple linear regression is used to calculate archertype growth.
+However, there is archetpye specfic growth is limited or unavailable, for this reason a simple linear regression is used to calculate archetype growth.
 Table \ref{Table:DwellingGrowth} shows the growth rate for historical years (2016-2020) and the future years (2020 onwards).
-The growth rate of apartments is the highest and increasing up to 2040, while attached and detached dwelling types is decreasing, indicating a preference for smaller dwellings and compact urban planning to rise in the future.
+The growth rate of apartments is the highest and increasing up to 2040, while attached and detached dwellings are decreasing, indicating a preference for smaller dwellings to increase in the future.
 
@@
 
-Space and water heating, and lighting energy service demands for existing dwellings are average values of the base year dwelling stock and as all new dwellings are required to be A-rated, the new is the average for the existing A-rated dwellings.
-The Pumps \& Fans are similar, except in the new dwellings the value is halved and the remaining half is added to space cooling which is not seperately defined in the base year dwelling stock.
+Space and water heating, and lighting energy service demands for existing dwellings are the average values of the base year dwelling stock, and as all new dwellings are required to be A-rated, the new is the average for the existing A-rated dwellings.
+The Pumps \& Fans are similar, except in the new dwellings the value is halved and the remaining half is added to space cooling which is not separately defined in the base year dwelling stock.
 This is an assumption which was cross-checked with POTEnCIA space cooling projections per household.
 
-While the energy service demand changes, so to does the number of dwellings by archetype and age. The existing dwellings decrease, oblsense rate of 0.35\%
-1.7 million in 2018, to 2.1 million 2030 to almost 2.6 million in 2050
+While the energy service demand changes, so does the number of dwellings by archetype and age. The existing dwellings decrease the obsolescence rate of 0.35\%
+1.7 million in 2018, to 2.1 million, 2030 to almost 2.6 million in 2050
 Share of new to old 23\ of buildings in 2030 will be new and 41\% in 2050.
@@
 The new A-rated dwellings can be used as a guide to show the share of future energy service demands.
-From the BER database, it can be seen that A-rated dwellings use less than half the space heating energy of the average dwelling and A-rated water heating energy decreases by about 40\% compaorange to the average A-rated dwelling.
+From the BER database, it can be seen that A-rated dwellings use less than half the space heating energy of the average dwelling and A-rated water heating energy decreases by about 40\% compared to the average A-rated dwelling.
 However Pumps \& Fans energy increases as mechanical ventilation is used over natural ventilation and lighting usage in A-rated is more than double the average household.
@@
 
-The base year heating energy consumption uses census data \cite{2016CensusIreland} from the CSO, dataset E1008 is used to obtain the fuel used in heating systems per archetype.
+The base year heating energy consumption uses census data \cite{2016CensusIreland} from the CSO dataset E1008 is used to obtain the fuel used in heating systems per archetype.
 The fuel usage is calibrated using the SEAI Energy Balance 2018.
@@
 Lighting, appliances, and pumps \& fans are all fueled by electricity, whereas cooking can be fueled by electricity, gas or LPG.
-As shown in Figure \ref{fig:Ireland2018ServiceDemands}, these energy services accounted for 20\% of energy consumption in 2018, however this share of consumption is expected to increase.
-Refrigeration, cloth washing, cloth drying, dish washing are under the appliances category and data is sourced from Topten \cite{Michel2016EnergyREPORT} and SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
+As shown in Figure \ref{fig:Ireland2018ServiceDemands}, these energy services accounted for 20\% of energy consumption in 2018, however, this share of consumption is expected to increase.
+Refrigeration, cloth washing, cloth drying, and dish washing are under the appliances category and data is sourced from Topten \cite{Michel2016EnergyREPORT} and SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
 Cooking is also sourced from Eurostat's ``Energy consumption in households by type and end-used" excel datasheet and cross-checked with SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}  Appliances not included in this list are simply under the ``Other Appliances" category in TIM and these final values are cross-checked with POTEnCIA.
@@
 Lighting and pumps \& fans are archetype specific energy services as the data is sourced from the BER database.
-Equation \ref{Eq:Technology} is also used, but unlike the heating energy service demands there is little or no choice in different technologies (k) for some of the demand categories (i), therefore, the model has no choice but to invest in a technology, however  cost and efficiency of each technology does determine the overall investment.
+Equation \ref{Eq:Technology} is also used, but unlike the heating energy service demands, there is little or no choice in different technologies (k) for some of the demand categories (i), therefore, the model has no choice but to invest in a technology, however  cost and efficiency of each technology does determine the overall investment.
 
@@
 The total share of dwellings connected to the gas grid is an upper constraint to grow by 50\% for each archetype.
-\item Maximum fuel share in cooking for Gas and LPG is 40\% and 10\% respectively. Base year fuel share for Gas is 32\% and LPG is 1.5 \%.
+\item Maximum fuel share in cooking for gas and LPG is 40\% and 10\% respectively. Base year fuel share for Gas is 32\% and LPG is 1.5 \%.
 \item Maximum 10\% solar input to dual fuel heating technology
-\item Maximum 30\% gas input into gas hybrid heating pumps
+\item Maximum 30\% gas input to gas hybrid heating pumps
 \end{itemize}
 
-While other constraints are applied in different sectors, which affect the residetinal sector and vice-versa, they are not listed in this paper.
+While other constraints are applied in different sectors, which affect the residential sector and vice versa, they are not listed in this paper.
 The growth constraints listed above are calculated using the following \cite{Loulou2016DocumentationI} :
@@
 The last four scenarios comply with 2030 and 2050 targets and offer insights about different trajectories to achieve those targets.
-Under CAP 2019 agriculture, GHG emission reduction is 10-15\% on the projected levels in 2030 relative to 2017. The higher agriculture ambition explored GHG reductions of 25-51\%.
+Under CAP 2019 agriculture, GHG emission reduction is 10-15\% on the projected levels in 2030 relative to 2017.
+The higher agriculture ambition explored GHG reductions of 25-51\%.
+However, the energy consumed by agriculture is covered in this model.
 
@@
 
-The share of sectoral $CO_2$ emissions is not evenly spread in 2018 and the optimal share of sectoral $CO_2$ reductions in 2030 is also unevenly distributed.
+The share of sectoral $CO_2$ emissions is not evenly spread in 2018 and the optimal share of sectoral $CO_2$ reduction in 2030 is also unevenly distributed.
 In the A51E51 scenario, the optimal electricity annual emissions reduction is $7.1 MtCO_2$ in 2030, while transport currently emits the similiar $CO_2$ levels only reduces by $4.5 MtCO_2$.
@@
 The overall share of ambient heat increases from 1.4 \% in 2019 to between 13-26 \% in 2030.
-as a fuel  In climate bill compliant sceanrios peat and coal are completely phased out, while oil use decreases 90-93 \% by 2030.
-Gas energy increase by 34 \% in the A51E51 scenario, but remains relatively stable amoung other climate bill compliant sceanrios in 2030.
-By comparing compliant and noncompliant sceanrios, we can observe the favoured fuels and technologies, while also observing the incremental steps of emission reduction and overall cost change with each sceanrio.
-
-While Figure \ref{fig:ResFuel} shows the primary energy redcution required, it must be noted that retrofits are used to reduce the primary energy required ( reduced energy through behavioural change is not modelled ).
+as a fuel  In climate bill compliant scenarios, peat and coal are completely phased out, while oil use decreases 90-93 \% by 2030.
+Gas energy increase by 34 \% in the A51E51 scenario, but remains relatively stable among other climate bill compliant scenarios in 2030.
+By comparing compliant and noncompliant scenarios, we can observe the favoured fuels and technologies, while also observing the incremental steps of emission reduction and overall cost change in each scenario.
+
+While Figure \ref{fig:ResFuel} shows the primary energy reduction required, it must be noted that retrofits are used to reduce the primary energy required ( reduced energy through behavioural change is not modelled ).
 Figure \ref{fig:Retrofits} shows both the number of retrofits per type and the energy savings for each of the four climate bill compliant scenarios.
@@
 
-Where SW is space and water heating, HC is space heating and space cooling and A-to-W represents air-to-water type heat pumps.
+Where SW is space and water heating, HC is space heating and space cooling, and A-to-W represents air-to-water type heat pumps.
 The optimal number of heat pumps in each scenario ranges from 492,000 in A51E51 to 938,000 in A25E65, which equates to 23 - 45 \% of dwellings having a heat pump in 2030.
@@
 
-The advantage of energy sytem model is to oberve the intersectoral impacts for further insights...
+The advantage of an energy system model is to observe the intersectoral connections.
+Figure \ref{fig:SectorFuel} presents  the base year primary energy for four sectors - power generation, transport, services, and residential.
+Additionally, the primary energy of fuels in 2030 is included in the four scenarios which comply with the 51 \% reduction scenarios in 2030.
 \begin{figure}[!htbp]
  \centering
- \includegraphics[width=\textwidth]{Figures/SectorCost.png}
- \caption{Heat Pump by type per scenario}
+ \includegraphics[width=\textwidth]{Figures/IntesectoralFuels.png}
+ \caption{Sector Annual Fuel in 2018 and four scenarios in 2030}
+ \label{fig:SectorFuel}
+\end{figure}
+
+Some consistencies across the scenarios show the large growth of renewables in the power generation sector and consequently the uptake of electricity across the remaining sectors.
+The least ambitious A51E51 scenario shows 59 PJ of electricity used in residential in 2030, but as constraints tighten, the electricity is used more in transport, while the residential sector invests more in solar panels and other forms of renewable energy.
+While the use of bioenergy in transport and services also affects the residential sector as bioenergy is limited.
+Transport uses 16 PJ of natural gas in A51E51 to 0 PJ in A25E65, this allows the residential sector to use natural gas instead, which moves from 33 PJ in A51E51 to 24 PJ in A25E65.
+
+As a cheaper natural gas powered vehicle may effect the optimal energy mix and the annualised cost in other sectors.
+Figure \ref{fig:SectorCost} shows the annual costs of no mitigation in 2030 and the incremental increase of 2030 sectoral costs of the four scenarios which comply with the 51 \% reduction scenarios in 2030.
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[width=\textwidth]{Figures/CostWaterfall.png}
+ \caption{Sector Annual Cost for five scenarios in 2030}
  \label{fig:SectorCost}
@@
 
-Final Fuels share in each secotr Figure ....
+The no mitigation scenario is expected to cost $\text{\euro} 8.1 $ billion in 2030, that figure rises to $\text{\euro} 12.4 $ billion under the A51E51 scenarion, with electricity ( Power generation ) required an extra $\text{\euro} 2 $ billion in 2030.
+As the scenario constraints tighten, larger investments can be seen in the transport sector.
 
@@
 \label{Discussion}
-5\% increase of residential energy in 2030 when internal temperatures are adjusted. 
-Soft-linking with LEAP and PLEXOS
-Gas is a faoured fuel from Fig \ref{fig:RESFuel} this provides insights on potential to decarbonise
-%o	Discuss and interpret the results
+
+The methodology applied to the residential sector in TIM is transparent, up-to-date, and robust.
+The results outlined have summarised costs, fuel-mix, and intersectoral effects within the energy system.
+The results provide insights, further scenarios or sensitivity analysis can provide further insights from TIM.
+In addition, the results can be used in parallel with simulation models to examine specific climate policies or can alternative scenarios.
+
+The model brings alot of value to the sectroal carbon budgets and long-term climate targets, while also including more technologies and detail than any previous Irish ESOM.
+
+Gas is a favoured fuel from Fig \ref{fig:RESFuel} this provides insights on the potential to decarbonise
+
 %o	Focus on what added value the model brings
 %o	Also include future development priorities here
-%o	What can’t the model do? Where is it more suitable to use another model - Hannah Email
-Retrofitting ``rebound" saving orangeuces energy other is more difficult to measure like improved physical and mental health. 
+
 \subsection{Strengths}
@@
 \label{Dis:Limit}
+%o	What can’t the model do? Where is it more suitable to use another model - Hannah Email
 BER Database - No PV Solar capacity, No Heat Pump ready, Large defaults, wall types check?)
@@
 depending on usage \cite{Sunikka-Blank2012BuildingConsumption}
+
+Retrofitting ``rebound" saving  energy is more difficult to measure like improved physical and mental health.
+ArDEM SQL - exogenous behavioural effects 5\% increase of residential energy in 2030 when internal temperatures are adjusted. 
 \section{Conclusions}
@@
 \label{sec:appendixA3}
-Depth of Retrofit defined by expected energy savings. The red cells are unfeasible, orange cells show deep retrofits and yellow represent shallow retrofits. 
+Depth of retrofit defined by expected energy savings. The red cells are unfeasible, orange cells show deep retrofits, and yellow represent shallow retrofits. 
 %The model has two types of retrofitting options per archetype,  a shallow retrofit is when the primary energy ($kWh/m^2/yr$) is reduced by $ \leq 35 \%$, and deep retrofit is where the primary energy is reduced by $>35 \%$, a reduction of $ \> 75 \%$ is considered infeasible.

### 2021-07-23 manuscript.tex
Fix up the reference from mendely

@@
 %% with harvard style bibliographic references
-
-\documentclass[authoryear,preprint,12pt]{elsarticle}
+%\usepackage{cite}
+%\usepackage{harvard}
+%\usepackage{biblatex}
+%\usepackage[backend=biber]{biblatex}
+% \addbibresource{references.bib} %Imports bibliography file
+
+%%%%%%%%%%%%%%%%%%%%%%%
+%% Elsevier bibliography styles
+%%%%%%%%%%%%%%%%%%%%%%%
+%% To change the style, put a % in front of the second line of the current style and
+%% remove the % from the second line of the style you would like to use.
+%%%%%%%%%%%%%%%%%%%%%%%
+
+%% Numbered
+%\bibliographystyle{model1-num-names}
+
+%% Numbered without titles
+%\bibliographystyle{model1a-num-names}
+
+%% Harvard
+%\bibliographystyle{model2-names.bst}\biboptions{authoryear}
+
+%% Vancouver numbered
+%\usepackage{numcompress}\bibliographystyle{model3-num-names}
+
+%% Vancouver name/year
+%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}
+
+%% APA style
+%\bibliographystyle{model5-names}\biboptions{authoryear}
+
+%% AMA style
+%\usepackage{numcompress}\bibliographystyle{model6-num-names}
+
+%% `Elsevier LaTeX' style
+%\bibliographystyle{elsarticle-num}
+%%%%%%%%%%%%%%%%%%%%%%%
+
+\documentclass[preprint,12pt]{elsarticle}
 %\documentclass[10pt,final,journal,compsoc,a4paper]{IEEEtran}
@@
 %% for a journal layout:
-%% \documentclass[final,1p,times,authoryear]{elsarticle}
+%%\documentclass[final,1p,times,authoryear]{elsarticle}
 %% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,3p,times,authoryear]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
+%%\documentclass[final,3p,times,authoryear]{elsarticle}
+%%\documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
 %% \documentclass[final,5p,times,authoryear]{elsarticle}
@@
 \usepackage{subcaption}
-
+\usepackage{blindtext}
+\usepackage{lineno,hyperref}
+\modulolinenumbers[5]
 \DeclareSIUnit{\EUR}{\text{\euro}}
@@
 %\usepackage[table]{xcolor}
-%\usepackage{cite}
-%\usepackage{harvard}
-%\usepackage{biblatex}
-\addbibresource{references.bib} %Imports bibliography file
+\usepackage{biblatex}
 
@@
 %% \usepackage{amsthm}
-
+\addbibresource{references.bib}
 %% The lineno packages adds line numbers. Start line numbering with
@@
 
-\author[inst1,inst2]{Jason Mc Guire\footnote{Contact: j.mcguire@ucc.ie}}
-
+\author[inst1,inst2]{Jason Mc Guire \corref{j.mcguire@ucc.ie}}
+\cortext[j.mcguire@ucc.ie]{j.mcguire@ucc.ie}
+%footnote{E-mail: \url{j.mcguire@ucc.ie}}}
+
+%\footnote{Contact: j.mcguire@ucc.ie}}
+%\thanks{Email: \url{j.mcguire@ucc.ie}}}
 \affiliation[inst1]{organization={Energy Policy and Modelling Group, MaREI Centre},%Department and Organization
@@
 %The TIMES-Ireland Model (TIM) is a newly developed optimisation model of the Irish energy system, which calculates the  cost-optimal decarbonisation pathway to meet future energy service demands while achieving Ireland's legally binding 2030 and 2050 climate targets.
-\par Through historical settlement patterns, thermally inefficient building fabrics, high fuel poverty, and the social aspects of solid fuel heating, the Irish residential sector has proved hard to decarbonise.
-Despite this, Ireland has set one of the most ambitious climate targets into law, a 51 \% greenhouse gas reduction in 2030, compared to 2018.
-There is a need for a new state of the art model to provide insights on the optimal pathway for the energy system.
-This paper reviews residential and energy system models and describes the methodology applied in the residential sector in the newly developed TIMES Ireland Model (TIM).
-The paper includes some results to provide a discussion on some of the insights which need to be further investigated in the residential sector.
+\par The Irish residential sector has proved hard to decarbonise, yet recently Ireland has set one of the most ambitious climate targets into law, a 51 \% greenhouse gas reduction in 2030, compared to 2018.
+There is a need for a new state of the art model to provide insights on the optimal decarbonisation pathway for the Irish energy system.
+This paper reviews residential and energy system models and describes the methodology applied to the development of the residential sector in the new TIMES Ireland Model (TIM).
+The paper includes some results to provide a discussion on some of the insights and highlights the insights which may need to be further investigated in the residential sector.
 
@@
 
-%%Graphical abstract
-%%\begin{graphicalabstract}
-%%\includegraphics{grabs}
-%%\end{graphicalabstract}
-
-%%Research highlights
-%%\begin{highlights}
-%%\item Research highlight 1
-%%\item Research highlight 2
-%%\end{highlights}
-
 \begin{keyword}
-%% keywords here, in the form: keyword \sep keyword
+
 Energy Systems Optimisation Model \sep TIMES \sep Residential decarbonisation \sep Model description
-%% PACS codes here, in the form: \PACS code \sep code
-%%\PACS 0000 \sep 1111
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-%% \MSC 0000 \sep 1111
+
 \end{keyword}
@@
 Section \ref{LitR} describes the methodology used to develop the residential sector.
-The treatment of input data is described in the model,  while the calculations used to develop the base year residential dwelling stock, and cost, saving and demand estimates are outlined, as is the process of applying constraints in the model.
-Section \ref{Results} outlines the results obtained from the six selected scenarios.
+The treatment of input data in the model is described, the calculations are explained, and assumptions and constraints are highlighted.
+Section \ref{Results} outlines the six sceanrios used and displays the energy system and residential results.
 Section \ref{Discussion} discusses the results, strengths, and weakness of the newly developed model and how it differs from other models.
@@
 
-Energy System Optimisation Models (ESOMs) can be used to explore the least cost pathways, as they consider a large number of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
-Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOMs cover all energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section.
-
-\par For example, Li et al.
-\cite{Li2018IncorporatingModel} investigated the heating technology preference in UK using UK TIMES Model (UKTM).
-As TIMES \cite{Loulou2016DocumentationI} assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
-Li set out to incorporate a Discrete Choice Model (DCM) for TIMES to better reflect how homeowners choose heating technologies.
+ESOMs have been used to explore the least cost pathways, as they consider a large number of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
+Some of the main ESOMs currently used include MESSAGE \cite{Messner1995Model-BasedPlanning}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOMs cover all energy-related sectors, they can also be used to specifically explore certain economic sector in higher granularity.
+
+TIMES has been extensively used in both the UK and Ireland, and there are several examples using TIMES to focus on the residential sector.
+TIMES assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
+With this in mind, Li et al.
+\cite{Li2018IncorporatingModel} investigated the heating technology preference in the UK using UK TIMES Model (UKTM) by incorporating a Discrete Choice Model (DCM) to better reflect how homeowners choose heating technologies.
 The DCM was developed from a nationwide survey with 442 homeowners completing the survey.
 The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems.
-The DCM then estimated the probability of heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices.
+The DCM then estimated the probability of heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices, this identifies perfect foresight as a weakness in TIMES.
 While Li's research has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights through more surveys and analysis.
-
-Leibowicz et al.
-developed an expanded version of the Open Source Energy Modeling System (OSeMOSYS) \cite{Howells2011OSeMOSYS:Development.} that integrates electricity generation and storage, residential building energy service demands, and end-use technologies \cite{Leibowicz2018OptimalServices}.Endogenous constraints on the market penetration rate of each technology and hourly service demand profiles are established and applied to the residential sector of Austin, Texas.
-Austin was chosen because of the high granularity of available data \cite{EnergyInc.}.
-Leibowicz et al.
-results find net present costs in the \textit{``Policy scenarios"} ( emissions declining linearly toward 10\% of their current level by 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations imposed ).
-However, the percentage increase in net present cost due to the policy diminishes with a more thermally efficient building stock.
-Leibowicz et al.
-also found electrification to be the dominant strategy for cost-effectively decarbonizing urban residential dwellings.
-These insights show the strength of an ESOM, but not all costs are taken into account in Leibowicz study, the capital costs associated with more thermally efficient buildings are not included.
-
-Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised  Balmorel model \cite{Wiese2018BalmorelModel} to handle the integrated power and district heating system to analyse how heat pumps with complementing heat storage affect wind power investments in Denmark.
-The model was disaggreated between a western and eastern Denmark region and the model covers Norway, Sweden, Finland, and Germany to cover import/export.
-Exogenous user constraints applied to the model, based on the current Danish policy includes: Wind power must provide at least 50\% of the national electricity consumption by 2020, coal is phased out in heat and power plants by 2030, and individual oil boilers are phased out by 2030.
-The results find that when investments in individual heat pumps are allowed, air-to-water heat pumps are installed in non-district heating areas.
-This represents a significant electricity demand in identifying individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments and orangeucing system costs, fuel consumption, and $CO_2$ emissions.
-Hedegard and Münster found the system benefits of adding individual heat storage to the heat pumps are moderate, the main benefit is an reduced need for peak/reserve capacity investments.
-The Balmorel model has provided insights, but it does not cover all aspects of the power and heat system, such as start-up costs, minimum load requirements, or part load efficiencies.
-Nevertheless, the insights are valuable and the results can be soft-linked to other models for deeper analysis.
-
-Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model \cite{stergaard2015ReviewingSimulations} to explore the feasibility of photovoltaic (PV) systems installed at the household level in combination with batteries in Samsø, which is known for being Denmark’s renewable energy island.
-Two approaches are observed for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
+ \par Many sectors are increasingly reliant on electricity generation as lower carbon intensity electricity is considered a cornerstone in optimal decarbonisation.
+Using the Open Source Energy Modeling System (OSeMOSYS) \cite{Howells2011OSeMOSYS:Development.}
+Leibowicz et al.\cite{Leibowicz2018OptimalServices} explores the optimal decarbonisation of the residential sector by examining electricity generation and storage strategies.
+The results find net present costs in the \textit{``Policy scenarios"} ( $CO_2$ declines linearly by 90\% in 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations imposed ).
+However, the percentage increase in net present cost due to the policy diminishes with a more thermally efficient building stock, highlighting the importance of energy building regulation as thermally efficient building stock can reduce the decarbonisation cost.
+District Heating is a proven low carbon heating technology, however it is almost non-existant and has not been included in previous ESOM in Ireland.
+Denmark are leading experts in this area and Denmark's district heating system has been comprehensively examined.
+One study by
+Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} used the Balmorel model to analyse how heat pumps with complementing heat storage affect wind power investments in Denmark.
+Exogenous user constraints applied to the model, based on the current Danish policy.
+The results find that air-to-water heat pumps are installed in non-district heating areas, while the system benefits of adding individual heat storage to the heat pumps are only moderate, the main benefit is a reduced need for peak/reserve capacity investments.
+This identifies individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments in areas where district heating is not available.
+Another technology which is proven but under utilised in Ireland, is photovoltaic (PV) solar panels.
+While Ireland's policy and building regulations have encouraged the uptake of PVs, however the market share of PV systems remains low, and as the cost continues to fall they can alter optimal decarbonisation pathways.
+Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model to explore the feasibility of PV systems installed at the household level in combination with batteries in Samsø.
+Two approaches are explored for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
 Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy system perspective, one reason for this is the wind power could also be storage in communal batteries.
-When identical communal and individual battery capacities where compared, it was fond  64\% of residential electricity demand is satisfied by an individual battery and only 36\% by a communal battery.
-While the battery location was exploorange, the PV system location was not, further investigation into the optimal PV system location could provide further insights.
-The household occupancy behaviour with and without household batteries on the PV system could alter the optimal pathway.
+When identical communal and individual battery capacities where compared, it was found  64\% of residential electricity demand is satisfied by an individual battery and only 36\% by a communal battery.
+While the battery location was explored, the PV system location was not.
+This is a great example of the strength of energy system models, whereby the residential PV systems affect wind energy investment.
+However, a weakness in the model is the ability to comprehend household occupancy behaviour with and without household batteries on the PV system.
 
@@
 
-% NOT INCLUDE HAVE ENOUGH Petrovic and Karlsson \cite{Petrovic2016ResidentialSystem}  examined a new approached to represent residential heat pumps in Denmark
-
 \subsubsection{Ireland: Residential Modelling}
@@
 
-ESOM models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem,Connolly2021AalborgEnergy-System}, however few Irish studies use ESOM models to explore the residential sector, two examples of Irish ESOM focusing on the residential sector are detailed here.
+ESOM models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem,Connolly2009DevelopingUniversity.}, however few Irish studies use ESOM models to explore the residential sector, two examples of Irish ESOM focusing on the residential sector are detailed here.
 
@@
 Clinch and Healy analysed different aspects of thermal comfort in the Irish residential sector.
-This was the first comprehensive analysis of the residential sector, and their studies examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure} and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2000DomesticFailure,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}.
-In 2000, Clinch and Healy developed and published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Microsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating.
+This was the first comprehensive analysis of the Irish residential sector, and their studies examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure} and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}.
+In 2000, Clinch and Healy published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Microsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating.
 A cost-benefit model (CBM) was fitted to the EAM to facilitate the conversion of the physical estimates into monetary amounts.
 Under the predicted scenario, energy savings easily outweighed programme costs, representing a benefit-cost ratio of 1.7.The benefits of reduced sickness, hospitalisation, drugs, and mortality were quantified and this moved the  benefit-cost ratio up to 3.0.
-The same model was used by Clinch, Healy, and King \cite{Clinch2001ModellingEfficiency} in addition three models of householder spending on energy are developed here to to evaluate the human behavioural component
-into the computer model.
+The same model was used by Clinch, Healy, and King \cite{Clinch2001ModellingEfficiency} in addition three models of householder spending on energy are developed here to to evaluate the human behavioural component into the computer model.
 The modelling process involved developing a computer model of the energy performance of the existing dwelling stock.
@@
 This shows the the type of retrofit and type of house can have a large affect on the energy saved.
-Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F, and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is consideorange negligible.
+Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F, and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is negligible.
 
@@
 \item {Core Structural Model of the Irish Economy (COSMO)  is a macro-economic model used by the Economic and Social Research Institute (ESRI) to create a baseline forecast of energy consumption by sector} \cite{Bergin2017COSMO:Ireland}
-\item{PLEXOS is a power systems modeling tool used for electricity modeling and planning}\cite{PLEXOSExemplar}
-\item{Energy-efficiency policy analysis tool informs the energy efficiency estimates} \cite{Scheer2015UnlockingTob}
+\item{PLEXOS is a power systems modeling tool used for electricity modeling and planning}\cite{2021PLEXOSExemplar}
+\item{Energy-efficiency policy analysis tool informs the energy efficiency estimates} \cite{Scheer2015UnlockingOpportunity}
 \begin{itemize}
@@
 This is largely down to Ireland’s higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings, and rural settlement patterns.
-This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}.
-The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of CO\textsubscript{2} emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousingGov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx.
+This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{Eurostat2021DistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{Eurostat2021ShareSurveyilc_lvho50a}.
+The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of CO\textsubscript{2} emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousing2021Gov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx.
 85\% of residential buildings are 2005 or older.
-Furthermore, Ireland's fossil fuel dependency is underpinned by home heating oil, accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.
+Furthermore, Ireland's fossil fuel dependency is underpinned by home heating oil, accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020EnergySEAI}.
 
 In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential Greenhouse Gas (GHG) emissions.
-GHG emissions from coal in the residential sector also declined to 6.9\% in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
-However, the decline in peat and coal has been offset by a more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
-While, gas availability is limited to some urban areas and accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+GHG emissions from coal in the residential sector also declined to 6.9\% in 2018 \cite{Howley2005Energy-RelatedIreland.}.
+However, the decline in peat and coal has been offset by a more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005Energy-RelatedIreland.}.
+While, gas availability is limited to some urban areas and accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005Energy-RelatedIreland.}.
 The residential sector relied on fossil fuels for over 90\% of thermal energy in 2018, consequently Ireland currently has the lowest renewable heat among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
 Preliminary results from the Environmental Protection Agency (EPA) have indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.
+
+Space heating is the largest residential energy service, which accounted for 61.8\% of final energy usage in 2018.
+Other residential energy services and their share of final energy are shown in Fig.\ref{fig:Ireland2018ServiceDemands}.
+As building thermal efficiency and heating technology efficiency improves, both space and water heating energy would be expected to decline.
+
+\begin{figure}[!htbp]
+\centering
+    \begin{tikzpicture}
+    \tikzstyle{every node}=[font={\bf}]
+    \pie[   color = { yellow!90!black, green!60!black, blue!60,
+        orange!70,
+        gray!80,
+        teal!20
+    }, explode = 0.6, text=legend, radius=2.4]{61.8/Space Heating, 2.2/Cooking, 18.5/Water Heating, 3.8/Lighting, 12.8/Appliances,0.9/Pumps \& Fans}
+    \end{tikzpicture}
+    \caption{Ireland 2018 Energy Service Demands}
+    \label{fig:Ireland2018ServiceDemands}
+\end{figure}
 
@@
     \end{tikzpicture}
-    \caption{Ireland 2018 GHG Share \cite{Howley2005ENERGY-RELATEDIreland}}
+    \caption{Ireland 2018 GHG Share \cite{Howley2005Energy-RelatedIreland.}}
     \label{fig:Ireland2018GHG}
@@
 These targets account for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to make sectoral five-year carbon budgets a legal requirement to aid progress of the forementioned long-term targets.
-\par The EU's first Nationally Determined Contribution (NDC), which had a 40\% GHG reduction target (first NDC), has been succeeded by a new target of at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}.
+\par The EU's first Nationally Determined Contribution (NDC), which had a 40\% GHG reduction target (first NDC), has been succeeded by a new target of at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{2020SUBMISSIONSTATES}.
 Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
@@
 RESS supports renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target.
-% Almost 800 MW of solar has been approved in RESS-1 \cite{Eirgrid2020RenewableResults}, this greatly exceeds the current solar capacity of 16 MW, while making significant progress towards the CAP2019 target of 1500 MW solar by 2030.
+
 \par Sustainable Energy Authority of Ireland (SEAI) has a key role in the rollout of retrofitting and heat pumps in existing homes.
@@
 Non-archetype energy services fill in the lack of data on archetype energy services.
-The six non-archetype energy services are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service, this bottom-up data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA (Policy Oriented Tool for Energy and Climate Change Impact Assessment) database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{LitR:Lighting,Appliances}.
+The six non-archetype energy services are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service, this bottom-up data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{LitR:Lighting,Appliances}.
 The four archetype energy services are dependent on  building archetypes and data is sourced from the BER database, which is discussed in Section \ref{LitR:StockData}.
@@
 
-Energy service demands in the residential sector are driven by the total dwelling stock for non-archetype energy services and apartment, attached or detached dwelling stock for archetype energy services, which in turn is driven by population growth and the occupancy rate (the average number of people per dwelling).
-The residential stock projections up to 2040 are taken from Economic and Social Research Institute's (ESRI's) housing demand estimates \cite{BerginREGIONALLEVEL}.
-However, there is archetpye specfic growth is limited or unavailable, for this reason a simple linear regression is used to calculate archetype growth.
-Table \ref{Table:DwellingGrowth} shows the growth rate for historical years (2016-2020) and the future years (2020 onwards).
+The technical bottom-up development approach of the residential sector is combined with other economic sectors to meet macroeconomic defined demands.
+The energy service demands in the residential sector are driven by the total dwelling stock for non-archetype energy services and apartment, attached or detached dwelling stock for archetype energy services.
+The residential stock projections up to 2040 are taken from Economic and Social Research Institute's (ESRI's) housing demand estimates \cite{BerginREGIONALLEVEL}, beyond 2040 the growth rate is extrapolated.
+However, archetpye specfic growth is limited or unavailable, for this reason a simple linear regression, shown in Eq.\ref{Eq:RegArchetype} is used to calculate the share of an archetype in a given year (y).
+Figure \ref{fig:DwellingStoc} graphically represents the number of existing and new dwellings per archetype.
 The growth rate of apartments is the highest and increasing up to 2040, while attached and detached dwellings are decreasing, indicating a preference for smaller dwellings to increase in the future.
-
-\begin{table}[htbp]
-  \centering
-  \footnotesize
-  \caption{Dwelling Growth}
-\begin{tabular}{l|lcc}
-Years & Apartment & \multicolumn{1}{l}{Attached } & \multicolumn{1}{l}{Detached} \\ \hline
-2016-2020  & 2.89\%   & 1.69\%     & 1.58\%     \\
-\hline
-2020-2030     & 4.26\%   & 1.15\%     & 0.81\%   \\
-2030-2040     & 3.33\%   & 0.89\%     & 0.53\%   \\
-2040-2050     & 2.47\%   & 0.52\%     & 0.14\%   \\
-\end{tabular}
-\label{Table:DwellingGrowth}
-\end{table}
+The obsolescence rate of existing dwelling is 0.35\%.
+
+\begin{equation} 
+\begin{split}
+Apt(\%) =  7.5\ln{(y)} -56.97 \\
+Att(\%) = -2.53\ln{(y)} + 19.68 \\
+Det(\%) = -4.93\ln{(y)} + 37.91
+\end{split}
+\label{Eq:RegArchetype}
+\end{equation}
+
+\begin{figure}[!htbp]
+ \centering
+ \caption{Number of dwellings by type and age ('000)}
+ \includegraphics[scale=0.42]{Figures/TIM_DwellingStock.png}
+ \label{fig:DwellingStoc}
+\end{figure}
+
+%\begin{table}[htbp]
+%  \centering
+%  \footnotesize
+%  \caption{Dwelling Growth}
+%\begin{tabular}{l|lcc}
+%Years & Apartment & \multicolumn{1}{l}{Attached } & %\multicolumn{1}{l}{Detached} \\ \hline
+%2016-2020  & 2.89\%   & 1.69\%     & 1.58\%     \\
+%\hline
+%2020-2030     & 4.26\%   & 1.15\%     & 0.81\%   \\
+%2030-2040     & 3.33\%   & 0.89\%     & 0.53\%   \\
+%2040-2050     & 2.47\%   & 0.52\%     & 0.14\%   \\
+%\end{tabular}
+%\label{Table:DwellingGrowth}
+%\end{table}
+
 Archetype energy service demands are dependant on both the type of dwelling and whether it is existing or new.
-The average archetype energy service in new and existing is shown in Table \ref{Archetype Energy Service Demands}
+The average archetype energy service demand in new and existing is shown in Table \ref{Archetype Energy Service Demands}.
+
 \begin{table}[htbp]
@@
 
-Space and water heating, and lighting energy service demands for existing dwellings are the average values of the base year dwelling stock, and as all new dwellings are required to be A-rated, the new is the average for the existing A-rated dwellings.
-The Pumps \& Fans are similar, except in the new dwellings the value is halved and the remaining half is added to space cooling which is not separately defined in the base year dwelling stock.
-This is an assumption which was cross-checked with POTEnCIA space cooling projections per household.
-
-While the energy service demand changes, so does the number of dwellings by archetype and age. The existing dwellings decrease the obsolescence rate of 0.35\%
-1.7 million in 2018, to 2.1 million, 2030 to almost 2.6 million in 2050
-Share of new to old 23\ of buildings in 2030 will be new and 41\% in 2050.
-Apartment share moves dramitically from 12\% to 24\%. Detached 43\% to 35\% and Attached 45\% to 41\% 
-%%PICK UP HERE
-\begin{figure}[!htbp]
- \centering
- \caption{Number of dwellings by type and age ('000)}
- \includegraphics[scale=0.42]{Figures/TIM_DwellingStock.png}
- \label{fig:DwellingStoc}
-\end{figure}
-
- The stock is expected to increase by 40\% from 2018 level with a CAGR of 2\%.
-This results in an average of 27,600 new houses per annum between between 2021-2040.
-Beyond 2040, the population is used as a driver to project housing stock.
-The total housing stock obtained in 2050 is 2.57 million, which implies 8\% increase from 2040.
-% Table generated by Excel2LaTeX from sheet 'Sheet1'
-\begin{table}[htbp]
-  \centering
-  \footnotesize
-  \caption{Number of dwellings by type ('000)}
-    \begin{tabular}{rrrr}
-    \hline
-    \multicolumn{1}{l}{Year} & \multicolumn{1}{l}{Apartment} & \multicolumn{1}{l}{Attached } & \multicolumn{1}{l}{Detached} \\ \hline
-    2018  & 207   & 766   & 724 \\
-    2030  & 355   & 918   & 833 \\
-    2040  & 493   & 1003  & 878 \\
-    2050  & 629   & 1057  & 890 \\ \hline
-    \end{tabular}%
-  \label{tab:addlabel}%
-\end{table}%
-
-HDD and warmer weather to offset each other???
-
-ENERGY CONSUMPTION CHARACTERISTICS of House - Table?
-The higher thermal building efficiency in new dwellings is also expected to reduce energy running costs. The average existing space and water heating and water h
-% The technical bottom-up development approach of the residential sector is combined with other economic sectors to meet macroeconomic defined demands, to provide GHG targets insights both the energy system and residential sector.
-
-Heterogeneous Occupancy Behaviour > House Stock
-
-Residential Energy was >50\%higher before internal temperature adjustments
-Macroeconomic 
-no elasticity - fuel prices, and weather compensation?
-Time slices - 40 etc. 
-% textit{To accommodate an expanding population, the number of dwellings has also increased to currently stand at 1.7 million households, although this fgure remains signifcantly short of projected housing need, with annual average demand estimated to be up to 36,000 units for the next 30 years (IBEC 2018).
-Yet despite the number of new homes built to ever increasing energy performance standards, the Irish housing stock is among the poorest in Europe in terms of energy effciency (Goggins et al.
-2016).
-Current trends also show that households are getting bigger, with an average 15\% increase in foor area across all homes between 2000 and 2016 (SEAI 2018) \cite{FahyEnergyPractice}}
-%
-% The growth rate of the housing stock in the period to 2030 is obtained from the ESRI's macroeconomic estimates \citep{Yakut2020}. 
-% A multiple linear regression is calculated to project future housing stock based on population size and occupancy rate beyond 2030.
-The total housing stock is expected to increase by 79\% in 2050 relative to 2018 to accommodate the growing population and declining occupancy rates.
-The share of apartments changes from 12\% in 2018 to 24\% in 2050, while the shares of attached and detached dwellings decline by 4\% and 8\% respectively.
-
-% Energy service demands in the residential sector are driven by the size of the housing stock by type (apartment, attached or detached), which in turn is driven by population growth and the occupancy rate, the average number of people per dwelling.
-The occupancy rate is expected to fall to 2.12 in 2051 \citep{PropertyIndustryIreland2019}, from 2.75 in 2016 (CSO).
-Intermediate values are estimated using Equation \ref{eq:OR}.
-% \begin{equation}
-% \label{eq:OR}
-%  OR= a+b*log(\Delta t)
-% \end{equation}
-% here, $OR$ is the occupancy rate, $\Delta t$ is difference between years, parameters $a$ \& $b$ are estimated using historical values.
-
-Space heating is the largest residential energy service demand, accounted for 61.8\% of final energy usage in 2018.
-Other residential energy services and their share of final energy are shown in Fig.\ref{fig:Ireland2018ServiceDemands}.
-As building thermal efficiency and heating technology efficiency improves, both space and water heating energy would be expected to decline.
-The new A-rated dwellings can be used as a guide to show the share of future energy service demands.
-From the BER database, it can be seen that A-rated dwellings use less than half the space heating energy of the average dwelling and A-rated water heating energy decreases by about 40\% compared to the average A-rated dwelling.
-However Pumps \& Fans energy increases as mechanical ventilation is used over natural ventilation and lighting usage in A-rated is more than double the average household.
-Cooking and appliance energy usage would be expected to remain stable.
-
-\begin{figure}[!htbp]
-\centering
-    \begin{tikzpicture}
-    \tikzstyle{every node}=[font={\bf}]
-    \pie[   color = { yellow!90!black, green!60!black, blue!60,
-        orange!70,
-        gray!80,
-        teal!20
-    }, explode = 0.6, text=legend, radius=2.4]{61.8/Space Heating, 2.2/Cooking, 18.5/Water Heating, 3.8/Lighting, 12.8/Appliances,0.9/Pumps \& Fans}
-    \end{tikzpicture}
-    \caption{Ireland 2018 Energy Service Demands}
-    \label{fig:Ireland2018ServiceDemands}
-\end{figure}
-
-Notably absent from Fig.\ref{fig:Ireland2018ServiceDemands}, is space cooling, which is accounted for in appliances.
-The Policy Oriented Tool for Energy and Climate Change Impact Assessment (POTEnCIA) project disaggregated Ireland's residential space cooling from appliances and have output expected space cooling energy consumption projections are shown in Fig.\ref{fig:Cooling}.
-
-Demand TIMESLICE 40
+Space heating, water heating, and lighting energy service demands for existing dwellings are calculated from the average values of the base year dwelling stock, and as all new dwellings are required to be A-rated, the new dwellings are the average for the existing A-rated dwellings.
+The Pumps \& Fans are similar, except in the new dwellings the value is halved and the remaining half is added to space cooling.
+This is an assumption which was cross-checked with POTEnCIA (Policy Oriented Tool for Energy and Climate Change Impact Assessment) space cooling projections per household.
+
+%HDD and warmer weather to offset each other???
+%Heterogeneous Occupancy Behaviour > House Stock 
+%Residential Energy was >50\%higher before internal temperature adjustments
+%Macroeconomic 
+%no elasticity - fuel prices, and weather compensation?
+%Time slices - 40 etc.
 
@@
 %o	What can’t the model do? Where is it more suitable to use another model - Hannah Email
+Cost of buildings - bER ratings..
 BER Database - No PV Solar capacity, No Heat Pump ready, Large defaults, wall types check?)
@@
 Actual energy consumption for dwellings of
-similar or near identical construction may vary by up to 600%
+similar or near identical construction may vary by up to 600 \%
 depending on usage \cite{Sunikka-Blank2012BuildingConsumption}
@@
 \label{sec:Acknowledgements}
-Tomas %CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC
+Tomas
+Ankita%CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC
 % https://www.marei.ie/project/capacity/
@@
 
-%% If you have bibdatabase file and want bibtex to generate the
-%% bibitems, please use
-%%
 \newpage
@@
 
+%\bibliographystyle{elsarticle-harv}
+
+%\bibliography{references.bib}
+
 %% else use the following coding to input the bibitems directly in the

### 2021-07-26 manuscript.tex
Done for the day, rereading - at Ireland residential sector

@@
 
-\documentclass[preprint,12pt]{elsarticle}
+\documentclass[review]{elsarticle}
+
+\usepackage{lineno,hyperref}
+\modulolinenumbers[5]
+
+%% `Elsevier LaTeX' style
+%\bibliographystyle{elsarticle-num}
 %\documentclass[10pt,final,journal,compsoc,a4paper]{IEEEtran}
@@
 \usepackage{blindtext}
-\usepackage{lineno,hyperref}
-\modulolinenumbers[5]
+
 \DeclareSIUnit{\EUR}{\text{\euro}}
@@
 %\usepackage[table]{xcolor}
-\usepackage{biblatex}
+%\usepackage[backend=bibtex]{biblatex}
 
@@
 %% \usepackage{amsthm}
-\addbibresource{references.bib}
+%\bibliographystyle{elsarticle-num}
+%\usepackage{cleveref}
+%\usepackage{hyperref}
+%\usepackage[style=numeric, sorting=none]{biblatex}  %backend=biber is 'better'
+%
+\addbibresource{TIM_V11.bib}
 %% The lineno packages adds line numbers. Start line numbering with
@@
 \journal{Energy and Buildings}
+%\bibliographystyle{elsarticle-num}
 
@@
 \par The Irish residential sector has proved hard to decarbonise, yet recently Ireland has set one of the most ambitious climate targets into law, a 51 \% greenhouse gas reduction in 2030, compared to 2018.
-There is a need for a new state of the art model to provide insights on the optimal decarbonisation pathway for the Irish energy system.
-This paper reviews residential and energy system models and describes the methodology applied to the development of the residential sector in the new TIMES Ireland Model (TIM).
-The paper includes some results to provide a discussion on some of the insights and highlights the insights which may need to be further investigated in the residential sector.
+There is a need for a new up to date and state of the art model to provide insights on the optimal decarbonisation pathway for the Irish energy system.
+This paper describes the methodology applied to the development of the residential sector in the new TIMES Ireland Model (TIM).
+Results are analysed to provide input to the discuss insights from the residential sector in TIM.
 
@@
 
-Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways and interpreting the results.
-Section \ref{sec:Introduction} of this paper looks at Ireland's climate targets, energy policy, and distinctive residential sector.
+The layout of this paper is as follows: 
+\begin{itemize}
+    \item Section \ref{sec:Introduction} of this paper looks at Ireland's climate targets, energy policy, and distinctive residential sector.
 While also examining how researchers have previously used bottom-up residential models and Energy System Optimization Models (ESOMs) to analyse the residential sector.
-Section \ref{LitR} describes the methodology used to develop the residential sector.
+    \item Section \ref{LitR} is the core section of this paper and explains the methodology used to develop the residential sector.
 The treatment of input data in the model is described, the calculations are explained, and assumptions and constraints are highlighted.
-Section \ref{Results} outlines the six sceanrios used and displays the energy system and residential results.
-Section \ref{Discussion} discusses the results, strengths, and weakness of the newly developed model and how it differs from other models.
-Lastly, section \ref{Conclusion} draws some conclusions on the model, results and discussion on future work.
+    \item Section \ref{Results} outlines the scenarios used and displays results from the residential sector.
+    \item Section \ref{Discussion} discusses the results, strengths, and weakness of the newly developed model and how it differs from other models.
+    \item Lastly, section \ref{Conclusion} draws some conclusions and discusses possible future work.
+\end{itemize}
 
@@
  ESOMs have been used to explore the least cost pathways, as they consider a large number of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
-Some of the main ESOMs currently used include MESSAGE \cite{Messner1995Model-BasedPlanning}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOMs cover all energy-related sectors, they can also be used to specifically explore certain economic sector in higher granularity.
+Some of the main ESOMs currently used include MESSAGE \cite{Messner1995Model-basedPlanning}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOMs cover all energy-related sectors, they can also be used to specifically explore certain economic sectors in higher granularity.
+This section gives some examples of residential sector studies using ESOMs.
 
 TIMES has been extensively used in both the UK and Ireland, and there are several examples using TIMES to focus on the residential sector.
-TIMES assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
+TIMES assumes perfect foresight to enable a least-cost solution to be calculated, but \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
 With this in mind, Li et al.
 \cite{Li2018IncorporatingModel} investigated the heating technology preference in the UK using UK TIMES Model (UKTM) by incorporating a Discrete Choice Model (DCM) to better reflect how homeowners choose heating technologies.
-The DCM was developed from a nationwide survey with 442 homeowners completing the survey.
-The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems.
-The DCM then estimated the probability of heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices, this identifies perfect foresight as a weakness in TIMES.
-While Li's research has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights through more surveys and analysis.
- \par Many sectors are increasingly reliant on electricity generation as lower carbon intensity electricity is considered a cornerstone in optimal decarbonisation.
-Using the Open Source Energy Modeling System (OSeMOSYS) \cite{Howells2011OSeMOSYS:Development.}
-Leibowicz et al.\cite{Leibowicz2018OptimalServices} explores the optimal decarbonisation of the residential sector by examining electricity generation and storage strategies.
+A nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems.
+The DCM then estimated the probability of heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices.
+Despite the perfect foresight assumption being essential to finding a solution, it has the potential to be a weakness by widening the gap between the model and reality.
+\par A common theme across many ESOMs shows different sectors are increasingly reliant on electricity generation as lower carbon intensity electricity is considered a cornerstone in optimal decarbonisation.
+Using the Open Source Energy Modeling System (OSeMOSYS), Leibowicz et al.
+\cite{Leibowicz2018OptimalServices} explores the optimal decarbonisation of the residential sector by examining electricity generation and storage strategies.
 The results find net present costs in the \textit{``Policy scenarios"} ( $CO_2$ declines linearly by 90\% in 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations imposed ).
-However, the percentage increase in net present cost due to the policy diminishes with a more thermally efficient building stock, highlighting the importance of energy building regulation as thermally efficient building stock can reduce the decarbonisation cost.
-District Heating is a proven low carbon heating technology, however it is almost non-existant and has not been included in previous ESOM in Ireland.
-Denmark are leading experts in this area and Denmark's district heating system has been comprehensively examined.
-One study by
-Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} used the Balmorel model to analyse how heat pumps with complementing heat storage affect wind power investments in Denmark.
-Exogenous user constraints applied to the model, based on the current Danish policy.
+However, the percentage increase in net present cost diminishes with a more thermally efficient building stock, highlighting the importance of energy building regulation as thermally efficient building stock can reduce residential decarbonisation costs.
+\par District Heating is a proven low carbon heating technology, however it is rarely researched in Ireland and has not been included in previous ESOM in Ireland.
+However, Denmark's district heating system has been comprehensively examined.
+One study by Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} used the Balmorel model to analyse how heat pumps with complementing heat storage affect wind power investments in Denmark.
 The results find that air-to-water heat pumps are installed in non-district heating areas, while the system benefits of adding individual heat storage to the heat pumps are only moderate, the main benefit is a reduced need for peak/reserve capacity investments.
 This identifies individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments in areas where district heating is not available.
-Another technology which is proven but under utilised in Ireland, is photovoltaic (PV) solar panels.
-While Ireland's policy and building regulations have encouraged the uptake of PVs, however the market share of PV systems remains low, and as the cost continues to fall they can alter optimal decarbonisation pathways.
-Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model to explore the feasibility of PV systems installed at the household level in combination with batteries in Samsø.
+\par While Ireland's building regulations have encouraged the uptake of photovoltaics (PVs), the market share of PV solar panel systems remains low.
+The low amount of direct sunlight means Ireland's climate is not ideal for PVs but as the cost of PVs continues to fall, the potential role of PVs in  residential decarbonisation rises.
+In one study, Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model to explore the feasibility of PV systems installed at the household level in combination with batteries in Samsø.
 Two approaches are explored for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
@@
 
-%The Cambridge Housing Model (BEIS) ses English Housing Survey data, coupled to a SAP ( UK's BER equivalent ) energy calculator, to estimate energy use and CO2 emissions for all homes in England, broken down by final use.
-The CHM reads in the EHS dwelling for each case and performs building physics calculations to determine energy consumption and associated CO2 emissions, by use and by fuel type.
-Multiplying the energy use and CO2 emissions by the associated weighting and summing across all cases gives total values for England.
-Housing Data is cleaned, Climate data for region is 8 energy services - space heating ( main \& secondary ), water heating, space cooling, lighting, electrical appliances, cooking and pumps \& fan.
-6 fuels - gas, oil, solid, biomass, electriicty and renewable
-
 \subsubsection{Ireland: Residential Modelling}
@@
 
-ESOM models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem,Connolly2009DevelopingUniversity.}, however few Irish studies use ESOM models to explore the residential sector, two examples of Irish ESOM focusing on the residential sector are detailed here.
+ESOMs have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem,Connolly2009DevelopingUniversity.}, however few Irish studies use ESOM models to explore the residential sector, two examples of Irish ESOM focusing on the residential sector are outlined below.
 
 In 2014, Rogan et al.
-\cite{Rogan2014LEAPsSystem} linked OSeMOYOS, an ESOM, to the Long range Energy Alternatives Planning system (LEAP) simulation model, to explore specific energy efficiency policies not only in the residential sector, but also in the transport, industry and services sectors in Ireland.
-The bottom-up development of the residential sector in LEAP was combined with the cost optimisation capabilities of OSeMOYOS.
-Three scenarios are exploorange, namely, the reference scenario, energy efficiency scenario, and energy efficiency+ scenario.
-The reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020.
-In the reference scenario, the residential total energy consumption rose 6.7 \% in 2020, compared to 2008.
-The share of space and water heating decreased from 82 \% to 77 \% in that period.
+\cite{Rogan2014LEAPsSystem} linked OSeMOYOS to the Long Range Energy Alternatives Planning system (LEAP) simulation model to explore specific energy efficiency policies not only in the residential sector, but also in the transport, industry and services sectors in Ireland.
+Three scenarios were explored, namely, the reference scenario, energy efficiency scenario, and energy efficiency+ scenario.
+Where the reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020 and the results show that residential total energy consumption rose 6.7 \% in 2020, compared to 2008.
 In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulting in a 9.9 \% reduction of energy consumption in 2020, compared to the reference scenario.
 The energy efficiency+ scenario also allows for 800,000 residential dwellings to be retrofitting, but accounts for deeper retrofits.
-This results in an energy consumption reduction of 16.5 \% in 2020, compared to the reference scenario.Rogan et al.
+This results in an energy consumption reduction of 16.5 \% in 2020, compared to the reference scenario.
+Rogan et al.
 discussed the barriers, obstacles to investment, split incentives, and information gaps which restrict energy efficiency improvements in the residential sector.
@@
 The previous version of Irish TIMES did not account for district heating, so EnergyPLAN is used to investigate the operation and  feasibility of district heating in Ireland.
-A sensitivity analysis was performed on investment costs, heat grid losses, and discount rates, however in all cases the inclusion of district heating is optimal.
-The final results find that district heating compared fuel consumption by 3.5 \% and the conversion of 37 \% of the Irish heat demand to district heating is found to be optimal.
+A sensitivity analysis was performed on investment costs, heat grid losses, and discount rates.
+In all cases, the inclusion of district heating was found to be favourable.
+The results find the conversion of 37 \% of the Irish heat demand to district heating to be optimal.
 Both the TIMES model and climate target are superseded and the study did not include the future heat savings expected from retrofitting, nevertheless this study has provided insights on the potential of district heating.
@@
 \par  The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector models, rather than energy system models specifically examining the residential sector.
-Nevertheless, the residential models have provided insights on the input data and methodology approach used in Ireland, some of which are outlined below.
+Nevertheless, the residential models have provided insights on the input data, methodology, and results, and some studies are outlined below.
 
@@
 In 2000, Clinch and Healy published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Microsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating.
-A cost-benefit model (CBM) was fitted to the EAM to facilitate the conversion of the physical estimates into monetary amounts.
-Under the predicted scenario, energy savings easily outweighed programme costs, representing a benefit-cost ratio of 1.7.The benefits of reduced sickness, hospitalisation, drugs, and mortality were quantified and this moved the  benefit-cost ratio up to 3.0.
-The same model was used by Clinch, Healy, and King \cite{Clinch2001ModellingEfficiency} in addition three models of householder spending on energy are developed here to to evaluate the human behavioural component into the computer model.
-The modelling process involved developing a computer model of the energy performance of the existing dwelling stock.
-It was noted that the energy consumption predicted by the model is considerably higher than given in the national energy balance, due to internal temperature assumptions.
-The model showed that the energy efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compared to the reference scenarios.
-
+A cost-benefit model (CBM) was fitted to the EAM and the results under the predicted scenario, showed energy savings outweighed programme costs, representing a benefit-cost ratio of 1.7.
+The benefits of reduced sickness, hospitalisation, drugs, and mortality were later quantified and this moved the benefit-cost ratio up to 3.0.
+The same model was later used by Clinch, Healy, and King \cite{Clinch2001ModellingEfficiency} to represent the human behavioural component in the computer model.
+
+\par
 In 2011, Dineen and Ó Gallachóir \cite{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP).
@@
 The building fabric efficiency of newly occupied dwellings developed takes into account the theoretical increase in energy efficiency due to improved 2008 and 2010 building regulations.
-Space heating demand accounted for 70\% of household energy consumption in newly built dwellings in 1997, down to just under 60\% in 2006 and the model predicts a reduction to 27\% by 2015.The paper acknowledges retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperatures affect on energy efficiency improvements.
-\par In 2013, Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures to the existing detached, oil centrally heated, rural housing stock.
+The paper acknowledges that retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperature effect on energy efficiency improvements.
+\par In 2013, Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures of the existing detached, oil centrally heated, rural housing stock.
 A base geometry according to age bands and thermal characteristics was created in the study.
-The heat load for a statistical occupancy and year was established.
 The default values used in some of the BER database variables where highlighted.
-For example,  there are assumed U-values when the roof insulation thickness is unknown, this is based on age, but there are no U-values for uninsulated roofs despite a 2001 survey found 18\% of detached housing in Ireland had no roof insulation.
+For example, there are assumed U-values when the roof insulation thickness is unknown, this is based on age, but there are no U-values for uninsulated roofs despite a 2001 survey finding 18\% of detached housing in Ireland had no roof insulation.
 Ahern et al.
 found dwellings built post-2000 consume the greatest amount of energy (kWh/annum), due to their large floor area.
-Concluding that the majority of energy efficiency progress for thermal use in Irish dwellings is being offset.
-The model thermal improvement measures resulted in a reduction of running costs and CO2 emissions by an average of 65\% for houses constructed prior to 1979, and this resulted in an average 12-year payback.
+Concluding that the majority of thermal energy efficiency progress in Irish dwellings is being offset.
+The model results show that thermal improvement measures resulted in a reduction of running costs and $CO_2$ emissions by an average of 65\% for houses constructed prior to 1979, and this resulted in an average 12-year payback.
 Houses built in 1979 or after resulted in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures.
+This study highlights the low hanging fruit of pre-1980 oil centrally heated, rural housing stock.
 
@@
 The DEAP model used 175 archetype dwellings ( 5 building types, 7 energy bands, and 5 wall construction types ) and 6 different retrofit measures were applied.
-BER database filters were used in this model, the BER filters from this study were updated and applied in TIM.
-A sensitivity analysis is performed and it was found that there was approximately a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift.
+The BER filters from this study were updated and applied in TIM.
+A sensitivity analysis performed found that there was a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift.
 Concluding, BER standard assumptions overestimate dwelling energy consumption and that the average internal temperature across the Irish dwelling stock may be closer to 18$^{\circ}$C than 21$^{\circ}$C.
 
-The Archetype Dwelling Energy Model (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}.
-ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and then calculates the annual energy consumption to simulate the potential for improved energy efficiency gains within the existing retrofit programs.
+\par The Archetype Dwelling Energy Model (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}.
+ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and then calculates the annual energy consumption using EN 13790 to simulate the potential for improved energy efficiency gains within the existing retrofit programs.
 Mac Uidhir et al.
@@
 This shows the the type of retrofit and type of house can have a large affect on the energy saved.
-Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F, and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is negligible.
-
-Finally, it is worth mentioning SEAI’s National Energy Modelling Framework (NEMF), which is a suite of modelling tools used for modelling the Irish energy system.
+
+\par Finally, it is worth mentioning Sustainable Energy Authority of Ireland's (SEAI’s) National Energy Modelling Framework (NEMF), which is a suite of modelling tools used for modelling the Irish energy system.
+The SEAI is Ireland's national energy authority and they play a key role in assisting consumers to reduce energy.
 Some of the main models used in the NEMF include:
@@
 \item {Core Structural Model of the Irish Economy (COSMO)  is a macro-economic model used by the Economic and Social Research Institute (ESRI) to create a baseline forecast of energy consumption by sector} \cite{Bergin2017COSMO:Ireland}
-\item{PLEXOS is a power systems modeling tool used for electricity modeling and planning}\cite{2021PLEXOSExemplar}
+\item{PLEXOS is a power systems modeling tool used for electricity modeling and planning} \cite{2021PLEXOSExemplar}
 \item{Energy-efficiency policy analysis tool informs the energy efficiency estimates} \cite{Scheer2015UnlockingOpportunity}
-\begin{itemize}
+\begin{enumerate}[i.]
 \item{Consumer Choice Model determines the attractiveness of energy efficiency finance options}\cite{SustainableEnergyAuthorityofIreland2017BehaviouralSector}
-\end{itemize}
-\item{BioHEAT is a techno-economic simulation model of bioenergy and heat, used by SEAI to support policy decision in Ireland.}\cite{Durusut2018BioHEAT:Sectors}
+\end{enumerate}
+\item{BioHEAT is a techno-economic simulation model of bioenergy and heat} \cite{Durusut2018BioHEAT:Sectors}
 \end{itemize}
@@
 The models have varied objectives such as exploring consumer choices, PV battery system location, evaluating the health effects of retrofitting, and integration of bioenergy and heat.
-The thermal efficiency of dwellings is a common objective in Irish models, due to the poor thermal efficiency of the dwelling stock and the growing BER database is providing high data granularity input to these models.
-The internal temperature is also analysed in some of the studies, a recurring theme is the wide range of internal temperatures.
+The thermal efficiency of dwellings is commonly researched in Irish models, due to the poor thermal efficiency of the dwelling stock and the growing BER database is providing high data granularity input to these models.
+The internal temperature is becoming more frequently analysed in the studies, a recurring theme is the wide range of internal temperatures.
 The bottom-up residential models are typically simulation models looking at a specific policy.
-The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switching to Microsoft SQL.
+The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switching to Microsoft SQL, with much higher computational power.
 There has been few studies on the Irish residential sector using ESOMs, as residential bottom-up models are preferred when focusing on that sector.
@@
 The use of ESOM to explore DH in Ireland allows for insights which bottom-up simulation models cannot provide.
-The connection between waste heat in electricity and industry can show the optimal solution for district heating in commercial and residential sectors.
+The connection between waste heat in electricity and industry can provide better insights on the optimal solution for DH in commercial and residential sectors.
 
@@
 \label{Intro:IreRSD}
-
+%PICKUPHERE
 Ireland’s distinctive energy sectors are further separated from other EU member states when focusing on the residential sector.
-This is largely down to Ireland’s higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings, and rural settlement patterns.
+This is largely down to higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings, and rural settlement patterns in Ireland's residential sector.
 This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{Eurostat2021DistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{Eurostat2021ShareSurveyilc_lvho50a}.
-The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of CO\textsubscript{2} emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousing2021Gov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx.
+The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of $CO_2$ emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousing2021Gov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx.
 85\% of residential buildings are 2005 or older.
@@
 However, the decline in peat and coal has been offset by a more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005Energy-RelatedIreland.}.
-While, gas availability is limited to some urban areas and accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005Energy-RelatedIreland.}.
+While, gas availability is limited to some urban areas and gas still accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005Energy-RelatedIreland.}.
 The residential sector relied on fossil fuels for over 90\% of thermal energy in 2018, consequently Ireland currently has the lowest renewable heat among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
@@
 
-Under 2006 Intergovernmental Panel on Climate Change (IPCC) guidelines, Ireland’s residential sector accounted for 14.2\% of GHGs emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.
+Under 2006 Intergovernmental Panel on Climate Change (IPCC) guidelines, Ireland’s residential sector accounted for 14.2\% of GHGs emissions in 2018 \cite{Howley2005Energy-RelatedIreland.}.
 Focusing only on energy-related GHG emissions, the residential sector directly accounts for 24\% of GHG emissions.
@@
 
-Where the residential sector is the direct source 47\% of heat and 30\% of electricity GHG emissions, the residential sector is therefore a pivotal sector in decarbonising energy in Ireland.
+Where the residential sector is directly responsible for 47\% of heat and 30\% of electricity GHG emissions, the residential sector is therefore a pivotal sector in decarbonising energy in Ireland.
 
@@
 These targets account for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to make sectoral five-year carbon budgets a legal requirement to aid progress of the forementioned long-term targets.
-\par The EU's first Nationally Determined Contribution (NDC), which had a 40\% GHG reduction target (first NDC), has been succeeded by a new target of at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{2020SUBMISSIONSTATES}.
+\par The EU's first Nationally Determined Contribution (NDC), which had a 40\% GHG reduction target (first NDC), has been succeeded by a new target of at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels.
 Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
@@
 \label{Intro:Policy}
-Ireland has set out national policy objectives under 10 strategic investment priorities \cite{Ireland2018ProjectFramework}.
-The first strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}.
-The densification of Ireland's main cities, will provide opportunities to improve the efficiency of heating, electricity and transport within and between cities.
-\par Specifically focusing on the energy sector, the Marginal Abatement Cost Curve (MACC) produced in CAP2019  \cite{EPA2018EPAReport} found electricity \& transport are the most cost-effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
-CAP2019 findings show the most cost-effective solution to decarbonising the residential sector, is to receive higher amounts of low carbon electricity.
-
-%The CAP2019 MACC, shown in Fig.\ref{fig:CAP2019:MACC} provides $>300$ of the most cost-effective technologies options to orangeuce Non-ETS GHG by 30\% in 2030 compared to 2005 levels. 
-%The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_{eq}$.
-The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_{eq}$ over the 2021 to 2030 period.
-The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_{eq}$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-%The options specific to the residential sector are:
-
-%\begin{itemize}
-%    \item{Retrofit oil boiler existing dwellings to B2 equivalent}
- %   \item{Switch from oil boilers to heat pumps in existing dwellings}
- %   \item{Retrofit gas boiler and solid fuel stove existing dwellings to B2 equivalent}
- %   \item{Introduce CO2-free heating in new buildings}\cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
-%\end{itemize}
-
-%\begin{figure}[!htbp]
- %\centering
- %\includegraphics[scale=1]{Figures/CAP2019MACC.png} 
- %\caption{Climate Action Plan 2019 - Marginal Abatement Cost Curve}
-% \label{fig:CAP2019:MACC} \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
-%\end{figure}
-%For example, Figure \ref{fig:CAP2019:MACC} shows the retrofitting of oil boilers is expected to orangeuce the Total Cost of Ownership (TCO) (ie.
-save money), but only abate a small amount of carbon, whereas switching oil boilers to heat pumps will increase TCO (ie.
-cost money) but the expected abatement of carbon is far greater.
-
-In Ireland, the decarbonisation of electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}.
+The Irish government's long-term overarching strategy aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}\cite{Ireland2018ProjectFramework}.
+The densification of Ireland's main cities will provide opportunities to improve the efficiency of heating, electricity, and transport within and between cities.
+\par The Marginal Abatement Cost Curve (MACC) produced in CAP2019, shows electricity \& transport are the most cost-effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}, therefore the most cost-effective residential decarbonisation is to receive higher amounts of low carbon electricity.
+
+\par In Ireland, the decarbonisation of electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005Energy-RelatedIreland.}.
 Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private investment towards low carbon or renewable energy.
@@
 \par Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019VisionIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-GNI anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}.
-The gas network currently reaches 40\% or 680,000 dwellings in urban areas, but with the ``Compact Growth" policy strategy a higher share of building may have access to the gas network in the future.
+GNI anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020EnergySEAI}.
 A major barrier to increasing biomethane from an agricultural source in the gas network is the cost, it was the most expensive option in the Marginal Abatement Cost Curve (MACC) produced in CAP2019.
 \par Ireland has one of the lowest shares of District Heating (DH) in Europe at less than 1\%  \cite{Gartland2016AIreland}.
-The ``Compact Growth" policy strategy and advancement in DH control will strengthen the case for DH networks to play a part in decarbonisaing the residential sector.
+Advancement in DH control and awareness of the technology will strengthen the case for DH networks to play a part in decarbonisaing the residential sector.
 CAP2019 plans to have approx.
-0.4\% of heating provided by DH in 2030, research using TIMES and EnergyPLAN has shown that 37\% of heating provided by DH is optimal if Ireland is to orangeuce GHG emissions by 80\% in 2050, compared to 1990 \cite{Thellufsen2019ImplementingIreland}.
+0.4\% of heating provided by DH in 2030, research using TIMES and EnergyPLAN has shown that 37\% of heating provided by DH is optimal if Ireland is to reduce GHG emissions by 80\% in 2050, compared to 1990 \cite{Thellufsen2019ImplementingIreland}.
 While Irish District Energy Association (IrDEA) suggest 57\% of heat can be sourced from DH, if supporting policy and regulation was in place.
-District Heating will likely play a role, however Gartland points out a range of organisational, technical, regulatory, and economic barriers to DH growth in Ireland \cite{Gartland2016AIreland}.
-\par While it maybe optimal to use Hydrotreated Vegetable Oil (HVO) in transport, it is also consideorange in the residential sector, as it can provide a cheap alternative for the 700,000 Irish dwellings using home heating oil ( mostly kerosene).
-HVO, which meets the EU sustainability criteria, reduces GHG emissions by 60\% as compared to fossil fuels \cite{Soam2019FactorsSweden}.
-The capacity of HVO is growing in Europe, but HVO is currently about twice the price of kerosene.
-HVO can be used to replace kerosene, albeit at a cost of up to €400 per oil boiler, however there is no policy for HVO in the residential sector.
+District Heating will likely play a role, however, Gartland points out a range of organisational, technical, regulatory, and economic barriers to DH growth in Ireland \cite{Gartland2016AIreland}.
 
@@
 The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.
-There is a need for a new ESOM to be built specifically to explore Ireland's new 2030 and 2050 climate targets and sectoral five-year carbon budgets, the new ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
+The introduction of new climate policies and targets combined with technological advancements, means there is a need for a new ESOM to be built to explore Ireland's new climate targets and sectoral five-year carbon budgets, the new ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
 \par TIMES is a bottom-up optimisation energy-environment model with perfect foresight which is used to analyse various levels of spatial, temporal and sectoral resolution.
@@
 \end{equation}
+
 Where, $NPV$ is the net present value of the total cost for all regions ; $ANNCOST(r,y)$ is the total annual cost in region r and year y; $d_{r,y}$ is the general discount rate; $REFYR$ is the reference year for discounting; YEARS is the set of years for which there are costs; and $R$ is the set of regions in the area of study.
-
-TIM is an optimisation model of the Irish energy system, which calculates the cost-optimal fuel and technology mix to meet future energy service demands.
-The residential, power generation and transport sectors are rebuilt with high granularity from the bottom-up, while industry, agriculture, and services sectors have been updated.
-The model operates within the limits of GHG constraints, resource constraints, and feasible deployment rates.
 
@@
 \label{LitR:RES}
-The new structure and updated technological and macroeconomic data will provide more robust insights on sectoral five-year carbon budgets towards 2030 and 2050 than previous models.
-The residential sector was developed in parallel with other sectors in TIM, and concurrently with the new Irish LEAP model, which allows for more accessible soft-linking to provide more robust analytical insights by combining simulation and optimisation, and to exchange expertise between modelling teams.
-Fig.\ref{fig:RES} illustrates the flow of energy from fuel to demand in the residential sector.
+The residential sector was developed in parallel with other sectors in TIM, and concurrently with the new Irish LEAP model, which allows for more accessible soft-linking to provide robust analytical insights by combining simulation and optimisation, and to exchange expertise between modelling teams.
+Fig.\ref{fig:RES} illustrates the flow of energy from fuel to demand in the TIM residential sector.
 
@@
  \centering
- \includegraphics[scale=0.45]{Figures/TIM_RES3.png}
- \caption{Residential Sector - Reference Energy System}
+ \includegraphics[width=\textwidth]{Figures/TIM_RES3.png}
+ \caption{TIM Residential Sector - Reference Energy System}
  \label{fig:RES}
@@
 The BER database provides the base year bottom-up energy consumption data, this is scaled up to represent the entire stock and cross-checked with the top-down energy consumption from the 2018 SEAI Energy Balance.
-To compare from the two databases, fuels were were combined and condensed down to 13 fuels in TIM.
+To compare the two databases, fuels were were combined and condensed down to 13 fuels in TIM.
 Table \ref{Table:Fuels} shows an overview of the fuels combined from the different data sources.
 
+\par In TIM, fuels such as coal, represent a combination of fuels from the SEAI Energy Balance and BER database.
+Gas oil which is accounted for in the SEAI Energy Balance, has similar properties to kerosene and there was no evidence to support keeping separate fuels would provide further insight into residential decarbonisation.
+However, dissagreatting ambient heat could potentially provide insights to residential decarbonisation.
+The solid multi-fuel from the BER database represents an open fire or stove which is heated by either coal, peat, or wood.
+Through examining the SEAI Energy Balance and BER database, the fuel share assigned for coal, peat, and wood is 19.4 \%, 74.4 \% and 6.2 \% respectively.
+\par There are two types of energy services represented in Table \ref{fig:RES} - Non-Archetype and Archetype.
+Non-archetype energy services fill in the lack of data on archetype energy services.
+The six non-archetype energy services are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service, this bottom-up data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{LitR:Lighting,Appliances}.
+The four archetype energy services are dependent on  building archetypes and data is sourced from the BER database, which is discussed in Section \ref{LitR:StockData}.
+The internal temperature adjustment for space heating is detailed in Section \ref{LitR:StockData}.
+The demand for archetype and non-archetype energy services is based on the growth of residential dwellings, which is detailed in section \ref{LitR:Demands}.
+
+\subsubsection{Dwelling Stock Data}
+\label{LitR:StockData}
+
+The dwelling stock data is obtained mainly from the SEAI BER Public database, which is firstly filteorange, then scaled up using Central Statistics Office (CSO) data. 
+The BER database is Ireland's national Energy Performance Certificate (EPC) register, which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD).
+The BERs are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
+Over half of Irish dwellings are captuorange in the BER database, every BER displays over 200 energy related variables.
+The DEAP model, which was outlined by Dineen et al.\cite{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM} provides the results for the energy rating in the BER database.
+However, the space heating calculation is used to estimate the expected energy consumption per dwelling, which makes some assumptions.
+DEAP assumes that each dwelling is single zoned, the heating season is from October to May inclusive, and the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season.
+One assumption, the internal temperature assumption, results in a large error, where space and water heating are overestimated in DEAP.
+Space heating consumption is reduced by 26 \% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance.
+DEAP assumes all living areas ( living room, sitting room, and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C.
+Through the analysis of previous residential modelling in Ireland this is not the case, as the internal temperatures are often much lower, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions applied in TIM.
+
+\begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Internal Temperature Assumptions}
+    \begin{tabular}{rrr}
+    \hline
+    \multicolumn{1}{l}{BER Rating} & \multicolumn{1}{l}{Living Area} & \multicolumn{1}{l}{Non-Living Area }  \\ \hline
+    A  & 23$^{\circ}$C   & 20$^{\circ}$C  \\
+    B  & 21$^{\circ}$C   & 18$^{\circ}$C  \\
+    C  & 18$^{\circ}$C  & 15$^{\circ}$C  \\
+    D  & 18$^{\circ}$C   & 15$^{\circ}$C  \\
+    E  & 18$^{\circ}$C   & 15$^{\circ}$C  \\ 
+    F  & 18$^{\circ}$C   & 15$^{\circ}$C  \\ 
+    G  & 18$^{\circ}$C   & 13$^{\circ}$C  \\ \hline
+    \end{tabular}
+  \label{Table:InternalTemps}
+\end{table}
+
+The ArDEM-SQL, which was described in Section \ref{Intro:ESOMIreland} is a simulation model built on EN 13790.
+ArDEM-SQL alters the primary energy consumption of the BER database by changing the internal temperatures.
+
+Other concerns for uncertainties in the database, which are not related to DEAP calculation are statistically anomalous spikes in the data, caused by applying default values \cite{Ahern2020ADatabases} and different companies generating variations of ±20\% in the assessment of energy consumption \cite{Harsman2016OnCertificates}.While the BER database provides high granularity data, it must be filteorange to improve the representation of the dwelling stock data.
+The filters applied to the raw BER database are outlined in \ref{sec:appendixA}.
+
+The Census 2016 data \cite{CentralStatisticsOffice2016CensusIreland} from the CSO is also used in calculating the dwelling stock.
+Dataset E1005 provides a split of seven archetypes per county, which are aggregated to form three archetypes.
+The total number of building archetypes per county in 2016 is used to scale up the BER database.
+The gap in total dwellings between the census year in 2016 and the model base year in 2018 was initially ignoored  due to lack of data.
+However, later it was found that data on new dwelling electrical connections per archetype and per county are available from CSO dataset NDQ06, this would have filled the gap, the first iteration of the dwelling stock has a 2 \% error.
+
+Once both databases were finalised, the next step in this process was to scale up the BER database.
+
+This was done by first finding the ratio of the building archetype per county between the BER database and CSO data, as shown in Equation \ref{Eq:BERtoCSO1}.
+
+\begin{equation} 
+\begin{split}
+R_{a,c} = {\frac{CSO_{a-nsa,c}} {BER_{a,c}} + \frac{CSO_{nsa,c}} {CSO_{a,c}}}
+\end{split}
+\label{Eq:BERtoCSO1}
+\end{equation}
+
+$R_{a,c}$ is the ratio per archetype (a) and per county (c), there are 26 counties and 3 archetypes.
+$CSO_{a-nsa,c}$ represents the number of dwelling archetypes minus the non-stated archetypes (nsa) per county in the CSO database.
+$BER_{a,c}$ is the number of archetypes per county  in the BER database.
+The next step is to multiply the number of BERs per archetype and per county by the relevant ratio.
+\begin{equation} 
+\begin{split}
+BER_{a,c} * R_{a,c} = TIM_{a,c}
+\end{split}
+\label{Eq:BERtoCSO2}
+\end{equation}
+
+$TIM_{a,c}$ represents the final dwelling stock data per archetype and per county, the first iteration of TIM does not disaggregate the dwellings per county, instead a national total per archetype is input to TIM ( $TIM_a$ ).
+
+Once the final dwelling stock is complete, the next step is to assign a BER rating to all dwellings.
+This was done by comparing 9 different dwelling age ranges for each of the 3 archetypes between the BER and CSO datasets.
+
+\begin{equation} 
+\begin{split}
+TIM_{r,y,a} = CSO_{y,a} * \frac{BER_{r,y,a}} {BER_{y,a}} * \frac{TIM_{a}} {BER_{a}}
+\end{split}
+\label{Eq:AgeBER}
+\end{equation}
+
+Where $TIM_{r,y,a}$ is the total dwellings in TIM by BER rating (r), year of construction (y), and archetype (a).
+$CSO_{y,a}$ is the number of dwellings by year of construction and archetype in the CSO dataset, the CSO dataset does not contain BER rating types.
+$BER_{r,y,a}$ is the number of BER ratings by year of construction and archetype from the BER database and dividing this value by $BER_{y,a}$ results in a probability of a certain age type falling within a BER rating, and the final part of the equation $\frac{TIM_{a}} {BER_{a}}$ compensates for the non-stated archetypes in the CSO data, $TIM_a$ was already calculated in equation \ref{Eq:BERtoCSO2}.
+The results of each of the correlations are in  \ref{sec:appendixA2}.
+Table \ref{Table:FinalDwellingStock} outlines the final dwelling stock used in TIM.
+
+\begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Final Dwelling Stock}
+\begin{tabular}{l|lll}
+BER Rating & Apartment & Attached & Detached \\ \hline
+A          & 9,419     & 15,472   & 20,379   \\
+B1         & 7,459     & 9,538    & 9,434    \\
+B2         & 14,042    & 17,545   & 20,091   \\
+B3         & 19,924    & 49,769   & 53,466   \\
+C          & 58,905    & 282,152  & 251,319  \\
+D          & 43,739    & 187,627  & 166,668  \\
+E          & 25,768    & 101,419  & 81,397   \\
+F          & 12,331    & 50,124   & 43,241   \\
+G          & 15,211    & 52,707   & 78,436   \\ \hline
+Total      & 206,799   & 766,352  & 724,430 
+\end{tabular}
+\label{Table:FinalDwellingStock}
+\end{table}
+
+\subsubsection{Costs}
+\label{LitR:Costs}
+
+This section only covers the costs applied in the residential sector, which includes space and water heating technologies, retrofitting, cooking, lighting, pumps, fans, space cooling, and other electrical appliances.
+The fuel costs are not applied in the residential sector, there are applied in the supply sector, and the data is sourced from World Energy Outlook reports from 2019 and 2020. 
+The majority of the space and water heating cost data is obtained from the
+\textit{``Techno-economic projections until 2050 for smaller heating and cooling technologies in the residential and tertiary sectors in the EU”} report \cite{JointResearchCentre2017Techno-economicEU}.
+The cost data used in TIM was based on the north EU region (Denmark) and where data was missing, the central EU region (Germany) cost data was used.
+The high granularity of data allows for the cost of a technology to be disaggregated between apartments and non-apartments, costs and efficiencies are projected every five years from 2020 to 2050.
+The lifetime and emissions/pollutants are also included, while the cost data is further disaggregated between capital expenditure ( equipment, installation \& additional capital costs ) and Operational \& Maintenance (O\&M) costs.
+\par The model has two types of retrofitting options per archetype,  a shallow retrofit is when the primary energy ($kWh/m^2/yr$) is reduced by $ \leq 35 \%$, and deep retrofit is where the primary energy is reduced by $>35 \%$, a reduction of $ \> 75 \%$ is considered infeasible.
+The different between the pre-retrofit and post-retrofit primary energy determines the type of retrofit, as shown in \ref{sec:appendixA3}.
+
+The retrofit cost data was sourced from Ali et.
+al \cite{Ali2020ABuildings} who derived costs from SEAI's Better Energy Homes (BEH) and Better Energy Warmer Homes (BEWH) datasets, combined this is Ireland's largest retrofitting dataset ( over 285,000 retrofits ).
+To provide more robust retrofit costs, data is also used from the Typology Approach for Building Stock Energy Assessment (TABULA) report \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} which has more detailed cost data with two stages of retrofit (standard and advanced) for a sample of 31 dwellings.
+The formulas used to find the average cost of a retrofit per type of retrofit and archetype are shown in Equations \ref{Eq:RetrofitCost1} and \ref{Eq:RetrofitCost2}.
+
+\begin{equation} 
+\begin{split}
+ \text{\euro}_{rf-rt,a} = Average ( (Ali_{rf-rt} * AF),TABULA_{rf-rt,a} )
+\end{split}
+\label{Eq:RetrofitCost1}
+\end{equation}
+
+Where $ {\text{\euro}}{}_{rf-rt,a}$ is the retrofit cost from pre-retrofit BER rating (rf) to post-retrofit rating (rt) and archetype (a).
+$Ali_{r,t}$ is the cost data from Ali et al.
+\cite{Ali2020ABuildings}, it contains all ratings and types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments, 1 for attached, and 1.1 for detached dwellings.
+$TABULA_{rf-rt,a}$ is the average retrofit cost from pre-retrofit BER rating to post-retrofit BER rating and archetype sourced from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland}.
+However with only 62 samples in total not all retrofits are included.
+In this case, TABULA is ignored and Ali et al.
+is the only source of retrofit costs.
+The retrofit costs are cross-checked with the residential cost-optimal report \cite{AECOM2020ReportBuildings}.
+\par To simplify the model, the cost of retrofit per BER rating is not used in TIM. Instead, an average weighted cost per archetype and per retrofit is calculated.
+
+\begin{equation} 
+\begin{split}
+\text{\euro}_{a,t} = Average(\text{\euro}_{rf,a}) * \frac{TIM_{r,a}} {TIM_{a}}
+\end{split}
+\label{Eq:RetrofitCost2}
+\end{equation}
+
+${TIM_{r,a}}$ is the number of dwellings by rating and archetype and ${TIM_{a}}$ is the total number of dwellings of that archetype, however A-rated dwellings are excluded from the total, as it is assumed they cannot be retrofitted further.
+
+\par The cost data for different electrical appliances was sourced from Topten EU database.
+Topten has a large database covering electrical appliances available in the European market, the data collected includes cost and efficiency.
+While some Topten reports have looked at sales and market shares \cite{Michel2016EnergyREPORT}.
+
+\subsubsection{Savings}
+\label{LitR:Savings}
+
+The previous section outlined the cost of retrofitting, and this section looks at the expected savings of retrofitting existing dwellings and building new dwellings.  
+The annual heating costs per archetype and per BER rating are sourced from SEAI's indicative energy running costs and the internal temperature assumptions in Table \ref{Table:InternalTemps} are included to calculate the costs, which are outlined in Table \ref{Table:FuelCost}
+
+\begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Heating fuel cost per BER rating and archetype }
+    \begin{tabular}{l|lcc}
+    \hline
+    BER Rating & Attached & \multicolumn{1}{l}{Detached} & \multicolumn{1}{l}{Apartment} \\ \hline
+A1         & $\text{\euro}$323      & $\text{\euro}$539                          & $\text{\euro}$200               \\
+A2         & $\text{\euro}$646      & $\text{\euro}$1078                         &$\text{\euro}$400                \\
+A3         & $\text{\euro}$804      & $\text{\euro}$1213                         & $\text{\euro}$500                 \\
+B1         & $\text{\euro}$745      & $\text{\euro}$1200                         & $\text{\euro}$440              \\
+B2         & $\text{\euro}$950      & $\text{\euro}$1500                         & $\text{\euro}$570                \\
+B3         & $\text{\euro}$1150     & $\text{\euro}$1900                         & $\text{\euro}$700                \\
+C1         & $\text{\euro}$862      & $\text{\euro}$1432                         & $\text{\euro}$499              \\
+C2         & $\text{\euro}$1022     & $\text{\euro}$1693                         & $\text{\euro}$624           \\
+C3         & $\text{\euro}$1181     & $\text{\euro}$1888                         & $\text{\euro}$686                \\
+D1         & $\text{\euro}$1434     & $\text{\euro}$2364                         & $\text{\euro}$836                \\
+D2         & $\text{\euro}$1701     & $\text{\euro}$2769                         & $\text{\euro}$964               \\
+E1         & $\text{\euro}$1979     & $\text{\euro}$3233                         & $\text{\euro}$1187               \\
+E2         & $\text{\euro}$2252     & $\text{\euro}$3646                         & $\text{\euro}$1319                \\
+F          & $\text{\euro}$2740     & $\text{\euro}$4403                         & $\text{\euro}$1626              \\
+G          & $\text{\euro}$2702     & $\text{\euro}$4398                         & $\text{\euro}$1687        \\                  \hline
+    \end{tabular}
+  \label{Table:FuelCost}
+\end{table}
+
+As TIM only inputs the savings per type of retrofit and archetype, some calculations must be performed on Table \ref{Table:FuelCost}.
+The first calculation is to find the savings from pre-retrofit running cost to post-retrofit running cost, as shown in Equation \ref{Eq:RetrofitSaving1}
+
+\begin{equation} 
+\begin{split}
+ Saving(\%)_{a,rf-rt} = 1 - ( \frac {Average(\text{\euro}_{rt,a}/year) } {Average(\text{\euro}_{rf,a}/year) } )
+\end{split}
+\label{Eq:RetrofitSaving1}
+\end{equation}
+
+Where $Saving(\%)_{a,rf-rt}$ are the savings ( or reduced running cost) per archetype (a), where the pre-retrofit BER rating is defined as the rating from (rf), and the post-retrofit BER rating is defined as the rating to (rt).
+
+Allocating each retrofit change to either shallow or deep is the next step, this is shown in \ref{sec:appendixA3}.
+The next calculation is to find the weighted average of each retrofit depth per archetype, this is then a TIM input.
+
+\begin{equation} 
+\begin{split}
+Saving(\%)_{a,d} = Average(Saving(\%)_{a,rf}) * \frac{TIM_{r,a}} {TIM_{a}}
+\end{split}
+\label{Eq:RetrofitSaving2}
+\end{equation}
+
+Where t is the type of retrofit.
+${TIM_{r,a}}$ is the number of dwellings by rating and archetype and ${TIM_{a}}$ is the total number of dwellings of that archetype.The expected savings from each type of retrofit and for each archetype are outlined in Table \ref{Table:RetrofitSavings}.
+
+\begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Retrofit Savings}
+\begin{tabular}{l|lcc}
+Retrofit & Attached & \multicolumn{1}{l}{Detached} & \multicolumn{1}{l}{Apartment} \\ \hline
+Shallow  & 20\%   & 20\%     & 22\%     \\
+Deep     & 44\%   & 42\%     & 37\%               
+\end{tabular}
+\label{Table:RetrofitSavings}
+\end{table}
+
+\subsubsection{Energy Service Demand Projections}
+\label{LitR:Demands}
+
+The technical bottom-up development approach of the residential sector is combined with other economic sectors to meet macroeconomic defined demands.
+The energy service demands in the residential sector are driven by the total dwelling stock for non-archetype energy services and apartment, attached or detached dwelling stock for archetype energy services.
+The residential stock projections up to 2040 are taken from Economic and Social Research Institute's (ESRI's) housing demand estimates \cite{Bergin2020RegionalLevel}, beyond 2040 the growth rate is extrapolated.
+However, archetpye specfic growth is limited or unavailable, for this reason a simple linear regression, shown in Eq.\ref{Eq:RegArchetype} is used to calculate the share of an archetype in a given year (y).
+Figure \ref{fig:DwellingStoc} graphically represents the number of existing and new dwellings per archetype.
+The growth rate of apartments is the highest and increasing up to 2040, while attached and detached dwellings are decreasing, indicating a preference for smaller dwellings to increase in the future.
+The obsolescence rate of existing dwelling is 0.35\%.
+
+\begin{equation} 
+\begin{split}
+Apt(\%) =  7.5\ln{(y)} -56.97 \\
+Att(\%) = -2.53\ln{(y)} + 19.68 \\
+Det(\%) = -4.93\ln{(y)} + 37.91
+\end{split}
+\label{Eq:RegArchetype}
+\end{equation}
+
+\begin{figure}[!htbp]
+ \centering
+ \caption{Number of dwellings by type and age ('000)}
+ \includegraphics[scale=0.42]{Figures/TIM_DwellingStock.png}
+ \label{fig:DwellingStoc}
+\end{figure}
+
+%\begin{table}[htbp]
+%  \centering
+%  \footnotesize
+%  \caption{Dwelling Growth}
+%\begin{tabular}{l|lcc}
+%Years & Apartment & \multicolumn{1}{l}{Attached } & %\multicolumn{1}{l}{Detached} \\ \hline
+%2016-2020  & 2.89\%   & 1.69\%     & 1.58\%     \\
+%\hline
+%2020-2030     & 4.26\%   & 1.15\%     & 0.81\%   \\
+%2030-2040     & 3.33\%   & 0.89\%     & 0.53\%   \\
+%2040-2050     & 2.47\%   & 0.52\%     & 0.14\%   \\
+%\end{tabular}
+%\label{Table:DwellingGrowth}
+%\end{table}
+
+Archetype energy service demands are dependant on both the type of dwelling and whether it is existing or new.
+The average archetype energy service demand in new and existing is shown in Table \ref{Archetype Energy Service Demands}.
+
+\begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Archetype Energy Service Demands}
+ \begin{tabular}{c|cc|cc|cc}
+ \hline
+ \multicolumn{1}{c|}{Energy Service}  & \multicolumn{2}{c|}{Apartments} & \multicolumn{2}{c|}{Attached} & \multicolumn{2}{c}{Detached}  \\
+  \multicolumn{1}{c|}{}  & \multicolumn{2}{c|}{($kWh/house/yr$)} & \multicolumn{2}{c|}{($kWh/house/yr$)} & \multicolumn{2}{c}{($kWh/house/yr$)}  \\
+  & Existing & New & Existing & New & Existing & New \\
+ \hline
+ Space Heating & 3,853 & 1,228 & 7,852 & 2,845 & 14,405 & 6,745 \\
+ Water Heating & 3,409 & 2,197 & 3,011 & 2,351 & 3,806 & 2,493 \\
+ Lighting & 179 & 156 & 254 & 229 & 419 & 392  \\
+ Pumps \& Fans  & 55  & 60 & 106 & 104 & 131 & 110  \\
+ Space Cooling  & 0 & 60 & 0 & 104 & 0 & 110  \\
+ \hline
+ \end{tabular}
+ \label{Archetype Energy Service Demands}
+\end{table}
+
+Space heating, water heating, and lighting energy service demands for existing dwellings are calculated from the average values of the base year dwelling stock, and as all new dwellings are required to be A-rated, the new dwellings are the average for the existing A-rated dwellings.
+The Pumps \& Fans are similar, except in the new dwellings the value is halved and the remaining half is added to space cooling.
+This is an assumption which was cross-checked with POTEnCIA (Policy Oriented Tool for Energy and Climate Change Impact Assessment) space cooling projections per household.
+
+%HDD and warmer weather to offset each other???
+%Heterogeneous Occupancy Behaviour > House Stock 
+%Residential Energy was >50\%higher before internal temperature adjustments
+%Macroeconomic 
+%no elasticity - fuel prices, and weather compensation?
+%Time slices - 40 etc.
+
+\subsubsection{Space and Water Heating}
+\label{LitR:Heating}
+
+The demand for space and water heating is outlined in the previous section, and this section explains how water and space heating are achieved through technology and retrofits.
+While the demands are exogenous, the choice of technologies to at least satisfy the demand are endogenous and is captured in Equation \ref{Eq:Technology}.
+
+\begin{equation} 
+\begin{split}
+Min \text{\euro} \sum_k VAR\_ACT_{k,i}(y)  \geq DM_i(y) 
+\end{split}
+\label{Eq:Technology}
+\end{equation}
+
+Where VAR\_ACT are the activity levels of end-use technology (k) for demand category (i) in year (y).
+DM is the exogenous demand per energy service per year.
+The cost of the technology used is minimised while also satisfying all constraints.
+
+All technologies are archetype specific and some technologies deliver both water and space heating, while other technologies can only deliver one type of heating, the list of available technologies is outlined in \ref{sec:appendixA4}.
+Retrofits are included as they satisfy the residential space heating demand, without consuming fuel.
+The Danish Energy Agency's Technology Data for Individual Heat Production dataset was input to TIM, this dataset is also used in the EU TIMES model, the dataset has high granularity and is continuously updated.
+
+The number of main heating systems equals the number of dwellings, this assumption ignores any impact of secondary system heating.
+One expection in the rule is , which are only secondary heating systems.
+
+The base year heating energy consumption uses census data \cite{2016CensusIreland} from the CSO dataset E1008 is used to obtain the fuel used in heating systems per archetype.
+The fuel usage is calibrated using the SEAI Energy Balance 2018.
+
+\subsubsection{Lighting, Appliances and Other}
+\label{LitR:Lighting,Appliances}
+
+Lighting, appliances, and pumps \& fans are all fueled by electricity, whereas cooking can be fueled by electricity, gas or LPG.
+As shown in Figure \ref{fig:Ireland2018ServiceDemands}, these energy services accounted for 20\% of energy consumption in 2018, however, this share of consumption is expected to increase.
+Refrigeration, cloth washing, cloth drying, and dish washing are under the appliances category and data is sourced from Topten \cite{Michel2016EnergyREPORT} and SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
+Cooking is also sourced from Eurostat's ``Energy consumption in households by type and end-used" excel datasheet and cross-checked with SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}  Appliances not included in this list are simply under the ``Other Appliances" category in TIM and these final values are cross-checked with POTEnCIA.
+Other Appliances include television, multimedia, and computer technology.
+Lighting and pumps \& fans are archetype specific energy services as the data is sourced from the BER database.
+Equation \ref{Eq:Technology} is also used, but unlike the heating energy service demands, there is little or no choice in different technologies (k) for some of the demand categories (i), therefore, the model has no choice but to invest in a technology, however  cost and efficiency of each technology does determine the overall investment.
+
+\subsubsection{User Constraints}
+\label{LitR:Constraints}
+The TIM model operates on the basis of perfect foresight, but
+\textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel} for this reason ``User Constraints" are applied in the residential sector to better reflect reality.
+The User Constraints applied to the residential sector include:
+\begin{itemize}
+ \item One Space and Water Heating System per household ( stoves excluded ) 
+ \item Maximum 60\% of existing buildings to be retrofitted by 2030, and 95 \% by 2070. Maximum \% interpolated between 2030 and 2070. 
+ \item A prerequisite electrical heat pump requirement is applied to each archetype, the percentage of dwellings which requires a shallow or deep retrofit before a heat pump can be installed (  B2 rating for heat pump ready dwelling ).
+Hybrid heat pumps are not included in this constraint.
+ \begin{table}[htbp]
+  \centering
+  \footnotesize
+  \caption{Prerequisite Heat Pump Requirement}
+    \begin{tabular}{r|rrr}
+    \hline
+    \multicolumn{1}{l}{Requirement} & \multicolumn{1}{l}{Apartment} & \multicolumn{1}{l}{Attached } & \multicolumn{1}{l}{Detached} \\ \hline
+    No Retrofit  & 24.5 \%   & 14.3 \%   & 12 \% \\
+    Shallow Retrofit  & 28.5 \%   & 34.7 \%  & 36.8 \% \\
+    Deep Retrofit  & 47 \%   & 51 \%  & 51.2 \% \\
+    \end{tabular}%
+  \label{tab:HeatPumpConstraint}%
+\end{table}%
+ \item The natural gas grid is currently connected to 39\% of apartments, 52\% of attached dwellings and 12\% of detached dwellings.
+The total share of dwellings connected to the gas grid is an upper constraint to grow by 50\% for each archetype.
+\item Maximum fuel share in cooking for gas and LPG is 40\% and 10\% respectively. Base year fuel share for Gas is 32\% and LPG is 1.5 \%.
+\item Maximum 10\% solar input to dual fuel heating technology
+\item Maximum 30\% gas input to gas hybrid heating pumps
+\end{itemize}
+
+While other constraints are applied in different sectors, which affect the residential sector and vice versa, they are not listed in this paper.
+The growth constraints listed above are calculated using the following \cite{Loulou2016DocumentationI} :
+\begin{equation} 
+\begin{split}
+VAR\_CAP(t + 1) \leq (1 + G^{M(t+1)-M(t)}) \cdot VAR\_CAP(t)+K
+\end{split}
+\label{Eq:GrowthConstraint}
+\end{equation}
+
+Where $VAR\_CAP$ is the technology capacity in time period (t), G is the growth coefficient, defined as a new attribute of the technology and represents the maximum annual growth allowed for the capacity.
+The quantity $M(t+1)-M(t)$ is the number of years between the milestones of time periods t and t+1.
+The constant K is useful whenever the technology has no capacity initially, to allow the capacity to build over time.
+
+\section{Results}
+\label{Results}
+
+The newly developed TIM provides some valuable insights on each sector and the sectoral interconnections.
+The results outlined in this paper are based upon six scenarios.
+While TIM can only analysis the energy system, the other large source of GHG emissions is from agriculture, therefore some scenarios are labelled based on agriculture GHG reduction percentage and Energy System GHG reduction percentage, the six scenarios are:
+\begin{itemize}
+    \item \textbf{No Mitigation} Least cost solution, where no emissions constraints are imposed
+    \item \textbf{With Additional Measures (WAM)} includes the implementation of Ireland’s CAP 2019.
+    \item \textbf{A25E65} Agriculture GHG emissions reduce by 25\%, Energy system GHG emissions reduces by 65\% by 2030 compared to 2018.
+    \item \textbf{A33E61} Agriculture GHGs reduces by 33\% and Energy system GHGs by 61\%
+    \item \textbf{A40E57} Agriculture GHGs reduces by 33\% and Energy system GHGs by 61\%
+    \item \textbf{A51E51} Equal GHG reduction share in of 51 \% in Agriculture and Energy Systems. 
+\end{itemize}
+
+While the first two scenarios do not comply with the Climate Action and Low Carbon Development (Amendment) Act 2021 target of an overall 51 \% GHG reduction in 2030 compared to 2018 or net-zero by 2050, however they do provide insights on the incremental steps required.
+The last four scenarios comply with 2030 and 2050 targets and offer insights about different trajectories to achieve those targets.
+Under CAP 2019 agriculture, GHG emission reduction is 10-15\% on the projected levels in 2030 relative to 2017.
+The higher agriculture ambition explored GHG reductions of 25-51\%.
+However, the energy consumed by agriculture is covered in this model.
+
+Figure \ref{fig:$CO_2$ Waterfall} provides an overview of  scenario affects on each sector in 2030, as the energy system ambition increases incrementally from left to right.
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[width=\textwidth]{Figures/CO2Waterfall.png}
+ \caption{$CO_2$ Waterfall}
+ \label{fig:$CO_2$ Waterfall}
+\end{figure}
+
+The share of sectoral $CO_2$ emissions is not evenly spread in 2018 and the optimal share of sectoral $CO_2$ reduction in 2030 is also unevenly distributed.
+In the A51E51 scenario, the optimal electricity annual emissions reduction is $7.1 MtCO_2$ in 2030, while transport currently emits the similiar $CO_2$ levels only reduces by $4.5 MtCO_2$.
+The residential sector reduces $CO_2$ emission by 63\% in A51E51 in 2030, electricity is 60 \% and transport is 38\%.
+Not only does Figure \ref{fig:$CO_2$ Waterfall} show the different optimal sectoral reductions, but also the ``low hanging fruit" in the large electricity reduction to the    ``high coconuts" in the step from A33E61 to A25E65, transport mitigates the majority of $CO_2$ emissions required.
+
+Zooming in and focusing on the fuel-mix within the residential sector gives further insights on the optimal pathways to achieve 2030 targets, Figure \ref{fig:ResFuel} shows this, with the base year stacked chart on the top row, the four climate bill compliant scenarios on rows 2-5 and the last two rows are non-compliant.
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[width=\textwidth]{Figures/RSD_Fuel2.png}
+ \caption{Residential Fuel per scenario}
+ \label{fig:ResFuel}
+\end{figure}
+
+In 2018, the residential sector had 124 PJ of primary energy that figure needs to decrease by 8-13 \% in 2030, to comply with climate targets.
+While Figure \ref{fig:$CO_2$ Waterfall} shows the large $CO_2$ reduction in electricity, this translates to a Figure \ref{fig:ResFuel} representing a large uptake of electricity in the residential sector within climate bill compliant scenarios.
+The large increase in ambient energy indicates the type of technology deployed in the residential sector, more on this in Figure XXX.
+The overall share of ambient heat increases from 1.4 \% in 2019 to between 13-26 \% in 2030.
+as a fuel  In climate bill compliant scenarios, peat and coal are completely phased out, while oil use decreases 90-93 \% by 2030.
+Gas energy increase by 34 \% in the A51E51 scenario, but remains relatively stable among other climate bill compliant scenarios in 2030.
+By comparing compliant and noncompliant scenarios, we can observe the favoured fuels and technologies, while also observing the incremental steps of emission reduction and overall cost change in each scenario.
+
+While Figure \ref{fig:ResFuel} shows the primary energy reduction required, it must be noted that retrofits are used to reduce the primary energy required ( reduced energy through behavioural change is not modelled ).
+Figure \ref{fig:Retrofits} shows both the number of retrofits per type and the energy savings for each of the four climate bill compliant scenarios.
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[width=\textwidth]{Figures/Retrofits.png}
+ \caption{Retrofits by type per scenario}
+ \label{fig:Retrofits}
+\end{figure}
+
+The number of retrofits required varies from 393,000 in A51E51 to 615,000 in A25E65, meaning between 23 - 36 \% of existing dwellings will require some form of retrofit.  
+As mentioned in Table \ref{tab:HeatPumpConstraint}, the rollout of retrofits constraints the amount of heat pumps which can be installed, however, a certain share of existing dwellings and all new dwellings are heat pump ready, i.e B2 rating or above.
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[width=\textwidth]{Figures/HeatPumps.png}
+ \caption{Heat Pump by type per scenario}
+ \label{fig:HeatPump}
+\end{figure}
+
+Where SW is space and water heating, HC is space heating and space cooling, and A-to-W represents air-to-water type heat pumps.
+The optimal number of heat pumps in each scenario ranges from 492,000 in A51E51 to 938,000 in A25E65, which equates to 23 - 45 \% of dwellings having a heat pump in 2030.
+Close to 45,000 air to water heat pumps are installed in apartments across all scenarios, as there are no retrofits in apartments, these heat pumps are installed in the existing B2 rating or above apartments or new apartments.
+The detached air-to-water heat pumps are the most prevalent across scenarios and maybe considered the low hanging fruit.
+
+The advantage of an energy system model is to observe the intersectoral connections.
+Figure \ref{fig:SectorFuel} presents  the base year primary energy for four sectors - power generation, transport, services, and residential.
+Additionally, the primary energy of fuels in 2030 is included in the four scenarios which comply with the 51 \% reduction scenarios in 2030.
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[width=\textwidth]{Figures/IntesectoralFuels.png}
+ \caption{Sector Annual Fuel in 2018 and four scenarios in 2030}
+ \label{fig:SectorFuel}
+\end{figure}
+
+Some consistencies across the scenarios show the large growth of renewables in the power generation sector and consequently the uptake of electricity across the remaining sectors.
+The least ambitious A51E51 scenario shows 59 PJ of electricity used in residential in 2030, but as constraints tighten, the electricity is used more in transport, while the residential sector invests more in solar panels and other forms of renewable energy.
+While the use of bioenergy in transport and services also affects the residential sector as bioenergy is limited.
+Transport uses 16 PJ of natural gas in A51E51 to 0 PJ in A25E65, this allows the residential sector to use natural gas instead, which moves from 33 PJ in A51E51 to 24 PJ in A25E65.
+
+As a cheaper natural gas powered vehicle may effect the optimal energy mix and the annualised cost in other sectors.
+Figure \ref{fig:SectorCost} shows the annual costs of no mitigation in 2030 and the incremental increase of 2030 sectoral costs of the four scenarios which comply with the 51 \% reduction scenarios in 2030.
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[width=\textwidth]{Figures/CostWaterfall.png}
+ \caption{Sector Annual Cost for five scenarios in 2030}
+ \label{fig:SectorCost}
+\end{figure}
+
+The no mitigation scenario is expected to cost $\text{\euro} 8.1 $ billion in 2030, that figure rises to $\text{\euro} 12.4 $ billion under the A51E51 scenarion, with electricity ( Power generation ) required an extra $\text{\euro} 2 $ billion in 2030.
+As the scenario constraints tighten, larger investments can be seen in the transport sector.
+
+\section{Discussion}
+\label{Discussion}
+
+The methodology applied to the residential sector in TIM is transparent, up-to-date, and robust.
+The results outlined have summarised costs, fuel-mix, and intersectoral effects within the energy system.
+The results provide insights, further scenarios or sensitivity analysis can provide further insights from TIM.
+In addition, the results can be used in parallel with simulation models to examine specific climate policies or can alternative scenarios.
+
+The model brings alot of value to the sectroal carbon budgets and long-term climate targets, while also including more technologies and detail than any previous Irish ESOM.
+
+Gas is a favoured fuel from Fig \ref{fig:RESFuel} this provides insights on the potential to decarbonise
+
+%o	Focus on what added value the model brings
+%o	Also include future development priorities here
+
+\subsection{Strengths}
+\label{Dis:Strengths}
+
+First iteration - of the model will need further development, DH, BEhavour, higher granularity, and spatial will be easily done with the dataset. 
+%What does new model do that previous didn’t?
+%TIM residential improvements - DH, HVO, Solar ( in progress ), better graps the cost and savings of retrofits and heat pumps, which can infrom MACC.Build with 2018 baseyear, not 2010 as previous versions had - this suits the 2030 target
+%THe implications of net-zero
+%Does new policy context require new modelling approach?
+%Yes, do account for ongoing investments in climate mitigation opitions, the approach should provide higher granularity for likely pathways
+%How does new modelling structure provide insights (give examples) in a way that previous model didn’t?
+%Demands - 
+%Improved more detailed Energy Service Demands
+Detailed bottom-up model - using the 2018 base year ( in line with policy )
+Updated technology
+Combines energy system effects
+Robust filteorange input data about housing stock
+
+\subsection{Limitations}
+\label{Dis:Limit}
+%o	What can’t the model do? Where is it more suitable to use another model - Hannah Email
+Cost of buildings - bER ratings..
+BER Database - No PV Solar capacity, No Heat Pump ready, Large defaults, wall types check?)
+Community Retrofits - Better Energy Communities is our national retrofit initiative with grant support of up to €28 million for 2021.
+We support new approaches to achieving energy efficiency in Irish communities.
+%Discrepancies were investigated, and assumptions reviewed intuitively in order to achieve a good match.=
+Consumers’ final investment decision can be
+represented by “willingness-to-pay” curves – for
+instance, around 10\% of owner-occupiers are willing
+to invest in energy efficiency measures for a payback
+time of 4 years, falling to 0\% for a payback time of
+6 years for some consumer cohorts
+Type of Tenure is ignoorange
+BER aggeragateed
+housing stock u-values
+behaviour and occupancy
+Area
+expected energy consumption weakness of BER database as 
+Actual energy consumption for dwellings of
+similar or near identical construction may vary by up to 600 \%
+depending on usage \cite{Sunikka-Blank2012BuildingConsumption}
+
+Retrofitting ``rebound" saving  energy is more difficult to measure like improved physical and mental health.
+ArDEM SQL - exogenous behavioural effects 5\% increase of residential energy in 2030 when internal temperatures are adjusted. 
+\section{Conclusions}
+\label{Conclusion}
+This paper reports results on an ambitious mitigation target in the period to 2050 for the Irish energy system. The analysis has been performed using the Irish TIMES model
+\textit{We stress the importance of differentiating between technical con-
+straints and behavioural constraints in optimisation models.
+Including both conflates what is consideorange desirable under the economic welfare paradigm (i.e.
+cost-minimising decisions) with “realistic decisions”, complicating the interpretation of the results.
+Optimisation models are by design well suited to study cost-optimal behaviour, and running scenarios just with technical constraints facilitates the identification of areas where market barriers are more entrenched, and interventions to accelerate the uptake of low-carbon practices may be needed.} \cite{Astudillo2017CanModelb}
+\section{Acknowledgements}
+\label{sec:Acknowledgements}
+Tomas
+Ankita%CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC
+% https://www.marei.ie/project/capacity/
+CHIMERA is supported by a research grant from Science Foundation Ireland (SFI) and the National Natural Science Foundation of China (NSFC) under the SFI-NSFC Partnership Programme, grant no.
+17/NSFC/5181 and
+supported by MaREI, the SFI Research Centre for Energy, Climate, and Marine [Grant No: 12/RC/2302\_P2]
+
+https://www.marei.ie/project/chimera/
+\appendix
+
+\newpage
+\section{}
+\label{sec:appendix}
+%% The Appendices part is started with the command \appendix;
+%% appendix sections are then done as normal sections
+\subsection{}
+\label{sec:appendixA}
+Residential Fuel data is sourced from SEAI Energy Balance and BER Database, the right column shows how the fuels are structured in TIM. 
 \begin{table}[htbp]
@@
 
-In TIM, fuels such as coal, represent a combination of coal fuels from the SEAI Energy Balance and BER database.
-Kerosene accounts for over 70 \% of home heating oil usage in 2018, while Gasoil accounts for most of the remaining share, the fuels have similar properties, and there was no evidence to support keeping separate fuels would provide further insight into residential decarbonisation.
-The exception is ambient heat, which could potentially provide insights to residential decarbonisation by separating it into ambient heat for space heating and ambient heat for water heating.
-The solid multi-fuel from the BER database represents an open fire or stove which is heated by either coal, peat, or wood.
-Through examining the SEAI Energy Balance and BER database, the fuel share assigned for coal, peat, and wood is 19.4 \%, 74.4 \% and 6.2 \% respectively.
-There are two types of energy services represented in Table \ref{fig:RES} - Non-Archetype and Archetype.
-Non-archetype energy services fill in the lack of data on archetype energy services.
-The six non-archetype energy services are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service, this bottom-up data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{LitR:Lighting,Appliances}.
-The four archetype energy services are dependent on  building archetypes and data is sourced from the BER database, which is discussed in Section \ref{LitR:StockData}.
-The internal temperature adjustment for space heating is detailed in Section \ref{LitR:StockData}.
-The demand for archetype and non-archetype energy services is based on the growth of residential dwellings, which is detailed in section \ref{LitR:Demands}.
-
-\subsubsection{Dwelling Stock Data}
-\label{LitR:StockData}
-
-The dwelling stock data is obtained mainly from the SEAI BER Public database, which is firstly filteorange, then scaled up using Central Statistics Office (CSO) data. 
-The BER database is Ireland's national Energy Performance Certificate (EPC) register, which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD).
-The BERs are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
-Over half of Irish dwellings are captuorange in the BER database, every BER displays over 200 energy related variables.
-The DEAP model, which was outlined by Dineen et al.\cite{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM} provides the results for the energy rating in the BER database.
-However, the space heating calculation is used to estimate the expected energy consumption per dwelling, which makes some assumptions.
-DEAP assumes that each dwelling is single zoned, the heating season is from October to May inclusive, and the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season.
-One assumption, the internal temperature assumption, results in a large error, where space and water heating are overestimated in DEAP.
-Space heating consumption is reduced by 26 \% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance.
-DEAP assumes all living areas ( living room, sitting room, and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C.
-Through the analysis of previous residential modelling in Ireland this is not the case, as the internal temperatures are often much lower, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions applied in TIM.
-
-\begin{table}[htbp]
-  \centering
-  \footnotesize
-  \caption{Internal Temperature Assumptions}
-    \begin{tabular}{rrr}
-    \hline
-    \multicolumn{1}{l}{BER Rating} & \multicolumn{1}{l}{Living Area} & \multicolumn{1}{l}{Non-Living Area }  \\ \hline
-    A  & 23$^{\circ}$C   & 20$^{\circ}$C  \\
-    B  & 21$^{\circ}$C   & 18$^{\circ}$C  \\
-    C  & 18$^{\circ}$C  & 15$^{\circ}$C  \\
-    D  & 18$^{\circ}$C   & 15$^{\circ}$C  \\
-    E  & 18$^{\circ}$C   & 15$^{\circ}$C  \\ 
-    F  & 18$^{\circ}$C   & 15$^{\circ}$C  \\ 
-    G  & 18$^{\circ}$C   & 13$^{\circ}$C  \\ \hline
-    \end{tabular}
-  \label{Table:InternalTemps}
-\end{table}
-
-The ArDEM-SQL, which was described in Section \ref{Intro:ESOMIreland} is a simulation model built on EN 13790.
-ArDEM-SQL alters the primary energy consumption of the BER database by changing the internal temperatures.
-
-Other concerns for uncertainties in the database, which are not related to DEAP calculation are statistically anomalous spikes in the data, caused by applying default values \cite{Ahern2020ADatabases} and different companies generating variations of ±20\% in the assessment of energy consumption \cite{Harsman2016OnCertificates}.While the BER database provides high granularity data, it must be filteorange to improve the representation of the dwelling stock data.
-The filters applied to the raw BER database are outlined in \ref{sec:appendixA}.
-
-The Census 2016 data \cite{2016CensusIreland} from the CSO is also used in calculating the dwelling stock.
-Dataset E1005 provides a split of seven archetypes per county, which are aggregated to form three archetypes.
-The total number of building archetypes per county in 2016 is used to scale up the BER database.
-The gap in total dwellings between the census year in 2016 and the model base year in 2018 was initially ignoored  due to lack of data.
-However, later it was found that data on new dwelling electrical connections per archetype and per county are available from CSO dataset NDQ06, this would have filled the gap, the first iteration of the dwelling stock has a 2 \% error.
-
-Once both databases were finalised, the next step in this process was to scale up the BER database.
-
-This was done by first finding the ratio of the building archetype per county between the BER database and CSO data, as shown in Equation \ref{Eq:BERtoCSO1}.
-
-\begin{equation} 
-\begin{split}
-R_{a,c} = {\frac{CSO_{a-nsa,c}} {BER_{a,c}} + \frac{CSO_{nsa,c}} {CSO_{a,c}}}
-\end{split}
-\label{Eq:BERtoCSO1}
-\end{equation}
-
-$R_{a,c}$ is the ratio per archetype (a) and per county (c), there are 26 counties and 3 archetypes.
-$CSO_{a-nsa,c}$ represents the number of dwelling archetypes minus the non-stated archetypes (nsa) per county in the CSO database.
-$BER_{a,c}$ is the number of archetypes per county  in the BER database.
-The next step is to multiply the number of BERs per archetype and per county by the relevant ratio.
-\begin{equation} 
-\begin{split}
-BER_{a,c} * R_{a,c} = TIM_{a,c}
-\end{split}
-\label{Eq:BERtoCSO2}
-\end{equation}
-
-$TIM_{a,c}$ represents the final dwelling stock data per archetype and per county, the first iteration of TIM does not disaggregate the dwellings per county, instead a national total per archetype is input to TIM ( $TIM_a$ ).
-
-Once the final dwelling stock is complete, the next step is to assign a BER rating to all dwellings.
-This was done by comparing 9 different dwelling age ranges for each of the 3 archetypes between the BER and CSO datasets.
-
-\begin{equation} 
-\begin{split}
-TIM_{r,y,a} = CSO_{y,a} * \frac{BER_{r,y,a}} {BER_{y,a}} * \frac{TIM_{a}} {BER_{a}}
-\end{split}
-\label{Eq:AgeBER}
-\end{equation}
-
-Where $TIM_{r,y,a}$ is the total dwellings in TIM by BER rating (r), year of construction (y), and archetype (a).
-$CSO_{y,a}$ is the number of dwellings by year of construction and archetype in the CSO dataset, the CSO dataset does not contain BER rating types.
-$BER_{r,y,a}$ is the number of BER ratings by year of construction and archetype from the BER database and dividing this value by $BER_{y,a}$ results in a probability of a certain age type falling within a BER rating, and the final part of the equation $\frac{TIM_{a}} {BER_{a}}$ compensates for the non-stated archetypes in the CSO data, $TIM_a$ was already calculated in equation \ref{Eq:BERtoCSO2}.
-The results of each of the correlations are in  \ref{sec:appendixA2}.
-Table \ref{Table:FinalDwellingStock} outlines the final dwelling stock used in TIM.
-
-\begin{table}[htbp]
-  \centering
-  \footnotesize
-  \caption{Final Dwelling Stock}
-\begin{tabular}{l|lll}
-BER Rating & Apartment & Attached & Detached \\ \hline
-A          & 9,419     & 15,472   & 20,379   \\
-B1         & 7,459     & 9,538    & 9,434    \\
-B2         & 14,042    & 17,545   & 20,091   \\
-B3         & 19,924    & 49,769   & 53,466   \\
-C          & 58,905    & 282,152  & 251,319  \\
-D          & 43,739    & 187,627  & 166,668  \\
-E          & 25,768    & 101,419  & 81,397   \\
-F          & 12,331    & 50,124   & 43,241   \\
-G          & 15,211    & 52,707   & 78,436   \\ \hline
-Total      & 206,799   & 766,352  & 724,430 
-\end{tabular}
-\label{Table:FinalDwellingStock}
-\end{table}
-
-\subsubsection{Costs}
-\label{LitR:Costs}
-
-This section only covers the costs applied in the residential sector, which includes space and water heating technologies, retrofitting, cooking, lighting, pumps, fans, space cooling, and other electrical appliances.
-The fuel costs are not applied in the residential sector, there are applied in the supply sector, and the data is sourced from World Energy Outlook reports from 2019 and 2020. 
-The majority of the space and water heating cost data is obtained from the
-\textit{``Techno-economic projections until 2050 for smaller heating and cooling technologies in the residential and tertiary sectors in the EU”} report \cite{JointResearchCentre2017Techno-economicEU}.
-The cost data used in TIM was based on the north EU region (Denmark) and where data was missing, the central EU region (Germany) cost data was used.
-The high granularity of data allows for the cost of a technology to be disaggregated between apartments and non-apartments, costs and efficiencies are projected every five years from 2020 to 2050.
-The lifetime and emissions/pollutants are also included, while the cost data is further disaggregated between capital expenditure ( equipment, installation \& additional capital costs ) and Operational \& Maintenance (O\&M) costs.
-\par The model has two types of retrofitting options per archetype,  a shallow retrofit is when the primary energy ($kWh/m^2/yr$) is reduced by $ \leq 35 \%$, and deep retrofit is where the primary energy is reduced by $>35 \%$, a reduction of $ \> 75 \%$ is considered infeasible.
-The different between the pre-retrofit and post-retrofit primary energy determines the type of retrofit, as shown in \ref{sec:appendixA3}.
-
-The retrofit cost data was sourced from Ali et.
-al \cite{Ali2020ABuildings} who derived costs from SEAI's Better Energy Homes (BEH) and Better Energy Warmer Homes (BEWH) datasets, combined this is Ireland's largest retrofitting dataset ( over 285,000 retrofits ).
-To provide more robust retrofit costs, data is also used from the Typology Approach for Building Stock Energy Assessment (TABULA) report \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} which has more detailed cost data with two stages of retrofit (standard and advanced) for a sample of 31 dwellings.
-The formulas used to find the average cost of a retrofit per type of retrofit and archetype are shown in Equations \ref{Eq:RetrofitCost1} and \ref{Eq:RetrofitCost2}.
-
-\begin{equation} 
-\begin{split}
- \text{\euro}_{rf-rt,a} = Average ( (Ali_{rf-rt} * AF),TABULA_{rf-rt,a} )
-\end{split}
-\label{Eq:RetrofitCost1}
-\end{equation}
-
-Where $ {\text{\euro}}{}_{rf-rt,a}$ is the retrofit cost from pre-retrofit BER rating (rf) to post-retrofit rating (rt) and archetype (a).
-$Ali_{r,t}$ is the cost data from Ali et al.
-\cite{Ali2020ABuildings}, it contains all ratings and types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments, 1 for attached, and 1.1 for detached dwellings.
-$TABULA_{rf-rt,a}$ is the average retrofit cost from pre-retrofit BER rating to post-retrofit BER rating and archetype sourced from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland}.
-However with only 62 samples in total not all retrofits are included.
-In this case, TABULA is ignored and Ali et al.
-is the only source of retrofit costs.
-The retrofit costs are cross-checked with the residential cost-optimal report \cite{AECOM2020ReportBuildings}.
-\par To simplify the model, the cost of retrofit per BER rating is not used in TIM. Instead, an average weighted cost per archetype and per retrofit is calculated.
-
-\begin{equation} 
-\begin{split}
-\text{\euro}_{a,t} = Average(\text{\euro}_{rf,a}) * \frac{TIM_{r,a}} {TIM_{a}}
-\end{split}
-\label{Eq:RetrofitCost2}
-\end{equation}
-
-${TIM_{r,a}}$ is the number of dwellings by rating and archetype and ${TIM_{a}}$ is the total number of dwellings of that archetype, however A-rated dwellings are excluded from the total, as it is assumed they cannot be retrofitted further.
-
-\par The cost data for different electrical appliances was sourced from Topten EU database.
-Topten has a large database covering electrical appliances available in the European market, the data collected includes cost and efficiency.
-While some Topten reports have looked at sales and market shares \cite{Michel2016EnergyREPORT}.
-
-\subsubsection{Savings}
-\label{LitR:Savings}
-
-The previous section outlined the cost of retrofitting, and this section looks at the expected savings of retrofitting existing dwellings and building new dwellings.  
-The annual heating costs per archetype and per BER rating are sourced from SEAI's indicative energy running costs and the internal temperature assumptions in Table \ref{Table:InternalTemps} are included to calculate the costs, which are outlined in Table \ref{Table:FuelCost}
-
-\begin{table}[htbp]
-  \centering
-  \footnotesize
-  \caption{Heating fuel cost per BER rating and archetype }
-    \begin{tabular}{l|lcc}
-    \hline
-    BER Rating & Attached & \multicolumn{1}{l}{Detached} & \multicolumn{1}{l}{Apartment} \\ \hline
-A1         & $\text{\euro}$323      & $\text{\euro}$539                          & $\text{\euro}$200               \\
-A2         & $\text{\euro}$646      & $\text{\euro}$1078                         &$\text{\euro}$400                \\
-A3         & $\text{\euro}$804      & $\text{\euro}$1213                         & $\text{\euro}$500                 \\
-B1         & $\text{\euro}$745      & $\text{\euro}$1200                         & $\text{\euro}$440              \\
-B2         & $\text{\euro}$950      & $\text{\euro}$1500                         & $\text{\euro}$570                \\
-B3         & $\text{\euro}$1150     & $\text{\euro}$1900                         & $\text{\euro}$700                \\
-C1         & $\text{\euro}$862      & $\text{\euro}$1432                         & $\text{\euro}$499              \\
-C2         & $\text{\euro}$1022     & $\text{\euro}$1693                         & $\text{\euro}$624           \\
-C3         & $\text{\euro}$1181     & $\text{\euro}$1888                         & $\text{\euro}$686                \\
-D1         & $\text{\euro}$1434     & $\text{\euro}$2364                         & $\text{\euro}$836                \\
-D2         & $\text{\euro}$1701     & $\text{\euro}$2769                         & $\text{\euro}$964               \\
-E1         & $\text{\euro}$1979     & $\text{\euro}$3233                         & $\text{\euro}$1187               \\
-E2         & $\text{\euro}$2252     & $\text{\euro}$3646                         & $\text{\euro}$1319                \\
-F          & $\text{\euro}$2740     & $\text{\euro}$4403                         & $\text{\euro}$1626              \\
-G          & $\text{\euro}$2702     & $\text{\euro}$4398                         & $\text{\euro}$1687        \\                  \hline
-    \end{tabular}
-  \label{Table:FuelCost}
-\end{table}
-
-As TIM only inputs the savings per type of retrofit and archetype, some calculations must be performed on Table \ref{Table:FuelCost}.
-The first calculation is to find the savings from pre-retrofit running cost to post-retrofit running cost, as shown in Equation \ref{Eq:RetrofitSaving1}
-
-\begin{equation} 
-\begin{split}
- Saving(\%)_{a,rf-rt} = 1 - ( \frac {Average(\text{\euro}_{rt,a}/year) } {Average(\text{\euro}_{rf,a}/year) } )
-\end{split}
-\label{Eq:RetrofitSaving1}
-\end{equation}
-
-Where $Saving(\%)_{a,rf-rt}$ are the savings ( or reduced running cost) per archetype (a), where the pre-retrofit BER rating is defined as the rating from (rf), and the post-retrofit BER rating is defined as the rating to (rt).
-
-Allocating each retrofit change to either shallow or deep is the next step, this is shown in \ref{sec:appendixA3}.
-The next calculation is to find the weighted average of each retrofit depth per archetype, this is then a TIM input.
-
-\begin{equation} 
-\begin{split}
-Saving(\%)_{a,d} = Average(Saving(\%)_{a,rf}) * \frac{TIM_{r,a}} {TIM_{a}}
-\end{split}
-\label{Eq:RetrofitSaving2}
-\end{equation}
-
-Where t is the type of retrofit.
-${TIM_{r,a}}$ is the number of dwellings by rating and archetype and ${TIM_{a}}$ is the total number of dwellings of that archetype.The expected savings from each type of retrofit and for each archetype are outlined in Table \ref{Table:RetrofitSavings}.
-
-\begin{table}[htbp]
-  \centering
-  \footnotesize
-  \caption{Retrofit Savings}
-\begin{tabular}{l|lcc}
-Retrofit & Attached & \multicolumn{1}{l}{Detached} & \multicolumn{1}{l}{Apartment} \\ \hline
-Shallow  & 20\%   & 20\%     & 22\%     \\
-Deep     & 44\%   & 42\%     & 37\%               
-\end{tabular}
-\label{Table:RetrofitSavings}
-\end{table}
-
-\subsubsection{Energy Service Demand Projections}
-\label{LitR:Demands}
-
-The technical bottom-up development approach of the residential sector is combined with other economic sectors to meet macroeconomic defined demands.
-The energy service demands in the residential sector are driven by the total dwelling stock for non-archetype energy services and apartment, attached or detached dwelling stock for archetype energy services.
-The residential stock projections up to 2040 are taken from Economic and Social Research Institute's (ESRI's) housing demand estimates \cite{BerginREGIONALLEVEL}, beyond 2040 the growth rate is extrapolated.
-However, archetpye specfic growth is limited or unavailable, for this reason a simple linear regression, shown in Eq.\ref{Eq:RegArchetype} is used to calculate the share of an archetype in a given year (y).
-Figure \ref{fig:DwellingStoc} graphically represents the number of existing and new dwellings per archetype.
-The growth rate of apartments is the highest and increasing up to 2040, while attached and detached dwellings are decreasing, indicating a preference for smaller dwellings to increase in the future.
-The obsolescence rate of existing dwelling is 0.35\%.
-
-\begin{equation} 
-\begin{split}
-Apt(\%) =  7.5\ln{(y)} -56.97 \\
-Att(\%) = -2.53\ln{(y)} + 19.68 \\
-Det(\%) = -4.93\ln{(y)} + 37.91
-\end{split}
-\label{Eq:RegArchetype}
-\end{equation}
-
-\begin{figure}[!htbp]
- \centering
- \caption{Number of dwellings by type and age ('000)}
- \includegraphics[scale=0.42]{Figures/TIM_DwellingStock.png}
- \label{fig:DwellingStoc}
-\end{figure}
-
-%\begin{table}[htbp]
-%  \centering
-%  \footnotesize
-%  \caption{Dwelling Growth}
-%\begin{tabular}{l|lcc}
-%Years & Apartment & \multicolumn{1}{l}{Attached } & %\multicolumn{1}{l}{Detached} \\ \hline
-%2016-2020  & 2.89\%   & 1.69\%     & 1.58\%     \\
-%\hline
-%2020-2030     & 4.26\%   & 1.15\%     & 0.81\%   \\
-%2030-2040     & 3.33\%   & 0.89\%     & 0.53\%   \\
-%2040-2050     & 2.47\%   & 0.52\%     & 0.14\%   \\
-%\end{tabular}
-%\label{Table:DwellingGrowth}
-%\end{table}
-
-Archetype energy service demands are dependant on both the type of dwelling and whether it is existing or new.
-The average archetype energy service demand in new and existing is shown in Table \ref{Archetype Energy Service Demands}.
-
-\begin{table}[htbp]
-  \centering
-  \footnotesize
-  \caption{Archetype Energy Service Demands}
- \begin{tabular}{c|cc|cc|cc}
- \hline
- \multicolumn{1}{c|}{Energy Service}  & \multicolumn{2}{c|}{Apartments} & \multicolumn{2}{c|}{Attached} & \multicolumn{2}{c}{Detached}  \\
-  \multicolumn{1}{c|}{}  & \multicolumn{2}{c|}{($kWh/house/yr$)} & \multicolumn{2}{c|}{($kWh/house/yr$)} & \multicolumn{2}{c}{($kWh/house/yr$)}  \\
-  & Existing & New & Existing & New & Existing & New \\
- \hline
- Space Heating & 3,853 & 1,228 & 7,852 & 2,845 & 14,405 & 6,745 \\
- Water Heating & 3,409 & 2,197 & 3,011 & 2,351 & 3,806 & 2,493 \\
- Lighting & 179 & 156 & 254 & 229 & 419 & 392  \\
- Pumps \& Fans  & 55  & 60 & 106 & 104 & 131 & 110  \\
- Space Cooling  & 0 & 60 & 0 & 104 & 0 & 110  \\
- \hline
- \end{tabular}
- \label{Archetype Energy Service Demands}
-\end{table}
-
-Space heating, water heating, and lighting energy service demands for existing dwellings are calculated from the average values of the base year dwelling stock, and as all new dwellings are required to be A-rated, the new dwellings are the average for the existing A-rated dwellings.
-The Pumps \& Fans are similar, except in the new dwellings the value is halved and the remaining half is added to space cooling.
-This is an assumption which was cross-checked with POTEnCIA (Policy Oriented Tool for Energy and Climate Change Impact Assessment) space cooling projections per household.
-
-%HDD and warmer weather to offset each other???
-%Heterogeneous Occupancy Behaviour > House Stock 
-%Residential Energy was >50\%higher before internal temperature adjustments
-%Macroeconomic 
-%no elasticity - fuel prices, and weather compensation?
-%Time slices - 40 etc.
-
-\subsubsection{Space and Water Heating}
-\label{LitR:Heating}
-
-The demand for space and water heating is outlined in the previous section, and this section explains how water and space heating are achieved through technology and retrofits.
-While the demands are exogenous, the choice of technologies to at least satisfy the demand are endogenous and is captured in Equation \ref{Eq:Technology}.
-
-\begin{equation} 
-\begin{split}
-Min \text{\euro} \sum_k VAR\_ACT_{k,i}(y)  \geq DM_i(y) 
-\end{split}
-\label{Eq:Technology}
-\end{equation}
-
-Where VAR\_ACT are the activity levels of end-use technology (k) for demand category (i) in year (y).
-DM is the exogenous demand per energy service per year.
-The cost of the technology used is minimised while also satisfying all constraints.
-
-All technologies are archetype specific and some technologies deliver both water and space heating, while other technologies can only deliver one type of heating, the list of available technologies is outlined in \ref{sec:appendixA4}.
-Retrofits are included as they satisfy the residential space heating demand, without consuming fuel.
-The Danish Energy Agency's Technology Data for Individual Heat Production dataset was input to TIM, this dataset is also used in the EU TIMES model, the dataset has high granularity and is continuously updated.
-
-The number of main heating systems equals the number of dwellings, this assumption ignores any impact of secondary system heating.
-One expection in the rule is , which are only secondary heating systems.
-
-The base year heating energy consumption uses census data \cite{2016CensusIreland} from the CSO dataset E1008 is used to obtain the fuel used in heating systems per archetype.
-The fuel usage is calibrated using the SEAI Energy Balance 2018.
-
-\subsubsection{Lighting, Appliances and Other}
-\label{LitR:Lighting,Appliances}
-
-Lighting, appliances, and pumps \& fans are all fueled by electricity, whereas cooking can be fueled by electricity, gas or LPG.
-As shown in Figure \ref{fig:Ireland2018ServiceDemands}, these energy services accounted for 20\% of energy consumption in 2018, however, this share of consumption is expected to increase.
-Refrigeration, cloth washing, cloth drying, and dish washing are under the appliances category and data is sourced from Topten \cite{Michel2016EnergyREPORT} and SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
-Cooking is also sourced from Eurostat's ``Energy consumption in households by type and end-used" excel datasheet and cross-checked with SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}  Appliances not included in this list are simply under the ``Other Appliances" category in TIM and these final values are cross-checked with POTEnCIA.
-Other Appliances include television, multimedia, and computer technology.
-Lighting and pumps \& fans are archetype specific energy services as the data is sourced from the BER database.
-Equation \ref{Eq:Technology} is also used, but unlike the heating energy service demands, there is little or no choice in different technologies (k) for some of the demand categories (i), therefore, the model has no choice but to invest in a technology, however  cost and efficiency of each technology does determine the overall investment.
-
-\subsubsection{User Constraints}
-\label{LitR:Constraints}
-The TIM model operates on the basis of perfect foresight, but
-\textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel} for this reason ``User Constraints" are applied in the residential sector to better reflect reality.
-The User Constraints applied to the residential sector include:
-\begin{itemize}
- \item One Space and Water Heating System per household ( stoves excluded ) 
- \item Maximum 60\% of existing buildings to be retrofitted by 2030, and 95 \% by 2070. Maximum \% interpolated between 2030 and 2070. 
- \item A prerequisite electrical heat pump requirement is applied to each archetype, the percentage of dwellings which requires a shallow or deep retrofit before a heat pump can be installed (  B2 rating for heat pump ready dwelling ).
-Hybrid heat pumps are not included in this constraint.
- \begin{table}[htbp]
-  \centering
-  \footnotesize
-  \caption{Prerequisite Heat Pump Requirement}
-    \begin{tabular}{r|rrr}
-    \hline
-    \multicolumn{1}{l}{Requirement} & \multicolumn{1}{l}{Apartment} & \multicolumn{1}{l}{Attached } & \multicolumn{1}{l}{Detached} \\ \hline
-    No Retrofit  & 24.5 \%   & 14.3 \%   & 12 \% \\
-    Shallow Retrofit  & 28.5 \%   & 34.7 \%  & 36.8 \% \\
-    Deep Retrofit  & 47 \%   & 51 \%  & 51.2 \% \\
-    \end{tabular}%
-  \label{tab:HeatPumpConstraint}%
-\end{table}%
- \item The natural gas grid is currently connected to 39\% of apartments, 52\% of attached dwellings and 12\% of detached dwellings.
-The total share of dwellings connected to the gas grid is an upper constraint to grow by 50\% for each archetype.
-\item Maximum fuel share in cooking for gas and LPG is 40\% and 10\% respectively. Base year fuel share for Gas is 32\% and LPG is 1.5 \%.
-\item Maximum 10\% solar input to dual fuel heating technology
-\item Maximum 30\% gas input to gas hybrid heating pumps
-\end{itemize}
-
-While other constraints are applied in different sectors, which affect the residential sector and vice versa, they are not listed in this paper.
-The growth constraints listed above are calculated using the following \cite{Loulou2016DocumentationI} :
-\begin{equation} 
-\begin{split}
-VAR\_CAP(t + 1) \leq (1 + G^{M(t+1)-M(t)}) \cdot VAR\_CAP(t)+K
-\end{split}
-\label{Eq:GrowthConstraint}
-\end{equation}
-
-Where $VAR\_CAP$ is the technology capacity in time period (t), G is the growth coefficient, defined as a new attribute of the technology and represents the maximum annual growth allowed for the capacity.
-The quantity $M(t+1)-M(t)$ is the number of years between the milestones of time periods t and t+1.
-The constant K is useful whenever the technology has no capacity initially, to allow the capacity to build over time.
-
-\section{Results}
-\label{Results}
-
-The newly developed TIM provides some valuable insights on each sector and the sectoral interconnections.
-The results outlined in this paper are based upon six scenarios.
-While TIM can only analysis the energy system, the other large source of GHG emissions is from agriculture, therefore some scenarios are labelled based on agriculture GHG reduction percentage and Energy System GHG reduction percentage, the six scenarios are:
-\begin{itemize}
-    \item \textbf{No Mitigation} Least cost solution, where no emissions constraints are imposed
-    \item \textbf{With Additional Measures (WAM)} includes the implementation of Ireland’s CAP 2019.
-    \item \textbf{A25E65} Agriculture GHG emissions reduce by 25\%, Energy system GHG emissions reduces by 65\% by 2030 compared to 2018.
-    \item \textbf{A33E61} Agriculture GHGs reduces by 33\% and Energy system GHGs by 61\%
-    \item \textbf{A40E57} Agriculture GHGs reduces by 33\% and Energy system GHGs by 61\%
-    \item \textbf{A51E51} Equal GHG reduction share in of 51 \% in Agriculture and Energy Systems. 
-\end{itemize}
-
-While the first two scenarios do not comply with the Climate Action and Low Carbon Development (Amendment) Act 2021 target of an overall 51 \% GHG reduction in 2030 compared to 2018 or net-zero by 2050, however they do provide insights on the incremental steps required.
-The last four scenarios comply with 2030 and 2050 targets and offer insights about different trajectories to achieve those targets.
-Under CAP 2019 agriculture, GHG emission reduction is 10-15\% on the projected levels in 2030 relative to 2017.
-The higher agriculture ambition explored GHG reductions of 25-51\%.
-However, the energy consumed by agriculture is covered in this model.
-
-Figure \ref{fig:$CO_2$ Waterfall} provides an overview of  scenario affects on each sector in 2030, as the energy system ambition increases incrementally from left to right.
-
-\begin{figure}[!htbp]
- \centering
- \includegraphics[width=\textwidth]{Figures/CO2Waterfall.png}
- \caption{$CO_2$ Waterfall}
- \label{fig:$CO_2$ Waterfall}
-\end{figure}
-
-The share of sectoral $CO_2$ emissions is not evenly spread in 2018 and the optimal share of sectoral $CO_2$ reduction in 2030 is also unevenly distributed.
-In the A51E51 scenario, the optimal electricity annual emissions reduction is $7.1 MtCO_2$ in 2030, while transport currently emits the similiar $CO_2$ levels only reduces by $4.5 MtCO_2$.
-The residential sector reduces $CO_2$ emission by 63\% in A51E51 in 2030, electricity is 60 \% and transport is 38\%.
-Not only does Figure \ref{fig:$CO_2$ Waterfall} show the different optimal sectoral reductions, but also the ``low hanging fruit" in the large electricity reduction to the    ``high coconuts" in the step from A33E61 to A25E65, transport mitigates the majority of $CO_2$ emissions required.
-
-Zooming in and focusing on the fuel-mix within the residential sector gives further insights on the optimal pathways to achieve 2030 targets, Figure \ref{fig:ResFuel} shows this, with the base year stacked chart on the top row, the four climate bill compliant scenarios on rows 2-5 and the last two rows are non-compliant.
-
-\begin{figure}[!htbp]
- \centering
- \includegraphics[width=\textwidth]{Figures/RSD_Fuel2.png}
- \caption{Residential Fuel per scenario}
- \label{fig:ResFuel}
-\end{figure}
-
-In 2018, the residential sector had 124 PJ of primary energy that figure needs to decrease by 8-13 \% in 2030, to comply with climate targets.
-While Figure \ref{fig:$CO_2$ Waterfall} shows the large $CO_2$ reduction in electricity, this translates to a Figure \ref{fig:ResFuel} representing a large uptake of electricity in the residential sector within climate bill compliant scenarios.
-The large increase in ambient energy indicates the type of technology deployed in the residential sector, more on this in Figure XXX.
-The overall share of ambient heat increases from 1.4 \% in 2019 to between 13-26 \% in 2030.
-as a fuel  In climate bill compliant scenarios, peat and coal are completely phased out, while oil use decreases 90-93 \% by 2030.
-Gas energy increase by 34 \% in the A51E51 scenario, but remains relatively stable among other climate bill compliant scenarios in 2030.
-By comparing compliant and noncompliant scenarios, we can observe the favoured fuels and technologies, while also observing the incremental steps of emission reduction and overall cost change in each scenario.
-
-While Figure \ref{fig:ResFuel} shows the primary energy reduction required, it must be noted that retrofits are used to reduce the primary energy required ( reduced energy through behavioural change is not modelled ).
-Figure \ref{fig:Retrofits} shows both the number of retrofits per type and the energy savings for each of the four climate bill compliant scenarios.
-
-\begin{figure}[!htbp]
- \centering
- \includegraphics[width=\textwidth]{Figures/Retrofits.png}
- \caption{Retrofits by type per scenario}
- \label{fig:Retrofits}
-\end{figure}
-
-The number of retrofits required varies from 393,000 in A51E51 to 615,000 in A25E65, meaning between 23 - 36 \% of existing dwellings will require some form of retrofit.  
-As mentioned in Table \ref{tab:HeatPumpConstraint}, the rollout of retrofits constraints the amount of heat pumps which can be installed, however, a certain share of existing dwellings and all new dwellings are heat pump ready, i.e B2 rating or above.
-\begin{figure}[!htbp]
- \centering
- \includegraphics[width=\textwidth]{Figures/HeatPumps.png}
- \caption{Heat Pump by type per scenario}
- \label{fig:HeatPump}
-\end{figure}
-
-Where SW is space and water heating, HC is space heating and space cooling, and A-to-W represents air-to-water type heat pumps.
-The optimal number of heat pumps in each scenario ranges from 492,000 in A51E51 to 938,000 in A25E65, which equates to 23 - 45 \% of dwellings having a heat pump in 2030.
-Close to 45,000 air to water heat pumps are installed in apartments across all scenarios, as there are no retrofits in apartments, these heat pumps are installed in the existing B2 rating or above apartments or new apartments.
-The detached air-to-water heat pumps are the most prevalent across scenarios and maybe considered the low hanging fruit.
-
-The advantage of an energy system model is to observe the intersectoral connections.
-Figure \ref{fig:SectorFuel} presents  the base year primary energy for four sectors - power generation, transport, services, and residential.
-Additionally, the primary energy of fuels in 2030 is included in the four scenarios which comply with the 51 \% reduction scenarios in 2030.
-\begin{figure}[!htbp]
- \centering
- \includegraphics[width=\textwidth]{Figures/IntesectoralFuels.png}
- \caption{Sector Annual Fuel in 2018 and four scenarios in 2030}
- \label{fig:SectorFuel}
-\end{figure}
-
-Some consistencies across the scenarios show the large growth of renewables in the power generation sector and consequently the uptake of electricity across the remaining sectors.
-The least ambitious A51E51 scenario shows 59 PJ of electricity used in residential in 2030, but as constraints tighten, the electricity is used more in transport, while the residential sector invests more in solar panels and other forms of renewable energy.
-While the use of bioenergy in transport and services also affects the residential sector as bioenergy is limited.
-Transport uses 16 PJ of natural gas in A51E51 to 0 PJ in A25E65, this allows the residential sector to use natural gas instead, which moves from 33 PJ in A51E51 to 24 PJ in A25E65.
-
-As a cheaper natural gas powered vehicle may effect the optimal energy mix and the annualised cost in other sectors.
-Figure \ref{fig:SectorCost} shows the annual costs of no mitigation in 2030 and the incremental increase of 2030 sectoral costs of the four scenarios which comply with the 51 \% reduction scenarios in 2030.
-
-\begin{figure}[!htbp]
- \centering
- \includegraphics[width=\textwidth]{Figures/CostWaterfall.png}
- \caption{Sector Annual Cost for five scenarios in 2030}
- \label{fig:SectorCost}
-\end{figure}
-
-The no mitigation scenario is expected to cost $\text{\euro} 8.1 $ billion in 2030, that figure rises to $\text{\euro} 12.4 $ billion under the A51E51 scenarion, with electricity ( Power generation ) required an extra $\text{\euro} 2 $ billion in 2030.
-As the scenario constraints tighten, larger investments can be seen in the transport sector.
-
-\section{Discussion}
-\label{Discussion}
-
-The methodology applied to the residential sector in TIM is transparent, up-to-date, and robust.
-The results outlined have summarised costs, fuel-mix, and intersectoral effects within the energy system.
-The results provide insights, further scenarios or sensitivity analysis can provide further insights from TIM.
-In addition, the results can be used in parallel with simulation models to examine specific climate policies or can alternative scenarios.
-
-The model brings alot of value to the sectroal carbon budgets and long-term climate targets, while also including more technologies and detail than any previous Irish ESOM.
-
-Gas is a favoured fuel from Fig \ref{fig:RESFuel} this provides insights on the potential to decarbonise
-
-%o	Focus on what added value the model brings
-%o	Also include future development priorities here
-
-\subsection{Strengths}
-\label{Dis:Strengths}
-
-First iteration - of the model will need further development, DH, BEhavour, higher granularity, and spatial will be easily done with the dataset. 
-%What does new model do that previous didn’t?
-%TIM residential improvements - DH, HVO, Solar ( in progress ), better graps the cost and savings of retrofits and heat pumps, which can infrom MACC.Build with 2018 baseyear, not 2010 as previous versions had - this suits the 2030 target
-%THe implications of net-zero
-%Does new policy context require new modelling approach?
-%Yes, do account for ongoing investments in climate mitigation opitions, the approach should provide higher granularity for likely pathways
-%How does new modelling structure provide insights (give examples) in a way that previous model didn’t?
-%Demands - 
-%Improved more detailed Energy Service Demands
-Detailed bottom-up model - using the 2018 base year ( in line with policy )
-Updated technology
-Combines energy system effects
-Robust filteorange input data about housing stock
-
-\subsection{Limitations}
-\label{Dis:Limit}
-%o	What can’t the model do? Where is it more suitable to use another model - Hannah Email
-Cost of buildings - bER ratings..
-BER Database - No PV Solar capacity, No Heat Pump ready, Large defaults, wall types check?)
-Community Retrofits - Better Energy Communities is our national retrofit initiative with grant support of up to €28 million for 2021.
-We support new approaches to achieving energy efficiency in Irish communities.
-%Discrepancies were investigated, and assumptions reviewed intuitively in order to achieve a good match.=
-Consumers’ final investment decision can be
-represented by “willingness-to-pay” curves – for
-instance, around 10\% of owner-occupiers are willing
-to invest in energy efficiency measures for a payback
-time of 4 years, falling to 0\% for a payback time of
-6 years for some consumer cohorts
-Type of Tenure is ignoorange
-BER aggeragateed
-housing stock u-values
-behaviour and occupancy
-Area
-expected energy consumption weakness of BER database as 
-Actual energy consumption for dwellings of
-similar or near identical construction may vary by up to 600 \%
-depending on usage \cite{Sunikka-Blank2012BuildingConsumption}
-
-Retrofitting ``rebound" saving  energy is more difficult to measure like improved physical and mental health.
-ArDEM SQL - exogenous behavioural effects 5\% increase of residential energy in 2030 when internal temperatures are adjusted. 
-\section{Conclusions}
-\label{Conclusion}
-This paper reports results on an ambitious mitigation target in the period to 2050 for the Irish energy system. The analysis has been performed using the Irish TIMES model
-\textit{We stress the importance of differentiating between technical con-
-straints and behavioural constraints in optimisation models.
-Including both conflates what is consideorange desirable under the economic welfare paradigm (i.e.
-cost-minimising decisions) with “realistic decisions”, complicating the interpretation of the results.
-Optimisation models are by design well suited to study cost-optimal behaviour, and running scenarios just with technical constraints facilitates the identification of areas where market barriers are more entrenched, and interventions to accelerate the uptake of low-carbon practices may be needed.} \cite{Astudillo2017CanModelb}
-\section{Acknowledgements}
-\label{sec:Acknowledgements}
-Tomas
-Ankita%CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC
-% https://www.marei.ie/project/capacity/
-CHIMERA is supported by a research grant from Science Foundation Ireland (SFI) and the National Natural Science Foundation of China (NSFC) under the SFI-NSFC Partnership Programme, grant no.
-17/NSFC/5181 and
-supported by MaREI, the SFI Research Centre for Energy, Climate, and Marine [Grant No: 12/RC/2302\_P2]
-
-https://www.marei.ie/project/chimera/
-\appendix
-
-\section{}
-\label{sec:appendix}
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
+\newpage
 \subsection{}
-\label{sec:appendixA}
+\label{sec:appendixA1}
+
 BER Database Filters
@@
 
-\newpage
-
 \subsection{}
 \label{sec:appendixA4}
-
 List of water and space heating technologies available per archetype. The ``???" denotes either APT for Apartments, ATT for Attached, or DET for detached dwellings.
+
 \begin{table}[htbp]
@@
   \caption{Space and Water Heating Technologies}
-  \resizebox{\textwidth}{!}{
- \begin{tabular}{cc|cc|cc|cc}
- \hline
+  \resizebox{\textwidth}{!}{\begin{tabular}{cc|cc|cc|cc}
+ \hline
+
   \multicolumn{2}{c|}{Technology} & \multicolumn{2}{c|}{Apartments} & \multicolumn{2}{c|}{Attached} & \multicolumn{2}{c}{Detached}  \\
@@
 \printbibliography
-
+%\bibliography{TIM_V1.1}
+%\bibliographystyle{elsarticle-num}
+%\bibliography{TIM_V11}
 %\bibliographystyle{elsarticle-harv}
@@
 \end{document}
-\endinput
+%\endinput
 %%

### 2021-07-28 manuscript.tex
Version 1.03 - ready to send to Brian OG

@@
 
-\documentclass[review]{elsarticle}
+\documentclass[preprint]{elsarticle}
 
@@
 %% Text of abstract
-%Ireland's residential sector has particularly underachieved in previous climate targets.
+\par Ireland's residential sector has particularly underachieved in previous climate targets.
 Located on the peripherals of Europe, with a high share of low thermal efficient detached dwellings with carbon intensive heating fuels, provides some distinctive barriers to decarbonising Ireland's residential sector.
@@
 \par
-%Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges which provides insights to decision makers on issues related to climate and energy policy.
+Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges, which provides insights to decision makers on issues related to climate and energy policy.
 \par
-%The TIMES-Ireland Model (TIM) is a newly developed optimisation model of the Irish energy system, which calculates the  cost-optimal decarbonisation pathway to meet future energy service demands while achieving Ireland's legally binding 2030 and 2050 climate targets.
-\par The Irish residential sector has proved hard to decarbonise, yet recently Ireland has set one of the most ambitious climate targets into law, a 51 \% greenhouse gas reduction in 2030, compared to 2018.
+The TIMES-Ireland Model (TIM) is a newly developed optimisation model for the Irish energy system.
+This paper describes the methodology applied to the development of the residential sector in TIM.
+Some results are analysed to provide input to discuss insights from the residential sector in TIM.
+%\par The Irish residential sector has proved hard to decarbonise, yet recently Ireland has set one of the most ambitious climate targets into law, a 51 \% greenhouse gas reduction in 2030, compared to 2018.
 There is a need for a new up to date and state of the art model to provide insights on the optimal decarbonisation pathway for the Irish energy system.
 This paper describes the methodology applied to the development of the residential sector in the new TIMES Ireland Model (TIM).
-Results are analysed to provide input to the discuss insights from the residential sector in TIM.
+Results are analysed to provide input to discuss insights from the residential sector in TIM.
 
@@
     \item Section \ref{Discussion} discusses the results, strengths, and weakness of the newly developed model and how it differs from other models.
-    \item Lastly, section \ref{Conclusion} draws some conclusions and discusses possible future work.
+    \item Section \ref{Conclusion} draws some conclusions and discusses possible future work.
 \end{itemize}
@@
 
-TIMES has been extensively used in both the UK and Ireland, and there are several examples using TIMES to focus on the residential sector.
+TIMES has been extensively used in both the UK and Ireland, and there are several examples of using TIMES to focus on the residential sector.
 TIMES assumes perfect foresight to enable a least-cost solution to be calculated, but \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
@@
 \par While Ireland's building regulations have encouraged the uptake of photovoltaics (PVs), the market share of PV solar panel systems remains low.
-The low amount of direct sunlight means Ireland's climate is not ideal for PVs but as the cost of PVs continues to fall, the potential role of PVs in  residential decarbonisation rises.
-In one study, Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model to explore the feasibility of PV systems installed at the household level in combination with batteries in Samsø.
+The low amount of direct sunlight means Ireland's climate is not ideal for PVs, but as the cost of PVs continues to fall, the potential role of PVs in  residential decarbonisation rises.
+In one study, Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} used EnergyPLAN model to explore the feasibility of PV systems installed at the household level in combination with batteries in Samsø.
 Two approaches are explored for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
@@
 While the battery location was explored, the PV system location was not.
-This is a great example of the strength of energy system models, whereby the residential PV systems affect wind energy investment.
+This is a great example of the strength of an ESOM, whereby the residential PV systems affect wind energy investment.
 However, a weakness in the model is the ability to comprehend household occupancy behaviour with and without household batteries on the PV system.
@@
 In 2000, Clinch and Healy published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Microsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating.
-A cost-benefit model (CBM) was fitted to the EAM and the results under the predicted scenario, showed energy savings outweighed programme costs, representing a benefit-cost ratio of 1.7.
+A cost-benefit model (CBM) was fitted to the EAM and the results under the predicted scenario showed energy savings outweighing programme costs, representing a benefit-cost ratio of 1.7.
 The benefits of reduced sickness, hospitalisation, drugs, and mortality were later quantified and this moved the benefit-cost ratio up to 3.0.
@@
 The bottom-up residential models are typically simulation models looking at a specific policy.
-The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switching to Microsoft SQL, with much higher computational power.
+The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switched to Microsoft SQL, with much higher computational power.
 There has been few studies on the Irish residential sector using ESOMs, as residential bottom-up models are preferred when focusing on that sector.
@@
 \label{Intro:IreRSD}
-%PICKUPHERE
+
 Ireland’s distinctive energy sectors are further separated from other EU member states when focusing on the residential sector.
@@
 85\% of residential buildings are 2005 or older.
-Furthermore, Ireland's fossil fuel dependency is underpinned by home heating oil, accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020EnergySEAI}.
+Additionally, Ireland's fossil fuel dependency is underpinned by home heating oil, which accounts for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020EnergySEAI}.
 
@@
 However, the decline in peat and coal has been offset by a more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005Energy-RelatedIreland.}.
-While, gas availability is limited to some urban areas and gas still accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005Energy-RelatedIreland.}.
+Despite being limited to some urban areas, gas still accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005Energy-RelatedIreland.}.
 The residential sector relied on fossil fuels for over 90\% of thermal energy in 2018, consequently Ireland currently has the lowest renewable heat among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
-Preliminary results from the Environmental Protection Agency (EPA) have indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.
+Preliminary results from the Environmental Protection Agency (EPA) have indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions, proving some consumers are economically rational by filling oil tanks when prices are low.
 
@@
 
-Where the residential sector is directly responsible for 47\% of heat and 30\% of electricity GHG emissions, the residential sector is therefore a pivotal sector in decarbonising energy in Ireland.
+Where the residential sector is directly responsible for 47\% of heat and 30\% of electricity GHG emissions.
+The residential sector would also be indirectly responsible for a share of transport GHG emissions, the residential sector is therefore a pivotal sector in decarbonising energy in Ireland.
 
@@
 
-Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to reduce Greenhouse Gas (GHG) emissions by 51\% in 2030 compared to 2018.
+Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to reduce GHG emissions by 51\% in 2030 compared to 2018.
 The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}.
@@
 \par The EU's first Nationally Determined Contribution (NDC), which had a 40\% GHG reduction target (first NDC), has been succeeded by a new target of at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels.
-Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-Climate Action Plan 2021 will have to be more ambitious than its porangeecessor, CAP2019 as it must reflect EU's updated NDC target and take account of the national target also.
+Ireland's Climate Action Plan 2019 (CAP2019) complies with the EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+Climate Action Plan 2021 will have to be more ambitious than its predecessor, CAP2019 as it must reflect EU's updated NDC target and take account of the national target also.
 There are also renewable energy targets to consider for transport (RES-T), heat (RES-H), and electricity (RES-E) under the renewable directive (EU) 2018/2001, and the energy efficiency targets under (EU) 2018/2002.
@@
 Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private investment towards low carbon or renewable energy.
-A €15/ $tCO_2$ carbon tax was introduced in 2009 to orangeuce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}.
+A €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}.
 Ireland's current carbon tax of €33.50/ $tCO_2$ will rise to €100/ $tCO_2$ in 2030.
@@
 
-\par Sustainable Energy Authority of Ireland (SEAI) has a key role in the rollout of retrofitting and heat pumps in existing homes.
+\par SEAI has a key role in the rollout of retrofitting and heat pumps in existing homes.
 In parallel, new dwellings must achieve a minimum share of 20\% renewable energy.
-The ambient heat used in heat pumps is included as a renewable energy, heat pumps are therefore the preferred heating technology despite the high capital costs.
+The ambient heat used in heat pumps is included as a renewable energy and are accounted for in the RES-H target, heat pumps are therefore the preferred heating technology despite the high capital costs.
 As mentioned in Section \ref{Intro:IreRSD}, new buildings in Ireland produce 70\% less GHG emissions than buildings in 2005, along with CAP2019 plans to retrofit 500,000 existing homes to at least a  ‘B2’ energy rating by 2030, which highlights the energy efficiency first approach in the residential sector.
@@
 \par In TIM, fuels such as coal, represent a combination of fuels from the SEAI Energy Balance and BER database.
-Gas oil which is accounted for in the SEAI Energy Balance, has similar properties to kerosene and there was no evidence to support keeping separate fuels would provide further insight into residential decarbonisation.
-However, dissagreatting ambient heat could potentially provide insights to residential decarbonisation.
+Gas oil, which is accounted for in the SEAI Energy Balance, has similar properties to kerosene and there was no evidence to support keeping separate fuels would provide further insight into residential decarbonisation.
+However, dissagreatting ambient heat could potentially provide insights on residential decarbonisation.
 The solid multi-fuel from the BER database represents an open fire or stove which is heated by either coal, peat, or wood.
-Through examining the SEAI Energy Balance and BER database, the fuel share assigned for coal, peat, and wood is 19.4 \%, 74.4 \% and 6.2 \% respectively.
-\par There are two types of energy services represented in Table \ref{fig:RES} - Non-Archetype and Archetype.
-Non-archetype energy services fill in the lack of data on archetype energy services.
+Through examining the SEAI Energy Balance and BER database, the fuel share  for coal, peat, and wood in solid multi-fuel is 19.4 \%, 74.4 \% and 6.2 \% respectively.
+\par There are two types of energy services represented in Fig.
+\ref{fig:RES} - Non-Archetype and Archetype.
+Essentially Non-archetype energy services have a lack of data and could not be accurately assigned to an archetype.
 The six non-archetype energy services are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service, this bottom-up data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{LitR:Lighting,Appliances}.
@@
 
-The dwelling stock data is obtained mainly from the SEAI BER Public database, which is firstly filteorange, then scaled up using Central Statistics Office (CSO) data. 
+The dwelling stock data is obtained from the SEAI BER Public database, which is firstly filtered, then scaled up using Central Statistics Office (CSO) data.
 The BER database is Ireland's national Energy Performance Certificate (EPC) register, which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD).
 The BERs are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
-Over half of Irish dwellings are captuorange in the BER database, every BER displays over 200 energy related variables.
-The DEAP model, which was outlined by Dineen et al.\cite{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM} provides the results for the energy rating in the BER database.
-However, the space heating calculation is used to estimate the expected energy consumption per dwelling, which makes some assumptions.
-DEAP assumes that each dwelling is single zoned, the heating season is from October to May inclusive, and the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season.
+\par Over half of Irish dwellings are captured in the BER database and every BER displays over 200 energy related variables.
+The DEAP model which is used to assign BER labels,  is outlined by Dineen et al.\cite{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM}, In terms of space heating, DEAP assumes that each dwelling is single zoned, the heating season is from October to May inclusive, and the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season.
 One assumption, the internal temperature assumption, results in a large error, where space and water heating are overestimated in DEAP.
 Space heating consumption is reduced by 26 \% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance.
-DEAP assumes all living areas ( living room, sitting room, and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C.
+DEAP assumes that all living areas ( living room, sitting room, and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C.
 Through the analysis of previous residential modelling in Ireland this is not the case, as the internal temperatures are often much lower, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions applied in TIM.
@@
 
-The ArDEM-SQL, which was described in Section \ref{Intro:ESOMIreland} is a simulation model built on EN 13790.
+The ArDEM-SQL, which was described in Section \ref{Intro:ESOMIreland} is a simulation model built on space heating calculation EN 13790.
 ArDEM-SQL alters the primary energy consumption of the BER database by changing the internal temperatures.
 
-Other concerns for uncertainties in the database, which are not related to DEAP calculation are statistically anomalous spikes in the data, caused by applying default values \cite{Ahern2020ADatabases} and different companies generating variations of ±20\% in the assessment of energy consumption \cite{Harsman2016OnCertificates}.While the BER database provides high granularity data, it must be filteorange to improve the representation of the dwelling stock data.
-The filters applied to the raw BER database are outlined in \ref{sec:appendixA}.
+Other concerns for uncertainties in the database, which are not related to DEAP calculation are statistically anomalous spikes in the data, caused by applying default values \cite{Ahern2020ADatabases} and different companies generating variations of ±20\% in the assessment of energy consumption \cite{Harsman2016OnCertificates}.
+While the BER database provides high granularity data, it must be filtered to improve the representation of the dwelling stock data.
+The filters applied to the raw BER database are outlined in \ref{sec:appendixA1}.
 
@@
 Dataset E1005 provides a split of seven archetypes per county, which are aggregated to form three archetypes.
-The total number of building archetypes per county in 2016 is used to scale up the BER database.
-The gap in total dwellings between the census year in 2016 and the model base year in 2018 was initially ignoored  due to lack of data.
-However, later it was found that data on new dwelling electrical connections per archetype and per county are available from CSO dataset NDQ06, this would have filled the gap, the first iteration of the dwelling stock has a 2 \% error.
-
-Once both databases were finalised, the next step in this process was to scale up the BER database.
-
+The total number of dwellings per archetype per county in 2016 is used to scale up the BER database.
+The gap in total dwellings between the census year in 2016 and the model base year in 2018 was initially ignored due to lack of data.
+However, later it was found that data on new dwelling electrical connections per archetype and per county are available from CSO dataset NDQ06, this would have filled the gap, the first iteration of the dwelling stock in TIM has a 2 \% error.
+\par Once the BER database and CSO database tables have been filled to include dwellings per archetype per county, the next step was to scale up the BER database.
 This was done by first finding the ratio of the building archetype per county between the BER database and CSO data, as shown in Equation \ref{Eq:BERtoCSO1}.
@@
   \footnotesize
-  \caption{Final Dwelling Stock}
+  \caption{TIM Base Year Dwelling Stock}
 \begin{tabular}{l|lll}
@@
 The lifetime and emissions/pollutants are also included, while the cost data is further disaggregated between capital expenditure ( equipment, installation \& additional capital costs ) and Operational \& Maintenance (O\&M) costs.
-\par The model has two types of retrofitting options per archetype,  a shallow retrofit is when the primary energy ($kWh/m^2/yr$) is reduced by $ \leq 35 \%$, and deep retrofit is where the primary energy is reduced by $>35 \%$, a reduction of $ \> 75 \%$ is considered infeasible.
-The different between the pre-retrofit and post-retrofit primary energy determines the type of retrofit, as shown in \ref{sec:appendixA3}.
+\par TIM has two types of retrofitting options per archetype,  a shallow retrofit is when the primary energy ($kWh/m^2/yr$) is reduced by $ \leq 35 \%$, and deep retrofit is where the primary energy is reduced by $>35 \%$, a reduction of $ \ >75 \%$ is considered infeasible.
+The difference between the pre-retrofit and post-retrofit primary energy determines the type of retrofit, as shown in \ref{sec:appendixA3}.
 
@@
 
-Where $ {\text{\euro}}{}_{rf-rt,a}$ is the retrofit cost from pre-retrofit BER rating (rf) to post-retrofit rating (rt) and archetype (a).
+Where $ {\text{\euro}}{}_{rf-rt,a}$ is the retrofit cost from pre-retrofit BER rating (rf) to post-retrofit rating (rt) for archetype (a).
 $Ali_{r,t}$ is the cost data from Ali et al.
-\cite{Ali2020ABuildings}, it contains all ratings and types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments, 1 for attached, and 1.1 for detached dwellings.
-$TABULA_{rf-rt,a}$ is the average retrofit cost from pre-retrofit BER rating to post-retrofit BER rating and archetype sourced from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland}.
+\cite{Ali2020ABuildings}, it contains all BER ratings and retrofit types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments, 1 for attached, and 1.1 for detached dwellings.
+$TABULA_{rf-rt,a}$ is the average retrofit cost from pre-retrofit BER rating to post-retrofit BER rating for an archetype, data is sourced from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland}.
 However with only 62 samples in total not all retrofits are included.
@@
 ${TIM_{r,a}}$ is the number of dwellings by rating and archetype and ${TIM_{a}}$ is the total number of dwellings of that archetype, however A-rated dwellings are excluded from the total, as it is assumed they cannot be retrofitted further.
-
-\par The cost data for different electrical appliances was sourced from Topten EU database.
+So $ {\text{\euro}}{}_{a,t}$ is the average cost per type of retrofit per archetype, this value is used in TIM.
+
+\par The cost data for different electrical appliances is sourced from Topten EU database.
 Topten has a large database covering electrical appliances available in the European market, the data collected includes cost and efficiency.
@@
 The previous section outlined the cost of retrofitting, and this section looks at the expected savings of retrofitting existing dwellings and building new dwellings.  
-The annual heating costs per archetype and per BER rating are sourced from SEAI's indicative energy running costs and the internal temperature assumptions in Table \ref{Table:InternalTemps} are included to calculate the costs, which are outlined in Table \ref{Table:FuelCost}
+The annual heating costs per archetype and per BER rating are sourced from SEAI's indicative energy running costs and the internal temperature assumptions in Table \ref{Table:InternalTemps} is included to calculate the costs, which are outlined in Table \ref{Table:FuelCost}
 
@@
   \footnotesize
-  \caption{Heating fuel cost per BER rating and archetype }
+  \caption{Annual heating fuel cost per BER rating and archetype }
     \begin{tabular}{l|lcc}
@@
 \begin{split}
-Saving(\%)_{a,d} = Average(Saving(\%)_{a,rf}) * \frac{TIM_{r,a}} {TIM_{a}}
+Saving(\%)_{a,t} = Average(Saving(\%)_{a,rf}) * \frac{TIM_{r,a}} {TIM_{a}}
 \end{split}
@@
 The energy service demands in the residential sector are driven by the total dwelling stock for non-archetype energy services and apartment, attached or detached dwelling stock for archetype energy services.
-The residential stock projections up to 2040 are taken from Economic and Social Research Institute's (ESRI's) housing demand estimates \cite{Bergin2020RegionalLevel}, beyond 2040 the growth rate is extrapolated.
-However, archetpye specfic growth is limited or unavailable, for this reason a simple linear regression, shown in Eq.\ref{Eq:RegArchetype} is used to calculate the share of an archetype in a given year (y).
-Figure \ref{fig:DwellingStoc} graphically represents the number of existing and new dwellings per archetype.
+\par The residential stock projections up to 2040 are taken from Economic and Social Research Institute's (ESRI's) housing demand estimates \cite{Bergin2020RegionalLevel}, beyond 2040 the growth rate is extrapolated.
+However, archetpye specfic growth is limited or unavailable, for this reason a simple linear regression, as shown in Eq.
+\ref{Eq:RegArchetype} is used to calculate the share of an archetype in a given year (y).
+This is graphically represented in Figure \ref{fig:DwellingStoc}.
 The growth rate of apartments is the highest and increasing up to 2040, while attached and detached dwellings are decreasing, indicating a preference for smaller dwellings to increase in the future.
-The obsolescence rate of existing dwelling is 0.35\%.
+The obsolescence rate of the existing dwellings is 0.35\%.
+
+\begin{figure}[!htbp]
+ \centering
+ \caption{Number of dwellings by type and age ('000)}
+ \includegraphics[width=\textwidth]{Figures/TIM_DwellingStock.png}
+ \label{fig:DwellingStoc}
+\end{figure}
 
@@
 
-\begin{figure}[!htbp]
- \centering
- \caption{Number of dwellings by type and age ('000)}
- \includegraphics[scale=0.42]{Figures/TIM_DwellingStock.png}
- \label{fig:DwellingStoc}
-\end{figure}
-
-%\begin{table}[htbp]
-%  \centering
-%  \footnotesize
-%  \caption{Dwelling Growth}
-%\begin{tabular}{l|lcc}
-%Years & Apartment & \multicolumn{1}{l}{Attached } & %\multicolumn{1}{l}{Detached} \\ \hline
-%2016-2020  & 2.89\%   & 1.69\%     & 1.58\%     \\
-%\hline
-%2020-2030     & 4.26\%   & 1.15\%     & 0.81\%   \\
-%2030-2040     & 3.33\%   & 0.89\%     & 0.53\%   \\
-%2040-2050     & 2.47\%   & 0.52\%     & 0.14\%   \\
-%\end{tabular}
-%\label{Table:DwellingGrowth}
-%\end{table}
-
 Archetype energy service demands are dependant on both the type of dwelling and whether it is existing or new.
@@
   \footnotesize
-  \caption{Archetype Energy Service Demands}
+  \caption[width=\textwidth]{Archetype Energy Service Demands}
  \begin{tabular}{c|cc|cc|cc}
@@
 
-Space heating, water heating, and lighting energy service demands for existing dwellings are calculated from the average values of the base year dwelling stock, and as all new dwellings are required to be A-rated, the new dwellings are the average for the existing A-rated dwellings.
-The Pumps \& Fans are similar, except in the new dwellings the value is halved and the remaining half is added to space cooling.
+Space heating, water heating, and lighting energy service demands for existing dwellings are calculated from the average values of the base year dwelling stock, and as all new dwellings are required to be A-rated, the energy service demand for new dwellings in TIM is calculated from the average of the existing A-rated dwellings.
+The Pumps \& Fans energy demands are calculated in a similar method, except in the new dwellings the value is halved and the remaining half is added to space cooling.
 This is an assumption which was cross-checked with POTEnCIA (Policy Oriented Tool for Energy and Climate Change Impact Assessment) space cooling projections per household.
-
-%HDD and warmer weather to offset each other???
-%Heterogeneous Occupancy Behaviour > House Stock 
-%Residential Energy was >50\%higher before internal temperature adjustments
-%Macroeconomic 
-%no elasticity - fuel prices, and weather compensation?
-%Time slices - 40 etc.
 
@@
 
-All technologies are archetype specific and some technologies deliver both water and space heating, while other technologies can only deliver one type of heating, the list of available technologies is outlined in \ref{sec:appendixA4}.
-Retrofits are included as they satisfy the residential space heating demand, without consuming fuel.
-The Danish Energy Agency's Technology Data for Individual Heat Production dataset was input to TIM, this dataset is also used in the EU TIMES model, the dataset has high granularity and is continuously updated.
-
-The number of main heating systems equals the number of dwellings, this assumption ignores any impact of secondary system heating.
-One expection in the rule is , which are only secondary heating systems.
-
-The base year heating energy consumption uses census data \cite{2016CensusIreland} from the CSO dataset E1008 is used to obtain the fuel used in heating systems per archetype.
-The fuel usage is calibrated using the SEAI Energy Balance 2018.
+All space and water heating technologies are archetype specific and some technologies deliver both water and space heating, while other technologies can only deliver one type of heating, the list of available technologies is outlined in \ref{sec:appendixA4}.
+Retrofits are included as they satisfy the residential space heating demand, even though retrofits do not consume fuel similar to a PV system, from that perspective.
+Danish Energy Agency's Technology Data for Individual Heat Production dataset was used  was input to TIM \cite{JointResearchCentre2017Techno-economicEU}, this dataset is also used in the EU TIMES model, the dataset has high granularity and is continuously updated.
+
+In TIM, the number of main heating systems equals the number of dwellings, this assumption ignores any impact of secondary system heating.
+One exception to the rule is stoves, which can only be used as secondary heating systems.
+
+The base year heating energy consumption is derived from the BER database, but uses census data \cite{CentralStatisticsOffice2016CensusIreland} from the CSO dataset E1008 to calibrate the fuel used in heating systems per archetype.
 
@@
 Refrigeration, cloth washing, cloth drying, and dish washing are under the appliances category and data is sourced from Topten \cite{Michel2016EnergyREPORT} and SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
-Cooking is also sourced from Eurostat's ``Energy consumption in households by type and end-used" excel datasheet and cross-checked with SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}  Appliances not included in this list are simply under the ``Other Appliances" category in TIM and these final values are cross-checked with POTEnCIA.
+Cooking is also sourced from Eurostat's ``Energy consumption in households by type and end-used" excel datasheet \cite{Eurostat2021Disaggregatednrg_d_hhq} and cross-checked with SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
+\par Appliances not included in this list are simply under the ``Other Appliances" category in TIM and these final values are cross-checked with POTEnCIA.
 Other Appliances include television, multimedia, and computer technology.
 Lighting and pumps \& fans are archetype specific energy services as the data is sourced from the BER database.
-Equation \ref{Eq:Technology} is also used, but unlike the heating energy service demands, there is little or no choice in different technologies (k) for some of the demand categories (i), therefore, the model has no choice but to invest in a technology, however  cost and efficiency of each technology does determine the overall investment.
+Equation \ref{Eq:Technology} is also used to find the technology in the energy service demand, but unlike the heating energy service demands, there is little or no choice in different technologies (k) for some of the demand categories (i), therefore, the model is restricted in lighting, appliances and pumps \& fans.
+However, the cost and efficiency of each technology does determine the overall investment.
 
@@
  \item Maximum 60\% of existing buildings to be retrofitted by 2030, and 95 \% by 2070. Maximum \% interpolated between 2030 and 2070. 
- \item A prerequisite electrical heat pump requirement is applied to each archetype, the percentage of dwellings which requires a shallow or deep retrofit before a heat pump can be installed (  B2 rating for heat pump ready dwelling ).
-Hybrid heat pumps are not included in this constraint.
+ \item A prerequisite requirement for electrical heat pump installation is applied to each archetype.
+The percentage of dwellings which requires a shallow or deep retrofit before a heat pump can be installed (  B2 rating for heat pump ready dwelling ).
+Hybrid heat pumps are not included in this constraint, as they can supply higher water temperatures without significantly reducing performance.
  \begin{table}[htbp]
@@
 \end{table}%
- \item The natural gas grid is currently connected to 39\% of apartments, 52\% of attached dwellings and 12\% of detached dwellings.
-The total share of dwellings connected to the gas grid is an upper constraint to grow by 50\% for each archetype.
+ \item For space and water heating, each archetype has an upper gas growth constraint of 40\% higher than the base year value.
 \item Maximum fuel share in cooking for gas and LPG is 40\% and 10\% respectively. Base year fuel share for Gas is 32\% and LPG is 1.5 \%.
@@
 
-The newly developed TIM provides some valuable insights on each sector and the sectoral interconnections.
-The results outlined in this paper are based upon six scenarios.
-While TIM can only analysis the energy system, the other large source of GHG emissions is from agriculture, therefore some scenarios are labelled based on agriculture GHG reduction percentage and Energy System GHG reduction percentage, the six scenarios are:
-\begin{itemize}
+The results from the newly developed TIM provide some valuable insights on each sector and the sectoral interconnections.
+The results outlined in this paper are based upon six scenarios, which are focused around the 2030 target.
+While TIM can only analysis the energy system, the other large source of GHG emissions is from agriculture.
+Therefore, some scenarios are labelled based on agriculture 2030 GHG reduction percentage and energy system 2030 GHG reduction percentage, below is a brief description of the six scenarios:
+\begin{enumerate}[(a)]
     \item \textbf{No Mitigation} Least cost solution, where no emissions constraints are imposed
-    \item \textbf{With Additional Measures (WAM)} includes the implementation of Ireland’s CAP 2019.
-    \item \textbf{A25E65} Agriculture GHG emissions reduce by 25\%, Energy system GHG emissions reduces by 65\% by 2030 compared to 2018.
-    \item \textbf{A33E61} Agriculture GHGs reduces by 33\% and Energy system GHGs by 61\%
-    \item \textbf{A40E57} Agriculture GHGs reduces by 33\% and Energy system GHGs by 61\%
-    \item \textbf{A51E51} Equal GHG reduction share in of 51 \% in Agriculture and Energy Systems. 
-\end{itemize}
-
-While the first two scenarios do not comply with the Climate Action and Low Carbon Development (Amendment) Act 2021 target of an overall 51 \% GHG reduction in 2030 compared to 2018 or net-zero by 2050, however they do provide insights on the incremental steps required.
+    \item \textbf{With Additional Measures (WAM)} includes the implementation of Ireland’s CAP 2019 towards 2030 and net-zero in 2050.
+    \item \textbf{A25E65} Agriculture GHG emissions reduce by 25\%, energy system GHG emissions reduces by 65\% by 2030 compared to 2018 \& net-zero in 2050 
+    \item \textbf{A33E61} Agriculture GHGs reduces by 33\% and Energy system GHGs by 61\% \& net-zero in 2050 
+    \item \textbf{A40E57} Agriculture GHGs reduces by 40\% and energy system GHGs by 57\% \& net-zero in 2050 
+    \item \textbf{A51E51} Equal GHG reduction share in of 51 \% in Agriculture and energy Systems \& net-zero in 2050 
+\end{enumerate}
+
+While the first two scenarios do not comply with the Climate Action and Low Carbon Development (Amendment) Act 2021 target of an overall 51 \% GHG reduction in 2030 compared to 2018, they do provide insights on the incremental steps required to achieve these targets.
 The last four scenarios comply with 2030 and 2050 targets and offer insights about different trajectories to achieve those targets.
-Under CAP 2019 agriculture, GHG emission reduction is 10-15\% on the projected levels in 2030 relative to 2017.
-The higher agriculture ambition explored GHG reductions of 25-51\%.
-However, the energy consumed by agriculture is covered in this model.
-
-Figure \ref{fig:$CO_2$ Waterfall} provides an overview of  scenario affects on each sector in 2030, as the energy system ambition increases incrementally from left to right.
-
-\begin{figure}[!htbp]
- \centering
- \includegraphics[width=\textwidth]{Figures/CO2Waterfall.png}
- \caption{$CO_2$ Waterfall}
- \label{fig:$CO_2$ Waterfall}
-\end{figure}
-
-The share of sectoral $CO_2$ emissions is not evenly spread in 2018 and the optimal share of sectoral $CO_2$ reduction in 2030 is also unevenly distributed.
-In the A51E51 scenario, the optimal electricity annual emissions reduction is $7.1 MtCO_2$ in 2030, while transport currently emits the similiar $CO_2$ levels only reduces by $4.5 MtCO_2$.
-The residential sector reduces $CO_2$ emission by 63\% in A51E51 in 2030, electricity is 60 \% and transport is 38\%.
-Not only does Figure \ref{fig:$CO_2$ Waterfall} show the different optimal sectoral reductions, but also the ``low hanging fruit" in the large electricity reduction to the    ``high coconuts" in the step from A33E61 to A25E65, transport mitigates the majority of $CO_2$ emissions required.
-
-Zooming in and focusing on the fuel-mix within the residential sector gives further insights on the optimal pathways to achieve 2030 targets, Figure \ref{fig:ResFuel} shows this, with the base year stacked chart on the top row, the four climate bill compliant scenarios on rows 2-5 and the last two rows are non-compliant.
+
+\subsection{Residential Results}
+\label{Results:Residential }
+The residential sector reduces $CO_2$ emissions by between 63\% - 71\% in 2030 for the four scenarios which comply with the Climate Bill.
+Compare this to the ``No Mitigation" scenario, which shows an increase of 4 \% $CO_2$ in 2030 and the  ``WAM" scenario which shows a 8 \% reduction in $CO_2$ by 2030.
+These results assist in representing the scale of decarbonisation required in the residential sector.
+
+Zooming in on the residential fuel-mix gives further insights on the optimal pathways to achieve 2030 targets.
+Figure \ref{fig:ResFuel} shows this, with the base year being represented on the top row, the four Climate Bill compliant scenarios on rows 2-5, and the last two rows are not compliant with the Climate Bill.
 
@@
 
-In 2018, the residential sector had 124 PJ of primary energy that figure needs to decrease by 8-13 \% in 2030, to comply with climate targets.
-While Figure \ref{fig:$CO_2$ Waterfall} shows the large $CO_2$ reduction in electricity, this translates to a Figure \ref{fig:ResFuel} representing a large uptake of electricity in the residential sector within climate bill compliant scenarios.
-The large increase in ambient energy indicates the type of technology deployed in the residential sector, more on this in Figure XXX.
-The overall share of ambient heat increases from 1.4 \% in 2019 to between 13-26 \% in 2030.
-as a fuel  In climate bill compliant scenarios, peat and coal are completely phased out, while oil use decreases 90-93 \% by 2030.
-Gas energy increase by 34 \% in the A51E51 scenario, but remains relatively stable among other climate bill compliant scenarios in 2030.
-By comparing compliant and noncompliant scenarios, we can observe the favoured fuels and technologies, while also observing the incremental steps of emission reduction and overall cost change in each scenario.
-
-While Figure \ref{fig:ResFuel} shows the primary energy reduction required, it must be noted that retrofits are used to reduce the primary energy required ( reduced energy through behavioural change is not modelled ).
-Figure \ref{fig:Retrofits} shows both the number of retrofits per type and the energy savings for each of the four climate bill compliant scenarios.
+In 2018, the residential sector had 124 PJ of primary energy, which is needed to decrease by 8-13 \% in 2030, to comply with the Climate Bill.
+Results show a large uptake of electricity in the residential sector within Climate Bill compliant scenarios.
+The large increase in ambient energy indicates the type of technology deployed in the residential sector,
+the overall share of ambient heat increases from 1.4 \% in 2018 to between 13-26 \% in 2030, more on this in Figure \ref{fig:HeatPump}.
+In Climate Bill compliant scenarios, peat and coal are completely phased out, while oil use decreases 90-93 \% by 2030.
+Gas energy increase by 34 \% in the A51E51 scenario, but shows little change in other Climate Bill compliant scenarios in 2030.
+
+Figure \ref{fig:Retrofits} shows both the number of retrofits per type and the energy savings up to 2030, for each of the four Climate Bill compliant scenarios.
 
@@
 
-The number of retrofits required varies from 393,000 in A51E51 to 615,000 in A25E65, meaning between 23 - 36 \% of existing dwellings will require some form of retrofit.  
+The number of retrofits varies from 393,000 in A51E51 to 615,000 in A25E65, meaning between 23 - 36 \% of existing dwellings will require some form of retrofit.  
 As mentioned in Table \ref{tab:HeatPumpConstraint}, the rollout of retrofits constraints the amount of heat pumps which can be installed, however, a certain share of existing dwellings and all new dwellings are heat pump ready, i.e B2 rating or above.
+
 \begin{figure}[!htbp]
@@
 
-Where SW is space and water heating, HC is space heating and space cooling, and A-to-W represents air-to-water type heat pumps.
+In Figure \ref{fig:HeatPump}, SW represents space and water heating, HC is space heating and space cooling, and A-to-W represents air-to-water type heat pumps.
 The optimal number of heat pumps in each scenario ranges from 492,000 in A51E51 to 938,000 in A25E65, which equates to 23 - 45 \% of dwellings having a heat pump in 2030.
 Close to 45,000 air to water heat pumps are installed in apartments across all scenarios, as there are no retrofits in apartments, these heat pumps are installed in the existing B2 rating or above apartments or new apartments.
-The detached air-to-water heat pumps are the most prevalent across scenarios and maybe considered the low hanging fruit.
-
+The detached air-to-water heat pumps are the most prevalent across scenarios and maybe considered the ``low hanging fruit".
+
+\subsection{Intersectoral Results}
+\label{Results:Intersectoral }
 The advantage of an energy system model is to observe the intersectoral connections.
-Figure \ref{fig:SectorFuel} presents  the base year primary energy for four sectors - power generation, transport, services, and residential.
-Additionally, the primary energy of fuels in 2030 is included in the four scenarios which comply with the 51 \% reduction scenarios in 2030.
+Figure \ref{fig:SectorFuel} presents  the base year primary energy fuel-mix for four sectors - power generation, transport, services, and residential and the 2030 fuel-mix for the four Climate Bill compliant scenarios.
+
 \begin{figure}[!htbp]
@@
 The least ambitious A51E51 scenario shows 59 PJ of electricity used in residential in 2030, but as constraints tighten, the electricity is used more in transport, while the residential sector invests more in solar panels and other forms of renewable energy.
-While the use of bioenergy in transport and services also affects the residential sector as bioenergy is limited.
-Transport uses 16 PJ of natural gas in A51E51 to 0 PJ in A25E65, this allows the residential sector to use natural gas instead, which moves from 33 PJ in A51E51 to 24 PJ in A25E65.
-
-As a cheaper natural gas powered vehicle may effect the optimal energy mix and the annualised cost in other sectors.
-Figure \ref{fig:SectorCost} shows the annual costs of no mitigation in 2030 and the incremental increase of 2030 sectoral costs of the four scenarios which comply with the 51 \% reduction scenarios in 2030.
+Because bioenergy is limited, the use of bioenergy in transport and services also affects the residential sector.
+
+By examining the same results further along the time horizon in 2050, we can begin to get a sense of the action required pre-2030 to help achieve net-zero in 2050.
 
@@
  \centering
- \includegraphics[width=\textwidth]{Figures/CostWaterfall.png}
- \caption{Sector Annual Cost for five scenarios in 2030}
- \label{fig:SectorCost}
+ \includegraphics[width=\textwidth]{Figures/Intersectoralfuel2050.png}
+ \caption{Sector Annual Fuel in 2018 and four scenarios in 2050}
+ \label{fig:SectorFuel2050}
 \end{figure}
 
-The no mitigation scenario is expected to cost $\text{\euro} 8.1 $ billion in 2030, that figure rises to $\text{\euro} 12.4 $ billion under the A51E51 scenarion, with electricity ( Power generation ) required an extra $\text{\euro} 2 $ billion in 2030.
-As the scenario constraints tighten, larger investments can be seen in the transport sector.
+The results in Fig.
+\ref{fig:SectorFuel2050} show a growth of low carbon electricity in power generation in 2050 compared to 2030 results in Fig.
+\ref{fig:SectorFuel}.
+The results indicate increased electrification pre-2030 across all sectors and a substantial increase of energy efficiency, as primary energy does not correlate with expected growth.
+
+The ``No Mitigation" scenario is expected to cost $\text{\euro} 8.1 $ billion in 2030, that figure rises to $\text{\euro} 12.4 $ billion under the A51E51 scenario, with electricity ( Power generation ) requiring an extra $\text{\euro} 2 $ billion in annual investment in 2030.
+\par Moving from ``No Mitigation" to A51E51, the residential sector requires $\text{\euro} 0.7 $ billion annual extra investment in 2030 or 16.3 \% of the extra investment across the energy system.
+While moving from ``No Mitigation" to the most ambitious energy system scenario, A25E65, would require  $\text{\euro} 1.5 $ billion annual extra residential investment in 2030 or 15 \% of the extra investment across the energy system.
 
@@
 The methodology applied to the residential sector in TIM is transparent, up-to-date, and robust.
-The results outlined have summarised costs, fuel-mix, and intersectoral effects within the energy system.
-The results provide insights, further scenarios or sensitivity analysis can provide further insights from TIM.
-In addition, the results can be used in parallel with simulation models to examine specific climate policies or can alternative scenarios.
-
-The model brings alot of value to the sectroal carbon budgets and long-term climate targets, while also including more technologies and detail than any previous Irish ESOM.
-
-Gas is a favoured fuel from Fig \ref{fig:RESFuel} this provides insights on the potential to decarbonise
-
-%o	Focus on what added value the model brings
-%o	Also include future development priorities here
+The previous version of TIM, namely, the Ireland TIMES model, which was used for almost 10 years in Ireland, and was continuously tweaked to react to the ever changing climate policy, climate targets, and technological advancements.
+\par TIM is developed to provide insights on sectoral carbon budgets and 2030 and 2050 climate targets.
+As the first iteration of TIM, the core model is expected to be improved, while specific research will tweak the core model to specifically analysis certain sectors.
+The large increase in electricity may require some soft-linking with PLEXOS to investigate power generation operating constraints and reliability of supply.
+While LEAP is another powerful tool within the Energy Policy and Modelling Group of MaREI, which can specifically examine climate policy.
+\par The model will require some future developments in the residential sector, district heating, and renewable gas will require improvements.
+The large growth of data centres expected in Ireland can provide waste heat to DH networks.
+While the capacity of biometheane in Ireland can effect the potential decarbonisation of natural gas, adding biomethane to the natural gas grid has not been accounted for in the model.
+\par The dissagreation of archetypes by BER ratings will add to the model complexity, but help provide insights into optimal BER ratings to target in retrofitting and heat pump installation.
+While the dissageration of the model into 26 regions will provide spatial insights.
+The data in TIM is setup to allow for this, but due to the complexity the model was simplified for the initial release.
 
@@
 
-First iteration - of the model will need further development, DH, BEhavour, higher granularity, and spatial will be easily done with the dataset. 
-%What does new model do that previous didn’t?
-%TIM residential improvements - DH, HVO, Solar ( in progress ), better graps the cost and savings of retrofits and heat pumps, which can infrom MACC.Build with 2018 baseyear, not 2010 as previous versions had - this suits the 2030 target
-%THe implications of net-zero
-%Does new policy context require new modelling approach?
-%Yes, do account for ongoing investments in climate mitigation opitions, the approach should provide higher granularity for likely pathways
-%How does new modelling structure provide insights (give examples) in a way that previous model didn’t?
-%Demands - 
-%Improved more detailed Energy Service Demands
-Detailed bottom-up model - using the 2018 base year ( in line with policy )
-Updated technology
-Combines energy system effects
-Robust filteorange input data about housing stock
+The model will be open sourced and be used to develop research and reports, while also used for insights by the Climate Change Advisory Council and the Irish government.
+During development of the model, the developers reached out to experts for feedback on results and the process will continue as the model is improved by a resourced modelling team.
+The model provides key insights into the cost and scale of retrofits and heat pumps required, which are currently being addressed under CAP2019 and are therefore are likely pathways towards achieving climate targets.
+TIM also takes into account new technologies such as DH, Hydrotreated Vegetable Oil (HVO), and solar.
+The model horizon is up to 2050, so the incremental steps and targets on the way to net-zero can be analysed.  
+Moreover, the usual advantage of using ESOM over bottom-up residential models applies to TIM, as the intersectoral effects can be explored.
+The joint effort of different researchers across different universities in Ireland, to output a dwelling stock, using a new filtering process to the BER database helps provide a more robust base year dwelling stock, this filtration process is expected to be improved and TIM can benefit from this process.
+The internal temperature calculation applied in TIM reflects the actual residential energy consumption.
 
@@
 \label{Dis:Limit}
-%o	What can’t the model do? Where is it more suitable to use another model - Hannah Email
-Cost of buildings - bER ratings..
-BER Database - No PV Solar capacity, No Heat Pump ready, Large defaults, wall types check?)
-Community Retrofits - Better Energy Communities is our national retrofit initiative with grant support of up to €28 million for 2021.
-We support new approaches to achieving energy efficiency in Irish communities.
-%Discrepancies were investigated, and assumptions reviewed intuitively in order to achieve a good match.=
-Consumers’ final investment decision can be
-represented by “willingness-to-pay” curves – for
-instance, around 10\% of owner-occupiers are willing
-to invest in energy efficiency measures for a payback
-time of 4 years, falling to 0\% for a payback time of
-6 years for some consumer cohorts
-Type of Tenure is ignoorange
-BER aggeragateed
-housing stock u-values
-behaviour and occupancy
-Area
-expected energy consumption weakness of BER database as 
-Actual energy consumption for dwellings of
-similar or near identical construction may vary by up to 600 \%
-depending on usage \cite{Sunikka-Blank2012BuildingConsumption}
-
-Retrofitting ``rebound" saving  energy is more difficult to measure like improved physical and mental health.
-ArDEM SQL - exogenous behavioural effects 5\% increase of residential energy in 2030 when internal temperatures are adjusted. 
+
+TIM is not a model to look at specific policies nor is it ideally suited to explore occupancy behaviour, instead it provides insights on optimal technological pathways.
+TIMES is an ESOM which assumes perfect foresight, we know this not to be entirely true, consumer choice is not only based on costs.
+The results in TIM show large changes from oil, peat, and coal to heat pumps pre-2030, while it maybe economical beneficial over the lifetime of a heat pump, there maybe financial or social barriers associated with investing in a heat pump and the hurdle rates in the residential sector need more focus to represent consumer choice.
+As previously mentioned, residential oil tanks were filled with low oil prices in 2020, because each consumer is responsible for their own investments and similarly consumers will unlikely invest in new low carbon heating technologies when their existing heating technologies are not at the end of life.
+While it may be optimal for the energy system, it may not be optimal for the consumer.
+\par Primarily due to the lack of reliable data surrounding heat pump performance in B3 or lower rated dwellings, heat pumps can only be installed in B2 or above rated dwellings, similarly heat pump grants supplied by SEAI are for B2 rated or above dwellings.
+In future iterations of TIM, it could be helpful to account for heat pumps in C or D rated dwellings, while the efficiency will be less, it could still provide part of the optimal pathway.
+
+TIM does not compensate for the warmer weather predicted, the change in floor area of a new dwelling, change in occupancy behaviour, or type of tenure.  
+Nor does TIM take account of the effect of higher BER ratings on property value, this could underestimate the role of retrofits in optimal decarbonisation. 
+Despite the filtration process in the BER database, limitations of the data exist, such as PV capacity, internal temperature assumption, default values, and heat loss indicator value ( which is a more accurate way of predicting a heat pump performance ).
+
 \section{Conclusions}
 \label{Conclusion}
-This paper reports results on an ambitious mitigation target in the period to 2050 for the Irish energy system. The analysis has been performed using the Irish TIMES model
-\textit{We stress the importance of differentiating between technical con-
-straints and behavioural constraints in optimisation models.
-Including both conflates what is consideorange desirable under the economic welfare paradigm (i.e.
-cost-minimising decisions) with “realistic decisions”, complicating the interpretation of the results.
-Optimisation models are by design well suited to study cost-optimal behaviour, and running scenarios just with technical constraints facilitates the identification of areas where market barriers are more entrenched, and interventions to accelerate the uptake of low-carbon practices may be needed.} \cite{Astudillo2017CanModelb}
+The discussion section outlines the improvements which will be made to the residential sector in TIM and some of the strengths and weaknesses.
+To conclude, the residential sector in TIM will be able to provide valuable insight into future research questions like ``How do changes to future technology developments affect the optimal residential heating decarbonisation pathway?"
+
 \section{Acknowledgements}
 \label{sec:Acknowledgements}
-Tomas
-Ankita%CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC
+Tomás Mac Uidhir worked in parallel with the TIM residential sector development on the LEAP residential model development, and he also started a new Ireland-wide group, which I joined with the aim of filtering the BER database.
+Ankita Singh Gaur worked on the residential demands and provided some time to assist with the demand calculations.
+CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC.
 % https://www.marei.ie/project/capacity/
-CHIMERA is supported by a research grant from Science Foundation Ireland (SFI) and the National Natural Science Foundation of China (NSFC) under the SFI-NSFC Partnership Programme, grant no.
+%CHIMERA is supported by a research grant from Science Foundation Ireland (SFI) and the National Natural Science Foundation of China (NSFC) under the SFI-NSFC Partnership Programme, grant no.
 17/NSFC/5181 and
-supported by MaREI, the SFI Research Centre for Energy, Climate, and Marine [Grant No: 12/RC/2302\_P2]
-
-https://www.marei.ie/project/chimera/
+%supported by MaREI, the SFI Research Centre for Energy, Climate, and Marine [Grant No: 12/RC/2302\_P2]
+
+https://www.marei.ie/project/capacity
+\newpage
+
 \appendix
 
-\newpage
 \section{}

### 2021-08-09 manuscript.tex
9 Aug 2021

@@
 
-\title{Sustainable Pathways for Irish homes in changing TIMES}
-%TIMES Ireland Model: Residential Sector
-% The changing TIMES of Irish Homes
-% Developing sustainable homes in changing TIMES
+\title{Developing Decarbonisation Pathways for Irish homes in changing TIMES}
 
@@
 \author[inst1,inst2]{Olexandr Balyk}
-\author[inst1,inst2]{Hannah Daly}
-\author[inst1,inst2]{ \& Brian Ó Gallachóir}
+\author[inst1,inst2]{Tomás Mac Uidhir}
+\author[inst1,inst2]{Ankita Gaur}
+\author[inst1,inst2]{Brian Ó Gallachóir}
+\author[inst1,inst2]{\& Hannah Daly}
 
@@
 %% Text of abstract
-\par Ireland's residential sector has particularly underachieved in previous climate targets.
-Located on the peripherals of Europe, with a high share of low thermal efficient detached dwellings with carbon intensive heating fuels, provides some distinctive barriers to decarbonising Ireland's residential sector.
-The disadvantageous starting position has impeded residential decarbonisation progress thus far.
+
+Ireland has committed to one of the most ambitious decarbonisation targets in the world to 2030.
+The residential sector, responsible for 14.2 \% of total greenhouse gas (GHG) emissions \cite{Howley2005Energy-RelatedIreland.} , is among the most carbon intensive per dwelling in the EU, emitting 60 \% more $CO_2$ than the national EU average \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, because of a high share of low thermal efficient detached dwellings coupled with a high dependency on carbon-intensive heating fuels.
+This disadvantageous starting position has provided some distinctive barriers to Irish residential decarbonisation progress thus far.
 \par
@@
 The TIMES-Ireland Model (TIM) is a newly developed optimisation model for the Irish energy system.
-This paper describes the methodology applied to the development of the residential sector in TIM.
-Some results are analysed to provide input to discuss insights from the residential sector in TIM.
-%\par The Irish residential sector has proved hard to decarbonise, yet recently Ireland has set one of the most ambitious climate targets into law, a 51 \% greenhouse gas reduction in 2030, compared to 2018.
-There is a need for a new up to date and state of the art model to provide insights on the optimal decarbonisation pathway for the Irish energy system.
-This paper describes the methodology applied to the development of the residential sector in the new TIMES Ireland Model (TIM).
-Results are analysed to provide input to discuss insights from the residential sector in TIM.
+This paper describes the methodology applied to the development of the residential sector in TIM and explores the results, which provides insights on optimal residential decarbonisation pathways.
 
@@
 
-The layout of this paper is as follows: 
-\begin{itemize}
-    \item Section \ref{sec:Introduction} of this paper looks at Ireland's climate targets, energy policy, and distinctive residential sector.
-While also examining how researchers have previously used bottom-up residential models and Energy System Optimization Models (ESOMs) to analyse the residential sector.
-    \item Section \ref{LitR} is the core section of this paper and explains the methodology used to develop the residential sector.
-The treatment of input data in the model is described, the calculations are explained, and assumptions and constraints are highlighted.
-    \item Section \ref{Results} outlines the scenarios used and displays results from the residential sector.
-    \item Section \ref{Discussion} discusses the results, strengths, and weakness of the newly developed model and how it differs from other models.
-    \item Section \ref{Conclusion} draws some conclusions and discusses possible future work.
-\end{itemize}
+The Introduction section of this paper reviews both bottom-up residential models and ESOMs, and analyses Ireland's distinctive residential sector, climate targets, and current energy policy.
 
@@
 
+The layout of this paper is as follows: 
+\begin{itemize}
+    \item Section \ref{LitR} is the core section of this paper and explains the methodology used to develop the residential sector.
+The treatment of input data in the model is described, the calculations are explained, and assumptions and constraints are highlighted.
+    \item Section \ref{Results} outlines the scenarios used and displays results from the residential sector.
+    \item Section \ref{Discussion} discusses the results, strengths, and weakness of the newly developed model and how it differs from other models.
+    \item Section \ref{Conclusion} draws some conclusions and discusses possible future work.
+\end{itemize}
 \section{Methodology}
@@
 \label{sec:Acknowledgements}
-Tomás Mac Uidhir worked in parallel with the TIM residential sector development on the LEAP residential model development, and he also started a new Ireland-wide group, which I joined with the aim of filtering the BER database.
-Ankita Singh Gaur worked on the residential demands and provided some time to assist with the demand calculations.
+The authors wish to acknowledge the contributions of Tomás Mac Uidhir, who supported filtering of the BER database, and Ankita Singh Gaur who developed the residential demands. 
 CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC.
+Alessandro and Maurizio provided technical assistants on the development of the residential sector in TIMES. 
 % https://www.marei.ie/project/capacity/

### 2021-08-10 manuscript.tex
Update - 1st draft changes in Intro

@@
 Ireland has committed to one of the most ambitious decarbonisation targets in the world to 2030.
-The residential sector, responsible for 14.2 \% of total greenhouse gas (GHG) emissions \cite{Howley2005Energy-RelatedIreland.} , is among the most carbon intensive per dwelling in the EU, emitting 60 \% more $CO_2$ than the national EU average \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, because of a high share of low thermal efficient detached dwellings coupled with a high dependency on carbon-intensive heating fuels.
-This disadvantageous starting position has provided some distinctive barriers to Irish residential decarbonisation progress thus far.
+The residential sector, responsible for 14.2 \% of total greenhouse gas (GHG) emissions \cite{Howley2020Energy-RelatedIreland.} , is among the most carbon intensive per dwelling in the EU, emitting 60 \% more $CO_2$ than the national EU average \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, because of a high share of low thermal efficient detached dwellings coupled with a high dependency on carbon-intensive heating fuels.
+These factors have served as barriers to the decarbonisation of the residential sector in Ireland to date.
 \par
-Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges, which provides insights to decision makers on issues related to climate and energy policy.
+Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges, which provides insights to decision-makers on issues related to climate and energy policy.
 \par
 The TIMES-Ireland Model (TIM) is a newly developed optimisation model for the Irish energy system.
-This paper describes the methodology applied to the development of the residential sector in TIM and explores the results, which provides insights on optimal residential decarbonisation pathways.
+This paper describes the methodology applied to the development of the residential sector in TIM and explores the results, which provides new insights on optimal residential decarbonisation pathways.
 
@@
 
-\subsection{Residential Modelling}
+\subsection{Energy Systems Modelling and the Residential Sector}
 \label{Intro:ESOM}
 
- ESOMs have been used to explore the least cost pathways, as they consider a large number of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
-Some of the main ESOMs currently used include MESSAGE \cite{Messner1995Model-basedPlanning}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOMs cover all energy-related sectors, they can also be used to specifically explore certain economic sectors in higher granularity.
-This section gives some examples of residential sector studies using ESOMs.
-
-TIMES has been extensively used in both the UK and Ireland, and there are several examples of using TIMES to focus on the residential sector.
+Technological advancements in residential heating appliances are being assisted by the exponentially increasing computational capacity over time, and with new low-carbon fuel alternatives and new heating technologies becoming available, the residential decarbonisation pathways are dynamic.
+ESOMs have been used to explore economically optimal or least cost decarbonisation pathways, ESOMs have a rich representation of existing and future technological variables such as efficiency, cost, and fuel.
+ESOMs provide enough final energy to at least satisfy energy service demands, while also accounting for complex sectoral interconnections.
+\par Some of the main ESOMs currently used include MESSAGE \cite{Messner1995Model-basedPlanning}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.
+While ESOMs cover all energy-related sectors, they can also be used to explore certain economic sectors in higher granularity.
+
+Different levels of disaggregation are applied to the residential sector in models and research by Natarajana et al.
+\cite{Natarajana2011ModellingApproach} highlighted a common approach used by defining an average performance for a number of dwelling categories, and this approach was adopted here.
+The heterogeneous mix of dwelling types means each dwelling is categorised based on dwelling type to better reflect the housing stock.
+While some models also disaggregate the year of construction \cite{Ahern2013StateStock}, wall type \cite{Dineen2015ImprovedSource} , level of insulation \cite{Clinch2001Cost-benefitEfficiency}, to keep the model as keep as simple as possible but as complex as necessary, these levels of disaggregation were not applied here, however the energy rating of archetypes will be used in future iterations.
+Currently, a weighted average value combines values for each of the three dwelling archetypes used - detached ( stand-alone ), attached ( connected to another dwelling, eg.
+semi-detached, terrace ) and apartment.
+Space and water heating are the main energy services, accounting for over 80 \% of the energy used in dwellings and typically have a lifetime of at least 15 years, consequently leading to a slow turnover of technology in the residential sector.
+The remaining energy services are space cooling, cooking, lighting \& appliances, Other (mainly outdoor appliances), the annual trends of these energy services are shown in Figure \ref{fig:IrelandServiceDemands}.
+
+% Average dwellings -	Would it be worth having a paragraph early in this section that describes some relevant characteristics of the residential sector, e.g.
+very heterogeneous mix of dwelling types (hence the archetype approach), a slow turnover of building stock (hence challenge of very ambitious near term targets), a broad distinction between new and existing dwellings (former very efficient, latter often not efficient), different types of energy use (space heating, water heating, appliances), drivers of energy demand (weather, affluence, demographics), non-economic rationality of dwelling occupants (hence caution required with techno-economic modelling approaches)?
+This might help frame your review.
+- Fionn
+
+%This section gives some examples of how the residential sector has been explored in previous studies using ESOMs.
+
+The Integrated MARKAL-EFOM System (TIMES) has been extensively used in both the UK and Ireland, and there are several examples of using TIMES to focus on the residential sector.
 TIMES assumes perfect foresight to enable a least-cost solution to be calculated, but \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
@@
 Despite the perfect foresight assumption being essential to finding a solution, it has the potential to be a weakness by widening the gap between the model and reality.
-\par A common theme across many ESOMs shows different sectors are increasingly reliant on electricity generation as lower carbon intensity electricity is considered a cornerstone in optimal decarbonisation.
+A growing trend across ESOMs is the focus on retrofitting to thermally improve the residential stock, Leibowicz et al.
+\cite{Leibowicz2018OptimalServices} highlights the importance of energy building regulation as thermally efficient dwellings can reduce residential decarbonisation costs.
+The other main trend observed shows sectors increasingly reliant on low carbon intensity electricity  \cite{Hedegaard2013InfluenceOperation} this is especially true in Ireland as the electrical grid has already begun rapid decarbonisation, while a focus on PhotoVoltaic (PV) systems can be examined in parallel \cite{Marczinkowski2018ResidentialSystems}
+
+%\par A common theme across many ESOMs shows different sectors are increasingly reliant on electricity generation as lower carbon intensity electricity is considered a cornerstone in optimal decarbonisation.
 Using the Open Source Energy Modeling System (OSeMOSYS), Leibowicz et al.
@@
 However, the percentage increase in net present cost diminishes with a more thermally efficient building stock, highlighting the importance of energy building regulation as thermally efficient building stock can reduce residential decarbonisation costs.
-\par District Heating is a proven low carbon heating technology, however it is rarely researched in Ireland and has not been included in previous ESOM in Ireland.
+%\par District Heating is a proven low carbon heating technology, however it is rarely researched in Ireland and has not been included in previous ESOM in Ireland.
 However, Denmark's district heating system has been comprehensively examined.
@@
 This identifies individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments in areas where district heating is not available.
-\par While Ireland's building regulations have encouraged the uptake of photovoltaics (PVs), the market share of PV solar panel systems remains low.
+%\par While Ireland's building regulations have encouraged the uptake of photovoltaics (PVs), the market share of PV solar panel systems remains low.
 The low amount of direct sunlight means Ireland's climate is not ideal for PVs, but as the cost of PVs continues to fall, the potential role of PVs in  residential decarbonisation rises.
-In one study, Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} used EnergyPLAN model to explore the feasibility of PV systems installed at the household level in combination with batteries in Samsø.
-Two approaches are explored for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
+%In one study, Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} used EnergyPLAN model to explore the feasibility of PV systems installed at the household level in combination with batteries in Samsø.
+%Two approaches are explored for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
 Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy system perspective, one reason for this is the wind power could also be storage in communal batteries.
@@
 
-\par The Archetype Dwelling Energy Model (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}.
-ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and then calculates the annual energy consumption using EN 13790 to simulate the potential for improved energy efficiency gains within the existing retrofit programs.
+\par The Archetype Dwelling Energy Model - SQL (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}.
+ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and calculates the annual energy consumption for space/ water heating, and lighting using EN 13790 to simulate the potential for improved energy efficiency gains within an existing retrofit program.
 Mac Uidhir et al.
-specifically examined the Better Energy Homes (BEH) scheme.
-The five most common retrofit combinations are simulated for each of the 9 archetypes ( 3 energy rating types and 3 building types ).
-The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\% can be achieved when comparing the baseline scenario.
-This shows the the type of retrofit and type of house can have a large affect on the energy saved.
-
-\par Finally, it is worth mentioning Sustainable Energy Authority of Ireland's (SEAI’s) National Energy Modelling Framework (NEMF), which is a suite of modelling tools used for modelling the Irish energy system.
+examined the Better Energy Homes (BEH) retrofit grant scheme.
+The five most common retrofit combinations were simulated for 9 distinct archetypes (3 BER energy rating groups, and 3 building types) and compared these to alternative retrofit combinations which delivered additional efficiency improvements.
+The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\%, relative to the set of retrofit combinations actually delivered from the program.
+This shows the the type of retrofit(s) and archetype can have a large affect on the energy saved.
+
+\par Finally, the Sustainable Energy Authority of Ireland's (SEAI’s) National Energy Modelling Framework (NEMF) is a suite of modelling tools used for modelling the Irish energy system.
 The SEAI is Ireland's national energy authority and they play a key role in assisting consumers to reduce energy.
@@
 
-\begin{figure}[!htbp]
+%, Commission Regulation (EU) No 431/2014 amends the Energy Statistics Regulation(ESR) to require all Member States to supply data on energy consumption in households split by the following end-uses: space heating; space cooling; water heating; cooking; lighting and appliances; other uses.
+%The following data concerning the final energy consumption in households by type of end-use are reported according to Regulation (EC) 1099/2008 on energy statistics and its subsequent amendments.
+%Implementation of Commission Regulation (EU) No 431/2014 of 24 April 2014 amending Regulation (EC) No 1099/2008 of the European Parliament and of the Council on energy statistics
+
+\begin{figure}[!ht]
 \centering
 \begin{tikzpicture}
-    \tikzstyle{every node}=[font={\bf}]
-    \pie[   color = { yellow!90!black, green!60!black, blue!60,
-        orange!70,
-        gray!80,
-        teal!20
-    }, explode = 0.6, text=legend, radius=2.4]{61.8/Space Heating, 2.2/Cooking, 18.5/Water Heating, 3.8/Lighting, 12.8/Appliances,0.9/Pumps \& Fans}
+\begin{axis}[
+    scale=1,
+    x tick label style={/pgf/number format/1000 sep=,rotate=45},
+    legend style={at={(0.5,-0.25)},anchor=north,legend columns=2},
+    xlabel={Year},
+    ylabel={Petajoules [PJ]},
+    xmin=2015, xmax=2019,
+    ymin=0, ymax=80,
+    xtick={2015,2016,2017,2018,2019},
+    ytick={0,20,40,60,80},
+    ymajorgrids=true,
+    grid style=dashed,
+]
+
+\addplot[
+    color=red,
+    ]
+    coordinates {
+    (2015,72.5)(2016,72.9)(2017,70.76)(2018,78.59)(2019,73)
+    };
+
+\addplot[
+    color=blue,
+    ]
+    coordinates {
+    (2015,21.54)(2016,22)(2017,23.25)(2018,24.12)(2019,23.84)
+    };
+
+\addplot[
+    color=green!60!black,
+    ]
+    coordinates {
+    (2015,2.55)(2016,2.57)(2017,2.64)(2018,2.7)(2019,2.7)
+    };
+\addplot[
+    color=blue!60,orange!70,
+    ]
+    coordinates {
+    (2015,19.23)(2016,19.211)(2017,19.59)(2018,19.98)(2019,19.95)
+    };
+\addplot[
+    color=black,
+    ]
+    coordinates {
+    (2015,1.046)(2016,1.045)(2017,1.065)(2018,1.086)(2019,1.084)
+    };
+    \legend{Space Heating,Water Heating,Cooking,Lighting \& Appliances, Other}
+
+\end{axis}
 \end{tikzpicture}
-    \caption{Ireland 2018 Energy Service Demands}
-    \label{fig:Ireland2018ServiceDemands}
+\caption{Ireland Residential Energy Service Demands}
+    \label{fig:IrelandServiceDemands}\cite{Eurostat2021Disaggregatednrg_d_hhq}
   \end{figure}
+
+% Other is  mainly covering uses of energy by households outside the dwellings themselves - this is included in Appliances in TIM..
 
@@
 Focusing only on energy-related GHG emissions, the residential sector directly accounts for 24\% of GHG emissions.
-Fig.
-\ref{fig:Ireland2018GHG} shows the energy-related GHG disaggregated between transport, heat and electricity.
-
-\begin{figure}[!htbp]
-    \begin{tikzpicture}[every node/.style={align=center}, pin distance=17mm]
-    \pie[cloud,text=inside,radius = 2.4]{34/Agriculture,7/Other,59/Energy-related}
-    \draw [->, line width=4pt] (5.4,-0.3) -- (6.5,-.3);
-    \hspace{8.5cm}
-    \pie[rotate =30,text=inside,radius=1.95,color={black!30,orange!70,yellow!40}]{40/Transport,33/Heat,27/Electricity}
-    \end{tikzpicture}
-    \caption{Ireland 2018 GHG Share \cite{Howley2005Energy-RelatedIreland.}}
-    \label{fig:Ireland2018GHG}
-\end{figure}
-
-Where the residential sector is directly responsible for 47\% of heat and 30\% of electricity GHG emissions.
-The residential sector would also be indirectly responsible for a share of transport GHG emissions, the residential sector is therefore a pivotal sector in decarbonising energy in Ireland.
+The residential sector is directly responsible for 47\% of heat and 30\% of electricity GHG emissions.
+The residential sector would also be indirectly responsible for a share of the remaining energy-related GHGs in the transport sector, the residential sector is therefore a pivotal sector in decarbonising energy in Ireland.
 
@@
 \label{Overview}
+
 \par  The expertise among energy policy researchers in Ireland combined with the global network of Energy Technology Systems Analysis Program (ETSAP) modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways.
@@
 The ArDEM-SQL, which was described in Section \ref{Intro:ESOMIreland} is a simulation model built on space heating calculation EN 13790.
-ArDEM-SQL alters the primary energy consumption of the BER database by changing the internal temperatures.
+ArDEM-SQL alters the primary energy consumption of the BER database by changing the internal temperatures assumed by DEAP (ref to table).
 
@@
 
-Zooming in on the residential fuel-mix gives further insights on the optimal pathways to achieve 2030 targets.
+Focusing on the residential fuel-mix gives further insights on the optimal pathways to achieve 2030 targets.
 Figure \ref{fig:ResFuel} shows this, with the base year being represented on the top row, the four Climate Bill compliant scenarios on rows 2-5, and the last two rows are not compliant with the Climate Bill.
@@
 The methodology applied to the residential sector in TIM is transparent, up-to-date, and robust.
-The previous version of TIM, namely, the Ireland TIMES model, which was used for almost 10 years in Ireland, and was continuously tweaked to react to the ever changing climate policy, climate targets, and technological advancements.
+The previous version of TIM, namely, the Ireland TIMES model, which was used for almost 10 years, and was continuously tweaked to react to the ever changing climate policy, climate targets, and technological advancements.
 \par TIM is developed to provide insights on sectoral carbon budgets and 2030 and 2050 climate targets.
-As the first iteration of TIM, the core model is expected to be improved, while specific research will tweak the core model to specifically analysis certain sectors.
-The large increase in electricity may require some soft-linking with PLEXOS to investigate power generation operating constraints and reliability of supply.
-While LEAP is another powerful tool within the Energy Policy and Modelling Group of MaREI, which can specifically examine climate policy.
-\par The model will require some future developments in the residential sector, district heating, and renewable gas will require improvements.
+As the first iteration of TIM, the core model is expected to be improved, while on-going research will continue to  analyse certain sectors in greater detail.
+The large increase in electricity ?demand?
+may require some soft-linking with PLEXOS to investigate power generation operating constraints and reliability of supply.
+While LEAP is another powerful tool within the Energy Policy and Modelling Group of MaREI, which can be used to analyse individual climate policies.
+\par The model will require some future developments in the residential sector, district heating, and renewable gas.
 The large growth of data centres expected in Ireland can provide waste heat to DH networks.
-While the capacity of biometheane in Ireland can effect the potential decarbonisation of natural gas, adding biomethane to the natural gas grid has not been accounted for in the model.
+While the capacity of biomethane in Ireland can effect the potential decarbonisation of natural gas, adding biomethane to the natural gas grid has not been accounted for in the model.
 \par The dissagreation of archetypes by BER ratings will add to the model complexity, but help provide insights into optimal BER ratings to target in retrofitting and heat pump installation.

### 2021-08-13 manuscript.tex
Done

@@
 
-The Introduction section of this paper reviews both bottom-up residential models and ESOMs, and analyses Ireland's distinctive residential sector, climate targets, and current energy policy.
-
-\subsection{Energy Systems Modelling and the Residential Sector}
+The Introduction section of this paper reviews bottom-up residential models, which have high technological granularity accounting for the residential sector only and ESOMs which have a bottom-up or hybrid approach and includes all energy-related sectors.
+The hybrid approach combines bottom-up technological detail and top-down macroeconomic detail.
+The introduction also analyses Ireland's distinctive residential sector, climate targets, and current energy policy.
+
+\subsection{Energy Systems Optimisation Modelling and the Residential Sector}
 \label{Intro:ESOM}
 
-Technological advancements in residential heating appliances are being assisted by the exponentially increasing computational capacity over time, and with new low-carbon fuel alternatives and new heating technologies becoming available, the residential decarbonisation pathways are dynamic.
-ESOMs have been used to explore economically optimal or least cost decarbonisation pathways, ESOMs have a rich representation of existing and future technological variables such as efficiency, cost, and fuel.
-ESOMs provide enough final energy to at least satisfy energy service demands, while also accounting for complex sectoral interconnections.
-\par Some of the main ESOMs currently used include MESSAGE \cite{Messner1995Model-basedPlanning}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.
-While ESOMs cover all energy-related sectors, they can also be used to explore certain economic sectors in higher granularity.
-
+Economically optimal or least cost residential decarbonisation pathways are dynamic, with technological advancements in heating appliances being assisted by exponentially increasing computational capacity over time, combined with new low-carbon fuel alternatives and new heating technologies becoming available.
+ESOMs have a rich representation of existing and future technological variables such as efficiency, cost, and fuel, while also accounting for macroeconomic effects.
+ESOMs supply enough energy to at least satisfy energy service demands so that the lights stay on, while also considering sectoral interconnections to explore future decarbonisation pathways.
+There are a number of widely used ESOMs, such as MESSAGE \cite{Messner1995Model-basedPlanning}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.
+
+These models can be used to explore the residential sector in higher granularity.
 Different levels of disaggregation are applied to the residential sector in models and research by Natarajana et al.
 \cite{Natarajana2011ModellingApproach} highlighted a common approach used by defining an average performance for a number of dwelling categories, and this approach was adopted here.
-The heterogeneous mix of dwelling types means each dwelling is categorised based on dwelling type to better reflect the housing stock.
-While some models also disaggregate the year of construction \cite{Ahern2013StateStock}, wall type \cite{Dineen2015ImprovedSource} , level of insulation \cite{Clinch2001Cost-benefitEfficiency}, to keep the model as keep as simple as possible but as complex as necessary, these levels of disaggregation were not applied here, however the energy rating of archetypes will be used in future iterations.
-Currently, a weighted average value combines values for each of the three dwelling archetypes used - detached ( stand-alone ), attached ( connected to another dwelling, eg.
-semi-detached, terrace ) and apartment.
-Space and water heating are the main energy services, accounting for over 80 \% of the energy used in dwellings and typically have a lifetime of at least 15 years, consequently leading to a slow turnover of technology in the residential sector.
-The remaining energy services are space cooling, cooking, lighting \& appliances, Other (mainly outdoor appliances), the annual trends of these energy services are shown in Figure \ref{fig:IrelandServiceDemands}.
+To account for the heterogeneous mix of dwellings, each dwelling is categorised based on building type to better reflect the housing stock and provide building specfic insights.
+Some models also disaggregate the year of construction \cite{Ahern2013StateStock}, wall type \cite{Dineen2015ImprovedSource} and level of insulation \cite{Clinch2001Cost-benefitEfficiency}.
 
@@
 
-%This section gives some examples of how the residential sector has been explored in previous studies using ESOMs.
-
 The Integrated MARKAL-EFOM System (TIMES) has been extensively used in both the UK and Ireland, and there are several examples of using TIMES to focus on the residential sector.
 TIMES assumes perfect foresight to enable a least-cost solution to be calculated, but \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
-With this in mind, Li et al.
-\cite{Li2018IncorporatingModel} investigated the heating technology preference in the UK using UK TIMES Model (UKTM) by incorporating a Discrete Choice Model (DCM) to better reflect how homeowners choose heating technologies.
-A nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems.
-The DCM then estimated the probability of heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices.
-Despite the perfect foresight assumption being essential to finding a solution, it has the potential to be a weakness by widening the gap between the model and reality.
-A growing trend across ESOMs is the focus on retrofitting to thermally improve the residential stock, Leibowicz et al.
-\cite{Leibowicz2018OptimalServices} highlights the importance of energy building regulation as thermally efficient dwellings can reduce residential decarbonisation costs.
-The other main trend observed shows sectors increasingly reliant on low carbon intensity electricity  \cite{Hedegaard2013InfluenceOperation} this is especially true in Ireland as the electrical grid has already begun rapid decarbonisation, while a focus on PhotoVoltaic (PV) systems can be examined in parallel \cite{Marczinkowski2018ResidentialSystems}
-
-%\par A common theme across many ESOMs shows different sectors are increasingly reliant on electricity generation as lower carbon intensity electricity is considered a cornerstone in optimal decarbonisation.
-Using the Open Source Energy Modeling System (OSeMOSYS), Leibowicz et al.
-\cite{Leibowicz2018OptimalServices} explores the optimal decarbonisation of the residential sector by examining electricity generation and storage strategies.
-The results find net present costs in the \textit{``Policy scenarios"} ( $CO_2$ declines linearly by 90\% in 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations imposed ).
-However, the percentage increase in net present cost diminishes with a more thermally efficient building stock, highlighting the importance of energy building regulation as thermally efficient building stock can reduce residential decarbonisation costs.
-%\par District Heating is a proven low carbon heating technology, however it is rarely researched in Ireland and has not been included in previous ESOM in Ireland.
-However, Denmark's district heating system has been comprehensively examined.
-One study by Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} used the Balmorel model to analyse how heat pumps with complementing heat storage affect wind power investments in Denmark.
-The results find that air-to-water heat pumps are installed in non-district heating areas, while the system benefits of adding individual heat storage to the heat pumps are only moderate, the main benefit is a reduced need for peak/reserve capacity investments.
-This identifies individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments in areas where district heating is not available.
-%\par While Ireland's building regulations have encouraged the uptake of photovoltaics (PVs), the market share of PV solar panel systems remains low.
-The low amount of direct sunlight means Ireland's climate is not ideal for PVs, but as the cost of PVs continues to fall, the potential role of PVs in  residential decarbonisation rises.
-%In one study, Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} used EnergyPLAN model to explore the feasibility of PV systems installed at the household level in combination with batteries in Samsø.
-%Two approaches are explored for the location of the batteries in the PV system; communal battery system or an individual residential battery approach.
-Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy system perspective, one reason for this is the wind power could also be storage in communal batteries.
-When identical communal and individual battery capacities where compared, it was found  64\% of residential electricity demand is satisfied by an individual battery and only 36\% by a communal battery.
-While the battery location was explored, the PV system location was not.
-This is a great example of the strength of an ESOM, whereby the residential PV systems affect wind energy investment.
-However, a weakness in the model is the ability to comprehend household occupancy behaviour with and without household batteries on the PV system.
+In light of this, Li et al.
+\cite{Li2018IncorporatingModel} investigated the heating technology preference in the UK using UK TIMES Model (UKTM) by incorporating a Discrete Choice Model (DCM) to better reflect how homeowners choose heating technologies and thus correcting for error in the perfect foresight assumption.
+
+There is a rising tendency for ESOMs to focus on new building fabric \cite{Leibowicz2018OptimalServices,Dineen2011ModellingHeating} and retrofitting to thermally improve the existing residential stock \cite{Dineen2015ImprovedSource}.
+Another trend observed is the electrification of energy-related sectors \cite{Chiodi2013ModellingSystem}, while a focus on photovoltaic (PV) systems can be examined in parallel \cite{Marczinkowski2018ResidentialSystems}.
+Retrofitting and electrification are a fundamental part of residential decarbonisation in Ireland, and therefore play a central role in the development of the model described here.
+Furthermore, the new model includes technologies, which were not previously considered, one such technology is District Heating (DH), some initial studies have found it may be optimal to have 37 \% of heat demand to be provided by DH in Ireland \cite{Thellufsen2019ImplementingIreland}, the full list of space and water heating technologies considered in the model can be found in \ref{sec:appendixA4}.
+The strength of an ESOM can be used to examine the sectoral interactions, one study by Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} examined the interaction between heat pumps and wind generation investments in Denmark, the results find it is optimal to install air-to-water heat pumps in non-district heating areas.
 
@@
 
-ESOMs have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem,Connolly2009DevelopingUniversity.}, however few Irish studies use ESOM models to explore the residential sector, two examples of Irish ESOM focusing on the residential sector are outlined below.
+ESOMs have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem,Connolly2009DevelopingUniversity.}, however no Irish studies use ESOM models to solely explore the residential sector, as many models look across all sectors.
+However, two examples of Irish ESOMs focusing the residential sector in higher granularity are outlined below.
 
 In 2014, Rogan et al.
-\cite{Rogan2014LEAPsSystem} linked OSeMOYOS to the Long Range Energy Alternatives Planning system (LEAP) simulation model to explore specific energy efficiency policies not only in the residential sector, but also in the transport, industry and services sectors in Ireland.
+\cite{Rogan2014LEAPsSystem} linked OSeMOYOS to the Long Range Energy Alternatives Planning system (LEAP), which is a scenario-based simulation modeling tool used to analysis energy policy in all sectors of the economy.
+The models were linked to explore specific energy efficiency policies in the residential and other sectors.
 Three scenarios were explored, namely, the reference scenario, energy efficiency scenario, and energy efficiency+ scenario.
-Where the reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020 and the results show that residential total energy consumption rose 6.7 \% in 2020, compared to 2008.
+The reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020, and the results show that residential total energy consumption rose 6.7 \% in 2020, compared to 2008.
 In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulting in a 9.9 \% reduction of energy consumption in 2020, compared to the reference scenario.
@@
 Rogan et al.
-discussed the barriers, obstacles to investment, split incentives, and information gaps which restrict energy efficiency improvements in the residential sector.
-The importance for the policy to not only look at the number of retrofits, but also the depth of retrofit work, is shown by the difference in scenarios.
+discussed the barriers, obstacles to investment, split incentives, and information gaps which restrict energy efficiency improvements.
+The insights highlighted the importance for the policy to not only look at the number of retrofits, but also the depth of retrofit, as shown by the different results in the scenarios.
 The model also found considerable scope for energy savings to be made through improved building regulations.
@@
 
-\par  The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector models, rather than energy system models specifically examining the residential sector.
-Nevertheless, the residential models have provided insights on the input data, methodology, and results, and some studies are outlined below.
-
-Clinch and Healy analysed different aspects of thermal comfort in the Irish residential sector.
-This was the first comprehensive analysis of the Irish residential sector, and their studies examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure} and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}.
-In 2000, Clinch and Healy published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Microsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating.
-A cost-benefit model (CBM) was fitted to the EAM and the results under the predicted scenario showed energy savings outweighing programme costs, representing a benefit-cost ratio of 1.7.
-The benefits of reduced sickness, hospitalisation, drugs, and mortality were later quantified and this moved the benefit-cost ratio up to 3.0.
-The same model was later used by Clinch, Healy, and King \cite{Clinch2001ModellingEfficiency} to represent the human behavioural component in the computer model.
+\par  The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector models.
+These models have provided insights on the input data, methodology, and results, and some studies are outlined below.
+\par In the first comprehensive analysis of the Irish residential sector, Clinch and Healy examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure}, occupancy behaviour \cite{Clinch2001ModellingEfficiency}, social benefit \& cost \cite{Clinch2001Cost-benefitEfficiency}  and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}.
 
@@
 In 2011, Dineen and Ó Gallachóir \cite{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP).
-The model focused on the energy efficiency of space and water heating by examining technology and building fabric efficiency.
-The building fabric efficiency of newly occupied dwellings developed takes into account the theoretical increase in energy efficiency due to improved 2008 and 2010 building regulations.
-The paper acknowledges that retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperature effect on energy efficiency improvements.
+The model focused on new buildings and examined the energy efficiency of space and water heating by examining technology and building fabric thermal efficiency.
+%The paper acknowledges that retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperature effect on energy efficiency improvements.
+Later in 2015, Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examining new dwellings, by examining the energy savings potential of energy efficiency improvement measures of the existing dwelling stock.
+The Dwelling Energy Assessment Procedure (DEAP) model was used, this is the national tool for Building Energy Rating (BER) calculation.
+The BER filters from this study were updated and applied in the model developed here.
+Dineen et al.
+performed and sensitivity analysis and found that there was a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift.
+Concluding, BER standard assumptions are poor as they vastly overestimate dwelling energy consumption, the average internal temperature across the Irish dwelling stock may be closer to 18$^{\circ}$C than 21$^{\circ}$C.
 \par In 2013, Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures of the existing detached, oil centrally heated, rural housing stock.
-A base geometry according to age bands and thermal characteristics was created in the study.
+%A base geometry according to age bands and thermal characteristics was created in the study.
 The default values used in some of the BER database variables where highlighted.
@@
 Concluding that the majority of thermal energy efficiency progress in Irish dwellings is being offset.
-The model results show that thermal improvement measures resulted in a reduction of running costs and $CO_2$ emissions by an average of 65\% for houses constructed prior to 1979, and this resulted in an average 12-year payback.
-Houses built in 1979 or after resulted in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures.
-This study highlights the low hanging fruit of pre-1980 oil centrally heated, rural housing stock.
-
-\par In 2015, Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examining new dwellings, by examining the energy savings potential of energy efficiency improvement measures of the existing dwelling stock.
-The Dwelling Energy Assessment Procedure (DEAP) model was used, this is the national tool for BER calculation and checking building regulation compliance.
-The DEAP model used 175 archetype dwellings ( 5 building types, 7 energy bands, and 5 wall construction types ) and 6 different retrofit measures were applied.
-The BER filters from this study were updated and applied in TIM.
-A sensitivity analysis performed found that there was a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift.
-Concluding, BER standard assumptions overestimate dwelling energy consumption and that the average internal temperature across the Irish dwelling stock may be closer to 18$^{\circ}$C than 21$^{\circ}$C.
-
+The model results show that thermal improvement measures resulted in a reduction of running costs and $CO_2$ emissions by an average of 65\% for houses constructed prior to 1979, compared to houses built in 1979 or after resulting in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures.
+This study highlights the low hanging fruit of pre-1979 oil centrally heated, rural housing stock.
 \par The Archetype Dwelling Energy Model - SQL (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}.
-ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and calculates the annual energy consumption for space/ water heating, and lighting using EN 13790 to simulate the potential for improved energy efficiency gains within an existing retrofit program.
+ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and calculates the annual energy consumption for space/ water heating, and lighting using EN 13790:2004 to simulate the potential for improved energy efficiency gains within an existing retrofit program.
 Mac Uidhir et al.
 examined the Better Energy Homes (BEH) retrofit grant scheme.
-The five most common retrofit combinations were simulated for 9 distinct archetypes (3 BER energy rating groups, and 3 building types) and compared these to alternative retrofit combinations which delivered additional efficiency improvements.
+%The five most common retrofit combinations were simulated for 9 distinct archetypes (3 BER energy rating groups, and 3 building types) and compared these to alternative retrofit combinations which delivered additional efficiency improvements.
 The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\%, relative to the set of retrofit combinations actually delivered from the program.
@@
 The SEAI is Ireland's national energy authority and they play a key role in assisting consumers to reduce energy.
-Some of the main models used in the NEMF include:
+The models listed are some of the models used within SEAI's NEMF to model the impact of incentives and opportunities created by specific policy instruments.
+
 \begin{itemize}
@@
 
-The listed models are some of the models used within SEAI's NEMF to model the impact of incentives and opportunities created by specific policy instruments.
-
-The models described in this section include a range of national ESOMs and bottom-up residential models.
-The models have varied objectives such as exploring consumer choices, PV battery system location, evaluating the health effects of retrofitting, and integration of bioenergy and heat.
-The thermal efficiency of dwellings is commonly researched in Irish models, due to the poor thermal efficiency of the dwelling stock and the growing BER database is providing high data granularity input to these models.
-The internal temperature is becoming more frequently analysed in the studies, a recurring theme is the wide range of internal temperatures.
-The bottom-up residential models are typically simulation models looking at a specific policy.
+The models described in this section include a range of national ESOMs and bottom-up residential models with varied objectives.
 The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switched to Microsoft SQL, with much higher computational power.
-There has been few studies on the Irish residential sector using ESOMs, as residential bottom-up models are preferred when focusing on that sector.
-The studies on Irish ESOM typically do not focus on the residential sector, instead they focus on the overall energy system, electricity generation, or transport.
-The use of ESOM to explore DH in Ireland allows for insights which bottom-up simulation models cannot provide.
-The connection between waste heat in electricity and industry can provide better insights on the optimal solution for DH in commercial and residential sectors.
-
-\subsection{Ireland: Residential Sector}
+The growing BER database is providing high data granularity input to Irish models.
+%The internal temperature is becoming more frequently analysed in Irish residential studies, a recurring theme is the wide range of internal temperatures which Dineen et al.
+\cite{Dineen2015ImprovedSource} found this assumption significantly changes the final residential energy consumption.
+
+\subsection{Case study for Ireland}
 \label{Intro:IreRSD}
@@
 Ireland’s distinctive energy sectors are further separated from other EU member states when focusing on the residential sector.
-This is largely down to higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings, and rural settlement patterns in Ireland's residential sector.
+This is largely down to rural settlement patterns, larger dwellings, lower building thermal efficiency, and higher reliance on fossil fuels.
 This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{Eurostat2021DistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{Eurostat2021ShareSurveyilc_lvho50a}.
@@
 
-In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential Greenhouse Gas (GHG) emissions.
-GHG emissions from coal in the residential sector also declined to 6.9\% in 2018 \cite{Howley2005Energy-RelatedIreland.}.
-However, the decline in peat and coal has been offset by a more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005Energy-RelatedIreland.}.
-Despite being limited to some urban areas, gas still accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005Energy-RelatedIreland.}.
+In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential GHG emissions.
+GHG emissions from coal in the residential sector also declined to 6.9\% in 2018 \cite{Howley2020Energy-RelatedIreland.}.
+However, the decline in peat and coal has been offset by a more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2020Energy-RelatedIreland.}.
+Despite being limited to some urban areas, gas still accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2020Energy-RelatedIreland.}.
 The residential sector relied on fossil fuels for over 90\% of thermal energy in 2018, consequently Ireland currently has the lowest renewable heat among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
-Preliminary results from the Environmental Protection Agency (EPA) have indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions, proving some consumers are economically rational by filling oil tanks when prices are low.
-
-Space heating is the largest residential energy service, which accounted for 61.8\% of final energy usage in 2018.
+%Preliminary results from the Environmental Protection Agency (EPA) have indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions, proving some consumers are economically rational by filling oil tanks when prices are low.
+
+Space and water heating are the main energy services, accounting for 80 \% of the energy used in residential dwellings \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and typically have a lifetime of at least 15 years, consequently leading to a slow turnover of technology in the residential sector.
+Apart from space and water heating, all EU member states are required under commission regulation (EU) 431/2014 to supply energy consumption data on space cooling, cooking, lighting \& appliances, and others (mainly outdoor appliances), the annual trend of these energy services is shown in Figure \ref{fig:IrelandServiceDemands}.
+These energy services were also used in TIM, however there is no space cooling data in Ireland and lighting \& appliances were separated while others was combined with appliances.
+
+% Space heating is the largest residential energy service, which accounted for 61.8\% of final energy usage in 2018.
 Other residential energy services and their share of final energy are shown in Fig.\ref{fig:Ireland2018ServiceDemands}.
 As building thermal efficiency and heating technology efficiency improves, both space and water heating energy would be expected to decline.
-
-%, Commission Regulation (EU) No 431/2014 amends the Energy Statistics Regulation(ESR) to require all Member States to supply data on energy consumption in households split by the following end-uses: space heating; space cooling; water heating; cooking; lighting and appliances; other uses.
-%The following data concerning the final energy consumption in households by type of end-use are reported according to Regulation (EC) 1099/2008 on energy statistics and its subsequent amendments.
-%Implementation of Commission Regulation (EU) No 431/2014 of 24 April 2014 amending Regulation (EC) No 1099/2008 of the European Parliament and of the Council on energy statistics
 
@@
     scale=1,
+    legend cell align={left},
     x tick label style={/pgf/number format/1000 sep=,rotate=45},
@@
     xmin=2015, xmax=2019,
-    ymin=0, ymax=80,
+    ymin=0, ymax=25,
     xtick={2015,2016,2017,2018,2019},
-    ytick={0,20,40,60,80},
+    ytick={5,10,15,20,25},
     ymajorgrids=true,
@@
 \addplot[
-    color=red,
-    ]
-    coordinates {
-    (2015,72.5)(2016,72.9)(2017,70.76)(2018,78.59)(2019,73)
-    };
-
-\addplot[
-    color=blue,
+    color=blue,mark=oplus*
     ]
@@
 \addplot[
-    color=green!60!black,
+    color=green!60!black,mark=oplus*
     ]
@@
 \addplot[
-    color=blue!60,orange!70,
+    color=blue!60,orange!70, mark=oplus*
     ]
@@
 \addplot[
-    color=black,
+    color=black,mark=oplus*
     ]
@@
 \end{axis}
+%
+\begin{axis}[
+scale=1,
+    scale=1,
+    legend cell align={left},
+    x tick label style={/pgf/number format/1000 sep=,rotate=45},
+    legend style={at={(0.5,-0.45)},anchor=north,legend columns=2},
+    xmin=2015, xmax=2019,
+    ymin=0, ymax=100,
+    xtick={2015,2016,2017,2018,2019},
+    ytick={20,40,60,80,100},
+    ymajorgrids=true,
+    grid style=dashed,
+    axis y line*=right,
+    axis x line=none,
+    ylabel style = {align=center},
+    ylabel= {Petajoules [PJ]}
+]
+
+\addplot[
+    color=red,mark=oplus*
+    ]
+    coordinates {
+    (2015,72.5)(2016,72.9)(2017,70.76)(2018,78.59)(2019,73)
+    };
+
+     \legend{Space Heating(on Secondary axis)}
+    \end{axis}
 \end{tikzpicture}
-\caption{Ireland Residential Energy Service Demands}
-    \label{fig:IrelandServiceDemands}\cite{Eurostat2021Disaggregatednrg_d_hhq}
+\caption{Ireland Residential Energy Service Demands \cite{Eurostat2021Disaggregatednrg_d_hhq}}
+\label{fig:IrelandServiceDemands}
   \end{figure}
 
-% Other is  mainly covering uses of energy by households outside the dwellings themselves - this is included in Appliances in TIM..
-
-\subsubsection{Current Greenhouse Gas Emissions}
+\subsubsection{Residential sector characteristics}
 \label{Intro:GHG}
 
-Under 2006 Intergovernmental Panel on Climate Change (IPCC) guidelines, Ireland’s residential sector accounted for 14.2\% of GHGs emissions in 2018 \cite{Howley2005Energy-RelatedIreland.}.
-Focusing only on energy-related GHG emissions, the residential sector directly accounts for 24\% of GHG emissions.
-The residential sector is directly responsible for 47\% of heat and 30\% of electricity GHG emissions.
-The residential sector would also be indirectly responsible for a share of the remaining energy-related GHGs in the transport sector, the residential sector is therefore a pivotal sector in decarbonising energy in Ireland.
-
-\subsubsection{Climate Targets}
-
-Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to reduce GHG emissions by 51\% in 2030 compared to 2018.
+Under 2006 Intergovernmental Panel on Climate Change (IPCC) guidelines, Ireland’s residential sector accounted for 14.2\% of GHGs emissions in 2018 \cite{Howley2020Energy-RelatedIreland.}.
+Focusing only on energy-related GHG emissions from heat, electricity and transport, the residential sector directly accounts for 24\% of GHG emissions.
+The residential sector is directly responsible for 47\% of heat and 30\% of electricity GHG emissions and indirectly responsible for a share of the remaining GHGs, especially in the transport sector, as the rural settlement patterns increase transport energy.
+
+Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has adopted legally binding national targets to reduce GHG emissions by 51\% in 2030 compared to 2018.
 The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}.
@@
 Ireland's Climate Action Plan 2019 (CAP2019) complies with the EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-Climate Action Plan 2021 will have to be more ambitious than its predecessor, CAP2019 as it must reflect EU's updated NDC target and take account of the national target also.
+Climate Action Plan 2021 will be more ambitious than its predecessor, CAP2019 as it must reflect EU's updated NDC target and take account of the national target also.
 There are also renewable energy targets to consider for transport (RES-T), heat (RES-H), and electricity (RES-E) under the renewable directive (EU) 2018/2001, and the energy efficiency targets under (EU) 2018/2002.
 
-\subsubsection{Current Energy Policy}
+\subsubsection{Policy context}
 \label{Intro:Policy}
@@
 The densification of Ireland's main cities will provide opportunities to improve the efficiency of heating, electricity, and transport within and between cities.
-\par The Marginal Abatement Cost Curve (MACC) produced in CAP2019, shows electricity \& transport are the most cost-effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}, therefore the most cost-effective residential decarbonisation is to receive higher amounts of low carbon electricity.
-
-\par In Ireland, the decarbonisation of electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005Energy-RelatedIreland.}.
+\par The Marginal Abatement Cost Curve (MACC) produced in CAP2019, shows electricity \& transport are the most cost-effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
+In Ireland, the decarbonisation of electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2020Energy-RelatedIreland.}.
 Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private investment towards low carbon or renewable energy.
 A €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}.
-Ireland's current carbon tax of €33.50/ $tCO_2$ will rise to €100/ $tCO_2$ in 2030.
+Ireland's current carbon tax of €33.50/ $tCO_2$ is set to rise to €100/ $tCO_2$ in 2030.
 The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise renewable electricity investment to achieve legally binding renewable targets under 2009/28/EC.
-REFIT operates by guaranteeing a minimum price for electricity over a 15 year period.
 The Renewable Electricity Support Scheme (RESS) replaced REFIT in 2020.
@@
 
-\par SEAI has a key role in the rollout of retrofitting and heat pumps in existing homes.
-In parallel, new dwellings must achieve a minimum share of 20\% renewable energy.
-The ambient heat used in heat pumps is included as a renewable energy and are accounted for in the RES-H target, heat pumps are therefore the preferred heating technology despite the high capital costs.
-As mentioned in Section \ref{Intro:IreRSD}, new buildings in Ireland produce 70\% less GHG emissions than buildings in 2005, along with CAP2019 plans to retrofit 500,000 existing homes to at least a  ‘B2’ energy rating by 2030, which highlights the energy efficiency first approach in the residential sector.
+\par As the grant provider, SEAI plays a key role in the rollout of retrofitting and heat pumps.
+However, the retrofitting and heat pump grants do require large upfront costs, which may hinder the ambitious targets as this wont be a viable option for many households.
+% are therefore the preferred heating technology despite the high capital costs.
+CAP2019 planned to retrofit 500,000 existing homes to at least a  ‘B2’ energy rating by 2030, which highlights the energy efficiency first approach in the residential sector.
 \par Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019VisionIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
-GNI anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020EnergySEAI}.
-A major barrier to increasing biomethane from an agricultural source in the gas network is the cost, it was the most expensive option in the Marginal Abatement Cost Curve (MACC) produced in CAP2019.
-\par Ireland has one of the lowest shares of District Heating (DH) in Europe at less than 1\%  \cite{Gartland2016AIreland}.
-Advancement in DH control and awareness of the technology will strengthen the case for DH networks to play a part in decarbonisaing the residential sector.
+%GNI anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020EnergySEAI}.
+A major barrier to increasing biomethane from an agricultural source in the gas network is the cost, it was the most expensive option in the MACC produced in CAP2019.
+\par Ireland has one of the lowest shares of District Heating (DH) in Europe at less than 1\%  \cite{Gartland2016AIreland}. %
+%Advancement in DH control and awareness of the technology will strengthen the case for DH networks to play a part in decarbonising the residential sector.
 CAP2019 plans to have approx.
-0.4\% of heating provided by DH in 2030, research using TIMES and EnergyPLAN has shown that 37\% of heating provided by DH is optimal if Ireland is to reduce GHG emissions by 80\% in 2050, compared to 1990 \cite{Thellufsen2019ImplementingIreland}.
-While Irish District Energy Association (IrDEA) suggest 57\% of heat can be sourced from DH, if supporting policy and regulation was in place.
-District Heating will likely play a role, however, Gartland points out a range of organisational, technical, regulatory, and economic barriers to DH growth in Ireland \cite{Gartland2016AIreland}.
-
-The layout of this paper is as follows: 
-\begin{itemize}
-    \item Section \ref{LitR} is the core section of this paper and explains the methodology used to develop the residential sector.
+0.4\% of heating provided by DH in 2030.
+The Irish District Energy Association (IrDEA) suggest 57\% of heat can be sourced from DH, if supporting policy and regulation was in place.
+CAP2019 sets out a need to develop a policy framework for the development of district heating in Ireland
+%District Heating will likely play a role, however, Gartland points out a range of organisational, technical, regulatory, and economic barriers to DH growth in Ireland \cite{Gartland2016AIreland}.
+
+\subsection{Motivation \& Paper Overview}
+\label{Intro:PaperOver}
+
+The distinctive Irish residential sector provides an interesting case study and new Irish climate targets now means previous Irish models which relied on superseded climate targets are out-dated but not redundant, new up to date models are needed to explore the decarbonisation pathways and technical limits.
+No such ESOM with a focus on Irish residential sector exists, the strength of an ESOM is to account for sectoral interactions which bottom-up simulation models cannot provide.
+Structural improvements and archetypes in the model aligns with existing retrofitting and electrification strategies.
+Other model improvements include filtering input data and calculating new energy service demands, while also considering internal temperatures.
+TIM will be a critical tool in exploring
+\textit{How changes to future technology developments affect the optimal residential heating decarbonisation pathway in Ireland? }
+
+The layout of the rest of this paper is as follows; section \ref{LitR} is the core section of this paper and explains the methodology used to develop the residential sector.
 The treatment of input data in the model is described, the calculations are explained, and assumptions and constraints are highlighted.
-    \item Section \ref{Results} outlines the scenarios used and displays results from the residential sector.
-    \item Section \ref{Discussion} discusses the results, strengths, and weakness of the newly developed model and how it differs from other models.
-    \item Section \ref{Conclusion} draws some conclusions and discusses possible future work.
-\end{itemize}
+Section \ref{Results} outlines the scenarios used and displays results from the residential sector.
+Section \ref{Discussion} discusses the results, strengths, and weakness of the newly developed model and how it differs from other models.
+Section \ref{Conclusion} draws some conclusions and discusses possible future work.
+
+%brings together sections 1.1 and 1.2 – one paragraph justifying this study (improving residential modelling; applying to important case study), and include the objectives of the study (introduce the model and any research questions) then bring the paragraph on the structure from the start.
+This is a bridging section between the introduction and methodology.
+- Hannah
+
 \section{Methodology}
@@
 
-\subsection{TIMES Energy System Optimisation Model}
+\subsection{The TIMES-Ireland Model (TIM)}
 \label{Overview}
@@
 The residential sector was developed in parallel with other sectors in TIM, and concurrently with the new Irish LEAP model, which allows for more accessible soft-linking to provide robust analytical insights by combining simulation and optimisation, and to exchange expertise between modelling teams.
-Fig.\ref{fig:RES} illustrates the flow of energy from fuel to demand in the TIM residential sector.
+Figure \ref{fig:RES} illustrates the flow of energy from fuel to demand in the TIM residential sector.
 
@@
  \includegraphics[width=\textwidth]{Figures/TIM_RES3.png}
- \caption{TIM Residential Sector - Reference Energy System}
+ \caption{Structure of TIM residential sector }
  \label{fig:RES}
@@
 The BER database provides the base year bottom-up energy consumption data, this is scaled up to represent the entire stock and cross-checked with the top-down energy consumption from the 2018 SEAI Energy Balance.
-To compare the two databases, fuels were were combined and condensed down to 13 fuels in TIM.
+To compare from the two databases, fuels were were combined and condensed down to 13 fuels in TIM.
 Table \ref{Table:Fuels} shows an overview of the fuels combined from the different data sources.
 
-\par In TIM, fuels such as coal, represent a combination of fuels from the SEAI Energy Balance and BER database.
-Gas oil, which is accounted for in the SEAI Energy Balance, has similar properties to kerosene and there was no evidence to support keeping separate fuels would provide further insight into residential decarbonisation.
-However, dissagreatting ambient heat could potentially provide insights on residential decarbonisation.
+\par In TIM, fuels such as ``coal", represent a combination of fuels from the SEAI Energy Balance and BER database.
+Gasoil which has similar properties to kerosene, is accounted for in the SEAI Energy Balance,  but there is no evidence to support keeping gasoil and kerosene separate would provide further insight into residential decarbonisation, therefore these fuels are combined in TIM.
+However, disaggregating ambient heat could potentially provide insights on residential decarbonisation, so the model has separated space ambient heat and water ambient heat.
 The solid multi-fuel from the BER database represents an open fire or stove which is heated by either coal, peat, or wood.
-Through examining the SEAI Energy Balance and BER database, the fuel share  for coal, peat, and wood in solid multi-fuel is 19.4 \%, 74.4 \% and 6.2 \% respectively.
+Through examining the SEAI Energy Balance and BER database, the fuel share for coal, peat, and wood in solid multi-fuel was set to 19.4 \%, 74.4 \% and 6.2 \% respectively.
 \par There are two types of energy services represented in Fig.
@@
 Essentially Non-archetype energy services have a lack of data and could not be accurately assigned to an archetype.
-The six non-archetype energy services are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service, this bottom-up data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{LitR:Lighting,Appliances}.
-The four archetype energy services are dependent on  building archetypes and data is sourced from the BER database, which is discussed in Section \ref{LitR:StockData}.
-The internal temperature adjustment for space heating is detailed in Section \ref{LitR:StockData}.
+The six non-archetype energy services are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service.
+Non-Archetype data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{LitR:Lighting,Appliances}.
+The four archetype energy services are dependent on  building type and data is sourced from the BER database, which is discussed in Section \ref{LitR:StockData}.
+The internal temperature adjustment for space heating is also detailed in Section \ref{LitR:StockData}.
 The demand for archetype and non-archetype energy services is based on the growth of residential dwellings, which is detailed in section \ref{LitR:Demands}.
 
+%PICKUPHERE-Jason13Aug
 \subsubsection{Dwelling Stock Data}

### 2021-08-17 manuscript.tex
Got rid of errors - references OK.

@@
-%% 
-%% Copyright 2007-2020 Elsevier Ltd
-%% 
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%% 
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%% 
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%% 
-%% Template article for Elsevier's document class `elsarticle'
-%% with harvard style bibliographic references
-%\usepackage{cite}
-%\usepackage{harvard}
-%\usepackage{biblatex}
-%\usepackage[backend=biber]{biblatex}
-% \addbibresource{references.bib} %Imports bibliography file
-
-%%%%%%%%%%%%%%%%%%%%%%%
-%% Elsevier bibliography styles
-%%%%%%%%%%%%%%%%%%%%%%%
-%% To change the style, put a % in front of the second line of the current style and
-%% remove the % from the second line of the style you would like to use.
-%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Numbered
-%\bibliographystyle{model1-num-names}
-
-%% Numbered without titles
-%\bibliographystyle{model1a-num-names}
-
-%% Harvard
-%\bibliographystyle{model2-names.bst}\biboptions{authoryear}
-
-%% Vancouver numbered
-%\usepackage{numcompress}\bibliographystyle{model3-num-names}
-
-%% Vancouver name/year
-%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}
-
-%% APA style
-%\bibliographystyle{model5-names}\biboptions{authoryear}
-
-%% AMA style
-%\usepackage{numcompress}\bibliographystyle{model6-num-names}
-
-%% `Elsevier LaTeX' style
-%\bibliographystyle{elsarticle-num}
-%%%%%%%%%%%%%%%%%%%%%%%
-
-\documentclass[preprint]{elsarticle}
-
-\usepackage{lineno,hyperref}
-\modulolinenumbers[5]
-
-%% `Elsevier LaTeX' style
-%\bibliographystyle{elsarticle-num}
-%\documentclass[10pt,final,journal,compsoc,a4paper]{IEEEtran}
-%% Use the option review to obtain double line spacing
-%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%%\documentclass[final,1p,times,authoryear]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
-%%\documentclass[final,3p,times,authoryear]{elsarticle}
-%%\documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,5p,times,authoryear]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}
-
-%% For including figures, graphicx.sty has been loaded in
-%% elsarticle.cls. If you prefer to use the old commands
-%% please give \usepackage{epsfig}
+\documentclass[final,3p,sort&compress,12pt]{elsarticle}
 
@@
 \usepackage{amssymb}
+\usepackage{setspace}
 \usepackage{pifont}
@@
 
+\bibliographystyle{elsarticle-num-names}
+
 \DeclareSIUnit{\EUR}{\text{\euro}}
  \pgfplotsset{compat=1.17}
-%\usepackage[table]{xcolor}
-%\usepackage[backend=bibtex]{biblatex}
-
-%% The amsthm package provides extended theorem environments
-%% \usepackage{amsthm}
-%\bibliographystyle{elsarticle-num}
-%\usepackage{cleveref}
-%\usepackage{hyperref}
-%\usepackage[style=numeric, sorting=none]{biblatex}  %backend=biber is 'better'
-%
-\addbibresource{TIM_V11.bib}
-%% The lineno packages adds line numbers. Start line numbering with
-%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-%% for the whole article with \linenumbers.
 
 \journal{Energy and Buildings}
-%\bibliographystyle{elsarticle-num}
 
@@
 %% Title, authors and addresses
-
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for theassociated footnote;
-%% use the fnref command within \author or \affiliation for footnotes;
-%% use the fntext command for theassociated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for theassociated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \affiliation{organization={},
-%%            addressline={}, 
-%%            city={},
-%%            postcode={}, 
-%%            state={},
-%%            country={}}
-%% \fntext[label3]{}
 
@@
 \cortext[j.mcguire@ucc.ie]{j.mcguire@ucc.ie}
-%footnote{E-mail: \url{j.mcguire@ucc.ie}}}
-
-%\footnote{Contact: j.mcguire@ucc.ie}}
-%\thanks{Email: \url{j.mcguire@ucc.ie}}}
-\affiliation[inst1]{organization={Energy Policy and Modelling Group, MaREI Centre},%Department and Organization
+
+\affiliation[inst1]{organization={Energy Policy and Modelling Group, MaREI Centre},
             addressline={Environmental Research Institute}, 
@@
 
-\affiliation[inst2]{organization={School of Engineering},%Department and Organization
+\affiliation[inst2]{organization={School of Engineering},
             addressline={University College Cork}, 
@@
 
+ \onehalfspacing
 \begin{abstract}
@@
 
+\vspace{0.25cm}
+\noindent
+
 \end{abstract}
@@
 %% \linenumbers
-
+\newpage
 %% main text
+ \onehalfspacing
 \section{Introduction}
@@
 
+ \onehalfspacing
 \subsection{Energy Systems Optimisation Modelling and the Residential Sector}
@@
 
-Space and water heating are the main energy services, accounting for 80 \% of the energy used in residential dwellings \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and typically have a lifetime of at least 15 years, consequently leading to a slow turnover of technology in the residential sector.
+Space and water heating are the main energy services, accounting for 80\% of the energy used in residential dwellings \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and typically have a lifetime of at least 15 years, consequently leading to a slow turnover of technology in the residential sector.
 Apart from space and water heating, all EU member states are required under commission regulation (EU) 431/2014 to supply energy consumption data on space cooling, cooking, lighting \& appliances, and others (mainly outdoor appliances), the annual trend of these energy services is shown in Figure \ref{fig:IrelandServiceDemands}.
@@
 Where t is the type of retrofit.
-${TIM_{r,a}}$ is the number of dwellings by rating and archetype and ${TIM_{a}}$ is the total number of dwellings of that archetype.The expected savings from each type of retrofit and for each archetype are outlined in Table \ref{Table:RetrofitSavings}.
+${TIM_{r,a}}$ is the number of dwellings by rating and archetype and ${TIM_{a}}$ is the total number of dwellings of that archetype.
+The expected savings from each type of retrofit and for each archetype are outlined in Table \ref{Table:RetrofitSavings}.
 
@@
 CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC.
-Alessandro and Maurizio provided technical assistants on the development of the residential sector in TIMES. 
+Alessandro and Maurizio provided technical assistance on the development of the residential sector in TIMES. 
 % https://www.marei.ie/project/capacity/
@@
 
-\newpage
-
-\printbibliography
-%\bibliography{TIM_V1.1}
-%\bibliographystyle{elsarticle-num}
-%\bibliography{TIM_V11}
-%\bibliographystyle{elsarticle-harv}
-
-%\bibliography{references.bib}
-
-%% else use the following coding to input the bibitems directly in the
-%% TeX file.
-
-%%\begin{thebibliography}{00}
-
-% %% \bibitem[Author(year)]{label}
-% %% Text of bibliographic item
-
-% \bibitem[ ()]{}
-
-% \end{thebibliography}
+\bibliography{TIMref.bib}
+
 \end{document}

### 2021-08-17 manuscript.tex
On first draft - 2.2.2 Costs

@@
 Ireland has committed to one of the most ambitious decarbonisation targets in the world to 2030.
-The residential sector, responsible for 14.2 \% of total greenhouse gas (GHG) emissions \cite{Howley2020Energy-RelatedIreland.} , is among the most carbon intensive per dwelling in the EU, emitting 60 \% more $CO_2$ than the national EU average \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, because of a high share of low thermal efficient detached dwellings coupled with a high dependency on carbon-intensive heating fuels.
+The residential sector, responsible for 14.2\% of total greenhouse gas (GHG) emissions \cite{Howley2020Energy-RelatedIreland.} , is among the most carbon intensive per dwelling in the EU, emitting 60\% more $CO_2$ than the national EU average \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, because of a high share of low thermal efficient detached dwellings coupled with a high dependency on carbon-intensive heating fuels.
 These factors have served as barriers to the decarbonisation of the residential sector in Ireland to date.
@@
 Economically optimal or least cost residential decarbonisation pathways are dynamic, with technological advancements in heating appliances being assisted by exponentially increasing computational capacity over time, combined with new low-carbon fuel alternatives and new heating technologies becoming available.
-ESOMs have a rich representation of existing and future technological variables such as efficiency, cost, and fuel, while also accounting for macroeconomic effects.
+ESOMs have a rich representation of existing and future technological variables such as efficiency, cost, and fuel, while also accounting for macroeconomic variables such as economic growth and the social discount rate.
 ESOMs supply enough energy to at least satisfy energy service demands so that the lights stay on, while also considering sectoral interconnections to explore future decarbonisation pathways.
@@
 These models can be used to explore the residential sector in higher granularity.
-Different levels of disaggregation are applied to the residential sector in models and research by Natarajana et al.
-\cite{Natarajana2011ModellingApproach} highlighted a common approach used by defining an average performance for a number of dwelling categories, and this approach was adopted here.
+Different levels of disaggregation are applied to the residential sector in models and research by \citet{Natarajana2011ModellingApproach}  highlighted a common approach used by defining an average performance for a number of dwelling categories, and this approach was adopted here.
 To account for the heterogeneous mix of dwellings, each dwelling is categorised based on building type to better reflect the housing stock and provide building specfic insights.
@@
 TIMES assumes perfect foresight to enable a least-cost solution to be calculated, but \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
-In light of this, Li et al.
-\cite{Li2018IncorporatingModel} investigated the heating technology preference in the UK using UK TIMES Model (UKTM) by incorporating a Discrete Choice Model (DCM) to better reflect how homeowners choose heating technologies and thus correcting for error in the perfect foresight assumption.
+In light of this,  \citet{Li2018IncorporatingModel} investigated the heating technology preference in the UK using UK TIMES Model (UKTM) by incorporating a Discrete Choice Model (DCM) to better reflect how homeowners choose heating technologies and thus correcting for error in the perfect foresight assumption.
 
@@
 Retrofitting and electrification are a fundamental part of residential decarbonisation in Ireland, and therefore play a central role in the development of the model described here.
-Furthermore, the new model includes technologies, which were not previously considered, one such technology is District Heating (DH), some initial studies have found it may be optimal to have 37 \% of heat demand to be provided by DH in Ireland \cite{Thellufsen2019ImplementingIreland}, the full list of space and water heating technologies considered in the model can be found in \ref{sec:appendixA4}.
-The strength of an ESOM can be used to examine the sectoral interactions, one study by Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} examined the interaction between heat pumps and wind generation investments in Denmark, the results find it is optimal to install air-to-water heat pumps in non-district heating areas.
+Furthermore, the new model includes technologies, which were not previously considered, one such technology is District Heating (DH), some initial studies have found it may be optimal to have 37\% of heat demand to be provided by DH in Ireland \cite{Thellufsen2019ImplementingIreland}, the full list of space and water heating technologies considered in the model can be found in \ref{sec:appendixA4}.
+The strength of an ESOM can be used to examine the sectoral interactions, one study by Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} examined the interaction between heat pumps and wind generation investments in Denmark, the results find it is optimal to install air-to-water heat pumps in non-DH areas.
 
@@
 
-ESOMs have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem,Connolly2009DevelopingUniversity.}, however no Irish studies use ESOM models to solely explore the residential sector, as many models look across all sectors.
+ESOMs have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem,Connolly2009DevelopingUniversity.}, however no Irish studies use ESOMs to solely explore the residential sector, as many models look across multiple sectors.
 However, two examples of Irish ESOMs focusing the residential sector in higher granularity are outlined below.
 
-In 2014, Rogan et al.
-\cite{Rogan2014LEAPsSystem} linked OSeMOYOS to the Long Range Energy Alternatives Planning system (LEAP), which is a scenario-based simulation modeling tool used to analysis energy policy in all sectors of the economy.
+In 2014, \citet{Rogan2014LEAPsSystem} linked OSeMOYOS to the Long Range Energy Alternatives Planning system (LEAP), which is a scenario-based simulation modeling tool used to analysis energy policy in all sectors of the economy.
 The models were linked to explore specific energy efficiency policies in the residential and other sectors.
 Three scenarios were explored, namely, the reference scenario, energy efficiency scenario, and energy efficiency+ scenario.
-The reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020, and the results show that residential total energy consumption rose 6.7 \% in 2020, compared to 2008.
-In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulting in a 9.9 \% reduction of energy consumption in 2020, compared to the reference scenario.
+The reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020, and the results show that residential total energy consumption rose 6.7\% in 2020, compared to 2008.
+In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and low energy lighting, resulting in a 9.9\% reduction of energy consumption in 2020, compared to the reference scenario.
 The energy efficiency+ scenario also allows for 800,000 residential dwellings to be retrofitting, but accounts for deeper retrofits.
-This results in an energy consumption reduction of 16.5 \% in 2020, compared to the reference scenario.
-Rogan et al.
-discussed the barriers, obstacles to investment, split incentives, and information gaps which restrict energy efficiency improvements.
+This results in an energy consumption reduction of 16.5\% in 2020, compared to the reference scenario.
+\citet{Rogan2014LEAPsSystem} discussed the barriers, obstacles to investment, split incentives, and information gaps which restrict energy efficiency improvements.
 The insights highlighted the importance for the policy to not only look at the number of retrofits, but also the depth of retrofit, as shown by the different results in the scenarios.
@@
 
-In 2018, Thellufsen et al.
-\cite{Thellufsen2019ImplementingIreland} investigated the potential for the implementation of district heating in both the commercial and residential sectors of Ireland.
-The paper analyses the $CO_2-80$ scenario ( 80 \% $CO_2$ reduction in 2050, compared to 1990 ) with and without district heating using the previous Irish TIMES model and EnergyPLAN.
-The previous version of Irish TIMES did not account for district heating, so EnergyPLAN is used to investigate the operation and  feasibility of district heating in Ireland.
+In 2018, \citet{Thellufsen2019ImplementingIreland} investigated the potential for the implementation of DH in both the commercial and residential sectors of Ireland.
+The paper analyses the $CO_2-80$ scenario ( 80\% $CO_2$ reduction in 2050, compared to 1990 ) with and without DH using the previous Irish TIMES model and EnergyPLAN.
+The previous version of Irish TIMES did not account for DH, so EnergyPLAN is used to investigate the operation and  feasibility of DH in Ireland.
 A sensitivity analysis was performed on investment costs, heat grid losses, and discount rates.
-In all cases, the inclusion of district heating was found to be favourable.
-The results find the conversion of 37 \% of the Irish heat demand to district heating to be optimal.
-Both the TIMES model and climate target are superseded and the study did not include the future heat savings expected from retrofitting, nevertheless this study has provided insights on the potential of district heating.
+In all cases, the inclusion of DH was found to be favourable.
+The results find the conversion of 37\% of the Irish heat demand to DH to be optimal.
+Both the TIMES model and climate target used in this study are superseded and the study did not include the future heat savings expected from retrofitting, nevertheless this study has provided insights on the potential of DH.
 
@@
 \par
-In 2011, Dineen and Ó Gallachóir \cite{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP).
+In 2011,  \citet{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP).
 The model focused on new buildings and examined the energy efficiency of space and water heating by examining technology and building fabric thermal efficiency.
 %The paper acknowledges that retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperature effect on energy efficiency improvements.
-Later in 2015, Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examining new dwellings, by examining the energy savings potential of energy efficiency improvement measures of the existing dwelling stock.
+Later in 2015, \citet{Dineen2015ImprovedSource} built on previous work in examining new dwellings, by examining the energy savings potential of energy efficiency improvement measures of the existing dwelling stock.
 The Dwelling Energy Assessment Procedure (DEAP) model was used, this is the national tool for Building Energy Rating (BER) calculation.
 The BER filters from this study were updated and applied in the model developed here.
-Dineen et al.
-performed and sensitivity analysis and found that there was a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift.
+\citet{Dineen2015ImprovedSource} performed and sensitivity analysis and found that there was a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift.
 Concluding, BER standard assumptions are poor as they vastly overestimate dwelling energy consumption, the average internal temperature across the Irish dwelling stock may be closer to 18$^{\circ}$C than 21$^{\circ}$C.
-\par In 2013, Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures of the existing detached, oil centrally heated, rural housing stock.
+\par In 2013, \citet{Ahern2013StateStock} investigated thermal retrofit measures of the existing detached, oil centrally heated, rural housing stock.
 %A base geometry according to age bands and thermal characteristics was created in the study.
@@
 For example, there are assumed U-values when the roof insulation thickness is unknown, this is based on age, but there are no U-values for uninsulated roofs despite a 2001 survey finding 18\% of detached housing in Ireland had no roof insulation.
-Ahern et al.
-found dwellings built post-2000 consume the greatest amount of energy (kWh/annum), due to their large floor area.
+\citet{Ahern2013StateStock} found dwellings built post-2000 consume the greatest amount of energy (kWh/annum), due to their large floor area.
 Concluding that the majority of thermal energy efficiency progress in Irish dwellings is being offset.
@@
 This study highlights the low hanging fruit of pre-1979 oil centrally heated, rural housing stock.
-\par The Archetype Dwelling Energy Model - SQL (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}.
-ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and calculates the annual energy consumption for space/ water heating, and lighting using EN 13790:2004 to simulate the potential for improved energy efficiency gains within an existing retrofit program.
-Mac Uidhir et al.
-examined the Better Energy Homes (BEH) retrofit grant scheme.
+\par The Archetype Dwelling Energy Model - SQL (ArDEM-SQL) was developed by \citet{Uidhir2020ImprovingDecisions} ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and calculates the annual energy consumption for space/ water heating, and lighting using EN 13790:2004 to simulate the potential for improved energy efficiency gains within an existing retrofit program.
+Mac \citet{Uidhir2020ImprovingDecisions} examined the Better Energy Homes (BEH) retrofit grant scheme.
 %The five most common retrofit combinations were simulated for 9 distinct archetypes (3 BER energy rating groups, and 3 building types) and compared these to alternative retrofit combinations which delivered additional efficiency improvements.
@@
 \par Finally, the Sustainable Energy Authority of Ireland's (SEAI’s) National Energy Modelling Framework (NEMF) is a suite of modelling tools used for modelling the Irish energy system.
-The SEAI is Ireland's national energy authority and they play a key role in assisting consumers to reduce energy.
+The SEAI is Ireland's national energy authority and they play a key role in bridging the gap between grant/policy and consumer, by assisting consumers to reduce energy.
 The models listed are some of the models used within SEAI's NEMF to model the impact of incentives and opportunities created by specific policy instruments.
 
-\begin{itemize}
+\begin{enumerate}[1.]
 \item {Core Structural Model of the Irish Economy (COSMO)  is a macro-economic model used by the Economic and Social Research Institute (ESRI) to create a baseline forecast of energy consumption by sector} \cite{Bergin2017COSMO:Ireland}
@@
 \begin{enumerate}[i.]
-\item{Consumer Choice Model determines the attractiveness of energy efficiency finance options}\cite{SustainableEnergyAuthorityofIreland2017BehaviouralSector}
+\item{Complementary to the Energy-efficiency policy analysis tool is the Consumer Choice Model which determines the attractiveness of energy efficiency finance options}\cite{SustainableEnergyAuthorityofIreland2017BehaviouralSector}
 \end{enumerate}
 \item{BioHEAT is a techno-economic simulation model of bioenergy and heat} \cite{Durusut2018BioHEAT:Sectors}
-\end{itemize}
+\end{enumerate}
 
@@
 Apart from space and water heating, all EU member states are required under commission regulation (EU) 431/2014 to supply energy consumption data on space cooling, cooking, lighting \& appliances, and others (mainly outdoor appliances), the annual trend of these energy services is shown in Figure \ref{fig:IrelandServiceDemands}.
-These energy services were also used in TIM, however there is no space cooling data in Ireland and lighting \& appliances were separated while others was combined with appliances.
+These energy services were also used in TIM, however there is no space cooling data available in Ireland and lighting \& appliances were separated while others was combined with appliances.
 
@@
     x tick label style={/pgf/number format/1000 sep=,rotate=45},
-    legend style={at={(0.5,-0.25)},anchor=north,legend columns=2},
+    legend style={at={(0.5,-0.35)},anchor=north,legend columns=2},
     xlabel={Year},
@@
     x tick label style={/pgf/number format/1000 sep=,rotate=45},
-    legend style={at={(0.5,-0.45)},anchor=north,legend columns=2},
+    legend style={at={(0.5,-0.55)},anchor=north,legend columns=2},
     xmin=2015, xmax=2019,
@@
 Focusing only on energy-related GHG emissions from heat, electricity and transport, the residential sector directly accounts for 24\% of GHG emissions.
-The residential sector is directly responsible for 47\% of heat and 30\% of electricity GHG emissions and indirectly responsible for a share of the remaining GHGs, especially in the transport sector, as the rural settlement patterns increase transport energy.
-
-Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has adopted legally binding national targets to reduce GHG emissions by 51\% in 2030 compared to 2018.
+The residential sector is directly responsible for 47\% of heat and 30\% of electricity and indirectly responsible for a share of the remaining GHGs, especially in the transport sector, as the rural settlement patterns increase transport demands.
+
+Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has adopted legally binding national targets to reduce GHG emissions by 51\% in 2030, compared to 2018.
 The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}.
-These targets account for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to make sectoral five-year carbon budgets a legal requirement to aid progress of the forementioned long-term targets.
+These targets account for all emissions - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to make sectoral five-year carbon budgets a legal requirement to aid progress of the forementioned long-term targets.
 \par The EU's first Nationally Determined Contribution (NDC), which had a 40\% GHG reduction target (first NDC), has been succeeded by a new target of at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels.
@@
 \par As the grant provider, SEAI plays a key role in the rollout of retrofitting and heat pumps.
-However, the retrofitting and heat pump grants do require large upfront costs, which may hinder the ambitious targets as this wont be a viable option for many households.
+The retrofitting and heat pump grants do still require large upfront costs from the consumer, which may hinder the ambitious targets as this wont be a viable option for many households, green loans are currently available to bridge this gap.
 % are therefore the preferred heating technology despite the high capital costs.
@@
 A major barrier to increasing biomethane from an agricultural source in the gas network is the cost, it was the most expensive option in the MACC produced in CAP2019.
-\par Ireland has one of the lowest shares of District Heating (DH) in Europe at less than 1\%  \cite{Gartland2016AIreland}. %
+\par Ireland has one of the lowest shares of DH in Europe at less than 1\%  \cite{Gartland2016AIreland}. %
 %Advancement in DH control and awareness of the technology will strengthen the case for DH networks to play a part in decarbonising the residential sector.
@@
 The Irish District Energy Association (IrDEA) suggest 57\% of heat can be sourced from DH, if supporting policy and regulation was in place.
-CAP2019 sets out a need to develop a policy framework for the development of district heating in Ireland
-%District Heating will likely play a role, however, Gartland points out a range of organisational, technical, regulatory, and economic barriers to DH growth in Ireland \cite{Gartland2016AIreland}.
+CAP2019 sets out a need to develop a policy framework for the development of DH in Ireland
+%DH will likely play a role, however, Gartland points out a range of organisational, technical, regulatory, and economic barriers to DH growth in Ireland \cite{Gartland2016AIreland}.
 
@@
 TIM will be a critical tool in exploring
-\textit{How changes to future technology developments affect the optimal residential heating decarbonisation pathway in Ireland? }
+\textit{how changes to future technology developments affect the optimal residential heating decarbonisation pathway in Ireland? }
 
@@
 The solid multi-fuel from the BER database represents an open fire or stove which is heated by either coal, peat, or wood.
-Through examining the SEAI Energy Balance and BER database, the fuel share for coal, peat, and wood in solid multi-fuel was set to 19.4 \%, 74.4 \% and 6.2 \% respectively.
+Through examining the SEAI Energy Balance and BER database, the fuel share for coal, peat, and wood in solid multi-fuel was set to 19.4\%, 74.4\% and 6.2\% respectively.
 \par There are two types of energy services represented in Fig.
@@
 
-%PICKUPHERE-Jason13Aug
 \subsubsection{Dwelling Stock Data}
@@
 The dwelling stock data is obtained from the SEAI BER Public database, which is firstly filtered, then scaled up using Central Statistics Office (CSO) data.
-The BER database is Ireland's national Energy Performance Certificate (EPC) register, which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD).
+The SEAI BER public database is Ireland's national Energy Performance Certificate (EPC) register, which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD).
 The BERs are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
 \par Over half of Irish dwellings are captured in the BER database and every BER displays over 200 energy related variables.
-The DEAP model which is used to assign BER labels,  is outlined by Dineen et al.\cite{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM}, In terms of space heating, DEAP assumes that each dwelling is single zoned, the heating season is from October to May inclusive, and the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season.
+The DEAP model which is used to assign BER labels,  is outlined by Dineen et al.\cite{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM}.
+In terms of space heating, DEAP assumes that each dwelling is single zoned, the heating season is from October to May inclusive, and the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season.
 One assumption, the internal temperature assumption, results in a large error, where space and water heating are overestimated in DEAP.
-Space heating consumption is reduced by 26 \% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance.
+Space heating consumption is reduced by 26\% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance.
 DEAP assumes that all living areas ( living room, sitting room, and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C.
@@
 
-The ArDEM-SQL, which was described in Section \ref{Intro:ESOMIreland} is a simulation model built on space heating calculation EN 13790.
-ArDEM-SQL alters the primary energy consumption of the BER database by changing the internal temperatures assumed by DEAP (ref to table).
+The ArDEM-SQL, which was described in section \ref{Intro:ESOMIreland} is a simulation model built on space heating calculation EN 13790.
+ArDEM-SQL alters the primary energy consumption of the BER database by changing the internal temperatures assumed by DEAP in table \ref{Table:InternalTemps}.
 
@@
 The Census 2016 data \cite{CentralStatisticsOffice2016CensusIreland} from the CSO is also used in calculating the dwelling stock.
-Dataset E1005 provides a split of seven archetypes per county, which are aggregated to form three archetypes.
+Dataset E1005 provides a split of seven building archetypes per county, which are aggregated to form three building archetypes.
 The total number of dwellings per archetype per county in 2016 is used to scale up the BER database.
 The gap in total dwellings between the census year in 2016 and the model base year in 2018 was initially ignored due to lack of data.
-However, later it was found that data on new dwelling electrical connections per archetype and per county are available from CSO dataset NDQ06, this would have filled the gap, the first iteration of the dwelling stock in TIM has a 2 \% error.
+However, later it was found that data on new dwelling electrical connections per archetype and per county are available from CSO dataset NDQ06, this would have filled the gap, the first iteration of the dwelling stock in TIM has a 2\% error.
 \par Once the BER database and CSO database tables have been filled to include dwellings per archetype per county, the next step was to scale up the BER database.
@@
 \begin{split}
-BER_{a,c} * R_{a,c} = TIM_{a,c}
+TIM_{a,c} = BER_{a,c} * R_{a,c} 
 \end{split}
@@
 This section only covers the costs applied in the residential sector, which includes space and water heating technologies, retrofitting, cooking, lighting, pumps, fans, space cooling, and other electrical appliances.
-The fuel costs are not applied in the residential sector, there are applied in the supply sector, and the data is sourced from World Energy Outlook reports from 2019 and 2020. 
+The fuel costs are not applied in the residential sector, there are applied in the supply sector, and the data is sourced from World Energy Outlook reports from 2019 \cite{/content/publication/caf32f3b-en} and 2020 \cite{/content/publication/557a761b-en}.
 The majority of the space and water heating cost data is obtained from the
@@
 The lifetime and emissions/pollutants are also included, while the cost data is further disaggregated between capital expenditure ( equipment, installation \& additional capital costs ) and Operational \& Maintenance (O\&M) costs.
-\par TIM has two types of retrofitting options per archetype,  a shallow retrofit is when the primary energy ($kWh/m^2/yr$) is reduced by $ \leq 35 \%$, and deep retrofit is where the primary energy is reduced by $>35 \%$, a reduction of $ \ >75 \%$ is considered infeasible.
+\par TIM has two types of retrofitting options per archetype,  a shallow retrofit is when the primary energy ($kWh/m^2/yr$) is reduced by $ \leq 35\%$, and deep retrofit is where the primary energy is reduced by $>35\%$, while a reduction of $ \ >75\%$ is considered infeasible.
 The difference between the pre-retrofit and post-retrofit primary energy determines the type of retrofit, as shown in \ref{sec:appendixA3}.
@@
 ${TIM_{r,a}}$ is the number of dwellings by rating and archetype and ${TIM_{a}}$ is the total number of dwellings of that archetype, however A-rated dwellings are excluded from the total, as it is assumed they cannot be retrofitted further.
-So $ {\text{\euro}}{}_{a,t}$ is the average cost per type of retrofit per archetype, this value is used in TIM.
+So $ {\text{\euro}}{}_{a,t}$ is the average weighted cost per type of retrofit per archetype, which is used in TIM.
 
@@
 Lighting, appliances, and pumps \& fans are all fueled by electricity, whereas cooking can be fueled by electricity, gas or LPG.
-As shown in Figure \ref{fig:Ireland2018ServiceDemands}, these energy services accounted for 20\% of energy consumption in 2018, however, this share of consumption is expected to increase.
+As shown in Figure \ref{fig:IrelandServiceDemands}, these energy services accounted for 20\% of energy consumption in 2018, however, this share of consumption is expected to increase.
 Refrigeration, cloth washing, cloth drying, and dish washing are under the appliances category and data is sourced from Topten \cite{Michel2016EnergyREPORT} and SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
@@
  \item One Space and Water Heating System per household ( stoves excluded ) 
- \item Maximum 60\% of existing buildings to be retrofitted by 2030, and 95 \% by 2070. Maximum \% interpolated between 2030 and 2070. 
+ \item Maximum 60\% of existing buildings to be retrofitted by 2030, and 95\% by 2070. Maximum interpolated between 2030 and 2070. 
  \item A prerequisite requirement for electrical heat pump installation is applied to each archetype.
@@
     \multicolumn{1}{l}{Requirement} & \multicolumn{1}{l}{Apartment} & \multicolumn{1}{l}{Attached } & \multicolumn{1}{l}{Detached} \\ \hline
-    No Retrofit  & 24.5 \%   & 14.3 \%   & 12 \% \\
-    Shallow Retrofit  & 28.5 \%   & 34.7 \%  & 36.8 \% \\
-    Deep Retrofit  & 47 \%   & 51 \%  & 51.2 \% \\
+    No Retrofit  & 24.5\%   & 14.3\%   & 12\% \\
+    Shallow Retrofit  & 28.5\%   & 34.7\%  & 36.8\% \\
+    Deep Retrofit  & 47\%   & 51\%  & 51.2\% \\
     \end{tabular}%
@@
  \item For space and water heating, each archetype has an upper gas growth constraint of 40\% higher than the base year value.
-\item Maximum fuel share in cooking for gas and LPG is 40\% and 10\% respectively. Base year fuel share for Gas is 32\% and LPG is 1.5 \%.
+\item Maximum fuel share in cooking for gas and LPG is 40\% and 10\% respectively. Base year fuel share for Gas is 32\% and LPG is 1.5\%.
 \item Maximum 10\% solar input to dual fuel heating technology
@@
     \item \textbf{A40E57} Agriculture GHGs reduces by 40\% and energy system GHGs by 57\% \& net-zero in 2050 
-    \item \textbf{A51E51} Equal GHG reduction share in of 51 \% in Agriculture and energy Systems \& net-zero in 2050 
+    \item \textbf{A51E51} Equal GHG reduction share in of 51\% in Agriculture and energy Systems \& net-zero in 2050 
 \end{enumerate}
 
-While the first two scenarios do not comply with the Climate Action and Low Carbon Development (Amendment) Act 2021 target of an overall 51 \% GHG reduction in 2030 compared to 2018, they do provide insights on the incremental steps required to achieve these targets.
+While the first two scenarios do not comply with the Climate Action and Low Carbon Development (Amendment) Act 2021 target of an overall 51\% GHG reduction in 2030 compared to 2018, they do provide insights on the incremental steps required to achieve these targets.
 The last four scenarios comply with 2030 and 2050 targets and offer insights about different trajectories to achieve those targets.
@@
 \label{Results:Residential }
-The residential sector reduces $CO_2$ emissions by between 63\% - 71\% in 2030 for the four scenarios which comply with the Climate Bill.
-Compare this to the ``No Mitigation" scenario, which shows an increase of 4 \% $CO_2$ in 2030 and the  ``WAM" scenario which shows a 8 \% reduction in $CO_2$ by 2030.
+The residential sector reduces $CO_2$ emissions by between 63-71\% in 2030 for the four scenarios which comply with the Climate Bill.
+Compare this to the ``No Mitigation" scenario, which shows an increase of 4\% $CO_2$ in 2030 and the  ``WAM" scenario which shows a 8\% reduction in $CO_2$ by 2030.
 These results assist in representing the scale of decarbonisation required in the residential sector.
@@
 
-In 2018, the residential sector had 124 PJ of primary energy, which is needed to decrease by 8-13 \% in 2030, to comply with the Climate Bill.
+In 2018, the residential sector had 124 PJ of primary energy, which is needed to decrease by 8-13\% in 2030, to comply with the Climate Bill.
 Results show a large uptake of electricity in the residential sector within Climate Bill compliant scenarios.
 The large increase in ambient energy indicates the type of technology deployed in the residential sector,
-the overall share of ambient heat increases from 1.4 \% in 2018 to between 13-26 \% in 2030, more on this in Figure \ref{fig:HeatPump}.
-In Climate Bill compliant scenarios, peat and coal are completely phased out, while oil use decreases 90-93 \% by 2030.
-Gas energy increase by 34 \% in the A51E51 scenario, but shows little change in other Climate Bill compliant scenarios in 2030.
+the overall share of ambient heat increases from 1.4\% in 2018 to between 13-26\% in 2030, more on this in Figure \ref{fig:HeatPump}.
+In Climate Bill compliant scenarios, peat and coal are completely phased out, while oil use decreases 90-93\% by 2030.
+Gas energy increase by 34\% in the A51E51 scenario, but shows little change in other Climate Bill compliant scenarios in 2030.
 
@@
 
-The number of retrofits varies from 393,000 in A51E51 to 615,000 in A25E65, meaning between 23 - 36 \% of existing dwellings will require some form of retrofit.  
+The number of retrofits varies from 393,000 in A51E51 to 615,000 in A25E65, meaning between 23-36\% of existing dwellings will require some form of retrofit.  
 As mentioned in Table \ref{tab:HeatPumpConstraint}, the rollout of retrofits constraints the amount of heat pumps which can be installed, however, a certain share of existing dwellings and all new dwellings are heat pump ready, i.e B2 rating or above.
@@
 In Figure \ref{fig:HeatPump}, SW represents space and water heating, HC is space heating and space cooling, and A-to-W represents air-to-water type heat pumps.
-The optimal number of heat pumps in each scenario ranges from 492,000 in A51E51 to 938,000 in A25E65, which equates to 23 - 45 \% of dwellings having a heat pump in 2030.
+The optimal number of heat pumps in each scenario ranges from 492,000 in A51E51 to 938,000 in A25E65, which equates to 23-45\% of dwellings having a heat pump in 2030.
 Close to 45,000 air to water heat pumps are installed in apartments across all scenarios, as there are no retrofits in apartments, these heat pumps are installed in the existing B2 rating or above apartments or new apartments.
@@
 The ``No Mitigation" scenario is expected to cost $\text{\euro} 8.1 $ billion in 2030, that figure rises to $\text{\euro} 12.4 $ billion under the A51E51 scenario, with electricity ( Power generation ) requiring an extra $\text{\euro} 2 $ billion in annual investment in 2030.
-\par Moving from ``No Mitigation" to A51E51, the residential sector requires $\text{\euro} 0.7 $ billion annual extra investment in 2030 or 16.3 \% of the extra investment across the energy system.
-While moving from ``No Mitigation" to the most ambitious energy system scenario, A25E65, would require  $\text{\euro} 1.5 $ billion annual extra residential investment in 2030 or 15 \% of the extra investment across the energy system.
+\par Moving from ``No Mitigation" to A51E51, the residential sector requires $\text{\euro} 0.7 $ billion annual extra investment in 2030 or 16.3\% of the extra investment across the energy system.
+While moving from ``No Mitigation" to the most ambitious energy system scenario, A25E65, would require  $\text{\euro} 1.5 $ billion annual extra residential investment in 2030 or 15\% of the extra investment across the energy system.
 
@@
 While LEAP is another powerful tool within the Energy Policy and Modelling Group of MaREI, which can be used to analyse individual climate policies.
-\par The model will require some future developments in the residential sector, district heating, and renewable gas.
+\par The model will require some future developments in the residential sector, DH, and renewable gas.
 The large growth of data centres expected in Ireland can provide waste heat to DH networks.
@@
 \end{table}
-
+\newpage
 \bibliography{TIMref.bib}
@@
 \end{document}
-%\endinput
-%%
-%% End of file `elsarticle-template-harv.tex'.

### 2021-08-19 manuscript.tex
Most warning gone and need to work on results next

@@
 Exogenous technical, physical, and regulatory constraints can be applied and technologies have specified fuel types, efficiency, lifetime, availability, environmental characteristics, and fixed, variable and operation costs.
+TIM is set up with regional options, either one national region or 26 county regions, only the annual results were explored in this study.
 TIMES uses a linear programming optimizer matrix in General Algebraic Modeling System (GAMS), to compute all calculations.
@@
 \par Over half of Irish dwellings are captured in the BER database and every BER displays over 200 energy related variables.
-The DEAP model which is used to assign BER labels,  is outlined by Dineen et al.\cite{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM}.
-In terms of space heating, DEAP assumes that each dwelling is single zoned, the heating season is from October to May inclusive, and the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season.
+The DEAP model which is used to assign BER labels,  is outlined by \citet{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM}.
+In terms of space heating, DEAP assumes that each dwelling is single zoned, the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season, which is from October to May inclusive and assumes monthly external temperatures, the average external temperature for the heating season is 7.6$^{\circ}$C.
 One assumption, the internal temperature assumption, results in a large error, where space and water heating are overestimated in DEAP.
@@
 Where $ {\text{\euro}}{}_{rf-rt,a}$ is the retrofit cost from pre-retrofit BER rating (rf) to post-retrofit rating (rt) for archetype (a).
-$Ali_{r,t}$ is the cost data from Ali et al.
-\cite{Ali2020ABuildings}, it contains all BER ratings and retrofit types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments, 1 for attached, and 1.1 for detached dwellings.
+$Ali_{r,t}$ is the cost data from  \citet{Ali2020ABuildings}, it contains all BER ratings and retrofit types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments, 1 for attached, and 1.1 for detached dwellings.
 $TABULA_{rf-rt,a}$ is the average retrofit cost from pre-retrofit BER rating to post-retrofit BER rating for an archetype, data is sourced from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland}.
 However with only 62 samples in total not all retrofits are included.
-In this case, TABULA is ignored and Ali et al.
-is the only source of retrofit costs.
+In this case, TABULA is ignored and \citet{Ali2020ABuildings} is the only source of retrofit costs.
 The retrofit costs are cross-checked with the residential cost-optimal report \cite{AECOM2020ReportBuildings}.
@@
 
-The previous section outlined the cost of retrofitting, and this section looks at the expected savings of retrofitting existing dwellings and building new dwellings.  
-The annual heating costs per archetype and per BER rating are sourced from SEAI's indicative energy running costs and the internal temperature assumptions in Table \ref{Table:InternalTemps} is included to calculate the costs, which are outlined in Table \ref{Table:FuelCost}
+The previous section outlined the cost of retrofitting, and this section looks at the expected savings of retrofitting existing dwellings and building regulations in new dwellings.  
+The annual heating costs per archetype and per BER rating are sourced from SEAI's indicative energy running costs.
+The internal temperature assumptions in Table \ref{Table:InternalTemps} are included to calculate the costs, which are outlined in Table \ref{Table:FuelCost}
 
@@
 The  technical  bottom-up  development  approach  of  the  residential  sector  is  combined with other economic sectors to meet macroeconomic defined demands.
-The energy service demands in the residential sector are driven by the total dwelling stock for non-archetype energy services and apartment, attached or detached dwelling stock for archetype energy services.
-\par The residential stock projections up to 2040 are taken from Economic and Social Research Institute's (ESRI's) housing demand estimates \cite{Bergin2020RegionalLevel}, beyond 2040 the growth rate is extrapolated.
-However, archetpye specfic growth is limited or unavailable, for this reason a simple linear regression, as shown in Eq.
+The energy service demands in the residential sector are driven by the total dwelling stock for non-archetype energy services and apartment, attached or detached for archetype energy services.
+The residential stock projections up to 2040 are taken from ESRI's housing demand estimates \cite{Bergin2020RegionalLevel}, beyond 2040 the growth rate is extrapolated.
+However, the archetype specific growth is limited or unavailable, for this reason a simple linear regression, as shown in Eq.
 \ref{Eq:RegArchetype} is used to calculate the share of an archetype in a given year (y).
 This is graphically represented in Figure \ref{fig:DwellingStoc}.
-The growth rate of apartments is the highest and increasing up to 2040, while attached and detached dwellings are decreasing, indicating a preference for smaller dwellings to increase in the future.
-The obsolescence rate of the existing dwellings is 0.35\%.
+The growth rate of apartments is the highest and increasing up to 2040, while the growth in attached and detached dwellings are lagging behind.
+The obsolescence rate of the existing dwellings is taken as 0.35\%.
 
@@
 The demand for space and water heating is outlined in the previous section, and this section explains how water and space heating are achieved through technology and retrofits.
-While the demands are exogenous, the choice of technologies to at least satisfy the demand are endogenous and is captured in Equation \ref{Eq:Technology}.
+While the demands are exogenous, the choice of technologies to at least satisfy the demand are endogenous and is captured in Eq.
+\ref{Eq:Technology}.
 
@@
 All space and water heating technologies are archetype specific and some technologies deliver both water and space heating, while other technologies can only deliver one type of heating, the list of available technologies is outlined in \ref{sec:appendixA4}.
-Retrofits are included as they satisfy the residential space heating demand, even though retrofits do not consume fuel similar to a PV system, from that perspective.
+Retrofits are included as they satisfy the residential space heating demand, even though retrofits do not consume fuel similar to a solar water heating system, from that perspective.
 Danish Energy Agency's Technology Data for Individual Heat Production dataset was used  was input to TIM \cite{JointResearchCentre2017Techno-economicEU}, this dataset is also used in the EU TIMES model, the dataset has high granularity and is continuously updated.
@@
 In TIM, the number of main heating systems equals the number of dwellings, this assumption ignores any impact of secondary system heating.
-One exception to the rule is stoves, which can only be used as secondary heating systems.
-
+One exception to the rule is stoves are considered as secondary heating systems only.
 The base year heating energy consumption is derived from the BER database, but uses census data \cite{CentralStatisticsOffice2016CensusIreland} from the CSO dataset E1008 to calibrate the fuel used in heating systems per archetype.
@@
 
-Lighting, appliances, and pumps \& fans are all fueled by electricity, whereas cooking can be fueled by electricity, gas or LPG.
 As shown in Figure \ref{fig:IrelandServiceDemands}, these energy services accounted for 20\% of energy consumption in 2018, however, this share of consumption is expected to increase. 
 Refrigeration, cloth washing, cloth drying, and dish washing are under the appliances category and data is sourced from Topten \cite{Michel2016EnergyREPORT} and SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
-Cooking is also sourced from Eurostat's ``Energy consumption in households by type and end-used" excel datasheet \cite{Eurostat2021Disaggregatednrg_d_hhq} and cross-checked with SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
-\par Appliances not included in this list are simply under the ``Other Appliances" category in TIM and these final values are cross-checked with POTEnCIA.
-Other Appliances include television, multimedia, and computer technology.
+Cooking data is sourced from Eurostat's ``Energy consumption in households by type and end-used"  \cite{Eurostat2021Disaggregatednrg_d_hhq} and cross-checked with SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
+\par Appliances not included in this list are simply under the ``Other Appliances" category in TIM.
+Other Appliances include television, multimedia, outdoor appliances, and computer technology.
 Lighting and pumps \& fans are archetype specific energy services as the data is sourced from the BER database.
-Equation \ref{Eq:Technology} is also used to find the technology in the energy service demand, but unlike the heating energy service demands, there is little or no choice in different technologies (k) for some of the demand categories (i), therefore, the model is restricted in lighting, appliances and pumps \& fans.
+\par Equation \ref{Eq:Technology} is also used to optimal choose the technology in the energy service demand, but unlike in the heating energy service demands, there is little or no choice in different technologies (k) for some of the demand categories (i), therefore, the model is restricted in lighting, appliances and pumps \& fans.
 However, the cost and efficiency of each technology does determine the overall investment.
@@
 The User Constraints applied to the residential sector include:
-\begin{itemize}
- \item One Space and Water Heating System per household ( stoves excluded ) 
+\begin{enumerate}[1.]
+ \item Only one space and water heating system per household ( stoves excluded ) 
  \item Maximum 60\% of existing buildings to be retrofitted by 2030, and 95\% by 2070. Maximum interpolated between 2030 and 2070. 
@@
 \item Maximum 30\% gas input to gas hybrid heating pumps
-\end{itemize}
+\end{enumerate}
 
@@
 \begin{split}
-VAR\_CAP(t + 1) \leq (1 + G^{M(t+1)-M(t)}) \cdot VAR\_CAP(t)+K
+VAR\_CAP(t + 1) \leq (1 + G^{M(t+1)-M(t)}) \cdot VAR\_CAP(t)
 \end{split}
@@
 The quantity $M(t+1)-M(t)$ is the number of years between the milestones of time periods t and t+1.
-The constant K is useful whenever the technology has no capacity initially, to allow the capacity to build over time.
-
+
+%PICKUPHERE
 \section{Results}

### 2021-08-23 manuscript.tex
23 Aug

@@
 
-%PICKUPHERE
 \section{Results}
@@
 
-The results from the newly developed TIM provide some valuable insights on each sector and the sectoral interconnections.
-The results outlined in this paper are based upon six scenarios, which are focused around the 2030 target.
-While TIM can only analysis the energy system, the other large source of GHG emissions is from agriculture.
-Therefore, some scenarios are labelled based on agriculture 2030 GHG reduction percentage and energy system 2030 GHG reduction percentage, below is a brief description of the six scenarios:
+The results from TIM can provide some valuable insights on the residential sector and the sectoral interconnections when targeting carbon budgets, and long-term climate targets.
+While the purpose of this paper was to describe the residential methodology, the results used here only scrape the surface in terms of exploring residential decarbonisation.
+However, the results provide some initial insights into residential decarbonisation and are discussed in section \ref{Discussion}.
+Three different scenarios are used to explore the results, those scenarios and descriptions are:
+
 \begin{enumerate}[(a)]
     \item \textbf{No Mitigation} Least cost solution, where no emissions constraints are imposed
-    \item \textbf{With Additional Measures (WAM)} includes the implementation of Ireland’s CAP 2019 towards 2030 and net-zero in 2050.
-    \item \textbf{A25E65} Agriculture GHG emissions reduce by 25\%, energy system GHG emissions reduces by 65\% by 2030 compared to 2018 \& net-zero in 2050 
-    \item \textbf{A33E61} Agriculture GHGs reduces by 33\% and Energy system GHGs by 61\% \& net-zero in 2050 
-    \item \textbf{A40E57} Agriculture GHGs reduces by 40\% and energy system GHGs by 57\% \& net-zero in 2050 
-    \item \textbf{A51E51} Equal GHG reduction share in of 51\% in Agriculture and energy Systems \& net-zero in 2050 
+    \item \textbf{Previous Effort } complies with Climate Action Plan 2019, reduces Non-ETS GHG by 30\% in 2030 compared to 2005, and net-zero in 2050.
+    \item \textbf{Current Effort} complies with Climate Action Bill 2021, energy system GHG emissions reduces by 65\% by 2030 compared to 2018 \& net-zero in 2050 ( complies if agriculture GHG emissions reduces by 25\% in 2030 ).
 \end{enumerate}
@@
 While the first two scenarios do not comply with the Climate Action and Low Carbon Development (Amendment) Act 2021 target of an overall 51\% GHG reduction in 2030 compared to 2018, they do provide insights on the incremental steps required to achieve these targets.
-The last four scenarios comply with 2030 and 2050 targets and offer insights about different trajectories to achieve those targets.
 
@@
 \label{Results:Residential }
-The residential sector reduces $CO_2$ emissions by between 63-71\% in 2030 for the four scenarios which comply with the Climate Bill.
-Compare this to the ``No Mitigation" scenario, which shows an increase of 4\% $CO_2$ in 2030 and the  ``WAM" scenario which shows a 8\% reduction in $CO_2$ by 2030.
-These results assist in representing the scale of decarbonisation required in the residential sector.
-
-Focusing on the residential fuel-mix gives further insights on the optimal pathways to achieve 2030 targets.
-Figure \ref{fig:ResFuel} shows this, with the base year being represented on the top row, the four Climate Bill compliant scenarios on rows 2-5, and the last two rows are not compliant with the Climate Bill.
+
+The quantity and mix of fuel going into the residential for each scenario and year is  shown in Figure \ref{fig:ResFuel}.
+The base year is shown on the left, while annual snapshots of the fuel-mix for the other scenarios are shown for 2030, 2040, and 2050.
 
@@
  \centering
- \includegraphics[width=\textwidth]{Figures/RSD_Fuel2.png}
- \caption{Residential Fuel per scenario}
+ \includegraphics[width=\textwidth]{Figures/residential_fuelmix.png}
+ \caption{Residential fuel-mix per scenario per year}
  \label{fig:ResFuel}
@@
 
-In 2018, the residential sector had 124 PJ of primary energy, which is needed to decrease by 8-13\% in 2030, to comply with the Climate Bill.
-Results show a large uptake of electricity in the residential sector within Climate Bill compliant scenarios.
-The large increase in ambient energy indicates the type of technology deployed in the residential sector,
-the overall share of ambient heat increases from 1.4\% in 2018 to between 13-26\% in 2030, more on this in Figure \ref{fig:HeatPump}.
-In Climate Bill compliant scenarios, peat and coal are completely phased out, while oil use decreases 90-93\% by 2030.
-Gas energy increase by 34\% in the A51E51 scenario, but shows little change in other Climate Bill compliant scenarios in 2030.
-
-Figure \ref{fig:Retrofits} shows both the number of retrofits per type and the energy savings up to 2030, for each of the four Climate Bill compliant scenarios.
+In 2018, the residential sector required 124 PJ of primary energy, the increase in households, and energy efficiency improvements resulted in a 9\% primary energy increase in 2030, under the Previous Effort scenario.
+Whereas under the Current Effort scenario, the primary energy is required to reduce by 13\% in 2030, this highlights the large incremental step in ambition required when moving from Climate Action Plan 2019 to the Climate Action Bill 2021.
+In 2050, under the Previous Effort scenario, the total primary energy shows a continued growth 17\% higher than 2018, whereas the Current Effort scenario shows an overall reduction of 11\% which indicates a post-2030 growth.
+The fluctuation in the fuel-mix required is significant, especially to comply with 2030 targets, for example, oil reduces 60\% in the Previous Effort and 90\% in the Current Effort scenario in 2030.
+Furthermore, peat and coal are phased out in the residential sector in both scenarios.
+The lesser carbon intensive natural gas increases 211\% in Previous Effort without the upper constraint of the gas infrastructure and decreases 4\% in Current Effort scenario.
+The electricity continues to decarbonise rapidly in all scenarios, resulting in an increase in electricity in the residential sector.
+In both scenarios, the carbon intensity of electricity ($gCO_2/kWh$) falls by two-thirds, with the Current Effort scenario only marginally lower than the Previous Effort scenario.
+New emerging fuels which are prevalent in the Current Effort include space ambient heat and water ambient heat due to the increase of heat pumps in the residential sector, more on this in Figure \ref{fig:HeatPump}.
+
+The optimal number of retrofits by type and quantity up to 2030, 2040 and 2050 are summarised in Figure \ref{fig:Retrofits}.
 
@@
  \centering
- \includegraphics[width=\textwidth]{Figures/Retrofits.png}
+ \includegraphics[width=\textwidth]{Figures/NumberofRetrofits.png}
  \caption{Retrofits by type per scenario}
@@
 
-The number of retrofits varies from 393,000 in A51E51 to 615,000 in A25E65, meaning between 23-36\% of existing dwellings will require some form of retrofit.  
+While the Climate Action Plan 2019 targeted 500,000 retrofits, the model only retrofitted 90,000 up to 2030, however, all these retrofits are deep and on larger detached dwellings.
+The number of retrofits jumps up to 615,000 in the Current Effort scenario, there is a mix of types of depth of retrofits with different dwelling types, but no apartments are retrofitted.
+The cumulative energy savings are displayed on the secondary axis.
+
 As mentioned in Table \ref{tab:HeatPumpConstraint}, the rollout of retrofits constraints the amount of heat pumps which can be installed, however, a certain share of existing dwellings and all new dwellings are heat pump ready, i.e B2 rating or above.
@@
 
+The residential sector reduces $CO_2$ emissions by between 63-71\% in 2030 for the four scenarios which comply with the Climate Bill.
+Compare this to the ``No Mitigation" scenario, which shows an increase of 4\% $CO_2$ in 2030 and the  ``WAM" scenario which shows a 8\% reduction in $CO_2$ by 2030.
+These results assist in representing the scale of decarbonisation required in the residential sector.
+
 \subsection{Intersectoral Results}
@@
 \label{Discussion}
-
+SOLAR??
 The methodology applied to the residential sector in TIM is transparent, up-to-date, and robust.

### 2021-08-24 manuscript.tex
New Version for talk with Olex and Fionn

@@
 Retrofitting and electrification are a fundamental part of residential decarbonisation in Ireland, and therefore play a central role in the development of the model described here.
-Furthermore, the new model includes technologies, which were not previously considered, one such technology is District Heating (DH), some initial studies have found it may be optimal to have 37\% of heat demand to be provided by DH in Ireland \cite{Thellufsen2019ImplementingIreland}, the full list of space and water heating technologies considered in the model can be found in \ref{sec:appendixA4}.
+Furthermore, the new model includes technologies, which were not previously considered, one such technology is District Heating (DH), some initial studies have found it may be optimal to have 37\% of heat demand to be provided by DH in Ireland \cite{Thellufsen2019ImplementingIreland}, the full list of space and water heating technologies considered in the model can be found in Table \ref{AppHeatingTech}.
 The strength of an ESOM can be used to examine the sectoral interactions, one study by Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} examined the interaction between heat pumps and wind generation investments in Denmark, the results find it is optimal to install air-to-water heat pumps in non-DH areas.
@@
 To compare from the two databases, fuels were were combined and condensed down to 13 fuels in TIM.
-Table \ref{Table:Fuels} shows an overview of the fuels combined from the different data sources.
+Table \ref{AppTable:Fuels} shows an overview of the fuels combined from the different data sources.
 
@@
 While the BER database provides high granularity data, it must be filtered to improve the representation of the dwelling stock data.
-The filters applied to the raw BER database are outlined in \ref{sec:appendixA1}.
+The filters applied to the raw BER database are outlined in Table \ref{AppBERFilters}.
 
@@
 $BER_{r,y,a}$ is the number of BER ratings by year of construction and archetype from the BER database and dividing this value by $BER_{y,a}$ results in a probability of a certain age type falling within a BER rating, and the final part of the equation $\frac{TIM_{a}} {BER_{a}}$ compensates for the non-stated archetypes in the CSO data, $TIM_a$ was already calculated in equation \ref{Eq:BERtoCSO2}.
-The results of each of the correlations are in  \ref{sec:appendixA2}.
-Table \ref{Table:FinalDwellingStock} outlines the final dwelling stock used in TIM.
+The results of each of the correlations are in  Table \ref{AppBERCor},\ref{AttBERCor} \& \ref{DetBERCor} and Table \ref{Table:FinalDwellingStock} outlines the final dwelling stock used in TIM.
 
@@
 \par TIM has two types of retrofitting options per archetype,  a shallow retrofit is when the primary energy ($kWh/m^2/yr$) is reduced by $ \leq 35\%$, and deep retrofit is where the primary energy is reduced by $>35\%$, while a reduction of $ \ >75\%$ is considered infeasible.
-The difference between the pre-retrofit and post-retrofit primary energy determines the type of retrofit, as shown in \ref{sec:appendixA3}.
+The difference between the pre-retrofit and post-retrofit primary energy determines the type of retrofit, as shown in Table \ref{AppDepthofRetrofit}.
 
@@
 
-Allocating each retrofit change to either shallow or deep is the next step, this is shown in \ref{sec:appendixA3}.
+Allocating each retrofit change to either shallow or deep is the next step, this is shown in Table \ref{AppDepthofRetrofit}.
 The next calculation is to find the weighted average of each retrofit depth per archetype, this is then a TIM input.
@@
 
-All space and water heating technologies are archetype specific and some technologies deliver both water and space heating, while other technologies can only deliver one type of heating, the list of available technologies is outlined in \ref{sec:appendixA4}.
+All space and water heating technologies are archetype specific and some technologies deliver both water and space heating, while other technologies can only deliver one type of heating, the list of available technologies is outlined in Table \ref{AppHeatingTech}.
 Retrofits are included as they satisfy the residential space heating demand, even though retrofits do not consume fuel similar to a solar water heating system, from that perspective.
@@
 
-The results from TIM can provide some valuable insights on the residential sector and the sectoral interconnections when targeting carbon budgets, and long-term climate targets.
+The results from TIM can provide some valuable insights on the residential sector and the sectoral interconnections in achieving long-term climate targets.
 While the purpose of this paper was to describe the residential methodology, the results used here only scrape the surface in terms of exploring residential decarbonisation.
 However, the results provide some initial insights into residential decarbonisation and are discussed in section \ref{Discussion}.
-Three different scenarios are used to explore the results, those scenarios and descriptions are:
+Two different scenarios are used to explore the results and those scenarios and descriptions detailed here:
 
 \begin{enumerate}[(a)]
-    \item \textbf{No Mitigation} Least cost solution, where no emissions constraints are imposed
-    \item \textbf{Previous Effort } complies with Climate Action Plan 2019, reduces Non-ETS GHG by 30\% in 2030 compared to 2005, and net-zero in 2050.
+
+    \item \textbf{Previous Effort } complies with Climate Action Plan 2019, reduces Non-ETS GHG by 30\% in 2030 compared to 2005, and post-2030 targets are forward extrapolated.
     \item \textbf{Current Effort} complies with Climate Action Bill 2021, energy system GHG emissions reduces by 65\% by 2030 compared to 2018 \& net-zero in 2050 ( complies if agriculture GHG emissions reduces by 25\% in 2030 ).
@@
 
-While the first two scenarios do not comply with the Climate Action and Low Carbon Development (Amendment) Act 2021 target of an overall 51\% GHG reduction in 2030 compared to 2018, they do provide insights on the incremental steps required to achieve these targets.
-
 \subsection{Residential Results}
@@
 The base year is shown on the left, while annual snapshots of the fuel-mix for the other scenarios are shown for 2030, 2040, and 2050.
+
+In 2018, the residential sector required 124 PJ of primary energy, the increase in households, and energy efficiency improvements resulted in a 9\% primary energy increase in 2030, under the Previous Effort scenario.
+Whereas under the Current Effort scenario, the primary energy is required to reduce by 13\% in 2030, this highlights the large incremental step in ambition required when moving from Climate Action Plan 2019 to the Climate Action Bill 2021.
+In 2050, under the Previous Effort scenario, the total primary energy shows a continued growth 17\% higher than 2018, whereas the Current Effort scenario shows an overall reduction of 11\% which indicates a post-2030 growth, highlighting the high ambition in the 2030 target.
 
@@
 
-In 2018, the residential sector required 124 PJ of primary energy, the increase in households, and energy efficiency improvements resulted in a 9\% primary energy increase in 2030, under the Previous Effort scenario.
-Whereas under the Current Effort scenario, the primary energy is required to reduce by 13\% in 2030, this highlights the large incremental step in ambition required when moving from Climate Action Plan 2019 to the Climate Action Bill 2021.
-In 2050, under the Previous Effort scenario, the total primary energy shows a continued growth 17\% higher than 2018, whereas the Current Effort scenario shows an overall reduction of 11\% which indicates a post-2030 growth.
 The fluctuation in the fuel-mix required is significant, especially to comply with 2030 targets, for example, oil reduces 60\% in the Previous Effort and 90\% in the Current Effort scenario in 2030.
-Furthermore, peat and coal are phased out in the residential sector in both scenarios.
-The lesser carbon intensive natural gas increases 211\% in Previous Effort without the upper constraint of the gas infrastructure and decreases 4\% in Current Effort scenario.
+Furthermore, peat and coal are phased out completely in the residential sector in both scenarios in 2030.
+The lesser carbon intensive natural gas increases by 211\% in Previous Effort without the upper constraint of the gas infrastructure and decreases 4\% in Current Effort scenario.
 The electricity continues to decarbonise rapidly in all scenarios, resulting in an increase in electricity in the residential sector.
@@
 
-The optimal number of retrofits by type and quantity up to 2030, 2040 and 2050 are summarised in Figure \ref{fig:Retrofits}.
+The optimal cumulative number of retrofits by type and quantity up to 2030, 2040, and 2050 are summarised in Figure \ref{fig:Retrofits}.
 
@@
 While the Climate Action Plan 2019 targeted 500,000 retrofits, the model only retrofitted 90,000 up to 2030, however, all these retrofits are deep and on larger detached dwellings.
-The number of retrofits jumps up to 615,000 in the Current Effort scenario, there is a mix of types of depth of retrofits with different dwelling types, but no apartments are retrofitted.
-The cumulative energy savings are displayed on the secondary axis.
-
-As mentioned in Table \ref{tab:HeatPumpConstraint}, the rollout of retrofits constraints the amount of heat pumps which can be installed, however, a certain share of existing dwellings and all new dwellings are heat pump ready, i.e B2 rating or above.
+The cumulative energy saving is displayed on the secondary axis, the savings from the 90,000 retrofits continues to increase, despite no new retrofits post-2030.
+The number of retrofits jumps up to 615,000 in the Current Effort scenario up to 2030, there is a mix of retrofits with different dwelling types, but no apartments are retrofitted.
+Again, the difference in ambition is very apparent and the expected savings up to 2030, which is four times higher in the Current Effort compared to the Previous Effort scenario stretching to almost five times higher by 2050.
+
+As mentioned in Table \ref{tab:HeatPumpConstraint}, retrofits are sometimes required before a heat pump can be installed.
+A dwelling must hold approx.
+B2 BER to be ``heat pump ready".
+The model assumes the lifetime of a Heat Pump is 20 years, so it maybe possible for one dwelling to get two new heat pumps over the next 30 years, unlike retrofits, heat pumps can be applied to the new dwellings.
+In Figure \ref{fig:HeatPump} the total cumulative number of heat pumps installed up to a given year is shown.
 
@@
 In Figure \ref{fig:HeatPump}, SW represents space and water heating, HC is space heating and space cooling, and A-to-W represents air-to-water type heat pumps.
-The optimal number of heat pumps in each scenario ranges from 492,000 in A51E51 to 938,000 in A25E65, which equates to 23-45\% of dwellings having a heat pump in 2030.
-Close to 45,000 air to water heat pumps are installed in apartments across all scenarios, as there are no retrofits in apartments, these heat pumps are installed in the existing B2 rating or above apartments or new apartments.
-The detached air-to-water heat pumps are the most prevalent across scenarios and maybe considered the ``low hanging fruit".
-
-The residential sector reduces $CO_2$ emissions by between 63-71\% in 2030 for the four scenarios which comply with the Climate Bill.
-Compare this to the ``No Mitigation" scenario, which shows an increase of 4\% $CO_2$ in 2030 and the  ``WAM" scenario which shows a 8\% reduction in $CO_2$ by 2030.
-These results assist in representing the scale of decarbonisation required in the residential sector.
+The optimal number of heat pumps in each scenario ranges from 39,000 to 938,000, which equates to 2-45\% of dwellings having a heat pump by 2030 depending on which scenario is used.
+Similar to the retrofit results, the Current Effort scenario shows both a higher quantity and mix.
+While A-to-W heat pumps are the predominant type, gas hybrid heat pumps begin to gain some momentum and account for 7\% of all new heat pumps by 2030 and increases to 14\% by 2040.
+
+The residential sector reduces $CO_2$ emissions by between 8-70\% in 2030, depending on the scenario explored, again highlighting the scale in step of ambition required in the residential sector.
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[width=\textwidth]{Figures/RSDCO2.png}
+ \caption{Residential Emissions}
+ \label{fig:RSDCO2}
+\end{figure}
 
@@
 The advantage of an energy system model is to observe the intersectoral connections.
-Figure \ref{fig:SectorFuel} presents  the base year primary energy fuel-mix for four sectors - power generation, transport, services, and residential and the 2030 fuel-mix for the four Climate Bill compliant scenarios.
+Figure \ref{fig:SectorFuel} presents the base year primary energy fuel-mix for four sectors - power generation, transport, services, and residential in 2030, 2040, and 2050 for both scenarios.
 
@@
  \centering
- \includegraphics[width=\textwidth]{Figures/IntesectoralFuels.png}
+ \includegraphics[width=\textwidth]{Figures/Intersectoralresults.png}
  \caption{Sector Annual Fuel in 2018 and four scenarios in 2030}
@@
 
-Some consistencies across the scenarios show the large growth of renewables in the power generation sector and consequently the uptake of electricity across the remaining sectors.
-The least ambitious A51E51 scenario shows 59 PJ of electricity used in residential in 2030, but as constraints tighten, the electricity is used more in transport, while the residential sector invests more in solar panels and other forms of renewable energy.
-Because bioenergy is limited, the use of bioenergy in transport and services also affects the residential sector.
-
-By examining the same results further along the time horizon in 2050, we can begin to get a sense of the action required pre-2030 to help achieve net-zero in 2050.
-
-\begin{figure}[!htbp]
- \centering
- \includegraphics[width=\textwidth]{Figures/Intersectoralfuel2050.png}
- \caption{Sector Annual Fuel in 2018 and four scenarios in 2050}
- \label{fig:SectorFuel2050}
-\end{figure}
-
-The results in Fig.
-\ref{fig:SectorFuel2050} show a growth of low carbon electricity in power generation in 2050 compared to 2030 results in Fig.
-\ref{fig:SectorFuel}.
-The results indicate increased electrification pre-2030 across all sectors and a substantial increase of energy efficiency, as primary energy does not correlate with expected growth.
-
-The ``No Mitigation" scenario is expected to cost $\text{\euro} 8.1 $ billion in 2030, that figure rises to $\text{\euro} 12.4 $ billion under the A51E51 scenario, with electricity ( Power generation ) requiring an extra $\text{\euro} 2 $ billion in annual investment in 2030.
-\par Moving from ``No Mitigation" to A51E51, the residential sector requires $\text{\euro} 0.7 $ billion annual extra investment in 2030 or 16.3\% of the extra investment across the energy system.
-While moving from ``No Mitigation" to the most ambitious energy system scenario, A25E65, would require  $\text{\euro} 1.5 $ billion annual extra residential investment in 2030 or 15\% of the extra investment across the energy system.
+Some consistencies across the scenarios show the large growth of renewables in the power generation sector and consequently the uptake of electricity across the remaining sectors, especially in the Current Effort scenario.
+The Previous Effort scenario opts for natural gas, backed by low carbon intensity electricity, the increase in ambition and tightening of constraints requires natural gas to decrease, with electricity becoming the cornerstone of energy-system decarbonisation.
 
@@
 \label{Discussion}
-SOLAR??
-The methodology applied to the residential sector in TIM is transparent, up-to-date, and robust.
-The previous version of TIM, namely, the Ireland TIMES model, which was used for almost 10 years, and was continuously tweaked to react to the ever changing climate policy, climate targets, and technological advancements.
+
+The previous version of TIM, namely, the Ireland TIMES model, was used for almost 10 years and was continuously tweaked to react to the ever-changing climate policy, climate targets, and technological advancements.
 \par TIM is developed to provide insights on sectoral carbon budgets and 2030 and 2050 climate targets.
 As the first iteration of TIM, the core model is expected to be improved, while on-going research will continue to  analyse certain sectors in greater detail.
-The large increase in electricity ?demand?
-may require some soft-linking with PLEXOS to investigate power generation operating constraints and reliability of supply.
+The large increase in electricity demand may require some soft-linking with PLEXOS to investigate power generation operating constraints and reliability of supply.
 While LEAP is another powerful tool within the Energy Policy and Modelling Group of MaREI, which can be used to analyse individual climate policies.
-\par The model will require some future developments in the residential sector, DH, and renewable gas.
-The large growth of data centres expected in Ireland can provide waste heat to DH networks.
-While the capacity of biomethane in Ireland can effect the potential decarbonisation of natural gas, adding biomethane to the natural gas grid has not been accounted for in the model.
 \par The dissagreation of archetypes by BER ratings will add to the model complexity, but help provide insights into optimal BER ratings to target in retrofitting and heat pump installation.
@@
 
-The model will be open sourced and be used to develop research and reports, while also used for insights by the Climate Change Advisory Council and the Irish government.
-During development of the model, the developers reached out to experts for feedback on results and the process will continue as the model is improved by a resourced modelling team.
-The model provides key insights into the cost and scale of retrofits and heat pumps required, which are currently being addressed under CAP2019 and are therefore are likely pathways towards achieving climate targets.
-TIM also takes into account new technologies such as DH, Hydrotreated Vegetable Oil (HVO), and solar.
+The methodology applied to the development of the residential sector in TIM is transparent, up-to-date, and robust.
+The model will be open sourced and be used to develop research and reports, while also used for insight by the Climate Change Advisory Council and the Irish government.
+During the development of the model, TIM developers reached out to experts for feedback on results and the process will continue as the model is improved by a resourced modelling team.
+The model provides key insights into the cost and scale of the retrofits and heat pumps required, which are currently being addressed under CAP2019 and are therefore are likely pathways towards achieving climate targets.
+TIM takes into account new technologies such as DH, Hydrotreated Vegetable Oil (HVO), and solar.
 The model horizon is up to 2050, so the incremental steps and targets on the way to net-zero can be analysed.  
 Moreover, the usual advantage of using ESOM over bottom-up residential models applies to TIM, as the intersectoral effects can be explored.
-The joint effort of different researchers across different universities in Ireland, to output a dwelling stock, using a new filtering process to the BER database helps provide a more robust base year dwelling stock, this filtration process is expected to be improved and TIM can benefit from this process.
-The internal temperature calculation applied in TIM reflects the actual residential energy consumption.
+%The joint effort of different researchers across different universities in Ireland, to output a dwelling stock, using a new filtering process to the BER database helps provide a more robust base year dwelling stock, this filtration process is expected to be improved and TIM can benefit from this process.
+The internal temperature calculation applied in TIM reflects will strengthen the model by better representing the actual residential energy consumption.
 
@@
 
-TIM is not a model to look at specific policies nor is it ideally suited to explore occupancy behaviour, instead it provides insights on optimal technological pathways.
+TIM is not a model to look at specific policies nor is it ideally suited to explore behaviour, instead it provides insights on optimal technological pathways.
 TIMES is an ESOM which assumes perfect foresight, we know this not to be entirely true, consumer choice is not only based on costs.
-The results in TIM show large changes from oil, peat, and coal to heat pumps pre-2030, while it maybe economical beneficial over the lifetime of a heat pump, there maybe financial or social barriers associated with investing in a heat pump and the hurdle rates in the residential sector need more focus to represent consumer choice.
+The pre-2030 results in TIM show large changes in end use from oil, peat, and coal to heat pumps.
+While it maybe economical beneficial over the lifetime of a heat pump, there maybe financial or social barriers associated with investing in a heat pump and the hurdle rates in the residential sector need more focus to represent consumer choice.
 As previously mentioned, residential oil tanks were filled with low oil prices in 2020, because each consumer is responsible for their own investments and similarly consumers will unlikely invest in new low carbon heating technologies when their existing heating technologies are not at the end of life.
@@
 \par Primarily due to the lack of reliable data surrounding heat pump performance in B3 or lower rated dwellings, heat pumps can only be installed in B2 or above rated dwellings, similarly heat pump grants supplied by SEAI are for B2 rated or above dwellings.
-In future iterations of TIM, it could be helpful to account for heat pumps in C or D rated dwellings, while the efficiency will be less, it could still provide part of the optimal pathway.
+In future iterations of TIM, it would be beneficial to account for heat pumps in C or D rated dwellings, while the efficiency will be less, it could still provide part of the optimal pathway.
 
@@
 Despite the filtration process in the BER database, limitations of the data exist, such as PV capacity, internal temperature assumption, default values, and heat loss indicator value ( which is a more accurate way of predicting a heat pump performance ).
+The model does not account for any new renewable gas within natural gas, which may lower the carbon intensity of gas and improve the viability of gas in decarbonisation pathways.
 
@@
 \label{Conclusion}
-The discussion section outlines the improvements which will be made to the residential sector in TIM and some of the strengths and weaknesses.
-To conclude, the residential sector in TIM will be able to provide valuable insight into future research questions like ``How do changes to future technology developments affect the optimal residential heating decarbonisation pathway?"
+The transparency and soft-linking capabilities will allow the residential sector in TIM will be able to provide valuable insight into future decarbonisation research questions.
+The residential sector has been developed with current climate policies such as retrofitting and heat pumps providing the cornerstone.
+Higher granularity in the model will be explored, while the current version of the model will provide insights  into how  changes to future technology developments affect the optimal residential heating decarbonisation pathway.
 
@@
 \label{sec:Acknowledgements}
-The authors wish to acknowledge the contributions of Tomás Mac Uidhir, who supported filtering of the BER database, and Ankita Singh Gaur who developed the residential demands. 
+
 CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC.
-Alessandro and Maurizio provided technical assistance on the development of the residential sector in TIMES. 
+Alessandro Chiodiand and Maurizio Gargiulo from E4sma provided technical assistance on the development of the residential sector in TIM. 
 % https://www.marei.ie/project/capacity/
@@
 
-\section{}
+\section{Additional input data and assumption}
 \label{sec:appendix}
@@
 %% appendix sections are then done as normal sections
-\subsection{}
-\label{sec:appendixA}
-Residential Fuel data is sourced from SEAI Energy Balance and BER Database, the right column shows how the fuels are structured in TIM. 
+
+%Residential Fuel data is sourced from SEAI Energy Balance and BER Database, the right column shows how the fuels are structured in TIM. 
 \begin{table}[htbp]
@@
 \end{tabular}
-\label{Table:Fuels}
+\label{AppTable:Fuels}
 \end{table}
 
-\newpage
-\subsection{}
-\label{sec:appendixA1}
-
-BER Database Filters
+%BER Database Filters
 \begin{table}[htbp]
@@
  \end{tabular}%
- \label{BERFilters}%
+ \label{AppBERFilters}%
 \end{table}%
 
-\subsection{}
-\label{sec:appendixA2}
-Age to BER Rating Correlation Index
+\begin{table}[htbp]
+ \centering
+ \footnotesize
+ \caption{Energy Savings per BER rating} 
+ \begin{tabular}{|cl|llllllll}
+\hline   
+&  \multicolumn{8}{c}{Pre-Retrofit Rating}     \\ 
+ & & G  & F & E & D  & C & B3  & B2  & B1 \\ \hline 
+\multirow{8}{*}{\rotatebox[origin=c]{90}{Post-Retrofit Rating}} & A       & \cellcolor{red}83\% & \cellcolor{orange}73\% & \cellcolor{orange}67\% & \cellcolor{orange}57\% & \cellcolor{orange}38\%  & \cellcolor{orange}46\% & \cellcolor{yellow}35\% & \cellcolor{yellow}17\% \\
+ & B1 & \cellcolor{red}79\% & \cellcolor{orange}68\%  & \cellcolor{orange}61\% & \cellcolor{orange}48\% & \cellcolor{yellow}25\%  & \cellcolor{yellow}35\% & \cellcolor{yellow}22\% &      \\
+  & B2      & \cellcolor{orange}73\% & \cellcolor{orange}59\% & \cellcolor{orange}49\% & \cellcolor{yellow}33\% & \cellcolor{yellow}4\%   & \cellcolor{yellow}17\% &      &      \\
+  & B3      & \cellcolor{orange}68\% & \cellcolor{orange}50\% & \cellcolor{orange}39\% & \cellcolor{yellow}19\% & \cellcolor{yellow}0\% &      &      &      \\
+  & C       & \cellcolor{orange}72\% & \cellcolor{orange}57\% & \cellcolor{orange}47\% & \cellcolor{yellow}30\% &       &      &      &      \\
+ & D       & \cellcolor{orange}60\% & \cellcolor{orange}\cellcolor{orange}38\% & \cellcolor{yellow}25\% &      &       &      &      &      \\
+ & E       & \cellcolor{orange}47\% & \cellcolor{yellow}18\% &      &      &       &      &      &      \\
+  & F       & \cellcolor{yellow}35\% &      &      &      &       &      &      &
+
+ \end{tabular}%
+ \label{AppDepthofRetrofit}%
+\end{table}
+
+\newpage
+%Age to BER Rating Correlation Index
 \begin{sidewaystable}
@@
  \caption{Apartment Age to BER Rating Correlation}
+ \label{AppBERCor}
 \begin{tabular}{l|lllllllll}
@@
  \caption{Attached Age to BER Rating Correlation}
+ \label{AttBERCor}
 \begin{tabular}{l|lllllllll}
@@
  \caption{Detached Age to BER Rating Correlation}
+ \label{DetBERCor}
 \begin{tabular}{l|lllllllll}
@@
 
-\newpage
-
-\subsection{}
-\label{sec:appendixA3}
-Depth of retrofit defined by expected energy savings. The red cells are unfeasible, orange cells show deep retrofits, and yellow represent shallow retrofits. 
-%The model has two types of retrofitting options per archetype,  a shallow retrofit is when the primary energy ($kWh/m^2/yr$) is reduced by $ \leq 35 \%$, and deep retrofit is where the primary energy is reduced by $>35 \%$, a reduction of $ \> 75 \%$ is considered infeasible.
-The different between the pre-retrofit and post-retrofit primary energy determines the type of retrofit, as shown i
-
 \begin{table}[htbp]
   \centering
- \footnotesize
- \caption{Energy Savings per BER rating}
- \begin{tabular}{|cl|llllllll}
-\hline   
-&  \multicolumn{8}{c}{Pre-Retrofit Rating}     \\ 
- & & G  & F & E & D  & C & B3  & B2  & B1 \\ \hline 
-\multirow{8}{*}{\rotatebox[origin=c]{90}{Post-Retrofit Rating}} & A       & \cellcolor{red}83\% & \cellcolor{orange}73\% & \cellcolor{orange}67\% & \cellcolor{orange}57\% & \cellcolor{orange}38\%  & \cellcolor{orange}46\% & \cellcolor{yellow}35\% & \cellcolor{yellow}17\% \\
- & B1 & \cellcolor{red}79\% & \cellcolor{orange}68\%  & \cellcolor{orange}61\% & \cellcolor{orange}48\% & \cellcolor{yellow}25\%  & \cellcolor{yellow}35\% & \cellcolor{yellow}22\% &      \\
-  & B2      & \cellcolor{orange}73\% & \cellcolor{orange}59\% & \cellcolor{orange}49\% & \cellcolor{yellow}33\% & \cellcolor{yellow}4\%   & \cellcolor{yellow}17\% &      &      \\
-  & B3      & \cellcolor{orange}68\% & \cellcolor{orange}50\% & \cellcolor{orange}39\% & \cellcolor{yellow}19\% & \cellcolor{yellow}0\% &      &      &      \\
-  & C       & \cellcolor{orange}72\% & \cellcolor{orange}57\% & \cellcolor{orange}47\% & \cellcolor{yellow}30\% &       &      &      &      \\
- & D       & \cellcolor{orange}60\% & \cellcolor{orange}\cellcolor{orange}38\% & \cellcolor{yellow}25\% &      &       &      &      &      \\
- & E       & \cellcolor{orange}47\% & \cellcolor{yellow}18\% &      &      &       &      &      &      \\
-  & F       & \cellcolor{yellow}35\% &      &      &      &       &      &      &
-
- \end{tabular}%
- \label{DepthofRetrofit}%
-\end{table}
-
-\subsection{}
-\label{sec:appendixA4}
-List of water and space heating technologies available per archetype. The ``???" denotes either APT for Apartments, ATT for Attached, or DET for detached dwellings.
-
-\begin{table}[htbp]
-  \centering
-  \caption{Space and Water Heating Technologies}
+  \caption{Space and Water Heating Technologies, where ``???" denotes either APT for Apartments, ATT for Attached, or DET for detached dwellings.}
   \resizebox{\textwidth}{!}{\begin{tabular}{cc|cc|cc|cc}
@@
  }
- \label{HeatingTech}%
+ \label{AppHeatingTech}%
 \end{table}
-\newpage
+
+\clearpage
+
 \bibliography{TIMref.bib}

### 2021-08-26 manuscript.tex
26 Aug

@@
 
-\title{Developing Decarbonisation Pathways for Irish homes in changing TIMES}
+\title{Developing decarbonisation pathways for Irish homes in changing TIMES}
 
@@
 These factors have served as barriers to the decarbonisation of the residential sector in Ireland to date.
-\par
-Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges, which provides insights to decision-makers on issues related to climate and energy policy.
-\par
 The TIMES-Ireland Model (TIM) is a newly developed optimisation model for the Irish energy system.
-This paper describes the methodology applied to the development of the residential sector in TIM and explores the results, which provides new insights on optimal residential decarbonisation pathways.
+This paper describes the innovative methodology applied to the development of the residential sector in TIM and explores the results, which provides new insights on optimal residential decarbonisation pathways.
 
@@
 
-The Introduction section of this paper reviews bottom-up residential models, which have high technological granularity accounting for the residential sector only and ESOMs which have a bottom-up or hybrid approach and includes all energy-related sectors.
+The Introduction section of this paper reviews Energy System Optimisation Models (ESOMs) which use a bottom-up or hybrid approach and include all energy-related sectors, and bottom-up residential sector models which have have a detailed level of technological representation.
 The hybrid approach combines bottom-up technological detail and top-down macroeconomic detail.
@@
 
-Economically optimal or least cost residential decarbonisation pathways are dynamic, with technological advancements in heating appliances being assisted by exponentially increasing computational capacity over time, combined with new low-carbon fuel alternatives and new heating technologies becoming available.
+Economically optimal or least cost residential decarbonisation pathways are dynamic, advancements in heating technologies, new low-carbon fuel alternatives and new heating technologies continuously alter the optimal pathway.
 ESOMs have a rich representation of existing and future technological variables such as efficiency, cost, and fuel, while also accounting for macroeconomic variables such as economic growth and the social discount rate.
-ESOMs supply enough energy to at least satisfy energy service demands so that the lights stay on, while also considering sectoral interconnections to explore future decarbonisation pathways.
+ESOMs supply enough energy to at least satisfy each energy service, while also considering sectoral interconnections to explore optimal decarbonisation pathways.
 There are a number of widely used ESOMs, such as MESSAGE \cite{Messner1995Model-basedPlanning}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.
 
-These models can be used to explore the residential sector in higher granularity.
-Different levels of disaggregation are applied to the residential sector in models and research by \citet{Natarajana2011ModellingApproach}  highlighted a common approach used by defining an average performance for a number of dwelling categories, and this approach was adopted here.
-To account for the heterogeneous mix of dwellings, each dwelling is categorised based on building type to better reflect the housing stock and provide building specfic insights.
+These models can be used to explore the residential sector in high spatial, temporal and technological granularity.
+Different levels of disaggregation are applied to the residential sector in models, \citet{Natarajana2011ModellingApproach}  highlighted a common approach used by defining an average performance for a number of dwelling categories.
+To account for the heterogeneous mix of dwellings, each dwelling is categorised based on building type to better reflect the housing stock and provide building specific insights.
 Some models also disaggregate the year of construction \cite{Ahern2013StateStock}, wall type \cite{Dineen2015ImprovedSource} and level of insulation \cite{Clinch2001Cost-benefitEfficiency}.
@@
 TIMES assumes perfect foresight to enable a least-cost solution to be calculated, but \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}.
-In light of this,  \citet{Li2018IncorporatingModel} investigated the heating technology preference in the UK using UK TIMES Model (UKTM) by incorporating a Discrete Choice Model (DCM) to better reflect how homeowners choose heating technologies and thus correcting for error in the perfect foresight assumption.
+In light of this,  \citet{Li2018IncorporatingModel} investigated the heating technology preference in the UK using UK TIMES Model (UKTM) by incorporating a Discrete Choice Model (DCM) to better reflect how homeowners choose heating technologies and thus addressing for simplification of the perfect foresight assumption.
 
@@
 Another trend observed is the electrification of energy-related sectors \cite{Chiodi2013ModellingSystem}, while a focus on photovoltaic (PV) systems can be examined in parallel \cite{Marczinkowski2018ResidentialSystems}.
-Retrofitting and electrification are a fundamental part of residential decarbonisation in Ireland, and therefore play a central role in the development of the model described here.
-Furthermore, the new model includes technologies, which were not previously considered, one such technology is District Heating (DH), some initial studies have found it may be optimal to have 37\% of heat demand to be provided by DH in Ireland \cite{Thellufsen2019ImplementingIreland}, the full list of space and water heating technologies considered in the model can be found in Table \ref{AppHeatingTech}.
 The strength of an ESOM can be used to examine the sectoral interactions, one study by Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} examined the interaction between heat pumps and wind generation investments in Denmark, the results find it is optimal to install air-to-water heat pumps in non-DH areas.
@@
 The models were linked to explore specific energy efficiency policies in the residential and other sectors.
-Three scenarios were explored, namely, the reference scenario, energy efficiency scenario, and energy efficiency+ scenario.
+Three scenarios were explored, namely, the reference scenario, energy efficiency scenario, and energy efficiency+ scenario, while three residential policies were analysed retrofitting, building regulation and lighting.
+At the time of this study little to no data on retrofits existed and the Building Energy Rating (BER) public database was much smaller.
+\citet{Rogan2014LEAPsSystem} research focuses on the theoretical energy savings from retrofits, \textit{with both the
+number of dwellings and the scale of improvement
+works specified exogenously from the model} \cite{Rogan2014LEAPsSystem}, the model limited retrofits to BER bands D to G while not accounting for internal temperature shifts.
 The reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020, and the results show that residential total energy consumption rose 6.7\% in 2020, compared to 2008.
-In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and low energy lighting, resulting in a 9.9\% reduction of energy consumption in 2020, compared to the reference scenario.
-The energy efficiency+ scenario also allows for 800,000 residential dwellings to be retrofitting, but accounts for deeper retrofits.
-This results in an energy consumption reduction of 16.5\% in 2020, compared to the reference scenario.
-\citet{Rogan2014LEAPsSystem} discussed the barriers, obstacles to investment, split incentives, and information gaps which restrict energy efficiency improvements.
-The insights highlighted the importance for the policy to not only look at the number of retrofits, but also the depth of retrofit, as shown by the different results in the scenarios.
+In the energy efficiency scenario, 800,000 residential dwellings are retrofitted, resulting in a 9.9\% reduction of energy consumption compared to the reference scenario.
+The energy efficiency+ scenario allows for the same quantity of retrofits, but accounts for deeper retrofits resulting in an energy consumption reduction of 16.5\%, compared to the reference scenario.
+The insights highlight the importance of the depth of retrofit, as shown by the results.
 The model also found considerable scope for energy savings to be made through improved building regulations.
 
-In 2018, \citet{Thellufsen2019ImplementingIreland} investigated the potential for the implementation of DH in both the commercial and residential sectors of Ireland.
+In 2018, \citet{Thellufsen2019ImplementingIreland} investigated the potential for the implementation of District Heating (DH) in both the commercial and residential sectors of Ireland.
 The paper analyses the $CO_2-80$ scenario ( 80\% $CO_2$ reduction in 2050, compared to 1990 ) with and without DH using the previous Irish TIMES model and EnergyPLAN.
@@
 
-\par  The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector models.
-These models have provided insights on the input data, methodology, and results, and some studies are outlined below.
-\par In the first comprehensive analysis of the Irish residential sector, Clinch and Healy examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure}, occupancy behaviour \cite{Clinch2001ModellingEfficiency}, social benefit \& cost \cite{Clinch2001Cost-benefitEfficiency}  and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}.
+\par  The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector models, which are not optimisation models.
+These models have provided insights on the input data, methodology, and results.
+In the first comprehensive analysis of the Irish residential sector, Clinch and Healy examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure}, occupancy behaviour \cite{Clinch2001ModellingEfficiency}, social benefit \& cost \cite{Clinch2001Cost-benefitEfficiency}  and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}.
 
@@
 %The paper acknowledges that retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperature effect on energy efficiency improvements.
-Later in 2015, \citet{Dineen2015ImprovedSource} built on previous work in examining new dwellings, by examining the energy savings potential of energy efficiency improvement measures of the existing dwelling stock.
-The Dwelling Energy Assessment Procedure (DEAP) model was used, this is the national tool for Building Energy Rating (BER) calculation.
-The BER filters from this study were updated and applied in the model developed here.
+Later in 2015, \citet{Dineen2015ImprovedSource} built on previous work in examining new dwellings, by examining the energy savings potential of energy efficiency improvement measures in the existing dwelling stock.
+The Dwelling Energy Assessment Procedure (DEAP) model was used, this is the national tool for the BER calculation.
+A complete set of filters was applied to the BER database before use in the model, erroneous and outlining data was removed.
 \citet{Dineen2015ImprovedSource} performed and sensitivity analysis and found that there was a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift.
@@
 The model results show that thermal improvement measures resulted in a reduction of running costs and $CO_2$ emissions by an average of 65\% for houses constructed prior to 1979, compared to houses built in 1979 or after resulting in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures.
-This study highlights the low hanging fruit of pre-1979 oil centrally heated, rural housing stock.
-\par The Archetype Dwelling Energy Model - SQL (ArDEM-SQL) was developed by \citet{Uidhir2020ImprovingDecisions} ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and calculates the annual energy consumption for space/ water heating, and lighting using EN 13790:2004 to simulate the potential for improved energy efficiency gains within an existing retrofit program.
+This study highlights the potential energy savings within pre-1979 oil centrally heated, rural housing stock.
+\par The Archetype Dwelling Energy Model - SQL (ArDEM-SQL) developed by Mac \citet{Uidhir2020ImprovingDecisions} converts the Irish BER public database into Microsoft SQL language to calculate the annual energy consumption for space/ water heating, and lighting using EN 13790:2004 to simulate the potential for improved energy efficiency gains within an existing retrofit program.
 Mac \citet{Uidhir2020ImprovingDecisions} examined the Better Energy Homes (BEH) retrofit grant scheme.
@@
 \par Finally, the Sustainable Energy Authority of Ireland's (SEAI’s) National Energy Modelling Framework (NEMF) is a suite of modelling tools used for modelling the Irish energy system.
-The SEAI is Ireland's national energy authority and they play a key role in bridging the gap between grant/policy and consumer, by assisting consumers to reduce energy.
+The SEAI is Ireland's national energy authority and they \textit{provide evidence-based advice across the government, state and semi-state bodies, in addition to delivering a range of grant and support programmes to promote sustainable energy} .
 The models listed are some of the models used within SEAI's NEMF to model the impact of incentives and opportunities created by specific policy instruments.
@@
 The models described in this section include a range of national ESOMs and bottom-up residential models with varied objectives.
-The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switched to Microsoft SQL, with much higher computational power.
-The growing BER database is providing high data granularity input to Irish models.
+The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switched to Microsoft SQL, the higher computational power allows for deeper analyses.
+The growing BER database is provides detailed open sourced data for input to Irish models.
 %The internal temperature is becoming more frequently analysed in Irish residential studies, a recurring theme is the wide range of internal temperatures which Dineen et al.
 \cite{Dineen2015ImprovedSource} found this assumption significantly changes the final residential energy consumption.
-
 \subsection{Case study for Ireland}
 \label{Intro:IreRSD}
+
+\subsubsection{Residential sector characteristics}
+\label{Intro:GHG}
+
+Under 2006 Intergovernmental Panel on Climate Change (IPCC) guidelines, Ireland’s residential sector accounted for 14.2\% of GHGs emissions in 2018 \cite{Howley2020Energy-RelatedIreland.}.
+Focusing only on energy-related GHG emissions from heat, electricity and transport, the residential sector directly accounts for 24\% of GHG emissions.
+The residential sector is directly responsible for 47\% of heat and 30\% of electricity and indirectly responsible for a share of the remaining GHGs, especially in the transport sector, as the rural settlement patterns increase transport demands.
 
@@
 
-\subsubsection{Residential sector characteristics}
-\label{Intro:GHG}
-
-Under 2006 Intergovernmental Panel on Climate Change (IPCC) guidelines, Ireland’s residential sector accounted for 14.2\% of GHGs emissions in 2018 \cite{Howley2020Energy-RelatedIreland.}.
-Focusing only on energy-related GHG emissions from heat, electricity and transport, the residential sector directly accounts for 24\% of GHG emissions.
-The residential sector is directly responsible for 47\% of heat and 30\% of electricity and indirectly responsible for a share of the remaining GHGs, especially in the transport sector, as the rural settlement patterns increase transport demands.
+\subsubsection{Policy context}
+\label{Intro:Policy}
 
@@
 
-\subsubsection{Policy context}
-\label{Intro:Policy}
 The Irish government's long-term overarching strategy aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}\cite{Ireland2018ProjectFramework}.
@@
 
-The distinctive Irish residential sector provides an interesting case study and new Irish climate targets now means previous Irish models which relied on superseded climate targets are out-dated but not redundant, new up to date models are needed to explore the decarbonisation pathways and technical limits.
+The distinctive Irish residential sector provides an interesting case study and new Irish climate targets now means previous Irish models which relied on superseded climate targets are out-dated, but not redundant.
+A new iteratively built transparent ESOM with wide stakeholder involvement, which is reflective of current climate policy is needed to robustly explore optimal decarbonisation pathways and technical limits.
+The new model will provide opportunity for deeper analysis of the energy system sectors, with accessible soft-linking capabilities.
+The model development process can be applied to any national or regional ESOM.
 No such ESOM with a focus on Irish residential sector exists, the strength of an ESOM is to account for sectoral interactions which bottom-up simulation models cannot provide.
 Structural improvements and archetypes in the model aligns with existing retrofitting and electrification strategies.
-Other model improvements include filtering input data and calculating new energy service demands, while also considering internal temperatures.
+Other model innovations include filtering the BER database, other EU countries with publicly available BER database under directive (EU) 2018/844 can apply similar methods, while TIM also considers internal temperatures shifts and new energy service demands were calculated.
 TIM will be a critical tool in exploring
@@
 
-The layout of the rest of this paper is as follows; section \ref{LitR} is the core section of this paper and explains the methodology used to develop the residential sector.
+The layout of the rest of this paper is as follows; section \ref{Meth} is the core section of this paper and explains the methodology used to develop the residential sector.
 The treatment of input data in the model is described, the calculations are explained, and assumptions and constraints are highlighted.
@@
 \section{Methodology}
-\label{LitR}
+\label{Meth}
 
@@
 
-\subsection{The TIMES-Ireland Model (TIM)}
-\label{Overview}
+\subsection{TIMES}
+\label{MethTIMES}
 
@@
 
-\subsection{Residential}
-\label{LitR:RES}
-The residential sector was developed in parallel with other sectors in TIM, and concurrently with the new Irish LEAP model, which allows for more accessible soft-linking to provide robust analytical insights by combining simulation and optimisation, and to exchange expertise between modelling teams.
-Figure \ref{fig:RES} illustrates the flow of energy from fuel to demand in the TIM residential sector.
+\subsection{TIMES-Ireland Model (TIM)}
+\label{Meth:TIM}
+
+TIM was developed completely online during national lockdowns, with no face to face meetings, the model relied on git-centred model development process with different developers leading different sectors.
+Weekly meetings were combined with intensive daily ``sprint" meetings to keep model milestones on track.
+The transparency of TIM improved the iterative development process, external energy experts and other stakeholders contributed using web-based dashboards \footnote{https://tim-review1.netlify.app/results/}.
+The schematic in Figure \ref{fig:TIMOverviewRES} shows the main sectors of the model, represented in the demand-side modules and how they link with the supply-side modules, to create a full picture of the national energy system \textbf{SOURCE}.
+
+\begin{figure}[!htbp]
+ \centering
+ \includegraphics[width=\textwidth]{Figures/TIMOverviewRES.jpg}
+ \caption{Structure of TIM }
+ \label{fig:TIMOverviewRES}
+\end{figure}
+
+\subsection{TIM: Residential Sector}
+\label{Meth:TIMRes}
+
+The residential sector was developed with awareness of the ongoing national retrofitting and electrification strategies and investments.
+The residential energy services were developed based upon regulation (EU) 431/2014, the BER database and EU databases.
+Figure \ref{fig:RES} illustrates the flow of energy from fuel to demand in the residential sector.
 
@@
 Table \ref{AppTable:Fuels} shows an overview of the fuels combined from the different data sources.
+The full list of space and water heating technologies considered in the model can be found in Table \ref{AppHeatingTech}.
 
@@
 The six non-archetype energy services are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service.
-Non-Archetype data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{LitR:Lighting,Appliances}.
-The four archetype energy services are dependent on  building type and data is sourced from the BER database, which is discussed in Section \ref{LitR:StockData}.
-The internal temperature adjustment for space heating is also detailed in Section \ref{LitR:StockData}.
-The demand for archetype and non-archetype energy services is based on the growth of residential dwellings, which is detailed in section \ref{LitR:Demands}.
-
-\subsubsection{Dwelling Stock Data}
-\label{LitR:StockData}
-
-The dwelling stock data is obtained from the SEAI BER Public database, which is firstly filtered, then scaled up using Central Statistics Office (CSO) data.
+Non-Archetype data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{Meth:Data}.
+The four archetype energy services are dependent on  building type and data is sourced from the BER database, which is discussed in Section \ref{Meth:Data}.
+The internal temperature adjustment for space heating is also detailed in Section \ref{Meth:Data}.
+The demand for archetype and non-archetype energy services is based on the growth of residential dwellings, which is detailed in section \ref{Meth:Data}.
+
+\subsubsection{Data}
+\label{Meth:Data}
+%
+The dwelling stock data is obtained from the SEAI BER Public database \footnote{https://ndber.seai.ie/BERResearchTool/ber/search.aspx}, which is firstly filtered, then scaled up using Central Statistics Office (CSO) data.
 The SEAI BER public database is Ireland's national Energy Performance Certificate (EPC) register,  which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD).
@@
 Space heating consumption is reduced by 26\% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance.
-DEAP assumes that all living areas ( living room, sitting room, and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C.
+DEAP assumes that living areas ( living room, sitting room, and possibly kitchen) in all dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C.
 Through the analysis of previous residential modelling in Ireland this is not the case, as the internal temperatures are often much lower, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions applied in TIM.
@@
 The ArDEM-SQL, which was described in section \ref{Intro:ESOMIreland} is a simulation model built on space heating calculation EN 13790.
-ArDEM-SQL alters the primary energy consumption of the BER database by changing the internal temperatures assumed by DEAP in table \ref{Table:InternalTemps}.
+ArDEM-SQL alters the primary energy consumption of the BER database by changing the internal temperatures assumed by DEAP in Table \ref{Table:InternalTemps}.
 
@@
 Once the final dwelling stock is complete, the next step is to assign a BER rating to all dwellings.
-This was done by comparing 9 different dwelling age ranges for each of the 3 archetypes between the BER and CSO datasets.
+This was done by comparing nine different dwelling age ranges for each of the three archetypes between the BER and CSO datasets.
 
@@
 
-\subsubsection{Costs}
-\label{LitR:Costs}
-
-This section only covers the costs applied in the residential sector, which includes space and water heating technologies, retrofitting, cooking, lighting, pumps, fans, space cooling, and other electrical appliances.
+The costs applied in the residential sector, which includes space and water heating technologies, retrofitting, cooking, lighting, pumps, fans, space cooling, and other electrical appliances. 
 The fuel costs are not applied in the residential sector, there are applied in the supply sector, and the data is sourced from World Energy Outlook reports from 2019 \cite{/content/publication/caf32f3b-en} and 2020 \cite{/content/publication/557a761b-en}.
@@
 
-\subsubsection{Savings}
-\label{LitR:Savings}
-
-The previous section outlined the cost of retrofitting, and this section looks at the expected savings of retrofitting existing dwellings and building regulations in new dwellings.  
+The previous paragraphs outlined the cost of retrofitting, and this following looks at the expected savings of retrofitting existing dwellings and building regulations in new dwellings.  
 The annual heating costs per archetype and per BER rating are sourced from SEAI's indicative energy running costs.
@@
 
-\subsubsection{Energy Service Demand Projections}
-\label{LitR:Demands}
-
 The  technical  bottom-up  development  approach  of  the  residential  sector  is  combined with other economic sectors to meet macroeconomic defined demands.
@@
 
-\subsubsection{Space and Water Heating}
-\label{LitR:Heating}
-
 The demand for space and water heating is outlined in the previous section, and this section explains how water and space heating are achieved through technology and retrofits.
@@
 The base year heating energy consumption is derived from the BER database, but uses census data \cite{CentralStatisticsOffice2016CensusIreland} from the CSO dataset E1008 to calibrate the fuel used in heating systems per archetype.
-
-\subsubsection{Lighting, Appliances and Other}
-\label{LitR:Lighting,Appliances}
 
@@
 
-\subsubsection{User Constraints}
-\label{LitR:Constraints}
-The TIM model operates on the basis of perfect foresight, but
-\textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel} for this reason ``User Constraints" are applied in the residential sector to better reflect reality.
-The User Constraints applied to the residential sector include:
+``User Constraints" are defined as any combination of constraints including fixed, maximum or minimum bounds on growth, investment or shares on technologies or fuels.
+The ``User Constraints" applied to better align with the residential sector include:
 \begin{enumerate}[1.]
- \item Only one space and water heating system per household ( stoves excluded ) 
- \item Maximum 60\% of existing buildings to be retrofitted by 2030, and 95\% by 2070. Maximum interpolated between 2030 and 2070. 
+ \item Only one space and water heating system per household ( stoves excluded ), this user constraint is applied to represent individual heating systems and calibrate results.
+ \item Maximum 60\% of existing buildings to be retrofitted by 2030, and 95\% by 2070, the maximum value is interpolated between 2030 and 2070.
+This is user constraint is applied to align with maximum available labour and was applied during the model development to calibrate results.
  \item A prerequisite requirement for electrical heat pump installation is applied to each archetype.
-The percentage of dwellings which requires a shallow or deep retrofit before a heat pump can be installed (  B2 rating for heat pump ready dwelling ).
+Table \ref{tab:HeatPumpConstraint} shows the percentage of dwellings which requires a shallow or deep retrofit before a heat pump can be installed (  A heat pump ready dwelling must be at least a B2 rating).
 Hybrid heat pumps are not included in this constraint, as they can supply higher water temperatures without significantly reducing performance.
+This user constraint was applied because of the lack of data regrading electrical heat pumps in less than B2 rated dwellings, without this constraint heat pumps are installed in poorly rated dwellings, this constraint aligns with known heat pump performance data.
  \begin{table}[htbp]
@@
  \item For space and water heating, each archetype has an upper gas growth constraint of 40\% higher than the base year value.
-\item Maximum fuel share in cooking for gas and LPG is 40\% and 10\% respectively. Base year fuel share for Gas is 32\% and LPG is 1.5\%.
-\item Maximum 10\% solar input to dual fuel heating technology
-\item Maximum 30\% gas input to gas hybrid heating pumps
+This user constraint was applied in advance of the model development to align with gas infrastructure development.
+\item Maximum fuel share in cooking for gas and LPG is 40\% and 10\% respectively.
+Base year fuel share for Gas is 32\% and LPG is 1.5\%.
+This user constraint was applied in advance of the model development to align with gas infrastructure development.
+\item Maximum 10\% solar input to dual fuel heating technology. This constraint was applied based on technological constraints.
+\item Maximum 30\% gas input to gas hybrid heating pumps. This constraint was applied based on technological constraints.
 \end{enumerate}
@@
 Some consistencies across the scenarios show the large growth of renewables in the power generation sector and consequently the uptake of electricity across the remaining sectors, especially in the Current Effort scenario.
-The Previous Effort scenario opts for natural gas, backed by low carbon intensity electricity, the increase in ambition and tightening of constraints requires natural gas to decrease, with electricity becoming the cornerstone of energy-system decarbonisation.
+The Previous Effort scenario opts for natural gas, backed by low carbon intensity electricity, the increase in ambition and tightening of constraints requires natural gas to decrease, with electricity becoming the cornerstone of energy system decarbonisation.
 
@@
 
-The previous version of TIM, namely, the Ireland TIMES model, was used for almost 10 years and was continuously tweaked to react to the ever-changing climate policy, climate targets, and technological advancements.
+The predecessor of TIM, namely, the Ireland TIMES model, was used for almost 10 years and was continuously tweaked to react to the ever-changing climate policy, climate targets, and technological advancements.
 \par TIM is developed to provide insights on sectoral carbon budgets and 2030 and 2050 climate targets.
@@
 
+%in parallel with other sectors in TIM, and concurrently with the new Irish LEAP model, which allows for more accessible soft-linking to provide robust analytical insights by combining simulation and optimisation, and to exchange expertise between modelling teams.
 The methodology applied to the development of the residential sector in TIM is transparent, up-to-date, and robust.

### 2021-08-27 manuscript.tex
Done for 27 Aug 2021

@@
 \cite{Dineen2015ImprovedSource} found this assumption significantly changes the final residential energy consumption.
+
 \subsection{Case study for Ireland}
@@
 
-Space and water heating are the main energy services, accounting for 80\% of the energy used in residential dwellings \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and typically have a lifetime of at least 15 years, consequently leading to a slow turnover of technology in the residential sector.
-Apart from space and water heating, all EU member states are required under commission regulation (EU) 431/2014 to supply energy consumption data on space cooling, cooking, lighting \& appliances, and others (mainly outdoor appliances), the annual trend of these energy services is shown in Figure \ref{fig:IrelandServiceDemands}.
-These energy services were also used in TIM, however there is no space cooling data available in Ireland and lighting \& appliances were separated while others was combined with appliances.
+Space and water heating are the main energy services, accounting for 80\% of the energy used in residential dwellings \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and typical heating systems have a lifetime of at least 15 years, consequently leading to a slow turnover of technology in the residential sector.
+Apart from space and water heating, all EU member states are required under commission regulation (EU) 431/2014 to supply energy consumption data on space cooling, cooking, lighting \& appliances, and others (mainly outdoor appliances), the annual trend of these energy services is shown in Fig.
+\ref{fig:IrelandServiceDemands}.
 
@@
 The densification of Ireland's main cities will provide opportunities to improve the efficiency of heating, electricity, and transport within and between cities.
-\par The Marginal Abatement Cost Curve (MACC) produced in CAP2019, shows electricity \& transport are the most cost-effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
+The Marginal Abatement Cost Curve (MACC) produced in CAP2019, shows electricity \& transport are the most cost-effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}.
 In Ireland, the decarbonisation of electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2020Energy-RelatedIreland.}.
@@
 \par As the grant provider, SEAI plays a key role in the rollout of retrofitting and heat pumps.
-The retrofitting and heat pump grants do still require large upfront costs from the consumer, which may hinder the ambitious targets as this wont be a viable option for many households, green loans are currently available to bridge this gap.
+The retrofitting and heat pump grants do still require large upfront costs from the consumer, which may hinder the ambitious targets as this wont be a viable option for many households, green loans are currently available to bridge this gap and other options are being explored.
 % are therefore the preferred heating technology despite the high capital costs.
 CAP2019 planned to retrofit 500,000 existing homes to at least a  ‘B2’ energy rating by 2030, which highlights the energy efficiency first approach in the residential sector.
-\par Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019VisionIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
+Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019VisionIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
 %GNI anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020EnergySEAI}.
 A major barrier to increasing biomethane from an agricultural source in the gas network is the cost, it was the most expensive option in the MACC produced in CAP2019.
-\par Ireland has one of the lowest shares of DH in Europe at less than 1\%  \cite{Gartland2016AIreland}. %
+Ireland has one of the lowest shares of DH in Europe at less than 1\%  \cite{Gartland2016AIreland}. %
 %Advancement in DH control and awareness of the technology will strengthen the case for DH networks to play a part in decarbonising the residential sector.
-CAP2019 plans to have approx.
-0.4\% of heating provided by DH in 2030.
+CAP2019 plans to have 0.4\% of heating provided by DH in 2030.
 The Irish District Energy Association (IrDEA) suggest 57\% of heat can be sourced from DH, if supporting policy and regulation was in place.
-CAP2019 sets out a need to develop a policy framework for the development of DH in Ireland
+CAP2019 sets out a need to develop a policy framework for the development of DH in Ireland.
 %DH will likely play a role, however, Gartland points out a range of organisational, technical, regulatory, and economic barriers to DH growth in Ireland \cite{Gartland2016AIreland}.
@@
 The distinctive Irish residential sector provides an interesting case study and new Irish climate targets now means previous Irish models which relied on superseded climate targets are out-dated, but not redundant.
-A new iteratively built transparent ESOM with wide stakeholder involvement, which is reflective of current climate policy is needed to robustly explore optimal decarbonisation pathways and technical limits.
+A new iteratively built transparent ESOM with wide stakeholder involvement, which aligns with current climate policy is needed to robustly explore optimal decarbonisation pathways and technical limits.
 The new model will provide opportunity for deeper analysis of the energy system sectors, with accessible soft-linking capabilities.
@@
 Other model innovations include filtering the BER database, other EU countries with publicly available BER database under directive (EU) 2018/844 can apply similar methods, while TIM also considers internal temperatures shifts and new energy service demands were calculated.
-TIM will be a critical tool in exploring
-\textit{how changes to future technology developments affect the optimal residential heating decarbonisation pathway in Ireland? }
+TIM can provide a robust methodology into \textit{``how Ireland should set GHG targets in the residential sector”} and TIM will be a critical tool in exploring \textit{how changes to future technology developments affect the optimal residential heating decarbonisation pathway in Ireland?
+}.
 
@@
 
-%``How should Ireland set GHG targets in the residential sector?”
-
 \subsection{TIMES}
@@
 \par  The expertise among energy policy researchers in Ireland combined with the global network of Energy Technology Systems Analysis Program (ETSAP) modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways.
-Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model and then updated with specific national data and assumptions.
-The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.
-The introduction of new climate policies and targets combined with technological advancements, means there is a need for a new ESOM to be built to explore Ireland's new climate targets and sectoral five-year carbon budgets, the new ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
+Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model and then updated with specific national data and assumptions, to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.
+The previous Irish TIMES model was non-transparent and inflexible model, which was not purpose built for the Irish energy system.
+The introduction of new climate policies and targets combined with technological advancements pushed the need for a new model, however the need for a transparent, flexible and robust model, purpose built for the Irish energy system by Irish energy modellers was a key driver to develop a new model.The new ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
 \par TIMES is a bottom-up optimisation energy-environment model with perfect foresight which is used to analyse various levels of spatial, temporal and sectoral resolution.
 Exogenous technical, physical, and regulatory constraints can be applied and technologies have specified fuel types, efficiency, lifetime, availability, environmental characteristics, and fixed, variable and operation costs.
-TIM is set up with regional options, either one national region or 26 county regions, only the annual results were explored in this study.
 TIMES uses a linear programming optimizer matrix in General Algebraic Modeling System (GAMS), to compute all calculations.
@@
 Weekly meetings were combined with intensive daily ``sprint" meetings to keep model milestones on track.
-The transparency of TIM improved the iterative development process, external energy experts and other stakeholders contributed using web-based dashboards \footnote{https://tim-review1.netlify.app/results/}.
-The schematic in Figure \ref{fig:TIMOverviewRES} shows the main sectors of the model, represented in the demand-side modules and how they link with the supply-side modules, to create a full picture of the national energy system \textbf{SOURCE}.
+The transparency of TIM improved the iterative development process, external energy experts and other stakeholders contributed using web-based dashboards\footnote{https://tim-review1.netlify.app/results/}.
+The schematic in Fig.
+\ref{fig:TIMOverviewRES} shows the main sectors of the model, represented in the demand-side modules and how they link with the supply-side modules, to create a full picture of the national energy system \textbf{SOURCE}.
 
@@
 
+TIM is set up with regional options, either one national region or 26 county regions, and with different temporal resolution options, only the national annual results were explored in this study.
+
 \subsection{TIM: Residential Sector}
@@
 The residential sector was developed with awareness of the ongoing national retrofitting and electrification strategies and investments.
-The residential energy services were developed based upon regulation (EU) 431/2014, the BER database and EU databases.
-Figure \ref{fig:RES} illustrates the flow of energy from fuel to demand in the residential sector.
+The residential energy services were developed based upon regulation (EU) 431/2014, shown in Fig.
+\ref{fig:IrelandServiceDemands}, however there is no space cooling data available in Ireland and lighting \& appliances are separated in TIM, while ``others" was combined with appliances.
+An overview of the flow of energy from fuels to demand, through the energy services is shown in Fig.
+\ref{fig:RES}.
 
@@
 The BER database provides the base year bottom-up energy consumption data, this is scaled up to represent the entire stock and cross-checked with the top-down energy consumption from the 2018 SEAI Energy Balance.
-To compare from the two databases, fuels were were combined and condensed down to 13 fuels in TIM.
-Table \ref{AppTable:Fuels} shows an overview of the fuels combined from the different data sources.
-The full list of space and water heating technologies considered in the model can be found in Table \ref{AppHeatingTech}.
-
-\par In TIM, fuels such as ``coal", represent a combination of fuels from the SEAI Energy Balance and BER database.
+To compare the two databases, fuels were were merged and separated, with 13 fuels adequately covering the residential sector in TIM, Table \ref{AppTable:Fuels} summarizes the fuels from the different data sources.
+In TIM, fuels such as ``coal", represent a combination of fuels from the SEAI Energy Balance and BER database.
 Gasoil which has similar properties to kerosene, is accounted for in the SEAI Energy Balance,  but there is no evidence to support keeping gasoil and kerosene separate would provide further insight into residential decarbonisation, therefore these fuels are combined in TIM.
@@
 Non-Archetype data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{Meth:Data}.
-The four archetype energy services are dependent on  building type and data is sourced from the BER database, which is discussed in Section \ref{Meth:Data}.
-The internal temperature adjustment for space heating is also detailed in Section \ref{Meth:Data}.
-The demand for archetype and non-archetype energy services is based on the growth of residential dwellings, which is detailed in section \ref{Meth:Data}.
+The four archetype energy services and the internal temperature adjustment are dependent on  building type and data is sourced from the BER database, which is discussed in Section \ref{Meth:Data}.
+The demand for archetype and non-archetype energy services is based on the growth of residential dwellings, which is also detailed in Section \ref{Meth:Data}.
 
@@
 %
-The dwelling stock data is obtained from the SEAI BER Public database \footnote{https://ndber.seai.ie/BERResearchTool/ber/search.aspx}, which is firstly filtered, then scaled up using Central Statistics Office (CSO) data.
+The dwelling stock data is obtained from the SEAI BER Public database\footnote{https://ndber.seai.ie/BERResearchTool/ber/search.aspx}, is filtered then scaled up using Central Statistics Office (CSO) data.
 The SEAI BER public database is Ireland's national Energy Performance Certificate (EPC) register,  which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD).
 The BERs are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$).
-\par Over half of Irish dwellings are captured in the BER database and every BER displays over 200 energy related variables.
+Over half of all Irish dwellings are captured in the BER database and every BER displays over 200 energy related variables.
 The DEAP model which is used to assign BER labels,  is outlined by \citet{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM}.
-In terms of space heating, DEAP assumes that each dwelling is single zoned, the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season, which is from October to May inclusive and assumes monthly external temperatures, the average external temperature for the heating season is 7.6$^{\circ}$C.
-One assumption, the internal temperature assumption, results in a large error, where space and water heating are overestimated in DEAP.
-Space heating consumption is reduced by 26\% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance.
+In terms of space heating, DEAP assumes that each dwelling is single zoned, the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season, which is from October to May inclusive and assumes monthly external temperatures, the average external temperature assumed for the heating season is 7.6$^{\circ}$C.
+However, one assumption, the internal temperature assumption, results in a large unsupportable error, where space heating is overestimated by 26\%, this is a simliar finding to \citet{Dineen2015ImprovedSource} who found a 7.5\% change in energy consumption for every 1$^{\circ}$C shift in internal temperatures.
+After internal temperature adjustments in Table \ref{Table:InternalTemps} were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance.
 DEAP assumes that living areas ( living room, sitting room, and possibly kitchen) in all dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C.
-Through the analysis of previous residential modelling in Ireland this is not the case, as the internal temperatures are often much lower, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions applied in TIM.
+Through the analysis of previous residential modelling in Ireland internal temperatures are often much lower than 21$^{\circ}$C in living areas and 18$^{\circ}$C in non-living areas \cite{Dineen2015ImprovedSource,Healy2002FuelIreland,Clinch2003ValuingModel,Clinch2000HousingMortality}.
 
@@
 ArDEM-SQL alters the primary energy consumption of the BER database by changing the internal temperatures assumed by DEAP in Table \ref{Table:InternalTemps}.
-
 Other concerns for uncertainties in the database, which are not related to DEAP calculation are statistically anomalous spikes in the data, caused by applying default values \cite{Ahern2020ADatabases} and different companies generating variations of ±20\% in the assessment of energy consumption \cite{Harsman2016OnCertificates}.
@@
 The filters applied to the raw BER database are outlined in Table \ref{AppBERFilters}.
-
-The Census 2016 data \cite{CentralStatisticsOffice2016CensusIreland} from the CSO is also used in calculating the dwelling stock.
-Dataset E1005 provides a split of seven building archetypes per county, which are aggregated to form three building archetypes.
+%PICKUPHERE
+The Census 2016 data \cite{CentralStatisticsOffice2016CensusIreland} from the CSO is also used in calculating the dwelling stock, dataset E1005 provides a split of seven building archetypes per county, which are aggregated to form three building archetypes in TIM.
 The total number of dwellings per archetype per county in 2016 is used to scale up the BER database.
 The gap in total dwellings between the census year in 2016 and the model base year in 2018 was initially ignored due to lack of data.
-However, later it was found that data on new dwelling electrical connections per archetype and per county are available from CSO dataset NDQ06, this would have filled the gap, the first iteration of the dwelling stock in TIM has a 2\% error.
+However, later it was found that data on new dwelling electrical connections per archetype and per county are available from CSO dataset NDQ06, this would have filled the gap, the first iteration of the dwelling stock in TIM does not account for the 2\% of existing dwellings build in 2017 and 2018.
 \par Once the BER database and CSO database tables have been filled to include dwellings per archetype per county, the next step was to scale up the BER database.
@@
 Where $TIM_{r,y,a}$ is the total dwellings in TIM by BER rating (r), year of construction (y), and archetype (a).
-$CSO_{y,a}$ is the number of dwellings by year of construction and archetype in the CSO dataset, the CSO dataset does not contain BER rating types.
-$BER_{r,y,a}$ is the number of BER ratings by year of construction and archetype from the BER database and dividing this value by $BER_{y,a}$ results in a probability of a certain age type falling within a BER rating, and the final part of the equation $\frac{TIM_{a}} {BER_{a}}$ compensates for the non-stated archetypes in the CSO data, $TIM_a$ was already calculated in equation \ref{Eq:BERtoCSO2}.
-The results of each of the correlations are in  Table \ref{AppBERCor},\ref{AttBERCor} \& \ref{DetBERCor} and Table \ref{Table:FinalDwellingStock} outlines the final dwelling stock used in TIM.
+$CSO_{y,a}$ is the number of dwellings by year of construction and archetype in the CSO dataset, no BER rating types exist in the CSO dataset.
+$BER_{r,y,a}$ is the number of BER ratings by year of construction and archetype from the BER database and dividing this value by $BER_{y,a}$ results in a probability of a certain age type falling within a BER rating, and the final part of the equation $\frac{TIM_{a}} {BER_{a}}$ compensates for the non-stated archetypes in the CSO data, $TIM_a$ was already calculated in Equation \ref{Eq:BERtoCSO2}.
+The results of each of the BER probablities are summarized for each archetype in  Table \ref{AppBERCor},\ref{AttBERCor} \& \ref{DetBERCor}, while Table \ref{Table:FinalDwellingStock} outlines the final dwelling stock used in TIM.
 
@@
 
-The costs applied in the residential sector, which includes space and water heating technologies, retrofitting, cooking, lighting, pumps, fans, space cooling, and other electrical appliances. 
-The fuel costs are not applied in the residential sector, there are applied in the supply sector, and the data is sourced from World Energy Outlook reports from 2019 \cite{/content/publication/caf32f3b-en} and 2020 \cite{/content/publication/557a761b-en}.
+The costs applied in the residential sector, includes space and water heating technologies, retrofitting, cooking, lighting, pumps, fans, space cooling, and other electrical appliances. 
 The majority of the space and water heating cost data is obtained from the
-\textit{``Techno-economic projections until 2050 for smaller heating and cooling technologies in the residential and tertiary sectors in the EU”} report \cite{JointResearchCentre2017Techno-economicEU}.
-The cost data used in TIM was based on the north EU region (Denmark) and where data was missing, the central EU region (Germany) cost data was used.
+\textit{``Techno-economic projections until 2050 for smaller heating and cooling technologies in the residential and tertiary sectors in the EU”} report \cite{JointResearchCentre2017Techno-economicEU}, the report has spatial cost data and the north EU region (Denmark) was mainly used \cite{JointResearchCentre2017Techno-economicEU}, this cost data is sourced from the continuously updated and expanding Danish Energy Agency's Technology Data for Individual Heat Production dataset and where data was missing, the central EU region (Germany) cost data was used \cite{JointResearchCentre2017Techno-economicEU}.
 The high granularity of data allows for the cost of a technology to be disaggregated between apartments and non-apartments, costs and efficiencies are projected every five years from 2020 to 2050.
 The lifetime and emissions/pollutants are also included, while the cost data is further disaggregated between capital expenditure ( equipment, installation \& additional capital costs ) and Operational \& Maintenance (O\&M) costs.
+The cost data for different electrical appliances is sourced from Topten EU database \cite{Michel2016EnergyREPORT} and SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
+Topten has a large database covering electrical appliances available in the European market, the data collected includes cost and efficiency of refrigeration, cloth washing, cloth drying, and dish washing, while some Topten reports have looked at sales and European market shares \cite{Michel2016EnergyREPORT}.
+Cooking data is sourced from Eurostat's ``Energy consumption in households by type and end-used"  \cite{Eurostat2021Disaggregatednrg_d_hhq} and cross-checked with SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
+Appliances not included in this list are simply under the ``Electrical Appliances" category in TIM.
+Electrical Appliances include television, multimedia, outdoor appliances, and computer technology and data is sourced from SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and Eurostat \cite{Eurostat2021Disaggregatednrg_d_hhq}.
+The fuel costs are not applied in the residential sector, there are applied in the supply sector, and the data is sourced from World Energy Outlook reports from 2019 \cite{/content/publication/caf32f3b-en} and 2020 \cite{/content/publication/557a761b-en}.
 \par TIM has two types of retrofitting options per archetype,  a shallow retrofit is when the primary energy ($kWh/m^2/yr$) is reduced by $ \leq 35\%$, and deep retrofit is where the primary energy is reduced by $>35\%$, while a reduction of $ \ >75\%$ is considered infeasible.
 The difference between the pre-retrofit and post-retrofit primary energy determines the type of retrofit, as shown in Table \ref{AppDepthofRetrofit}.
-
 The retrofit cost data was sourced from Ali et.
@@
 
-Where $ {\text{\euro}}{}_{rf-rt,a}$ is the retrofit cost from pre-retrofit BER rating (rf) to post-retrofit rating (rt) for archetype (a).
+Where $ {\text{\euro}}{}_{rf-rt,a}$ is the retrofit cost from pre-retrofit BER rating or rating from (rf) to post-retrofit rating or rating to (rt) for archetype (a).
 $Ali_{r,t}$ is the cost data from  \citet{Ali2020ABuildings}, it contains all BER ratings and retrofit types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments, 1 for attached, and 1.1 for detached dwellings.
@@
 
-\par The cost data for different electrical appliances is sourced from Topten EU database.
-Topten has a large database covering electrical appliances available in the European market, the data collected includes cost and efficiency.
-While some Topten reports have looked at sales and market shares \cite{Michel2016EnergyREPORT}.
-
-The previous paragraphs outlined the cost of retrofitting, and this following looks at the expected savings of retrofitting existing dwellings and building regulations in new dwellings.  
+The previous paragraphs outlined the cost of retrofitting and the expected savings of retrofitting in existing dwellings is now examined, along with building regulations savings in new dwellings.  
 The annual heating costs per archetype and per BER rating are sourced from SEAI's indicative energy running costs.
@@
 
-As TIM only inputs the savings per type of retrofit and archetype, some calculations must be performed on Table \ref{Table:FuelCost}.
+As TIM only inputs the savings per type of retrofit and archetype, some calculations must be performed on Table \ref{Table:FuelCost} to estimate the savings from retrofits.
 The first calculation is to find the savings from pre-retrofit running cost to post-retrofit running cost, as shown in Equation \ref{Eq:RetrofitSaving1}
@@
 Where $Saving(\%)_{a,rf-rt}$ are the savings ( or reduced running cost) per archetype (a), where the pre-retrofit BER rating is defined as the rating from (rf), and the post-retrofit BER rating is defined as the rating to (rt).
-
 Allocating each retrofit change to either shallow or deep is the next step, this is shown in Table \ref{AppDepthofRetrofit}.
@@
 \ref{Eq:RegArchetype} is used to calculate the share of an archetype in a given year (y).
-This is graphically represented in Figure \ref{fig:DwellingStoc}.
+This is graphically represented in Fig.
+\ref{fig:DwellingStoc}.
 The growth rate of apartments is the highest and increasing up to 2040, while the growth in attached and detached dwellings are lagging behind.
@@
 
+The base year heating energy consumption is derived from the BER database, but uses census data \cite{CentralStatisticsOffice2016CensusIreland} from the CSO dataset E1008 to calibrate the fuel used in heating systems per archetype.
 Space heating, water heating, and lighting energy service demands for existing dwellings are calculated from the average values of the base year dwelling stock, and as all new dwellings are required to be A-rated, the energy service demand for new dwellings in TIM is calculated from the average of the existing A-rated dwellings.
-The Pumps \& Fans energy demands are calculated in a similar method, except in the new dwellings the value is halved and the remaining half is added to space cooling.
-This is an assumption which was cross-checked with POTEnCIA (Policy Oriented Tool for Energy and Climate Change Impact Assessment) space cooling projections per household.
-
-The demand for space and water heating is outlined in the previous section, and this section explains how water and space heating are achieved through technology and retrofits.
+The pumps \& fans energy demands are calculated in a similar method, except in the new dwellings the value is halved and the remaining half is added to space cooling, this aligns with POTEnCIA's (Policy Oriented Tool for Energy and Climate Change Impact Assessment) space cooling and pump \& fan projections per household.
 While the demands are exogenous, the choice of technologies to at least satisfy the demand are endogenous and is captured in Eq.
@@
 DM is the exogenous demand per energy service per year.
-The cost of the technology used is minimised while also satisfying all constraints.
-
+The cost of the technology is minimised while also satisfying all constraints.
 All space and water heating technologies are archetype specific and some technologies deliver both water and space heating, while other technologies can only deliver one type of heating, the list of available technologies is outlined in Table \ref{AppHeatingTech}.
 Retrofits are included as they satisfy the residential space heating demand, even though retrofits do not consume fuel similar to a solar water heating system, from that perspective.
-Danish Energy Agency's Technology Data for Individual Heat Production dataset was used  was input to TIM \cite{JointResearchCentre2017Techno-economicEU}, this dataset is also used in the EU TIMES model, the dataset has high granularity and is continuously updated.
-
-In TIM, the number of main heating systems equals the number of dwellings, this assumption ignores any impact of secondary system heating.
-One exception to the rule is stoves are considered as secondary heating systems only.
-The base year heating energy consumption is derived from the BER database, but uses census data \cite{CentralStatisticsOffice2016CensusIreland} from the CSO dataset E1008 to calibrate the fuel used in heating systems per archetype.
-
-As shown in Figure \ref{fig:IrelandServiceDemands}, these energy services accounted for 20\% of energy consumption in 2018, however, this share of consumption is expected to increase. 
-Refrigeration, cloth washing, cloth drying, and dish washing are under the appliances category and data is sourced from Topten \cite{Michel2016EnergyREPORT} and SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
-Cooking data is sourced from Eurostat's ``Energy consumption in households by type and end-used"  \cite{Eurostat2021Disaggregatednrg_d_hhq} and cross-checked with SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
-\par Appliances not included in this list are simply under the ``Other Appliances" category in TIM.
-Other Appliances include television, multimedia, outdoor appliances, and computer technology.
-Lighting and pumps \& fans are archetype specific energy services as the data is sourced from the BER database.
-\par Equation \ref{Eq:Technology} is also used to optimal choose the technology in the energy service demand, but unlike in the heating energy service demands, there is little or no choice in different technologies (k) for some of the demand categories (i), therefore, the model is restricted in lighting, appliances and pumps \& fans.
+In Appliances  unlike in the heating energy service demands, there is little or no choice in different technologies (k) for some of the demand categories (i), therefore, the model is restricted in lighting, appliances and pumps \& fans.
 However, the cost and efficiency of each technology does determine the overall investment.
@@
 
-The quantity and mix of fuel going into the residential for each scenario and year is  shown in Figure \ref{fig:ResFuel}.
+The quantity and mix of fuel going into the residential for each scenario and year is  shown in Fig.
+\ref{fig:ResFuel}.
 The base year is shown on the left, while annual snapshots of the fuel-mix for the other scenarios are shown for 2030, 2040, and 2050.
@@
 In both scenarios, the carbon intensity of electricity ($gCO_2/kWh$) falls by two-thirds, with the Current Effort scenario only marginally lower than the Previous Effort scenario.
-New emerging fuels which are prevalent in the Current Effort include space ambient heat and water ambient heat due to the increase of heat pumps in the residential sector, more on this in Figure \ref{fig:HeatPump}.
-
-The optimal cumulative number of retrofits by type and quantity up to 2030, 2040, and 2050 are summarised in Figure \ref{fig:Retrofits}.
+New emerging fuels which are prevalent in the Current Effort include space ambient heat and water ambient heat due to the increase of heat pumps in the residential sector, more on this in Fig.
+\ref{fig:HeatPump}.
+
+The optimal cumulative number of retrofits by type and quantity up to 2030, 2040, and 2050 are summarised in Fig. \ref{fig:Retrofits}.
 
@@
 The model assumes the lifetime of a Heat Pump is 20 years, so it maybe possible for one dwelling to get two new heat pumps over the next 30 years, unlike retrofits, heat pumps can be applied to the new dwellings.
-In Figure \ref{fig:HeatPump} the total cumulative number of heat pumps installed up to a given year is shown.
+In Fig.
+\ref{fig:HeatPump} the total cumulative number of heat pumps installed up to a given year is shown.
 
@@
 
-In Figure \ref{fig:HeatPump}, SW represents space and water heating, HC is space heating and space cooling, and A-to-W represents air-to-water type heat pumps.
+In Fig.
+\ref{fig:HeatPump}, SW represents space and water heating, HC is space heating and space cooling, and A-to-W represents air-to-water type heat pumps.
 The optimal number of heat pumps in each scenario ranges from 39,000 to 938,000, which equates to 2-45\% of dwellings having a heat pump by 2030 depending on which scenario is used.
@@
 The advantage of an energy system model is to observe the intersectoral connections.
-Figure \ref{fig:SectorFuel} presents the base year primary energy fuel-mix for four sectors - power generation, transport, services, and residential in 2030, 2040, and 2050 for both scenarios.
+Fig.
+\ref{fig:SectorFuel} presents the base year primary energy fuel-mix for four sectors - power generation, transport, services, and residential in 2030, 2040, and 2050 for both scenarios.
 
